### 2018-01-21 MEM.tex
Initial commit.

\documentclass{article}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{blkarray}
\usepackage[font=small]{caption}
\usepackage{cite}
\usepackage{graphicx}

% ---- Author affiliations ---- %

\renewcommand\Affilfont{\itshape\small}

% ---- Propositions, lemmas, defintions... ---- %

\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}
%\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
%\newtheorem{remark}{Remark}

% ---- Special environments (examples and remarks) ---- %

\newcounter{examplecounter}
\newenvironment{example}
{\small\vspace{0.5\baselineskip}
  \refstepcounter{examplecounter}%
  \noindent\textbf{Example \arabic{examplecounter}.}%
}{\vspace{-0.2\baselineskip}\begin{center}%
  $\star$\end{center}\vspace{0.5\baselineskip}}

\newcounter{remarkcounter}
\newenvironment{remark}
{\small\it\vspace{0.5\baselineskip}
  \refstepcounter{remarkcounter}%
  \noindent\textbf{Remark \arabic{remarkcounter}.}%
}{\vspace{0.5\baselineskip}}

\newenvironment{inset}
{\vspace{0.5\baselineskip}\begin{center}}
{\end{center}\vspace{0.5\baselineskip}}

% ---- Macros ---- %

\newcommand{\smA}{\scriptstyle{A}}
\newcommand{\smC}{\scriptstyle{C}}
\newcommand{\smD}{\scriptstyle{D}}
\newcommand{\smE}{\scriptstyle{E}}
\newcommand{\smG}{\scriptstyle{G}}
\newcommand{\smI}{\scriptstyle{I}}
\newcommand{\smS}{\scriptstyle{S}}
\newcommand{\smT}{\scriptstyle{T}}
\newcommand{\smDELz}{\scriptstyle{\Delta_0}}
\newcommand{\smDELs}{\scriptstyle{\Delta_*}}

%---------------------------------------------------------------

\title{Calibrating MEM seeding heuristics}

\author[1,2]{Guillaume J. Filion}
\affil[1]{Genome Architecture, Gene Regulation, Stem Cells and Cancer
Programme, Center for Genomic Regulation (CRG), The Barcelona Institute of
Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
\affil[2]{University Pompeu Fabra, Doctor Aiguader, 08003 Barcelona,
Spain.}

\date{\today}

%---------------------------------------------------------------
%---------------------------------------------------------------

\begin{document}

\maketitle

\begin{abstract}
The abstract will come later.
\end{abstract}

%---------------------------------------------------------------
%---------------------------------------------------------------

\section{Introduction}
Some introduction

\section{MEM seeds}

We can describe reads as sequences of four possible symbols representing
the state of the local match with the genome. The first symbol,
represented by a black square indicates a sequencing error. This is a
mismatch between the read and the genome so the nucleotide cannot belong
to a correct MEM seed. It either belongs to an incorrect MEM seed, or it
marks the end of a correct MEM seed.

The second symbol, represented by a black square with a white cross
indicates a sequencing error that is also a mismatch for \emph{all} the
duplicates. The nucleotide cannot belong to any MEM seed (correct or not)
so the next MEM seed must start at the following nucleotide. The symbol
`resets' the seeding process, which becomes in the same state as at the
beginning of the read.

The third symbol, represented by an exclamation mark indicates that the
target sequence is the only local match between the read and the genome.
This means that there is at least one mismatch between the read and any
duplicate since the beginning of the current MEM. From that nucleotide,
the MEM will extend forward until the next sequencing error. This means
that two exclamation marks must be separated by either a black square or a
black square with a cross.

The last symbol, represented as a white square indicates all the other
situations. This can mark the beginning but not the end of a MEM.

\begin{figure}[h]
\centering
\includegraphics[scale=.9]{sketch_MEM.pdf}
\caption{\textbf{Reads and MEM seeds}. 
Some legend.}
\label{fig:sketch_MEM}
\end{figure}

\begin{equation*}
M(z) =
\begin{blockarray}{ccccccc}
   & \scriptstyle{\blacksquare} & \scriptstyle{\times} 
   & \scriptstyle{(!)_1} & \scriptstyle{(!)_2}
   & \ldots & \scriptstyle{(!)_{\gamma-1}} \\
\begin{block}{c[cccccc]}
\scriptstyle{\blacksquare} & \tilde{A}_\infty(z) & \tilde{B}_\infty(z) &
    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
\scriptstyle{\times} & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) &
    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
\scriptstyle{(!)_1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & 0 &
    \ldots & 0 \\
\scriptstyle{(!)_2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\scriptstyle{(!)_{\gamma-1}} & A_1(z) & B_1(z) & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

\begin{eqnarray*}
A_n(z) &=& p (1-(1-\kappa/3)^N) \sum_{m=0}^{n-1} q^mz^{m+1}, \\
\tilde{A}_n(z) &=& p (1-(1-\kappa/3)^N)
    \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
B_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} q^mz^{m+1}, \\
\tilde{B}_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
r_n(z) &=& (C_{n-1}-C_n) (qz)^n, \\
F_n(z) &=& \sum_{m=0}^n C_m (qz)^m.
\end{eqnarray*}

Define $C_m = \big( 1- (1-(1-\kappa)^m)^N \big)$ as the probability that
during $m$ nucleotides, not all duplicates have collapsed.

\section{Conclusion}
Some conclusion.

\section*{Acknowledgements}

I acknowledge the financial support of the Spanish Ministry of Economy and
Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan
Nacional BFU2012-37168), of the CERCA Programme~/~Generalitat de
Catalunya, and of the European Research Council (Synergy Grant 609989).

%---------------------------------------------------------------
%---------------------------------------------------------------

\bibliography{references,pubmed}
\bibliographystyle{plain}

%----------------------------------------------------------------

\end{document}

%gs -dNoOutputFonts -sDEVICE=pdfwrite -o out.pdf latex.pdf

### 2018-01-21 C/sim.c

@@ -1,0 +1,123 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+#include <time.h>
+
+#include "mt.h"
+
+// Global variables.
+unsigned int K;
+unsigned int GAMMA;
+unsigned int NDUP;
+
+
+void
+update_dup
+(
+         unsigned int * dup,
+   const unsigned long int mt_match
+)
+{
+   for (int j = 0 ; j < NDUP ; j++)
+      dup[j] = randomMT() < mt_match ? dup[j] + 1 : 0;
+   return;
+}
+
+
+int
+there_is_a_seed
+(
+   const unsigned int   target,
+   const unsigned int * dup
+)
+{
+   if (target < GAMMA) return 0;
+   for (int j = 0 ; j < NDUP ; j++)
+      if (dup[j] >= target) return 0;
+   return 1;
+}
+
+
+int
+simulate_read
+(
+   unsigned long int   mt_error,
+   unsigned long int   mt_match1,
+   unsigned long int   mt_match2,
+   unsigned int      * dup
+)
+{
+   int read_error;
+
+   // Initialize counters.
+   bzero(dup, NDUP * sizeof(int));
+   int target = 0;
+
+   for (int i = 0 ; i < K ; i++) {
+
+      read_error = randomMT() < mt_error;
+      update_dup(dup, read_error ? mt_match2 : mt_match1);
+
+      // The code of 'there_is_a_seed()' works for both strict and
+      // shared seeds, but we need to update 'target' before when
+      // there is no error, and after when there is one.
+      if (!read_error) target++;
+      if (there_is_a_seed(target, dup)) return 1;
+      if (read_error) target = 0;
+
+   }
+
+   // Check if there is a shared seed at the end of the read.
+   if (target >= GAMMA && !read_error)
+      return there_is_a_seed(target+1, dup);
+
+   return 0;
+
+}
+
+
+
+int main(int argc, char **argv) {
+
+                K     = atoi(argv[1]);            // Read size.
+                GAMMA = atoi(argv[2]);            // Seed length.
+   double       prob  = strtod(argv[3], NULL);    // Read error.
+   double       kappa = strtod(argv[4], NULL);    // Divergence.
+                NDUP  = atoi(argv[5]);            // Duplicate number.
+   long int     ITER  = atoi(argv[6]);
+
+   // Check arguments.
[…]

### 2018-04-06 MEM.tex
Clearing the mess.

@@
 
-\newcommand{\smA}{\scriptstyle{A}}
-\newcommand{\smC}{\scriptstyle{C}}
-\newcommand{\smD}{\scriptstyle{D}}
-\newcommand{\smE}{\scriptstyle{E}}
-\newcommand{\smG}{\scriptstyle{G}}
-\newcommand{\smI}{\scriptstyle{I}}
-\newcommand{\smS}{\scriptstyle{S}}
-\newcommand{\smT}{\scriptstyle{T}}
-\newcommand{\smDELz}{\scriptstyle{\Delta_0}}
-\newcommand{\smDELs}{\scriptstyle{\Delta_*}}
+\newcommand{\DN}{\scriptstyle{\Downarrow}}
+\newcommand{\dn}{\scriptstyle{\downarrow}}
+\newcommand{\up}[1]{\scriptstyle{(\uparrow)_{#1}}}
+\newcommand{\eq}[1]{\scriptstyle{(\sim)_{#1}}}
+
+\newcommand{\upg}{\scriptstyle{\uparrow}}
+\newcommand{\eqg}{\scriptstyle{\sim}}
 
@@
 
+\section{Some definitions etc.}
+
 \begin{figure}[h]
@@
 
+We assume that the target sequence has $N$ duplicates that are potential
+false positives for the seeding process. We further assume that
+duplication was instantaenous and that all $N+1$ sequences diverge
+independently from each other at a constant rate. In other words, we
+ignore the complications due to the genealogy of the duplication events
+and we simply assume that at each position, the target and a duplicate are
+identical with probability $1-\mu$. In case they are not, we assume that
+the duplicate has any of the remaining three nucleotides with equal
+probability.
+
+Each of the $N+1$ sequences of interest corresponds to a Bernoulli process
+where `success' stands for a match between the sequence and the read, and
+`failure' stands for a mismatch. To describe those processes, we will use
+the `match function' $\varphi(n)$ that gives the length of the success
+streak at position $n$. The processes are indexed such that $\varphi_0$
+describes the matches of the target sequence, whereas $\varphi_j$ $(1 \leq
+j \leq N)$ describes the matches of the duplicates. For instance,
+$\varphi_0(5) = 5$ and $\varphi_1(5) = 3$ in the example of
+Fig.~\ref{fig:sketch_MEM}.
+
+For simplicity, we will refer to the Bernoulli processes as `threads', and
+we will distinguish the `target thread' from the `duplicate threads'. At
+any position $n$ of the read, a duplicate thread $j$ $(1 \leq j \leq N)$
+is said to be `strictly masking' if $\varphi_j(n) > \varphi_0(n)$,
+`potentially masking' if $\varphi_j(n) \geq \varphi_0(n)$ and
+`non-masking' if $\varphi_j(n) < \varphi_0(n)$. There can be no MEM seed
+at positions where there is a stricly masking thread. At positions where
+there is a potentially masking thread any MEM seed must be a shared MEM
+seed.
+
+With these definitions, the decision tree for choosing the next symbol to
+insert in a read can be summarized as in Fig.~\ref{fig:decision_tree}.
+This sketch will be useful as a reference to compute the probabilities of
+the blocks and their weighted generating functions.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.7]{decision_tree.pdf}
+\caption{\textbf{Decision tree}. 
+Some legend.}
+\label{fig:decision_tree}
+\end{figure}
+
 \begin{equation*}
-M(z) =
-\begin{blockarray}{ccccccc}
-   & \scriptstyle{\blacksquare} & \scriptstyle{\times} 
-   & \scriptstyle{(!)_1} & \scriptstyle{(!)_2}
-   & \ldots & \scriptstyle{(!)_{\gamma-1}} \\
-\begin{block}{c[cccccc]}
-\scriptstyle{\blacksquare} & \tilde{A}_\infty(z) & \tilde{B}_\infty(z) &
-    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
-\scriptstyle{\times} & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) &
-    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
-\scriptstyle{(!)_1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & 0 &
+\begin{blockarray}{ccccccccc}
+   & \dn & \DN &\eq{1} & \ldots & \eq{\gamma-1} &
+    \up{1} & \ldots & \up{\gamma-1} \\
+\begin{block}{c[cccccccc]}
+\dn & * & * & * & \ldots & * & * & \ldots & * \\
+\DN & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) & 0 & \ldots & 0 &
+    r_1(z) & \ldots & r_{\gamma-1}(z) \\
+\eq{1} & * & * & 0 & \ldots & 0 & 0 & \ldots & G_{1,\gamma-1}(z) \\
+\eq{2} & * & * & 0 & \ldots & 0 & 0 & \ldots & G_{2,\gamma-1}(z) \\
+\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
+    \ddots & \vdots \\
+\eq{\gamma-1} & * & * & 0 & \ldots & 0 & 0 & \ldots & 0 \\
+\up{1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
     \ldots & 0 \\
-\scriptstyle{(!)_2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & 0 &
+\up{2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
     \ldots & 0  \\
-\vdots & \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
-\scriptstyle{(!)_{\gamma-1}} & A_1(z) & B_1(z) & 0 & 0 & \ldots & 0 \\
+\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
+    \ddots & \vdots \\
+\up{\gamma-1} & A_1(z) & B_1(z) & 0 & \ldots & 0 & 0 & \ldots & 0 \\
 \end{block}
@@
 
+\begin{definition}
+The main thread is dominant at nucleotide $n$ if $\varphi_0(n) >
+\varphi_i(n)$ for all $i$ $(1 \leq i \leq N)$. The main thread is
+co-dominant if inequality is not strict and there exists $i$ $(1\leq i
+\leq N)$ such that $\varphi_0(n) = \varphi_i(n)$.
+\end{definition}
+
+\subsection{Head vector}
+
+At the beginning of the read, the match lengths of all the threads are
+$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
+consider that the read starts in the $\Downarrow$ state, which is
+equivalent to adding an edge labelled with the empty object $\varepsilon$
+between the head vertex and the $\Downarrow$ vertex. The head vector is
+thus
+\begin{equation}
+H(z) = (1,0, \ldots, 0).
+\end{equation}
+
+\subsection{Transitions from $\uparrow$}
+\label{sec:trans_from_up}
+
+After a $\uparrow$ symbol, the main thread becomes dominant. It will
+remain so until the first read error, which can be a mismatch for all the
+threads ($\Downarrow$ symbol), or for only some of them ($\downarrow$
+symbol).
+
+Since the read has no MEM seed, the $\Downarrow$ symbol or the
+$\downarrow$ symbol must come before $\varphi_0(n) \geq \gamma$, which
+imposes a limit on the number of $\square$ symbols that can follow a
+$\square$ symbol. If nucleotide at position $n$ is the $\uparrow$ symbol
+and $\varphi_0(n) = j$ $(1 \leq j \leq \gamma-1)$, the state of the read
+is represented as $(\uparrow)_j$, meaning that the maximum number of
+$\square$ symbols that can follow is $\gamma-1-j$.
+
+Whether the nucleotides match the alternative threads is irrelevant until
+the end of the block. The $\square$ symbols only mean that the nucleotides
+are not error so their weighted generating function is $qz$. The
+$\Downarrow$ symbol indicates an error -- probability $p$ -- that is also
+a mismatch for the $N$ alternative threads -- probability $(1-\mu/3)^N$ --
+so its weighted generating function is $pz(1-\mu/3)^N$. Likewise, the
+weighted generating function of the $\downarrow$ symbol is seen to be
+$pz(1-(1-\mu/3)^N)$.
+
+The weighted generating function of the transition from $(\uparrow)_j$ to
+$\Downarrow$ is
+\begin{equation}
+B_{\gamma-j}(z) = pz (1-\mu/3)^N \sum_{n=0}^{\gamma-j-1} (qz)^n.
+\end{equation}
+
+And the weighted generating function of the transition from
+$(\uparrow)_j$ to $\downarrow$ is
+\begin{equation}
+A_{\gamma-j}(z) = pz (1-(1-\mu/3)^N) \sum_{n=0}^{\gamma-j-1} (qz)^n.
+\end{equation}
+
+\subsection{Transitions from $\Downarrow$}
+\label{sec:trans_from_Down}
+
+At a $\Downarrow$ symbol, all the threads have match length $0$ and the
+read is `reset' to its initial state. The main thread is co-dominant with
+all the alternative threads. The next terminator symbol depends on what
+happens first. If all the alternative threads fail before the main thread,
+the latter becomes dominant ($\uparrow$ symbol). If not all alternative
+threads have failed when the main thread does, we need to distinguish two
+cases: if all threads fail together with the main thread, the read is
+reset ($\Downarrow$ symbol), otherwise the main read becomes dominated
+($\downarrow$ symbol).
+
+If the next terminator is the $\uparrow$ symbol, we need to distinguish
+the cases based on the value of $\varphi_0$ (see
+section~\ref{sec:trans_from_up}), which is just the position of
+the $\uparrow$ symbol after the $\Downarrow$ symbol in this case. A
+transition from $\Downarrow$ to $(\uparrow)_j$ consists of a succession of
+$j$ correct nucleotides -- each with probability $q$. But we must also
+make sure that no other $\uparrow$ symbol is inserted before position $j$.
+We thus introduce $\xi_j$, the probability that at least one of the $N$
+alternative threads survives at position $j$ after the $\Downarrow$ symbol. At
+any correct nucleotide, the probability of failure of an alternative thread is
+$\mu$, so
+\begin{equation}
+\xi_j = 1-(1-(1-\mu)^j)^N.
+\end{equation}
+
+The probablity of transition from $\Downarrow$ to $(\uparrow)_j$ is thus
+$(\xi_{j-1}-\xi_j)q^j$ and the associated weighted generating function is
+\begin{equation}
+r_j(z) = (\xi_{j-1}-\xi_j)(qz)^j.
+\end{equation}
+
+If the next terminator is $\downarrow$, there is no limit on the number of
+$\square$ symbols that can be inserted. Indeed, the main thread will not
+be part of the MEM in this case. As mentioned in
+section~\ref{sec:trans_from_up}, the weighted generating function of the
+$\downarrow$ symbol is $pz(1-(1-\mu/3)^N)$. The probability that it first
+occurs at position $n+1$ after the $\Downarrow$ symbol is the probability
+that the first $n$ nucleotides are correct -- each with probability $q$
+-- and that at least one alternative thread survives -- probability
+$\xi_n$. The weighted generating function of the transition from
+$\Downarrow$ to $\downarrow$ is thus
+\begin{equation}
+\tilde{A}_\infty(z) = pz(1-(1-\mu/3)^N)
+  \sum_{n=0}^\infty \xi_n \cdot (qz)^n.
+\end{equation}
+
+Finally, if the next terminator is $\Downarrow$, there is a limit on the
+number of $\square$ symbols. Indeed, if there are $\gamma$ or more
+nucleotides between the two $\Downarrow$ symbols, they constitute a shared
+MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
+generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
+same rationale as above, the weighted generating function of the
+transition from $\Downarrow$ to $\Downarrow$ is
+\begin{equation}
+\tilde{B}_{\gamma}(z) = pz(1-\mu/3)^N
+  \sum_{n=0}^{\gamma-1} \xi_n \cdot (qz)^n.
+\end{equation}
+
+\subsection{Transitions from $\sim$}
+
+At a $\sim$ symbol, the main thread becomes co-dominant, together with $1$
+to $N-1$ alternative threads. As in section~\ref{sec:trans_from_Down}, the
+next terminator symbol depends on whether all the alternative co-dominant
+threads fail before the main thread. If this is the case, the main thread
+will become dominant ($\uparrow$ symbol), otherwise we again need to
+distinguish the case that the read is reset ($\Downarrow$ symbol) from the
+case that the main thread becomes dominated ($\downarrow$ symbol).
+
+An important difficulty is that the $\sim$ symbol does not inform us about
+the number of surviving threads (except when $N = 1$). So we need to
+marginalize the probabilities of interest over this number. The first
+thing to note is that a $\sim$ terminator can only follow a $\downarrow$
+terminator. Indeed, after a $\Downarrow$ symbol the threads are already
+co-dominant, and after a $\uparrow$ symbol the alternative threads cannot
+`catch up' with the main thread until a read error occurs, \textit{i.e.}
+until a $\Downarrow$ symbol or a $\downarrow$ symbol is inserted. The
+number of surviving threads depends on the position of the $\sim$ symbol
+relative to the preceding $\downarrow$ symbol, so as in
+section~\ref{sec:trans_from_up} we need to distinguish states based on the
+value of $\varphi_0$ at the position of the $\sim$ symbol.
+
+If nucleotide at position $n$ is $\sim$ and $\varphi_0(n) = j$ $(1 \leq j
+\leq \gamma-1)$, the state of the read is represented as $(\sim)_j$.
+
+\subsection{Transitions from $\downarrow$}
+
+Like... more complicated.
+
+By conditioning on how many duplicate threads fail at the $\dn$ symbol, we
+can compute the probability that a $\eqg$ symbol occurs at position $k$
+after an error as $q^kE_k / (1-(1-\mu/3)^N)$, where
+
+\begin{equation*}
+\begin{split}
+E_k = \sum_{m=1}^{N-1} {N \choose m} (1-\mu/3)^m (\mu/3)^{N-m} \times
+\Big( 1 - (1-(1-\mu)^k)^m \Big) \times \\
+\Big( (1-(1-\mu)^k)^{N-m} - (1-(1-\mu)^{k-1})^{N-m} \Big).
+\end{split}
+\end{equation*}
+
+\begin{equation*}
+\begin{split}
+E_k = [1-\mu/3(1-\mu)^k]^N - [1-\mu/3(1-\mu)^{k-1}]^N
+-[1-(1-\mu)^k]^N + \\
+[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1}]^N.
+\end{split}
+\end{equation*}
+
+The calculation above says that $m$ threads fail at the $\dn$ symbol, that
+they do not all fail in the following $k$ nucleotides, and that the last
+survivor of the $\dn$ symbol fails at nucleotide $k$. Note that $E_k=0$
+for $N=0$ and $N=1$, as expected. To find the probability, we need to
+multiply by $q^k$ to account for the fact that there is no error in $k$
+nucleotides, and we need to divide by $(1-(1-\mu/3)^N)$ to account for the
+fact that not all threads can fail at symbol $\dn$ (otherwise it would be
+symbol $\DN$). The sum does not include the case $m=0$ because in this
+case, none of the duplicate threads can have the same length as the main
+thread, so there cannot be a $\eqg$ symbol.
+
 \begin{eqnarray*}
-A_n(z) &=& p (1-(1-\kappa/3)^N) \sum_{m=0}^{n-1} q^mz^{m+1}, \\
-\tilde{A}_n(z) &=& p (1-(1-\kappa/3)^N)
-    \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
-B_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} q^mz^{m+1}, \\
-\tilde{B}_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
-r_n(z) &=& (C_{n-1}-C_n) (qz)^n, \\
-F_n(z) &=& \sum_{m=0}^n C_m (qz)^m.
+\sum_{m=n}^N {N \choose m} {m \choose n} x^m y^{N-m}
+&=& x^n \sum_{r=0}^{N-n} {N \choose n+r} {n+r \choose n}a^r y^{N-n-r} \\
+&=& {N \choose n} x^n \sum_{r=0}^{N-n} {N-n \choose r}x^r y^{N-n-r} \\
+&=& {N \choose n} x^n (x+y)^{N-n}.
 \end{eqnarray*}
 
-Define $C_m = \big( 1- (1-(1-\kappa)^m)^N \big)$ as the probability that
-during $m$ nucleotides, not all duplicates have collapsed.
+\begin{equation*}
+\sum_{m=n}^{N-1} {N \choose m} {m \choose n} x^m y^{N-m} =
+{N \choose n} x^n \Big[ (x+y)^{N-n} - x^{N-n} \Big].
+\end{equation*}
+
+The probability of symbol $\eqg$ with $n$ surviving threads after
+$\scriptstyle{\blacksquare}$ in $k$ steps is
+\begin{equation*}
+\begin{split}
+\sum_{m=n}^{N-1} {N \choose m} (1-\mu/3)^m (\mu/3)^{N-m} \times
+{m \choose n} (1-\mu)^{kn} \big(1-(1-\mu)^k\big)^{m-n} \times \\
+\Big( (1-(1-\mu)^k)^{N-m} - (1-(1-\mu)^{k-1})^{N-m} \Big).
+\end{split}
+\end{equation*}
+
+We obtain a first term equal to
+\begin{equation*}
+(1-\mu)^{kn} (1-(1-\mu)^k)^{N-n}
+\sum_{m=n}^{N-1} {N \choose m} {m \choose n} (1-\mu/3)^m (\mu/3)^{N-m}.
+\end{equation*}
+
+Using the formula above with $x=1-\mu/3$ and $y=\mu/3$, we see that this
+first term is equal to
+\begin{equation*}
+{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \bigg( \big[ 1-(1-\mu)^k \big]^{N-n}
+-\big[ (1-(1-\mu)^k)(1-\mu/3)\big]^{N-n} \bigg)
+\end{equation*}
+
+The second term is
+\begin{equation*}
+\begin{split}
+\left(\frac{(1-\mu)^k}{1-(1-\mu)^k}\right)^n \cdot
+\sum_{m=n}^{N-1} {N \choose m} {m \choose n} (1-\mu/3)^m (\mu/3)^{N-m}
+\times \\
+(1-(1-\mu)^k)^m (1-(1-\mu)^{k-1})^{N-m}.
+\end{split}
+\end{equation*}
+
+Using the formula with $x=(1-(1-\mu)^k)(1-\mu/3)$ and
+$y = (1-(1-\mu)^{k-1})\mu/3$, we obtain
+\begin{equation*}
+{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n
+\bigg( \big[ x+y \big]^{N-n}
+-\big[ (1-(1-\mu)^k)(1-\mu/3)\big]^{N-n} \bigg).
+\end{equation*}
+
+Computing the difference between the terms, the probability we are
+looking for comes out as
+\begin{equation*}
+{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \bigg(
+\big[ 1-(1-\mu)^k \big]^{N-n} -
+\big[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^{N-n} \bigg).
+\end{equation*}
+
+One can verify that the sum of the terms above for $n=1, 2, \ldots, N-1$
+is equal to $E_k$.
+
+To find the probabilities of interest, it is useful to precompute the
+terms
+\begin{equation*}
+\begin{split}
+E_{k,m} = \sum_{n-1}^{N-1}{N \choose n} (1-\mu)^{kn} (1-\mu/3)^n \times \\
+\bigg( \big[ 1-(1-\mu)^k \big]^{N-n} -
+\big[1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^{N-n} \bigg) \times
+(1-(1-\mu)^m)^n.
+\end{split}
+\end{equation*}
+
+Using Newton's binomial formula we find
+\begin{equation*}
+\begin{split}
+E_{k,m} = \big[ (1-\mu)^k(1-\mu/3)(1-(1-\mu)^m) + 1-(1-\mu)^k \big]^N
+- \big[1-(1-\mu)^k \big]^N + \\
+\big[ 1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^N - \\
+\big[(1-\mu)^k(1-\mu/3)(1-(1-\mu)^m) +
+1-(1-\mu+\mu^2/3)(1-\mu)^{k-1} \big]^N.
+\end{split}
+\end{equation*}
+
+Note that $E_{k,0} = 0$.
+
+The probability that there is a $\eqg$ symbol among the first $k$
+nucleotides after a $\dn$ symbol is $q^k H_k / (1-(1-\mu/3)^N)$, where
+\begin{eqnarray*}
+H_k &=&
+\sum_{m=1}^{N-1}{N \choose m}
+(1-\mu/3)^m(\mu/3)^{N-m} \big(1-(1-\mu)^k\big)^{N-m}
+\big(1 - \big( 1-(1-\mu)^k \big)^m\big) \\
+&=& [1-\mu/3(1-\mu)^k]^N - [1-\mu/3]^N - [1-(1-\mu)^k]^N +
+[(1-\mu/3)(1-(1-\mu)^k)]^N.
+\end{eqnarray*}
+
+The probability that the $k$-th symbol after a $\dn$ symbol is also $\dn$
+is
+\begin{equation*}
+\big( 1-q^kH_k/(1-(1-\mu/3)^N) \big) \cdot p(1-(1-\mu/3)^N) =
+p(1-(1-\mu/3)^N) - pq^kH_k.
+\end{equation*}
+
+The probability that the $k$-th symbol is $\DN$ is
+\begin{equation*}
+\big( 1-q^kH_k/(1-(1-\mu/3)^N) \big) \cdot p(1-\mu/3)^N =
+p(1-\mu/3)^N - pq^kH_k \frac{(1-\mu/3)^N}{1-(1-\mu/3)^N}.
+\end{equation*}
 

### 2018-04-06 C/kpoly/memprob.c

@@ -529,7 +529,7 @@ void print_kpoly (kpoly_t *p) {
    fprintf(stderr, "%f", p->coeff[0]);
    fprintf(stderr, " + %fz", p->coeff[1]);
    for (int i = 2 ; i <= K ; i++) {
-      fprintf(stderr, " + %fz^%d", p->coeff[i], i);
+      fprintf(stderr, " + %.10fz^%d", p->coeff[i], i);
    }
    fprintf(stderr, "\n");
 }
@@ -540,7 +540,7 @@ int main(void) {
    memprob_init(0.01, 0.05, 100);
 
    uint gamma = 17;
-   uint N = 5;
+   uint N = 2;
 
    kpoly_t *w = new_zero_kpoly();
    mat_t *M = new_matrix_M(gamma, N);
@@ -557,7 +557,7 @@ int main(void) {
    kpoly_update_add(w, powM1->term[gamma+2]);
 
 
-   for (int i = 0 ; i < 8 ; i++) {
+   for (int i = 0 ; i < 10 ; i++) {
       matrix_mult(powM2, powM1, M);
       kpoly_update_add(w, powM2->term[gamma+2]);
       matrix_mult(powM1, powM2, M);

### 2018-04-06 C/misc/print_mini.py

@@ -1,0 +1,83 @@
+#!/usr/bin/env python
+# -*- coding:utf-8 -*-
+
+import sys
+
+'''This script is used to print the relatively complex transfer matrix
+of the model with a main target and a duplicate.'''
+
+class MatrixAsString:
+
+   def __init__(self, gamma, k, p, kappa, N):
+      self.gamma = gamma
+      self.k     = k    
+      self.p     = p
+      self.kappa = kappa
+      self.N     = N
+
+   def power(self, txt, k):
+      if k == 0: return '1'
+      if k == 1: return txt
+      return '(%s)^%d' % (txt, k)
+
+   def AB_type_terms(self, k):
+      terms = [self.power('%f*z' % (1-self.p), m) for m in range(k)]
+      return '+'.join(terms)
+
+   def tilde_AB_type_terms(self, k):
+      qz = '%f*z' % (1-self.p)
+      terms = [self.Cm(m) + '*' + self.power(qz, m) for m in range(k)]
+      return '+'.join(terms)
+
+   def A(self, k):
+      cst_term = '%f*z*(1-(1-%f/3)^%d)' % (self.p, self.kappa, self.N)
+      if k == 1: return cst_term
+      return cst_term + '*(' + self.AB_type_terms(k) + ')'
+
+   def B(self, k):
+      cst_term = '%f*z*(1-%f/3)^%d' % (self.p, self.kappa, self.N)
+      if k == 1: return cst_term
+      return cst_term + '*(' + self.AB_type_terms(k) + ')'
+
+   def Cm(self, k):
+      if k == 0: return '1'
+      if k == 1: return '(1-%f^%d)' % (self.kappa, self.N)
+      return '(1-(1-(1-%f)^%d)^%d)' % (self.kappa, k, self.N)
+
+   def tildeA(self, k):
+      cst_term = '%f*z*(1-(1-%f/3)^%d)' % (self.p, self.kappa, self.N)
+      if k == 1: return cst_term
+      return cst_term + '*(' + self.tilde_AB_type_terms(k) + ')'
+
+   def tildeB(self, k):
+      cst_term = '%f*z*(1-%f/3)^%d' % (self.p, self.kappa, self.N)
+      if k == 1: return cst_term
+      return cst_term + '*(' + self.tilde_AB_type_terms(k) + ')'
+
+   def r(self, k):
+      if k == 1:
[…]

### 2018-04-06 C/misc/print_transfer_matrix.py

@@ -1,0 +1,73 @@
+#!/usr/bin/env python
+# -*- coding:utf-8 -*-
+
+import sys
+
+'''This script is used to print the relatively complex transfer matrix
+of the model with a main target and a duplicate.'''
+
+class MatrixAsString:
+
+   def sqbr(txt):
+      return '[' + txt + ']'
+
+   def rdbr(txt):
+      return '(' + txt + ')'
+
+   def pow(txt, m):
+      if m == 0: return '1'
+      if m == 1: return txt
+      return self.rdbr('%s^%d' % txt, m)
+
+   def A(k, p, kappa, N):
+      cst_term = '%f*(1-(1-%f/3)^%d)*z' % (p, kappa, N)
+      if k == 1: return cst_term
+
+      other_terms = [self.pow('%f*z' % 1-p, m) for m in range(k)]
+      self.sqbr('+'.join(other_terms))
+
+
+   def first_term(n):
+      '''Return a string of the first term of the row.'''
+      if n == 0: return 'd*z'
+      return 'd*z*(' + '+'.join([az(a) for a in range(n+1)]) + ')'
+
+   def shift_right(L, shift):
+      '''Shift the terms to the right.'''
+      if shift == 0: return L
+      return ['0']*shift + L[:-shift]
+
+   def overwrite_tail(L, n):
+      '''Replace the tail elements by 0.'''
+      if n == 0: return L
+      return L[:-n] + ['0']*n 
+
+   # Two complete lists of terms.
+   L1 = ['c*z'] + ['c*z*%s' % az(a) for a in range(1,d-1)]
+   L2 = ['b*z'] + ['b*z*%s' % az(a) for a in range(1,d-1)]
+
+   # First row.
+   row_terms = [first_term(d-1)] + L1 + L2
+   rows = [brackets(', '.join(row_terms))]
+
+   # Next series of rows.
+   for i in range(1,d):
+      row_terms = [first_term(d-1-i)] + \
+            shift_right(L1,i) + overwrite_tail(L2,i-1)
+      rows.append(brackets(', '.join(row_terms)))
+
+   # Next series of rows.
+   for i in range(1,d):
+      row_terms = [first_term(d-1-i)] + \
+            overwrite_tail(L1,i-1) + shift_right(L2,i)
+      rows.append(brackets(', '.join(row_terms)))
+
+   return brackets(', '.join(rows))
+
+def write_assignment_line(d):
+   head = 'M%d := Matrix(' %d
[…]

### 2018-04-06 make_figure_simulation_mem_fp.R

@@ -1,0 +1,34 @@
+d = 17
+k = 30:100
+k3 = rep(30:100, each=3)
+
+L = list()
+M = list()
+
+M17_2 = as.matrix(read.table("../False_Positives/M17.txt"))
+z3_2 = M17_2[151:200,c(1,3,5)]
+C3_2 = M17_2[151:200,c(2,4,6)]
+
+M17_3 = as.matrix(read.table("M17.txt"))
+z3_3 = M17_3[151:200,c(1,3,5)]
+C3_3 = M17_3[151:200,c(2,4,6)]
+
+for (i in 1:50) {
+   # All the terms
+   #L[[i]] = Re(tapply(X = C3_3[i,]/z3_3[i,]^(k3+1) - C3_2[i,]/z3_2[i,]^(k3+1), INDEX=k3, sum))
+   # P2 simplified
+   #L[[i]] = Re(tapply(X = C3_3[i,]/z3_3[i,]^(k3+1), INDEX=k3, sum)) - Re(C3_2[i,1]/z3_2[i,1]^(k+1))
+   # P3 simplified
+   #L[[i]] = Re(C3_3[i,1]/z3_3[i,1]^(k+1)) - Re(tapply(X = C3_2[i,]/z3_2[i,]^(k3+1), INDEX=k3, sum)) 
+   # Both simplified
+   L[[i]] = Re(C3_3[i,1]/z3_3[i,1]^(k+1)) - Re(C3_2[i,1]/z3_2[i,1]^(k+1))
+}
+
+pdf("sht.pdf")
+ymin = 0
+ymax = max(unlist(L))
+plot(L[[1]], type='l', ylim=c(ymin,ymax))
+for (i in 33:50) {
+   lines(L[[i]])
+}
+dev.off()

### 2018-04-19 MEM.tex
First clean version of the transfer matrix with proper notations.

--- 2018-01-21 MEM.tex
@@
 
-We can describe reads as sequences of four possible symbols representing
-the state of the local match with the genome. The first symbol,
-represented by a black square indicates a sequencing error. This is a
-mismatch between the read and the genome so the nucleotide cannot belong
-to a correct MEM seed. It either belongs to an incorrect MEM seed, or it
-marks the end of a correct MEM seed.
+\subsection{Seeds}
+DEFINE MEM SEED AND SHARED MEM SEED.
 
-The second symbol, represented by a black square with a white cross
-indicates a sequencing error that is also a mismatch for \emph{all} the
-duplicates. The nucleotide cannot belong to any MEM seed (correct or not)
-so the next MEM seed must start at the following nucleotide. The symbol
-`resets' the seeding process, which becomes in the same state as at the
-beginning of the read.
+\subsection{The seeding alphabet}
 
-The third symbol, represented by an exclamation mark indicates that the
-target sequence is the only local match between the read and the genome.
-This means that there is at least one mismatch between the read and any
-duplicate since the beginning of the current MEM. From that nucleotide,
-the MEM will extend forward until the next sequencing error. This means
-that two exclamation marks must be separated by either a black square or a
-black square with a cross.
+By nature, reads are sequences of \texttt{A}, \texttt{T}, \texttt{G} and
+\texttt{C}, but for the purpose of the seeding process, it is more useful
+to consider them as a sequence of symbols telling if the nucleotides match
+the target and~/ or the duplicates. Here, we will consider that reads are
+sequences of the five symbols $\Downarrow$, $\downarrow$, $\uparrow$, $\sim$
+and $\square$.
 
-The last symbol, represented as a white square indicates all the other
-situations. This can mark the beginning but not the end of a MEM.
+The $\Downarrow$ symbol indicates that the nucleotide is not only a
+mismatch for the target, but also for every duplicate. The $\downarrow$
+symbol indicates that the nucleotide is a mismatch for the target, but is
+a match for at least one dupulicate. These two symbols are referred to as
+the \emph{error symbols}, because they imply the occurrence of a
+sequencing error (this is the only way the read may not match the target).
+
+The remaing three symbols have a more complex usage because they describe
+events of the MEM seeding process that are best explained by way of a
+concrete example. Fig.~\ref{fig:sketch_MEM} illustrates the case of a read
+of size 25 from a sequence with three duplicates. The four files of
+squares at the bottom are the target and duplicate sequences, where a
+black square means that the decoded nucleotide is a mismatch and a grey
+square means that it is a match.
 
@@
 
+Now consider streaks of matches until a particular position of the read.
+At the 4-th nucleotide, the target scores 4 matches in a row versus 2, 0
+and 1 for the duplicates. The target just acquired the strictly longest
+streak, which is indicated by the $\uparrow$ symbol. This symbol also
+appears at position 17, where the target sequence acquires the strictly
+longest streak again (6 matches in a row, versus 4, 0 and 4 for the
+dupicates).
+
+At position 23, the target and the first duplicate both acquire the
+longest streak (2 matches in a row), which is indicated by the $\sim$
+symbol. This symbol is used only at positions where the target
+\emph{acquires} the longest streak (together with at least one duplicate).
+For instance, it is not used at position 1 nor at position 12 because the
+target already had the longest streak (0 in both cases).
+
+Finally, the $\square$ symbol is used for all the other cases,
+\textit{i.e.}, when there is no sequencing error, and when the target does
+not acquire the longest streak of matches (shared or not with a
+duplicate).
+
+Each sequence corresponds to a Bernoulli process where `success' stands
+for a match between the sequence and the read, and `failure' stands for a
+mismatch. For simplicity, we will refer to the Bernoulli processes as
+`threads'.
+
+\begin{definition}
+At a given position of the read, a duplicate thread is said to be
+\emph{strictly masking} if its match streak is strictly longer than the
+match streak of the target. A duplicate thread is said to be
+\emph{potentially masking} if its match streak is the same as the target.
+\end{definition}
+
+Duplicate threads are thus classified into strictly masking, potentially
+masking and not masking, which allows us to describe the usage of the five
+symbols.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.7]{decision_tree.pdf}
+\caption{\textbf{Decision tree}. 
+Some legend.}
+\label{fig:decision_tree}
+\end{figure}
+
+We assume that the target sequence has $N$ duplicates that are potential
+false positives for the seeding process. We further assume that
+duplication was instantaenous and that all $N+1$ sequences diverge
+independently from each other at a constant rate. In other words, we
+ignore the complications due to the genealogy of the duplication events
+and we simply assume that at each position, any given duplicate has the
+same nucleotide as the target with probability $1-\mu$. If it does not,
+we assume that the duplicate has any of the remaining three
+nucleotides with equal probability.
+
+To describe those processes, we will
+often use the function $\varphi(n)$ that gives the length of the success
+streak at position $n$. The processes are indexed such that $\varphi_0$
+refers to the target sequence, and functions $\varphi_j$ $(1 \leq j \leq
+N)$ refer to the duplicates. For a concrete example, $\varphi_0(5) = 5$
+and $\varphi_1(5) = 3$ in Fig.~\ref{fig:sketch_MEM}.
+
+With these definitions, the decision tree for choosing the next symbol to
+insert in a read can be summarized as in Fig.~\ref{fig:decision_tree}.
+This sketch will be useful as a reference to compute the probabilities of
+the blocks and their weighted generating functions.
+
+\begin{definition}
+The main thread is dominant at nucleotide $n$ if $\varphi_0(n) >
+\varphi_i(n)$ for all $i$ $(1 \leq i \leq N)$. The main thread is
+co-dominant if inequality is not strict and there exists $i$ $(1\leq i
+\leq N)$ such that $\varphi_0(n) = \varphi_i(n)$.
+\end{definition}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{no_3MEM_graph.pdf}
+\caption{\textbf{Transfer graph of a read without 3-MEM}. 
+Some legend.}
+\label{fig:noMEM_graph}
+\end{figure}
+
+\section{The transfer matrix}
+
+\subsection{Transitions from $\uparrow$}
+\label{sec:trans_from_up}
+
+On a $\uparrow$ symbol, the main thread becomes dominant. It will remain
+so until the first sequencing error, so the next block will be terminated
+by an error symbol (\textit{i.e.} $\Downarrow$ or $\downarrow$).
+
+\begin{definition}
+$\omega$ and $\tilde{\omega}$ are the probabilities of the $\Downarrow$
+and of the $\downarrow$ error symbols, respectively, \textit{i.e.}
+\begin{eqnarray}
+\omega &=& p \cdot (1-\mu/3)^N \\
+\tilde{\omega} &=& p \cdot \big( 1-(1-\mu/3)^N \big).
+\end{eqnarray}
+\end{definition}
+
+The weighted generating function of the transition from $(\uparrow)_j$ to
+$\Downarrow$ is
+\begin{equation}
+D_{\gamma-j}(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
+\end{equation}
+
+The weighted generating function of the transition from
+$(\uparrow)_j$ to $\downarrow$ is
+\begin{equation}
+\tilde{D}_{\gamma-j}(z) = \tilde{\omega} z
+  \sum_{i=0}^{\gamma-j-1} (qz)^i.
+\end{equation}
+
+\subsection{Transitions from $\Downarrow$}
+\label{sec:trans_from_Down}
+
+At a $\Downarrow$ symbol, all the threads have match length $0$ and the
+read is `reset' to its initial state. The main thread is co-dominant with
+all the alternative threads. The next terminator symbol depends on what
+happens first. If all the alternative threads fail before the main thread,
+the latter becomes dominant ($\uparrow$ symbol). If not all alternative
+threads have failed when the main thread does, we need to distinguish two
+cases: if all threads fail together with the main thread, the read is
+reset ($\Downarrow$ symbol), otherwise the main read becomes dominated
+($\downarrow$ symbol).
+
+If the next terminator is the $\uparrow$ symbol, we need to distinguish
+the cases based on the value of $\varphi_0$ (see
+section~\ref{sec:trans_from_up}), which is just the position of the
+$\uparrow$ symbol after the $\Downarrow$ symbol in this case. A transition
+from $\Downarrow$ to $(\uparrow)_j$ consists of a succession of $j$
+correct nucleotides -- each with probability $q$. But we must also make
+sure that no other $\uparrow$ symbol is inserted before position $j$.
+
+\begin{definition}
+$\xi_{i,m}$ is the probability that at least one of $m$ alternative
+threads does not fail in $i$ nucleotides, \textit{i.e.}
+\begin{equation}
+  \xi_{i,m} = 1-(1-(1-\mu)^i)^m.
+\end{equation}
+\end{definition}
+
+The weighted generating function of the transition from $\Downarrow$ to
+$(\uparrow)_i$ is
+\begin{equation}
+u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
+\end{equation}
+
+If the next terminator is $\Downarrow$, there is a limit on the
+number of $\square$ symbols. Indeed, if there are $\gamma$ or more
+nucleotides between the two $\Downarrow$ symbols, they constitute a shared
+MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
+generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
+same rationale as above, the weighted generating function of the
+transition from $\Downarrow$ to $\Downarrow$ is
+\begin{equation}
+A_{\gamma}(z) = \omega z
+  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
+\end{equation}
+
+If the next terminator is $\downarrow$, there is no limit on the number of
+$\square$ symbols that can be inserted. Indeed, the main thread will not
+be part of the MEM in this case. As mentioned in
+section~\ref{sec:trans_from_up}, the weighted generating function of the
+$\downarrow$ symbol is $pz(1-(1-\mu/3)^N)$. The probability that it first
+occurs at position $n+1$ after the $\Downarrow$ symbol is the probability
+that the first $n$ nucleotides are correct -- each with probability $q$
+-- and that at least one alternative thread survives -- probability
+$\xi_{n,N}$. The weighted generating function of the transition from
+$\Downarrow$ to $\downarrow$ is thus
+\begin{equation}
+\tilde{A}_\infty(z) = \tilde{\omega} z
+  \sum_{i=0}^\infty \xi_{i,N} \cdot (qz)^i.
+\end{equation}
+
+\subsection{Transitions from $\downarrow$}
+\label{sec:trans_from_down}
+
+At a $\downarrow$ symbol, the main thread fails and at least one duplicate
+thread does not (othwerwise the symbol would be $\Downarrow$). By
+definition, the duplicate threads that do not fail are strictly masking,
+and those that fail are potentially masking. A $\downarrow$ symbol implies
+the presence of a sequencing error, so duplicate threads fail with
+probability $1-\mu/3$. The probability that $m$ duplicate threads fail and
+$N-m$ do not $(0 \leq m \leq N-1)$ is thus
+\begin{equation}
+  Q_m = \frac{1}{1-(1-\mu/3)^N}{N \choose m} (1-\mu/3)^m(\mu/3)^{N-m}.
+\end{equation}
+
+The $\downarrow$ symbol can be followed by all the terminators. We need to
+start with the case of the $\sim$ terminator, because it will be required
+for the other ones. Referring to the decision tree of
+Fig.~\ref{fig:decision_tree}, we see that the $\sim$ symbol occurs in an
+error-free stretch, if all the strictly masking threads fail before the
+potentially masking threads. In section~\ref{sec:trans_from_sim} it will
+be important to distinguish different cases based on the value of
+$\varphi_0$ at the position of the $\sim$ symbol, so as in
+section~\ref{sec:trans_from_up}, we introduce the states $(\sim)_j$ $(1
+\leq j \leq \gamma-1)$ such that $\varphi_0(n) = j$ if the nucleotide at
+position $n$ is $(\sim)_j$. In this case, the value of $\varphi_0(n) = j$
+is simply the position of the $\sim$ symbol relative to the preceding
+$\downarrow$ symbol.
+
+If there are $m = 0$ potentially masking threads, the next terminator
+cannot be $\sim$. Assuming that there are $m > 0$ potentially masking
+threads at the $\downarrow$ symbol, the probability that the have not all
+faied at position $i$ is $\xi_{i,m}$. Similarly, the probability that the
+last strictly masking thread fails at position $i$ is
+$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
+is $(\sim)_i$ is $q^j V_i$, where
+\begin{eqnarray}
+\label{eq:Si}
+V_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
+  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
+  &=& \frac{\alpha_i^N-\alpha_{i-1}^N-\gamma_i^N+\delta_{i-1}^N}
+{1-(1-\mu/3)^N} \text{, where}
+\end{eqnarray}
+
+\begin{eqnarray}
+\alpha_i &=& 1-(1-\mu)^i\mu/3 \\
+\gamma_i &=& 1-(1-\mu)^i \\
+\delta_i &=& 1-(1-\mu+\mu^2/3)(1-\mu)^i.
+\end{eqnarray}
+
+We have assumed $m>0$, but the sum above can start at $m=0$ because
+$\xi_{i,0}=0$. So the weighted generating function of the transition from
+$\downarrow$ to $(\sim)_i$ is
+\begin{equation}
+v_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
+  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
+\end{equation}
+
+Referring again to the decision of Fig.~\ref{fig:decision_tree}, we see
+that the $\uparrow$ symbol occurs in an error-free stretch, if all the
+potentially masking threads fail before or at the same nucleotide as the
+strictly masking threads. Assuming that there are $m \geq 0$ potentially
+masking threads at the $\downarrow$ symbol, the probability that they have
+all failed at position $i$ is $1-\xi_{i,m}$. As we have seen above, the
+probability that the last strictly masking thread fails at position $j$ is
+$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
+is the $\uparrow$ symbol at position $i$ is $q^i W_i$, where
+\begin{eqnarray}
+  W_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
+  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
+    &=& \frac{\gamma_i^N-\delta_{i-1}^N}{1-(1-\mu/3)^N}.
+\end{eqnarray}
+
+The weighted generating function of the transition from $\downarrow$ to
+$(\uparrow)_i$ is
+\begin{equation}
+w_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
+\end{equation}
+
+In order to compute the weighted generating functions of the other
+transitions, first observe that
+\begin{equation}
+V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
+\end{equation}
+So the probability that an error-free stretch of length $i \geq 1$
+following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
+$\uparrow$ symbol is
+\begin{equation}
+\sum_{j=1}^i V_j+W_j = \frac{\alpha_i^N -
+\alpha_0^N}{1-(1-\mu/3)^N}.
+\end{equation}
+
+The probability that the next terminator after a $\downarrow$ symbol is a
+$\Downarrow$ symbol at position $j$ is the probability of an error-free
+stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
+$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
+by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
+weighted generating function of the transition from $\downarrow$ to
+$\Downarrow$ is
+\begin{equation}
+B_\infty(z) = \omega z \cdot \Big(1 + \sum_{i=1}^\infty
+\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
+\end{equation}
+
+By the same rationale, we see that the weighted generating function of the
+transition from $\downarrow$ to $\downarrow$ is
+\begin{equation}
+\tilde{B}_\infty(z) = \tilde{\omega} z \cdot \Big(1 + \sum_{i=1}^\infty
+\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
+\end{equation}
+
+\subsection{Transitions from $\sim$}
+\label{sec:trans_from_sim}
+
+At a $\sim$ symbol, the main thread becomes co-dominant, together with $1$
+to $N-1$ alternative threads (if there were $N$, the symbol would be
+$\Downarrow$). As in section~\ref{sec:trans_from_Down}, the next
+terminator symbol depends on whether all the alternative co-dominant
+threads fail before the main thread. If this is the case, the main thread
+will become dominant ($\uparrow$ symbol), otherwise we again need to
+distinguish the case that the read is reset ($\Downarrow$ symbol) from the
+case that the main thread becomes dominated ($\downarrow$ symbol).
+
+An important difficulty is that the $\sim$ symbol does not inform us about
+the number of surviving threads (except when $N = 1$). So we need to
+marginalize the probabilities of interest over this number. The first
+thing to note is that a $\sim$ terminator can only follow a $\downarrow$
+terminator. Indeed, after a $\Downarrow$ symbol the threads are already
+co-dominant, and after a $\uparrow$ symbol the alternative threads cannot
+`catch up' with the main thread until a sequencing error occurs,
+\textit{i.e.} until a $\Downarrow$ symbol or a $\downarrow$ symbol is
+inserted. The number of surviving threads depends on the position of the
+$\sim$ symbol relative to the preceding $\downarrow$ symbol, so we need
+to treat the states $(\sim)_j$ separately.
+
+We first establish an equality that will turn out to be useful in the
+subsequent calculations.
+
+\begin{lemma}
+For any real numbers $x$ and $y$,
+\begin{eqnarray*}
+\sum_{n=m}^N {N \choose n} {n \choose m} x^n y^{N-n}
+&=& x^m \sum_{r=0}^{N-m} {N \choose m+r} {m+r \choose m}a^r y^{N-m-r} \\
+&=& {N \choose m} x^m \sum_{r=0}^{N-m} {N-m \choose r}x^r y^{N-m-r} \\
+&=& {N \choose m} x^m (x+y)^{N-m}.
+\end{eqnarray*}
+
+The equality above implies
+\begin{equation}
+\label{eq:double_binom}
+\sum_{n=m}^{N-1} {N \choose n} {n \choose m} x^n y^{N-n} =
+{N \choose m} x^m \Big[ (x+y)^{N-m} - x^{N-m} \Big].
+\end{equation}
+\end{lemma}
+
+The probability that there are $m>1$ potentially masking threads upon
+entering state $(\sim)_j$ is the probability that there were $n \geq m$
+potentially masking threads at the $\downarrow$ symbol -- probability
+$Q_n$ --, that exactly $m$ have survived until position $j$ -- probability
+${n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}$ -- and that the last of
+the $N-n$ strictly masking threads fails at position $j$ -- probability
+$\xi_{j-1,N-n}-\xi_{j,N-n}$. Summing over $n = m, \ldots, N-1$, the
+probability is seen to be
+\begin{equation}
+\label{eq:miscprob}
+\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
+\cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
+\end{equation}
+
+We write $\xi_{j-1,N-n}-\xi_{j,N-n}$ as $(1-\xi_{j,N-n}) -
+(1-\xi_{j-1,N-n})$ and split the sum of equation (\ref{eq:miscprob}) into
+two parts. The first term is equal to
 \begin{equation*}
-M(z) =
-\begin{blockarray}{ccccccc}
-   & \scriptstyle{\blacksquare} & \scriptstyle{\times} 
-   & \scriptstyle{(!)_1} & \scriptstyle{(!)_2}
-   & \ldots & \scriptstyle{(!)_{\gamma-1}} \\
-\begin{block}{c[cccccc]}
-\scriptstyle{\blacksquare} & \tilde{A}_\infty(z) & \tilde{B}_\infty(z) &
-    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
-\scriptstyle{\times} & \tilde{A}_\infty(z) & \tilde{B}_\gamma(z) &
-    r_1(z) & r_2(z) & \ldots & r_{\gamma-1}(z) \\
-\scriptstyle{(!)_1} & A_{\gamma-1}(z) & B_{\gamma-1}(z) & 0 & 0 &
-    \ldots & 0 \\
-\scriptstyle{(!)_2} & A_{\gamma-2}(z) & B_{\gamma-2}(z) & 0 & 0 &
-    \ldots & 0  \\
-\vdots & \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
-\scriptstyle{(!)_{\gamma-1}} & A_1(z) & B_1(z) & 0 & 0 & \ldots & 0 \\
+\begin{split}
+\sum_{n=m}^{N-1} & Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
+\cdot (1 - \xi_{j,N-n}) = \\
+& \frac{(1-\mu)^{jm} (1-(1-\mu)^j)^{N-m}}{1-(1-\mu/3)^N}
+\sum_{n=m}^{N-1} {N \choose n} {n \choose m} (1-\mu/3)^n (\mu/3)^{N-n}.
+\end{split}
+\end{equation*}
+
+Using formula (\ref{eq:double_binom}) with $x=1-\mu/3$ and $y=\mu/3$, we
+see that this term is equal to
+\begin{equation*}
+\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
+\bigg( \gamma_j^{N-m}
+-\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
+\end{equation*}
+
+The second term of equation (\ref{eq:miscprob}) is equal to
+\begin{multline*}
+\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
+\cdot (1 - \xi_{j-1,N-n}) = \\
+\frac{1}{1-(1-\mu/3)^N}
+\left(\frac{(1-\mu)^j}{1-(1-\mu)^j}\right)^m
+\; \sum_{n=m}^{N-1} {N \choose n} {n \choose m} \times \\
+(1-\mu/3)^n (\mu/3)^{N-n}
+(1-(1-\mu)^j)^n (1-(1-\mu)^{j-1})^{N-n}.
+\end{multline*}
+
+Formula (\ref{eq:double_binom}) with $x=(1-(1-\mu)^j)(1-\mu/3)$ and
+$y = (1-(1-\mu)^{j-1})\mu/3$ yields
+\begin{equation*}
+\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m}
+\bigg( \delta_{j-1}^{N-m}
+-\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
+\end{equation*}
+
+Computing the difference between the terms, we see that the probability
+that there are $m$ potentially masking threads upon entering state
+$(\sim)_j$ is
+\begin{equation*}
+P_{m,j} = \frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
+\big( \gamma_j^{N-m} - \delta_{j-1}^{N-m} \big).
+\end{equation*}
+
+The sum of the terms above for $m = 1, \ldots, N-1$ is $V_j$ from equation
+(\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.
+
+Now, referring to the decision tree of Fig.~\ref{fig:decision_tree}, we
+see that if the $m$ potentially masking threads fail before the main
+thread, the next terminator will be the $\uparrow$ symbol, otherwise it
+will be either the $\Downarrow$ symbol or the $\downarrow$ symbol.
+
+The probability that the last of the $m$ potentially masking threads fails
+at position $i$ after entering state $(\sim)_j$ is
+$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $V_j$ and summing over $m = 1,
+\ldots, N-1$, we find that the weighted generating function of the
+transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
+\begin{eqnarray}
+y_{j,j+i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
+\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
+&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
+-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
+  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
+\end{eqnarray}
+where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
+\begin{equation}
+\beta_{j,i} = 1 - (1-\mu)^j\mu/3 + (1-\mu)^i(1-\mu/3).
+\end{equation}
+
+The probability that the next terminator is a $\Downarrow$ symbol or a
+$\downarrow$ symbol at position $i$ is the probability that at least one
+of the potentially masking threads survives until position $i-1$ --
+probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
+error -- probability $p$. Dividing by $V_j$ and summing over $m = 1,
+\ldots, N-1$, we find that the weighted generating function of the
+transtion from $(\sim)_j$ to
+$\Downarrow$ is
+\begin{eqnarray}
+C_{\gamma-j}(z) &=& \omega z \sum_{i=1}^{\gamma-j-1} \sum_{m=1}^{N-1}
+  \frac{P_{m,j}}{V_j} \xi_{i-1,m} \cdot (qz)^{i-1} \notag \\
+  &=& \omega z \sum_{i=1}^{\gamma-j-1}
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
+  +\beta_{j-1,i+j-1}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
+\end{eqnarray}
+
+With the same notations, the weighted generating function from $(\sim)_j$
+to $\downarrow$ is seen to be 
+\begin{equation}
+\tilde{C}_{\gamma-j}(z) = \tilde{\omega} z \sum_{i=1}^{\gamma-j-1}
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
+  +\beta_{j-1,i+j-1}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
+\end{equation}
+
+\subsection{Head and tail vectors}
+
+At the beginning of the read, the match lengths of all the threads are
+$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
+consider that the read starts in the $\Downarrow$ state, which is
+equivalent to adding an edge labelled with the empty object $\varepsilon$
+between the head vertex and the $\Downarrow$ vertex. The head vector is
+thus
+\begin{equation*}
+H(z) = 
+\begin{blockarray}{cccccccc}
+   \;\; \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
+    \up{1} & \ldots & \up{\gamma-1} \\
+\begin{block}{[cccccccc]}
+\;\; 1 & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 \\
+\end{block}
+\end{blockarray}.
+\end{equation*}
+The tail vector contains the weighted generating functions of
+non-terminated stretches of $\square$ symbols. Referring to the decision
+tree of Fig.~\ref{fig:decision_tree}, we see that the probability of
+occurrence of $\square$ symbols depends on whether there are potentially
+masking and / or strictly masking threads. All the symbols except
+$\downarrow$ impose a maximum on the number of $\square$ symbols at the
+end of the reads.
+
+For the $\downarrow$ symbol, the weighted generating function is simply
+\begin{equation}
+T_{\downarrow}(z) = \sum_{i=0}^\infty (qz)^i = \frac{1}{1-qz}.
+\end{equation}
+
+For the $(\uparrow)_j$ state, the weighted generating function is
+\begin{equation}
+T_{(\uparrow)_j}(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
+\end{equation}
+
+For the other symbols, we must take care that the potentially masking
+threads do not all fail (otherwise this would imply the presence of a
+$\uparrow$ terminator in the tail). For the $\Downarrow$ symbol, the
+weighted generating function is
+\begin{equation}
+T_{\Downarrow}(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
+\end{equation}
+
+For the $(\sim)_j$ state, the weighted generating function is
+\begin{equation}
+T_{(\sim)_j}(z) = \sum_{i=0}^{\gamma-j-1}
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
+\end{equation}
+
+In summary,
+\begin{equation*}
+T(z) = 
+\begin{blockarray}{cc}
+\begin{block}{c[c]}
+\dn & T_\downarrow(z) \\
+\DN & T_\Downarrow(z) \\
+\eq{1} & T_{(\sim)_1}(z) \\
+\vdots & \vdots \\
+\eq{\gamma-1} & T_{(\sim)_{\gamma-1}}(z) \\
+\up{1} & T_{(\uparrow)_1}(z) \\
+\vdots & \vdots \\
+\up{\gamma-1} & \; T_{(\uparrow)_{\gamma-1}}(z) \; \\
 \end{block}
@@
 
-\begin{eqnarray*}
-A_n(z) &=& p (1-(1-\kappa/3)^N) \sum_{m=0}^{n-1} q^mz^{m+1}, \\
-\tilde{A}_n(z) &=& p (1-(1-\kappa/3)^N)
-    \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
-B_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} q^mz^{m+1}, \\
-\tilde{B}_n(z) &=& p(1-\kappa/3)^N \sum_{m=0}^{n-1} C_m q^mz^{m+1}, \\
-r_n(z) &=& (C_{n-1}-C_n) (qz)^n, \\
-F_n(z) &=& \sum_{m=0}^n C_m (qz)^m.
-\end{eqnarray*}
+\section{Conclusion}
 
-Define $C_m = \big( 1- (1-(1-\kappa)^m)^N \big)$ as the probability that
-during $m$ nucleotides, not all duplicates have collapsed.
-
-\section{Conclusion}
-Some conclusion.
+\begin{equation*}
+\begin{blockarray}{ccccccccc}
+   & \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
+    \up{1} & \ldots & \up{\gamma-1} \\
+\begin{block}{c[cccccccc]}
+\DN & A_\gamma(z) & \tilde{A}_\infty(z) & 0 & \ldots & 0 & u_1(z)
+    & \ldots & u_{\gamma-1}(z) \\
+\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z) & \ldots &
+    v_{\gamma-1}(z) & w_1(z) & \ldots & w_{\gamma-1}(z) \\
+\eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
+    \ldots & y_{1,\gamma-1}(z) \\
+\eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
+    \ldots & y_{2,\gamma-1}(z) \\
+\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
+    \ddots & \vdots \\
+\eq{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
+    \ldots & 0 & 0 & \ldots & 0 \\
+\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
+    \ldots & 0 \\
+\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
+    \ldots & 0  \\
+\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
+    \ddots & \vdots \\
+\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
+    \ldots & 0 \\
+\end{block}
+\end{blockarray}
+\end{equation*}
 

### 2018-04-22 C/kpoly/memmatrix.c
Update C code with new (correct) formulas. Many errors (debugging).

@@ -1,0 +1,755 @@
+#include <math.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+
+
+//  TYPE DECLARATIONS  //
+
+typedef unsigned int  uint;   // Just a shortcut.
+
+// Struct declarations and definitions //
+
+typedef struct kpoly_t  kpoly_t;
+typedef struct mat_t    mat_t;
+typedef struct mono_t   mono_t;
+
+struct mono_t {
+   size_t deg;
+   double coeff;
+};
+
+struct kpoly_t {
+   mono_t mono;              // Monomer (if applicable).
+   double coeff[];           // Terms of the polynomial.
+};
+
+struct mat_t {
+   const size_t    dim;      // Column / row number.
+         kpoly_t * term[];   // Terms of the matrix.
+};
+
+
+// GLOBAL VARIABLES / CONSTANTS //
+
+size_t    G;       // Minimum size of MEM seeds.
+size_t    N;       // Number of duplicate sequences.
+size_t    K;       // Max degree of the polynomials (read size).
+
+size_t    HIGH;    // Proxy for infinite degree polynomials.
+
+double    P;       // Probability of a read error.
+double    U;       // Divergence rate between duplicates.
+double    OMEGA;   // Probability of double-down arrow.
+double   _OMEGA;   // Probability of down arrow.
+
+size_t    KSZ;     // Size of the 'kpoly_t' struct.
+
+kpoly_t * TEMP = NULL;        // Matrix multipliciation.
+kpoly_t * ARRAY[1024] = {0};  // Store the results (indexed by N).
+
+int       ERRNO; // Error codes.
+
+
+// MACROS //
+#define YES 1
+#define NO  0
+
+// Creation of a new 'kpoly_t' is just a call to 'calloc()'.
+#define new_zero_kpoly() calloc(1, KSZ)
+
+// Prob one of m altnerative threads survives i steps.
+#define xi(i,m) ( 1.0 - pow( 1.0 - pow(1.0-U,(i)), (m) ))
+
+// Calculation intermediates (one index).
+#define aN(i) pow( 1.0 - pow(1.0-U,(i)) * U/3.0, N )
[…]

### 2018-04-22 C/kpoly/unittest/runtests.c

@@ -1,0 +1,33 @@
+#include "unittest.h"
+
+char open_banner[] = "";
+char close_banner[] = "---\n";
+
+int
+main(
+   int     argc,
+   char ** argv
+)
+{
+
+   // Import test cases.
+   extern test_case_t test_cases_dunno[];
+
+   // Register test cases.
+   const test_case_t *list_of_test_cases[] = {
+      test_cases_dunno,
+      NULL,
+   };
+
+   // Display banner.
+   fprintf(stderr, "%s", open_banner);
+
+   // Run the tests.
+   int nfailed = run_unittest(argc, argv, list_of_test_cases);
+
+   // Display banner.
+   fprintf(stderr, "%s", close_banner);
+
+   return nfailed;
+
+}

### 2018-04-22 C/kpoly/unittest/unittests_hmm.c

@@ -1,0 +1,19 @@
+#include "unittest.h"
+#include "memmatrix.c"
+
+void
+test_kpoly_mult
+(void)
+{
+
+   test_assert(0);
+
+   return;
+
+}
+
+// Test cases for export.
+const test_case_t test_cases_dunno[] = {
+   {"dunno/kpoly_mult",         test_kpoly_mult},
+   {NULL, NULL},
+};

### 2018-04-23 C/kpoly/memmatrix.c
Fixing some bugs (still plenty of issues).

@@ -71,7 +71,7 @@ int       ERRNO; // Error codes.
                            pow(1-U,(i))*(1-U/3.0), N)
 
 
-void print_kpoly (kpoly_t *p);
+void print_kpoly (kpoly_t *p, size_t);
 
 
 // FUNCTION DEFINITIONS //
@@ -280,6 +280,7 @@ new_kpoly_u
       // See definition of polynomial u.
       new->mono.deg = deg;
       new->mono.coeff = (xi(deg-1,N) - xi(deg,N)) * pow(1.0-P,deg);
+      new->coeff[deg] = new->mono.coeff;
    }
 
    return new; // NULL in case of failure.
@@ -310,6 +311,7 @@ new_kpoly_v
       double numer =  aN(deg) - aN(deg-1) - gN(deg) + dN(deg-1);
       double denom = 1.0 - pow(1-U/3.0, N);
       new->mono.coeff = numer / denom * pow(1.0-P, deg);
+      new->coeff[deg] = new->mono.coeff;
    }
 
    return new; // NULL in case of failure.
@@ -340,6 +342,7 @@ new_kpoly_w
       double numer = gN(deg) - dN(deg-1);
       double denom = 1.0 - pow(1-U/3.0, N);
       new->mono.coeff = numer / denom * pow(1.0-P, deg);
+      new->coeff[deg] = new->mono.coeff;
    }
 
    return new; // NULL in case of failure.
@@ -374,6 +377,7 @@ new_kpoly_y
       double numer = bN(j,j+i) - bN(j,j+i-1) - bN(j-1,i+j) + bN(j-1,j+i-1);
       double denom_j = aN(j) - aN(j-1) - gN(j) + dN(j-1);
       new->mono.coeff = numer / denom_j * pow(1.0-P, deg);
+      new->coeff[deg] = new->mono.coeff;
    }
 
    return new; // NULL in case of failure.
@@ -573,12 +577,13 @@ new_matrix_M
       // First row.
       M->term[0*dim+1] = new_zero_kpoly();
       M->term[0*dim+1]->coeff[0] = 1.0;
+      M->term[0*dim+1]->mono.coeff = 1.0;
 
       // Second row.
       M->term[1*dim+1] = new_kpoly_A(G, NO);
       M->term[1*dim+2] = new_kpoly_A(HIGH, YES);
       for (int j = 1 ; j <= G-1 ; j++)
-         M->term[1*dim+j+G+1] = new_kpoly_u(j);
[…]

### 2018-04-24 C/kpoly/memmatrix.c
Fixed the errors. Results seem correct.

@@ -67,8 +67,8 @@ int       ERRNO; // Error codes.
 #define dN(i) pow( 1.0 - (1.0 - U + U*U/3.0) * pow(1.0-U,(i)), N )
 
 // Calculation intermediates (two indices).
-#define bN(j,i) pow( 1.0 - pow(1.0-U,(j))*U/3.0 + \
-                           pow(1-U,(i))*(1-U/3.0), N)
+#define bN(j,i) pow( 1.0 - pow(1.0-U,(j))*U/3.0 - \
+                           pow(1.0-U,(i))*(1.0-U/3.0), N)
 
 
 void print_kpoly (kpoly_t *p, size_t);
@@ -141,10 +141,16 @@ new_kpoly_A
    }
    else {
       // See definition of polynomial A.
+      const int d = deg <= G ? deg : G;
       const double cst = tilde ? _OMEGA : OMEGA;
       double pow_of_q = 1.0;
-      for (int i = 1 ; i <= deg ; i++) {
+      for (int i = 1 ; i <= d ; i++) {
          new->coeff[i] = cst * (xi(i-1,N)) * pow_of_q;
+         pow_of_q *= (1.0-P);
+      }
+      // Terms of the polynomials with degree higher than 'G' (if any).
+      for (int i = d+1 ; i <= deg ; i++) {
+         new->coeff[i] = P * (1-aN(i-1)) * pow_of_q;
          pow_of_q *= (1.0-P);
       }
    }
@@ -176,10 +182,9 @@ new_kpoly_B
       // See definition of polynomial B.
       const double cst = tilde ? _OMEGA : OMEGA;
       const double denom = 1.0 - pow(1-U/3.0, N);
-      double pow_of_q = 1.0-P;
-      new->coeff[1] = cst;
-      for (int i = 2 ; i <= deg ; i++) {
-         double numer = aN(i-1) - aN(0);
+      double pow_of_q = 1.0;
+      for (int i = 1 ; i <= deg ; i++) {
+         double numer = 1.0 - aN(i-1);
          new->coeff[i] = cst * numer / denom * pow_of_q;
          pow_of_q *= (1.0-P);
       }
@@ -198,6 +203,9 @@ new_kpoly_C
 )
 {
 
+   // FIXME //
+   if (N == 1) return NULL;
+
    if (deg > K || deg == 0) {
       ERRNO = __LINE__;
       return NULL;
[…]

### 2018-04-24 MEM.tex
Text is a mess but formula are mostly correct except typos.

@@
 $\xi_{i,m}$ is the probability that at least one of $m$ alternative
-threads does not fail in $i$ nucleotides, \textit{i.e.}
+threads does not fail in $i$ error-free nucleotides, \textit{i.e.}
 \begin{equation}
@@
 \begin{equation}
-A_{\gamma}(z) = \omega z
+A_\gamma(z) = \omega z
   \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
@@
 
-If the next terminator is $\downarrow$, there is no limit on the number of
-$\square$ symbols that can be inserted. Indeed, the main thread will not
-be part of the MEM in this case. As mentioned in
-section~\ref{sec:trans_from_up}, the weighted generating function of the
-$\downarrow$ symbol is $pz(1-(1-\mu/3)^N)$. The probability that it first
-occurs at position $n+1$ after the $\Downarrow$ symbol is the probability
-that the first $n$ nucleotides are correct -- each with probability $q$
--- and that at least one alternative thread survives -- probability
-$\xi_{n,N}$. The weighted generating function of the transition from
-$\Downarrow$ to $\downarrow$ is thus
+If the next terminator is $\downarrow$ we have to dinstinguish two cases.
+If there are $\gamma-1$ or fewer error-free nucleotides, we must only make
+sure that there is no $\uparrow$ symbol among them. The weighted
+generating function is
 \begin{equation}
-\tilde{A}_\infty(z) = \tilde{\omega} z
-  \sum_{i=0}^\infty \xi_{i,N} \cdot (qz)^i.
+\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
+\end{equation}
+
+There can also be more than $\gamma-1$ error-free nucleotides, but in this
+case, we must make sure that at least one potentially masking thread masks
+the main thread.
+
+\begin{definition}
+$\alpha_i$ is the probability that the first thread has at least one
+failure in $i$ error-free nucleotides, followed by a sequencing error,
+\textit{i.e.}
+\begin{equation}
+\alpha_i = 1 - (1-\mu)^i\mu/3
+\end{equation}
+\end{definition}
+
+The probability that at least one duplicate thread masks the main thread
+if the first error occurs at position $i$ is thus $1 - \alpha_{i-1}^N$.
+The weighted generatin function is
+\begin{equation}
+pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
+\end{equation}
+
+Finally, the weighted generating function of the transition from
+$\Downarrow$ to $\downarrow$ is
+\begin{equation}
+\tilde{A}_\infty(z) =
+\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
+pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
 \end{equation}
@@
 \begin{eqnarray}
-\alpha_i &=& 1-(1-\mu)^i\mu/3 \\
 \gamma_i &=& 1-(1-\mu)^i \\
@@
 We have assumed $m>0$, but the sum above can start at $m=0$ because
-$\xi_{i,0}=0$. So the weighted generating function of the transition from
+$\xi_{i,0}=0$. The weighted generating function of the transition from
 $\downarrow$ to $(\sim)_i$ is
@@
 \end{equation}
+
+\begin{remark}
+When $N = 1$, $V_i = 0$ and none of the states $(\sim)_i$ is accessible.
+The symbol $\sim$ can exist only if there is more than one duplicate.
+\end{remark}
 
@@
 
-In order to compute the weighted generating functions of the other
-transitions, first observe that
+%In order to compute the weighted generating functions of the other
+%transitions, first observe that
+%\begin{equation}
+%V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
+%\end{equation}
+%So the probability that an error-free stretch of length $i \geq 1$
+%following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
+%$\uparrow$ symbol is
+%\begin{equation}
+%1 - \sum_{j=1}^i V_j+W_j = 1 - \frac{\alpha_i^N -
+%\alpha_0^N}{1-(1-\mu/3)^N}.
+%\end{equation}
+%
+%The probability that the next terminator after a $\downarrow$ symbol is a
+%$\Downarrow$ symbol at position $j$ is the probability of an error-free
+%stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
+%$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
+%by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
+%weighted generating function of the transition from $\downarrow$ to
+%$\Downarrow$ is
+%\begin{equation}
+%B_\infty(z) = \omega z \sum_{i=0}^\infty
+%\Big(1 - \frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} \Big) (qz)^i.
+%\end{equation}
+
+The probability that there is neither $\sim$ nor $\uparrow$ symbol is just
+the probability that at least one stricty masking thread is still alive
 \begin{equation}
-V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
-\end{equation}
-So the probability that an error-free stretch of length $i \geq 1$
-following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
-$\uparrow$ symbol is
-\begin{equation}
-\sum_{j=1}^i V_j+W_j = \frac{\alpha_i^N -
-\alpha_0^N}{1-(1-\mu/3)^N}.
+\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} = \frac{1-\alpha_i^N}{1-(1-\mu/3)^N}
 \end{equation}
 
-The probability that the next terminator after a $\downarrow$ symbol is a
-$\Downarrow$ symbol at position $j$ is the probability of an error-free
-stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
-$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
-by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
-weighted generating function of the transition from $\downarrow$ to
+So the weighted generating function of the transition from $\downarrow$ to
 $\Downarrow$ is
 \begin{equation}
-B_\infty(z) = \omega z \cdot \Big(1 + \sum_{i=1}^\infty
-\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
+B_\infty(z) = \omega z\sum_{i=0}^\infty
+\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
@@
 \begin{equation}
-\tilde{B}_\infty(z) = \tilde{\omega} z \cdot \Big(1 + \sum_{i=1}^\infty
-\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
+\tilde{B}_\infty(z) = \tilde{\omega} z\sum_{i=0}^\infty
+\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
@@
 \label{eq:miscprob}
-\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
-\cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
+P_{m,j} = \sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm}
+(1-(1-\mu)^j)^{n-m} \cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
 \end{equation}
@@
 \begin{eqnarray}
-y_{j,j+i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
+y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
 \big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
@@
 \begin{equation}
-\beta_{j,i} = 1 - (1-\mu)^j\mu/3 + (1-\mu)^i(1-\mu/3).
+\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
 \end{equation}
@@
 
-For the $\downarrow$ symbol, the weighted generating function is simply
+For the $\downarrow$ symbol, the weighted generating function is
 \begin{equation}
-T_{\downarrow}(z) = \sum_{i=0}^\infty (qz)^i = \frac{1}{1-qz}.
+T_{\downarrow}(z) = \sum_{i=0}^\infty
+\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
@@
 \begin{block}{c[c]}
+\DN & T_\Downarrow(z) \\
 \dn & T_\downarrow(z) \\
-\DN & T_\Downarrow(z) \\
 \eq{1} & T_{(\sim)_1}(z) \\
@@
 \eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & y_{1,\gamma-1}(z) \\
+    \ldots & y_{1,\gamma-2}(z) \\
 \eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & y_{2,\gamma-1}(z) \\
+    \ldots & y_{2,\gamma-3}(z) \\
 \vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &

### 2018-04-24 C/kpoly/compute_mem_prob.c
Now truncated polynomials and made N a parameter instead of a constant.

@@ -6,27 +6,27 @@
 
 //  TYPE DECLARATIONS  //
 
-typedef unsigned int  uint;   // Just a shortcut.
-
 // Struct declarations and definitions //
 
-typedef struct kpoly_t  kpoly_t;
-typedef struct mat_t    mat_t;
-typedef struct mono_t   mono_t;
+typedef struct trunc_pol_t  trunc_pol_t;
+typedef struct matrix_t     matrix_t;
+typedef struct monomial_t   monomial_t;
 
-struct mono_t {
-   size_t deg;
-   double coeff;
+// TODO : explain that monomials must not be just deg and coeff //
+
+struct monomial_t {
+   size_t deg;           // Degree of the coefficent.
+   double coeff;         // Value of the coefficient
 };
 
-struct kpoly_t {
-   mono_t mono;              // Monomer (if applicable).
-   double coeff[];           // Terms of the polynomial.
+struct trunc_pol_t {
+   monomial_t mono;      // Monomial (if applicable).
+   double coeff[];       // Terms of the polynomial.
 };
 
-struct mat_t {
-   const size_t    dim;      // Column / row number.
-         kpoly_t * term[];   // Terms of the matrix.
+struct matrix_t {
+   const size_t    dim;  // Column / row number.
+   trunc_pol_t * term[]; // Terms of the matrix.
 };
 
 
@@ -40,13 +40,11 @@ size_t    HIGH;    // Proxy for infinite degree polynomials.
 
 double    P;       // Probability of a read error.
 double    U;       // Divergence rate between duplicates.
-double    OMEGA;   // Probability of double-down arrow.
-double   _OMEGA;   // Probability of down arrow.
 
-size_t    KSZ;     // Size of the 'kpoly_t' struct.
+size_t    KSZ;     // Size of the 'trunc_pol_t' struct.
 
-kpoly_t * TEMP = NULL;        // Matrix multipliciation.
-kpoly_t * ARRAY[1024] = {0};  // Store the results (indexed by N).
+trunc_pol_t * TEMP = NULL;        // For matrix multipliciation.
[…]

### 2018-04-25 C/compute_mem_prob/unittest/runtests.c
librarified the code. Not tested.

@@ -11,11 +11,11 @@ main(
 {
 
    // Import test cases.
-   extern test_case_t test_cases_dunno[];
+   extern test_case_t test_cases_compute_mem_prob[];
 
    // Register test cases.
    const test_case_t *list_of_test_cases[] = {
-      test_cases_dunno,
+      test_cases_compute_mem_prob,
       NULL,
    };
 

### 2018-04-25 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -1,0 +1,19 @@
+#include "unittest.h"
+#include "compute_mem_prob.c"
+
+void
+test_trunc_pol_mult
+(void)
+{
+
+   test_assert(0);
+
+   return;
+
+}
+
+// Test cases for export.
+const test_case_t test_cases_compute_mem_prob[] = {
+   {"compute_mem_prob/kpoly_mult",      test_trunc_pol_mult},
+   {NULL, NULL},
+};

### 2018-04-25 C/kpoly/kpoly.c

@@ -1,513 +1,0 @@
-#include <math.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <strings.h>
-
-
-typedef unsigned int uint;
-
-typedef struct kpoly_t kpoly_t;
-typedef struct mat_t mat_t;
-typedef struct mono_t mono_t;
-
-struct mono_t {
-  size_t deg;
-  double coeff;
-};
-
-struct kpoly_t {
-  const size_t k;            // Max degree of the polynomial.
-        mono_t mono;         // Monomer (if applicable).
-        double coeff[];      // Terms of the polynomial.
-};
-
-struct mat_t {
-  const size_t    dim;       // Column / row number.
-  const size_t    k;         // Max degree of the polynomials.
-        kpoly_t * term[];    // Terms of the matrix.
-};
-
-
-
-void critical_error (char *, int) __attribute__ ((noreturn));
-
-
-void print_kpoly (kpoly_t *P) {
-   if (P == NULL) {
-      fprintf(stderr, "0\n");
-      return;
-   }
-   fprintf(stderr, "%f", P->coeff[0]);
-   fprintf(stderr, " + %fz", P->coeff[1]);
-   for (int i = 2 ; i <= P->k ; i++) {
-      fprintf(stderr, " + %fz^%d", P->coeff[i], i);
-   }
-   fprintf(stderr, "\n");
-}
-
-
-void
-critical_error
-(
-   char * msg,
-   int    lineno
-)
-{
-   fprintf(stderr, "error line %d: %s\n", lineno, msg);
-   abort();
-}
-
-
-void
-destroy_mat
-(
-   mat_t *mat
-)
-{
-
-   size_t dim = mat->dim;
-   for (int i = 0 ; i < dim*dim ; i++) free(mat->term[i]);
-   free(mat);
-
-}
-
-
-kpoly_t *
-new_zero_kpoly
-(
-   size_t k
-)
-{
-
-   // Initialize to zero.
-   kpoly_t *new = calloc(1, sizeof(kpoly_t) + (k+1) * sizeof(double));
-
-   if (new == NULL) critical_error("memory error", __LINE__);
-
-   // The size of the k-polynomial is set upon
-   // creation and must never change afterwards.
-   *(size_t *)&new->k = k;
-
-   return new;
-
-}
-
-
-mat_t *
-new_null_matrix
-(
[…]

### 2018-04-25 C/kpoly/memprob.c

@@ -1,575 +1,0 @@
-#include <math.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <strings.h>
-
-
-//  TYPE DECLARATIONS  //
-
-typedef unsigned int  uint;   // Just a shortcut.
-
-// Struct declarations and definitions //
-
-typedef struct kpoly_t  kpoly_t;
-typedef struct mat_t    mat_t;
-typedef struct mono_t   mono_t;
-
-struct mono_t {
-   size_t deg;
-   double coeff;
-};
-
-struct kpoly_t {
-   mono_t mono;              // Monomer (if applicable).
-   double coeff[];           // Terms of the polynomial.
-};
-
-struct mat_t {
-   const size_t    dim;      // Column / row number.
-         kpoly_t * term[];   // Terms of the matrix.
-};
-
-
-// GLOBAL VIARABLES //
-
-double    P;     // Probability of a read error.
-double    U;     // Divergence rate between duplicates.
-size_t    K;     // Max degree of the polynomials (read size).
-
-size_t    KSZ;   // Size of the 'kpoly_t' struct.
-
-kpoly_t * TEMP = NULL;        // Matrix multipliciation.
-kpoly_t * ARRAY[1024] = {0};  // Store the results.
-
-int       ERRNO; // Error codes.
-
-
-
-// FUNCTION DEFINITIONS //
-
-int
-memprob_init
-(
-   double p,
-   double u,
-   size_t k
-)
-{
-
-   P = p;
-   U = u;
-   K = k; 
-
-   KSZ = sizeof(kpoly_t) + (K+1) * sizeof(double);
-
-   for (int i = 0 ; i < 1024 ; i++) free(ARRAY[i]);
-   bzero(ARRAY, 1024 * sizeof(kpoly_t *));
-
-   free(TEMP);
-   TEMP = calloc(1, KSZ);
-
-   ERRNO = TEMP == NULL ? __LINE__ : 0;
-
-   // Return 1 on error, 0 on success.
-   return TEMP != NULL;
-
-}
-
-
-
-kpoly_t *
-new_zero_kpoly (void)
-{
-   return calloc(1, KSZ);
-}
-
-
-
-kpoly_t *
-new_kpoly_A
-(
-   size_t n,    // Degree of A.
-   uint   N     // Number of duplicates.
-)
-{
-
-   if (n > K || n == 0) {
[…]

### 2018-04-25 C/kpoly/unittest/unittests_dunno.c

@@ -1,19 +1,0 @@
-#include "unittest.h"
-#include "memmatrix.c"
-
-void
-test_kpoly_mult
-(void)
-{
-
-   test_assert(0);
-
-   return;
-
-}
-
-// Test cases for export.
-const test_case_t test_cases_dunno[] = {
-   {"dunno/kpoly_mult",         test_kpoly_mult},
-   {NULL, NULL},
-};

### 2018-04-26 C/compute_mem_prob/compute_mem_prob.c
Added first tests.

@@ -4,9 +4,50 @@
 #include <strings.h>
 
 
-//  TYPE DECLARATIONS  //
+// MACROS //
 
-// Struct declarations and definitions //
+#define LIBNAME "compute_mem_prob"
+#define VERSION "0.9 04-27-2018"
+
+#define YES 1
+#define NO  0
+
+#define SUCCESS 1
+#define FAILURE 0
+
+// Maximum allowed number of duplicates.
+#define MAXN 1024
+
+// Compute omega and tilde-omega.
+#define  OMEGA ( P*pow(1.0-U/3, N) )
+#define _OMEGA ( P*(1-pow(1.0-U/3, N)) )
+
+// Creation of a new 'trunc_pol_t' is just a call to 'calloc()'.
+#define new_zero_trunc_pol() calloc(1, KSZ)
+
+// Prob that one of m altnerative threads survives i steps.
+#define xi(i,m) ( 1.0 - pow( 1.0 - pow(1.0-U,(i)), (m) ))
+
+// Calculation intermediates (one index).
+#define aN(i) pow( 1.0 - pow(1.0-U,(i)) * U/3.0, N )
+#define gN(i) pow( 1.0 - pow(1.0-U,(i)), N )
+#define dN(i) pow( 1.0 - (1.0 - U + U*U/3.0) * pow(1.0-U,(i)), N )
+
+// Calculation intermediates (two indices).
+#define bN(j,i) pow( 1.0 - pow(1.0-U,(j))*U/3.0 - \
+                           pow(1.0-U,(i))*(1.0-U/3.0), N)
+
+
+// Macro to simplify error handling.
+#define handle_memory_error(x) do { \
+   if ((x) == NULL) { \
+      warning("memory_error", __func__, __LINE__); \
+      ERRNO = __LINE__; \
+      goto in_case_of_failure; \
+}} while (0)
+
+
+//  TYPE DECLARATIONS  //
 
 typedef struct trunc_pol_t  trunc_pol_t;
 typedef struct matrix_t     matrix_t;
@@ -30,6 +71,7 @@ struct matrix_t {
 };
 
 
+
 // GLOBAL VARIABLES / CONSTANTS //
 
 size_t    G;       // Minimum size of MEM seeds.
@@ -44,42 +86,42 @@ double    U;       // Divergence rate between duplicates.
 size_t    KSZ;     // Size of the 'trunc_pol_t' struct.
 
 trunc_pol_t * TEMP = NULL;        // For matrix multipliciation.
[…]

### 2018-04-26 C/compute_mem_prob/compute_mem_prob.h

@@ -1,0 +1,13 @@
+#ifndef __COMPUTE_MEM_PROB_HEADER
+#define __COMPUTE_MEM_PROB_HEADER
+
+// Visible functions.
+
+void   clean_mem_prob (void);
+double compute_mem_prob(size_t, size_t);
+int    get_mem_prob_error_code (void);
+char * get_mem_prob_error_msg (void);
+void   reset_mem_prob_error (void);
+int    set_params_mem_prob (size_t, size_t, double, double);
+
+#endif

### 2018-04-26 C/compute_mem_prob/unittest/lib/unittest.c

@@ -1,39 +1,30 @@
 #include "unittest.h"
 
-// Standard malloc() and realloc() //
-extern void *__libc_malloc(size_t);
-extern void *__libc_realloc(void *, size_t);
-extern void *__libc_calloc(size_t, size_t);
+//     Constants     //
+#define UTEST_BUFFER_SIZE 1024
+#define MAX_N_ERROR_MSG 20
+
 // Private functions // 
 void   redirect_stderr (void);
-void * run_testset (void *); 
+void * run_test (void *); 
 char * sent_to_stderr (void);
 void   unit_test_clean (void);
 void   unit_test_init (void);
 void   unredirect_stderr (void);
 void   test_case_clean (void);
 void   test_case_init (void);
-void * fail_countdown_calloc (size_t, size_t);
-void * fail_countdown_malloc (size_t);
-void * fail_countdown_realloc (void *, size_t);
-void * fail_prone_calloc (size_t, size_t);
-void * fail_prone_malloc (size_t);
-void * fail_prone_realloc (void *, size_t);
-
 
 //     Global variables     //
-void *(*MALLOC_CALL)(size_t) = __libc_malloc;
-void *(*REALLOC_CALL)(void *, size_t) = __libc_realloc;
-void *(*CALLOC_CALL)(size_t, size_t) = __libc_calloc;
-
 static double ALLOC_ERR_PROB;
 static int    ALLOC_FAIL_COUNTER;
 static FILE * DEBUG_DUMP_FILE;
-static int    N_ERROR_MESSAGES;
+static int    N_ERROR_MSG;
 static int    ORIG_STDERR_DESCRIPTOR;
 static char * STDERR_BUFFER;
 static int    STDERR_OFF;
 static int    TEST_CASE_FAILED;
+
+static int    SHOWALL = 0;
 
 
 //     Function definitions     //
@@ -78,56 +69,9 @@ terminate_thread
 
 }
 
-int
-run_unittest
-(
-         int            argc,
-         char        ** argv,
-   const test_case_t  * test_case_list[]
-)
-{
-
-   // Initialize test set. //
-   unit_test_init();
-
-   pthread_t tid;
-
-   int nbad = 0;
-
-   // Run test cases in sequential order.
-   for (int j = 0 ; test_case_list[j] != NULL ; j++) {
-
-      const test_case_t *tests = test_case_list[j];
[…]

### 2018-04-26 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -2,6 +2,97 @@
 #include "compute_mem_prob.c"
 
 void
+test_set_params_mem_prob
+(void)
+{
+
+   int success;
+
+   // Case 1. //
+   success = set_params_mem_prob(17, 50, 0.01, 0.05);
+   test_assert(success);
+
+   test_assert(G == 17);
+   test_assert(K == 50);
+   test_assert(P == 0.01);
+   test_assert(U == 0.05);
+
+   test_assert(HIGH == 50);
+   test_assert(KSZ == sizeof(trunc_pol_t) + 51*sizeof(double));
+
+   for (int i = 0 ; i < MAXN ; i++) {
+      test_assert(ARRAY[i] == NULL);
+   }
+
+   // Case 2. //
+   set_params_mem_prob(50, 20, 0.99, 0.95);
+   test_assert(success);
+
+   test_assert(G == 50);
+   test_assert(K == 20);
+   test_assert(P == 0.99);
+   test_assert(U == 0.95);
+
+   test_assert(HIGH == 50);
+   test_assert(KSZ == sizeof(trunc_pol_t) + 21*sizeof(double));
+
+   for (int i = 0 ; i < MAXN ; i++) {
+      test_assert(ARRAY[i] == NULL);
+   }
+
+   return;
+
+}
+
+
+void
+test_error_set_params_mem_prob
+(void)
+{
+
+   int success;
+
+   // Case 1. //
+   redirect_stderr();
+   success = set_params_mem_prob(17, 50, 0.00, 0.05);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
+   // Case 2. //
+   redirect_stderr();
+   success = set_params_mem_prob(17, 50, 0.01, 0.00);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
+   // Case 3. //
+   redirect_stderr();
+   success = set_params_mem_prob(17, 50, 1.00, 0.05);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
+   // Case 4. //
+   redirect_stderr();
+   success = set_params_mem_prob(17, 50, 0.01, 1.00);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
[…]

### 2018-04-27 C/compute_mem_prob/compute_mem_prob.c
Added tests and checking for initialization.

@@ -22,9 +22,6 @@
 #define  OMEGA ( P*pow(1.0-U/3, N) )
 #define _OMEGA ( P*(1-pow(1.0-U/3, N)) )
 
-// Creation of a new 'trunc_pol_t' is just a call to 'calloc()'.
-#define new_zero_trunc_pol() calloc(1, KSZ)
-
 // Prob that one of m altnerative threads survives i steps.
 #define xi(i,m) ( 1.0 - pow( 1.0 - pow(1.0-U,(i)), (m) ))
 
@@ -74,22 +71,20 @@ struct matrix_t {
 
 // GLOBAL VARIABLES / CONSTANTS //
 
-size_t    G;       // Minimum size of MEM seeds.
-size_t    N;       // Number of duplicate sequences.
-size_t    K;       // Max degree of the polynomials (read size).
+static size_t    G = 0;       // Minimum size of MEM seeds.
+static size_t    K = 0;       // Max degree of the polynomials (read size).
 
-size_t    HIGH;    // Proxy for infinite degree polynomials.
+static size_t    HIGH = 0;    // Proxy for infinite degree polynomials.
 
-double    P;       // Probability of a read error.
-double    U;       // Divergence rate between duplicates.
+static double    P = 0.0;     // Probability of a read error.
+static double    U = 0.0;     // Divergence rate between duplicates.
 
-size_t    KSZ;     // Size of the 'trunc_pol_t' struct.
+static size_t    KSZ = 0;     // Size of the 'trunc_pol_t' struct.
 
-trunc_pol_t * TEMP = NULL;        // For matrix multipliciation.
-trunc_pol_t * ARRAY[MAXN] = {0};  // Store the results (indexed by N).
+static trunc_pol_t * TEMP = NULL;        // For matrix multipliciation.
+static trunc_pol_t * ARRAY[MAXN] = {0};  // Store the results (indexed by N).
 
-int       ERRNO = 0;
-char      ERRMSG[1024] = {0};
+static int       ERRNO = 0;
 
 // Error message.
 const char internal_error[] =
@@ -101,7 +96,6 @@ const char internal_error[] =
 
 // VISLBLE IO and error report functions.
 int    get_mem_prob_error_code (void) { return ERRNO; }
-char * get_mem_prob_error_msg (void) { return ERRMSG; }
[…]

### 2018-04-27 C/compute_mem_prob/compute_mem_prob.h

@@ -6,7 +6,6 @@
 void   clean_mem_prob (void);
 double compute_mem_prob(size_t, size_t);
 int    get_mem_prob_error_code (void);
-char * get_mem_prob_error_msg (void);
 void   reset_mem_prob_error (void);
 int    set_params_mem_prob (size_t, size_t, double, double);
 

### 2018-04-27 C/compute_mem_prob/unittest/lib/unittest.c

@@ -304,11 +304,9 @@ fail_debug_dump
    const char * function
 )
 // When a test fails, write a gdb file with a breakpoint at
-// the function that failed, and another one at the line
-// where the failured happened.
+// the line where the failured happened.
 {
    fprintf(DEBUG_DUMP_FILE, "b %s:%d\n", file, lineno);
-   fprintf(DEBUG_DUMP_FILE, "b %s\n", function);
    fflush(DEBUG_DUMP_FILE);
 }
 

### 2018-04-27 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -40,6 +40,8 @@ test_set_params_mem_prob
       test_assert(ARRAY[i] == NULL);
    }
 
+   clean_mem_prob();
+
    return;
 
 }
@@ -80,6 +82,20 @@ test_error_set_params_mem_prob
    test_assert(!success);
    test_assert_stderr("[compute_mem_prob] error in function `set_");
 
+   // Case 5. //
+   redirect_stderr();
+   success = set_params_mem_prob(0, 50, 0.01, 1.00);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
+   // Case 6. //
+   redirect_stderr();
+   success = set_params_mem_prob(17, 0, 0.01, 1.00);
+   unredirect_stderr();
+   test_assert(!success);
+   test_assert_stderr("[compute_mem_prob] error in function `set_");
+
    // Test memory error //
    set_alloc_failure_rate_to(1.0);
    redirect_stderr();
@@ -93,20 +109,184 @@ test_error_set_params_mem_prob
 
 
 void
-test_trunc_pol_mult
+test_uninitialized_error
 (void)
 {
 
-   test_assert(0);
+   // Do not call 'set_params_mem_prob()'.
+   redirect_stderr();
+   double x = compute_mem_prob(5, 20);
+   unredirect_stderr();
+   test_assert_stderr("[compute_mem_prob] error in function `comp");
+   test_assert(x != x);
 
    return;
 
 }
 
+
+void
+test_new_zero_trunc_pol
+(void)
+{
+
+   size_t ksz = 50;
+   int success = set_params_mem_prob(17, ksz, 0.01, 0.05);
+   test_assert(success);
+
+   trunc_pol_t *a = new_zero_trunc_pol();
+   test_assert_critical(a);
+
+   test_assert(a->mono.deg == 0);
+   test_assert(a->mono.coeff == 0);
+   for (int i = 0 ; i <= ksz ; i++) {
+      test_assert(a->coeff[i] == 0);
+   }
+
+   free(a);
+   clean_mem_prob();
+
+   return;
+
+}
+
+
+void
+test_trunc_pol_mult
+(void)
+{
+   
+   // Test multiplications between zero polynomials.
+
+   size_t ksz = 50;
+   set_params_mem_prob(17, ksz, 0.01, 0.05);
[…]

### 2018-05-01 C/compute_mem_prob/compute_mem_prob.c
Added test cases for truncated polynomials.

@@ -232,9 +232,20 @@ new_trunc_pol_A
       ERRNO = __LINE__;
       goto in_case_of_failure;
    }
-
+   
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
+
+   if (N == 0) {
+      // In the special case N = 0, A(z) = pz and ~A(z) = 0.
+      if (!tilde) {
+         new->mono.deg = 1;
+         new->mono.coeff = P;
+         new->coeff[1] = P;
+         return new;
+      }
+      return new;
+   }
 
    // See definition of polynomial A.
    const int d = deg <= G ? deg : G;
@@ -276,6 +287,9 @@ new_trunc_pol_B
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
 
+   // Polynomial B is null when N = 0.
+   if (N == 0) return new;
+
    // See definition of polynomial B.
    const double cst = tilde ? _OMEGA : OMEGA;
    const double denom = 1.0 - pow(1-U/3.0, N);
@@ -303,9 +317,6 @@ new_trunc_pol_C
 )
 {
 
-   // Avoid division by zero when N = 1 (not a failure).
-   if (N == 1) return NULL;
-
    if (deg > K || deg == 0) {
       warning(internal_error, __func__, __LINE__);
       ERRNO = __LINE__;
@@ -314,6 +325,9 @@ new_trunc_pol_C
 
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
+
+   // Avoid division by zero when N < 2 (not a failure).
+   if (N < 2) return new;
 
    // See definition of polynomial C.
    const int j = G - deg;
@@ -349,6 +363,9 @@ new_trunc_pol_D
       goto in_case_of_failure;
    }
 
+   // NB: The special case N = 0 is implicit for the
+   // polynomial D (it implies omega = p and ~omega = 0).
+   
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
 
@@ -387,7 +404,11 @@ new_trunc_pol_u
 
    // See definition of polynomial u.
    new->mono.deg = deg;
-   new->mono.coeff = (xi(deg-1,N) - xi(deg,N)) * pow(1.0-P,deg);
+   // The case N = 0 involves the term 'xi(0,0)', which should be
+   // equal to 0 but is equal to 1 because 'pow(0,0)' is 1 as per
[…]

### 2018-05-01 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -8,9 +8,9 @@ test_set_params_mem_prob
 
    int success;
 
-   // Case 1. //
+   // Case 1.
    success = set_params_mem_prob(17, 50, 0.01, 0.05);
-   test_assert(success);
+   test_assert_critical(success);
 
    test_assert(G == 17);
    test_assert(K == 50);
@@ -24,9 +24,9 @@ test_set_params_mem_prob
       test_assert(ARRAY[i] == NULL);
    }
 
-   // Case 2. //
+   // Case 2.
    set_params_mem_prob(50, 20, 0.99, 0.95);
-   test_assert(success);
+   test_assert_critical(success);
 
    test_assert(G == 50);
    test_assert(K == 20);
@@ -54,53 +54,54 @@ test_error_set_params_mem_prob
 
    int success;
 
-   // Case 1. //
+   // Case 1.
    redirect_stderr();
    success = set_params_mem_prob(17, 50, 0.00, 0.05);
    unredirect_stderr();
    test_assert(!success);
    test_assert_stderr("[compute_mem_prob] error in function `set_");
 
-   // Case 2. //
+   // Case 2.
    redirect_stderr();
    success = set_params_mem_prob(17, 50, 0.01, 0.00);
    unredirect_stderr();
    test_assert(!success);
    test_assert_stderr("[compute_mem_prob] error in function `set_");
 
-   // Case 3. //
+   // Case 3.
    redirect_stderr();
    success = set_params_mem_prob(17, 50, 1.00, 0.05);
    unredirect_stderr();
    test_assert(!success);
    test_assert_stderr("[compute_mem_prob] error in function `set_");
 
-   // Case 4. //
+   // Case 4.
    redirect_stderr();
    success = set_params_mem_prob(17, 50, 0.01, 1.00);
    unredirect_stderr();
    test_assert(!success);
    test_assert_stderr("[compute_mem_prob] error in function `set_");
 
-   // Case 5. //
+   // Case 5.
    redirect_stderr();
    success = set_params_mem_prob(0, 50, 0.01, 1.00);
    unredirect_stderr();
    test_assert(!success);
[…]

### 2018-05-03 C/compute_mem_prob/compute_mem_prob.c
Wrote most test cases.

@@ -469,10 +469,12 @@ new_trunc_pol_w
       goto in_case_of_failure;
    }
 
-   // TODO: treat special case N = 0. //
-
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
+
+   // In the special case N = 0, the polynomial w is
+   // undefined, but we return a zero polynomial.
+   if (N == 0) return new;
 
    // See definition of polynomial w.
    new->mono.deg = deg;
@@ -493,25 +495,23 @@ in_case_of_failure:
 trunc_pol_t *
 new_trunc_pol_y
 (
-   const int j,      // Initial state.
-   const int i,      // Degree of the polynomial.
+   const size_t j,      // Initial state.
+   const size_t i,      // Degree of the polynomial.
    const size_t N    // Number of duplicates.
 )
 {
 
-   if (i > K || i >= G || i == 0) {
+   if (j == 0 || j >= G-1 || i > K || i >= G-1 || i == 0 || i+j > G-1) {
       warning(internal_error, __func__, __LINE__);
       ERRNO = __LINE__;
       goto in_case_of_failure;
    }
 
-   // Avoid division by zero when N = 1 (not a failure).
-   if (N == 1) return NULL;
-
-   // TODO: treat special case N = 0. //
-
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
+
+   // Avoid division by zero when N < 2 (not a failure).
+   if (N < 2) return new;
 
    // See definition of polynomial y.
    new->mono.deg = i;
@@ -537,6 +537,10 @@ new_trunc_pol_T_down
 
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
+
+   // In the special case N = 0, the polynomial T is
+   // undefined, but we return a zero polynomial.
+   if (N == 0) return new;
 
    const double denom = 1.0 - pow(1-U/3.0, N);
    double pow_of_q = 1.0;
@@ -564,10 +568,17 @@ new_trunc_pol_T_double_down
    trunc_pol_t *new = new_zero_trunc_pol();
    handle_memory_error(new);
 
-   double pow_of_q = 1.0;
-   for (int i = 0 ; i <= G-1 ; i++) {
-      new->coeff[i] = xi(i,N) * pow_of_q;
-      pow_of_q *= (1-P);
+   if (N == 0) {
[…]

### 2018-05-03 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -283,6 +283,22 @@ test_trunc_pol_mult
 
 
 void
+test_matrix_mult
+(void)
+{
+   test_assert(0);
+}
+
+
+void
+test_error_matrix_mult
+(void)
+{
+   test_assert(0);
+}
+
+
+void
 test_new_trunc_pol_A
 (void)
 {
@@ -379,7 +395,7 @@ test_error_new_trunc_pol_A
 
    set_alloc_failure_rate_to(1);
    redirect_stderr();
-   new_trunc_pol_A(18, 2, NO);
+   A = new_trunc_pol_A(18, 2, NO);
    unredirect_stderr();
    reset_alloc();
 
@@ -483,7 +499,7 @@ test_error_new_trunc_pol_B
 
    set_alloc_failure_rate_to(1);
    redirect_stderr();
-   new_trunc_pol_B(18, 2, NO);
+   B = new_trunc_pol_B(18, 2, NO);
    unredirect_stderr();
    reset_alloc();
 
@@ -612,7 +628,7 @@ test_error_new_trunc_pol_C
 
    set_alloc_failure_rate_to(1);
    redirect_stderr();
-   new_trunc_pol_C(18, 2, NO);
+   C = new_trunc_pol_C(18, 2, NO);
    unredirect_stderr();
    reset_alloc();
 
@@ -721,7 +737,7 @@ test_error_new_trunc_pol_D
 
    set_alloc_failure_rate_to(1);
    redirect_stderr();
-   new_trunc_pol_D(18, 2, NO);
+   D = new_trunc_pol_D(18, 2, NO);
    unredirect_stderr();
    reset_alloc();
 
@@ -808,7 +824,7 @@ test_error_new_trunc_pol_u
 
    set_alloc_failure_rate_to(1);
    redirect_stderr();
-   new_trunc_pol_u(10, 2);
+   u = new_trunc_pol_u(10, 2);
    unredirect_stderr();
    reset_alloc();
 
@@ -828,7 +844,7 @@ test_new_trunc_pol_v
    int success = set_params_mem_prob(17, 50, 0.01, 0.05);
    test_assert_critical(success);
 
-   // Test a u polynomial of degree 10 with N = 2.
+   // Test a v polynomial of degree 10 with N = 2.
    trunc_pol_t *v = new_trunc_pol_v(10, 2);
    test_assert_critical(v != NULL);
 
@@ -895,7 +911,7 @@ test_error_new_trunc_pol_v
 
[…]

### 2018-05-03 C/compute_mem_prob/compute_mem_prob.c
Improved Makefile.

@@ -821,17 +821,15 @@ trunc_pol_mult
 )
 {
 
+   // Erase destination.
+   bzero(dest, KSZ);
 
    // If any of the two k-polynomials is zero,
-   // set 'dest' to zero and return 'NULL'.
-   if (a == NULL || b == NULL) {
-      bzero(dest, KSZ);
-      return NULL;
-   }
+   // set 'dest' to zero and return 'NULL' (not dest).
+   if (a == NULL || b == NULL) return NULL;
 
    if (a->mono.coeff && b->mono.coeff) {
       // Both are monomials, just do one multiplication.
-      bzero(dest, KSZ);
       // If degree is too high, all coefficients are zero.
       if (a->mono.deg + b->mono.deg > K) return NULL;
       // Otherwise do the multiplication.
@@ -841,13 +839,11 @@ trunc_pol_mult
    }
    else if (a->mono.coeff) {
       // 'a' is a monomial, do one "row" of multiplications.
-      bzero(dest, KSZ);
       for (int i = a->mono.deg ; i <= K ; i++)
          dest->coeff[i] = a->mono.coeff * b->coeff[i-a->mono.deg];
    }
    else if (b->mono.coeff) {
       // 'b' is a monomial, do one "row" of multiplications.
-      bzero(dest, KSZ);
       for (int i = b->mono.deg ; i <= K ; i++)
          dest->coeff[i] = b->mono.coeff * a->coeff[i-b->mono.deg];
    }

### 2018-05-03 MEM.tex
About to swap v and w.

@@
 
+EXPLAIN THAT THE TAIL IS AN UNTERMINATED SEGMENT.
+
 \subsection{The seeding alphabet}
 
-By nature, reads are sequences of \texttt{A}, \texttt{T}, \texttt{G} and
-\texttt{C}, but for the purpose of the seeding process, it is more useful
-to consider them as a sequence of symbols telling if the nucleotides match
-the target and~/ or the duplicates. Here, we will consider that reads are
-sequences of the five symbols $\Downarrow$, $\downarrow$, $\uparrow$, $\sim$
-and $\square$.
+In order to compute seeding probabilities, we need to consider that reads
+are sequences of letters from the \emph{seeding alphabet} rather than from
+the nucleotide alphabet. them as a sequence of symbols telling if the
+nucleotides match the target and~/ or the duplicates. Here, we will
+consider that reads are sequences of the five symbols $\Downarrow$,
+$\downarrow$, $\uparrow$, $\sim$ and $\square$.
 
@@
 At a given position of the read, a duplicate thread is said to be
-\emph{strictly masking} if its match streak is strictly longer than the
+\emph{strictly masking} if its success streak is strictly longer than the
 match streak of the target. A duplicate thread is said to be
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[scale=1]{no_3MEM_graph.pdf}
-\caption{\textbf{Transfer graph of a read without 3-MEM}. 
-Some legend.}
-\label{fig:noMEM_graph}
-\end{figure}
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=1]{no_3MEM_graph.pdf}
+%\caption{\textbf{Transfer graph of a read without 3-MEM}. 
+%Some legend.}
+%\label{fig:noMEM_graph}
+%\end{figure}
 
@@
 \end{equation}
+
+\begin{remark}
+If $N = 0$, the states $(\downarrow)_2, \ldots, (\downarrow)_{\gamma-1}$
+are inacessible. Consistently, the mononomial $u_i(z)$ is equal to $qz$ if
+$i = 1$ and $0$ otherwise.
+\end{remark}
 
@@
 
+\begin{remark}
+The term $\alpha_i^N-\alpha_i^N$ is the probability that the last dupicate
+thread fails in one error followed by $i$ error-free nucleotides. The term
+$\gamma_i^N - \delta_{i-1}^N$ is the probability that all threads fail in
+$i$ error-free nucleotides...
+\end{remark}
+
 We have assumed $m>0$, but the sum above can start at $m=0$ because
@@
 \begin{remark}
-When $N = 1$, $V_i = 0$ and none of the states $(\sim)_i$ is accessible.
+When $N = 1$, $V_i = 0$ and the states $(\sim)_i$ are inaccessible.
 The symbol $\sim$ can exist only if there is more than one duplicate.
@@
 
+In the case $N = 1$, reads never contain the symbol $\sim$ and the matrix
+is
+\begin{equation*}
+\begin{blockarray}{cccccc}
+   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
+\begin{block}{c[ccccc]}
+\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_1(z)
+    & \ldots & u_{\gamma-1}(z) \\
+\dn & B_\infty(z) & \tilde{B}_\infty(z) & w_1(z)
+    & \ldots & w_{\gamma-1}(z) \\
+\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &
+    \ldots & 0 \\
+\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 &
+    \ldots & 0  \\
+\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
+\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 \\
+\end{block}
+\end{blockarray}
+\end{equation*}
+
+In the case $N = 0$, the symbol $\downarrow$ also disappears. Some
+polynomials have a particularly simple form in this case, and the
+body of the transfer matrix is
+\begin{equation*}
+\begin{blockarray}{ccc}
+   & \DN & \up{1} \\
+\begin{block}{c[cc]}
+\DN    & pz              & qz \\
+\up{1} & D_{\gamma-1}(z) & 0  \\
+\end{block}
+\end{blockarray}
+\end{equation*}
+
 \section*{Acknowledgements}

### 2018-05-03 C/compute_mem_prob/compute_mem_prob.c
Swapped v and w. Removed head row from M.

@@ -420,7 +420,7 @@ in_case_of_failure:
 
 
 trunc_pol_t *
-new_trunc_pol_v
+new_trunc_pol_w
 (
    const size_t deg,  // Degree of polynomial v.
    const size_t N     // Number of duplicates.
@@ -456,7 +456,7 @@ in_case_of_failure:
 
 
 trunc_pol_t *
-new_trunc_pol_w
+new_trunc_pol_v
 (
    const size_t deg,  // Degree of polynomial w.
    const size_t N     // Number of duplicates.
@@ -739,15 +739,11 @@ new_matrix_M
 )
 {
 
-   const size_t dim = 2*G+2;
+   const size_t dim = 2*G+1;
    matrix_t *M = new_null_matrix(dim);
    handle_memory_error(M);
 
-   // First row.
-   M->term[0*dim+1] = new_zero_trunc_pol();
-   handle_memory_error(M->term[0*dim+1]);
-   M->term[0*dim+1]->coeff[0] = 1.0;
-   M->term[0*dim+1]->mono.coeff = 1.0;
+   // First row is null.
 
    // Second row.
    M->term[1*dim+1] = new_trunc_pol_A(G, N, NO);
@@ -755,11 +751,11 @@ new_matrix_M
    M->term[1*dim+2] = new_trunc_pol_A(HIGH, N, YES);
    handle_memory_error(M->term[1*dim+2]);
    for (int j = 1 ; j <= G-1 ; j++) {
-      M->term[1*dim+(j+G+1)] = new_trunc_pol_u(j, N);
-      handle_memory_error(M->term[1*dim+(j+G+1)]);
+      M->term[1*dim+(j+2)] = new_trunc_pol_u(G-j, N);
+      handle_memory_error(M->term[1*dim+(j+2)]);
    }
-   M->term[1*dim+dim-1] = new_trunc_pol_T_double_down(N);
-   handle_memory_error(M->term[1*dim+dim-1]);
+   M->term[1*dim+0] = new_trunc_pol_T_double_down(N);
+   handle_memory_error(M->term[1*dim+0]);
 
    // Third row.
    M->term[2*dim+1] = new_trunc_pol_B(HIGH, N, NO);
@@ -767,41 +763,39 @@ new_matrix_M
    M->term[2*dim+2] = new_trunc_pol_B(HIGH, N, YES);
    handle_memory_error(M->term[2*dim+2]);
[…]

### 2018-05-03 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -837,93 +837,6 @@ test_error_new_trunc_pol_u
 
 
 void
-test_new_trunc_pol_v
-(void)
-{
-
-   int success = set_params_mem_prob(17, 50, 0.01, 0.05);
-   test_assert_critical(success);
-
-   // Test a v polynomial of degree 10 with N = 2.
-   trunc_pol_t *v = new_trunc_pol_v(10, 2);
-   test_assert_critical(v != NULL);
-
-   const double denom = 1-pow(1-.05/3,2);
-   const double num = pow(1-pow(1-.05,10)*.05/3,2) - \
-      pow(1-pow(1-.05,9)*.05/3,2) - pow(1-pow(1-.05,10),2) + \
-      pow(1-(1-.05+.05*.05/3)*pow(1-.05,9),2);
-   const double target = pow(1-.01,10) * num / denom;
-   test_assert(v->mono.deg == 10);
-   test_assert(fabs(v->mono.coeff-target) < 1e-9);
-   for (int i = 0 ; i <= 50 ; i++) {
-      if (i == 10)
-         test_assert(fabs(v->coeff[i]-target) < 1e-9);
-      else
-         test_assert(v->coeff[i] == 0);
-   }
-
-   // Test the special case N = 0.
-   trunc_pol_t *v0 = new_trunc_pol_v(1, 0);
-   test_assert_critical(v0 != NULL);
-   test_assert(v0->mono.deg == 0);
-   test_assert(v0->mono.coeff == 0);
-   for (int i = 0 ; i <= 50 ; i++) {
-      test_assert(v0->coeff[i] == 0);
-   }
-   
-   free(v);
-   free(v0);
-   clean_mem_prob();
-
-}
-
-
-void
-test_error_new_trunc_pol_v
-(void)
-{
-
-   int success = set_params_mem_prob(17, 50, 0.01, 0.05);
-   test_assert_critical(success);
-
-   trunc_pol_t *v;
-
-   redirect_stderr();
-   v = new_trunc_pol_v(0, 0);
-   unredirect_stderr();
-
-   test_assert(v == NULL);
-   test_assert_stderr("[compute_mem_prob] error in function `new_t");
-
-   redirect_stderr();
-   v = new_trunc_pol_v(0, 2);
-   unredirect_stderr();
-
-   test_assert(v == NULL);
-   test_assert_stderr("[compute_mem_prob] error in function `new_t");
-
-   redirect_stderr();
-   new_trunc_pol_v(51, 2);
[…]

### 2018-05-03 C/compute_mem_prob/compute_mem_prob.c
Full coverage, fixed all known issues.

@@ -205,7 +205,14 @@ in_case_of_failure:
 trunc_pol_t *
 new_zero_trunc_pol
 (void)
+// IMPORTANT: do not call before set_params_mem_prob().
 {
+
+   if (G == 0 || K == 0 || P == 0 || U == 0 || KSZ == 0) {
+      warning("parameters unset: call `set_params_mem_prob'",
+            __func__, __LINE__);
+      goto in_case_of_failure;
+   }
 
    trunc_pol_t *new = calloc(1, KSZ);
    handle_memory_error(new);
@@ -856,7 +863,6 @@ trunc_pol_mult
 }
 
 
-
 void
 trunc_pol_update_add
 (
@@ -868,12 +874,14 @@ trunc_pol_update_add
    // No update when adding zero.
    if (a == NULL) return;
 
+   // No monomial after update.
+   dest->mono = (monomial_t) {0};
+
    for (int i = 0 ; i <= K ; i++) {
       dest->coeff[i] += a->coeff[i];
    }
 
 }
-
 
 
 matrix_t *
@@ -888,7 +896,7 @@ matrix_mult
    if (a->dim != dest->dim || b->dim != dest->dim) {
       warning(internal_error, __func__, __LINE__);
       ERRNO = __LINE__;
-      return NULL;
+      goto in_case_of_failure;
    }
 
    size_t dim = dest->dim;
@@ -898,6 +906,9 @@ matrix_mult
       // Erase destination entry.
       bzero(dest->term[i*dim+j], KSZ);
       for (int m = 0 ; m < dim ; m++) {
+         // By construction, the first NULL term of the row
+         // indicates that all the remaining terms are zero.
+         if (a->term[i*dim+m] == NULL) break;
          trunc_pol_update_add(dest->term[i*dim+j],
             trunc_pol_mult(TEMP, a->term[i*dim+m], b->term[m*dim+j]));
       }
@@ -905,6 +916,9 @@ matrix_mult
    }
 
    return dest;
+
+in_case_of_failure:
+   return NULL;
 
 }
 
@@ -924,7 +938,7 @@ compute_mem_prob // VISIBLE //
    matrix_t *powM1 = NULL;
    matrix_t *powM2 = NULL;
 
-   if (G == 0 || K == 0 || P == 0 || U == 0) {
+   if (G == 0 || K == 0 || P == 0 || U == 0 || KSZ == 0) {
       warning("parameters unset: call `set_params_mem_prob'",
[…]

### 2018-05-03 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -286,7 +286,80 @@ void
 test_matrix_mult
 (void)
 {
-   test_assert(0);
+
+
+   int success = set_params_mem_prob(17, 50, 0.01, 0.05);
+   test_assert_critical(success);
+
+   matrix_t *mat1 = new_zero_matrix(2);
+   matrix_t *mat2 = new_zero_matrix(2);
+   test_assert_critical(mat1 != NULL);
+   test_assert_critical(mat2 != NULL);
+
+   
+   //  | 1   z |    |  0   z  |
+   //  |z^2 z^3|    |2z^2 3z^3|
+
+   // Fill dummy matrices mat1 and mat2.
+   for (int i = 0; i < 4 ; i++) {
+      trunc_pol_t *w1 = mat1->term[i];
+      trunc_pol_t *w2 = mat2->term[i];
+      w1->mono.deg = i;
+      w1->mono.coeff = 1;
+      w1->coeff[w1->mono.deg] = w1->mono.coeff;
+      w2->mono.deg = i;
+      w2->mono.coeff = i;
+      w2->coeff[w2->mono.deg] = w2->mono.coeff;
+   }
+
+   matrix_t *tmp1 = new_zero_matrix(2);
+   test_assert_critical(tmp1 != NULL);
+
+   matrix_mult(tmp1, mat1, mat2);
+
+   const double tmp1_array1[51] = {0,0,0,2};
+   const double tmp1_array2[51] = {0,1,0,0,3};
+   const double tmp1_array3[51] = {0,0,0,0,0,2};
+   const double tmp1_array4[51] = {0,0,0,1,0,0,3};
+
+   for (int i = 0 ; i < 4 ; i++) {
+      test_assert(tmp1->term[i]->mono.deg == 0);
+      test_assert(tmp1->term[i]->mono.coeff == 0);
+   }
+   for (int i = 0 ; i <= 50 ; i++) {
+      test_assert(tmp1->term[0]->coeff[i] == tmp1_array1[i]);
+      test_assert(tmp1->term[1]->coeff[i] == tmp1_array2[i]);
+      test_assert(tmp1->term[2]->coeff[i] == tmp1_array3[i]);
+      test_assert(tmp1->term[3]->coeff[i] == tmp1_array4[i]);
+   }
+
+   matrix_t *tmp2 = new_zero_matrix(2);
+   test_assert_critical(tmp2 != NULL);
+
+   matrix_mult(tmp2, mat2, mat1);
+
+   const double tmp2_array1[51] = {0,0,0,1};
[…]

### 2018-05-13 MEM.tex
Swapped v and w. Removed head row from M.
Fixed errors in the theory. Cleaned symbols to math code.

@@
 \begin{equation}
-D_{\gamma-j}(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
+C^\Downarrow_j(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
 \end{equation}
@@
 \begin{equation}
-\tilde{D}_{\gamma-j}(z) = \tilde{\omega} z
+C^\downarrow_j(z) = \tilde{\omega} z
   \sum_{i=0}^{\gamma-j-1} (qz)^i.
@@
 \begin{equation}
-A_\gamma(z) = \omega z
+A^\Downarrow(z) = \omega z
   \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
@@
 If the next terminator is $\downarrow$ we have to dinstinguish two cases.
-If there are $\gamma-1$ or fewer error-free nucleotides, we must only make
-sure that there is no $\uparrow$ symbol among them. The weighted
-generating function is
+In the example below, there is a shared MEM seed only if $\gamma \leq
+11$. The main thread is unmasked, so the number of nucleotides between the
+error symbols matters.
+\begin{inset}
+\includegraphics[scale=.9]{ddown_to_down_case_1.pdf}
+\end{inset}
+In this other example, there can be no MEM seed because the first
+duplicate masks the main thread. This depends neither on the value of
+$\gamma$ nor on the number of nucleotides between the error symbols.
+\begin{inset}
+\includegraphics[scale=.9]{ddown_to_down_case_2.pdf}
+\end{inset}
+
+So if the $\Downarrow$ and the $\downarrow$ symbols are separated by
+$\gamma-1$ or fewer nucleotides, there is no MEM seed and we must only
+rule out the presence of other terminators in the segment (in this case
+the $\uparrow$ symbol). The weighted generating function is
 \begin{equation}
@@
 
-There can also be more than $\gamma-1$ error-free nucleotides, but in this
-case, we must make sure that at least one potentially masking thread masks
-the main thread.
+On the other hand, if the $\Downarrow$ and the $\downarrow$ symbols are
+separated by $\gamma$ or more nucleotides, there could be a shared MEM
+seed, so we must enforce the condition that at least one potentially
+masking thread masks the main thread up to and including the $\downarrow$
+symbol (this incidentally rules out the presence of other terminators in
+the segment).
 
@@
 \begin{equation}
-\tilde{A}_\infty(z) =
+A^\downarrow(z) =
 \tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
@@
 $\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
-is $(\sim)_i$ is $q^j V_i$, where
+is $(\sim)_i$ is $q^j W_i$, where
 \begin{eqnarray}
 \label{eq:Si}
-V_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
+W_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
   (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
@@
 \begin{equation}
-v_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
+w_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
   \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
@@
 \begin{eqnarray}
-  W_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
+  V_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
   (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
@@
 \begin{equation}
-w_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
+v_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
 \end{equation}
-
-%In order to compute the weighted generating functions of the other
-%transitions, first observe that
-%\begin{equation}
-%V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
-%\end{equation}
-%So the probability that an error-free stretch of length $i \geq 1$
-%following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
-%$\uparrow$ symbol is
-%\begin{equation}
-%1 - \sum_{j=1}^i V_j+W_j = 1 - \frac{\alpha_i^N -
-%\alpha_0^N}{1-(1-\mu/3)^N}.
-%\end{equation}
-%
-%The probability that the next terminator after a $\downarrow$ symbol is a
-%$\Downarrow$ symbol at position $j$ is the probability of an error-free
-%stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
-%$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
-%by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
-%weighted generating function of the transition from $\downarrow$ to
-%$\Downarrow$ is
-%\begin{equation}
-%B_\infty(z) = \omega z \sum_{i=0}^\infty
-%\Big(1 - \frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} \Big) (qz)^i.
-%\end{equation}
 
@@
 \begin{equation}
-B_\infty(z) = \omega z\sum_{i=0}^\infty
+B^\Downarrow(z) = \omega z\sum_{i=0}^\infty
 \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
@@
 
-By the same rationale, we see that the weighted generating function of the
-transition from $\downarrow$ to $\downarrow$ is
+If the next terminator is $\downarrow$, we again have to distinguish two
+cases. In the example below there is no MEM seed, but the nature of the
+terminator depends on $\gamma$. If $\gamma < 10$, the state $(\sim)_9$
+does not exist, so there is a single segment of size 12, terminated by
+$\downarrow$. If $\gamma \geq 10$, the state $(\sim)_9$ exists and there
+is a segment of size 9 terminated by $\sim$, followed by a segment of size
+3 terminated by $\downarrow$.
+\begin{inset}
+\includegraphics[scale=.9]{down_to_down_case_1.pdf}
+\end{inset}
+
+So if the two $\downarrow$ symbols are separated by $\gamma-1$ or fewer
+nucleotides, there is no MEM seed and we must only rule out the presence
+of other terminators in the segment (in this case $\sim$ and $\uparrow$).
+With the probability computed above that there are neither $\sim$ nor
+$\uparrow$ symbols, the weighted generating function is
 \begin{equation}
-\tilde{B}_\infty(z) = \tilde{\omega} z\sum_{i=0}^\infty
+\tilde{\omega} z\sum_{i=0}^{\gamma-1}
 \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
+
+If the two $\downarrow$ symbols are separated by $\gamma$ or more
+nucleotides, there could be a shared MEM seed, so we must enforce the
+condition that the main thread is masked. In addition, we must also rule
+out the presence of the $\sim$ terminator among the first $\gamma-1$
+nucleotides (but not among the remaining nucleotides). Finally, we must
+rule out the presence of the $\uparrow$ terminator throughout the segment.
+This is satisfied if and only if $i.$ a strictly masking thread survives
+until the penultimate nucleotide of the fragment or $ii.$ all the strictly
+masking threads fail before the penultimate nucleotide but not before the
+$\gamma$-th nucleotide, and a potentially masking thread survives until
+the last nucleotide of the fragment.
+
+If there are $m$ potentially masking threads and $N-m$ strictly masking
+threads, the respective probabilities of conditions $i.$ and $ii.$ are
+$\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N)$ and $(\xi_{i-1,N-m}
+-\xi_{\gamma-1,N-m})(1-\alpha_{i-1}^m)$. Since the conditions are
+are mutually exclusive, the full probability is
+\begin{equation}
+\sum_{m=0}^{N-1}Q_m \bigg(\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N) +
+(\xi_{\gamma-1,N-m}-\xi_{i-1,N-m})(1-\alpha_{i-1}^m) \bigg).
+\end{equation}
+
+The first term is equal to $1-\alpha_{i-1}^N$, the second is equal to
+\begin{equation}
+\frac{\alpha_{i-1}^N - \alpha_{\gamma-1}^N -
+  \zeta_{i-1,i-1}^N + \zeta_{\gamma-1,i-1}^N}{1-(1-\mu/3)^N}.
+\end{equation}
+where
+\begin{equation}
+\zeta_{j,i} = 1-(1-\mu)^j\mu/3-(1-\mu)^i\mu/3(1-\mu/3)
+\end{equation}
+
+Finally, the weighted generating function of the transitions from
+$\downarrow$ to $\downarrow$ is
+
+\begin{equation*}
+\begin{split}
+B^\downarrow(z) = \tilde{\omega} z \sum_{i=0}^{\gamma-1}
+& \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i + \\
+pz\sum_{i=\gamma}^\infty & \left( 1-\alpha_i^N + 
+\frac{\alpha_i^N - \alpha_{\gamma-1}^N -
+  \zeta_{i,i}^N + \zeta_{\gamma-1,i}^N}{1-(1-\mu/3)^N} \right)
+(qz)^i
+\end{split}
+\end{equation*}
 
@@
 
-The sum of the terms above for $m = 1, \ldots, N-1$ is $V_j$ from equation
+The sum of the terms above for $m = 1, \ldots, N-1$ is $W_j$ from equation
 (\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.
@@
 at position $i$ after entering state $(\sim)_j$ is
-$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $V_j$ and summing over $m = 1,
+$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $W_j$ and summing over $m = 1,
 \ldots, N-1$, we find that the weighted generating function of the
@@
 \begin{eqnarray}
-y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
+y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{W_j}
 \big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
@@
 probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
-error -- probability $p$. Dividing by $V_j$ and summing over $m = 1,
+error -- probability $p$. Dividing by $W_j$ and summing over $m = 1,
 \ldots, N-1$, we find that the weighted generating function of the
@@
 \begin{eqnarray}
-C_{\gamma-j}(z) &=& \omega z \sum_{i=1}^{\gamma-j-1} \sum_{m=1}^{N-1}
-  \frac{P_{m,j}}{V_j} \xi_{i-1,m} \cdot (qz)^{i-1} \notag \\
-  &=& \omega z \sum_{i=1}^{\gamma-j-1}
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
-  +\beta_{j-1,i+j-1}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
+D^\Downarrow_j(z) &=& \omega z \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
+  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i \notag \\
+  &=& \omega z \sum_{i=0}^{\gamma-j-2}
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
+  +\beta_{j-1,i+j}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
 \end{eqnarray}
@@
 to $\downarrow$ is seen to be 
-\begin{equation}
-\tilde{C}_{\gamma-j}(z) = \tilde{\omega} z \sum_{i=1}^{\gamma-j-1}
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j-1}^N
-  +\beta_{j-1,i+j-1}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^{i-1}.
-\end{equation}
+\begin{eqnarray}
+D^\downarrow_j(z) = \tilde{\omega} z
+  \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
+  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i +
+  pz \sum_{i=\gamma-j-1}^\infty \sum_{m=1}^{N-1}
+  \frac{P_{m,j}}{W_j} (1-\alpha_i^m) \cdot (qz)^i \notag \\
+\tilde{\omega} z \sum_{i=0}^{\gamma-j-2}
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
+  +\beta_{j-1,i+j}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
+pz \sum_{i=\gamma-j-1}^\infty
+\frac{ \alpha_j^N-\alpha_{j-1}^N-\zeta_{j,i+j}^N
+  +\zeta_{j-1,i+j}^N }
+{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
+\end{eqnarray}
 
@@
 \begin{equation}
-T_{\downarrow}(z) = \sum_{i=0}^\infty
+T^\downarrow(z) = \sum_{i=0}^\infty
 \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
@@
 \begin{equation}
-T_{(\uparrow)_j}(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
+T^\uparrow_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
 \end{equation}
@@
 \begin{equation}
-T_{\Downarrow}(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
+T^\Downarrow(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
 \end{equation}
@@
 \begin{equation}
-T_{(\sim)_j}(z) = \sum_{i=0}^{\gamma-j-1}
+T^\sim_j(z) = \sum_{i=0}^{\gamma-j-1}
 \frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
@@
 \begin{block}{c[c]}
-\DN & T_\Downarrow(z) \\
-\dn & T_\downarrow(z) \\
-\eq{1} & T_{(\sim)_1}(z) \\
+\DN & T^\Downarrow(z) \\
+\dn & T^\downarrow(z) \\
+\up{\gamma-1} & T^\uparrow_{\gamma-1}(z) \\
 \vdots & \vdots \\
-\eq{\gamma-1} & T_{(\sim)_{\gamma-1}}(z) \\
-\up{1} & T_{(\uparrow)_1}(z) \\
+\up{1} & \; T^\uparrow_1(z) \; \\
+\eq{\gamma-1} & T^\sim_{\gamma-1}(z) \\
 \vdots & \vdots \\
-\up{\gamma-1} & \; T_{(\uparrow)_{\gamma-1}}(z) \; \\
+\eq{1} & T^\sim_1(z) \\
 \end{block}
@@
 \begin{blockarray}{ccccccccc}
-   & \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
-    \up{1} & \ldots & \up{\gamma-1} \\
+   & \DN & \dn &\up{\gamma-1} & \ldots & \up{1} &
+    \eq{\gamma-1} & \ldots & \eq{1} \\
 \begin{block}{c[cccccccc]}
-\DN & A_\gamma(z) & \tilde{A}_\infty(z) & 0 & \ldots & 0 & u_1(z)
-    & \ldots & u_{\gamma-1}(z) \\
-\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z) & \ldots &
-    v_{\gamma-1}(z) & w_1(z) & \ldots & w_{\gamma-1}(z) \\
-\eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & y_{1,\gamma-2}(z) \\
-\eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & y_{2,\gamma-3}(z) \\
+\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_{\gamma-1}(z) & \ldots &
+    u_1(z) & 0 & \ldots & 0 \\
+\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_{\gamma-1}(z) & \ldots &
+    v_1(z) & w_{\gamma-1}(z) & \ldots & w_1(z) \\
+\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
+    \ldots & 0 & 0 & \ldots & 0 \\
 \vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
     \ddots & \vdots \\
-\eq{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
+\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
+    \ldots & 0 \\
+\eq{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
+    \ldots & 0 \\
+\eq{\gamma-2} & D_2(z) & \tilde{D}_2(z) & y_{\gamma-2,1}(z) &
     \ldots & 0 & 0 & \ldots & 0 \\
-\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & 0 \\
-\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & 0  \\
 \vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
     \ddots & \vdots \\
-\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
-    \ldots & 0 \\
+\eq{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) &
+    y_{2,\gamma-3}(z) & \ldots & 0 & 0 & \ldots & 0 \\
+\eq{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) &
+    y_{1,\gamma-2}(z) & \ldots & 0 & 0 & \ldots & 0 \\
 \end{block}
@@
     & \ldots & u_{\gamma-1}(z) \\
-\dn & B_\infty(z) & \tilde{B}_\infty(z) & w_1(z)
-    & \ldots & w_{\gamma-1}(z) \\
+\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z)
+    & \ldots & v_{\gamma-1}(z) \\
 \up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &

### 2018-05-22 MEM.tex
Rewrote the theory without state ~ (seems to give identical results).

@@
 \begin{equation}
+\label{eq:u}
 u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
@@
 
-The $\downarrow$ symbol can be followed by all the terminators. We need to
-start with the case of the $\sim$ terminator, because it will be required
-for the other ones. Referring to the decision tree of
-Fig.~\ref{fig:decision_tree}, we see that the $\sim$ symbol occurs in an
-error-free stretch, if all the strictly masking threads fail before the
-potentially masking threads. In section~\ref{sec:trans_from_sim} it will
-be important to distinguish different cases based on the value of
-$\varphi_0$ at the position of the $\sim$ symbol, so as in
-section~\ref{sec:trans_from_up}, we introduce the states $(\sim)_j$ $(1
-\leq j \leq \gamma-1)$ such that $\varphi_0(n) = j$ if the nucleotide at
-position $n$ is $(\sim)_j$. In this case, the value of $\varphi_0(n) = j$
-is simply the position of the $\sim$ symbol relative to the preceding
-$\downarrow$ symbol.
+The probability that the next terminator is the $\uparrow$ symbol at
+position $i$ is the probability that the last of the $N$ duplicate threads
+fails at position $i$ is $\xi_{i-1,N-m}-\xi_{i,N-m}$ so the weighted
+generating function of the transition from $\downarrow$ to $(\uparrow)_i$
+is
+\begin{equation}
+\tag{\ref{eq:u}}
+u_i(z) = (\xi_{i-1,N}-\xi_{i,N}) \cdot (qz)^i.
+\end{equation}
 
-If there are $m = 0$ potentially masking threads, the next terminator
-cannot be $\sim$. Assuming that there are $m > 0$ potentially masking
-threads at the $\downarrow$ symbol, the probability that the have not all
-faied at position $i$ is $\xi_{i,m}$. Similarly, the probability that the
-last strictly masking thread fails at position $i$ is
-$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
-is $(\sim)_i$ is $q^j W_i$, where
-\begin{eqnarray}
-\label{eq:Si}
-W_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
-  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
-  &=& \frac{\alpha_i^N-\alpha_{i-1}^N-\gamma_i^N+\delta_{i-1}^N}
-{1-(1-\mu/3)^N} \text{, where}
-\end{eqnarray}
+To compute the weighted generating function of the transition from
+$\downarrow$ to $\Downarrow$, we need to separate two cases. In the first,
+the error symbols are separated by $\gamma-1$ or fewer error-free
+nucleotides. In this case, there can be no MEM seed and we only have to
+exclude the $\uparrow$ terminator. We obtain
+\begin{equation}
+\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i.
+\end{equation}
 
-\begin{eqnarray}
-\gamma_i &=& 1-(1-\mu)^i \\
-\delta_i &=& 1-(1-\mu+\mu^2/3)(1-\mu)^i.
-\end{eqnarray}
-
-\begin{remark}
-The term $\alpha_i^N-\alpha_i^N$ is the probability that the last dupicate
-thread fails in one error followed by $i$ error-free nucleotides. The term
-$\gamma_i^N - \delta_{i-1}^N$ is the probability that all threads fail in
-$i$ error-free nucleotides...
-\end{remark}
-
-We have assumed $m>0$, but the sum above can start at $m=0$ because
-$\xi_{i,0}=0$. The weighted generating function of the transition from
-$\downarrow$ to $(\sim)_i$ is
+In the case that $\downarrow$ and $\Downarrow$ are separated by $\gamma$
+or more error-free nucleotides, we must ensure that at least one strictly
+masking thread survives until the penultimate nucleotide of the segment.
+This probability is equal to
 \begin{equation}
-w_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
-  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
+\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} =
+\frac{1-\alpha_i^N}{1-(1-\mu/3)^N}.
 \end{equation}
@@
 \begin{remark}
-When $N = 1$, $V_i = 0$ and the states $(\sim)_i$ are inaccessible.
-The symbol $\sim$ can exist only if there is more than one duplicate.
+A strictly masking thread survives until the $\Downarrow$ symbol if and
+only if one of the $N$ duplicate threads reverse-survives from
+the penultimate nucleotide of the segment down to and including the
+$\downarrow$ symbol.
 \end{remark}
 
-Referring again to the decision of Fig.~\ref{fig:decision_tree}, we see
-that the $\uparrow$ symbol occurs in an error-free stretch, if all the
-potentially masking threads fail before or at the same nucleotide as the
-strictly masking threads. Assuming that there are $m \geq 0$ potentially
-masking threads at the $\downarrow$ symbol, the probability that they have
-all failed at position $i$ is $1-\xi_{i,m}$. As we have seen above, the
-probability that the last strictly masking thread fails at position $j$ is
-$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
-is the $\uparrow$ symbol at position $i$ is $q^i W_i$, where
-\begin{eqnarray}
-  V_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
-  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
-    &=& \frac{\gamma_i^N-\delta_{i-1}^N}{1-(1-\mu/3)^N}.
-\end{eqnarray}
-
-The weighted generating function of the transition from $\downarrow$ to
-$(\uparrow)_i$ is
+In sum, the weighted generating function of the transition from
+$\downarrow$ to $\Downarrow$ is
 \begin{equation}
-v_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
-\end{equation}
-
-The probability that there is neither $\sim$ nor $\uparrow$ symbol is just
-the probability that at least one stricty masking thread is still alive
-\begin{equation}
-\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} = \frac{1-\alpha_i^N}{1-(1-\mu/3)^N}
-\end{equation}
-
-So the weighted generating function of the transition from $\downarrow$ to
-$\Downarrow$ is
-\begin{equation}
-B^\Downarrow(z) = \omega z\sum_{i=0}^\infty
-\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
+B^\Downarrow(z) = 
+\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
+\omega z\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
@@
 If the next terminator is $\downarrow$, we again have to distinguish two
-cases. In the example below there is no MEM seed, but the nature of the
-terminator depends on $\gamma$. If $\gamma < 10$, the state $(\sim)_9$
-does not exist, so there is a single segment of size 12, terminated by
-$\downarrow$. If $\gamma \geq 10$, the state $(\sim)_9$ exists and there
-is a segment of size 9 terminated by $\sim$, followed by a segment of size
-3 terminated by $\downarrow$.
-\begin{inset}
-\includegraphics[scale=.9]{down_to_down_case_1.pdf}
-\end{inset}
-
-So if the two $\downarrow$ symbols are separated by $\gamma-1$ or fewer
-nucleotides, there is no MEM seed and we must only rule out the presence
-of other terminators in the segment (in this case $\sim$ and $\uparrow$).
-With the probability computed above that there are neither $\sim$ nor
-$\uparrow$ symbols, the weighted generating function is
+cases. If the two $\downarrow$ symbols are separated by $\gamma-1$ or
+fewer nucleotides, there is no MEM seed and we must only rule out the
+presence of the $\uparrow$ terminator. We obtain
 \begin{equation}
 \tilde{\omega} z\sum_{i=0}^{\gamma-1}
-\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
+\xi_{i,N} \cdot (qz)^i.
 \end{equation}
@@
 nucleotides, there could be a shared MEM seed, so we must enforce the
-condition that the main thread is masked. In addition, we must also rule
-out the presence of the $\sim$ terminator among the first $\gamma-1$
-nucleotides (but not among the remaining nucleotides). Finally, we must
-rule out the presence of the $\uparrow$ terminator throughout the segment.
-This is satisfied if and only if $i.$ a strictly masking thread survives
-until the penultimate nucleotide of the fragment or $ii.$ all the strictly
-masking threads fail before the penultimate nucleotide but not before the
-$\gamma$-th nucleotide, and a potentially masking thread survives until
-the last nucleotide of the fragment.
+condition that the main thread is masked. This is satisfied if and only if
+one of the following conditions holds:
+\begin{enumerate}
+\item A strictly masking thread survives until the penultimate
+nucleotide of the segment.
+\item All the strictly masking threads fail before the penultimate
+nucleotide but a potentially masking thread survives until the last
+nucleotide of the segment.
+\end{enumerate}
 
-If there are $m$ potentially masking threads and $N-m$ strictly masking
-threads, the respective probabilities of conditions $i.$ and $ii.$ are
-$\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N)$ and $(\xi_{i-1,N-m}
--\xi_{\gamma-1,N-m})(1-\alpha_{i-1}^m)$. Since the conditions are
-are mutually exclusive, the full probability is
+Assuming that there are $m \geq 0$ potentially masking threads, the
+probability of condition $1.$ is $\xi_{i-1,N-m}$ and that of condition 2.
+is $(1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m)$. Summing of $m = 0, \ldots, N-1$
+on obtains
 \begin{equation}
-\sum_{m=0}^{N-1}Q_m \bigg(\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N) +
-(\xi_{\gamma-1,N-m}-\xi_{i-1,N-m})(1-\alpha_{i-1}^m) \bigg).
+(1-\mu/3)\sum_{m=0}^{N-1}Q_m \cdot \xi_{i-1,N-m} =
+1-\alpha_{i-1}^N \text{, and}
 \end{equation}
-
-The first term is equal to $1-\alpha_{i-1}^N$, the second is equal to
 \begin{equation}
-\frac{\alpha_{i-1}^N - \alpha_{\gamma-1}^N -
-  \zeta_{i-1,i-1}^N + \zeta_{\gamma-1,i-1}^N}{1-(1-\mu/3)^N}.
+\sum_{m=0}^{N-1}Q_m (1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m) =
+\frac{\alpha_{i-1}^N - \alpha_0^N -
+  \zeta_{i-1,i-1}^N + \zeta_{0,i-1}^N}{1-(1-\mu/3)^N},
 \end{equation}
@@
 $\downarrow$ to $\downarrow$ is
-
 \begin{equation*}
-\begin{split}
-B^\downarrow(z) = \tilde{\omega} z \sum_{i=0}^{\gamma-1}
-& \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i + \\
-pz\sum_{i=\gamma}^\infty & \left( 1-\alpha_i^N + 
-\frac{\alpha_i^N - \alpha_{\gamma-1}^N -
-  \zeta_{i,i}^N + \zeta_{\gamma-1,i}^N}{1-(1-\mu/3)^N} \right)
-(qz)^i
-\end{split}
+B^\downarrow(z) = \tilde{\omega}z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot
+(qz)^i + pz \sum_{i=\gamma}^\infty
+\left( 1-\alpha_i^N +
+\frac{\alpha_i^N - \alpha_0^N - \zeta_{i,i}^N +
+\zeta_{0,i}^N}{1-(1-\mu/3)^N} \right) (qz)^i
 \end{equation*}
 
-\subsection{Transitions from $\sim$}
-\label{sec:trans_from_sim}
-
-At a $\sim$ symbol, the main thread becomes co-dominant, together with $1$
-to $N-1$ alternative threads (if there were $N$, the symbol would be
-$\Downarrow$). As in section~\ref{sec:trans_from_Down}, the next
-terminator symbol depends on whether all the alternative co-dominant
-threads fail before the main thread. If this is the case, the main thread
-will become dominant ($\uparrow$ symbol), otherwise we again need to
-distinguish the case that the read is reset ($\Downarrow$ symbol) from the
-case that the main thread becomes dominated ($\downarrow$ symbol).
-
-An important difficulty is that the $\sim$ symbol does not inform us about
-the number of surviving threads (except when $N = 1$). So we need to
-marginalize the probabilities of interest over this number. The first
-thing to note is that a $\sim$ terminator can only follow a $\downarrow$
-terminator. Indeed, after a $\Downarrow$ symbol the threads are already
-co-dominant, and after a $\uparrow$ symbol the alternative threads cannot
-`catch up' with the main thread until a sequencing error occurs,
-\textit{i.e.} until a $\Downarrow$ symbol or a $\downarrow$ symbol is
-inserted. The number of surviving threads depends on the position of the
-$\sim$ symbol relative to the preceding $\downarrow$ symbol, so we need
-to treat the states $(\sim)_j$ separately.
-
-We first establish an equality that will turn out to be useful in the
-subsequent calculations.
-
-\begin{lemma}
-For any real numbers $x$ and $y$,
-\begin{eqnarray*}
-\sum_{n=m}^N {N \choose n} {n \choose m} x^n y^{N-n}
-&=& x^m \sum_{r=0}^{N-m} {N \choose m+r} {m+r \choose m}a^r y^{N-m-r} \\
-&=& {N \choose m} x^m \sum_{r=0}^{N-m} {N-m \choose r}x^r y^{N-m-r} \\
-&=& {N \choose m} x^m (x+y)^{N-m}.
-\end{eqnarray*}
-
-The equality above implies
-\begin{equation}
-\label{eq:double_binom}
-\sum_{n=m}^{N-1} {N \choose n} {n \choose m} x^n y^{N-n} =
-{N \choose m} x^m \Big[ (x+y)^{N-m} - x^{N-m} \Big].
-\end{equation}
-\end{lemma}
-
-The probability that there are $m>1$ potentially masking threads upon
-entering state $(\sim)_j$ is the probability that there were $n \geq m$
-potentially masking threads at the $\downarrow$ symbol -- probability
-$Q_n$ --, that exactly $m$ have survived until position $j$ -- probability
-${n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}$ -- and that the last of
-the $N-n$ strictly masking threads fails at position $j$ -- probability
-$\xi_{j-1,N-n}-\xi_{j,N-n}$. Summing over $n = m, \ldots, N-1$, the
-probability is seen to be
-\begin{equation}
-\label{eq:miscprob}
-P_{m,j} = \sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm}
-(1-(1-\mu)^j)^{n-m} \cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
-\end{equation}
-
-We write $\xi_{j-1,N-n}-\xi_{j,N-n}$ as $(1-\xi_{j,N-n}) -
-(1-\xi_{j-1,N-n})$ and split the sum of equation (\ref{eq:miscprob}) into
-two parts. The first term is equal to
-\begin{equation*}
-\begin{split}
-\sum_{n=m}^{N-1} & Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
-\cdot (1 - \xi_{j,N-n}) = \\
-& \frac{(1-\mu)^{jm} (1-(1-\mu)^j)^{N-m}}{1-(1-\mu/3)^N}
-\sum_{n=m}^{N-1} {N \choose n} {n \choose m} (1-\mu/3)^n (\mu/3)^{N-n}.
-\end{split}
-\end{equation*}
-
-Using formula (\ref{eq:double_binom}) with $x=1-\mu/3$ and $y=\mu/3$, we
-see that this term is equal to
-\begin{equation*}
-\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
-\bigg( \gamma_j^{N-m}
--\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
-\end{equation*}
-
-The second term of equation (\ref{eq:miscprob}) is equal to
-\begin{multline*}
-\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
-\cdot (1 - \xi_{j-1,N-n}) = \\
-\frac{1}{1-(1-\mu/3)^N}
-\left(\frac{(1-\mu)^j}{1-(1-\mu)^j}\right)^m
-\; \sum_{n=m}^{N-1} {N \choose n} {n \choose m} \times \\
-(1-\mu/3)^n (\mu/3)^{N-n}
-(1-(1-\mu)^j)^n (1-(1-\mu)^{j-1})^{N-n}.
-\end{multline*}
-
-Formula (\ref{eq:double_binom}) with $x=(1-(1-\mu)^j)(1-\mu/3)$ and
-$y = (1-(1-\mu)^{j-1})\mu/3$ yields
-\begin{equation*}
-\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m}
-\bigg( \delta_{j-1}^{N-m}
--\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
-\end{equation*}
-
-Computing the difference between the terms, we see that the probability
-that there are $m$ potentially masking threads upon entering state
-$(\sim)_j$ is
-\begin{equation*}
-P_{m,j} = \frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
-\big( \gamma_j^{N-m} - \delta_{j-1}^{N-m} \big).
-\end{equation*}
-
-The sum of the terms above for $m = 1, \ldots, N-1$ is $W_j$ from equation
-(\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.
-
-Now, referring to the decision tree of Fig.~\ref{fig:decision_tree}, we
-see that if the $m$ potentially masking threads fail before the main
-thread, the next terminator will be the $\uparrow$ symbol, otherwise it
-will be either the $\Downarrow$ symbol or the $\downarrow$ symbol.
-
-The probability that the last of the $m$ potentially masking threads fails
-at position $i$ after entering state $(\sim)_j$ is
-$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $W_j$ and summing over $m = 1,
-\ldots, N-1$, we find that the weighted generating function of the
-transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
-\begin{eqnarray}
-y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{W_j}
-\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
-&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
--\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
-  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
-\end{eqnarray}
-where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
-\begin{equation}
-\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
-\end{equation}
-
-The probability that the next terminator is a $\Downarrow$ symbol or a
-$\downarrow$ symbol at position $i$ is the probability that at least one
-of the potentially masking threads survives until position $i-1$ --
-probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
-error -- probability $p$. Dividing by $W_j$ and summing over $m = 1,
-\ldots, N-1$, we find that the weighted generating function of the
-transtion from $(\sim)_j$ to
-$\Downarrow$ is
-\begin{eqnarray}
-D^\Downarrow_j(z) &=& \omega z \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
-  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i \notag \\
-  &=& \omega z \sum_{i=0}^{\gamma-j-2}
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
-  +\beta_{j-1,i+j}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
-\end{eqnarray}
-
-With the same notations, the weighted generating function from $(\sim)_j$
-to $\downarrow$ is seen to be 
-\begin{eqnarray}
-D^\downarrow_j(z) = \tilde{\omega} z
-  \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
-  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i +
-  pz \sum_{i=\gamma-j-1}^\infty \sum_{m=1}^{N-1}
-  \frac{P_{m,j}}{W_j} (1-\alpha_i^m) \cdot (qz)^i \notag \\
-\tilde{\omega} z \sum_{i=0}^{\gamma-j-2}
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
-  +\beta_{j-1,i+j}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
-pz \sum_{i=\gamma-j-1}^\infty
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\zeta_{j,i+j}^N
-  +\zeta_{j-1,i+j}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
-\end{eqnarray}
+\begin{remark}
+One can interpret $\zeta_{i,i}$ as the probability that
+\end{remark}
 
@@
 H(z) = 
-\begin{blockarray}{cccccccc}
-   \;\; \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
-    \up{1} & \ldots & \up{\gamma-1} \\
-\begin{block}{[cccccccc]}
-\;\; 1 & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 \\
+\begin{blockarray}{ccccc}
+   \;\; \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
+\begin{block}{[ccccc]}
+\;\; 1 & 0 & 0 & \ldots & 0 \\
 \end{block}
@@
 \begin{equation}
-T^\downarrow(z) = \sum_{i=0}^\infty
-\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
+T^\downarrow(z) = 
+\sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
+\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
 \end{equation}
@@
 
-For the $(\sim)_j$ state, the weighted generating function is
-\begin{equation}
-T^\sim_j(z) = \sum_{i=0}^{\gamma-j-1}
-\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
-{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
-\end{equation}
-
 In summary,
@@
 \up{1} & \; T^\uparrow_1(z) \; \\
-\eq{\gamma-1} & T^\sim_{\gamma-1}(z) \\
-\vdots & \vdots \\
-\eq{1} & T^\sim_1(z) \\
 \end{block}
@@
 \begin{equation*}
-\begin{blockarray}{ccccccccc}
-   & \DN & \dn &\up{\gamma-1} & \ldots & \up{1} &
-    \eq{\gamma-1} & \ldots & \eq{1} \\
-\begin{block}{c[cccccccc]}
-\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_{\gamma-1}(z) & \ldots &
-    u_1(z) & 0 & \ldots & 0 \\
-\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_{\gamma-1}(z) & \ldots &
-    v_1(z) & w_{\gamma-1}(z) & \ldots & w_1(z) \\
-\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
-    \ldots & 0 & 0 & \ldots & 0 \\
-\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
-    \ddots & \vdots \\
-\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
-    \ldots & 0 \\
-\eq{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
-    \ldots & 0 \\
-\eq{\gamma-2} & D_2(z) & \tilde{D}_2(z) & y_{\gamma-2,1}(z) &
-    \ldots & 0 & 0 & \ldots & 0 \\
-\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
-    \ddots & \vdots \\
-\eq{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) &
-    y_{2,\gamma-3}(z) & \ldots & 0 & 0 & \ldots & 0 \\
-\eq{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) &
-    y_{1,\gamma-2}(z) & \ldots & 0 & 0 & \ldots & 0 \\
-\end{block}
-\end{blockarray}
-\end{equation*}
-
-In the case $N = 1$, reads never contain the symbol $\sim$ and the matrix
-is
-\begin{equation*}
 \begin{blockarray}{cccccc}
@@
 \begin{block}{c[ccccc]}
-\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_1(z)
+\DN & A^\Downarrow(z) & A^\downarrow(z) & u_1(z)
     & \ldots & u_{\gamma-1}(z) \\
-\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z)
-    & \ldots & v_{\gamma-1}(z) \\
-\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &
-    \ldots & 0 \\
-\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 &
+\dn & B^\Downarrow(z) & B^\downarrow(z) & u_1(z)
+    & \ldots & u_{\gamma-1}(z) \\
+\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 &
     \ldots & 0 \\
 \vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
-\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 \\
+\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & \ldots & 0 \\
 \end{block}
@@
 \DN    & pz              & qz \\
-\up{1} & D_{\gamma-1}(z) & 0  \\
+\up{1} & C_{\gamma-1}(z) & 0  \\
 \end{block}

### 2018-07-19 MEM.tex
Resketched the whole thing with newer theory and Monte Carlo.

@@
 
+\usepackage{algorithm2e}
 \usepackage{amsmath}
@@
 
+\setlength{\thickmuskip}{2mu plus 3mu minus 1mu}
+\setlength{\medmuskip}{1mu plus 2mu minus 1mu}
+
+\SetKwComment{Comment}{$\triangleright$\ }{}
+
 % ---- Author affiliations ---- %
@@
 
-\newtheorem{algorithm}{Algorithm}
+%\newtheorem{algorithm}{Algorithm}
 \newtheorem{corollary}{Corollary}
@@
 
-\newcommand{\DN}{\scriptstyle{\Downarrow}}
-\newcommand{\dn}{\scriptstyle{\downarrow}}
-\newcommand{\up}[1]{\scriptstyle{(\uparrow)_{#1}}}
-\newcommand{\eq}[1]{\scriptstyle{(\sim)_{#1}}}
-
-\newcommand{\upg}{\scriptstyle{\uparrow}}
-\newcommand{\eqg}{\scriptstyle{\sim}}
+\newcommand{\dn}[1]{\scriptstyle{\downarrow_{/#1}}}
+\newcommand{\up}[1]{\scriptstyle{\uparrow_{/#1}}}
+\newcommand{\nd}{\scriptstyle{|}}
 
@@
 \author[1,2]{Guillaume J. Filion}
+\author[1,2]{Eduard Zorita}
 \affil[1]{Genome Architecture, Gene Regulation, Stem Cells and Cancer
@@
 The abstract will come later.
+A popular mapping heuristic is called \emph{seeding}, and a recent variant
+is the \emph{MEM seeding} approach. Here we tackle the question of how to
+evaluate the probability that MEM seeding hits are on-target. This in turn
+expresses the confidence one should have in the location of the mapped DNA
+fragment and can be used to calibrate the MEM seeding heuristic, or to
+filter the hits at the desired level of confidence.
 \end{abstract}
@@
 \section{Introduction}
-Some introduction
 
-\section{MEM seeds}
+\subsection{Mapping}
 
-\subsection{Seeds}
-DEFINE MEM SEED AND SHARED MEM SEED.
+Consider the following task referred to as the \emph{true mapping
+problem}: a random DNA fragment from a known genome is sequenced with an
+imperfect instrument; can we identify the location of the fragment in the
+genome?
 
-EXPLAIN THAT THE TAIL IS AN UNTERMINATED SEGMENT.
+It seems that the answer depends only on the size of the fragment and the
+error rate of the instrument, but there is another important factor to
+consider. Most genomes contain repeated sequences. Intuitively, the
+problem is more difficult if the random DNA fragment is duplicated. If one
+of the duplicates is exactly identical to the original, then the DNA
+fragment cannot be mapped with certainty. Even if the duplicates are all
+different from the original (which happens because of mutations in the
+genome), the read may still be closer to one of the duplicates because of
+sequencing errors. So we can never be sure that the read is mapped to the
+\emph{true} location.
 
-\subsection{The seeding alphabet}
+For this reason, we introduce the \emph{best mapping problem}, where we
+search not the true, but the \emph{best} location of the DNA fragment,
+\textit{i.e.} the sequence of the genome that is the most similar to the
+read.
 
-In order to compute seeding probabilities, we need to consider that reads
-are sequences of letters from the \emph{seeding alphabet} rather than from
-the nucleotide alphabet. them as a sequence of symbols telling if the
-nucleotides match the target and~/ or the duplicates. Here, we will
-consider that reads are sequences of the five symbols $\Downarrow$,
-$\downarrow$, $\uparrow$, $\sim$ and $\square$.
+There are exact solutions to the best mapping problem, but the algorithms
+are too slow to process the large amount of data generated by modern
+sequencers. Instead, one uses heuristic methods, \textit{i.e.} algorithms
+that run fast, but are not guaranteed to return the correct answer. In the
+context of the best mapping problem, there are thus three possible
+outcomes: $i.$ the DNA fragment is mapped to the best location, which is
+called a \emph{true positive} (or an \emph{on-target hit}), $ii.$ the DNA
+fragment is mapped to another location, which is called a \emph{false
+positive} (or an \emph{off-target hit}), $iii.$ the DNA fragment is not
+mapped at all, which is called a \emph{false negative}. A \emph{true
+negative} would designate the outcome where a DNA fragment foreign to the
+genome is not mapped at all, but here we assume that all DNA fragments
+belong to the genome of interest.
 
-The $\Downarrow$ symbol indicates that the nucleotide is not only a
-mismatch for the target, but also for every duplicate. The $\downarrow$
-symbol indicates that the nucleotide is a mismatch for the target, but is
-a match for at least one dupulicate. These two symbols are referred to as
-the \emph{error symbols}, because they imply the occurrence of a
-sequencing error (this is the only way the read may not match the target).
+In a seeding heuristic, one uses exact or near exact matches between the
+read and the genome to collect a set of candidate locations. The read is
+only compared to the sequences at the candidate locations, but not to the
+rest of the genome. The advantage of this method is that there are few
+candidates and that they can be discovered fast. However, the best hit may
+not be in the candidate set, in which case it cannot be discovered
+(\textit{i.e.}, the outcome is either a false positive or a false
+negative).
 
-The remaing three symbols have a more complex usage because they describe
-events of the MEM seeding process that are best explained by way of a
-concrete example. Fig.~\ref{fig:sketch_MEM} illustrates the case of a read
-of size 25 from a sequence with three duplicates. The four files of
-squares at the bottom are the target and duplicate sequences, where a
-black square means that the decoded nucleotide is a mismatch and a grey
-square means that it is a match.
+A \emph{seed} designates a subsequence of fixed length that is shared
+between the read and the genome. In other words, a seed is a perfect match
+between a part of the read an a genomic site. The candidates consist of
+the genomic sites that contain one or more seeds (this depends on the
+specific heuristic).
+
+In the theory developed below, we will assume that the \emph{true}
+location of the read is also the \emph{best}. This is not always the case,
+but it is unfeasible to model the behavior of the best hit in the present
+context. We will later discuss how this relates to the probability that
+the best hit has a seed.
+
+\subsection{Maximal Exact Matches (MEMs)}
+
+Here we consider a seeding variant where seeds do not have a fixed
+length. Let us start by introducing the definition of a Maximal Exact
+Match.
+
+\begin{definition}
+A Maximal Exact Match (MEM) is a subsequence of a read that is present in
+the genome and that cannot be extended left or right --- either because
+the read ends or because the extended subsequence is not in the genome.
+\end{definition}
+
+This definition presents a computational challenge: to know if a given
+subsequence is a MEM, we need to know if it exists somewhere in the
+genome. This is a non trivial problem in itself, but fortunately there
+exists practical methods to answer the question, even for very large
+genomes. Here on, we will assume that we always know which subsequences of
+the read are MEMs.
+
+A practical consideration is that short matches between the read and the
+genome are irrelevant. For instance, $99.7\%$ of all the possible 12-mers
+are present in the reference human genome, so an exact match of size 12 or
+lower says nothing about the location of the DNA fragment. In what
+follows, we will simply disregard short matches. More specifically, we
+will assume that subsequences of the read can only match the original DNA
+sequence or its duplicates, but not the rest of the genome. This will
+simplify the discussion without loss of generality, because short
+uninformative matches are ignored in all practical applications.
+
+Fig.~\ref{fig:MEM_example} shows a concrete example of the relationship
+between read errors, on-target matches, off-target matches and MEMs. The
+sequence of the original DNA fragment is the first genomic hit (this is
+the \emph{true} location). The second, third and fourth genomic hits are
+duplicates. The mismatches between the genome and the read are
+highlighted in bold. The read has three errors indicated by stars, so
+\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
+\texttt{A} and \texttt{G} (because the first hit is the \emph{true}
+sequence).
 
@@
 \centering
-\includegraphics[scale=.9]{sketch_MEM.pdf}
+\includegraphics[scale=1]{MEM_example.pdf}
+\caption{\textbf{Errors, matches and MEMs.}
+Some legend.}
+\label{fig:MEM_example}
+\end{figure}
+
+The MEMs are indicated by grey boxes. Technically, MEMs are subsequences
+of the read, but they are shown on the genomic hits to highlight where
+they match. Two MEMs can overlap, in which case they must match distinct
+sequences of the genome (this follows from the definition). The last MEM
+corresponds to two matches of length 4, with the first and the second
+genomic hits. Such MEMs with several matches are referred to as
+\emph{shared}. MEMs with a single match in the genome are referred to as
+\emph{strict}.
+
+As the name suggests, the principle of MEM seeding is to use MEM as seeds
+to identify candidate genomic locations. Because short matches are
+irrelevant, MEM seeds are usually requested to have a length greater than
+or above a certain threshold denoted $\gamma$. We will assume that a
+single MEM seed is sufficient to consider that a genomic hit is a
+candidate location, which is why we will focus on the case that the read
+does not have a single on-target MEM seed.
+
+%In the example of Fig.~\ref{fig:MEM_example}, there is no MEM that matches
+%the third genomic hit. It is important to understand that this can also
+%happen to the best hit, as in the example shown in
+%Fig~\ref{fig:MEM_eclipse}. The undesirable consequence is that some reads
+%are bound to yield false positives, even if we tune the MEM seeding
+%heuristic by changing the minimum seed length $\gamma$. We will refer to
+%the case that the best hit is masked by off-target MEMs as \emph{MEM
+%eclipse}. 
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=1]{MEM_eclipse.pdf}
+%\caption{\textbf{MEM eclipse.}
+%Some legend.}
+%\label{fig:MEM_eclipse}
+%\end{figure}
+%
+%If you are not already familiar with MEM seeding, we recommend that you
+%take a moment to study the examples of Fig.~\ref{fig:MEM_example} and
+%\ref{fig:MEM_eclipse}. This will be essential to understand the next
+%sections. Before reading further, make sure that you can answer the
+%following questions:
+%\begin{enumerate}
+%\itemsep0em
+%\item What are the conditions for a MEM to be shared?
+%\item What is the minimum number of duplicates for MEM eclipse?
+%\end{enumerate}
+
+\section{Combinatorial construction of reads}
+
+In this section we give a representation of reads as combinatorial ojects.
+The construction has two essential purposes: the first is to give a
+generative model of reads without any on-target MEM seed, the second is to
+allow us to find their weighted generating function, which in turn will
+allow us to find their probability of occurrence.
+
+\subsection{The MEM alphabet}
+
+The nucleotide sequence of the read is not so useful to know where MEMs
+are and whether they are on-target. To develop a formal method to compute
+the probability that the read contains an on-target MEM seed, we will
+recode reads as sequences of letters from the so-called \emph{MEM
+alphabet} $\{\downarrow, \uparrow, \square\}$.
+
+Each nucleotide of the read is recoded with one of three letters. The
+$\downarrow$ symbol indicates that the nucleotide is a sequencing error.
+By definition, it is a mismatch for the target. The $\uparrow$ symbol
+characterizes the presence of an on-target MEM in a way that we are about
+to clarify. All the other nucleotides are represented by the $\square$
+symbol.
+
+To understand the meaning of the $\uparrow$ symbol, consider the following
+situation: at the left end of the read, as we try to extend the first
+match as much as possible to the right, the number of hits in the genome
+keeps decreasing. It may be the case that at some point, we are left with
+a single hit that is on-target. The very first nucleotid where that
+happens is called the \emph{pivot} and is labelled with the symbol
+$\uparrow$.
+
+Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
+encoded with the MEM alphabet. Below the read are the matches between the
+read and the genomic hit, as they are in Fig.~\ref{fig:MEM_example}. A
+black square means that the nucleotide is a mismatch for the genomic
+sequence, and a white or grey square means that it is a match. Grey
+squares are used to highlight MEMs, as in Fig.~\ref{fig:MEM_example}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.85]{sketch_MEM.pdf}
 \caption{\textbf{Reads and MEM seeds}. 
@@
 
-Now consider streaks of matches until a particular position of the read.
-At the 4-th nucleotide, the target scores 4 matches in a row versus 2, 0
-and 1 for the duplicates. The target just acquired the strictly longest
-streak, which is indicated by the $\uparrow$ symbol. This symbol also
-appears at position 17, where the target sequence acquires the strictly
-longest streak again (6 matches in a row, versus 4, 0 and 4 for the
-dupicates).
+Observe that the two on-target MEMs contain a pivot, but the shared MEM
+does not. This is always the case. Recall that the pivot indicates that
+the true location is the \emph{only} local match, so the $\uparrow$ symbol
+cannot appear in on-target MEM seeds when they are shared.
 
-At position 23, the target and the first duplicate both acquire the
-longest streak (2 matches in a row), which is indicated by the $\sim$
-symbol. This symbol is used only at positions where the target
-\emph{acquires} the longest streak (together with at least one duplicate).
-For instance, it is not used at position 1 nor at position 12 because the
-target already had the longest streak (0 in both cases).
+Strict on-target MEM seeds are easy to find when reads are represented in
+the MEM alphabet: they are the longest stretches of symbols containing the
+$\uparrow$ symbol and not containing the $\downarrow$ symbol. Shared
+on-target MEM seeds are trickier. They do not directly appear in this
+representation so we will have to keep that in mind throughout the theory.
+It is in a sense justified to put more emphasis on strict MEM seeds
+because they guarantee that the true location will be found. Shared MEM
+seeds can have thousands of hits in the genome in practice, so the user
+may not test them all. That said, the theory developed here fully takes
+shared MEM seeds into account, even if they are not visible in the
+representation of the MEM alphabet.
 
-Finally, the $\square$ symbol is used for all the other cases,
-\textit{i.e.}, when there is no sequencing error, and when the target does
-not acquire the longest streak of matches (shared or not with a
-duplicate).
-
-Each sequence corresponds to a Bernoulli process where `success' stands
-for a match between the sequence and the read, and `failure' stands for a
-mismatch. For simplicity, we will refer to the Bernoulli processes as
-`threads'.
+The point of the MEM seeding alphabet is that reads can be seen as
+sequences of letters, but also as sequences of \emph{segments} delimited
+by the $\uparrow$ and the $\downarrow$ symbols. Because such segments are
+intimately linked to the presence of on-target MEM seeds, we can give a
+construction of reads that do not contain any.
 
 \begin{definition}
-At a given position of the read, a duplicate thread is said to be
-\emph{strictly masking} if its success streak is strictly longer than the
-match streak of the target. A duplicate thread is said to be
-\emph{potentially masking} if its match streak is the same as the target.
+\label{def:segment}
+A segment is a sequence of $\square$ symbols terminated by either a
+$\uparrow$ symbol or by a $\downarrow$ symbol. The unterminated sequence
+of $\square$ symbols at the end of the read is called the tail --- which
+can thus be an empty sequence of symbols.
 \end{definition}
 
-Duplicate threads are thus classified into strictly masking, potentially
-masking and not masking, which allows us to describe the usage of the five
-symbols.
+Following definition~\ref{def:segment}, the $\uparrow$ and $\downarrow$
+symbols will be referrred to as \emph{terminators}.
+Fig.~\ref{fig:sketch_segment} shows the decomposition of the read from
+Fig.~\ref{fig:sketch_MEM} in segments. The size of each segment (the
+number of nucleotides that compose it) is indicated below.
 
@@
 \centering
-\includegraphics[scale=.7]{decision_tree.pdf}
-\caption{\textbf{Decision tree}. 
+\includegraphics[scale=.85]{sketch_segments.pdf}
+\caption{\textbf{Reads as sequences of segments}. 
 Some legend.}
-\label{fig:decision_tree}
+\label{fig:sketch_segment}
 \end{figure}
 
-We assume that the target sequence has $N$ duplicates that are potential
-false positives for the seeding process. We further assume that
-duplication was instantaenous and that all $N+1$ sequences diverge
-independently from each other at a constant rate. In other words, we
-ignore the complications due to the genealogy of the duplication events
-and we simply assume that at each position, any given duplicate has the
-same nucleotide as the target with probability $1-\mu$. If it does not,
-we assume that the duplicate has any of the remaining three
-nucleotides with equal probability.
+\subsection{The extended MEM alphabet}
 
-To describe those processes, we will
-often use the function $\varphi(n)$ that gives the length of the success
-streak at position $n$. The processes are indexed such that $\varphi_0$
-refers to the target sequence, and functions $\varphi_j$ $(1 \leq j \leq
-N)$ refer to the duplicates. For a concrete example, $\varphi_0(5) = 5$
-and $\varphi_1(5) = 3$ in Fig.~\ref{fig:sketch_MEM}.
+An issue with the MEM alphabet is that the probabilities of the symbols
+are difficult to evaluate. The $\downarrow$ symbol coincides with read
+errors, which is straightforward to model. The $\uparrow$ symbol coincides
+with pivots, which have a more complex behavior. We thus extend the MEM
+alphabet so that the occurrence of the $\uparrow$ becomes easier to
+evaluate.
 
-With these definitions, the decision tree for choosing the next symbol to
-insert in a read can be summarized as in Fig.~\ref{fig:decision_tree}.
-This sketch will be useful as a reference to compute the probabilities of
-the blocks and their weighted generating functions.
+Only the terminators are extended; the $\square$ symbols remain unchanged.
+Let us denote the number of duplicates of the target by $N$. The
+$\downarrow$ symbol is replaced by the $N+1$ symbols
+$\downarrow_{/m}$, where $m$ is the number of duplicates that
+\emph{match} the erroneous nucleotide $(0 \leq m \leq N)$. The $\uparrow$
+symbol is replaced by the symbols $\uparrow_{/j}$, where $i
+\geq 0$ is the size of the segment (including the
+$\uparrow_{/j}$ symbol). Finally, we add the special terminator
+$|$ that marks the end of the read. In other words, tails are considered
+normal segments terminated by the symbol $|$ (with size 0).
 
-\begin{definition}
-The main thread is dominant at nucleotide $n$ if $\varphi_0(n) >
-\varphi_i(n)$ for all $i$ $(1 \leq i \leq N)$. The main thread is
-co-dominant if inequality is not strict and there exists $i$ $(1\leq i
-\leq N)$ such that $\varphi_0(n) = \varphi_i(n)$.
-\end{definition}
+It is important to highlight that the numbers that decorate the symbols
+have a very different nature. In the case of the $\downarrow$ symbol, it
+is an information about the number of sequences matching the read; in the
+case of the $\uparrow$ symbol, it is the position of the pivot relative to
+the previous terminator (or to the beginning of the read).
 
-%\begin{figure}[h]
-%\centering
-%\includegraphics[scale=1]{no_3MEM_graph.pdf}
-%\caption{\textbf{Transfer graph of a read without 3-MEM}. 
-%Some legend.}
-%\label{fig:noMEM_graph}
-%\end{figure}
+Fig.~\ref{fig:sketch_extended} shows the decomposition of the read from
+Fig.~\ref{fig:sketch_segment} in segments using the extended MEM alphabet.
+The numbers decorating the symbols are shown above them to avoid crowding
+the figure.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.85]{sketch_extended.pdf}
+\caption{\textbf{Reads as sequences of segments in the extended MEM
+alphabet}. 
+Some legend.}
+\label{fig:sketch_extended}
+\end{figure}
+
+There is an infinite amount of $\uparrow_{/j}$ symbols in the
+extended MEM alphabet, but if the read has no on-target MEM seed, then
+only the $\uparrow_{/j}$ symbols where $1 \leq i \leq \gamma-1$
+can be used (if a pivot terminates a segment of size greater than or equal
+to $\gamma$, then the read contains a strict on-target MEM seed). The
+extended MEM alphabet is made in such a way that the numbers decorating
+the $\uparrow$ symbol will allow us to exclude strict on-target MEM seeds,
+and the numbers decorating the $\downarrow$ symbol will help us exclude
+shared on-target MEM seeds.
+
+\section{Symbolic and analytic representations}
+
+\subsection{Weighted generating functions}
+
+Let $\mathcal{A}$ be a set of combinatorial objects which have a
+\emph{probability of occurrence}, and a \emph{size} that is an integer
+number. The function defined as $A(z) = \sum_{k=0}^\infty a_kz^k$, where
+$a_k$ is the total probability of objects of size $k$, is called the
+\emph{weighted generating function} of the combinatorial objects.
+
+In our case, we are interested in the weighted generating function of
+reads without on-target MEM seed, where the coefficient $a_k$ is the
+probability that a read of size $k$ has no on-target MEM seed. As $k$
+becomes larger, it becomes more difficult to evaluate $a_k$, but the
+combinatorial construction developed above will allow us to do this by
+using established strategies involving weighted generating functions.
+
+\subsection{The symbolic method}
+
+The symbolic method is a strategy belonging to the general \emph{analytic
+combinatorics} framework. The idea is to combine simple combinatorial
+objects into more complex objects. Each combinatorial operation on the
+objects corresponds to a mathematical operation on their weighted
+generating functions. One can thus obtain the weighted generating function
+of complex objects, and from there extract their probability of
+occurrence.
+
+The combinatorial construction of reads using the extended MEM seeding
+alphabet gives us a method to express their weighted generating function
+from that of segments. More specifically, we can describe complex
+sequences of segments through a so-called \emph{transfer matrix} $M(z)$
+that specifies which segments can follow one another in the reads of
+interest.
+
+The segments types are identified by their terminators and arranged in a
+predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
+weighted generating function of segments with the $j$-th terminator that
+can be inserted immediately after segments with the $i$-th terminator. The
+matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$ contains the
+weighted generating functions of the reads of all sizes. The entry that
+corresponds to terminators $\downarrow_{/0}$ and $|$ is the weighted
+generating function of the reads of interest (the beginning of the read is
+equivalent to a ghost segment terminated by $\downarrow_{/0}$,
+\textit{i.e.} a mismatch for all the sequences).
+
+With the transfer matrix $M(z)$, with thus have a way of computing the
+weighted generating function of the reads without on-target MEM seed, and
+from there we can extract teir probability of occurrence. Our next task is
+thus to specify $M(z)$.
 
@@
 
-\subsection{Transitions from $\uparrow$}
-\label{sec:trans_from_up}
+\subsection{Match streak and masks}
 
-On a $\uparrow$ symbol, the main thread becomes dominant. It will remain
-so until the first sequencing error, so the next block will be terminated
-by an error symbol (\textit{i.e.} $\Downarrow$ or $\downarrow$).
+For each sequence, we define the \emph{match streak} at a given nucleotide
+as the number of nucleotides since the last mismatch on the left. For
+instance, in Fig.~\ref{fig:MEM_example}, the match streak of the first
+genomic hit at the ninth nucleotide is $1$ while for the second genomic
+hit, it is $7$.
 
 \begin{definition}
-$\omega$ and $\tilde{\omega}$ are the probabilities of the $\Downarrow$
-and of the $\downarrow$ error symbols, respectively, \textit{i.e.}
-\begin{eqnarray}
-\omega &=& p \cdot (1-\mu/3)^N \\
-\tilde{\omega} &=& p \cdot \big( 1-(1-\mu/3)^N \big).
-\end{eqnarray}
+At a given position of the read, a duplicate sequence is a \emph{hard
+mask} if its success streak is strictly longer than the match streak of
+the target. A duplicate thread is a \emph{soft mask} if its has the same
+match streak as the target.
 \end{definition}
 
-The weighted generating function of the transition from $(\uparrow)_j$ to
-$\Downarrow$ is
-\begin{equation}
-C^\Downarrow_j(z) = \omega z \sum_{i=0}^{\gamma-j-1} (qz)^i.
-\end{equation}
+For instance, the second genomic hit in the example above is a hard mask
+at the ninth nucleotide. The point of hard and soft masks is that they are
+directly linked with MEMs. A nucleotide with a hard mask cannot belong to
+an on-target MEM. A nucleotide without mask belongs to a strict on-target
+MEM. A nucleotide with a soft mask may belong to a strict on-target MEM,
+to a shared on-target MEM, or may not belong to any MEM. The distinction
+between the three cases will be clarified in the specification of the
+transfer matrix.
 
-The weighted generating function of the transition from
-$(\uparrow)_j$ to $\downarrow$ is
-\begin{equation}
-C^\downarrow_j(z) = \tilde{\omega} z
-  \sum_{i=0}^{\gamma-j-1} (qz)^i.
-\end{equation}
+The extended MEM alphabet is nothing more than a way to keep track of the
+hard and soft masks. For instance, a pivot (associated with the symbol
+$\uparrow$) is a nucleotide where the number of masks becomes is reduced
+to 0. If there are $N$ duplicates, the symbol $\downarrow_{/m}$ indicates
+that there $m$ hard masks and $N-m$ soft masks.
 
-\subsection{Transitions from $\Downarrow$}
-\label{sec:trans_from_Down}
+\subsection{Error model and divergence of the duplicates}
 
-At a $\Downarrow$ symbol, all the threads have match length $0$ and the
-read is `reset' to its initial state. The main thread is co-dominant with
-all the alternative threads. The next terminator symbol depends on what
-happens first. If all the alternative threads fail before the main thread,
-the latter becomes dominant ($\uparrow$ symbol). If not all alternative
-threads have failed when the main thread does, we need to distinguish two
-cases: if all threads fail together with the main thread, the read is
-reset ($\Downarrow$ symbol), otherwise the main read becomes dominated
-($\downarrow$ symbol).
+We assume that the target sequence has $N$ duplicates. We further assume
+that duplication was instantaenous and that all $N+1$ sequences diverge
+independently from each other at a constant rate. In other words, we
+ignore the complications due to the genealogy of the duplication events
+and we simply assume that at each position, any given duplicate is
+identical to the target with probability $1-\mu$. If it is not, we
+assume that the duplicate has any of the remaining three nucleotides with
+equal probability (\textit{i.e.} each is found with probability $\mu/3$).
 
-If the next terminator is the $\uparrow$ symbol, we need to distinguish
-the cases based on the value of $\varphi_0$ (see
-section~\ref{sec:trans_from_up}), which is just the position of the
-$\uparrow$ symbol after the $\Downarrow$ symbol in this case. A transition
-from $\Downarrow$ to $(\uparrow)_j$ consists of a succession of $j$
-correct nucleotides -- each with probability $q$. But we must also make
-sure that no other $\uparrow$ symbol is inserted before position $j$.
+We also assume that the sequencing instrument has a constant substitution
+rate $p$, and that insertions and deletions never occur. In case of a
+substitution, we again assume that the remaining three nucleotides are
+equally likely.
+
+On a read error the target is never identical to the read (because we
+assume that the target is the true sequence), and a duplicate is identical
+to the read with probability $mu/3$ (the probability that the duplicate is
+different from the target, times the probability that the nucleotide is
+the same as the error). On a correct nucleotide, the target is always
+identical to the read, and a dupicate is identical with probability
+$1-\mu$ (the probablity that the duplicate is identical to the target).
+
+With these assumptions, we are now ready to describe the entries of the
+transfer matrix $M(z)$, which we do in order of increasing difficulty.
+
+\subsection{Segments following $\uparrow_{/j}$}
+
+A $\uparrow_{/j}$ symbol marks a pivot and the absence of masks. The
+segment is thus the beginning of an on-target MEM of size at least $j$.
+The MEM will extend until the next sequencing error, or until the end of
+the read, so the number of $\square$ symbols in the next segment must be
+at most $\gamma-j-1$ and it must be terminated by a $\downarrow$ symbol or
+by the tail terminator $|$.
+
+In this case, the matches between the read and the duplicates are
+irrelevant until the terminator. The weighted genearting function of each
+$\square$ symbol is simply $qz$ ($q$ is the probability that the
+nucleotide is not an error and $z$ marks objects of size $1$).
 
 \begin{definition}
-$\xi_{i,m}$ is the probability that at least one of $m$ alternative
-threads does not fail in $i$ error-free nucleotides, \textit{i.e.}
+The probability that a symbol is $\downarrow_{/m}$ given that the
+nucleotide is a read error is
 \begin{equation}
-  \xi_{i,m} = 1-(1-(1-\mu)^i)^m.
+\label{eq:omega}
+\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m.
 \end{equation}
@@
 
-The weighted generating function of the transition from $\Downarrow$ to
-$(\uparrow)_i$ is
+The weighted generating function of the $\downarrow_{/m}$ terminator is
+thus $\omega_m pz$ (the probability of the $\downarrow_{/m}$ is $\omega_m
+p$ and $z$ marks objects of size $1$).
+
+The entries of the transfer matrix that correspond to transitions from
+$\uparrow_{/j}$ to other segments are now easy to find. The weighted
+generating function of the segments terminated by $\downarrow_{/m}$
+following a segment terminated by $\uparrow_{/j}$ is
 \begin{equation}
-\label{eq:u}
-u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
+\label{eq:D}
+D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i.
 \end{equation}
 
-\begin{remark}
-If $N = 0$, the states $(\downarrow)_2, \ldots, (\downarrow)_{\gamma-1}$
-are inacessible. Consistently, the mononomial $u_i(z)$ is equal to $qz$ if
-$i = 1$ and $0$ otherwise.
-\end{remark}
-
-If the next terminator is $\Downarrow$, there is a limit on the
-number of $\square$ symbols. Indeed, if there are $\gamma$ or more
-nucleotides between the two $\Downarrow$ symbols, they constitute a shared
-MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
-generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
-same rationale as above, the weighted generating function of the
-transition from $\Downarrow$ to $\Downarrow$ is
+And the weighted generating function of the tail segments following
+segments terminated by $\uparrow_{/j}$ is
 \begin{equation}
-A^\Downarrow(z) = \omega z
-  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
+\label{eq:E}
+E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
 \end{equation}
 
-If the next terminator is $\downarrow$ we have to dinstinguish two cases.
-In the example below, there is a shared MEM seed only if $\gamma \leq
-11$. The main thread is unmasked, so the number of nucleotides between the
-error symbols matters.
-\begin{inset}
-\includegraphics[scale=.9]{ddown_to_down_case_1.pdf}
-\end{inset}
-In this other example, there can be no MEM seed because the first
-duplicate masks the main thread. This depends neither on the value of
-$\gamma$ nor on the number of nucleotides between the error symbols.
-\begin{inset}
-\includegraphics[scale=.9]{ddown_to_down_case_2.pdf}
-\end{inset}
+\subsection{Segments following $\downarrow_{/m}$}
 
-So if the $\Downarrow$ and the $\downarrow$ symbols are separated by
-$\gamma-1$ or fewer nucleotides, there is no MEM seed and we must only
-rule out the presence of other terminators in the segment (in this case
-the $\uparrow$ symbol). The weighted generating function is
-\begin{equation}
-\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
-\end{equation}
+At a $\downarrow_{/m}$ symbol, there are $m$ hard masks and $N-m$ soft
+masks. If all the masks disappear before the first read error, the next
+terminator will be a $\uparrow$ symbol, otherwise it will be a
+$\downarrow$ symbol.
 
-On the other hand, if the $\Downarrow$ and the $\downarrow$ symbols are
-separated by $\gamma$ or more nucleotides, there could be a shared MEM
-seed, so we must enforce the condition that at least one potentially
-masking thread masks the main thread up to and including the $\downarrow$
-symbol (this incidentally rules out the presence of other terminators in
-the segment).
+\subsubsection*{Case 1: the terminator is $\uparrow_{/i}$}
 
 \begin{definition}
-$\alpha_i$ is the probability that the first thread has at least one
-failure in $i$ error-free nucleotides, followed by a sequencing error,
-\textit{i.e.}
+The probability that a given duplicate sequence contains a mismatch in a
+sequence of $i$ error-free nucleotides is
 \begin{equation}
-\alpha_i = 1 - (1-\mu)^i\mu/3
+\label{eq:xi}
+\xi_i = 1-(1-\mu)^i.
 \end{equation}
@@
 
-The probability that at least one duplicate thread masks the main thread
-if the first error occurs at position $i$ is thus $1 - \alpha_{i-1}^N$.
-The weighted generatin function is
+With this notation, the probability that at least one mask survives a
+sequence of $i$ error-free nucleotides is thus $1-\xi_i^N$, and the
+probability that there remains a mask at the $i-1$-th but not at the 
+$i$-th error-free nucleotide is $\xi_i^N - \xi_{i-1}^N$. From this we
+conclude that the weighted generating function of the segments terminated
+by $\uparrow_{/i}$ following a segment terminated by $\downarrow_{/m}$ is
 \begin{equation}
-pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
+\label{eq:B}
+B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i.
 \end{equation}
 
-Finally, the weighted generating function of the transition from
-$\Downarrow$ to $\downarrow$ is
+In the equation above, $q^i$ is the probability that there is no
+sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
+As mentioned above, $i < \gamma$ for reads without on-target MEM seed.
+Also note that this expression is the same for all $\downarrow$ symbols;
+it does not depend on $m$.
+
+\subsubsection*{Case 2a: the terminator $|$ comes before the $\gamma$-th
+nucleotide}
+
+In this case there can be no on-target MEM seed. We must just enforce the
+condition that at least one of the $N$ masks survives until the end of the
+segment in order to exclude the $\uparrow$ symbol. The weighted generating
+function is
+\begin{equation*}
+\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
+\end{equation*}
+
+\subsubsection*{Case 2b: the terminator $|$ comes after the $\gamma$-th
+nucleotide}
+
+In this case, the soft masks are irrelevant. Even if they survive until
+the end of the segment, there will be a shared on-target MEM seed. So we
+must enforce the condition that at least one hard mask survives until the
+end of the segment (which is impossible if $m = 0$).
+\begin{equation*}
+\sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
+\end{equation*}
+
+The transition from $\downarrow_{/m}$ to $|$ is
 \begin{equation}
-A^\downarrow(z) =
-\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
-pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
+\label{eq:C}
+C_m(z) =
+\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
+  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
 \end{equation}
 
-\subsection{Transitions from $\downarrow$}
-\label{sec:trans_from_down}
+\subsubsection*{Case 3a: the terminator $\downarrow_{/n}$ comes before the
+$\gamma$-th nucleotide}
 
-At a $\downarrow$ symbol, the main thread fails and at least one duplicate
-thread does not (othwerwise the symbol would be $\Downarrow$). By
-definition, the duplicate threads that do not fail are strictly masking,
-and those that fail are potentially masking. A $\downarrow$ symbol implies
-the presence of a sequencing error, so duplicate threads fail with
-probability $1-\mu/3$. The probability that $m$ duplicate threads fail and
-$N-m$ do not $(0 \leq m \leq N-1)$ is thus
+As above, there can be no on-target MEM seed and we must just exclude the
+$\uparrow$ terminator. The weighted generating function is
 \begin{equation}
-  Q_m = \frac{1}{1-(1-\mu/3)^N}{N \choose m} (1-\mu/3)^m(\mu/3)^{N-m}.
+\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
 \end{equation}
 
-The probability that the next terminator is the $\uparrow$ symbol at
-position $i$ is the probability that the last of the $N$ duplicate threads
-fails at position $i$ is $\xi_{i-1,N-m}-\xi_{i,N-m}$ so the weighted
-generating function of the transition from $\downarrow$ to $(\uparrow)_i$
-is
+\subsubsection*{Case 3a: the terminator $\downarrow_{/n}$ comes after the
+$\gamma$-th nucleotide}
+
+This case is by far the most convoluted. Since the segment contains at
+least $\gamma$ error-free nucleotides, we must take care that it does not
+generate an on-target MEM seed. This will be the case if any of the two
+following conditions is validated: $i)$ at least one hard mask covers all
+the error-free nucleotides, or $ii)$ none of the hard masks does but at
+least one soft mask covers the whole segment (including the terminator).
+
+The two conditions are mutually exclusive by construction. They are
+graphically represented in the diagram below. The left panel corresponds
+to case $i)$ and the panel to case $ii)$. The top row represents the
+target, and the bottom rows represent duplicates (using the same
+symbols as in Fig.~\ref{fig:sketch_MEM}).
+\begin{inset}
+\includegraphics{masks.pdf}
+\end{inset}
+
+Whenever a hard mask (here the first duplicate) covers the nucleotides
+as shown in the left panel, there can be no on-target MEM seed,
+irrespective of the value of $\gamma$. The positions marked with a
+question mark are irrelevant, they cannot change the fact that there is no
+on-target MEM seed. If the hard masks are interrupted, as in the right
+panel, then it depends on the soft masks. If a soft mask covers the whole
+segment, then there can be no on-target MEM seed, irrespective of the
+value of $\gamma$.
+
+Condition $i)$ has probability $\big(1 - \xi_i^m \big)$ and the
+weighted generating function is thus
+\begin{equation*}
+\omega_n pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
+\end{equation*}
+
+For condition $ii)$, we will need to introduce some new notations.
+\begin{definition}
+The probability that a given duplicate sequence contains a mismatch in a
+sequence of $i$ error-free nucleotides followed by an error is
 \begin{equation}
-\tag{\ref{eq:u}}
-u_i(z) = (\xi_{i-1,N}-\xi_{i,N}) \cdot (qz)^i.
+\label{eq:eta}
+\eta_i = 1-(1-\mu)^i\mu/3.
+\end{equation}
+\end{definition}
+
+The probability of condition $ii)$ is
+\begin{equation*}
+\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
+\end{equation*}
+but we need to break up this term among the terminators $\downarrow_{/n}$,
+($0 \leq n \leq N$). For this, we need to keep track of the number of soft
+masks that run to the end of the segment, including the terminator. The
+probability that there are $r \geq 1$ such soft masks is
+\begin{equation*}
+{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
+\end{equation*}
+
+Each of the terminator, so the total number of matches is $r$ plus the
+number of sequences, among the $N-r$ that remain, that also match the
+terminator. The probability that in total $n$ sequences match the
+terminator is
+\begin{eqnarray*}
+&\;& \sum_{r\geq1} {N-m \choose r} (1 - \eta_i)^r
+\eta_i^{N-m-r} {N-r \choose n-r} (\mu/3)^{n-r} (1-\mu/3)^{N-n} \\
+&=& \eta_i^{N-m} (\mu/3)^n(1-\mu/3)^{N-n}
+  \sum_{r\geq1} {N-m \choose r} {N-r \choose n-r}
+  \left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \\
+&=& \omega_n \cdot \eta_i^{N-m} \cdot \zeta_{i,m,n},
+\end{eqnarray*}
+where
+\begin{equation}
+\label{eq:zeta}
+\zeta_{i,m,n} = \sum_{r\geq1}
+{N-m \choose r} {N-r \choose n-r}
+\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
 \end{equation}
 
-To compute the weighted generating function of the transition from
-$\downarrow$ to $\Downarrow$, we need to separate two cases. In the first,
-the error symbols are separated by $\gamma-1$ or fewer error-free
-nucleotides. In this case, there can be no MEM seed and we only have to
-exclude the $\uparrow$ terminator. We obtain
+%\begin{remark}
+%It is possible to approximate $\zeta_{i,m,n}$ with a simpler expression as
+%follows. We start with general expressions of the form
+%\begin{equation*}
+%\sum_{r\geq0} {N-m \choose r} {N-r \choose n-r} x^r.
+%\end{equation*}
+%
+%Applying $m$ times the identity ${a \choose b} = {a-1 \choose b} + {a-1
+%\choose b-1}$ we obtain
+%\begin{equation*}
+%{N-m \choose n-r} = \sum_{s=0}^m {m \choose s} {N-r-m \choose n-r-s}.
+%\end{equation*}
+%
+%Combining the binomial terms, we can see that
+%\begin{equation*}
+%{N-m \choose r} {N-r-m \choose n-r-s} = {N-m \choose n-s} {n-s \choose r}.
+%\end{equation*}
+%
+%This finally leads to the following expression featuring a sum of
+%hypergeometric probabilities
+%\begin{equation*}
+%\sum_{s=0}^m {m \choose s} {N-m \choose n-s}  \sum_{r\geq0} {n-s
+%\choose r} x^r = \sum_{s=0}^m {m \choose s} {N-m \choose n-s} (1+x)^{n-s}.
+%\end{equation*}
+%
+%The hypergeometric distribution can be approximated by the binomial
+%distribution using the following formula
+%\begin{equation*}
+%{m \choose s} {N-m \choose n-s} \Bigg/ {N \choose n} \approx
+%{n \choose s} (m/N)^s (1-m/N)^{n-s}.
+%\end{equation*}
+%
+%We can thus simplify the sum above using Newton's binomial theorem. We
+%obtain
+%\begin{equation*}
+%{N \choose n} \big( 1+x \cdot (1-m/N) \big)^n.
+%\end{equation*}
+%
+%From this we conclude
+%\begin{equation*}
+%\sum_{r=1}^{N-m} {N-m \choose r} {N-r \choose n-r} x^r  \approx
+%{N \choose n} \Big[ \big( 1+x \cdot (1-m/N) \big)^n - 1 \Big].
+%\end{equation*}
+%
+%Substituting the value of $x$, we finally obtain
+%\begin{equation*}
+%\zeta_{i,m,n} \approx \omega_n \xi_i^m \cdot \lambda_{i,m,n},
+%\end{equation*}
+%where
+%\begin{equation}
+%\lambda_{i,m,n} = \eta_i^{N-m} \left[ \left( 1+ \frac{1-\eta_i}{\eta_i \mu
+%/ 3} \cdot \frac{N-m}{N} \right)^n - 1 \right].
+%\end{equation}
+%\end{remark}
+
+The weighted generating function we are looking for is thus
 \begin{equation}
-\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i.
+\label{eq:A}
+A_{m,n} =
+\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
+pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
+\zeta_{i,m,n}) \Big) (qz)^i.
 \end{equation}
 
-In the case that $\downarrow$ and $\Downarrow$ are separated by $\gamma$
-or more error-free nucleotides, we must ensure that at least one strictly
-masking thread survives until the penultimate nucleotide of the segment.
-This probability is equal to
-\begin{equation}
-\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} =
-\frac{1-\alpha_i^N}{1-(1-\mu/3)^N}.
-\end{equation}
+\subsection{The expression of the transfer matrix}
 
-\begin{remark}
-A strictly masking thread survives until the $\Downarrow$ symbol if and
-only if one of the $N$ duplicate threads reverse-survives from
-the penultimate nucleotide of the segment down to and including the
-$\downarrow$ symbol.
-\end{remark}
+The final expression for the transfer matrix $M(z)$ is
+\begin{equation*}
+\begin{blockarray}{cccccccc}
+   & \dn{0} & \ldots & \dn{N} & \up{1} & \ldots & \up{\gamma-1} & \nd \\
+\begin{block}{c[ccccccc]}
+\dn{0} & A_{0,0}(z) & \ldots & A_{0,N}(z) & B_1(z) & \ldots &
+    B_{\gamma-1}(z) & C_0(z) \\
+\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
+    \vdots & \vdots \\
+\dn{N} & A_{N,0}(z) & \ldots & A_{N,N}(z) & B_1(z) & \ldots &
+    B_{\gamma-1}(z) & C_N(z) \\
+\up{1} & D_{1,0}(z) & \ldots & D_{1,N}(z) & 0 & \ldots & 0 & E_1(z) \\
+\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
+    \vdots & \vdots \\
+\up{\gamma-1} & D_{\gamma-1,0}(z) & \ldots & D_{\gamma-1,N}(z) & 0 &
+  \ldots & 0 & E_{\gamma-1}(z) \\
+\nd & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}
+\end{equation*}
+where
+\begin{gather}
+\tag{\ref{eq:A}}
+A_{m,n} =
+\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
+pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
+\zeta_{i,m,n}) \Big) (qz)^i \\
+\tag{\ref{eq:B}}
+B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i \\
+\tag{\ref{eq:C}}
+C_m(z) =
+\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
+  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i \\
+\tag{\ref{eq:D}}
+D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i \\
+\tag{\ref{eq:E}}
+E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i
+\end{gather}
+and where
+\begin{gather}
+\tag{\ref{eq:omega}}
+\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m \\
+\tag{\ref{eq:xi}}
+\xi_i = 1-(1-\mu)^i \\
+\tag{\ref{eq:eta}}
+\eta_i = 1-(1-\mu)^i\mu/3 \\
+\tag{\ref{eq:zeta}}
+\zeta_{i,m,n} = \sum_{r\geq1}
+{N-m \choose r} {N-r \choose n-r}
+\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
+\end{gather}
 
-In sum, the weighted generating function of the transition from
-$\downarrow$ to $\Downarrow$ is
-\begin{equation}
-B^\Downarrow(z) = 
-\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
-\omega z\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
-\end{equation}
+\section{Computing the probabilities}
 
-If the next terminator is $\downarrow$, we again have to distinguish two
-cases. If the two $\downarrow$ symbols are separated by $\gamma-1$ or
-fewer nucleotides, there is no MEM seed and we must only rule out the
-presence of the $\uparrow$ terminator. We obtain
-\begin{equation}
-\tilde{\omega} z\sum_{i=0}^{\gamma-1}
-\xi_{i,N} \cdot (qz)^i.
-\end{equation}
+\subsection{Coefficient extraction}
 
-If the two $\downarrow$ symbols are separated by $\gamma$ or more
-nucleotides, there could be a shared MEM seed, so we must enforce the
-condition that the main thread is masked. This is satisfied if and only if
-one of the following conditions holds:
-\begin{enumerate}
-\item A strictly masking thread survives until the penultimate
-nucleotide of the segment.
-\item All the strictly masking threads fail before the penultimate
-nucleotide but a potentially masking thread survives until the last
-nucleotide of the segment.
-\end{enumerate}
+The purpose of constructing the weighted generating function $F(z) = a_0 +
+a_1z + a_2z^2 + \ldots$ is to extract the coefficient $a_k$, which
+represents the probability that a read of size $k$ does not contain an
+on-target MEM seed. $F(z)$ is the entry at coordinates $(1,N+\gamma+1)$ of
+the matrix $M(z) \cdot (I-M(z))^{-1}$. The expression of $M(z)$ is in
+general too complex to compute this function directly. Instead, we return
+to the expression $M(z) \cdot (I-M(z))^{-1} = M(z) + M(z)^2 + \ldots$ and
+observe that the terms $M(z)^{k+2}, M(z)^{k+3}, \ldots$ have no influence
+on the coefficients $a_0, a_1, \ldots, a_k$.
 
-Assuming that there are $m \geq 0$ potentially masking threads, the
-probability of condition $1.$ is $\xi_{i-1,N-m}$ and that of condition 2.
-is $(1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m)$. Summing of $m = 0, \ldots, N-1$
-on obtains
-\begin{equation}
-(1-\mu/3)\sum_{m=0}^{N-1}Q_m \cdot \xi_{i-1,N-m} =
-1-\alpha_{i-1}^N \text{, and}
-\end{equation}
-\begin{equation}
-\sum_{m=0}^{N-1}Q_m (1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m) =
-\frac{\alpha_{i-1}^N - \alpha_0^N -
-  \zeta_{i-1,i-1}^N + \zeta_{0,i-1}^N}{1-(1-\mu/3)^N},
-\end{equation}
-where
-\begin{equation}
-\zeta_{j,i} = 1-(1-\mu)^j\mu/3-(1-\mu)^i\mu/3(1-\mu/3)
-\end{equation}
+Indeed, a read of size $k$ has at most $k+1$ segments. Since the entries
+of $M(z)^k$ are the weighted generating functions of reads with exactly
+$k$ segments, $a_k$ cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More
+formally, we can prove by induction that all the entries of $M(z)^k$ are
+divisible by $z^{k-1}$, showing that the contribution of $M(z)^{k+2} +
+M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is $0$.
 
-Finally, the weighted generating function of the transitions from
-$\downarrow$ to $\downarrow$ is
-\begin{equation*}
-B^\downarrow(z) = \tilde{\omega}z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot
-(qz)^i + pz \sum_{i=\gamma}^\infty
-\left( 1-\alpha_i^N +
-\frac{\alpha_i^N - \alpha_0^N - \zeta_{i,i}^N +
-\zeta_{0,i}^N}{1-(1-\mu/3)^N} \right) (qz)^i
-\end{equation*}
+So we can extract the coefficients of $F(z)$ up to order $k$ by computing
+the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$, and compute the first
+terms of the Taylor expansion of the entry at coordinates
+$(1,N+\gamma+1)$.
 
-\begin{remark}
-One can interpret $\zeta_{i,i}$ as the probability that
-\end{remark}
+But we can do better than that. If we are only interested in the
+coefficients of order up to $k$, we can replace the infinite power series
+of the terms $A_{m,n}(z)$ and $C_m(z)$ by their truncated versions where
+we keep only the terms of order up to $k$. Likewise, we perform all
+algebraic operation on truncated polynomials of order $k$, \textit{i.e.}
+we discard at every stage the coefficients of order greater than $k$. With
+this approach, the entry of the matrix $M(z) + M(z)^2 + \ldots +
+M(z)^{k+1}$ at coordinates $(1,N+\gamma+1)$ is a truncated polynomial that
+directly provides the coefficients of interest.
 
-\subsection{Head and tail vectors}
+But we can do better than that. A read with $k$ segments contains at least
+$(k-1)/2$ errors. Indeed, two segments terminated by $\uparrow$ cannot
+follow each other, so at least (approximately) half of the segments must
+be terminated by $\downarrow$, which means that the segment contains a
+read error. Since read errors typically have a small probability of
+occurrence, all the coefficients of $M(z)^k$ rapidly converge to $0$ and
+become negligible as $k$ increases. So instead of computing $M(z) + M(z)^2
++ \ldots + M(z)^{k+1}$, we can interrupt the summation after a certain
+power of $M(z)$ because the terms are negligible.
 
-At the beginning of the read, the match lengths of all the threads are
-$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
-consider that the read starts in the $\Downarrow$ state, which is
-equivalent to adding an edge labelled with the empty object $\varepsilon$
-between the head vertex and the $\Downarrow$ vertex. The head vector is
-thus
-\begin{equation*}
-H(z) = 
-\begin{blockarray}{ccccc}
-   \;\; \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
-\begin{block}{[ccccc]}
-\;\; 1 & 0 & 0 & \ldots & 0 \\
-\end{block}
-\end{blockarray}.
-\end{equation*}
-The tail vector contains the weighted generating functions of
-non-terminated stretches of $\square$ symbols. Referring to the decision
-tree of Fig.~\ref{fig:decision_tree}, we see that the probability of
-occurrence of $\square$ symbols depends on whether there are potentially
-masking and / or strictly masking threads. All the symbols except
-$\downarrow$ impose a maximum on the number of $\square$ symbols at the
-end of the reads.
+This approach makes it feasible to compute the probability that a read of
+size $k$ has no on-target MEM seed when the matrix $M(z)$ is small, but
+recall that $M(z)$ has dimension $N+\gamma+1$. Some sequences can have
+million of duplicates in mammalian genomes (\textit{i.e.} $N > 10^6$),
+which makes the computations with this method unfeasible. Clearly, another
+method is needed.
 
-For the $\downarrow$ symbol, the weighted generating function is
-\begin{equation}
-T^\downarrow(z) = 
-\sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
-\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
-\end{equation}
+\subsection{Monte Carlo sampling}
 
-For the $(\uparrow)_j$ state, the weighted generating function is
-\begin{equation}
-T^\uparrow_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
-\end{equation}
+The symbolic representation of reads in the extended MEM alphabet can also
+be used as a basis for an efficient method to sample reads. Instead of
+generating the nucleotides of each sequence one by one and comparing them
+to the read, one can generate segments of the extended MEM alphabet. As
+argued above, a read contains few segments when the error rate is small.
+Since this property does not depend on the value of $N$, we can obtain a
+fast Monte Carlo method to sample millions of reads and count the
+proportion that contain an on-target MEM seed.
 
-For the other symbols, we must take care that the potentially masking
-threads do not all fail (otherwise this would imply the presence of a
-$\uparrow$ terminator in the tail). For the $\Downarrow$ symbol, the
-weighted generating function is
-\begin{equation}
-T^\Downarrow(z) = \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
-\end{equation}
+The principle is to first sample the position of the next read error, and
+then sample the number of masks that survive until this position. If none
+of the does, the read contains an on-target MEM seed, provided the next
+read error is at a distance greater than $\gamma$. Otherwise, the process
+is repeated for the next segment until an on-target MEM seed is generated,
+or until the read has size $k$ or greater (and is then truncated).
 
-In summary,
-\begin{equation*}
-T(z) = 
-\begin{blockarray}{cc}
-\begin{block}{c[c]}
-\DN & T^\Downarrow(z) \\
-\dn & T^\downarrow(z) \\
-\up{\gamma-1} & T^\uparrow_{\gamma-1}(z) \\
-\vdots & \vdots \\
-\up{1} & \; T^\uparrow_1(z) \; \\
-\end{block}
-\end{blockarray}.
-\end{equation*}
+The method is summarized by the algorithm below.
+
+\begin{algorithm}[H]
+\SetAlgoLined
+\KwResult {Sample a read, return 1 if it contains an on-target MEM seed,
+  otherwise return 0.}
+  $read.size \leftarrow 0$\;
+  $m \leftarrow 0$ \Comment*[r]{Current number of hard masks.}
+  \While {$read.size < k$}{
+    $i \leftarrow geom(p)$ \Comment*[r]{Position of next error.}
+    \eIf {$read.size + i > k$}{
+      $\lambda \leftarrow k - read.size$ \Comment*[r]{Size of the tail.}
+      \eIf{$\lambda < \gamma$} {
+        return 0\; }{
+        $h \leftarrow binom(m,(1-\mu)^\lambda)$
+            \Comment*[r]{Surviving hard masks.}
+        return 1 if $h = 0$, otherwise return 0\; }
+    }
+    {
+      $h \leftarrow binom(m,(1-\mu)^i)$
+          \Comment*[r]{Surviving hard masks.}
+      $s \leftarrow binom(N-m,(1-\mu)^i \mu/3)$
+          \Comment*[r]{Surviving soft masks.}
+      \eIf {$ i \geq \gamma$ and $h = 0$ and $s = 0$}{
+        return 1\;}{
+        $m \leftarrow s + binom(N-s, \mu/3)$\;
+        $read.size \leftarrow read.size + i$\;}
+    }
+ }
+\end{algorithm}
 
@@
 
-\begin{equation*}
-\begin{blockarray}{cccccc}
-   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
-\begin{block}{c[ccccc]}
-\DN & A^\Downarrow(z) & A^\downarrow(z) & u_1(z)
-    & \ldots & u_{\gamma-1}(z) \\
-\dn & B^\Downarrow(z) & B^\downarrow(z) & u_1(z)
-    & \ldots & u_{\gamma-1}(z) \\
-\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 &
-    \ldots & 0 \\
-\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
-\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & \ldots & 0 \\
-\end{block}
-\end{blockarray}
-\end{equation*}
-
-In the case $N = 0$, the symbol $\downarrow$ also disappears. Some
-polynomials have a particularly simple form in this case, and the
-body of the transfer matrix is
-\begin{equation*}
-\begin{blockarray}{ccc}
-   & \DN & \up{1} \\
-\begin{block}{c[cc]}
-\DN    & pz              & qz \\
-\up{1} & C_{\gamma-1}(z) & 0  \\
-\end{block}
-\end{blockarray}
-\end{equation*}
+This is awesome.
 
@@
 
-I acknowledge the financial support of the Spanish Ministry of Economy and
-Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan
+We acknowledge the financial support of the Spanish Ministry of Economy
+and Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan
 Nacional BFU2012-37168), of the CERCA Programme~/~Generalitat de

### 2018-08-10 MEM.tex
Fixed the theory (and checked exactness through implementation).

@@
 \label{eq:B}
-B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i.
+B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i.
 \end{equation}
@@
 but we need to break up this term among the terminators $\downarrow_{/n}$,
-($0 \leq n \leq N$). For this, we need to keep track of the number of soft
+($0 \leq n \leq N$). For this, we decompose the sum on the number of soft
 masks that run to the end of the segment, including the terminator. The
@@
 
-Each of the terminator, so the total number of matches is $r$ plus the
-number of sequences, among the $N-r$ that remain, that also match the
-terminator. The probability that in total $n$ sequences match the
-terminator is
+Each of them matches the nucleotide at the end of the segment, so the
+total number of matches is $r$ plus the number of sequences, among the
+remaining $N-m-r$ soft masks and $m$ hard masks, that also match the
+last nucleotide.
+
+Let us start with the $N-m-r$ that failed somewhere in the segment. The
+conditional probability that they match the last nucleotide given that
+they fail somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
+conditional probability that they do not match is $(1-\mu/3) / \eta_i$.
+For the $m$ hard masks, the probability that they match the last
+nucleotide given that they sail somewhere in the segment is simply
+$\mu/3$, and the probability that they do not is $1-\mu/3$.
+
+The probability that those $N-r$ duplicates contribute $n-r$ matches is
+\begin{equation*}
+\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_i^{N-m-r}}
+\psi_{i,m,n,r}\text{, where \;}
+\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
+\xi_i^{n-r-q}.
+\end{equation*}
+
+The probability that in
+total $n$ sequences match the terminator is
 \begin{eqnarray*}
-&\;& \sum_{r\geq1} {N-m \choose r} (1 - \eta_i)^r
-\eta_i^{N-m-r} {N-r \choose n-r} (\mu/3)^{n-r} (1-\mu/3)^{N-n} \\
-&=& \eta_i^{N-m} (\mu/3)^n(1-\mu/3)^{N-n}
-  \sum_{r\geq1} {N-m \choose r} {N-r \choose n-r}
-  \left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \\
-&=& \omega_n \cdot \eta_i^{N-m} \cdot \zeta_{i,m,n},
+&\;& \sum_{r\geq1} {N-m \choose r}
+(1 - \eta_i)^r (\mu/3)^{n-r} (1-\mu/3)^{N-n} \psi_{i,m,n,r} \\
+&=& (\mu/3)^n(1-\mu/3)^{N-n} \sum_{r\geq1} {N-m \choose r}
+  (1 - \mu)^{ri} \psi_{i,m,n,r} \\
+&=& \omega_n \cdot \zeta_{i,m,n},
 \end{eqnarray*}
@@
 \label{eq:zeta}
-\zeta_{i,m,n} = \sum_{r\geq1}
-{N-m \choose r} {N-r \choose n-r}
-\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
+\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
+(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n}.
 \end{equation}
-
-%\begin{remark}
-%It is possible to approximate $\zeta_{i,m,n}$ with a simpler expression as
-%follows. We start with general expressions of the form
-%\begin{equation*}
-%\sum_{r\geq0} {N-m \choose r} {N-r \choose n-r} x^r.
-%\end{equation*}
-%
-%Applying $m$ times the identity ${a \choose b} = {a-1 \choose b} + {a-1
-%\choose b-1}$ we obtain
-%\begin{equation*}
-%{N-m \choose n-r} = \sum_{s=0}^m {m \choose s} {N-r-m \choose n-r-s}.
-%\end{equation*}
-%
-%Combining the binomial terms, we can see that
-%\begin{equation*}
-%{N-m \choose r} {N-r-m \choose n-r-s} = {N-m \choose n-s} {n-s \choose r}.
-%\end{equation*}
-%
-%This finally leads to the following expression featuring a sum of
-%hypergeometric probabilities
-%\begin{equation*}
-%\sum_{s=0}^m {m \choose s} {N-m \choose n-s}  \sum_{r\geq0} {n-s
-%\choose r} x^r = \sum_{s=0}^m {m \choose s} {N-m \choose n-s} (1+x)^{n-s}.
-%\end{equation*}
-%
-%The hypergeometric distribution can be approximated by the binomial
-%distribution using the following formula
-%\begin{equation*}
-%{m \choose s} {N-m \choose n-s} \Bigg/ {N \choose n} \approx
-%{n \choose s} (m/N)^s (1-m/N)^{n-s}.
-%\end{equation*}
-%
-%We can thus simplify the sum above using Newton's binomial theorem. We
-%obtain
-%\begin{equation*}
-%{N \choose n} \big( 1+x \cdot (1-m/N) \big)^n.
-%\end{equation*}
-%
-%From this we conclude
-%\begin{equation*}
-%\sum_{r=1}^{N-m} {N-m \choose r} {N-r \choose n-r} x^r  \approx
-%{N \choose n} \Big[ \big( 1+x \cdot (1-m/N) \big)^n - 1 \Big].
-%\end{equation*}
-%
-%Substituting the value of $x$, we finally obtain
-%\begin{equation*}
-%\zeta_{i,m,n} \approx \omega_n \xi_i^m \cdot \lambda_{i,m,n},
-%\end{equation*}
-%where
-%\begin{equation}
-%\lambda_{i,m,n} = \eta_i^{N-m} \left[ \left( 1+ \frac{1-\eta_i}{\eta_i \mu
-%/ 3} \cdot \frac{N-m}{N} \right)^n - 1 \right].
-%\end{equation}
-%\end{remark}
 
@@
 \omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
-pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
-\zeta_{i,m,n}) \Big) (qz)^i.
+pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
+(1- \zeta_{i,m,n}) \Big) (qz)^i.
 \end{equation}
@@
 \omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
-pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
-\zeta_{i,m,n}) \Big) (qz)^i \\
+pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
+(1- \zeta_{i,m,n}) \Big) (qz)^i \\
 \tag{\ref{eq:B}}
-B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i \\
+B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i \\
 \tag{\ref{eq:C}}
@@
 \tag{\ref{eq:zeta}}
-\zeta_{i,m,n} = \sum_{r\geq1}
-{N-m \choose r} {N-r \choose n-r}
-\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
+\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
+(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n} \\
+\notag
+\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
+\xi_i^{n-r-q}.
 \end{gather}
@@
 
+\section{Estimating the parameters}
+
+Let us assume that we have a sequence of length $L$ that has $N$
+duplicates in the genome. Let us further assume that the instrument has a
+uniform error rate of $p$, and that at each nucleotide, duplicates are
+different from the original sequence with probability $\mu$. What are the
+values of $p$ and $\mu$ that maximize the probability that the read is
+different from the original, but identical to a duplicate?
+
+If the read has no error, this cannot happen. If we consider the reads
+with one error, the probability that this happens is
+\begin{equation}
+Q(p,\mu) = L \cdot p \cdot (1-p)^{L-1}
+\cdot \Big( 1- (1-(1-\mu)^{L-1} \mu/3)^N \Big).
+\end{equation}
+
+Observe that for every value of $p$, $\mu = 1/L$ maximizes $Q(p,\mu)$. So
+in this simple proble, there exists a ``worst'' value of $\mu$, that
+maximizes the probability of making a mistake irrespective of the
+precision of the sequencing instrument and of the number of duplicates.
+
+The MEM seeding heuristic is more complex than the toy problem above, but
+we still get the intuition that values of $\mu$ close to $1/L$ are the
+most problematic, where $L$ is the typical size of MEM seeds.
+
+Using the above methods to compute MEM seeding probabilities, we can test
+this claim with typical values of the parameters. For the human genome,
+seeds of length $\approx 20$ are the norm...
+
 \section{Conclusion}

### 2019-01-26 MEM.tex
Added references, started to write a clear document.

@@
 
-\subsection{Mapping}
-
-Consider the following task referred to as the \emph{true mapping
-problem}: a random DNA fragment from a known genome is sequenced with an
-imperfect instrument; can we identify the location of the fragment in the
-genome?
-
-It seems that the answer depends only on the size of the fragment and the
-error rate of the instrument, but there is another important factor to
-consider. Most genomes contain repeated sequences. Intuitively, the
-problem is more difficult if the random DNA fragment is duplicated. If one
-of the duplicates is exactly identical to the original, then the DNA
-fragment cannot be mapped with certainty. Even if the duplicates are all
-different from the original (which happens because of mutations in the
-genome), the read may still be closer to one of the duplicates because of
-sequencing errors. So we can never be sure that the read is mapped to the
-\emph{true} location.
-
-For this reason, we introduce the \emph{best mapping problem}, where we
-search not the true, but the \emph{best} location of the DNA fragment,
-\textit{i.e.} the sequence of the genome that is the most similar to the
-read.
-
-There are exact solutions to the best mapping problem, but the algorithms
-are too slow to process the large amount of data generated by modern
-sequencers. Instead, one uses heuristic methods, \textit{i.e.} algorithms
-that run fast, but are not guaranteed to return the correct answer. In the
-context of the best mapping problem, there are thus three possible
-outcomes: $i.$ the DNA fragment is mapped to the best location, which is
-called a \emph{true positive} (or an \emph{on-target hit}), $ii.$ the DNA
-fragment is mapped to another location, which is called a \emph{false
-positive} (or an \emph{off-target hit}), $iii.$ the DNA fragment is not
-mapped at all, which is called a \emph{false negative}. A \emph{true
-negative} would designate the outcome where a DNA fragment foreign to the
-genome is not mapped at all, but here we assume that all DNA fragments
-belong to the genome of interest.
-
-In a seeding heuristic, one uses exact or near exact matches between the
-read and the genome to collect a set of candidate locations. The read is
-only compared to the sequences at the candidate locations, but not to the
-rest of the genome. The advantage of this method is that there are few
-candidates and that they can be discovered fast. However, the best hit may
-not be in the candidate set, in which case it cannot be discovered
-(\textit{i.e.}, the outcome is either a false positive or a false
-negative).
-
-A \emph{seed} designates a subsequence of fixed length that is shared
-between the read and the genome. In other words, a seed is a perfect match
-between a part of the read an a genomic site. The candidates consist of
-the genomic sites that contain one or more seeds (this depends on the
-specific heuristic).
-
-In the theory developed below, we will assume that the \emph{true}
-location of the read is also the \emph{best}. This is not always the case,
-but it is unfeasible to model the behavior of the best hit in the present
-context. We will later discuss how this relates to the probability that
-the best hit has a seed.
-
-\subsection{Maximal Exact Matches (MEMs)}
-
-Here we consider a seeding variant where seeds do not have a fixed
-length. Let us start by introducing the definition of a Maximal Exact
-Match.
+\subsection{Mapping reads to genomes}
+
+Suppose that a DNA fragment drawn at random from a known
+genome\footnote{We will assume throughout that the sequence of the genome
+is known exactly and with absolute certainty. This is not the case in
+practical application, but this assumption will allow us to develop a
+theory centered on the read.} is sequenced with an imperfect instrument;
+how can we identify the location of the fragment in the genome?
+
+We will refer to this question as the \emph{true location problem}.
+It is central to countless applications, including discovering
+disease-causing mutations; diagnosing viral, bacterial or fungal
+infections; diagnosing cancer and choosing appropriate therapies;
+selecting breeds of interest in agriculture; tracing human migrations from
+ancient bones; identifying the donor of a biological sample; diagnosing
+genetic diseaes at birth; finding compatible graft donors \textit{etc}.
+
+The focus of this document is solve the true location problem \textit{per
+se}, but rather to evaluate whether a solution is correct. At first sight,
+whether we can map reads to their true location seems to depend only on
+their lengths and on the error rate of the instrument. However, there is
+another important factor to consider: most genomes contain repeated
+sequences, \textit{i.e.} relatively large subsequences (from $\sim20$~bp
+to more than 10~kb) that are present at more than one location. So we
+cannot map the fragment with certainty if there exists an exact copy of
+the sequence at somewhere else in the genome, because we cannot know which
+of the two copies was sequenced.
+
+This case is rare though. Repeated sequences are in general not identical,
+but merely homologous\footnote{By that we mean that drawing nucleotides
+uniformly at random is very unlikely to produce sequences with the
+observed degree of identity.}. So when the DNA fragment is duplicated, we
+can map it with high confidence provided we can distinguish it from its
+duplicates. This in turn depends on their similarity with the target
+fragment and on the error rate of the sequencer.
+
+Because sequencing errors can convert a DNA fragment into one of its
+duplicates, the \emph{true} location of a DNA fragment is not always the
+\emph{best} (as measured by the identity between the sequence and the
+candidate location). This leads us to consider the following question: how
+can we identify the \emph{best} location of the fragment in the genome?
+
+This is the \emph{best location problem}. It ammounts to finding the
+optimal alignment between two sequences, and for this reason has received
+substantial attention in bioinformatics. It is relevant for the present
+discussion because the candidate solution of the \emph{true} location
+problem is always the \emph{best} location. The next section will further
+clarify the relationship between the two problems.
+
+\subsection{Seeding heuristics}
+
+There exists exact algorithms to solve the best location
+problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
+the large amount of data generated by modern sequencers. Instead, one uses
+heuristic methods, \textit{i.e.} algorithms that run fast, but may not
+return the best location\cite{Waterman1984}. Heuristic mapping algorithms
+thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
+best location, $ii.$ the DNA fragment is mapped to a suboptimal location,
+$iii.$ the DNA fragment is not mapped at all.
+
+The most popular heuristic to address the best location problem is to
+first identify short regions of high similiarity between the read and the
+genome, and then run an exact alignment algorithm around those regions.
+The first step is usually called ``seeding'' and the regions of high
+similarity are called ``seeds''. This is for instance the general strategy
+of the popular alignment algorithm BLAST\cite{pmid2231712}.
+
+If seeds can be discovered fast, we can quickly zoom in on a small set of
+candidate locations, and use slower exact algorithms on a reduced data
+set. The disadvantage is that the target location is not always in the
+candidate set. In this case, it cannot be discovered and the outcome is
+either a false positive or a false negative.
+
+This imposes a trade-off between speed and accuracy between each candidate
+location must be tested with a slow algorithm. If the seeding step is set
+to produce many candidates, the target is likely to be among them and it
+will be discovered. The downside is that this will take long. Conversely,
+if the seeding step is set to produce few candidates, the process will
+terminate faster but the target is less likely to sit among the candidates
+so it is less likely to be discovered.
+
+Importantly, the terms of this trade-off are not set in stone. Some
+seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
+one can develop faster mapping algorithms at no cost for accuracy.
+
+This framework was developed for the best location problem, but it also
+holds for the true location problem. Actually, the process is exactly the
+same but the outcomes $i.$ and $ii.$ are different. The first, called a
+``true positive'' or an ``on-target hit'', is that the DNA fragment is
+mapped to the \emph{true} location; the second, called a ``false
+positive'' or an ``off-target hit'' is that the DNA fragment is mapped
+somewhere else in the genome. A true positive thus requires that the
+target (the true location) is in the candidate set, and that it is the
+optimum.
+
+\subsection{MEM seeds}
+
+In general, seeds do not have to be exact matches\cite{pmid20460430}. For
+instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
+seeds'' that tolerate mismatches\footnote{The method is actually covered
+by a patent (reference US20070088510A1).}. However, in this document, we
+will always assume that a seed is an exact match with no gap. In other
+words, a seed is a perfect match between a part of the read and a genomic
+site.
+
+The rest of the document will focus on a seeding heuristic known as ``MEM
+seeding''. This approach gives good empirical results when mapping short
+reads (it is for instance the strategy\footnote{To be exact, BWA-MEM uses
+SMEM seeds (Super Maximal Exact Match seeds).} of the popular mapper
+BWA-MEM\cite{li2013aligning}) but there is presently no theory to justify
+this performance. A deeper understanding of MEM seeding would thus be
+useful to calibrate the speed/accuracy trade-off of this heurisitic.
+
+Before developing a formal theory, let us clarify the meaning of the MEM
+acronym.
 
 \begin{definition}
-A Maximal Exact Match (MEM) is a subsequence of a read that is present in
-the genome and that cannot be extended left or right --- either because
-the read ends or because the extended subsequence is not in the genome.
+A Maximal Exact Match (MEM) seed is a subsequence of a read that is
+present in the genome of interest and that cannot be extended --- either
+because the read ends or because the extended subsequence is not in the
+genome.
 \end{definition}
@@
 This definition presents a computational challenge: to know if a given
-subsequence is a MEM, we need to know if it exists somewhere in the
+subsequence is a MEM seed, we need to know if it exists somewhere in the
 genome. This is a non trivial problem in itself, but fortunately there
-exists practical methods to answer the question, even for very large
-genomes. Here on, we will assume that we always know which subsequences of
-the read are MEMs.
-
-A practical consideration is that short matches between the read and the
-genome are irrelevant. For instance, $99.7\%$ of all the possible 12-mers
-are present in the reference human genome, so an exact match of size 12 or
-lower says nothing about the location of the DNA fragment. In what
-follows, we will simply disregard short matches. More specifically, we
-will assume that subsequences of the read can only match the original DNA
-sequence or its duplicates, but not the rest of the genome. This will
-simplify the discussion without loss of generality, because short
-uninformative matches are ignored in all practical applications.
-
-Fig.~\ref{fig:MEM_example} shows a concrete example of the relationship
-between read errors, on-target matches, off-target matches and MEMs. The
-sequence of the original DNA fragment is the first genomic hit (this is
-the \emph{true} location). The second, third and fourth genomic hits are
-duplicates. The mismatches between the genome and the read are
-highlighted in bold. The read has three errors indicated by stars, so
-\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
-\texttt{A} and \texttt{G} (because the first hit is the \emph{true}
-sequence).
+exists practical methods to identify all the MEM seeds from a read, even
+for very large genomes\cite{ferragina2000opportunistic,
+ferragina2005indexing}. Such algorithms are crucial for the present
+discussion, but describing them falls outside the scope of the document.
+We refer the interested reader to the relevant
+literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
+li2013aligning}, and here on we will assume that MEM seeds are simply
+available at all times.
+
+Before going further, we need to move a practical consideration out of the
+way. It stands to reason that very short matches between the read and the
+genome are uninformative. For instance, $99.7\%$ of all the possible
+12-mers are present in the reference human genome, so an exact match of
+size 12 or lower says nothing about the location of the DNA fragment.
+In what follows, we will assume that subsequences of the read can only
+match the original DNA sequence and/or its duplicates, but not the rest of
+the genome. This will simplify the discussion without loss of generality
+because non homologous matches are short and they can be ignored.
+
+Fig.~\ref{fig:MEM_example} shows a concrete example. The target (the true
+location of the read) is first potential location. The locations below
+correspond to duplicate sequences. The mismatches between the genome and
+the read are highlighted in bold. The read has three errors indicated by
+stars, so \texttt{T}, \texttt{T} and \texttt{C} should actually be
+\texttt{A}, \texttt{A} and \texttt{G}.
 
@@
 \includegraphics[scale=1]{MEM_example.pdf}
-\caption{\textbf{Errors, matches and MEMs.}
-Some legend.}
+\caption{\textbf{A concrete example of MEM seeding.}
+A DNA fragment is sequenced with an imperfect instrument. It comes from
+the top genomic location (Truth), but three errors were introduced
+(indicated by stars). The sequence of the DNA fragment has three
+duplicates at other genomic locations, shown below the target. Mismatches
+between the read and the candidate sequence are highlighted in bold. MEM
+seeds are indicated by grey boxes.}
 \label{fig:MEM_example}
@@
 
-The MEMs are indicated by grey boxes. Technically, MEMs are subsequences
-of the read, but they are shown on the genomic hits to highlight where
-they match. Two MEMs can overlap, in which case they must match distinct
-sequences of the genome (this follows from the definition). The last MEM
-corresponds to two matches of length 4, with the first and the second
-genomic hits. Such MEMs with several matches are referred to as
-\emph{shared}. MEMs with a single match in the genome are referred to as
-\emph{strict}.
-
-As the name suggests, the principle of MEM seeding is to use MEM as seeds
-to identify candidate genomic locations. Because short matches are
-irrelevant, MEM seeds are usually requested to have a length greater than
-or above a certain threshold denoted $\gamma$. We will assume that a
-single MEM seed is sufficient to consider that a genomic hit is a
-candidate location, which is why we will focus on the case that the read
-does not have a single on-target MEM seed.
-
-%In the example of Fig.~\ref{fig:MEM_example}, there is no MEM that matches
-%the third genomic hit. It is important to understand that this can also
-%happen to the best hit, as in the example shown in
-%Fig~\ref{fig:MEM_eclipse}. The undesirable consequence is that some reads
-%are bound to yield false positives, even if we tune the MEM seeding
-%heuristic by changing the minimum seed length $\gamma$. We will refer to
-%the case that the best hit is masked by off-target MEMs as \emph{MEM
-%eclipse}. 
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[scale=1]{MEM_eclipse.pdf}
-%\caption{\textbf{MEM eclipse.}
-%Some legend.}
-%\label{fig:MEM_eclipse}
-%\end{figure}
-%
-%If you are not already familiar with MEM seeding, we recommend that you
-%take a moment to study the examples of Fig.~\ref{fig:MEM_example} and
-%\ref{fig:MEM_eclipse}. This will be essential to understand the next
-%sections. Before reading further, make sure that you can answer the
-%following questions:
-%\begin{enumerate}
-%\itemsep0em
-%\item What are the conditions for a MEM to be shared?
-%\item What is the minimum number of duplicates for MEM eclipse?
-%\end{enumerate}
+The MEM seeds are indicated by grey boxes. Technically, they are part of
+the read, but they are shown on the genome to highlight the position of
+the match. One MEM seed cannot cover a smaller one (otherwise the smaller
+would not be a MEM seed), but two consecutive MEM seeds can overlap, as in
+the case of the frst two from the left. Note that overlapping MEM seeds
+must match distinct sequences of the genome (otherwise neither of them
+would be a MEM seed). Consecutive MEM seeds can also be disjoint, as in
+the case of the second and the third from the left. In this case, they
+must be separated by a nucleotide that is a mismatch against \emph{all}
+the duplicates.
+
+Observe that the rightmost MEM seed matches two genomic locations. This
+case motivates the following definition, which will play an important role
+for the development of the theory.
+
+\begin{definition}
+A \emph{strict} MEM seed has a single genomic match.
+A \emph{shared} MEM seed has several genomic matches.
+\end{definition}
+
+Without surprise, the principle of MEM seeding is to use the genomic
+matches of MEM seeds as candidate locations. Importantly, MEM seeds
+shorter than a certain threshold $\gamma$ are ignored (in line with the
+observation that short matches are uninformative). The parameter $\gamma$
+is the tuning knob on the speed/accuracy trade-off. In
+Fig.~\ref{fig:MEM_example} for instance, if $\gamma = 7$, there are three
+candidate locations (the first, the second and the fourth) and if $\gamma
+= 8$ there are only two (the first and the fourth).
+
+Not every MEM seed makes the target a candidate location, so we introduce
+the following terms to simplify the discussion.
+
+\begin{definition}
+Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
+on-target MEM seed of size $\gamma$ or greater. Every other MEM seed is a
+\emph{bad} seed. In other words, the true location is in the candidate set
+if and only if the read has a good seed.
+\end{definition}
+
+Compared to seeds of fixed size, MEM seeds have at least two
+counter-intuitive properties. The first is that there are cases where
+there is no good seed, even when $\gamma$ is decreased.
+Fig.~\ref{fig:full_masking_example} shows such an example. Even though
+there is a single sequencing error, the read does not have any MEM seed
+matching the target. Lowering $\gamma$ does not change this, so there is
+no way to discover the true location (even though it is also the best
+location).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{full_masking_example.pdf}
+\caption{\textbf{MEM masking case 1.}
+The read, the MEM seeds and the sequences are as in
+Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
+locations at the bottom effectively mask the target, so it cannot
+be discovered. MEM masking can occur even when the true location is the
+best and when there is a single error on the read.}
+\label{fig:full_masking_example}
+\end{figure}
+
+The second counter-intuitive property of MEM seeding is that there are
+cases where there would be a good seed if the read were shorter.
+Fig.~\ref{fig:short_vs_long} shows an example of this case. Again, there
+is no on-target MEM seed, but there would be if the read were two
+nucleotides shorter on the right side. Indeed, in this case there would be
+a shared MEM seed matching the first and the second locations. It would
+also be a good seed for $\gamma \leq 12$.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{short_vs_long_example.pdf}
+\caption{\textbf{MEM masking case 2.}
+The read, the MEM seeds and the sequences are as in
+Fig.~\ref{fig:MEM_example}. There is no on-target MEM seed, but there
+would be a shared MEM seed if the read were two nucleotides shorter. The
+true location has been masked by sequencing the last two nucleotides.}
+\label{fig:short_vs_long}
+\end{figure}
+
+These examples show that MEM seeds can perform worse than seeds of fixed
+size, again raising the question why they seem to give good performance in
+practice. The theory developed below will allow us to compute the
+probability that a read has no good seed and thus that the true location
+is missed. This probability is the key to evaluating and calibrating the
+MEM seeding heuristic.
+
+Before going further, we recommend the readers that are not already
+familiar with MEM seeding to take a moment to study the examples above.
+The complexity of the next sections will go \textit{crescendo} and they
+will draw heavily from the concepts and definitions above.
 
@@
 
-In this section we give a representation of reads as combinatorial ojects.
-The construction has two essential purposes: the first is to give a
-generative model of reads without any on-target MEM seed, the second is to
-allow us to find their weighted generating function, which in turn will
-allow us to find their probability of occurrence.
+In this section we represent reads as combinatorial ojects. The
+construction has two essential purposes: the first is to give a generative
+model of reads that do not have any on-target MEM seed; the second is to
+allow us to find the weighted generating function of those reads, which in
+turn will allow us to find their probability of occurrence.
 
@@
 
-The nucleotide sequence of the read is not so useful to know where MEMs
-are and whether they are on-target. To develop a formal method to compute
-the probability that the read contains an on-target MEM seed, we will
-recode reads as sequences of letters from the so-called \emph{MEM
-alphabet} $\{\downarrow, \uparrow, \square\}$.
-
-Each nucleotide of the read is recoded with one of three letters. The
-$\downarrow$ symbol indicates that the nucleotide is a sequencing error.
-By definition, it is a mismatch for the target. The $\uparrow$ symbol
-characterizes the presence of an on-target MEM in a way that we are about
-to clarify. All the other nucleotides are represented by the $\square$
-symbol.
-
-To understand the meaning of the $\uparrow$ symbol, consider the following
-situation: at the left end of the read, as we try to extend the first
-match as much as possible to the right, the number of hits in the genome
-keeps decreasing. It may be the case that at some point, we are left with
-a single hit that is on-target. The very first nucleotid where that
-happens is called the \emph{pivot} and is labelled with the symbol
-$\uparrow$.
+The nucleotide sequence of the read is inadequate to know where MEM seeds
+are and whether they are on target. For this purpose, we will recode reads
+as sequences of letters from the four letter \emph{MEM alphabet}
+$\{\downarrow, \uparrow, |, \square\}$.
+
+Each nucleotide is recoded with $\downarrow$, $\uparrow$, or $\square$.
+The $\downarrow$ symbol indicates that the nucleotide is a sequencing
+error. By definition, it is a mismatch for the target. The $\uparrow$
+symbol indicates that the nucleotide is the ``first'' nucleotide of an
+on-target MEM seed, which we are about to clarify. All the other
+nucleotides are represented by the $\square$ symbol. Finally, the $|$
+symbol is appended after the last nucleotide of the read. It is an
+obligatory terminator that does not correspond to an actual nucleotide.
+
+To explain the precise meaning of the $\uparrow$ symbol, let us consider
+the left end of the read: as we try to extend the first match as much as
+possible to the right, the number of hits in the genome decreases. It may
+be the case that at some point, there is a single hit and that it is on
+target. The very first nucleotid where that happens is called the
+\emph{pivot} and is labelled with the symbol $\uparrow$. An on-target MEM
+seed has exactly one pivot, so the number of $\uparrow$ symbols is the
+number of on-target MEM seeds in the read. Note that $\uparrow$ symbols
+marks all on-target MEM seeds, regardless of their size.
 
@@
 encoded with the MEM alphabet. Below the read are the matches between the
-read and the genomic hit, as they are in Fig.~\ref{fig:MEM_example}. A
-black square means that the nucleotide is a mismatch for the genomic
-sequence, and a white or grey square means that it is a match. Grey
-squares are used to highlight MEMs, as in Fig.~\ref{fig:MEM_example}.
+read and the potential locations from Fig.~\ref{fig:MEM_example}. A black
+square means that the nucleotide is a mismatch for the genomic sequence,
+and a white or grey square means that it is a match. Grey squares are used
+to highlight MEM seeds, as in Fig.~\ref{fig:MEM_example}.
 
@@
 \includegraphics[scale=.85]{sketch_MEM.pdf}
-\caption{\textbf{Reads and MEM seeds}. 
-Some legend.}
+\caption{\textbf{The MEM encoding}.
+The read of Fig.~\ref{fig:MEM_example} is recoded with the MEM alphabet.
+The mismatches between the read and the potential locations are indicated
+with black boxes and the matches with white boxes. The grey boxes are
+matches and they indicate the locations of the MEM seeds. The $\downarrow$
+symbols coincide exactly with the mismatches between the read and the
+target. The $\uparrow$ symbol is always a match for the target and a
+mismatch for some off-target location. It indicates the presence of a
+strict on-target MEM seed (flanked by the nearest $\downarrow$ symbols or
+by the end of the read). Also note the presence of the $|$ terminator at
+the end of the read.}
 \label{fig:sketch_MEM}
@@
 
-Observe that the two on-target MEMs contain a pivot, but the shared MEM
-does not. This is always the case. Recall that the pivot indicates that
-the true location is the \emph{only} local match, so the $\uparrow$ symbol
-cannot appear in on-target MEM seeds when they are shared.
-
-Strict on-target MEM seeds are easy to find when reads are represented in
-the MEM alphabet: they are the longest stretches of symbols containing the
-$\uparrow$ symbol and not containing the $\downarrow$ symbol. Shared
-on-target MEM seeds are trickier. They do not directly appear in this
-representation so we will have to keep that in mind throughout the theory.
-It is in a sense justified to put more emphasis on strict MEM seeds
-because they guarantee that the true location will be found. Shared MEM
-seeds can have thousands of hits in the genome in practice, so the user
-may not test them all. That said, the theory developed here fully takes
-shared MEM seeds into account, even if they are not visible in the
-representation of the MEM alphabet.
-
-The point of the MEM seeding alphabet is that reads can be seen as
-sequences of letters, but also as sequences of \emph{segments} delimited
-by the $\uparrow$ and the $\downarrow$ symbols. Because such segments are
-intimately linked to the presence of on-target MEM seeds, we can give a
-construction of reads that do not contain any.
+Observe that the two on-target MEM seeds contain a pivot, but the shared
+MEM seed does not. This is always the case. Recall that the pivot
+indicates that the true location is the \emph{only} local match, so the
+$\uparrow$ symbol cannot appear in on-target MEM seeds when they are
+shared.
+
+The MEM alphabet makes it easy to find strict on-target MEM seeds: they
+are the longest stretches of symbols containing $\uparrow$ and not
+containing $\downarrow$. However, shared on-target MEM seeds do not appear
+in this representation.
+
+\subsection{The extended MEM alphabet}
+
+To make shared on-target MEM seeds apparent, we decorate the $\uparrow$
+and $\downarrow$ symbols, effectively extending the alphabet. The
+$\downarrow$ symbol is replaced by the set of symbols $\downarrow_{/m}$,
+where $m$ is the number of duplicates that \emph{match} the erroneous
+nucleotide. The $\uparrow$ symbol is replaced by the set of symbols
+$\uparrow_{/j}$, where $j-1$ is the number of consecutive $\square$
+symbols preceding $\uparrow$.
+
+Note that the numbers decorating the symbols have a very different nature.
+In the case of the $\downarrow$ symbol, it is an information about the
+number of sequences matching the read; in the case of the $\uparrow$
+symbol, it is the position of the pivot relative to the previous
+sequencing error (or relative to the beginning of the read).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.85]{sketch_extended.pdf}
+\caption{\textbf{The extended MEM encoding}.
+The read of Fig.~\ref{fig:sketch_MEM} is represented in the extended MEM
+alphabet. The symbols $\uparrow$ and $\downarrow$ are now decorated by
+numbers. The arrows departing from those numbers help understand their
+meaning. The $\downarrow$ symbol is decorated by the number of locations
+that match the nucleotide. The $\uparrow$ symbol is decorated by the
+number of consecutive $\square$ symbols before it, plus one.}
+\label{fig:sketch_extended}
+\end{figure}
+
+Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
+Fig.~\ref{fig:sketch_MEM} in the extended MEM alphabet. The arrows bring
+the focus on how the numbers decorating the $\uparrow$ and $\downarrow$
+symbols are computed. For instance, the $\downarrow_{/0}$ symbol indicates
+that the nucleotide is an error (this is the meaning of the $\downarrow$
+symbol), and that it matches no other location (this is meaning of $/0)$.
+In other words, it is a mismatch against every potential location.
+
+In this new encoding, shared on-target MEM seeds are stretches of
+$\square$ symbols flanked either by $\downarrow_{/0}$, or by $|$, or by
+the left end of the read. Indeed, such a stretch is a MEM seed because it
+matches at least one sequence (the target) and it cannot be extended
+(because the end of the read is reached or because there is no match in
+the genome). Also, it cannot be a strict on-target MEM seed because it
+does not contain the $\uparrow$ symbol, so it must be a shared on-target
+MEM seed.
+
+The reason for introducing more than just the $\downarrow_{/0}$ symbol is
+not immediately clear. For the combinatorial construction to be useful, we
+need not only to identify on-target MEM seeds, but also to compute the
+probabilities of the reads of each type. This is precisely the role of the
+extended alphabet, as we will see in section~\ref{sec:transfer_mat}.
+
+\subsection{Reads as extended MEM segments}
+
+The final twist in the combinatorial construction is to consider reads not
+as sequences of symbols, but as sequences of \emph{segments}. The purpose
+here is merely to express that some on-target MEM seeds are good
+(\textit{i.e.}, their size is $\gamma$ or greater) and some are not.
 
@@
 \label{def:segment}
-A segment is a sequence of $\square$ symbols terminated by either a
-$\uparrow$ symbol or by a $\downarrow$ symbol. The unterminated sequence
-of $\square$ symbols at the end of the read is called the tail --- which
-can thus be an empty sequence of symbols.
+A segment is a maximal sequence of $\square$ symbols flanked on the right
+side by one of the symbols $\uparrow_{/j}$, $\downarrow_{/m}$, or $|$. The
+one and only segment flanked by $|$ is called the tail. It is the only
+segment that can have size 0.
 \end{definition}
 
-Following definition~\ref{def:segment}, the $\uparrow$ and $\downarrow$
-symbols will be referrred to as \emph{terminators}.
-Fig.~\ref{fig:sketch_segment} shows the decomposition of the read from
-Fig.~\ref{fig:sketch_MEM} in segments. The size of each segment (the
-number of nucleotides that compose it) is indicated below.
+In line with definition~\ref{def:segment}, the symbols $\uparrow_{/j}$,
+$\downarrow_{/m}$ and $|$ are referrred to as \emph{terminators}.
+Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
+from Fig.~\ref{fig:sketch_extended}. The numbers below indicate the size
+of each segment.
 
@@
 \includegraphics[scale=.85]{sketch_segments.pdf}
-\caption{\textbf{Reads as sequences of segments}. 
-Some legend.}
+\caption{\textbf{The extended MEM segment encoding}.
+The read from Fig.~\ref{fig:sketch_segment} is represented as a sequence
+of segments. Each segment is terminated by either $\uparrow_{/j}$, or
+$\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
+below indicate the size of each segment in nucleotides.}
 \label{fig:sketch_segment}
@@
 
-\subsection{The extended MEM alphabet}
-
-An issue with the MEM alphabet is that the probabilities of the symbols
-are difficult to evaluate. The $\downarrow$ symbol coincides with read
-errors, which is straightforward to model. The $\uparrow$ symbol coincides
-with pivots, which have a more complex behavior. We thus extend the MEM
-alphabet so that the occurrence of the $\uparrow$ becomes easier to
-evaluate.
-
-Only the terminators are extended; the $\square$ symbols remain unchanged.
-Let us denote the number of duplicates of the target by $N$. The
-$\downarrow$ symbol is replaced by the $N+1$ symbols
-$\downarrow_{/m}$, where $m$ is the number of duplicates that
-\emph{match} the erroneous nucleotide $(0 \leq m \leq N)$. The $\uparrow$
-symbol is replaced by the symbols $\uparrow_{/j}$, where $i
-\geq 0$ is the size of the segment (including the
-$\uparrow_{/j}$ symbol). Finally, we add the special terminator
-$|$ that marks the end of the read. In other words, tails are considered
-normal segments terminated by the symbol $|$ (with size 0).
-
-It is important to highlight that the numbers that decorate the symbols
-have a very different nature. In the case of the $\downarrow$ symbol, it
-is an information about the number of sequences matching the read; in the
-case of the $\uparrow$ symbol, it is the position of the pivot relative to
-the previous terminator (or to the beginning of the read).
-
-Fig.~\ref{fig:sketch_extended} shows the decomposition of the read from
-Fig.~\ref{fig:sketch_segment} in segments using the extended MEM alphabet.
-The numbers decorating the symbols are shown above them to avoid crowding
-the figure.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.85]{sketch_extended.pdf}
-\caption{\textbf{Reads as sequences of segments in the extended MEM
-alphabet}. 
-Some legend.}
-\label{fig:sketch_extended}
-\end{figure}
-
-There is an infinite amount of $\uparrow_{/j}$ symbols in the
-extended MEM alphabet, but if the read has no on-target MEM seed, then
-only the $\uparrow_{/j}$ symbols where $1 \leq i \leq \gamma-1$
-can be used (if a pivot terminates a segment of size greater than or equal
-to $\gamma$, then the read contains a strict on-target MEM seed). The
-extended MEM alphabet is made in such a way that the numbers decorating
-the $\uparrow$ symbol will allow us to exclude strict on-target MEM seeds,
-and the numbers decorating the $\downarrow$ symbol will help us exclude
-shared on-target MEM seeds.
+Shared on-target MEM seeds at the tail of the read are segments but all
+other on-target MEM seeds are not. Shared MEM seeds cannot contain the
+terminator $\downarrow_{/0}$ because it is a mismatch against the target.
+Strict on-target MEM seeds are split over at least two segments: one
+terminated by $\uparrow_{/j}$ and one right of it terminated by either
+$\downarrow_{/m}$ or $|$.
 
@@
 
+We now turn to the question of translating the extended MEM segment
+representation to a construction of reads without good seeds
+(\textit{i.e.}, without on-target MEM seeds of size $\gamma$ or greater).
+For this, we use tools borrowed from analytic combinatorics. We give the
+minimum amount of background that is necessary to expose the theory. A
+more detailed account of the strategy to compute seeding probabilities
+with analytic combinatorics has been published
+elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
+readers interested in learning more about analytic combinatorics to the
+textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.
+
 \subsection{Weighted generating functions}
 
-Let $\mathcal{A}$ be a set of combinatorial objects which have a
-\emph{probability of occurrence}, and a \emph{size} that is an integer
-number. The function defined as $A(z) = \sum_{k=0}^\infty a_kz^k$, where
-$a_k$ is the total probability of objects of size $k$, is called the
-\emph{weighted generating function} of the combinatorial objects.
+Let $\mathcal{A}$ be a set of combinatorial objects with a size $k \in
+\mathbb{N}$ and a probability of occurrence. Denoting $a_k$ the total
+probability of objects of size $k$, the function defined as $A(z) =
+\sum_{k\geq0} a_kz^k$ is called the \emph{weighted generating
+function} of the combinatorial objects.
 
 In our case, we are interested in the weighted generating function of
-reads without on-target MEM seed, where the coefficient $a_k$ is the
-probability that a read of size $k$ has no on-target MEM seed. As $k$
-becomes larger, it becomes more difficult to evaluate $a_k$, but the
-combinatorial construction developed above will allow us to do this by
-using established strategies involving weighted generating functions.
+reads without good seeds, so the coefficient $a_k$ is the probability
+that a read of size $k$ has no good seed. As $k$ becomes larger,
+it becomes more difficult to evaluate $a_k$, but the representation
+developed above will allow us to do this.
 
@@
 
-The combinatorial construction of reads using the extended MEM seeding
-alphabet gives us a method to express their weighted generating function
-from that of segments. More specifically, we can describe complex
-sequences of segments through a so-called \emph{transfer matrix} $M(z)$
-that specifies which segments can follow one another in the reads of
-interest.
-
-The segments types are identified by their terminators and arranged in a
+The combinatorial construction of reads using the extended MEM alphabet
+gives us a method to express their weighted generating function from the
+simpler generating function of segments. More specifically, we can
+describe complex sequences of segments through a so-called \emph{transfer
+matrix} $M(z)$ that specifies which pairs of consecutive segments are
+allowed and which are forbidden in the reads of interest.
+
+The segment types are identified by their terminators arranged in a
 predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
@@
 can be inserted immediately after segments with the $i$-th terminator. The
-matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$ contains the
-weighted generating functions of the reads of all sizes. The entry that
-corresponds to terminators $\downarrow_{/0}$ and $|$ is the weighted
-generating function of the reads of interest (the beginning of the read is
-equivalent to a ghost segment terminated by $\downarrow_{/0}$,
-\textit{i.e.} a mismatch for all the sequences).
-
-With the transfer matrix $M(z)$, with thus have a way of computing the
-weighted generating function of the reads without on-target MEM seed, and
-from there we can extract teir probability of occurrence. Our next task is
-thus to specify $M(z)$.
+entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
+are the weighted generating functions of the reads of all sizes using all
+the allowed combinations of segments.
+
+The entry of interest from $M(z)$ is the one that corresponds to the
+terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
+any size whose first segment can be inserted after $\downarrow_{/0}$, and
+whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
+the end of the read, so it must be the final symbol of every valid read.
+As for the first segment, observe that all the segments that can follow
+the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
+postion of the read\footnote{One could have introduced a special start
+symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
+the reads.}.
+
+In summary, the transfer matrix $M(z)$ contains the weighted generating
+function of interest. Our next task is thus to specify $M(z)$.
 
 \section{The transfer matrix}
-
-\subsection{Match streak and masks}
-
-For each sequence, we define the \emph{match streak} at a given nucleotide
-as the number of nucleotides since the last mismatch on the left. For
-instance, in Fig.~\ref{fig:MEM_example}, the match streak of the first
-genomic hit at the ninth nucleotide is $1$ while for the second genomic
-hit, it is $7$.
+\label{sec:transfer_mat}
+
+\subsection{Hard and soft masking}
+
+For each potential location including the target, we define the
+\emph{match streak} at a given nucleotide as the number of nucleotides
+since the last mismatch on the left. For instance, in
+Fig.~\ref{fig:MEM_example}, the match streak of the target at the ninth
+nucleotide is $1$ while for the second potential location, it is $7$.
 
 \begin{definition}
-At a given position of the read, a duplicate sequence is a \emph{hard
+At a given position of the read, an off-target location is a \emph{hard
 mask} if its success streak is strictly longer than the match streak of
-the target. A duplicate thread is a \emph{soft mask} if its has the same
-match streak as the target.
+the target. An off-target location is a \emph{soft mask} if its has the
+same match streak as the target.
 \end{definition}
 
-For instance, the second genomic hit in the example above is a hard mask
-at the ninth nucleotide. The point of hard and soft masks is that they are
-directly linked with MEMs. A nucleotide with a hard mask cannot belong to
-an on-target MEM. A nucleotide without mask belongs to a strict on-target
-MEM. A nucleotide with a soft mask may belong to a strict on-target MEM,
-to a shared on-target MEM, or may not belong to any MEM. The distinction
-between the three cases will be clarified in the specification of the
-transfer matrix.
+For instance, the second location in the example above is a hard mask at
+the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further check
+that all the off-target locations are soft masks at the second nucleotide
+and that the second location is a soft mask at the last nucleotide of the
+read.
+
+A nucleotide with a hard mask cannot be part of an on-target MEM seed. A
+nucleotide without mask belongs to a strict on-target MEM seed. A
+nucleotide with a soft mask may belong to a strict on-target MEM seed, to
+a shared on-target MEM seed, or may not belong to any MEM. This case is
+the most intricate and it will be teased apart below.
 
 The extended MEM alphabet is nothing more than a way to keep track of the
-hard and soft masks. For instance, a pivot (associated with the symbol
-$\uparrow$) is a nucleotide where the number of masks becomes is reduced
-to 0. If there are $N$ duplicates, the symbol $\downarrow_{/m}$ indicates
-that there $m$ hard masks and $N-m$ soft masks.
+hard and soft masks. For instance, a pivot (associated with the symbols
+$\uparrow_{/j}$) is a nucleotide where the number of masks becomes 0.
+Denoting the number of off-target locations by $N$, the symbol
+$\downarrow_{/m}$ indicates that there are $m$ hard masks and $N-m$ soft
+masks.
 
@@
 
-The method is summarized by the algorithm below.
+The method is summarized by the algorithm below (FIX THE ALGORITHM).
 
@@
 
+\section{Using the information of the backward search}
+
+Assume that the sequence to map has $N$ additional duplicates, all with a
+uniform divergence equal to $\mu$. We can use the number of hits
+obtained in the backard search to estimate $N$ and $\mu$.
+
+At every stage of the backward search, we obtain the number of sequences
+in the reference genome that match the end of the query. Ignoring
+sequencing errors and focusing only on the $N$ duplicates, the probability
+that the count of duplicates goes from $N_i$ to $N_{i+1}$ at each step of
+the search is given by
+
+\begin{equation*}
+P(N_{i+1}|N_i) = {N_i \choose N_{i+1}}
+\mu^{N_i-N_{i+1}}(1-\mu)^{N_{i+1}}.
+\end{equation*}
+
+The probability of the sequence $(N_1, N_2, \ldots,
+N_m)$ starting from $N_0$ is
+
+\begin{equation*}
+P(N_1, \ldots, N_m|N_0) = 
+{N_0 \choose N_1 } \ldots {N_{m-1} \choose N_m }
+\mu^{N_0-N_m}(1-\mu)^{N_1+\ldots+N_m}.
+\end{equation*}
+
+We can now use the maximum likelihood approach to find the value of
+$\mu$ that maximizes this expression. Taking the logarithm and
+differentiating with respect to $\mu$ we find
+
+\begin{gather*}
+\frac{N_0-N_m}{\mu}-\frac{N_1+\ldots+N_m}{1-\mu} = 0
+\text{, which implies} \\
+\mu = \frac{N_0-N_m}{N_0+\ldots+N_{m-1}}.
+\end{gather*}
+
+We also observe that by defining the loss at each step $\Delta_i = N_{i-1}
+- N_i$, the distribution of $(\Delta_1, \ldots, \Delta_m, N_m)$ given
+$N_0$ is multinomial with parameters $\big(\mu, \mu(1-\mu),
+\ldots, \mu(1-\mu)^{m-1}, (1-\mu)^m\big)$. More specifically, the
+probability of observing the sequence $(\Delta_1, \ldots, \Delta_m, N_m)$
+given $N_0$ is
+
+\begin{align*}
+{N_0 \choose \Delta_1, \ldots, \Delta_m, N_m}
+\mu^{\Delta_1}\big(\mu(1-\mu)\big)^{\Delta_2}
+\ldots \big(\mu(1-\mu)^{m-1}\big)^{\Delta_m}(1-\mu)^{mN_m}.
+\end{align*}
+
+Now, if $N_0, N_1, \ldots, N_{n-1}$ are missing, we can compute the
+probability of observing $(\Delta_{n+1}, \ldots, \Delta_m, N_m)$ from the
+marginal distribution as
+
+\begin{align*}
+\begin{split}
+{N_0 \choose N_0-N_n, \Delta_{n+1}, \ldots, \Delta_m, N_m}
+\big(1-(1-\mu)^n\big)^{N_0-N_n}
+\big(\mu(1-\mu)^n\big)^{\Delta_{n+1}} \times
+\ldots \\
+\ldots \times
+\big(\mu(1-\mu)^{m-1}\big)^{\Delta_m}(1-\mu)^{mN_m}.
+\end{split}
+\end{align*}
+
+This expression depends on $N_0$, which can be considered a parameter of
+the distribution. To estimate it, we take the logarithm of the probability
+and differentiate with respect to $N_0$. We obtain the two following
+equations
+
+\begin{gather*}
+\Psi(N_0+1)-\Psi(N_0-N_n+1) + \log\big( 1-(1-\mu)^n \big) = 0 \\
+\frac{N_n-N_m}{\mu}
+-\frac{nN_n + N_{n+1}+\ldots+N_m}{1-\mu} +
+(N_0-N_n)\frac{n(1-\mu)^{n-1}}{1-(1-\mu)^n} = 0.
+\end{gather*}
+
+In the expression above, $\Psi(\cdot)$ is the digamma function, defined as
+the derivative of $\log \Gamma(\cdot)$. We observe that the second
+equation gives an expression of $N_0$ as a function of $\mu$, namely
+
+\begin{equation*}
+N_0 = N_n +
+\left(\frac{nN_n+N_{n+1}+\ldots+N_m}{1-\mu}-
+\frac{N_n-N_m}{\mu} \right)
+\frac{1-(1-\mu)^n}{n(1-\mu)^{n-1}}.
+\end{equation*}
+
+We can substitute this expression in the first equation, obtaining an
+equation depending on $\mu$ only, which can be solved numerically by
+the Newton-Raphson method.
+
 \section{Conclusion}

### 2019-02-07 MEM.tex
Added the part on false positives.

@@
 
-\title{Calibrating MEM seeding heuristics}
+\title{Computing MEM seeding probabilities}
 
@@
 
-The focus of this document is solve the true location problem \textit{per
-se}, but rather to evaluate whether a solution is correct. At first sight,
-whether we can map reads to their true location seems to depend only on
-their lengths and on the error rate of the instrument. However, there is
-another important factor to consider: most genomes contain repeated
-sequences, \textit{i.e.} relatively large subsequences (from $\sim20$~bp
-to more than 10~kb) that are present at more than one location. So we
-cannot map the fragment with certainty if there exists an exact copy of
-the sequence at somewhere else in the genome, because we cannot know which
-of the two copies was sequenced.
+The focus of this document is not to solve the true location problem
+\textit{per se}, but rather to evaluate whether a solution is correct. At
+first sight, whether we can map reads to their true location seems to
+depend only on their lengths and on the error rate of the instrument.
+However, there is another important factor to consider: most genomes
+contain repeated sequences, \textit{i.e.} relatively large subsequences
+(from $\sim20$~bp to more than 10~kb) that are present at more than one
+location. So we cannot map the fragment with certainty if there exists an
+exact copy of the sequence at somewhere else in the genome, because we
+cannot know which of the two copies was sequenced.
 
@@
 \subsection{Seeding heuristics}
+\label{sec:seedheur}
 
@@
 li2013aligning}, and here on we will assume that MEM seeds are simply
-available at all times.
+available at all times. Looking ahead to section~\ref{sec:est}, let
+us just say that this can be done with an algorithm called the
+\emph{backward search}\cite{ferragina2000opportunistic}.
 
@@
 
+From now we will make the simplifying assumption that a read with a good
+seed is always mapped at the correct location. In other words, we assume
+that a good seed is necessary and sufficient for a true positive or
+outcome $i.$ In practice this is not always the case, but this assumption
+shifts the burden of mapping to the alignment algorithm and allows us
+focus on improving the seeding heuristic.
+
+We first explain a detailed strategy to compute the probablity of outcome
+$i.$ We then explain how to compute the probability of outcome $iii.$,
+\textit{i.e.} that the seeding step returns no hit. Combining these two
+quantities allows us to compute the probability of interest, which is that
+the read is mapped to an incorrect location or outcome $ii.$ Finally, we
+provide a computationally efficient method for estimating the parameters
+that are necessary to estimate the probabilities of interest.
+
 Before going further, we recommend the readers that are not already
@@
 
-\section{Combinatorial construction of reads}
-
-In this section we represent reads as combinatorial ojects. The
-construction has two essential purposes: the first is to give a generative
-model of reads that do not have any on-target MEM seed; the second is to
-allow us to find the weighted generating function of those reads, which in
-turn will allow us to find their probability of occurrence.
+\section{Combinatorial construction}
+
+In this section we represent reads as combinatorial ojects. The purpose of
+this construction is to give a generative model of reads without good
+seeds so that we can later compute their probability of occurrence.
+
+\subsection{Hard and soft masking}
+
+For each potential location including the target, we define the
+\emph{match streak} at a given nucleotide as the number of nucleotides
+since the last mismatch on the left, or since the beginning of the read.
+For instance, in Fig.~\ref{fig:MEM_example}, the match streak of the
+target at the ninth nucleotide is $1$ while for the second potential
+location, it is $7$.
+
+\begin{definition}
+At a given position of the read, an off-target location is a \emph{hard
+mask} if its success streak is strictly longer than the match streak of
+the target. An off-target location is a \emph{soft mask} if its has the
+same match streak as the target.
+\end{definition}
+
+For instance, the second location in the example above is a hard mask at
+the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further see
+that at the second nucleotide, all the off-target locations are soft masks
+and that at the last nucleotide, the second location is a soft mask.
+
+From the definition, the last nucleotide of every strict on-target MEM
+seed is always unmasked. Conversely, an unmasked nucleotide always belongs
+to exactly one strict on-target MEM seed. Also, the last nucleotide of
+every shared on-target MEM seed is always soft-masked, but a soft-masked
+nucleotide does not always belong to a shared on-target MEM seed.
+
+Since hard and soft masks inform us about the positions of on-target
+MEM seeds, we construct an alphabet that encodes the masking status of the
+nucleotides.
 
@@
 
-The nucleotide sequence of the read is inadequate to know where MEM seeds
-are and whether they are on target. For this purpose, we will recode reads
-as sequences of letters from the four letter \emph{MEM alphabet}
-$\{\downarrow, \uparrow, |, \square\}$.
-
-Each nucleotide is recoded with $\downarrow$, $\uparrow$, or $\square$.
-The $\downarrow$ symbol indicates that the nucleotide is a sequencing
-error. By definition, it is a mismatch for the target. The $\uparrow$
-symbol indicates that the nucleotide is the ``first'' nucleotide of an
-on-target MEM seed, which we are about to clarify. All the other
-nucleotides are represented by the $\square$ symbol. Finally, the $|$
-symbol is appended after the last nucleotide of the read. It is an
-obligatory terminator that does not correspond to an actual nucleotide.
-
-To explain the precise meaning of the $\uparrow$ symbol, let us consider
-the left end of the read: as we try to extend the first match as much as
-possible to the right, the number of hits in the genome decreases. It may
-be the case that at some point, there is a single hit and that it is on
-target. The very first nucleotid where that happens is called the
-\emph{pivot} and is labelled with the symbol $\uparrow$. An on-target MEM
-seed has exactly one pivot, so the number of $\uparrow$ symbols is the
-number of on-target MEM seeds in the read. Note that $\uparrow$ symbols
-marks all on-target MEM seeds, regardless of their size.
-
-Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
-encoded with the MEM alphabet. Below the read are the matches between the
-read and the potential locations from Fig.~\ref{fig:MEM_example}. A black
-square means that the nucleotide is a mismatch for the genomic sequence,
-and a white or grey square means that it is a match. Grey squares are used
-to highlight MEM seeds, as in Fig.~\ref{fig:MEM_example}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.85]{sketch_MEM.pdf}
-\caption{\textbf{The MEM encoding}.
-The read of Fig.~\ref{fig:MEM_example} is recoded with the MEM alphabet.
-The mismatches between the read and the potential locations are indicated
-with black boxes and the matches with white boxes. The grey boxes are
-matches and they indicate the locations of the MEM seeds. The $\downarrow$
-symbols coincide exactly with the mismatches between the read and the
-target. The $\uparrow$ symbol is always a match for the target and a
-mismatch for some off-target location. It indicates the presence of a
-strict on-target MEM seed (flanked by the nearest $\downarrow$ symbols or
-by the end of the read). Also note the presence of the $|$ terminator at
-the end of the read.}
-\label{fig:sketch_MEM}
-\end{figure}
-
-Observe that the two on-target MEM seeds contain a pivot, but the shared
-MEM seed does not. This is always the case. Recall that the pivot
-indicates that the true location is the \emph{only} local match, so the
-$\uparrow$ symbol cannot appear in on-target MEM seeds when they are
-shared.
-
-The MEM alphabet makes it easy to find strict on-target MEM seeds: they
-are the longest stretches of symbols containing $\uparrow$ and not
-containing $\downarrow$. However, shared on-target MEM seeds do not appear
-in this representation.
-
-\subsection{The extended MEM alphabet}
-
-To make shared on-target MEM seeds apparent, we decorate the $\uparrow$
-and $\downarrow$ symbols, effectively extending the alphabet. The
-$\downarrow$ symbol is replaced by the set of symbols $\downarrow_{/m}$,
-where $m$ is the number of duplicates that \emph{match} the erroneous
-nucleotide. The $\uparrow$ symbol is replaced by the set of symbols
-$\uparrow_{/j}$, where $j-1$ is the number of consecutive $\square$
-symbols preceding $\uparrow$.
-
-Note that the numbers decorating the symbols have a very different nature.
-In the case of the $\downarrow$ symbol, it is an information about the
-number of sequences matching the read; in the case of the $\uparrow$
-symbol, it is the position of the pivot relative to the previous
-sequencing error (or relative to the beginning of the read).
+We recode the reads as sequences of letters from the so-called MEM
+alphabet $\mathcal{A} = \{\square, |, \uparrow_{/1}, \uparrow_{/2},
+\uparrow_{/3} \ldots, \downarrow_{/0}, \downarrow_{/1}, \downarrow_{/2},
+\ldots\}$.
+
+The symbols $\downarrow_{/m}$ indicate that the nucleotide is a sequencing
+error and $m \geq 0$ is the number of off-target locations that
+\emph{match} the nucleotide. Since a sequencing error is always a mismatch
+against the target, the symbol $\downarrow_{/0}$ indicates that the
+ncleotide is a mismatch against \emph{every} candidate.
+
+The symbols $\uparrow_{/j}$ indicate a change in masking status: the
+nucleotide is not masked but its left neighbor is. The index $j \geq 1$ is
+the number of nucleotides since the last mismatch or since the beginning
+of the read.
+
+All the other nucleotides are represented by the $\square$ symbol.
+Finally, the $|$ symbol is appended after the last nucleotide of the read.
+It is an obligatory terminator that does not correspond to an actual
+nucleotide.
+
+Note that the numbers decorating $\uparrow$ and $\downarrow$ are different
+in nature. In the case of the symbol $\downarrow$ it is an information
+about the number of sequences matching the read; in the case of the symbol
+$\uparrow$ it is the position of the previous sequencing error.
+Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
+Fig.~\ref{fig:MEM_example} in the MEM alphabet.
 
@@
 \caption{\textbf{The extended MEM encoding}.
-The read of Fig.~\ref{fig:sketch_MEM} is represented in the extended MEM
-alphabet. The symbols $\uparrow$ and $\downarrow$ are now decorated by
-numbers. The arrows departing from those numbers help understand their
-meaning. The $\downarrow$ symbol is decorated by the number of locations
-that match the nucleotide. The $\uparrow$ symbol is decorated by the
-number of consecutive $\square$ symbols before it, plus one.}
+The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
+The arrows departing from those numbers help understand their meaning. The
+$\downarrow$ symbol is decorated by the number of locations that match the
+nucleotide. The $\uparrow$ symbol is decorated by the number of
+nucleotides from the last error or from the beginning of the read. Also
+note the presence of the terminator $|$ at the end of the read.}
 \label{fig:sketch_extended}
@@
 
-Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
-Fig.~\ref{fig:sketch_MEM} in the extended MEM alphabet. The arrows bring
-the focus on how the numbers decorating the $\uparrow$ and $\downarrow$
-symbols are computed. For instance, the $\downarrow_{/0}$ symbol indicates
-that the nucleotide is an error (this is the meaning of the $\downarrow$
-symbol), and that it matches no other location (this is meaning of $/0)$.
-In other words, it is a mismatch against every potential location.
-
-In this new encoding, shared on-target MEM seeds are stretches of
-$\square$ symbols flanked either by $\downarrow_{/0}$, or by $|$, or by
-the left end of the read. Indeed, such a stretch is a MEM seed because it
-matches at least one sequence (the target) and it cannot be extended
-(because the end of the read is reached or because there is no match in
-the genome). Also, it cannot be a strict on-target MEM seed because it
-does not contain the $\uparrow$ symbol, so it must be a shared on-target
-MEM seed.
-
-The reason for introducing more than just the $\downarrow_{/0}$ symbol is
-not immediately clear. For the combinatorial construction to be useful, we
-need not only to identify on-target MEM seeds, but also to compute the
-probabilities of the reads of each type. This is precisely the role of the
-extended alphabet, as we will see in section~\ref{sec:transfer_mat}.
+The MEM alphabet captures the masking status of the nucleotide as follows:
+If we denote the number of off-targets as $N$, the symbol
+$\downarrow_{/m}$ indicates that the nucleotide has $m$ hard masks and
+$N-m$ soft masks. In addition, the symbols $\uparrow_{/j}$ indicate that
+the nucleotide is unmasked.
+
+In the MEM alphabet, strict on-target MEM seeds are the longest stretches
+of symbols containing $\uparrow$ and not containing $\downarrow$. Indeed,
+such a stretch matches the target, it matches the target only because it
+contains at least one unmasked nucleotide (marked by $\uparrow$), and it
+cannot be extended because it is flanked by sequencing errors or by the
+ends of the reads. Note that there is exactly one symbol $\uparrow$ per
+strict on-target MEM seed, and therefore that two symbols $\uparrow$ must
+be separated by at least one symbol $\downarrow$.
+
+Shared on-target MEM seeds are the longest stretches of $\square$ symbols
+flanked by $\downarrow_{/0}$, or by the ends of the read. Indeed, such a
+stretch is a MEM seed because it matches the target and it cannot be
+extended ($\downarrow_{/0}$ is a mismatch against every sequence). Also,
+it cannot be a strict on-target MEM seed because it does not contain the
+$\uparrow$ symbol, so it must be shared.
 
@@
 
-The final twist in the combinatorial construction is to consider reads not
-as sequences of symbols, but as sequences of \emph{segments}. The purpose
-here is merely to express that some on-target MEM seeds are good
-(\textit{i.e.}, their size is $\gamma$ or greater) and some are not.
+Reads can be viewed as sequences of symbols, or as sequences of
+\emph{segments}. For the purpose of constructing reads without good seeds,
+the second point of view turns out to be essential.
 
@@
 Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
-from Fig.~\ref{fig:sketch_extended}. The numbers below indicate the size
-of each segment.
+from Fig.~\ref{fig:sketch_extended}.
 
@@
 \caption{\textbf{The extended MEM segment encoding}.
-The read from Fig.~\ref{fig:sketch_segment} is represented as a sequence
+The read from Fig.~\ref{fig:sketch_extended} is represented as a sequence
 of segments. Each segment is terminated by either $\uparrow_{/j}$, or
 $\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
-below indicate the size of each segment in nucleotides.}
+below indicate the size of each segment in number of nucleotides.}
 \label{fig:sketch_segment}
@@
 
-Shared on-target MEM seeds at the tail of the read are segments but all
-other on-target MEM seeds are not. Shared MEM seeds cannot contain the
-terminator $\downarrow_{/0}$ because it is a mismatch against the target.
-Strict on-target MEM seeds are split over at least two segments: one
-terminated by $\uparrow_{/j}$ and one right of it terminated by either
-$\downarrow_{/m}$ or $|$.
-
-\section{Symbolic and analytic representations}
-
-We now turn to the question of translating the extended MEM segment
-representation to a construction of reads without good seeds
-(\textit{i.e.}, without on-target MEM seeds of size $\gamma$ or greater).
-For this, we use tools borrowed from analytic combinatorics. We give the
-minimum amount of background that is necessary to expose the theory. A
-more detailed account of the strategy to compute seeding probabilities
-with analytic combinatorics has been published
+We now have a representation of reads as sequences of segments. Only some
+of these sequences are possible. For instance, a segment terminated by
+$\uparrow$ cannot follow another. The task, however, is not to give a
+construction of reads, but a construction of reads without good seeds. For
+this, we need to introduce a new set of tools.
+
+\section{The computational strategy}
+\label{sec:symbolic}
+
+We turn to the question of translating the extended MEM segment
+representation to a construction of reads without good seeds. For this, we
+use tools borrowed from analytic combinatorics. We give the minimum amount
+of background that is necessary to expose the theory. A more detailed
+account of the strategy to compute seeding probabilities with analytic
+combinatorics has been published
 elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
@@
 
-Let $\mathcal{A}$ be a set of combinatorial objects with a size $k \in
-\mathbb{N}$ and a probability of occurrence. Denoting $a_k$ the total
-probability of objects of size $k$, the function defined as $A(z) =
-\sum_{k\geq0} a_kz^k$ is called the \emph{weighted generating
-function} of the combinatorial objects.
-
-In our case, we are interested in the weighted generating function of
-reads without good seeds, so the coefficient $a_k$ is the probability
-that a read of size $k$ has no good seed. As $k$ becomes larger,
-it becomes more difficult to evaluate $a_k$, but the representation
-developed above will allow us to do this.
-
-\subsection{The symbolic method}
+The central object of analytic combinatorics is the \emph{generating
+function}.
+
+\begin{definition}
+\label{def:GF}
+Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
+size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
+generating function of $\mathcal{A}$ is defined as
+
+\begin{equation}
+\label{eq:GF1}
+A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
+\end{equation}
+where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
+respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
+\geq 0}$ such that
+
+\begin{equation*}
+A(z) = \sum_{k=0}^\infty a_k z^k.
+\end{equation*}
+
+By definition $a_k = \sum_{a \in A_k}w(a)$, where $A_k$ is the class of
+objects of size $k$ in $\mathcal{A}$. The number $a_k$ is the
+total weight of objects of size $k$.
+\end{definition}
+
+To give a concrete example, if the probability of the symbol
+$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
+containing only this symbol is $qz$. In our case, we are interested in the
+weighted generating function of reads without good seeds, so that the
+coefficient $a_k$ is equal to the probability that a read of size $k$ has
+no good seed.
+
+The motivation for introducing weighted generating functions is that
+operations on combinatorial objects translate into operations on their
+weighted generating functions. If $A(z)$ and $B(z)$ are the weighted
+generating functions of two mutually exclusive sets $\mathcal{A}$ and
+$\mathcal{B}$, the weighted generating function of $\mathcal{A} \cup
+\mathcal{B}$ is $A(z) + B(z)$, as evident from expression (\ref{eq:GF1}).
+Size and weight can be defined for pairs of objects in $\mathcal{A} \times
+\mathcal{B}$ as $|(a,b)| = |a| + |b|$ and $w(a,b) = w(a)w(b)$. In other
+words the sizes are added and the weights are multiplied.  With this
+convention, the weighted generating function of the Cartesian product
+$\mathcal{A} \times \mathcal{B}$ is $A(z)B(z)$. This simply follows from
+expression (\ref{eq:GF1}) and from
+
+\begin{equation*}
+A(z)B(z) =
+\sum_{a\in \mathcal{A}}w(a)z^{|a|} \sum_{b\in \mathcal{B}}w(b)z^{|b|}
+= \sum_{(a,b) \in \mathcal{A} \times \mathcal{B}} w(a)w(b)z^{|a|+|b|}.
+\end{equation*}
+
+These two operations are all we need in order to compute the weighted
+generating functions of the reads of interest. Addition corresponds to
+creating a new family by merging reads from families $\mathcal{A}$ and
+$\mathcal{B}$. Multiplication corresponds to creating a new family by
+concatenating reads from families $\mathcal{A}$ and $\mathcal{B}$.
+
+\subsection{Analytic representation}
 
@@
 
-\subsection{Hard and soft masking}
-
-For each potential location including the target, we define the
-\emph{match streak} at a given nucleotide as the number of nucleotides
-since the last mismatch on the left. For instance, in
-Fig.~\ref{fig:MEM_example}, the match streak of the target at the ninth
-nucleotide is $1$ while for the second potential location, it is $7$.
-
-\begin{definition}
-At a given position of the read, an off-target location is a \emph{hard
-mask} if its success streak is strictly longer than the match streak of
-the target. An off-target location is a \emph{soft mask} if its has the
-same match streak as the target.
-\end{definition}
-
-For instance, the second location in the example above is a hard mask at
-the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further check
-that all the off-target locations are soft masks at the second nucleotide
-and that the second location is a soft mask at the last nucleotide of the
-read.
-
-A nucleotide with a hard mask cannot be part of an on-target MEM seed. A
-nucleotide without mask belongs to a strict on-target MEM seed. A
-nucleotide with a soft mask may belong to a strict on-target MEM seed, to
-a shared on-target MEM seed, or may not belong to any MEM. This case is
-the most intricate and it will be teased apart below.
-
-The extended MEM alphabet is nothing more than a way to keep track of the
-hard and soft masks. For instance, a pivot (associated with the symbols
-$\uparrow_{/j}$) is a nucleotide where the number of masks becomes 0.
-Denoting the number of off-target locations by $N$, the symbol
-$\downarrow_{/m}$ indicates that there are $m$ hard masks and $N-m$ soft
-masks.
-
 \subsection{Error model and divergence of the duplicates}
 
-We assume that the target sequence has $N$ duplicates. We further assume
-that duplication was instantaenous and that all $N+1$ sequences diverge
-independently from each other at a constant rate. In other words, we
-ignore the complications due to the genealogy of the duplication events
-and we simply assume that at each position, any given duplicate is
-identical to the target with probability $1-\mu$. If it is not, we
-assume that the duplicate has any of the remaining three nucleotides with
-equal probability (\textit{i.e.} each is found with probability $\mu/3$).
+We assume that the target sequence has $N$ duplicates, so that there are
+$N$ off-target sequences. We further assume that duplication was
+instantaenous and that all $N+1$ sequences diverge independently of each
+other at a constant rate. In other words, we ignore the complications due
+to the genealogy of the duplication events and we simply assume that at
+each position, any given duplicate is identical to the target with
+probability $1-\mu$. If it is not, we assume that the duplicate has any of
+the remaining three nucleotides with equal probability (\textit{i.e.} each
+is found with probability $\mu/3$).
 
@@
 
-On a read error the target is never identical to the read (because we
-assume that the target is the true sequence), and a duplicate is identical
-to the read with probability $mu/3$ (the probability that the duplicate is
-different from the target, times the probability that the nucleotide is
-the same as the error). On a correct nucleotide, the target is always
-identical to the read, and a dupicate is identical with probability
-$1-\mu$ (the probablity that the duplicate is identical to the target).
-
-With these assumptions, we are now ready to describe the entries of the
-transfer matrix $M(z)$, which we do in order of increasing difficulty.
+Read errors are always mismatches against the target (because we assume
+that the target is the true sequence), and they match each off-target
+sequence with probability $\mu/3$. Correct nucleotides are always a match
+for the target, and they match each off-target sequence with probability
+$1-\mu$.
+
+We are now ready to describe the transfer matrix $M(z)$. We do this entry
+by entry, in order of increasing complexity.
 
@@
 
-A $\uparrow_{/j}$ symbol marks a pivot and the absence of masks. The
-segment is thus the beginning of an on-target MEM of size at least $j$.
-The MEM will extend until the next sequencing error, or until the end of
-the read, so the number of $\square$ symbols in the next segment must be
-at most $\gamma-j-1$ and it must be terminated by a $\downarrow$ symbol or
-by the tail terminator $|$.
-
-In this case, the matches between the read and the duplicates are
-irrelevant until the terminator. The weighted genearting function of each
-$\square$ symbol is simply $qz$ ($q$ is the probability that the
-nucleotide is not an error and $z$ marks objects of size $1$).
+A segment terminated by $\uparrow_{/j}$ is the beginning of an on-target
+MEM seed of size at least $j$. The seed reaches the next sequencing
+error or the end of the read, so the number of $\square$ symbols in the
+next segment must be at most $\gamma-j-1$ and it must be terminated by a
+$\downarrow$ symbol or by the tail terminator $|$.
+
+On this segment, the matches between the read and the off-target sequences
+are irrelevant. The weighted genearting function of each $\square$ symbol
+is simply $qz$ ($q$ is the probability that the nucleotide is not an error
+and $z$ marks objects of size $1$). The following definition will simplify
+the notations.
 
@@
 The weighted generating function of the $\downarrow_{/m}$ terminator is
-thus $\omega_m pz$ (the probability of the $\downarrow_{/m}$ is $\omega_m
-p$ and $z$ marks objects of size $1$).
+thus $\omega_m pz$ (the probability of the symbol $\downarrow_{/m}$ is
+$\omega_m p$ and $z$ marks objects of size $1$).
 
@@
 
-At a $\downarrow_{/m}$ symbol, there are $m$ hard masks and $N-m$ soft
-masks. If all the masks disappear before the first read error, the next
-terminator will be a $\uparrow$ symbol, otherwise it will be a
-$\downarrow$ symbol.
+The symbol $\downarrow_{/m}$ signifies that the nucleotide has $m$ hard
+masks and $N-m$ soft masks. If all the masks disappear before the first
+read error, the next terminator will be a $\uparrow$ symbol, otherwise it
+will be a $\downarrow$ symbol. We separate the cases based on the
+terminator of the segment.
 
@@
 \begin{definition}
-The probability that a given duplicate sequence contains a mismatch in a
+The probability that a given off-target sequence contains a mismatch in a
 sequence of $i$ error-free nucleotides is
@@
 sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
-As mentioned above, $i < \gamma$ for reads without on-target MEM seed.
-Also note that this expression is the same for all $\downarrow$ symbols;
-it does not depend on $m$.
+The fact that the reads have no good seeds imposes $i < \gamma$. Also note
+that this expression is the same for all $\downarrow$ symbols; it does not
+depend on $m$.
 
@@
 
-In this case there can be no on-target MEM seed. We must just enforce the
-condition that at least one of the $N$ masks survives until the end of the
-segment in order to exclude the $\uparrow$ symbol. The weighted generating
-function is
+In this case there can be no good seed because the read terminates too
+early. We must just enforce the condition that at least one of the $N$
+masks survives until the end of the segment in order to exclude the
+$\uparrow$ symbol. The weighted generating function is
 \begin{equation*}
@@
 
-In this case, the soft masks are irrelevant. Even if they survive until
-the end of the segment, there will be a shared on-target MEM seed. So we
-must enforce the condition that at least one hard mask survives until the
-end of the segment (which is impossible if $m = 0$).
+In this case, the soft masks do not hide the target: even if one of them
+survives until the end of the read, there will be a good seed (shared in
+this case). To make sure that the read has no good seed, we must enforce
+the condition that at least one hard mask survives until the end of the
+segment (which is impossible if $m = 0$). The weighted generating function
+is
 \begin{equation*}
@@
 
-The transition from $\downarrow_{/m}$ to $|$ is
+Finally, the weighted generating function of the tail segments following
+segments terminated by $\downarrow_{/m}$ is
 \begin{equation}
@@
 
-As above, there can be no on-target MEM seed and we must just exclude the
+In this case, there can be no good seed and we must only exclude the
 $\uparrow$ terminator. The weighted generating function is
@@
 
-\subsubsection*{Case 3a: the terminator $\downarrow_{/n}$ comes after the
+\subsubsection*{Case 3b: the terminator $\downarrow_{/n}$ comes after the
 $\gamma$-th nucleotide}
@@
 This case is by far the most convoluted. Since the segment contains at
-least $\gamma$ error-free nucleotides, we must take care that it does not
-generate an on-target MEM seed. This will be the case if any of the two
-following conditions is validated: $i)$ at least one hard mask covers all
-the error-free nucleotides, or $ii)$ none of the hard masks does but at
-least one soft mask covers the whole segment (including the terminator).
+least $\gamma$ error-free nucleotides, we must ensure that it does not
+contain a good seed. This will be the case if any of the two following
+conditions is validated: $i)$ at least one hard mask covers all the
+error-free nucleotides, or $ii)$ all the hard masks vanish but at least
+one soft mask covers the whole segment (including the terminator).
 
@@
 to case $i)$ and the panel to case $ii)$. The top row represents the
-target, and the bottom rows represent duplicates (using the same
-symbols as in Fig.~\ref{fig:sketch_MEM}).
+target, and the bottom rows represent off-target sequences (using the same
+symbols as in Fig.~\ref{fig:MEM_example}).
 \begin{inset}
@@
 
-Whenever a hard mask (here the first duplicate) covers the nucleotides
-as shown in the left panel, there can be no on-target MEM seed,
+Whenever a hard mask (here the first off-target sequence) covers the
+nucleotides as shown in the left panel, there can be no good seed,
 irrespective of the value of $\gamma$. The positions marked with a
 question mark are irrelevant, they cannot change the fact that there is no
-on-target MEM seed. If the hard masks are interrupted, as in the right
-panel, then it depends on the soft masks. If a soft mask covers the whole
-segment, then there can be no on-target MEM seed, irrespective of the
-value of $\gamma$.
+on-target MEM seed. If the hard masks vanish, as in the right panel, then
+it all depends on the soft masks. If a soft mask covers the whole segment,
+then there can be no good seed, irrespective of the value of $\gamma$.
 
@@
 
-For condition $ii)$, we will need to introduce some new notations.
+Condition $ii)$ is more convoluted, so we introduce some further
+notations to solve this sub-case.
 \begin{definition}
-The probability that a given duplicate sequence contains a mismatch in a
+The probability that a given off-target sequence contains a mismatch in a
 sequence of $i$ error-free nucleotides followed by an error is
@@
 \end{equation*}
-but we need to break up this term among the terminators $\downarrow_{/n}$,
-($0 \leq n \leq N$). For this, we decompose the sum on the number of soft
-masks that run to the end of the segment, including the terminator. The
-probability that there are $r \geq 1$ such soft masks is
+but we need to break up this term among all the terminators
+$\downarrow_{/n}$, ($0 \leq n \leq N$). For this, we decompose the sum on
+the number of soft masks that run to the end of the segment, including the
+terminator. The probability that there are $r \geq 1$ such soft masks is
 \begin{equation*}
@@
 Each of them matches the nucleotide at the end of the segment, so the
-total number of matches is $r$ plus the number of sequences, among the
-remaining $N-m-r$ soft masks and $m$ hard masks, that also match the
-last nucleotide.
-
-Let us start with the $N-m-r$ that failed somewhere in the segment. The
+total number of matches is $r$ plus the number of sequences that also
+match the last nucleotide, among the remaining $N-m-r$ soft masks and $m$
+hard masks.
+
+Let us start with the $N-m-r$ that vanished somewhere in the segment. The
 conditional probability that they match the last nucleotide given that
-they fail somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
-conditional probability that they do not match is $(1-\mu/3) / \eta_i$.
+they vanish somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
+conditional probability that they do not match it is $(1-\mu/3) / \eta_i$.
 For the $m$ hard masks, the probability that they match the last
-nucleotide given that they sail somewhere in the segment is simply
-$\mu/3$, and the probability that they do not is $1-\mu/3$.
-
-The probability that those $N-r$ duplicates contribute $n-r$ matches is
+nucleotide given that they vanish somewhere in the segment is simply
+$\mu/3$, and the probability that they do not match it is $1-\mu/3$.
+
+The probability that the last nucleotide of the segment matches $n-r$ of
+the $N-r$ masks that vanished before the terminator is
 \begin{equation*}
@@
 
-The probability that in
-total $n$ sequences match the terminator is
+Finally, the probability that a total of $n$ off-target sequences match
+the terminator and thus that the terminator is the symbol
+$\downarrow_{/n}$ is
 \begin{eqnarray*}
@@
 
-The weighted generating function we are looking for is thus
+The weighted generating function of the segments terminated by
+$\downarrow_{/n}$ following segments terminated by $\downarrow_{/m}$ is
 \begin{equation}
@@
 \subsection{The expression of the transfer matrix}
-
-The final expression for the transfer matrix $M(z)$ is
+\label{sec:expression_of_M}
+
+Collecting and arranging the results above, the final expression of the
+transfer matrix $M(z)$ appears as
 \begin{equation*}
@@
 \section{Computing the probabilities}
+\label{sec:compute}
 
 \subsection{Coefficient extraction}
+\label{sec:extract}
 
@@
 a_1z + a_2z^2 + \ldots$ is to extract the coefficient $a_k$, which
-represents the probability that a read of size $k$ does not contain an
-on-target MEM seed. $F(z)$ is the entry at coordinates $(1,N+\gamma+1)$ of
-the matrix $M(z) \cdot (I-M(z))^{-1}$. The expression of $M(z)$ is in
-general too complex to compute this function directly. Instead, we return
-to the expression $M(z) \cdot (I-M(z))^{-1} = M(z) + M(z)^2 + \ldots$ and
-observe that the terms $M(z)^{k+2}, M(z)^{k+3}, \ldots$ have no influence
-on the coefficients $a_0, a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments. Since the entries
-of $M(z)^k$ are the weighted generating functions of reads with exactly
-$k$ segments, $a_k$ cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More
-formally, we can prove by induction that all the entries of $M(z)^k$ are
-divisible by $z^{k-1}$, showing that the contribution of $M(z)^{k+2} +
-M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is $0$.
-
-So we can extract the coefficients of $F(z)$ up to order $k$ by computing
-the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$, and compute the first
-terms of the Taylor expansion of the entry at coordinates
-$(1,N+\gamma+1)$.
-
-But we can do better than that. If we are only interested in the
-coefficients of order up to $k$, we can replace the infinite power series
+represents the probability that a read of size $k$ does not contain a
+good seed.
+
+$F(z)$ is the entry of the matrix $M(z) \cdot (I-M(z))^{-1}$ at the first
+row and the last column, associated with terminators $\downarrow_{/0}$ and
+$|$. The expression of $M(z)$ is too complex to compute this function
+directly. Instead, we return to the definition $M(z) \cdot (I-M(z))^{-1} =
+M(z) + M(z)^2 + \ldots$ and observe that the terms $M(z)^{k+2},
+M(z)^{k+3}, \ldots$ have no influence on the coefficients $a_0, a_1,
+\ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k+1$ segments\footnote{Each
+segment contains a terminator, which has size 1, except the unique
+terminator $|$, which has size 0.}. Since the entries of $M(z)^n$ are the
+weighted generating functions of reads with exactly $n$ segments, $a_k$
+cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by
+induction that all the entries of $M(z)^k$ are divisible by $z^{k-1}$,
+showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to
+$a_0 + a_1z + \ldots +a_kz^k$ is $0$.
+
+So to extract the coefficients of $F(z)$ up to order $k$ we can compute
+the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$, isolate the entry at the
+first row and the last column and then compute the terms of the Taylor
+expansion up to order $k$.
+
+But we can do better than that. Since we are only interested in the
+coefficients up to order $k$, we can replace the infinite power series
 of the terms $A_{m,n}(z)$ and $C_m(z)$ by their truncated versions where
-we keep only the terms of order up to $k$. Likewise, we perform all
-algebraic operation on truncated polynomials of order $k$, \textit{i.e.}
-we discard at every stage the coefficients of order greater than $k$. With
-this approach, the entry of the matrix $M(z) + M(z)^2 + \ldots +
-M(z)^{k+1}$ at coordinates $(1,N+\gamma+1)$ is a truncated polynomial that
-directly provides the coefficients of interest.
-
-But we can do better than that. A read with $k$ segments contains at least
-$(k-1)/2$ errors. Indeed, two segments terminated by $\uparrow$ cannot
-follow each other, so at least (approximately) half of the segments must
-be terminated by $\downarrow$, which means that the segment contains a
-read error. Since read errors typically have a small probability of
-occurrence, all the coefficients of $M(z)^k$ rapidly converge to $0$ and
-become negligible as $k$ increases. So instead of computing $M(z) + M(z)^2
-+ \ldots + M(z)^{k+1}$, we can interrupt the summation after a certain
-power of $M(z)$ because the terms are negligible.
-
-This approach makes it feasible to compute the probability that a read of
-size $k$ has no on-target MEM seed when the matrix $M(z)$ is small, but
-recall that $M(z)$ has dimension $N+\gamma+1$. Some sequences can have
-million of duplicates in mammalian genomes (\textit{i.e.} $N > 10^6$),
-which makes the computations with this method unfeasible. Clearly, another
-method is needed.
+we keep only the terms up to order $k$. Likewise, we perform all
+algebraic operation on truncated polynomials of order $k$, \textit{i.e.},
+we discard at every stage the coefficients of order $k+1$ or greater.
+
+More formally, truncated polynomials are defined as polynomials whose
+coeffients of degree greater than $k$ are 0. Multiplying two truncated
+polynomials $P(z)$ and $Q(z)$ gives a truncated polynomial $R(z)$ whose
+coefficients up to order $k$ are those of $P(z)Q(z)$. Using truncated
+polynomials of order $k$, the entry of the matrix $M(z) + M(z)^2 + \ldots
++ M(z)^{k+1}$ at the first row and the last column directly provides the
+coefficients of interest.
+
+But we can do even better than that. A read with $n$ segments contains at
+least $m=\lfloor{(n-1)/2}\rfloor$ errors. Indeed, two segments terminated
+by $\uparrow$ cannot follow each other, so at least (approximately) half
+of the segments must be terminated by $\downarrow$, which means that the
+segment contains a read error. Since read errors typically have a small
+probability of occurrence, all the coefficients of $M(z)^n$ rapidly
+converge to $0$ as $n$ increases. So instead of computing the matrix $M(z)
++ M(z)^2 + \ldots + M(z)^{k+1}$, we can interrupt the summation after a
+certain power of $M(z)$ because the terms become negligible.
+
+The number of errors $X$ in a read of size $k$ has a Binomial
+distribution $X \sim B(k,p)$. From\cite{arratia1989tutorial} we can bound
+the probabilities of the tail with the expression
+
+\begin{equation}
+\label{eq:bound}
+Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-p)} -m\log
+\frac{m}{kp} \right).
+\end{equation}
+
+Using the formula above, we can thus bound the probability that a read has
+$n$ segments or more. We compute the matrices $M(z) + M(z)^2 + \ldots
++M(z)^n$ where the weighted generating functions have been replaced by
+truncated polynomials and we extract $F_n(z)$, the entry at the first
+column and the last row. When the bound is lower than a set fraction
+$\varepsilon$ of the coefficient of $z^k$, we stop the computations.
+Typically $\varepsilon = 0.01$ so this method ensures that the
+probabilities that a read of size $k$ has no good seeds are accurate to
+within 1\%.
+
+This approach is feasible when the matrix $M(z)$ is small, but recall that
+$M(z)$ has $N+\gamma+1$ rows/columns. Some sequences can have millions of
+duplicates in plant and animal genomes (\textit{i.e.} $N > 10^6$), which
+makes the computations with this method unfeasible because even a few
+matrix multiplications are prohibitive. For large $N$, another method is
+clearly needed.
 
@@
 
-The symbolic representation of reads in the extended MEM alphabet can also
-be used as a basis for an efficient method to sample reads. Instead of
-generating the nucleotides of each sequence one by one and comparing them
-to the read, one can generate segments of the extended MEM alphabet. As
-argued above, a read contains few segments when the error rate is small.
-Since this property does not depend on the value of $N$, we can obtain a
-fast Monte Carlo method to sample millions of reads and count the
-proportion that contain an on-target MEM seed.
-
-The principle is to first sample the position of the next read error, and
-then sample the number of masks that survive until this position. If none
-of the does, the read contains an on-target MEM seed, provided the next
-read error is at a distance greater than $\gamma$. Otherwise, the process
-is repeated for the next segment until an on-target MEM seed is generated,
-or until the read has size $k$ or greater (and is then truncated).
-
-The method is summarized by the algorithm below (FIX THE ALGORITHM).
+The symbolic representation as MEM segments can be used for an efficient
+method to sample reads. Instead of generating the nucleotides of the $N+1$
+sequences one by one, one can generate a single sequence of MEM segments.
+Since the number of segment does not depend on $N$, we can obtain a fast
+Monte Carlo method to sample millions of reads and count the proportion
+that contain a good seed.
+
+The principle is to proceed in cycles of two steps. We first sample the
+position of the next read error, which gives the position of the next
+symbol $\downarrow$. The second step is to determine whether there is a
+symbol $\uparrow$ before that. For this we sample the number of masks that
+vanish before the symbol $\downarrow$. If they all vanish, the read
+contains a good seed, provided the next read error is at a distance
+greater than $\gamma$. Otherwise, the process is repeated until we
+generate a good seed, or until the read has size $k$ or greater (in which
+case it has no good seed).
+
+The method is summarized by the algorithm below. It requires efficient
+methods to generate random samples with a geometric distribution and a
+binomial distribution. Sampling from a geometric distribution can be done
+by computing the logarithm of a uniform $(0,1)$ random variable. Sampling
+from a binomial distribution can be done by the method of
+Kachitvichyanukul and Schmeiser\cite{kachitvichyanukul1988binomial}.
 
 \begin{algorithm}[H]
+\label{alg:mcmc}
 \SetAlgoLined
-\KwResult {Sample a read, return 1 if it contains an on-target MEM seed,
-  otherwise return 0.}
-  $read.size \leftarrow 0$\;
+\KwResult {Sample a read at random. Return 1 if the read contains a
+good seed, otherwise return 0.}
+  $\lambda \leftarrow k$ \Comment*[r]{Size remaining to generate.}
   $m \leftarrow 0$ \Comment*[r]{Current number of hard masks.}
-  \While {$read.size < k$}{
-    $i \leftarrow geom(p)$ \Comment*[r]{Position of next error.}
-    \eIf {$read.size + i > k$}{
-      $\lambda \leftarrow k - read.size$ \Comment*[r]{Size of the tail.}
+  \While {$\lambda > 0$}{
+    $i \leftarrow geom(p)-1$ \Comment*[r]{Error-free nucleotides.}
+    \eIf {$i \geq \lambda$}{
       \eIf{$\lambda < \gamma$} {
@@
         $m \leftarrow s + binom(N-s, \mu/3)$\;
-        $read.size \leftarrow read.size + i$\;}
+        $\lambda \leftarrow \lambda - (i+1)$\;}
     }
@@
 
+In practice, we use a slightly modified version of this algorithm, that
+returns the position of the good seeds if they are generated. This allows
+us to estimate the probabilities for \emph{all} read sizes up to $k$. We
+typically run this Monte Carlo sampling scheme 10,000,000 times, so that
+we can estimate probabilities as low as $\sim1$ in 1 million.
+
+The running speed depends on the error rate of the instrument $p$, but not
+on the number of off-target sequences $N$. For values of $p$ around 0.01,
+the running time for 10,000,000 iterations is $1$-$10$ seconds on current
+hardware.
+
+\section{False positives}
+
+So far we have designed a method to estimate the probability that a
+sequencing read contains a good seed. We referred to this as outcome $i.$
+in section~\ref{sec:seedheur}. However, we are interested in the
+probability that a read is mapped at an incorrect location, which is
+outcome $ii.$
+
+Computing the probability of outcome $ii.$ is hard, but we can compute the
+probability that the read is not mapped at all, which is outcome $iii.$
+The probability of outcome $ii.$ is then found by subtraction.
+
+\subsection{The probability of outcome $iii.$}
+
+Outcome $iii.$ implies that the read contains no seed at all, neither
+on-target, nor off-target. This requires that there is no exact match of
+size $\geq \gamma$ for \emph{any} sequence. Indeed, if it were the case,
+then there would a MEM seed of size $\geq \gamma$ for at least one of the
+sequences.
+
+Let $E_0$ denote the event that there is no on-target match of size $\geq
+\gamma$, and $E_j$ denote the event that there is no match of size $\geq
+\gamma$ for off-target sequence $j$ ($1 \leq j \leq N$). We are thus
+interested in computing $P(E_0 \cap \ldots \cap E_N)$. Since we assume
+that the off-target sequences have evolved through the same mutagenesis
+process and independently of each other, we can write
+
+\begin{eqnarray}
+\notag
+P(E_0 \cap \ldots \cap E_N) &=&
+  P(E_0) \cdot P(E_1 \cap \ldots \cap E_N | E_0) \\
+\label{eq:PE0E1}
+  &=& P(E_0) \cdot P(E_1 | E_0)^N =
+  P(E_0) \cdot \left( \frac{P(E_0 \cap E_1)}{P(E_0)} \right)^N.
+\end{eqnarray}
+
+Expression (\ref{eq:PE0E1}) shows that we can compute the probability of
+outcome $iii.$ from two quantities, namely $P(E_0)$ and $P(E_0 \cap E_1)$.
+
+\subsection{Computing $P(E_0)$}
+\label{sec:cPE0}
+
+$P(E_0)$ is straightforward to evaluate because it involves a single
+sequence. Methods to do this efficiently have been explained in detail
+elsewhere\cite{filion2018analytic}. Here we only give the recurrence
+equation to compute $P(E_0)$. Assume that the probability of a sequencing
+error $p$ is known, that $q = 1-p$, and that $a_k$ is the probability that
+a read of size $k$ does not contain a match of size $\geq \gamma$ for the
+target. Then the following equation can be used to compute $a_k$ by
+recurrence:
+
+\begin{equation}
+\label{eq:PE0}
+a_k = 
+\begin{cases}
+1            &\quad\text{if } k < \gamma, \\
+1 -pq^\gamma &\quad\text{if } k = \gamma, \\
+a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
+\end{cases}
+\end{equation}
+
+Knowing the size of the read and the error rate of the sequencing
+instrument, we can thus compute $P(E_0)$ using expression (\ref{eq:PE0}).
+
+\subsection{Computing $P(E_0 \cap E_1)$}
+\label{sec:cPE0E1}
+
+The event $E_0 \cap E_1$ is that the read contains no match of size $\geq
+\gamma$ for either the target or the first off-target sequence (here
+``first'' refers to any fixed off-target sequence). We compute the
+probability of this event with the strategy described in
+section~\ref{sec:symbolic}: we first recode the reads using a different
+alphabet and then we specify the transfer matrix of the reads that have no
+match for any sequence. The powers of this matrix hold the weighted
+generating function of interest, from which we extract the coefficients
+that correspond to the terms $P(E_0 \cap E_1)$ for reads of different
+sizes.
+
+The reads are recoded as sequences of symbols from the following alphabet
+$\mathcal{A}_* = \{\square, |, \downarrow_{/1}^-, \downarrow_{/2}^-,
+\ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots, \Downarrow\}$. The
+symbols $\downarrow_{/j}^-$ signify that the nucleotide is a mismatch
+against the off-target sequence, the symbols $\downarrow_{/j}^+$ signify
+that it is a mismatch against the target, and the symbol $\Downarrow$
+signifies that it is a mismatch against both. As before, every other
+nucleotide is replaced by the symbol $\square$, and the terminator $|$
+is appended at the end of the read. We again define reads as sequences of
+segments, except that now the terminators are the symbols
+$\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
+
+The number $j$ in the symbol $\downarrow_{/j}^-$ indicates the relative
+position of the last mismatch against the target. Likewise, the number $j$
+in the symbol $\downarrow_{/j}^+$ indicates the relative position of the
+last mismatch against the off-target sequence. In other words, the index
+$j$ is the size of the match for the sequence that does not have a
+mismatch. The terminators thus capture the local state of the read. For
+instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
+mismatch against the off-target sequence, that it is a match for the
+target, and that the six previous nucleotides were also a match for the
+target.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.88]{sketch_dual.pdf}
+\caption{\textbf{Title.}
+Caption.}
+\label{fig:dual}
+\end{figure}
+
+We use the same error model as before: for every nucleotide, the read
+differs from the target with a fixed probability $p$ and the sequences
+differ from each other with a fixed probability $\mu$. With these
+definitions, the probability of occurrence of the symbol $\square$ is $a =
+(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
+symbols $\downarrow_{/j}^-$ is $b = p\mu/3$, that of the symbols 
+$\downarrow_{/j}^+$ is $c = (1-p)\mu$, and the probability of occurrence
+of the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
+$a+b+c+d=1$.
+
+These definitions allow us to specify the weighted generating function of
+the segments. For instance, the weighted generating function of the
+first segment of Fig.~\ref{fig:dual} is $(az)bz$, and that of the second is
+$(az)^5cz$.
+
+To specify the transfer matrix, we must ensure that we eliminate
+the concatenations of two segments that would create a match of size $\geq
+\gamma$ for any of the two sequences. After defining
+\begin{equation*}
+\begin{gathered}
+r_i^+(z) = (az)^icz, \\
+r_i^-(z) = (az)^ibz, \\
+R_i(z) = \big( 1 + az + \ldots + (az)^i \big)dz, \\
+F_i(z) = 1 + az + \ldots + (az)^i,
+\end{gathered}
+\end{equation*}
+we can verify that the expression of the transfer matrix $L(z)$ is
+
+\begin{equation*}
+\begin{blockarray}{ccccccccc}
+   & \scriptstyle{\Downarrow} & \scriptstyle{\downarrow_{/1}^+} & 
+    \ldots & \scriptstyle{\downarrow_{\gamma-1}^+} &
+    \scriptstyle{\downarrow_{/1}^-} & \ldots &
+    \scriptstyle{\downarrow_{/\gamma-1}^-} & \scriptstyle{|} \\
+\begin{block}{c[cccccccc]}
+\scriptstyle{\Downarrow} & R_{\gamma-1}(z)  & r_0^+(z) & \ldots &
+    r_{\gamma-2}^+(z) & r_0^-(z) & \ldots & r_{\gamma-2}^-(z) &
+    F_{\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/1}^+} & R_{\gamma-2}(z) & & & & & & &
+    F_{\gamma-2}(z) \\
+\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
+    F_{\gamma-3}(z) \\
+\vdots & \vdots & & A(z) & & & B(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
+\scriptstyle{\downarrow_{/1}^-} & R_{\gamma-2}(z) & & & & & & &
+    F_{\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
+    F_{\gamma-2}(z) \\
+\vdots & \vdots & & C(z) & & & D(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
+\scriptstyle{|} & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are square matrices with
+$\gamma-1$ rows/columns, that are defined as
+
+\begin{equation*}
+A(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & 0 & r_0^+(z) & \ldots &
+    r_{\gamma-4}^+(z) & r_{\gamma-3}^+(z) \\
+\scriptstyle{\downarrow_{/2}^+} & 0 & 0 & \ldots &
+    r_{\gamma-5}^+(z) & r_{\gamma-4}^+(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & 0 & 0 & \ldots & 0 & r_0^+(z) \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+B(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
+   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
+   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & r_0^-(z) & r_1^-(z) & \ldots &
+    r_{\gamma-3}^-(z) & r_{\gamma-2}^-(z) \\
+\scriptstyle{\downarrow_{/2}^+} & r_0^-(z) & r_1^-(z) & \ldots &
+    r_{\gamma-3}^-(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & r_0^-(z) & r_1^-(z) &
+    \ldots & 0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & r_0^-(z) & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+C(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & r_{\gamma-2}^+(z) \\
+\scriptstyle{\downarrow_{/2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & r_0^+(z) & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+D(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
+    \scriptstyle{\downarrow_{/\gamma-1}^-} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & 0 & r_0^-(z) & \ldots &
+    r_{\gamma-4}^-(z) & r_{\gamma-3}^-(z) \\
+\scriptstyle{\downarrow_{/2}^-} & 0 & 0 & \ldots &
+    r_{\gamma-5}^-(z) & r_{\gamma-4}^-(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & 0 & 0 & \ldots & 0 & r_0^-(z) \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}.
+\end{equation*}
+
+The matrix $L(z)$ is relatively sparse and most non-zero entries are
+monomials. In addition, the number of rows/columns is approximately
+$2\gamma$, so the dimension of the matrix rarely exceeds 50. This means
+that computing the powers of $L(z)$ is not prohibitive and we do not need
+to design a Monte Carlo sampling scheme as we did before.
+
+That being said, we do not need to compute all the powers up to $k$
+because the terms vanish and become negligible after a certain iteration.
+Indeed, the terms of $L(z)^n$ are the weighted generating function of the
+reads without a match for either sequences and that contain exactly $n$
+segments. Of those $n$ segments, $n-1$ are terminated by a mismatch
+against at least one of the sequences. Following the rationale of
+section~\ref{sec:extract}, we define $\tilde{p}$ as the upper bound on the
+probability of a mismatch $\tilde{p} = \max\{b,c,d\}$, and updating
+formula (\ref{eq:bound}) we bound the probability that a read of size $k$
+contains $m = n-1$ or more mismatches
+
+\begin{equation*}
+Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-\tilde{p})} -m\log
+\frac{m}{k\tilde{p}} \right).
+\end{equation*}
+
+The strategy is the same as in section~\ref{sec:extract}: we compute the
+sum of matrices $L(z) + L(z)^2 + \ldots + L(z)^n$ where the weighted
+generating functions have been replaced by truncated polynomials and we
+extract the entry at the first column and the last row. When the bound is
+lower than a set fraction $\varepsilon$ of the coefficient of $z^k$, we
+stop the computations.
+
+\subsection{The probability of outcome $ii.$}
+
+For given values of the parameters $p$, $\mu$ and $N$, we can now compute
+the probability of a false positive, which is outcome $ii$.
+
+For this, we compute the probability of outcome $i.$, either by computing
+the powers of $M(z)$ or by Monte Carlo sampling as explained in
+section~\ref{sec:compute}. Next we compute the probability of outcome
+$iii.$ using expression (\ref{eq:PE0E1}), where the terms are computed as
+explained in sections~\ref{sec:cPE0E1} and \ref{sec:cPE0}.
+
 \section{Estimating the parameters}
-
-Let us assume that we have a sequence of length $L$ that has $N$
-duplicates in the genome. Let us further assume that the instrument has a
-uniform error rate of $p$, and that at each nucleotide, duplicates are
-different from the original sequence with probability $\mu$. What are the
-values of $p$ and $\mu$ that maximize the probability that the read is
-different from the original, but identical to a duplicate?
-
-If the read has no error, this cannot happen. If we consider the reads
-with one error, the probability that this happens is
-\begin{equation}
-Q(p,\mu) = L \cdot p \cdot (1-p)^{L-1}
-\cdot \Big( 1- (1-(1-\mu)^{L-1} \mu/3)^N \Big).
-\end{equation}
-
-Observe that for every value of $p$, $\mu = 1/L$ maximizes $Q(p,\mu)$. So
-in this simple proble, there exists a ``worst'' value of $\mu$, that
-maximizes the probability of making a mistake irrespective of the
-precision of the sequencing instrument and of the number of duplicates.
-
-The MEM seeding heuristic is more complex than the toy problem above, but
-we still get the intuition that values of $\mu$ close to $1/L$ are the
-most problematic, where $L$ is the typical size of MEM seeds.
-
-Using the above methods to compute MEM seeding probabilities, we can test
-this claim with typical values of the parameters. For the human genome,
-seeds of length $\approx 20$ are the norm...
-
-\section{Using the information of the backward search}
-
-Assume that the sequence to map has $N$ additional duplicates, all with a
-uniform divergence equal to $\mu$. We can use the number of hits
-obtained in the backard search to estimate $N$ and $\mu$.
-
-At every stage of the backward search, we obtain the number of sequences
-in the reference genome that match the end of the query. Ignoring
-sequencing errors and focusing only on the $N$ duplicates, the probability
-that the count of duplicates goes from $N_i$ to $N_{i+1}$ at each step of
-the search is given by
+\label{sec:est}
+
+So far we have assumed that we know all the specifications of the problem,
+including the error rate of the instrument $p$, the number of off-target
+sequences $N$, and their rate of sequence divergence $\mu$ (or more
+accurately the probability that they differ from the target at every
+nucleotide).
+
+In general, $p$ is known from the vendor of the sequencing instrument, but
+$N$ and $\mu$ can change for every read and they are usually unknown. We
+thus need an estimation algorithm that can run fast enough that it does
+not slow the mapping process. The main idea here is to use the backward
+search, which is the method used to find the MEM seeds in the first place.
+
+The backward search is a method that was first proposed by Ferragina and
+Mazini\cite{ferragina2000opportunistic} to identify exact matches between
+a text query and an index. The query is extended one letter/nucleotide at
+a time, and at each iteration the search returns the number of matches in
+the index/genome. This is thus a very efficient method to identify MEM
+seeds, because one can keep extending the query until there is no match in
+the genome or until the end of the read.
+
+What matters for us is that we can run the backward search at some
+arbitrary position of the read. The $i$-th iteration returns the number
+$N_i$ of matches in the genome for a query sequence of length $i$.
+Ignoring sequencing errors and focusing only on the $N$ off-target
+sequences, the probability that the count goes from $N_i$ to $N_{i+1}$ is
 
@@
 
-The probability of the sequence $(N_1, N_2, \ldots,
-N_m)$ starting from $N_0$ is
+Assuming that there are $N = N_0$ duplicates and that we run $m$
+iterations of the backward search, the probability of the sequence of
+observations $(N_1, N_2, \ldots, N_m)$ is thus
 
@@
 
-We can now use the maximum likelihood approach to find the value of
-$\mu$ that maximizes this expression. Taking the logarithm and
-differentiating with respect to $\mu$ we find
-
-\begin{gather*}
-\frac{N_0-N_m}{\mu}-\frac{N_1+\ldots+N_m}{1-\mu} = 0
-\text{, which implies} \\
-\mu = \frac{N_0-N_m}{N_0+\ldots+N_{m-1}}.
-\end{gather*}
-
-We also observe that by defining the loss at each step $\Delta_i = N_{i-1}
-- N_i$, the distribution of $(\Delta_1, \ldots, \Delta_m, N_m)$ given
-$N_0$ is multinomial with parameters $\big(\mu, \mu(1-\mu),
-\ldots, \mu(1-\mu)^{m-1}, (1-\mu)^m\big)$. More specifically, the
-probability of observing the sequence $(\Delta_1, \ldots, \Delta_m, N_m)$
-given $N_0$ is
+This is a familiar distribution in disguise: by defining $\Delta_i =
+N_{i-1} - N_i$, the series of observations $(\Delta_1, \ldots, \Delta_m,
+N_m)$ given $N_0$ has a multinomial distribution with parameters
+$\big(\mu, \mu(1-\mu), \ldots, \mu(1-\mu)^{m-1}, (1-\mu)^m\big)$. More
+specifically, the probability of observing the sequence $(\Delta_1,
+\ldots, \Delta_m, N_m)$ given $N_0$ is
 
@@
 
-Now, if $N_0, N_1, \ldots, N_{n-1}$ are missing, we can compute the
-probability of observing $(\Delta_{n+1}, \ldots, \Delta_m, N_m)$ from the
-marginal distribution as
+At that point, we need to bring back into the light the practical
+consideration that short matches are uninformative. This implies that the
+first iterations of the backward search say nothing about the number of
+duplicates of the sequence. In practice, we are better off ignoring the
+output of the first $n-1$ iterations by just discarding them. We observed
+that running 30 iterations of the backward search and discarding the
+output of the first 20 gives good results (\textit{i.e.} we set $n = 21$
+and $m=30$).
+
+Now, if $N_0, N_1, \ldots, N_{n-1}$ are missing, we obtain the probability
+of observing $(\Delta_{n+1}, \ldots, \Delta_m, N_m)$ from the marginal
+distribution as
 
@@
 
-This expression depends on $N_0$, which can be considered a parameter of
-the distribution. To estimate it, we take the logarithm of the probability
-and differentiate with respect to $N_0$. We obtain the two following
-equations
-
-\begin{gather*}
-\Psi(N_0+1)-\Psi(N_0-N_n+1) + \log\big( 1-(1-\mu)^n \big) = 0 \\
+Now replacing again $\Delta_i$ by $N_{i-1}-N_i$, the log-likelihood can be
+expressed as
+
+\begin{equation}
+\begin{split}
+\ell = cst + \log \Gamma(N_0+1) - \log \Gamma(N_0-N_n-1)
++ (N_0-N_n) \log \big(1-(1-\mu)^n\big) + \\
+(N_n-N_m) \log(\mu) + (nN_n + N_{n+1} + \ldots + N_m) \log(1-\mu).
+\end{split}
+\end{equation}
+
+To estimate the parameters by maximum likelihood, we differentiate $\ell$
+with respect to $N_0$ and with respect to $\mu$, then we equate both terms
+to 0. We thus have to solve
+
+\begin{gather}
+\label{eq:psi}
+\Psi(N_0+1)-\Psi(N_0-N_n+1) +
+  \log\big( 1-(1-\mu)^n \big) = 0, \text{ and} \\
+\label{eq:N0}
 \frac{N_n-N_m}{\mu}
@@
 (N_0-N_n)\frac{n(1-\mu)^{n-1}}{1-(1-\mu)^n} = 0.
-\end{gather*}
-
-In the expression above, $\Psi(\cdot)$ is the digamma function, defined as
-the derivative of $\log \Gamma(\cdot)$. We observe that the second
-equation gives an expression of $N_0$ as a function of $\mu$, namely
+\end{gather}
+
+In the expressions above, $\Psi(\cdot)$ is the digamma function, defined
+as the derivative of $\log \Gamma(\cdot)$. We observe that equation
+(\ref{eq:N0}) gives an expression of $N_0$ as a function of $\mu$ and of
+the observations, namely
 
@@
 
-We can substitute this expression in the first equation, obtaining an
-equation depending on $\mu$ only, which can be solved numerically by
-the Newton-Raphson method.
+To find the maximum likelihood estimate of $\mu$, we can substitute this
+expression in equation (\ref{eq:psi}), obtaining an equation depending on
+$\mu$ and on the observations, which can be solved numerically. One can
+then substitute the solution into equation (\ref{eq:N0}) to obtain the
+estimate of the number of duplicate sequences $N_0$.
+
+However, the estimations of $\mu$ and $N_0$ show large deviations from the
+target values on simulated data. In practice, one does not need to know
+the parameters with exquisite precision, so to avoid unrealistic
+estimates, we pick the best $\mu$ from a short list of candidate values
+(\textit{e.g.}, 0.02, 0.04 and 0.06). To do this, we choose a value
+$\mu^*$ from the candidate set, we compute $N_0(\mu^*)$ from equation
+(\ref{eq:N0}) and we evaluate $\ell$ with these values. Finally, we select
+the $\mu$ that gives the higest value of $\ell$, and the estimate for
+$N$ is set to $N_0(\mu)$.
+
+Finding the estimates of $N$ and $\mu$ can be performed in negligible time
+compared to running the backward search. We can repeat this method at
+different areas of the read and finally average the estimates to get
+reasonable values of $N$ and $\mu$ for the whole read. With this
+information, we can compute the probability that the read has no good seed
+using the methods presented above.
 

### 2019-02-08 MEM.tex
First non-draft version.

@@
 \begin{abstract}
-The abstract will come later.
-A popular mapping heuristic is called \emph{seeding}, and a recent variant
-is the \emph{MEM seeding} approach. Here we tackle the question of how to
-evaluate the probability that MEM seeding hits are on-target. This in turn
-expresses the confidence one should have in the location of the mapped DNA
-fragment and can be used to calibrate the MEM seeding heuristic, or to
-filter the hits at the desired level of confidence.
+High throughput sequencing produces massive amounts of data that must be
+processed efficiently. Most reads are identified by mapping them to a
+reference, typically a genome. Modern mappers rely on heuristics to gain
+speed at a reasonable cost of accuracy. MEM seeding is a popular heuristic
+that shows good empirical performance but for which no theory is
+available. It is thus difficult to calibrate or optimize mapping
+strategies based on MEM seeding. Here we develop a theory to predict the
+performance of MEM seeding. More specifically, we design a method to
+compute the probability a read is mapped to the right location when using
+MEM seeds. Our computational strategy is based on the analytic
+combinatorics programme, but uses only part of it. By recoding the reads
+in a specialized combinatorial alphabet, we can implicitly specify their
+generating function and extract the probabilities of interest by direct
+methods or by Monte Carlo sampling. This yields an efficient algorithm to
+calibrate the MEM seeding heuristic for short read technologies.
 \end{abstract}
@@
 
-\section{The computational strategy}
+\section{Computational strategy}
 \label{sec:symbolic}
@@
 \subsection{Monte Carlo sampling}
+\label{sec:montecarlo}
 
@@
 
-The reads are recoded as sequences of symbols from the following alphabet
-$\mathcal{A}_* = \{\square, |, \downarrow_{/1}^-, \downarrow_{/2}^-,
-\ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots, \Downarrow\}$. The
-symbols $\downarrow_{/j}^-$ signify that the nucleotide is a mismatch
-against the off-target sequence, the symbols $\downarrow_{/j}^+$ signify
-that it is a mismatch against the target, and the symbol $\Downarrow$
-signifies that it is a mismatch against both. As before, every other
-nucleotide is replaced by the symbol $\square$, and the terminator $|$
-is appended at the end of the read. We again define reads as sequences of
-segments, except that now the terminators are the symbols
-$\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
+The reads are recoded as sequences of symbols from the so-called ``+/-
+alphabet'' $\mathcal{A}_{+/-} = \{\square, |, \downarrow_{/1}^-,
+\downarrow_{/2}^-, \ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots,
+\Downarrow\}$. The symbols $\downarrow_{/j}^-$ signify that the nucleotide
+is a mismatch against the off-target sequence, the symbols
+$\downarrow_{/j}^+$ signify that it is a mismatch against the target, and
+the symbol $\Downarrow$ signifies that it is a mismatch against both. As
+before, every other nucleotide is replaced by the symbol $\square$, and
+the terminator $|$ is appended at the end of the read. We again define
+reads as sequences of segments, except that now the terminators are the
+symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
 
@@
 \includegraphics[scale=0.88]{sketch_dual.pdf}
-\caption{\textbf{Title.}
-Caption.}
+\caption{\textbf{The +/- encoding.}
+A example read is represented in the +/- alphabet. The symbols
+$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
+the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
+and the symbol $\Downarrow$ represent a mismatch against both. The index
+$j$ is the number of matches since the last mismatch for the other
+sequence. Note the presence of the terminator $|$ at the end of the read.
+}
 \label{fig:dual}
@@
 
-\subsection{The probability of outcome $ii.$}
+\subsection{The probability of false positives}
 
@@
 
-\section{Conclusion}
-
-This is awesome.
+\section{Discussion}
+
+We have devised a method to compute the probability that the MEM seeding
+heuristic misguides the mapping algorithm so that it cannot find the true
+location of the read. The pillars of our strategy are borrowed from
+analytic combinatorics~\cite{flajolet2009analytic,
+sedgewick2013introduction}, even though we do not follow the standard
+programme. Constructing generating functions usually serves the purpose of
+finding their singularities in order to approximate the solution. In our
+case, however, the weighted generating functions cannot be computed so
+this strategy is not applicable.
+
+To find the probabilities of interest in the absence of a fully specified
+weighted generating function, we compute only the first terms using
+iterative methods as in sections~\ref{sec:extract}, \ref{sec:cPE0} and
+\ref{sec:cPE0E1}, or using Monte Carlo sampling as in
+section~\ref{sec:montecarlo}. In this regard, the breakthrough is the
+encoding of reads as MEM segments, which is a form of implicit Markov
+imbedding~\cite{fu1994distribution}.
+
+The method is general, but it is important to clearly state the
+assumptions it depends on. First and most importantly, we have ignored
+insertions and deletions. We assume that the sequencing errors are
+substitutions only, which makes the method adapted to the Illumina
+technology, but not to deletion-prone instruments such as the Oxford
+Nanopore technology. We also assume that insertions and deletions never
+occur among duplicated sequences. This is obviously incorrect, but our
+initial tests with real data suggest that this is a minor impediment. On
+real genomes, the estimations are only approximately correct and this is
+but one of several simplifying assumptions. The theory is substantially
+intrincate; insertions and deletions would make it intractable.
+It is presently unclear how gracefully incorporate them in the theory.
+
+The second assumption is that all the candidate locations with a MEM
+seed of size $\geq \gamma$ are tested with the exact sequence alignment
+algorithm, so that the best location is always discovered if it is in the
+candidate set. This is not completely unrealistic, but it is important to
+realize that for plant and animal genomes, a shared MEM seed may have tens
+of thousands of hits. Therefore, most mappers impose a limit on the number
+of alignments so the best hit may not be found even if it is in the
+candidate set. This is again a minor impediment, since we can correct the
+probability that the location is a false positive when only some
+candidates have been tested.
+
+The third assumption is that all the duplicate sequences evolve
+independently of each other and at the same rate. This is again incorrect,
+because duplication events can happen continuously, creating complex
+ancestry relationships. It is possible to infer the ancestry using tree
+reconstruction techniques, but it would be challenging to incorporate this
+information in the present theory. The symbols of the MEM alphabet
+implicitly assume that the sequences are exchangeable and the complexity
+of the calculations explodes if it is not the case.
+
+The last assumption is that MEM seeds can match only the target or one of
+its duplicates. This is not the case: MEM seeds overlapping a sequencing
+error often match random locations of the genome. When this occurs, the
+MEM seeds are typically small and the alignment score at the candidate
+location is very bad because the sequences are unrelated. So in practice,
+it is relatively easy to detect this case and reject the location, even if
+it is the best in the candidate set.
+
+The proposed methods are compliant with the demand for speed of the
+mapping algorithms. One needs to compute the probabilities only once for a
+given value of $N$ and $\mu$ (the error rate $p$ is known and constant).
+For $N > 20$, the iterative method is usually prohibitive and we need to
+use Monte Carlo sampling instead, which for reads of size $< 200$
+nucleotides keeps the computations around 1 second on modern hardware. The
+values of $N$ and $\mu$ can be binned in intervals of interest so that
+there are only $\approx 100$ pairs for a total cost of a few minutes per
+run. Considering that mappers can process at most $10,000$ reads per
+second, the computer time of mapping a sequencing run of 250 million reads
+is over 7 hours, two orders of magnitude larger than the time required to
+estimate the probabilities of error.
+
+In conclusion, we presented a practical solution to the problem of
+estimating the probability of false positives when using MEM seeds. This
+solution is adapted for mapping short reads sequenced with the Illumina
+technology. Being able to calibrate the MEM seeding heuristic not only
+allows the user to choose how to balance speed versus accuracy, but also
+opens new applications. For instance, one can map reads from contaminated
+samples in pools of closely related genomes (\textit{e.g.}, modern human
+and Neanderthal) in order to assign the reads to the organism they belong
+to. The probabilities of false positives give the right level of
+confidence in the assignment.
+
+More generally, the analytic combinatorics programme is a very powerful
+tool to address problems in bioinformatics. Here we have seen how this
+strategy can be useful even when the generating functions cannot be
+computed. Using the same ideas, one could calibrate heuristics used in
+other alignment methods, especially in expanding fields such as long-read
+technologies.
 
@@
 
-We acknowledge the financial support of the Spanish Ministry of Economy
-and Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan
-Nacional BFU2012-37168), of the CERCA Programme~/~Generalitat de
-Catalunya, and of the European Research Council (Synergy Grant 609989).
+We thank Ruggero Cortini for suggesting Monte Carlo methods as a way to
+tackle intractable cases. We acknowledge the financial support of the
+Spanish Ministry of Economy and Competitiveness (‘Centro de Excelencia
+Severo Ochoa 2013-2017’, Plan Nacional BFU2012-37168), of the CERCA
+Programme~/~Generalitat de Catalunya, and of the European Research Council
+(Synergy Grant 609989).
 

### 2019-02-13 MEM.tex
Added section about practical considerations.

@@
 The focus of this document is not to solve the true location problem
-\textit{per se}, but rather to evaluate whether a solution is correct. At
-first sight, whether we can map reads to their true location seems to
-depend only on their lengths and on the error rate of the instrument.
-However, there is another important factor to consider: most genomes
-contain repeated sequences, \textit{i.e.} relatively large subsequences
-(from $\sim20$~bp to more than 10~kb) that are present at more than one
-location. So we cannot map the fragment with certainty if there exists an
-exact copy of the sequence at somewhere else in the genome, because we
-cannot know which of the two copies was sequenced.
+\textit{per se}, but rather to evaluate whether a given solution is
+correct. At first sight, whether we can map a read to its true location
+seems to depend only on its length and on the error rate of the
+instrument. However, there is another important factor to consider: most
+genomes contain repeated sequences, \textit{i.e.} relatively large
+subsequences that are present at more than one location. So we cannot map
+the fragment with certainty if there exists an exact copy of the sequence
+somewhere else in the genome, because we cannot know which of the two
+copies was sequenced.
 
@@
 \emph{best} (as measured by the identity between the sequence and the
-candidate location). This leads us to consider the following question: how
-can we identify the \emph{best} location of the fragment in the genome?
-
-This is the \emph{best location problem}. It ammounts to finding the
-optimal alignment between two sequences, and for this reason has received
-substantial attention in bioinformatics. It is relevant for the present
-discussion because the candidate solution of the \emph{true} location
-problem is always the \emph{best} location. The next section will further
-clarify the relationship between the two problems.
+candidate location). This leads to the following question: how can we
+identify the \emph{best} location of the fragment in the genome?
+
+We will refer to this as the \emph{best location problem}. It amounts to
+finding the optimal alignment between two sequences, and for this reason
+has received substantial attention in bioinformatics. The distinction is
+relevant for the present discussion because the candidate solution of the
+\emph{true} location problem can only be the \emph{best} location. So in
+order to solve the true location problem, we must first solve the best
+location problem.
 
@@
 thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
-best location, $ii.$ the DNA fragment is mapped to a suboptimal location,
+best location, $ii.$ the DNA fragment is mapped to another location,
 $iii.$ the DNA fragment is not mapped at all.
@@
 If seeds can be discovered fast, we can quickly zoom in on a small set of
-candidate locations, and use slower exact algorithms on a reduced data
-set. The disadvantage is that the target location is not always in the
-candidate set. In this case, it cannot be discovered and the outcome is
-either a false positive or a false negative.
+candidate locations and reduce the input size of the exact but slow
+algorithm. The disadvantage is that the target location is not always in
+the candidate set. In this case, it cannot be discovered and the outcome
+is either a false positive or a false negative.
 
@@
 if the seeding step is set to produce few candidates, the process will
-terminate faster but the target is less likely to sit among the candidates
-so it is less likely to be discovered.
+run faster but the target is less likely to be in the candidate list so it
+is less likely to be discovered.
 
@@
 This framework was developed for the best location problem, but it also
-holds for the true location problem. Actually, the process is exactly the
+applies to the true location problem. Actually, the process is exactly the
 same but the outcomes $i.$ and $ii.$ are different. The first, called a
@@
 somewhere else in the genome. A true positive thus requires that the
-target (the true location) is in the candidate set, and that it is the
-optimum.
+target (the true location) is in the candidate set, and that it has the
+best score. For now, we will assume that the true location is always the
+best and we defer the discussion about this assumption until
+section~\ref{sec:not_the_best}.
 
@@
 instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
-seeds'' that tolerate mismatches\footnote{The method is actually covered
-by a patent (reference US20070088510A1).}. However, in this document, we
-will always assume that a seed is an exact match with no gap. In other
-words, a seed is a perfect match between a part of the read and a genomic
-site.
+seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
+covered by a patent (reference US20070088510A1).}. However, in this
+document, we will always assume that a seed is an exact match with no gap.
+In other words, a seed is a perfect match between a part of the read and a
+genomic site.
 
@@
 match the original DNA sequence and/or its duplicates, but not the rest of
-the genome. This will simplify the discussion without loss of generality
-because non homologous matches are short and they can be ignored.
+the genome. This will simplify the discussion and we defer the discussion
+of this assumption until section~\ref{sec:short_MEMs}
 
@@
 
+\section{Neglected practical considerations}
+
+To keep the focus on the MEM seeding process, we have ignored some
+practicalities that we now have to revisit in order to evaluate the
+feasibility of our approach.
+
+\subsection{Is the true location the best one?}
+\label{sec:not_the_best}
+
+So far we have assumed that the true location is also the best one, so
+that it is always discovered when it is in the candidate set. But how
+likely is it that the true location is not the best one, and thus that it
+is virtually impossible to discover?
+
+Obviously, it is possible that the target has an exact duplicate. When
+that happens, the read cannot be mapped unambiguously, but this case is
+relatively easy to detect, so the mapper can return a low confidence score
+and thus avoid making an error.
+
+From now we assume that the read has no strict duplicate. The probability
+that the read of size $k$ has $m$ errors is
+\begin{equation*}
+{k \choose m}p^m(1-p)^{k-m}.
+\end{equation*} 
+A given duplicate has a better score if its Hamming distance from the read
+is less than $k$. In a first order approximation, this happens if the
+duplicate matches one of the $k$ errors, and also the $L-k$ correct
+nucleotides. The probability of this even is
+\begin{equation*}
+m(1-\mu)^{k-m}\mu/3.
+\end{equation*}
+
+There are $N$ duplicates that evolve independently, so the probability
+that one of them has a better score than the target is
+\begin{equation*}
+1-\left(1-m(1-\mu)^{k-m}\mu/3\right)^N \approx Nm(1-\mu)^{k-m}\mu/3.
+\end{equation*}
+
+Combining the first and the last expression, we find that the probability
+that the target does not have the best score is approximately
+\begin{equation}
+\label{eq:not_the_best}
+\sum_{m=0}^k {k \choose m}p^m(1-p)^{k-m} Nm\mu/3(1-\mu)^{k-m} = 
+Nk p \left(1-\mu(1-p)\right)^{k-1}\mu/3.
+\end{equation}
+
+For instance, for reads of size $k=100$ with error rate $p=0.01$, where
+the target has $N=5$ duplicates with divergence $\mu=0.05$ we obtain an
+estimate close to $10^{-4}$. In comparison, the probability of a false
+positive when using MEM seeds with a minimum size $\gamma=17$ is
+approximately $4.4\cdot10^{-4}$.
+
+The probability that the target does not have the best score is usually
+small, but it is of the same order of magnitude as the probabilities of
+false positives with MEM seeds. The reason is that a common source of
+false positives are the cases when the target does not have the best
+score. Indeed, in this case the target is much less likely to have a MEM
+seed.
+
+In short, the event that the true location is not the best one is already
+incorporated in the false positive rate and we do not recommend further
+updating the estimates. In the case that one is interested in this event
+\textit{per se}, formula (\ref{eq:not_the_best}) gives an approximate
+probability of its occurrence. When $\mu$ is unknown, we can replace it by
+its worst case value $\mu = 1/qL$, where $q=1-p$. But $N$ must be known at
+least approximately because the proability increases with $N$.
+
+In summary, the case that the true location is not the best one has a
+small but not negligible probability. It is partly redundant with the
+false positives generated during the MEM seeding process, and therefore it
+is not straightforward to adjust the estimates accordingly. We do not
+adjust them at all and consider that this case is already implicitly
+accounted for.
+
+\subsection{MEM seeds at random locations}
+\label{sec:short_MEMs}
+
+To simplify the problem, we have assumed that MEM seeds can match only the
+target or one of its duplicates. In practice, MEM seeds that overlap a
+sequencing error often match random locations of the genome instead of the
+duplicates. When this occurs, the MEM seeds are typically small because 
+the match is random.
+
+We select MEM seeds of size at least $\gamma$, and it is important for the
+validity of the theory that all the MEM seeds above the threshold are
+considered candidate locations. If not, the estimates of the false
+positive rate are off. The hope is that spurious MEM seeds are shorter
+than $\gamma$ so that they are invisible, but this is not always the case.
+And if they qualify, we need to verify them.
+
+Such random hits can be problematic for two reasons: First, the time spent
+verifying them is wasted. Second, they can cause false positives.
+Fortunately, both issues can be addressed so that they do not constitute a
+major problem.
+
+To avoid spending too much time on small random MEM seeds, we can
+prioritize the seeds so that the best hit is likely to be discovered first
+and bail out from the alignments as soon as it appears that their score
+will be worse than the best hit so far. We can even put an upper bound on
+the quality of candidate so that the alignment can be skipped altogether.
+For instance, if we already have a candidate at distance 1 from the read,
+there is no need to verify candidates from MEM seeds that do not reach one
+end of the read. Indeed, the hit must be at distance at least 2 because
+there is a mismatch before the MEM seed and another one after it.
+
+These considerations depend on the implementation of the mapper so we will
+not further develop them. What matters is that random hits do not have to
+impose a significant burden on the time needed to verify the candidates.
+
+The second issue is that random hits can generate false positives. Such
+false positives happen when the read does not belong to the reference
+genome, or when it contains so many errors that no proper MEM seed was
+found. Since the hit is not homologous to the read, the alignment score is
+essentially random. The upshot is that it is easy to detect.
+
+Indeed, if the candidate location is random, mismatches occur with
+probability 3/4. If it is the right location, mismatches occur with
+probability $p$. We discard the MEM seed, which is an automatic match of
+size at least $\gamma$. Say that there remains $L$ nucleotides and that
+$m$ of them are mismatches for the candidate location. Using Bayes
+formula, we find that the probability that the location is random given
+the observed alignment score is
+\begin{equation}
+\label{eq:bayes}
+\frac{1}{1 + 4^L(p/3)^mq^{L-m}(1-\beta)/\beta},
+\end{equation}
+where $q=1-p$ and $\beta$ is the prior probability that the hit is
+random.
+
+One may have information about the level of contamination, allowing to
+choose a meaningful value for $\beta$. For instance, sequencing runs are
+often spiked by foreign DNA such as the phiX genome or other. If these
+reads are not filtered automatically, $\beta$ can be set to their expected
+frequency. For sequences that belong to the reference genome, random hits
+are rather rare because the read must contain so many error that neither
+the target nor its duplicates were discovered. The value of $\beta$ is
+much smaller in this case, but nevertheless difficult to estimate.
+
+That being said, the value of $\beta$ has little importance in practice.
+Say that a read generates a hit in the genome such that 33 nucleotides
+need to be aligned (the alignment status of the others was already known
+from the seeding step). If the hit is random there is a 99.99\% chance
+that at least 15 are mismatches. If $p = 0.01$, the denominator of
+expression (\ref{eq:bayes}) is approximately $1 +
+4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
+support for the hypothesis that the hit is random is overwhelming.
+
+Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
+less are mismathes. In this case, the denominator is approximately
+$1+6.8\cdot10^9(1-\beta)/\beta$, so unless $\beta > 1-10^9$, the support
+for the hypothesis that the hit is not random is overwhelming.
+
+In general, one does not need to worry about the value of $\beta$ because
+random hits are extremely visible when $p$ is small. For instance, one can
+choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears to
+simplify expression (\ref{eq:bayes}). We compute this expression for every
+read and add the result to the estimated probability of false positive for
+the read. Either the term is negligible and it does not affect the
+estimated probability, or it is strongly dominant and yields a very strong
+probability that the location is incorrect, regardless of the estimate
+from the MEM seeding process.
+
+In summary, MEM seeds in random sequences happen relatively frequently,
+but it is possible to minimize their computational burden. Also, they
+are no cause for concern regarding false positives because they can easily
+be detected using Bayes formula.
+
 \section{Discussion}
@@
 The last assumption is that MEM seeds can match only the target or one of
-its duplicates. This is not the case: MEM seeds overlapping a sequencing
-error often match random locations of the genome. When this occurs, the
-MEM seeds are typically small and the alignment score at the candidate
-location is very bad because the sequences are unrelated. So in practice,
-it is relatively easy to detect this case and reject the location, even if
-it is the best in the candidate set.
+its duplicates. This is not the case but we have shown how to gracefully
+deal with this in section~\ref{sec:short_MEMs}.
+
+The Monte Carlo approach of algorithm~\ref{alg:mcmc} is relatively
+straightforward, so one may wonder why the approach with weighted
+generating functions is necessary at all. The only reason is precision. To
+estimate the frequency of an event by Monte Carlo sampling, it must occur
+at least a few times in the simulation. For instance, with 10 million
+rounds of sampling, frequencies around $1/1,000,000$ or lower cannot be
+measured accurately. When not interested in rare events, it is thus a
+perfectly reasonable strategy to use only Monte Carlo sampling.
+
+For $N < 20$, the probability that a read does not contain a good seed is
+the smallest, so we need a method that is accurate in this range.
+Fortunately, the transfer matrix method is fast in this regime because the
+matrix $M(z)$ has few rows/columns.
 
@@
 For $N > 20$, the iterative method is usually prohibitive and we need to
-use Monte Carlo sampling instead, which for reads of size $< 200$
-nucleotides keeps the computations around 1 second on modern hardware. The
-values of $N$ and $\mu$ can be binned in intervals of interest so that
-there are only $\approx 100$ pairs for a total cost of a few minutes per
-run. Considering that mappers can process at most $10,000$ reads per
-second, the computer time of mapping a sequencing run of 250 million reads
-is over 7 hours, two orders of magnitude larger than the time required to
-estimate the probabilities of error.
+use Monte Carlo sampling instead. The running time depends on the desired
+number of iterations, on $p$ and on the size of the reads $k$. Since those
+are constant throughout the sequencing run, the method always takes the
+same time (around 1 second for reads of size $100$-$200$ nucleotides on
+modern hardware). The values of $N$ and $\mu$ can be binned in intervals
+of interest so that there are only around $100$ pairs for a total cost of
+a few minutes per run. Considering that mappers can process at most
+$10,000$ reads per second per core, the time of mapping a sequencing run
+of 250 million reads is over 7 hours, two orders of magnitude larger than
+the time required to estimate the probabilities of error.
 

### 2019-02-15 MEM.tex
Optimizing the figures and the text.

@@
 reference, typically a genome. Modern mappers rely on heuristics to gain
-speed at a reasonable cost of accuracy. MEM seeding is a popular heuristic
-that shows good empirical performance but for which no theory is
-available. It is thus difficult to calibrate or optimize mapping
-strategies based on MEM seeding. Here we develop a theory to predict the
-performance of MEM seeding. More specifically, we design a method to
-compute the probability a read is mapped to the right location when using
-MEM seeds. Our computational strategy is based on the analytic
-combinatorics programme, but uses only part of it. By recoding the reads
-in a specialized combinatorial alphabet, we can implicitly specify their
-generating function and extract the probabilities of interest by direct
-methods or by Monte Carlo sampling. This yields an efficient algorithm to
-calibrate the MEM seeding heuristic for short read technologies.
+speed at a reasonable cost of accuracy. MEM seeding (where MEM stands for
+Maximal Exact Match) is a popular heuristic that shows good empirical
+performance but for which no theory is available. It is thus difficult to
+calibrate or optimize mapping strategies based on MEM seeding. Here we
+develop a theory to predict the performance of MEM seeding. More
+specifically, we design a method to compute the probability a read is
+mapped to the right location when using MEM seeds. Our computational
+strategy is based on the analytic combinatorics programme, but uses only
+part of it. By recoding the reads in a specialized combinatorial alphabet,
+we can implicitly specify their generating function and extract the
+probabilities of interest by direct methods or by Monte Carlo sampling.
+This yields an efficient algorithm to calibrate the MEM seeding heuristic
+for short read technologies.
 \end{abstract}
@@
 Suppose that a DNA fragment drawn at random from a known
-genome\footnote{We will assume throughout that the sequence of the genome
-is known exactly and with absolute certainty. This is not the case in
-practical application, but this assumption will allow us to develop a
-theory centered on the read.} is sequenced with an imperfect instrument;
-how can we identify the location of the fragment in the genome?
+genome\footnote{We will assume that the sequence of the genome is known
+exactly and with absolute certainty. This is not the case in practical
+application, but this assumption will allow us to develop a theory
+centered on the read.} is sequenced with an imperfect instrument; how can
+we identify the location of the fragment in the genome?
 
 We will refer to this question as the \emph{true location problem}.
-It is central to countless applications, including discovering
+It has applications in countless domains, like discovering
 disease-causing mutations; diagnosing viral, bacterial or fungal
@@
 seems to depend only on its length and on the error rate of the
-instrument. However, there is another important factor to consider: most
+sequencer. However, there is another important factor to consider: most
 genomes contain repeated sequences, \textit{i.e.} relatively large
-subsequences that are present at more than one location. So we cannot map
-the fragment with certainty if there exists an exact copy of the sequence
-somewhere else in the genome, because we cannot know which of the two
-copies was sequenced.
+subsequences that are present at more than one location. For instance, we
+cannot map the fragment with certainty if there exists an exact copy of
+the sequence somewhere else in the genome, because we cannot know which of
+the two copies was sequenced.
 
@@
 uniformly at random is very unlikely to produce sequences with the
-observed degree of identity.}. So when the DNA fragment is duplicated, we
-can map it with high confidence provided we can distinguish it from its
-duplicates. This in turn depends on their similarity with the target
-fragment and on the error rate of the sequencer.
+observed degree of identity, meaning that the similarity is due some
+biological duplication process.}. So when the DNA fragment is duplicated,
+we can still map it, as long as we can distinguish it from its duplicates.
+This in turn depends on their similarity with the target fragment and on
+the error rate of the sequencer.
 
@@
 
-We will refer to this as the \emph{best location problem}. It amounts to
-finding the optimal alignment between two sequences, and for this reason
-has received substantial attention in bioinformatics. The distinction is
-relevant for the present discussion because the candidate solution of the
-\emph{true} location problem can only be the \emph{best} location. So in
-order to solve the true location problem, we must first solve the best
-location problem.
+We will refer to this task as the \emph{best location problem}. It amounts
+to finding the optimal alignment between two sequences, and for this
+reason has received substantial attention in bioinformatics. Finding the
+best location of the read is the only way to get to its true location, so
+we need to focus our attention on the best location problem.
 
@@
 thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
-best location, $ii.$ the DNA fragment is mapped to another location,
+best location, $ii.$ the DNA fragment is mapped to some other location,
 $iii.$ the DNA fragment is not mapped at all.
@@
 If seeds can be discovered fast, we can quickly zoom in on a small set of
-candidate locations and reduce the input size of the exact but slow
+candidate locations and reduce the input size of the (slow) alignment
 algorithm. The disadvantage is that the target location is not always in
 the candidate set. In this case, it cannot be discovered and the outcome
-is either a false positive or a false negative.
-
-This imposes a trade-off between speed and accuracy between each candidate
+is either $ii.$ or $iii.$
+
+This imposes a trade-off between speed and accuracy because each candidate
 location must be tested with a slow algorithm. If the seeding step is set
@@
 
-Importantly, the terms of this trade-off are not set in stone. Some
+Importantly, the terms of this trade-off depend on the algorithm. Some
 seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
@@
 
-This framework was developed for the best location problem, but it also
-applies to the true location problem. Actually, the process is exactly the
-same but the outcomes $i.$ and $ii.$ are different. The first, called a
-``true positive'' or an ``on-target hit'', is that the DNA fragment is
-mapped to the \emph{true} location; the second, called a ``false
-positive'' or an ``off-target hit'' is that the DNA fragment is mapped
-somewhere else in the genome. A true positive thus requires that the
-target (the true location) is in the candidate set, and that it has the
-best score. For now, we will assume that the true location is always the
-best and we defer the discussion about this assumption until
-section~\ref{sec:not_the_best}.
+The seeding framework was developed for the best location problem, but it
+also applies to the true location problem. Actually, the process is
+exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
+location. Outcome $i.$, called a ``true positive'', is that the DNA
+fragment is mapped to the \emph{true} location. Outcome $ii.$, called a
+``false positive'' is that the DNA fragment is mapped to an incorrect
+location. Outcome $iii.$, called a ``false negative'', does not change and
+still means that the DNA fragment is not mapped at all.
+
+We will make two assumptions that will simplify the discussion. The first
+is that the true location is always the best. We will further comment on
+this and explain how to relax this assumption in section
+section~\ref{sec:not_the_best}. Until then, we consider that the true
+location problem is the same as the best location problem. The second
+assumption is that the best location is always discovered if it is in the
+candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
+determined by the seeding process. For this reason, we will sometimes
+refer to them as the outcomes of the seeding step, even though they are
+the outcomes of the whole mapping process.
 
@@
 The rest of the document will focus on a seeding heuristic known as ``MEM
-seeding''. This approach gives good empirical results when mapping short
-reads (it is for instance the strategy\footnote{To be exact, BWA-MEM uses
-SMEM seeds (Super Maximal Exact Match seeds).} of the popular mapper
-BWA-MEM\cite{li2013aligning}) but there is presently no theory to justify
-this performance. A deeper understanding of MEM seeding would thus be
-useful to calibrate the speed/accuracy trade-off of this heurisitic.
-
-Before developing a formal theory, let us clarify the meaning of the MEM
-acronym.
+seeding'' (where MEM stands for Maximal Exact Match). This approach gives
+good empirical results when mapping short reads (it is for instance the
+strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
+Exact Match seeds).} of the popular mapper BWA-MEM\cite{li2013aligning})
+but there is presently no theory to justify its performance. A deeper
+understanding of MEM seeding would be useful to calibrate the
+speed/accuracy trade-off of this heurisitic.
+
+Before developing a formal theory, let us explain in more detail what
+Maximal Exact Matches are and how MEM seeding works.
 
@@
 This definition presents a computational challenge: to know if a given
-subsequence is a MEM seed, we need to know if it exists somewhere in the
+subsequence is a MEM seed we need to know if it exists somewhere in the
 genome. This is a non trivial problem in itself, but fortunately there
-exists practical methods to identify all the MEM seeds from a read, even
-for very large genomes\cite{ferragina2000opportunistic,
+exists practical methods to identify MEM seeds, even when the reference
+genome is very large\cite{ferragina2000opportunistic,
 ferragina2005indexing}. Such algorithms are crucial for the present
@@
 literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
-li2013aligning}, and here on we will assume that MEM seeds are simply
-available at all times. Looking ahead to section~\ref{sec:est}, let
-us just say that this can be done with an algorithm called the
-\emph{backward search}\cite{ferragina2000opportunistic}.
+li2013aligning}, and here on we will assume that MEM seeds are available
+at all times. Looking ahead to section~\ref{sec:est}, let us just mention
+that this can be done with an algorithm called the \emph{backward
+search}\cite{ferragina2000opportunistic}.
 
@@
 12-mers are present in the reference human genome, so an exact match of
-size 12 or lower says nothing about the location of the DNA fragment.
-In what follows, we will assume that subsequences of the read can only
-match the original DNA sequence and/or its duplicates, but not the rest of
-the genome. This will simplify the discussion and we defer the discussion
-of this assumption until section~\ref{sec:short_MEMs}
-
-Fig.~\ref{fig:MEM_example} shows a concrete example. The target (the true
-location of the read) is first potential location. The locations below
-correspond to duplicate sequences. The mismatches between the genome and
-the read are highlighted in bold. The read has three errors indicated by
-stars, so \texttt{T}, \texttt{T} and \texttt{C} should actually be
-\texttt{A}, \texttt{A} and \texttt{G}.
+size 12 or lower says nothing about the location of the DNA fragment. In
+what follows, we will assume that subsequences of the read can only match
+the original DNA sequence and/or its duplicates, but not the rest of the
+genome. This will greatly simplify the exposition of the theory. We will
+discuss how to relax this assumption in section~\ref{sec:short_MEMs}.
+
+Fig.~\ref{fig:MEM_example} shows a concrete example where a read is
+aligned to candidate locations. The target (the true location of the
+read) is first potential location. The locations below correspond to
+duplicate sequences. The mismatches between the genome and the read are
+highlighted in bold. The read has three errors indicated by stars, so
+\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
+\texttt{A} and \texttt{G} (\textit{i.e.} as in the target sequence).
 
@@
 \includegraphics[scale=1]{MEM_example.pdf}
-\caption{\textbf{A concrete example of MEM seeding.}
+\caption{\textbf{Example of MEM seeding.}
 A DNA fragment is sequenced with an imperfect instrument. It comes from
@@
 Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
-on-target MEM seed of size $\gamma$ or greater. Every other MEM seed is a
-\emph{bad} seed. In other words, the true location is in the candidate set
-if and only if the read has a good seed.
+on-target MEM seed of size $\gamma$ or greater. In other words, the true
+location is in the candidate set if and only if the read contains a
+good seed.
 \end{definition}
@@
 Compared to seeds of fixed size, MEM seeds have at least two
-counter-intuitive properties. The first is that there are cases where
-there is no good seed, even when $\gamma$ is decreased.
-Fig.~\ref{fig:full_masking_example} shows such an example. Even though
-there is a single sequencing error, the read does not have any MEM seed
-matching the target. Lowering $\gamma$ does not change this, so there is
-no way to discover the true location (even though it is also the best
-location).
+counter-intuitive (and undesirable) properties. The first is that there
+are cases where there cannot be any good seed, even when changing
+$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
+though there is a single sequencing error, the read has no good seed.
+Lowering $\gamma$ does not change this, so there is no way to discover the
+true location (even though it is the best location).
 
@@
 \includegraphics[scale=1]{full_masking_example.pdf}
-\caption{\textbf{MEM masking case 1.}
-The read, the MEM seeds and the sequences are as in
+\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
+The read, the MEM seeds and the sequences are represented as in
 Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
-locations at the bottom effectively mask the target, so it cannot
-be discovered. MEM masking can occur even when the true location is the
+locations at the bottom effectively hide the target, so it cannot
+be discovered. This can occur even when the true location is the
 best and when there is a single error on the read.}
@@
 
-The second counter-intuitive property of MEM seeding is that there are
-cases where there would be a good seed if the read were shorter.
-Fig.~\ref{fig:short_vs_long} shows an example of this case. Again, there
-is no on-target MEM seed, but there would be if the read were two
-nucleotides shorter on the right side. Indeed, in this case there would be
-a shared MEM seed matching the first and the second locations. It would
-also be a good seed for $\gamma \leq 12$.
+The second counter-intuitive property of MEM seeding is that shortening
+a read can sometimes create a good seed. Fig.~\ref{fig:short_vs_long}
+shows an example of this case. Again, there is no good seed, but there
+would be one if the read were two nucleotides shorter on the right side.
+Indeed, in this case there would be a shared MEM seed matching the first
+and the second locations, provided $\gamma \leq 12$.
 
@@
 \includegraphics[scale=1]{short_vs_long_example.pdf}
-\caption{\textbf{MEM masking case 2.}
-The read, the MEM seeds and the sequences are as in
-Fig.~\ref{fig:MEM_example}. There is no on-target MEM seed, but there
-would be a shared MEM seed if the read were two nucleotides shorter. The
-true location has been masked by sequencing the last two nucleotides.}
+\caption{\textbf{Issues with MEM seeds: too long reads.}
+The read, the MEM seeds and the sequences are represented as in
+Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
+read were two nucleotides shorter. The true location is hidden by
+the last two nucleotides.}
 \label{fig:short_vs_long}
@@
 These examples show that MEM seeds can perform worse than seeds of fixed
-size, again raising the question why they seem to give good performance in
-practice. The theory developed below will allow us to compute the
-probability that a read has no good seed and thus that the true location
-is missed. This probability is the key to evaluating and calibrating the
-MEM seeding heuristic.
-
-From now we will make the simplifying assumption that a read with a good
-seed is always mapped at the correct location. In other words, we assume
-that a good seed is necessary and sufficient for a true positive or
-outcome $i.$ In practice this is not always the case, but this assumption
-shifts the burden of mapping to the alignment algorithm and allows us
-focus on improving the seeding heuristic.
-
-We first explain a detailed strategy to compute the probablity of outcome
-$i.$ We then explain how to compute the probability of outcome $iii.$,
-\textit{i.e.} that the seeding step returns no hit. Combining these two
-quantities allows us to compute the probability of interest, which is that
-the read is mapped to an incorrect location or outcome $ii.$ Finally, we
-provide a computationally efficient method for estimating the parameters
-that are necessary to estimate the probabilities of interest.
-
-Before going further, we recommend the readers that are not already
-familiar with MEM seeding to take a moment to study the examples above.
-The complexity of the next sections will go \textit{crescendo} and they
-will draw heavily from the concepts and definitions above.
+size. MEM seeds yield fewer candidates and therefore speed up the mapping
+process, but the question is at what cost for accuracy? The theory
+developed below will allow us to compute the probability that a read has
+no good seed and thus that the true location is missed. This probability
+is the key to evaluating and calibrating the MEM seeding heuristic.
+
+We first expose a strategy to compute the probablity of true positives
+(outcome $i.$), then show how to compute the probability of false
+negatives (outcome $iii.$), and combine these two quantities to compute
+the probability of false positives (outcome $ii.$). Finally, we provide a
+computationally efficient method for estimating the parameters that are
+necessary to estimate the probabilities of interest.
 
@@
 since the last mismatch on the left, or since the beginning of the read.
-For instance, in Fig.~\ref{fig:MEM_example}, the match streak of the
-target at the ninth nucleotide is $1$ while for the second potential
-location, it is $7$.
 
 \begin{definition}
-At a given position of the read, an off-target location is a \emph{hard
+At a given position of the read, an off-target sequence is a \emph{hard
 mask} if its success streak is strictly longer than the match streak of
-the target. An off-target location is a \emph{soft mask} if its has the
+the target. An off-target sequence is a \emph{soft mask} if its has the
 same match streak as the target.
@@
 
-For instance, the second location in the example above is a hard mask at
-the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further see
-that at the second nucleotide, all the off-target locations are soft masks
-and that at the last nucleotide, the second location is a soft mask.
-
-From the definition, the last nucleotide of every strict on-target MEM
-seed is always unmasked. Conversely, an unmasked nucleotide always belongs
-to exactly one strict on-target MEM seed. Also, the last nucleotide of
-every shared on-target MEM seed is always soft-masked, but a soft-masked
-nucleotide does not always belong to a shared on-target MEM seed.
+Fig.~\ref{fig:hard_vs_soft_masks} illustrates hard and soft masks with an
+example. It is important to bear in mind that hard and soft masks depend
+on the position of interest: a sequence can be a mask at the left end of
+the read but not at the right end, or \textit{vice versa}. In the example
+of Fig.~\ref{fig:full_masking_example}, every nucleotide of the read has a
+hard mask, which is why the target cannot be discovered. In the example of
+Fig.~\ref{fig:short_vs_long}, the middle sequence is a soft mask on the
+right half of the read, but it turns into a hard mask two nucleotides
+before the right end, which is why the target could be discovered if the
+read were shorter.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.85]{hard_vs_soft_masks.pdf}
+\caption{\textbf{Example of hard and soft masks.}
+The nucleotides that do not match the read are reprsented as black boxes.
+The nucleotides that contribute to the match streak are represented as
+grey boxes. The remaining nucleotides are represented as open boxes. At
+the given position, the match streak of the target is 7. The first
+off-target is a hard mask because its match streak is $13>7$. The second
+off-target sequence is a soft mask because its match streak is 7 like the
+target. The third off-target sequence is not a mask because its match
+streak is $2<7$.}
+\label{fig:hard_vs_soft_masks}
+\end{figure}
+
+From the definition, we see that the last nucleotide of every strict
+on-target MEM seed is always unmasked. Conversely, an unmasked nucleotide
+always belongs to exactly one strict on-target MEM seed. Also, the last
+nucleotide of every shared on-target MEM seed is always soft-masked, but a
+soft-masked nucleotide does not always belong to a shared on-target MEM
+seed.
 
@@
 \includegraphics[scale=.85]{sketch_extended.pdf}
-\caption{\textbf{The extended MEM encoding}.
+\caption{\textbf{The MEM encoding}.
 The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
@@
 \centering
-\includegraphics[scale=0.88]{sketch_dual.pdf}
+\includegraphics[scale=0.85]{sketch_dual.pdf}
 \caption{\textbf{The +/- encoding.}
@@
 (1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
-symbols $\downarrow_{/j}^-$ is $b = p\mu/3$, that of the symbols 
-$\downarrow_{/j}^+$ is $c = (1-p)\mu$, and the probability of occurrence
-of the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
+symbols $\downarrow_{/j}^-$ is $b = (1-p)\mu$, that of the symbols
+$\downarrow_{/j}^+$ is $c = p\mu/3$, and the probability of occurrence of
+the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
 $a+b+c+d=1$.
@@
 Assuming that there are $N = N_0$ duplicates and that we run $m$
-iterations of the backward search, the probability of the sequence of
-observations $(N_1, N_2, \ldots, N_m)$ is thus
+iterations of the backward search, the probability of the observations
+$(N_1, N_2, \ldots, N_m)$ is thus
 
@@
 
-To avoid spending too much time on small random MEM seeds, we can
-prioritize the seeds so that the best hit is likely to be discovered first
-and bail out from the alignments as soon as it appears that their score
-will be worse than the best hit so far. We can even put an upper bound on
-the quality of candidate so that the alignment can be skipped altogether.
-For instance, if we already have a candidate at distance 1 from the read,
-there is no need to verify candidates from MEM seeds that do not reach one
-end of the read. Indeed, the hit must be at distance at least 2 because
-there is a mismatch before the MEM seed and another one after it.
+To save time during the alignment phase, we can prioritize the seeds so
+that the best hit is likely to be discovered first, allowing us to bail
+out from other alignments as early as possible. We can even bound
+the maximum quality of some candidates so that the alignment can be
+skipped altogether. For instance, if we already have a candidate at
+distance 1 from the read, there is no need to verify candidates from MEM
+seeds in the middle of the read. Indeed, the hit must be at distance at
+least 2 because there is a mismatch before the MEM seed and another one
+after it.
 
@@
 random hits are extremely visible when $p$ is small. For instance, one can
-choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears to
-simplify expression (\ref{eq:bayes}). We compute this expression for every
-read and add the result to the estimated probability of false positive for
-the read. Either the term is negligible and it does not affect the
-estimated probability, or it is strongly dominant and yields a very strong
-probability that the location is incorrect, regardless of the estimate
-from the MEM seeding process.
+choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears in
+expression (\ref{eq:bayes}).
+
+Because the probability that the best candidate is a random sequence is
+either very small or very large, it either has no influence on the
+probability of false positive, or completely dominates it. For every read,
+we thus take the maximum of this probability or the estimated probability
+that MEM seeding returned a false positive.
 

### 2019-02-15 MEM.tex
Added some references about seeding.

@@
 There exists exact algorithms to solve the best location
-problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
+problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
 the large amount of data generated by modern sequencers. Instead, one uses
 heuristic methods, \textit{i.e.} algorithms that run fast, but may not
-return the best location\cite{Waterman1984}. Heuristic mapping algorithms
+return the best location~\cite{Waterman1984}. Heuristic mapping algorithms
 thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
@@
 
-The most popular heuristic to address the best location problem is to
-first identify short regions of high similiarity between the read and the
-genome, and then run an exact alignment algorithm around those regions.
-The first step is usually called ``seeding'' and the regions of high
-similarity are called ``seeds''. This is for instance the general strategy
-of the popular alignment algorithm BLAST\cite{pmid2231712}.
+The most popular heuristic for this problem is to first identify short
+regions of high similiarity between the read and the genome, and then run
+an exact alignment algorithm around those regions. The first step is a
+filtration called ``seeding'' and the regions of high similarity
+are called ``seeds''. This is for instance the general strategy of the
+popular alignment algorithm BLAST~\cite{pmid2231712}.
 
@@
 Importantly, the terms of this trade-off depend on the algorithm. Some
-seeding methods are better than others\cite{pmid16533404,pmid20460430}, so
-one can develop faster mapping algorithms at no cost for accuracy.
+seeding methods are better than others (see~\cite{sun2005designing,
+pmid11934743,xu2006optimizing,pmid17044164,brejova2003vector,
+pmid18684737} for examples of design, see~\cite{pmid16533404,pmid20460430}
+for comparisons and discussions, and see~\cite{navarro2001guided} for a
+more general review of filtration methods in pattern matching). The upshot
+is that one can develop faster mapping algorithms at no cost for accuracy.
 
@@
 
-In general, seeds do not have to be exact matches\cite{pmid20460430}. For
-instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
+In general, seeds do not have to be exact matches~\cite{pmid20460430}. For
+instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
 seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
@@
 strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
-Exact Match seeds).} of the popular mapper BWA-MEM\cite{li2013aligning})
+Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
 but there is presently no theory to justify its performance. A deeper
@@
 exists practical methods to identify MEM seeds, even when the reference
-genome is very large\cite{ferragina2000opportunistic,
+genome is very large~\cite{ferragina2000opportunistic,
 ferragina2005indexing}. Such algorithms are crucial for the present
@@
 We refer the interested reader to the relevant
-literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
+literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
 li2013aligning}, and here on we will assume that MEM seeds are available
@@
 that this can be done with an algorithm called the \emph{backward
-search}\cite{ferragina2000opportunistic}.
+search}~\cite{ferragina2000opportunistic}.
 
@@
 combinatorics has been published
-elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
+elsewhere~\cite{filion2017analytic,filion2018analytic}. We refer the
 readers interested in learning more about analytic combinatorics to the
-textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.
+textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.
 
@@
 The number of errors $X$ in a read of size $k$ has a Binomial
-distribution $X \sim B(k,p)$. From\cite{arratia1989tutorial} we can bound
+distribution $X \sim B(k,p)$. From~\cite{arratia1989tutorial} we can bound
 the probabilities of the tail with the expression
@@
 from a binomial distribution can be done by the method of
-Kachitvichyanukul and Schmeiser\cite{kachitvichyanukul1988binomial}.
+Kachitvichyanukul and Schmeiser~\cite{kachitvichyanukul1988binomial}.
 
@@
 sequence. Methods to do this efficiently have been explained in detail
-elsewhere\cite{filion2018analytic}. Here we only give the recurrence
+elsewhere~\cite{filion2018analytic}. Here we only give the recurrence
 equation to compute $P(E_0)$. Assume that the probability of a sequencing
@@
 The backward search is a method that was first proposed by Ferragina and
-Mazini\cite{ferragina2000opportunistic} to identify exact matches between
+Mazini~\cite{ferragina2000opportunistic} to identify exact matches between
 a text query and an index. The query is extended one letter/nucleotide at
@@
 \bibliography{references,pubmed}
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 

### 2019-02-18 C/compute_mem_prob/compute_mem_prob.c
Added Makefile and removed C directory.

@@ -1,1021 +1,0 @@
-#include <math.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <strings.h>
-
-
-// MACROS //
-
-#define LIBNAME "compute_mem_prob"
-#define VERSION "0.9 04-27-2018"
-
-#define YES 1
-#define NO  0
-
-#define SUCCESS 1
-#define FAILURE 0
-
-// Maximum allowed number of duplicates.
-#define MAXN 1024
-
-// Compute omega and tilde-omega.
-#define  OMEGA ( P*pow(1.0-U/3, N) )
-#define _OMEGA ( P*(1-pow(1.0-U/3, N)) )
-
-// Prob that one of m altnerative threads survives i steps.
-#define xi(i,m) ( 1.0 - pow( 1.0 - pow(1.0-U,(i)), (m) ))
-
-// Calculation intermediates (one index).
-#define aN(i) pow( 1.0 - pow(1.0-U,(i)) * U/3.0, N )
-#define gN(i) pow( 1.0 - pow(1.0-U,(i)), N )
-#define dN(i) pow( 1.0 - (1.0 - U + U*U/3.0) * pow(1.0-U,(i)), N )
-
-// Calculation intermediates (two indices).
-#define bN(j,i) pow( 1.0 - pow(1.0-U,(j))*U/3.0 - \
-                           pow(1.0-U,(i))*(1.0-U/3.0), N)
-
-
-// Macro to simplify error handling.
-#define handle_memory_error(x) do { \
-   if ((x) == NULL) { \
-      warning("memory_error", __func__, __LINE__); \
-      ERRNO = __LINE__; \
-      goto in_case_of_failure; \
-}} while (0)
-
-
-//  TYPE DECLARATIONS  //
-
-typedef struct trunc_pol_t  trunc_pol_t;
-typedef struct matrix_t     matrix_t;
-typedef struct monomial_t   monomial_t;
-
-// TODO : explain that monomials must not be just deg and coeff //
-
-struct monomial_t {
-   size_t deg;           // Degree of the coefficent.
-   double coeff;         // Value of the coefficient
-};
-
-struct trunc_pol_t {
-   monomial_t mono;      // Monomial (if applicable).
-   double coeff[];       // Terms of the polynomial.
-};
-
-struct matrix_t {
-   const size_t    dim;  // Column / row number.
-   trunc_pol_t * term[]; // Terms of the matrix.
-};
-
-
-
-// GLOBAL VARIABLES / CONSTANTS //
-
[…]

### 2019-02-18 C/compute_mem_prob/main.c

@@ -1,34 +1,0 @@
-#include <stdio.h>
-#include "compute_mem_prob.h"
-
-int main(void) {
-
-   // First fix parameters to compute seeding probabilities.
-   // In this example, the parameters are
-   // 1. minimum MEM seed size: 17
-   // 2. maximum read size: 50
-   // 3. sequencing error rate: 0.01
-   // 4. divergence rate: 0.05
-   int success = set_params_mem_prob(17, 50, 0.01, 0.05);
-
-   // Success of initialization can be checked
-   // from the return value of the function.
-   if (!success) return -1;
-
-   double x = compute_mem_prob(2, 18);
-
-   // If something went wrong, the function returns 'nan'.
-   // This can be checked with the condition 'x != x'.
-   if (x != x) return -1;
-
-   // Alternatively, one can directly access the last error.
-   if (get_mem_prob_error_code()) return -1;
-
-   // After using the library, one needs to call 'clean_mem_prob()'
-   // to free internal variables and reset the internal state.
-   clean_mem_prob(); 
-
-   fprintf(stdout, "%.5f\n", x);
-   return 0;
-
-}

### 2019-02-18 C/compute_mem_prob/unittest/unittests_compute_mem_prob.c

@@ -1,1859 +1,0 @@
-#include "unittest.h"
-#include "compute_mem_prob.c"
-
-void
-test_set_params_mem_prob
-(void)
-{
-
-   int success;
-
-   // Case 1.
-   success = set_params_mem_prob(17, 50, 0.01, 0.05);
-   test_assert_critical(success);
-
-   test_assert(G == 17);
-   test_assert(K == 50);
-   test_assert(P == 0.01);
-   test_assert(U == 0.05);
-
-   test_assert(HIGH == 50);
-   test_assert(KSZ == sizeof(trunc_pol_t) + 51*sizeof(double));
-
-   for (int i = 0 ; i < MAXN ; i++) {
-      test_assert(ARRAY[i] == NULL);
-   }
-
-   // Case 2.
-   set_params_mem_prob(50, 20, 0.99, 0.95);
-   test_assert_critical(success);
-
-   test_assert(G == 50);
-   test_assert(K == 20);
-   test_assert(P == 0.99);
-   test_assert(U == 0.95);
-
-   test_assert(HIGH == 50);
-   test_assert(KSZ == sizeof(trunc_pol_t) + 21*sizeof(double));
-
-   for (int i = 0 ; i < MAXN ; i++) {
-      test_assert(ARRAY[i] == NULL);
-   }
-
-   clean_mem_prob();
-
-   return;
-
-}
-
-
-void
-test_error_set_params_mem_prob
-(void)
-{
-
-   int success;
-
-   // Case 1.
-   redirect_stderr();
-   success = set_params_mem_prob(17, 50, 0.00, 0.05);
-   unredirect_stderr();
-   test_assert(!success);
-   test_assert_stderr("[compute_mem_prob] error in function `set_");
-
-   // Case 2.
-   redirect_stderr();
-   success = set_params_mem_prob(17, 50, 0.01, 0.00);
-   unredirect_stderr();
-   test_assert(!success);
-   test_assert_stderr("[compute_mem_prob] error in function `set_");
-
-   // Case 3.
-   redirect_stderr();
-   success = set_params_mem_prob(17, 50, 1.00, 0.05);
-   unredirect_stderr();
-   test_assert(!success);
-   test_assert_stderr("[compute_mem_prob] error in function `set_");
-
-   // Case 4.
-   redirect_stderr();
[…]

### 2019-02-18 C/make_figure_simulation_mem_fp.R

@@ -1,34 +1,0 @@
-d = 17
-k = 30:100
-k3 = rep(30:100, each=3)
-
-L = list()
-M = list()
-
-M17_2 = as.matrix(read.table("../False_Positives/M17.txt"))
-z3_2 = M17_2[151:200,c(1,3,5)]
-C3_2 = M17_2[151:200,c(2,4,6)]
-
-M17_3 = as.matrix(read.table("M17.txt"))
-z3_3 = M17_3[151:200,c(1,3,5)]
-C3_3 = M17_3[151:200,c(2,4,6)]
-
-for (i in 1:50) {
-   # All the terms
-   #L[[i]] = Re(tapply(X = C3_3[i,]/z3_3[i,]^(k3+1) - C3_2[i,]/z3_2[i,]^(k3+1), INDEX=k3, sum))
-   # P2 simplified
-   #L[[i]] = Re(tapply(X = C3_3[i,]/z3_3[i,]^(k3+1), INDEX=k3, sum)) - Re(C3_2[i,1]/z3_2[i,1]^(k+1))
-   # P3 simplified
-   #L[[i]] = Re(C3_3[i,1]/z3_3[i,1]^(k+1)) - Re(tapply(X = C3_2[i,]/z3_2[i,]^(k3+1), INDEX=k3, sum)) 
-   # Both simplified
-   L[[i]] = Re(C3_3[i,1]/z3_3[i,1]^(k+1)) - Re(C3_2[i,1]/z3_2[i,1]^(k+1))
-}
-
-pdf("sht.pdf")
-ymin = 0
-ymax = max(unlist(L))
-plot(L[[1]], type='l', ylim=c(ymin,ymax))
-for (i in 33:50) {
-   lines(L[[i]])
-}
-dev.off()

### 2019-02-18 C/misc/print_mini.py

@@ -1,83 +1,0 @@
-#!/usr/bin/env python
-# -*- coding:utf-8 -*-
-
-import sys
-
-'''This script is used to print the relatively complex transfer matrix
-of the model with a main target and a duplicate.'''
-
-class MatrixAsString:
-
-   def __init__(self, gamma, k, p, kappa, N):
-      self.gamma = gamma
-      self.k     = k    
-      self.p     = p
-      self.kappa = kappa
-      self.N     = N
-
-   def power(self, txt, k):
-      if k == 0: return '1'
-      if k == 1: return txt
-      return '(%s)^%d' % (txt, k)
-
-   def AB_type_terms(self, k):
-      terms = [self.power('%f*z' % (1-self.p), m) for m in range(k)]
-      return '+'.join(terms)
-
-   def tilde_AB_type_terms(self, k):
-      qz = '%f*z' % (1-self.p)
-      terms = [self.Cm(m) + '*' + self.power(qz, m) for m in range(k)]
-      return '+'.join(terms)
-
-   def A(self, k):
-      cst_term = '%f*z*(1-(1-%f/3)^%d)' % (self.p, self.kappa, self.N)
-      if k == 1: return cst_term
-      return cst_term + '*(' + self.AB_type_terms(k) + ')'
-
-   def B(self, k):
-      cst_term = '%f*z*(1-%f/3)^%d' % (self.p, self.kappa, self.N)
-      if k == 1: return cst_term
-      return cst_term + '*(' + self.AB_type_terms(k) + ')'
-
-   def Cm(self, k):
-      if k == 0: return '1'
-      if k == 1: return '(1-%f^%d)' % (self.kappa, self.N)
-      return '(1-(1-(1-%f)^%d)^%d)' % (self.kappa, k, self.N)
-
-   def tildeA(self, k):
-      cst_term = '%f*z*(1-(1-%f/3)^%d)' % (self.p, self.kappa, self.N)
-      if k == 1: return cst_term
-      return cst_term + '*(' + self.tilde_AB_type_terms(k) + ')'
-
-   def tildeB(self, k):
-      cst_term = '%f*z*(1-%f/3)^%d' % (self.p, self.kappa, self.N)
-      if k == 1: return cst_term
-      return cst_term + '*(' + self.tilde_AB_type_terms(k) + ')'
-
-   def r(self, k):
-      if k == 1:
[…]

### 2019-02-18 C/misc/print_transfer_matrix.py

@@ -1,73 +1,0 @@
-#!/usr/bin/env python
-# -*- coding:utf-8 -*-
-
-import sys
-
-'''This script is used to print the relatively complex transfer matrix
-of the model with a main target and a duplicate.'''
-
-class MatrixAsString:
-
-   def sqbr(txt):
-      return '[' + txt + ']'
-
-   def rdbr(txt):
-      return '(' + txt + ')'
-
-   def pow(txt, m):
-      if m == 0: return '1'
-      if m == 1: return txt
-      return self.rdbr('%s^%d' % txt, m)
-
-   def A(k, p, kappa, N):
-      cst_term = '%f*(1-(1-%f/3)^%d)*z' % (p, kappa, N)
-      if k == 1: return cst_term
-
-      other_terms = [self.pow('%f*z' % 1-p, m) for m in range(k)]
-      self.sqbr('+'.join(other_terms))
-
-
-   def first_term(n):
-      '''Return a string of the first term of the row.'''
-      if n == 0: return 'd*z'
-      return 'd*z*(' + '+'.join([az(a) for a in range(n+1)]) + ')'
-
-   def shift_right(L, shift):
-      '''Shift the terms to the right.'''
-      if shift == 0: return L
-      return ['0']*shift + L[:-shift]
-
-   def overwrite_tail(L, n):
-      '''Replace the tail elements by 0.'''
-      if n == 0: return L
-      return L[:-n] + ['0']*n 
-
-   # Two complete lists of terms.
-   L1 = ['c*z'] + ['c*z*%s' % az(a) for a in range(1,d-1)]
-   L2 = ['b*z'] + ['b*z*%s' % az(a) for a in range(1,d-1)]
-
-   # First row.
-   row_terms = [first_term(d-1)] + L1 + L2
-   rows = [brackets(', '.join(row_terms))]
-
-   # Next series of rows.
-   for i in range(1,d):
-      row_terms = [first_term(d-1-i)] + \
-            shift_right(L1,i) + overwrite_tail(L2,i-1)
-      rows.append(brackets(', '.join(row_terms)))
-
-   # Next series of rows.
-   for i in range(1,d):
-      row_terms = [first_term(d-1-i)] + \
-            overwrite_tail(L1,i-1) + shift_right(L2,i)
-      rows.append(brackets(', '.join(row_terms)))
-
-   return brackets(', '.join(rows))
-
-def write_assignment_line(d):
-   head = 'M%d := Matrix(' %d
[…]

### 2019-02-18 C/mt.h

@@ -1,3 +1,0 @@
-typedef unsigned long uint32;
-void seedMT(uint32);
-uint32 randomMT(void);

### 2019-02-18 C/sim.c

@@ -1,123 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <strings.h>
-#include <time.h>
-
-#include "mt.h"
-
-// Global variables.
-unsigned int K;
-unsigned int GAMMA;
-unsigned int NDUP;
-
-
-void
-update_dup
-(
-         unsigned int * dup,
-   const unsigned long int mt_match
-)
-{
-   for (int j = 0 ; j < NDUP ; j++)
-      dup[j] = randomMT() < mt_match ? dup[j] + 1 : 0;
-   return;
-}
-
-
-int
-there_is_a_seed
-(
-   const unsigned int   target,
-   const unsigned int * dup
-)
-{
-   if (target < GAMMA) return 0;
-   for (int j = 0 ; j < NDUP ; j++)
-      if (dup[j] >= target) return 0;
-   return 1;
-}
-
-
-int
-simulate_read
-(
-   unsigned long int   mt_error,
-   unsigned long int   mt_match1,
-   unsigned long int   mt_match2,
-   unsigned int      * dup
-)
-{
-   int read_error;
-
-   // Initialize counters.
-   bzero(dup, NDUP * sizeof(int));
-   int target = 0;
-
-   for (int i = 0 ; i < K ; i++) {
-
-      read_error = randomMT() < mt_error;
-      update_dup(dup, read_error ? mt_match2 : mt_match1);
-
-      // The code of 'there_is_a_seed()' works for both strict and
-      // shared seeds, but we need to update 'target' before when
-      // there is no error, and after when there is one.
-      if (!read_error) target++;
-      if (there_is_a_seed(target, dup)) return 1;
-      if (read_error) target = 0;
-
-   }
-
-   // Check if there is a shared seed at the end of the read.
-   if (target >= GAMMA && !read_error)
-      return there_is_a_seed(target+1, dup);
-
-   return 0;
-
-}
-
-
-
-int main(int argc, char **argv) {
-
-                K     = atoi(argv[1]);            // Read size.
-                GAMMA = atoi(argv[2]);            // Seed length.
-   double       prob  = strtod(argv[3], NULL);    // Read error.
-   double       kappa = strtod(argv[4], NULL);    // Divergence.
-                NDUP  = atoi(argv[5]);            // Duplicate number.
-   long int     ITER  = atoi(argv[6]);
-
-   // Check arguments.
[…]

### 2019-02-18 MEM.tex
Improved the abstract.

@@
 \begin{abstract}
-High throughput sequencing produces massive amounts of data that must be
-processed efficiently. Most reads are identified by mapping them to a
-reference, typically a genome. Modern mappers rely on heuristics to gain
-speed at a reasonable cost of accuracy. MEM seeding (where MEM stands for
-Maximal Exact Match) is a popular heuristic that shows good empirical
-performance but for which no theory is available. It is thus difficult to
-calibrate or optimize mapping strategies based on MEM seeding. Here we
-develop a theory to predict the performance of MEM seeding. More
-specifically, we design a method to compute the probability a read is
-mapped to the right location when using MEM seeds. Our computational
-strategy is based on the analytic combinatorics programme, but uses only
-part of it. By recoding the reads in a specialized combinatorial alphabet,
-we can implicitly specify their generating function and extract the
-probabilities of interest by direct methods or by Monte Carlo sampling.
-This yields an efficient algorithm to calibrate the MEM seeding heuristic
-for short read technologies.
+The increasing throughput of DNA sequencing technologies demands faster
+algorithms to analyze sequencing data. The fate of most reads is to be
+mapped to a reference sequence, typically a genome. Modern mappers rely on
+heuristics to gain speed at a reasonable cost for accuracy. MEM seeding
+(where MEM stands for Maximal Exact Match) is a popular heuristic that
+shows good empirical performance but for which theory is lacking. It is
+thus difficult to calibrate or optimize mapping strategies based on MEM
+seeding. Here we develop a theory to estimate the probability that the MEM
+seeding heuristic fails to return the correct candidate. The major
+innovation is to use concepts from analytic combinatorics to represent
+reads as abstract sequences, from which we estimate the probabilities of
+interest. We provide two separate methods, which combined together give a
+practicial solution to the problem of calibrating MEM seeding when
+aligning short reads. These results have implications for MEM-based
+heuristics and lay the foundation of a general strategy to tackle sequence
+alignment problems.
 \end{abstract}
@@
 
-The focus of this document is not to solve the true location problem
-\textit{per se}, but rather to evaluate whether a given solution is
-correct. At first sight, whether we can map a read to its true location
-seems to depend only on its length and on the error rate of the
-sequencer. However, there is another important factor to consider: most
-genomes contain repeated sequences, \textit{i.e.} relatively large
-subsequences that are present at more than one location. For instance, we
-cannot map the fragment with certainty if there exists an exact copy of
-the sequence somewhere else in the genome, because we cannot know which of
-the two copies was sequenced.
+At first sight, whether a read read can be mapepd to its correct location
+seems to depend only on its length and on the error rate of the sequencer.
+However, there is another important factor to consider: most genomes
+contain repeated sequences, \textit{i.e.} relatively large subsequences
+that are present at more than one location. For instance, we cannot map
+the fragment with certainty if there exists an exact copy of the sequence
+somewhere else in the genome, because we cannot know which of the two
+copies was sequenced.
 
@@
 to finding the optimal alignment between two sequences, and for this
-reason has received substantial attention in bioinformatics. Finding the
-best location of the read is the only way to get to its true location, so
-we need to focus our attention on the best location problem.
+reason has received substantial attention in bioinformatics. It turns out
+that finding the best location of the read is the only way to get to its
+true location, so for now we need to focus our attention on the best
+location problem.
 
@@
 regions of high similiarity between the read and the genome, and then run
-an exact alignment algorithm around those regions. The first step is a
-filtration called ``seeding'' and the regions of high similarity
-are called ``seeds''. This is for instance the general strategy of the
-popular alignment algorithm BLAST~\cite{pmid2231712}.
+an exact alignment algorithm locally at those regions. The first step is a
+filtration called ``seeding'' and the regions of high similarity are
+called ``seeds''. This is for instance the general strategy of the popular
+alignment algorithm BLAST~\cite{pmid2231712}.
 
@@
 the candidate set. In this case, it cannot be discovered and the outcome
-is either $ii.$ or $iii.$
-
-This imposes a trade-off between speed and accuracy because each candidate
-location must be tested with a slow algorithm. If the seeding step is set
-to produce many candidates, the target is likely to be among them and it
-will be discovered. The downside is that this will take long. Conversely,
-if the seeding step is set to produce few candidates, the process will
-run faster but the target is less likely to be in the candidate list so it
-is less likely to be discovered.
+is either $ii.$ or $iii.$, which are both erroneous.
+
+The slowness of the alignment algorithm imposes a trade-off between speed
+and accuracy. If the seeding step is set to produce many candidates, the
+target is likely to be among them and it will be discovered. The downside
+is that this will take long. Conversely, if the seeding step is set to
+produce few candidates, the process will run faster but the target is less
+likely to be in the candidate list so it is less likely to be discovered.
 
 Importantly, the terms of this trade-off depend on the algorithm. Some
-seeding methods are better than others (see~\cite{sun2005designing,
-pmid11934743,xu2006optimizing,pmid17044164,brejova2003vector,
-pmid18684737} for examples of design, see~\cite{pmid16533404,pmid20460430}
-for comparisons and discussions, and see~\cite{navarro2001guided} for a
-more general review of filtration methods in pattern matching). The upshot
-is that one can develop faster mapping algorithms at no cost for accuracy.
+seeding methods are better than others (see~\cite{pmid20460430,
+sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
+brejova2003vector,pmid18684737,pmid15359419} for examples of design,
+see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
+see~\cite{navarro2001guided} for a more general review of filtration
+methods in pattern matching). The upshot is that we can develop faster
+mapping algorithms at no cost for accuracy if we find better seeding
+strategies.
 
@@
 exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
-location. Outcome $i.$, called a ``true positive'', is that the DNA
-fragment is mapped to the \emph{true} location. Outcome $ii.$, called a
-``false positive'' is that the DNA fragment is mapped to an incorrect
-location. Outcome $iii.$, called a ``false negative'', does not change and
-still means that the DNA fragment is not mapped at all.
+location. In outcome $i.$ called a ``true positive'', the DNA fragment is
+mapped to the \emph{true} location. In outcome $ii.$ called a ``false
+positive'', the DNA fragment is mapped to an incorrect location. Outcome
+$iii.$ called a ``false negative'' is the same as before and means that
+the DNA fragment is not mapped at all.
 
@@
 determined by the seeding process. For this reason, we will sometimes
-refer to them as the outcomes of the seeding step, even though they are
-the outcomes of the whole mapping process.
+refer to them as the outcomes of the seeding step, even though strictly
+speaking they are the outcomes of the whole mapping pipeline\footnote{This
+can be confusing because some authors refer to the off-target seeds as
+false positives. When we say that the outcome of the seeding step is a
+false positive, we mean that \emph{all} the seeds are off-target.}.
 
@@
 
-In general, seeds do not have to be exact matches~\cite{pmid20460430}. For
+In general, seeds do not have to be exact matches{}. For
 instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
@@
 \begin{definition}
-A \emph{strict} MEM seed has a single genomic match.
-A \emph{shared} MEM seed has several genomic matches.
+A \emph{strict} MEM seed has a single match in the genome.
+A \emph{shared} MEM seed has several matches in the genome.
 \end{definition}
@@
 
+Algorithm~\ref{alg:mcmc} is a major result. It gives a compact solution to
+the problem of estimating the probability that a read can be mapped when
+using MEM seeds. The algorithm is also much faster than the naive approach
+of sampling every nucleotide of every sequence, because it samples many
+nucleotidea from \emph{all} the sequences at each iteration.
+
 \section{False positives}
@@
 We have devised a method to compute the probability that the MEM seeding
-heuristic misguides the mapping algorithm so that it cannot find the true
-location of the read. The pillars of our strategy are borrowed from
-analytic combinatorics~\cite{flajolet2009analytic,
+heuristic fails and sets up the mapping process for a false positive. This
+fills a gap of theoretical knowledge to understand and calibrate the
+performance of the MEM seeding heuristic. The pillars of our strategy are
+borrowed from analytic combinatorics~\cite{flajolet2009analytic,
 sedgewick2013introduction}, even though we do not follow the standard

### 2019-03-03 MEM.tex
Adding theory for skip seeds and exact seeds.

@@
 \usepackage{cite}
+\usepackage{enumerate}
 \usepackage{graphicx}
@@
 \newcommand{\dn}[1]{\scriptstyle{\downarrow_{/#1}}}
+\newcommand{\Dn}[1]{\scriptstyle{\Downarrow_{#1}}}
 \newcommand{\up}[1]{\scriptstyle{\uparrow_{/#1}}}
 \newcommand{\nd}{\scriptstyle{|}}
+\newcommand{\modulo}[1]{\ (\mathrm{mod}\ #1)}
 
@@
 
-\title{Computing MEM seeding probabilities}
+\title{Estimating mismapping rates in seed-based algorithms
+for read mapping}
 
@@
 \begin{abstract}
+% Statement of the problem.
 The increasing throughput of DNA sequencing technologies demands faster
@@
 mapped to a reference sequence, typically a genome. Modern mappers rely on
-heuristics to gain speed at a reasonable cost for accuracy. MEM seeding
-(where MEM stands for Maximal Exact Match) is a popular heuristic that
-shows good empirical performance but for which theory is lacking. It is
-thus difficult to calibrate or optimize mapping strategies based on MEM
-seeding. Here we develop a theory to estimate the probability that the MEM
-seeding heuristic fails to return the correct candidate. The major
-innovation is to use concepts from analytic combinatorics to represent
-reads as abstract sequences, from which we estimate the probabilities of
-interest. We provide two separate methods, which combined together give a
-practicial solution to the problem of calibrating MEM seeding when
-aligning short reads. These results have implications for MEM-based
-heuristics and lay the foundation of a general strategy to tackle sequence
-alignment problems.
+heuristics to gain speed at a reasonable cost for accuracy. In the seeding
+approach, short matches between the reads and the genome are used to
+narrow the search to a set of candidate locations. Several seeding
+variants used in modern mappers show good empirical performance but
+theoretical results are still lacking, making them difficult to calibrate
+or top optimize.
+% Summary of the results.
+Here we develop a theory to estimate the probability that reads are
+mismapped due to a seeding failure. We describe the properties of simple
+exact seeds, skip-seeds used in Bowtie2 and MEM (Maximal Exact Match)
+seeds used in BWA-MEM. 
+% Innovation.
+The major innovation is to use concepts from analytic combinatorics to
+represent reads as abstract sequences, from which we estimate the
+probabilities of interest.
+% Durability.
+We provide several algorithms, which combined together give a practicial
+solution to the problem of calibrating seeding heuristic for short reads.
+% Scope.
+These results significantly improve current mapping algorithms and lay the
+foundation of a general strategy to tackle sequence alignment problems.
 \end{abstract}
@@
 to finding the optimal alignment between two sequences, and for this
-reason has received substantial attention in bioinformatics. It turns out
-that finding the best location of the read is the only way to get to its
-true location, so for now we need to focus our attention on the best
-location problem.
+reason has received substantial attention in bioinformatics. For
+simplicity we will assume that the true location is always the best. We
+will relax this assumption in section section~\ref{sec:not_the_best}, but
+until then we will consider that the solution of the true location problem
+is the best location.
 
@@
 
-There exists exact algorithms to solve the best location
+There exists algorithms to solve the best location
 problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
 the large amount of data generated by modern sequencers. Instead, one uses
-heuristic methods, \textit{i.e.} algorithms that run fast, but may not
-return the best location~\cite{Waterman1984}. Heuristic mapping algorithms
-thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
-best location, $ii.$ the DNA fragment is mapped to some other location,
-$iii.$ the DNA fragment is not mapped at all.
-
-The most popular heuristic for this problem is to first identify short
+heuristic methods, \textit{i.e.} algorithms that run fast, but that may
+return an incorrect result~\cite{Waterman1984}.
+
+The most popular heuristic for the best location problem is a filtration
+method called ``seeding''. The principle is to first identify short
 regions of high similiarity between the read and the genome, and then run
-an exact alignment algorithm locally at those regions. The first step is a
-filtration called ``seeding'' and the regions of high similarity are
-called ``seeds''. This is for instance the general strategy of the popular
-alignment algorithm BLAST~\cite{pmid2231712}.
-
-If seeds can be discovered fast, we can quickly zoom in on a small set of
-candidate locations and reduce the input size of the (slow) alignment
-algorithm. The disadvantage is that the target location is not always in
-the candidate set. In this case, it cannot be discovered and the outcome
-is either $ii.$ or $iii.$, which are both erroneous.
+an exact alignment algorithm at the candidate locations. This is for
+instance the general strategy of the popular alignment algorithm
+BLAST~\cite{pmid2231712}.
+
+If seeding is fast, we can quickly zoom in on a small set of candidate
+locations and reduce the input size of the (slow) alignment algorithm. The
+disadvantage is that the target location is not always in the candidate
+set. When that happens, the read cannot be mapped to the correct location
+because it has been filtered out at the seeding step.
 
 The slowness of the alignment algorithm imposes a trade-off between speed
-and accuracy. If the seeding step is set to produce many candidates, the
-target is likely to be among them and it will be discovered. The downside
-is that this will take long. Conversely, if the seeding step is set to
-produce few candidates, the process will run faster but the target is less
-likely to be in the candidate list so it is less likely to be discovered.
-
-Importantly, the terms of this trade-off depend on the algorithm. Some
-seeding methods are better than others (see~\cite{pmid20460430,
-sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
-brejova2003vector,pmid18684737,pmid15359419} for examples of design,
-see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
-see~\cite{navarro2001guided} for a more general review of filtration
-methods in pattern matching). The upshot is that we can develop faster
-mapping algorithms at no cost for accuracy if we find better seeding
-strategies.
-
-The seeding framework was developed for the best location problem, but it
-also applies to the true location problem. Actually, the process is
-exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
-location. In outcome $i.$ called a ``true positive'', the DNA fragment is
-mapped to the \emph{true} location. In outcome $ii.$ called a ``false
-positive'', the DNA fragment is mapped to an incorrect location. Outcome
-$iii.$ called a ``false negative'' is the same as before and means that
-the DNA fragment is not mapped at all.
-
-We will make two assumptions that will simplify the discussion. The first
-is that the true location is always the best. We will further comment on
-this and explain how to relax this assumption in section
-section~\ref{sec:not_the_best}. Until then, we consider that the true
-location problem is the same as the best location problem. The second
-assumption is that the best location is always discovered if it is in the
-candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
-determined by the seeding process. For this reason, we will sometimes
-refer to them as the outcomes of the seeding step, even though strictly
-speaking they are the outcomes of the whole mapping pipeline\footnote{This
-can be confusing because some authors refer to the off-target seeds as
-false positives. When we say that the outcome of the seeding step is a
-false positive, we mean that \emph{all} the seeds are off-target.}.
+and accuracy: If the seeding step is set to produce many candidates, it is
+likely to retrieve the candidate. The downside is that this will take
+long. Conversely, if the seeding step is set to produce few candidates,
+the process will run faster but the target is less likely to be in the
+candidate list.
+
+Importantly, the trade-off depends on the seeding method. This means that
+we can develop faster mapping algorithms at no cost for accuracy, as long
+as we can find better seeding strategies. Progress on this line of
+research has largely benefited from the improvement of computer hardware
+and from the deveopment of indexing algorithms. Literature on this topic
+is already abudant. Some examples of seeding algorithms are given in
+references~\cite{sun2005designing,pmid11934743,xu2006optimizing,
+pmid17044164,brejova2003vector,pmid18684737,pmid15359419}.
+References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
+comparisons of different designs, and reference~\cite{navarro2001guided}
+gives a global overview of filtration methods in pattern matching.
+
+\subsection{The two types of seeding failure}
+
+Filtering methods are considered to fail whenever the target is not in the
+candidate set. For seeding heuristics, we need to distinguish two
+different kinds of failure. In the first kind, the candidate set contains 
+a duplicate of the target but does not contain the target itself; in the
+second kind, the candidate set contains neither.
+
+The distinction is important because duplicates of the target are similar
+to the read (due to their similarity to the target), so a failure of the
+first kind is difficult to distinguish from a success. A failure of the
+second kind is easier to recognize because in this case the best hit is
+not similar to the read.
+
+Before going further, we introduce some vocabulary that will be useful to
+simplify the exposition of the theory.
+
+\begin{definition}
+The seeding process is said to be
+\begin{enumerate}[i)]
+\item on target if the candidate set contains the target,
+\item off target if the candidate set contains a duplicate but
+not the target, 
+\item null if the candidate set contains neither.
+\end{enumerate}
+\end{definition}
+
+If a seed-based algorithm returns an incorrect location, then the best
+location was never tested with an exact alignment. If the error is given a
+high confidence, then it must be the best in the candidate set and it must
+be similar to the read. This narrows down the cases of mismapping, or more
+exactly high confidence mismapping to off-target seeding.
+
+This observation is fundamental. In order to estimate the mismapping rate
+of an algorithm, we must know how often the associated seeding heuristic
+is off target. Yet, this information is missing for nearly all seeding
+methods, meaning that current mappers cannot be properly calibrated.
+
+The focus of this article is to develop a theory to estimate the
+probability of off-target seeding under the most popular seeding
+heuristics. Our previous work pioneered a method to compute seeding
+probabilities but it did not distinguish off-target from null
+seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
+not provide a way to control the mismapping rate. Other authors
+investigated the reliability of mapping algorithms~\cite{pmid23872968},
+but they focused on the probability that random sequences have a single
+hit, recognizing that solving the problem requires taking into
+consideration the repeat structure of the genome.
+
+\section{Types of seeds}
+
+The term ``seed'' does not have exactly the same meaning according to
+different authors. It can designate a part of the read, a part of the
+genome, a particular sequence motif or a structured pattern of matches.
+Also, the term is not always short for ``exact match''. For instance, the
+algorithm PatternHunter~\cite{pmid11934743} describes ``spaced seeds''
+that tolerate mismatches\footnote{For the anecdote, the method is covered
+by a patent (reference US20070088510A1).}. To avoid potential confusion,
+we will adopt the following convention.
+
+\begin{definition}
+A seed is a subsequence of the read that is above a certain size $\gamma$
+(defined by the context of the problem) and that has at least one perfect
+match in the reference genome.
+\end{definition}
+
+When a seed matches a particular location of the genome, we will say that
+it is a seed \emph{for} that location. For instance, we will often refer
+to a seed for the target or a seed for one of its duplicates.
+
+This definition presents a computational challenge: to know if a given
+subsequence is a seed we need to know if it exists somewhere in the
+genome. This is a non trivial problem in itself, but fortunately there
+exists practical methods to solve it, even when the reference genome is
+very large~\cite{ferragina2000opportunistic, ferragina2005indexing}. Such
+algorithms are crucial for the present theory, but describing them
+falls outside the scope of the document. Looking ahead to
+section~\ref{sec:est}, let us only mention that this can be done with an
+algorithm called the \emph{backward
+search}~\cite{ferragina2000opportunistic}. We refer the interested reader
+to the relevant
+literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736} and
+here on we will assume that seeds are always known without ambiguity [ALL
+THE CITED LITERATURE IS FOR MEM SEEDS, WE NEED TO FIND ARTICLES THAT TREAT
+THE CASE OF OTHER SEEDS].
+
+\subsection{Exact seeds}
+
+Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
+perfect matches of size $\gamma$ are collected to constitute the candidate
+set. This seeding heuristic was used in the first version of
+BLASTN~\cite{pmid2231712}, but is no longer used because they produce many
+overlapping matches in the same regions, as we are about to illustrate.
+
+Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
+by an imperfect sequencer. The sequenced DNA fragment comes from a region
+of the genome that has three extra duplicates, and the read has three
+miscalled nucleotides (indicated by stars).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{exact_seed_example.pdf}
+\caption{\textbf{Example of exact seeds.}
+A sequenced DNA fragment (Read) is shown above its original form in the
+genome (Target).}
+\label{fig:exact_seed_example}
+\end{figure}
+
+Fig.~\ref{fig:exact_seed_example} also illustrates the seeding process
+with exact seeds of size $\gamma=7$. It is clear that erroneous
+nucleotides can also belong to exact seeds because they sometimes match
+one of the duplicates. However, this is not the case of erroneous
+nucleotides that are also mismatches for \emph{all} the duplicates, like
+the second sequencing error in this example.
+
+Thie middle of the read shows some cluttering where consecutive seeds
+match consecutive sequences at the same location. This is typical for
+exact seeds and constitutes a nuisance for the implementation. Indeed, it
+is a waste of computer resources to discover matches in sequences that are
+already in the candidate set. In addition, this seeding method is not
+particularly sensitive~\cite{pmid11934743} so it is used only for specific
+applications. Nevertheless, it will be extremely useful for the
+development of the present theory.
+
+\subsection{Skip seeds}
+
+Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
+start at every nucleotide. Instead, a fixed amount of nucleotides is
+skipped between every seed. This is a way to reduce the amount of
+overlapping matches at the same location, but the cost is a loss of
+sensitivity. This seeding heuristic is used by
+Bowtie2~\cite{pmid22388286}, where seeds have size $\gamma=16$ and fall
+every 10 nucleotides (9 nucleotides are skipped between seeds).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{skip_seed_example.pdf}
+\caption{\textbf{Example of skip seeds.}
+The sequences and the annotations are the same as in
+Fig.~\ref{fig:exact_seed_example}, but here one nucleotide is skipped
+betwen the seeds. In other words, seeds can never start on a nucleotide
+with an even index. To highlight the difference with
+Fig.~\ref{fig:exact_seed_example}, the missing seeds are represented by
+dotted lines. Skipping nucleotides reduces the clutter at the center of
+the read, but it also removes the only seed for the second duplicate.}
+\label{fig:skip_seed_example}
+\end{figure}
+
+Fig.~\ref{fig:skip_seed_example} shows the seeding process on the read of
+Fig.~\ref{fig:exact_seed_example} when using skip seeds instead of exact
+seeds. Here the size is still $\gamma=7$ but 1 extra nucleotide is skipped
+between seeds. This amounts to removing every other seed. The consequence
+is that there are fewer overlapping matches at the center of the read, but
+the only seed for the second duplicate was lost. This is a rather positive
+outcome because there is one off-target location fewer in the candidate
+set, but in general the same might happen to the target.
+
+It is clear that skipping nucleotides reduces the sensitivity of the
+seeding, but to what extent? One could test this empirically, but it
+depends on the seed length, the amount of nucleotides that are skipped,
+the error rate of the sequencer and the size of the reads. The present
+theory will allow us to get the answer automatically.
 
@@
 
-In general, seeds do not have to be exact matches{}. For
-instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
-seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
-covered by a patent (reference US20070088510A1).}. However, in this
-document, we will always assume that a seed is an exact match with no gap.
-In other words, a seed is a perfect match between a part of the read and a
-genomic site.
-
-The rest of the document will focus on a seeding heuristic known as ``MEM
-seeding'' (where MEM stands for Maximal Exact Match). This approach gives
-good empirical results when mapping short reads (it is for instance the
+We will now focus on a seeding heuristic known as ``MEM seeding'' (where
+MEM stands for Maximal Exact Match). This approach gives good empirical
+results when mapping short reads (it is for instance the
 strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
@@
 \end{definition}
-
-This definition presents a computational challenge: to know if a given
-subsequence is a MEM seed we need to know if it exists somewhere in the
-genome. This is a non trivial problem in itself, but fortunately there
-exists practical methods to identify MEM seeds, even when the reference
-genome is very large~\cite{ferragina2000opportunistic,
-ferragina2005indexing}. Such algorithms are crucial for the present
-discussion, but describing them falls outside the scope of the document.
-We refer the interested reader to the relevant
-literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
-li2013aligning}, and here on we will assume that MEM seeds are available
-at all times. Looking ahead to section~\ref{sec:est}, let us just mention
-that this can be done with an algorithm called the \emph{backward
-search}~\cite{ferragina2000opportunistic}.
-
-Before going further, we need to move a practical consideration out of the
-way. It stands to reason that very short matches between the read and the
-genome are uninformative. For instance, $99.7\%$ of all the possible
-12-mers are present in the reference human genome, so an exact match of
-size 12 or lower says nothing about the location of the DNA fragment. In
-what follows, we will assume that subsequences of the read can only match
-the original DNA sequence and/or its duplicates, but not the rest of the
-genome. This will greatly simplify the exposition of the theory. We will
-discuss how to relax this assumption in section~\ref{sec:short_MEMs}.
 
@@
 \includegraphics[scale=1]{MEM_example.pdf}
-\caption{\textbf{Example of MEM seeding.}
+\caption{\textbf{Example of MEM seeds.}
 A DNA fragment is sequenced with an imperfect instrument. It comes from
@@
 the duplicates.
+
+\section{Model and computational strategy}
+\label{sec:symbolic}
+
+\subsection{Sequencing error and duplications}
+
+Now that this is clarified, we need to model the sequencing and the
+duplication processes so that we can compute the probabilities of the
+events of interest. We assume that the sequencing instrument has a
+constant substitution rate $p$, and that insertions and deletions never
+occur. When a substitution occurs, we assume that the instrument is
+equally likely to output any one of the remaining three nucleotides. This
+is a coarse-grain description of the error spectrum of the Illumina
+technology~\cite{pmid21576222}.
+
+Next, we assume that the target sequence has $N \geq 0$ duplicates, so
+that there are $N$ off-target sequences. We further assume that
+duplication happened instantaenously at some point in the past and that
+all $N+1$ sequences diverge independently of each other at a constant
+rate. In other words, we ignore the complications due to the genealogy of
+the duplication events and we simply assume that at each position, any
+given duplicate is identical to the target with probability $1-\mu$. If it
+is not, we assume that the duplicate has any of the remaining three
+nucleotides with equal probability (\textit{i.e.} each is found with
+probability $\mu/3$).
+
+Note that read errors are always mismatches against the target (because we
+assume that the target is the true sequence), and they match each
+off-target sequence with probability $\mu/3$. Correct nucleotides are
+always a match for the target, and they match each off-target sequence
+with probability $1-\mu$.
+
+\subsection{Weighted generating functions}
+
+The central object of analytic combinatorics is the generating function,
+and for our particular purpose we will use a special kind known as
+\emph{weighted generating function}.
+
+\begin{definition}
+\label{def:GF}
+Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
+size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
+generating function of $\mathcal{A}$ is defined as
+
+\begin{equation}
+\label{eq:GF1}
+A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
+\end{equation}
+where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
+respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
+\geq 0}$ such that
+
+\begin{equation*}
+A(z) = \sum_{k=0}^\infty a_k z^k.
+\end{equation*}
+
+By definition $a_k = \sum_{a \in A_k}w(a)$, where $A_k$ is the class of
+objects of size $k$ in $\mathcal{A}$. The number $a_k$ is the
+total weight of objects of size $k$.
+\end{definition}
+
+To give a concrete example, if the probability of the symbol
+$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
+containing only this symbol is $qz$. In our case, we are interested in the
+weighted generating function of reads without good seeds, so that the
+coefficient $a_k$ is equal to the probability that a read of size $k$ has
+no good seed.
+
+The motivation for introducing weighted generating functions is that
+operations on combinatorial objects translate into operations on their
+weighted generating functions. If $A(z)$ and $B(z)$ are the weighted
+generating functions of two mutually exclusive sets $\mathcal{A}$ and
+$\mathcal{B}$, the weighted generating function of $\mathcal{A} \cup
+\mathcal{B}$ is $A(z) + B(z)$, as evident from expression (\ref{eq:GF1}).
+Size and weight can be defined for pairs of objects in $\mathcal{A} \times
+\mathcal{B}$ as $|(a,b)| = |a| + |b|$ and $w(a,b) = w(a)w(b)$. In other
+words the sizes are added and the weights are multiplied.  With this
+convention, the weighted generating function of the Cartesian product
+$\mathcal{A} \times \mathcal{B}$ is $A(z)B(z)$. This simply follows from
+expression (\ref{eq:GF1}) and from
+
+\begin{equation*}
+A(z)B(z) =
+\sum_{a\in \mathcal{A}}w(a)z^{|a|} \sum_{b\in \mathcal{B}}w(b)z^{|b|}
+= \sum_{(a,b) \in \mathcal{A} \times \mathcal{B}} w(a)w(b)z^{|a|+|b|}.
+\end{equation*}
+
+These two operations are all we need in order to compute the weighted
+generating functions of the reads of interest. Addition corresponds to
+creating a new family by merging reads from families $\mathcal{A}$ and
+$\mathcal{B}$. Multiplication corresponds to creating a new family by
+concatenating reads from families $\mathcal{A}$ and $\mathcal{B}$.
+
+\subsection{Analytic representation}
+
+The symbolic method is a strategy belonging to the general \emph{analytic
+combinatorics} framework. The idea is to combine simple combinatorial
+objects into more complex objects. Each combinatorial operation on the
+objects corresponds to a mathematical operation on their weighted
+generating functions. One can thus obtain the weighted generating function
+of complex objects, and from there extract their probability of
+occurrence.
+
+The combinatorial construction of reads using the extended MEM alphabet
+gives us a method to express their weighted generating function from the
+simpler generating function of segments. More specifically, we can
+describe complex sequences of segments through a so-called \emph{transfer
+matrix} $M(z)$ that specifies which pairs of consecutive segments are
+allowed and which are forbidden in the reads of interest.
+
+The segment types are identified by their terminators arranged in a
+predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
+weighted generating function of segments with the $j$-th terminator that
+can be inserted immediately after segments with the $i$-th terminator. The
+entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
+are the weighted generating functions of the reads of all sizes using all
+the allowed combinations of segments.
+
+The entry of interest from $M(z)$ is the one that corresponds to the
+terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
+any size whose first segment can be inserted after $\downarrow_{/0}$, and
+whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
+the end of the read, so it must be the final symbol of every valid read.
+As for the first segment, observe that all the segments that can follow
+the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
+postion of the read\footnote{One could have introduced a special start
+symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
+the reads.}.
+
+In summary, the transfer matrix $M(z)$ contains the weighted generating
+function of interest. Our next task is thus to specify $M(z)$.
+
+More didactic descriptions of the strategy to compute seeding
+probabilities using analytic combinatorics have been published
+elsewhere~\cite{filion2017analytic,filion2018analytic}. The interested
+readers can also discover more about analytic combinatorics in the
+standard textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.
+
+\subsection{Example 1: on-target exact seeds}
+\label{sec:sample_exact}
+
+We highlight the strategy above with an example that will turn out to be
+central for the development of the theory. In addition, it is simple
+enough to provide a gentle introduction to the general methodology. This
+example was described in detail in~\cite{filion2017analytic} and
+in~\cite{filion2018analytic}, but we repeat it here with a formalism that
+fits the present manuscript.
+
+The first step is to note that the nucleotide sequence of the read is
+irrelevant. Indeed, the read has an on-target seed if and only if it
+contains a stretch of $\gamma$ nucleotides without error. So the main step
+is to recode the read as sequence of correct or erroneous nucleotides.
+
+We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
+\Downarrow,\}$, where $\square$ represents a correct nucleotide,
+$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
+appended after the last nucleotide to mark the end of the read.
+
+This recoding allows us to segment the read in \emph{segments} defined as
+a sequence of 0 or more symbols $\square$ followed by a symbol
+$\Downarrow$. The last part of the read is called the tail and it is
+terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
+not correspond to a nucleotide, it has size 0. This decomposition is
+unique, so we can view a read as a sequence of segments with a tail,
+instead of a sequence of nucleotides.
+
+Fig.~\ref{fig:simple} shows an example of decomposition in segments.
+On-target seeds cannot contain sequencing errors, therefore they must lie
+inside a segment or inside the tail. So the size of the segments and of
+the tail indicates whether the read contains an on-target seed.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.85]{sketch_simple.pdf}
+\caption{\textbf{The mismatch encoding.}
+A example read is represented in the match alphabet. The symbol
+$\Downarrow$ represents a mismatch against the target sequence and
+the symbol $\square$ represents a match.}
+\label{fig:simple}
+\end{figure}
+
+The probability of occurrence of the symbol $\Downarrow$ is $p$, the error
+rate of the sequencer, and the probability of occurrence of the symbol
+$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
+weighted generating function are $pz$ and $qz$. Using the rule for
+concatenation, we see that the weighted generating function of a segment
+with $m$ symbols $\square$ is $(qz)^mpz$. The symbol $|$ has size 0, so
+the weighted generating function of a tail segment with $m$ symbols
+$\square$ is $(qz)^m$.
+
+The key insight is that reads without on-target seed correspond exactly
+with the reads that are made of segments and tail with fewer than $\gamma$
+symbols $\square$. The weighted generating function of such segments is
+$pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail
+is $1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can
+be encoded as a transfer matrix.
+
+Since there are two kinds of bjects, the segments and the tails, the
+dimensions of the transfer matrix are $2 \times 2$. A segment can be
+followed by another segment or by the tail. The tail can be followed by
+nothing. The expression of the transfer matrix $M(z)$ is thus
+\begin{equation*}
+\begin{blockarray}{ccc}
+   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
+\begin{block}{c[cc]}
+\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
+\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
+\scriptstyle{|} & 0 & 0 \\
+\end{block}
+\end{blockarray}.
+\end{equation*}
+
+In this case, it is possible to compute $M(z) + M(z)^2 + \ldots = M(z)
+\cdot (I-M(z))^{-1}$ directly and to extract the entry of interest.
+Here it is the top right term because the read can start with a segment or
+the tail (as if the read were prepended by an invisible segment), and it
+must end with the tail. It is easy to check that the weighted generating
+function is equal to
+\begin{equation}
+\label{eq:F}
+\frac{1+qz+\ldots+(qz)^{\gamma-1}}
+  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
+\end{equation}
+
+The function above can be represented as a Taylor series $a_0 + a_1z +
+a_2z^2 + \ldots$ By construction, $a_k$ is the quantity of interest,
+\textit{i.e.} the probability that a read of size $k$ does not contain an
+on-target seed, so we now need to extract those coefficients from
+expression (\ref{eq:F}). There are several methods to do so, the one we
+choose here is to build a recurrence equation to compute $a_k$ from the
+previous terms of the series $a_0, a_1, \ldots, a_{k-1}$.
+
+For this, we observe that equation (\ref{eq:F}) is equivalent to
+\begin{equation*}
+(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
+(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
+\end{equation*}
+Equating the terms of same degree on both side of the equal signs leads
+to the following relationship between the coefficients of the series:
+\begin{equation}
+\label{eq:PE0}
+a_k = 
+\begin{cases}
+1            &\quad\text{if } k < \gamma, \\
+1 -pq^\gamma &\quad\text{if } k = \gamma, \\
+a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
+\end{cases}
+\end{equation}
+
+The term of interest can be computed recursively using expression
+(\ref{eq:PE0}). This approach is very efficient because every iteration
+involves at most one multiplication and one subtraction. Also, the default
+floating-point arithmetic on modern computers gives sufficient precision
+to not worry about numeric instability for the problems considered here
+(we rarely need to compute those probabilities for reads above 500
+nucleotides).
+
+This example shows how the symbolic approach gives a non-trivial and yet
+very simple method to compute the probability that a read of size $k$ does
+not contain an on-target exact seed.
+
+\subsection{Example 2: on-target skip seeds}
+
+Another example that will turn out to be essential is to compute the
+probability that a read contains no on-target skip seed. We tackle the
+question with the same strategy as in the previous example, so we first
+recode the reads using a more useful alphabet to solve this problem.
+
+Once again, we need to know whether a nucleotide is a sequencing error,
+but this time we also need to know its phase in the repeated cycles of
+skipped positions. For this, we define the skip-$n$ alphabet
+$\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1, \ldots,
+\Downarrow_n \}$. As before, the symbol $\square$ represents a correct
+nucleotide and the symbol $|$ is a terminator appended at the end of the
+read. The symbols $\Downarrow_m$ ($m \in \mathbb{N}$) represent sequencing
+errors and $m$ indicates the number of nucleotides till the next
+non-skipped position (\textit{i.e.} $m=0$ for nucleotides immediately
+before a non-skipped position and $m=n$ for nucleotides at a non-skipped
+position).
+
+The segments in this alphabet are defined as sequences of 0 or more
+symbols $\square$ followed by any of the symbols $\Downarrow_m$. The tail
+is again defined as a sequence of 0 or more symbols $\square$ followed by
+the symbol $|$. Given that this decomposition is unique, we can again view
+a read as a sequence of segments with a tail. The example of
+Fig.~\ref{fig:simple} is showed again in Fig.~\ref{fig:skip}, where the
+segments in the mismatch alphabet have been replaced by the segments in
+the skip-3 alphabet,
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.85]{sketch_skip.pdf}
+\caption{\textbf{The skip encoding.}
+Legend.}
+\label{fig:skip}
+\end{figure}
+
+The probability of occurrence of a sequencing error is $p$ so the weighted
+generating function of all the symbols $\Downarrow_m$ are identical and
+equal to $pz$. The weighted generating function of the symbol $\square$ is
+again $qz$, and so the weighted generating function of a segment with $m$
+symbols $\square$ is again $(qz)^mpz$, and the weighted generating
+function of a tail segment with $m$ symbols $\square$ is again $(qz)^m$.
+
+Reads that do not contain any on-target skip seed can still contain
+segments with $\gamma$ or more symbols $\square$, so this case is more
+complex than the previous one. For instance, if there is a sequencing
+error in phase $r$, the status of the next $r$ nucleotides does not have
+any influence on the presence of an on-target seed. Indeed, there is no
+on-target seed upstream of the error by construction, and the next seed is
+scheduled to start $r$ nucleotides downstream of it, so the following
+segment or tail can have up to $\gamma+r-1$ symbols $\square$. So in order
+to enforce the absence of on-target seed, we have to adjust the maximum
+size of the segments depending on the preceding terminator.
+
+In the skip-$n$ alphabet, reads are composed of $n+2$ different kinds of
+objects: the segments terminated by the $n+1$ symbols $\Downarrow_0,
+\Downarrow_1, \ldots, \Downarrow_n$, plus the tail. The dimensions of the
+transfer matrix are thus $(n+2) \times (n+2)$.
+
+As mentioned above, after the $\Downarrow_r$ terminator, we can append
+segments with up to $\gamma+r-1$ symbols $\square$. The terminators of
+those $\gamma+r$ possible segments are distributed according to the laws
+of the modulo $n+1$ arithmetic. For instance, if the next segment has no
+symbol $\square$ then it is terminated by $\Downarrow_{r+1}$, but also if
+it has $n+1$ symbols $\square$ (assuming $n+1 < \gamma$). Specifying the
+transfer matrix thus involves a substantial amount of bookkeeping.
+
+\begin{equation*}
+\begin{blockarray}{cccccc}
+   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
+\begin{block}{c[ccccc]}
+\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
+\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
+\nd & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}
+\end{equation*}
+
+\begin{eqnarray}
+H_{r,s}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
+  \ldots + (qz)^{m(n+1)} \big), \\
+\notag
+  &\;& \text{where } x = r-s-1 \modulo{n+1},
+  \text{ and } m = \left\lfloor
+  \frac{\gamma+r-1-x}{n+1} \right\rfloor, \\
+J_r(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+r-1}.
+\end{eqnarray}
+
+The transfer matrix is now too complex to compute a close expression
+for $M(z)\cdot(I-M(z))^{-1}$. We return to the definition $M(z) + M(z)^2 +
+\ldots$ and we observe that the terms $M(z)^{k+2}, M(z)^{k+3}, \ldots$
+have no influence on the coefficients $a_0, a_1, \ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k$ segments (plus one obligatory
+tail). Since $M(z)^{s+1}$ contains the weighted generating functions of
+reads with exactly $s$ segments and one tail, $a_k$ cannot depend on
+$M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by induction that
+all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
+contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
++a_kz^k$ is $0$.
+
+So we can compute the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
+extract the entry of interest and then compute the terms of the Taylor
+expansion up to order $k$. But we can do better. Since we are only
+interested in the coefficients up to order $k$, we can perform all
+algebraic operation on truncated polynomials of order $k$, \textit{i.e.},
+we discard the coefficients of order $k+1$ or greater when multiplying two
+polynomials.
+
+But we can do even better. A read with $s$ segments contains $s$ errors.
+Since read errors typically have a small probability of occurrence, all
+the coefficients of $M(z)^{s+1}$ rapidly converge to $0$ as $s$ increases.
+So instead of computing the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
+we can interrupt the summation after a certain power of $M(z)$ because the
+terms become negligible.
+
+The number of errors $X$ in a read of size $k$ has a Binomial
+distribution $X \sim B(k,p)$. From~\cite{arratia1989tutorial} we can bound
+the probabilities of the tail with the expression
+
+\begin{equation}
+\label{eq:bound}
+Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-p)} -s\log
+\frac{s}{kp} \right).
+\end{equation}
+
+Using the formula above, we can thus bound the probability that a read has
+$s$ segments or more. We compute the matrices $M(z) + M(z)^2 + \ldots
++M(z)^{s+1}$ where the weighted generating functions have been replaced by
+truncated polynomials and we extract the entry at the first column and the
+last row. When the bound is lower than a set fraction $\varepsilon$ of the
+smallest coefficient, we stop the computations. Typically $\varepsilon =
+0.01$ so this method ensures that the probabilities that a read of size
+$k$ has no good seeds are accurate to within 1\%.
+
+\section{Exact seeds}
+
+\subsection{Computing $P(E_0 \cap E_1)$}
+\label{sec:cPE0E1}
+
+The event $E_0 \cap E_1$ is that the read contains no match of size $\geq
+\gamma$ for either the target or the first off-target sequence (here
+``first'' refers to any fixed off-target sequence). We compute the
+probability of this event with the strategy described in
+section~\ref{sec:symbolic}: we first recode the reads using a different
+alphabet and then we specify the transfer matrix of the reads that have no
+match for any sequence. The powers of this matrix hold the weighted
+generating function of interest, from which we extract the coefficients
+that correspond to the terms $P(E_0 \cap E_1)$ for reads of different
+sizes.
+
+The reads are recoded as sequences of symbols from the so-called ``+/-
+alphabet'' $\mathcal{A}_{+/-} = \{\square, |, \downarrow_{/1}^-,
+\downarrow_{/2}^-, \ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots,
+\Downarrow\}$. The symbols $\downarrow_{/j}^-$ signify that the nucleotide
+is a mismatch against the off-target sequence, the symbols
+$\downarrow_{/j}^+$ signify that it is a mismatch against the target, and
+the symbol $\Downarrow$ signifies that it is a mismatch against both. As
+before, every other nucleotide is replaced by the symbol $\square$, and
+the terminator $|$ is appended at the end of the read. We again define
+reads as sequences of segments, except that now the terminators are the
+symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
+
+The number $j$ in the symbol $\downarrow_{/j}^-$ indicates the relative
+position of the last mismatch against the target. Likewise, the number $j$
+in the symbol $\downarrow_{/j}^+$ indicates the relative position of the
+last mismatch against the off-target sequence. In other words, the index
+$j$ is the size of the match for the sequence that does not have a
+mismatch. The terminators thus capture the local state of the read. For
+instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
+mismatch against the off-target sequence, that it is a match for the
+target, and that the six previous nucleotides were also a match for the
+target.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.85]{sketch_dual.pdf}
+\caption{\textbf{The +/- encoding.}
+A example read is represented in the +/- alphabet. The symbols
+$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
+the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
+and the symbol $\Downarrow$ represent a mismatch against both. The index
+$j$ is the number of matches since the last mismatch for the other
+sequence. Note the presence of the terminator $|$ at the end of the read.
+}
+\label{fig:dual}
+\end{figure}
+
+We use the same error model as before: for every nucleotide, the read
+differs from the target with a fixed probability $p$ and the sequences
+differ from each other with a fixed probability $\mu$. With these
+definitions, the probability of occurrence of the symbol $\square$ is $a =
+(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
+symbols $\downarrow_{/j}^-$ is $b = (1-p)\mu$, that of the symbols
+$\downarrow_{/j}^+$ is $c = p\mu/3$, and the probability of occurrence of
+the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
+$a+b+c+d=1$.
+
+These definitions allow us to specify the weighted generating function of
+the segments. For instance, the weighted generating function of the
+first segment of Fig.~\ref{fig:dual} is $(az)bz$, and that of the second is
+$(az)^5cz$.
+
+To specify the transfer matrix, we must ensure that we eliminate
+the concatenations of two segments that would create a match of size $\geq
+\gamma$ for any of the two sequences. After defining
+\begin{equation*}
+\begin{gathered}
+r_i^+(z) = (az)^icz, \\
+r_i^-(z) = (az)^ibz, \\
+R_i(z) = \big( 1 + az + \ldots + (az)^i \big)dz, \\
+F_i(z) = 1 + az + \ldots + (az)^i,
+\end{gathered}
+\end{equation*}
+we can verify that the expression of the transfer matrix $L(z)$ is
+
+\begin{equation*}
+\begin{blockarray}{ccccccccc}
+   & \scriptstyle{\Downarrow} & \scriptstyle{\downarrow_{/1}^+} & 
+    \ldots & \scriptstyle{\downarrow_{\gamma-1}^+} &
+    \scriptstyle{\downarrow_{/1}^-} & \ldots &
+    \scriptstyle{\downarrow_{/\gamma-1}^-} & \scriptstyle{|} \\
+\begin{block}{c[cccccccc]}
+\scriptstyle{\Downarrow} & R_{\gamma-1}(z)  & r_0^+(z) & \ldots &
+    r_{\gamma-2}^+(z) & r_0^-(z) & \ldots & r_{\gamma-2}^-(z) &
+    F_{\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/1}^+} & R_{\gamma-2}(z) & & & & & & &
+    F_{\gamma-2}(z) \\
+\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
+    F_{\gamma-3}(z) \\
+\vdots & \vdots & & A(z) & & & B(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
+\scriptstyle{\downarrow_{/1}^-} & R_{\gamma-2}(z) & & & & & & &
+    F_{\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
+    F_{\gamma-2}(z) \\
+\vdots & \vdots & & C(z) & & & D(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
+\scriptstyle{|} & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are square matrices with
+$\gamma-1$ rows/columns, that are defined as
+
+\begin{equation*}
+A(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & 0 & r_0^+(z) & \ldots &
+    r_{\gamma-4}^+(z) & r_{\gamma-3}^+(z) \\
+\scriptstyle{\downarrow_{/2}^+} & 0 & 0 & \ldots &
+    r_{\gamma-5}^+(z) & r_{\gamma-4}^+(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & 0 & 0 & \ldots & 0 & r_0^+(z) \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+B(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
+   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
+   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & r_0^-(z) & r_1^-(z) & \ldots &
+    r_{\gamma-3}^-(z) & r_{\gamma-2}^-(z) \\
+\scriptstyle{\downarrow_{/2}^+} & r_0^-(z) & r_1^-(z) & \ldots &
+    r_{\gamma-3}^-(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & r_0^-(z) & r_1^-(z) &
+    \ldots & 0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & r_0^-(z) & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+C(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & r_{\gamma-2}^+(z) \\
+\scriptstyle{\downarrow_{/2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & r_0^+(z) & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+
+\begin{equation*}
+D(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
+    \scriptstyle{\downarrow_{/\gamma-1}^-} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & 0 & r_0^-(z) & \ldots &
+    r_{\gamma-4}^-(z) & r_{\gamma-3}^-(z) \\
+\scriptstyle{\downarrow_{/2}^-} & 0 & 0 & \ldots &
+    r_{\gamma-5}^-(z) & r_{\gamma-4}^-(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & 0 & 0 & \ldots & 0 & r_0^-(z) \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}.
+\end{equation*}
+
+The matrix $L(z)$ is relatively sparse and most non-zero entries are
+monomials. In addition, the number of rows/columns is approximately
+$2\gamma$, so the dimension of the matrix rarely exceeds 50. This means
+that computing the powers of $L(z)$ is not prohibitive and we do not need
+to design a Monte Carlo sampling scheme as we did before.
+
+That being said, we do not need to compute all the powers up to $k$
+because the terms vanish and become negligible after a certain iteration.
+Indeed, the terms of $L(z)^n$ are the weighted generating function of the
+reads without a match for either sequences and that contain exactly $n$
+segments. Of those $n$ segments, $n-1$ are terminated by a mismatch
+against at least one of the sequences. Following the rationale of
+section~\ref{sec:extract}, we define $\tilde{p}$ as the upper bound on the
+probability of a mismatch $\tilde{p} = \max\{b,c,d\}$, and updating
+formula (\ref{eq:bound}) we bound the probability that a read of size $k$
+contains $m = n-1$ or more mismatches
+
+\begin{equation*}
+Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-\tilde{p})} -m\log
+\frac{m}{k\tilde{p}} \right).
+\end{equation*}
+
+The strategy is the same as in section~\ref{sec:extract}: we compute the
+sum of matrices $L(z) + L(z)^2 + \ldots + L(z)^n$ where the weighted
+generating functions have been replaced by truncated polynomials and we
+extract the entry at the first column and the last row. When the bound is
+lower than a set fraction $\varepsilon$ of the coefficient of $z^k$, we
+stop the computations.
+
+\subsection{Corollary: this title sucks}
+
+The results of this section allow us to derive the probabilities of false
+positives when using seeds of fixed size $\gamma$.
+
+The complementary of the event $E_0$ is that the read contains at least
+one match of size $\geq \gamma$ for the target, implying that there is a
+seed for the target\footnote{This is the major difference with MEM seeds,
+where the logical implication does not hold.}. With our assumptions this
+means that the read is mapped to the correct location. The probability of
+outcome $i.$, a true positive, is thus $1-P(E_0)$.
+
+The event $E_0 \cap \ldots \cap E_N$ is that the read contains no match of
+size $\geq \gamma$ for either the targer or any of the off-target
+sequences. In other words, the probability of outome $iii.$, a false
+negative, is $P(E_0 \cap \ldots \cap E_N)$.
+
+The event $E_0 \cap (E_1 \cap \ldots \cap E_N)^c$ is that the read
+contains no match of size $\geq \gamma$ for the target, but contains a
+match of size $\geq \gamma$ for at least one of the off-target sequences.
+In other words, the probability of outcome $ii.$, a false positive, is
+$P(E_0) - P(E_0 \cap \ldots \cap E_N)$.
+
+It is easy to see that the three probabilities sum to 1, as expected.
+Below is a summary of the probabilities of interest.
+
+\begin{eqnarray}
+P(i.) &=& 1-P(E_0) \\
+P(ii.) &=& P(E_0) \cdot \left( 1- \left( \frac{P(E_0
+  \cap E_1)}{P(E_0)} \right)^N\right) \\
+P(iii.) &=& P(E_0) \cdot \left( \frac{P(E_0 \cap E_1)}{P(E_0)} \right)^N
+\end{eqnarray}
+
+In the expressions above, we have used equation (\ref{eq:PE0E1}) to
+highlight that these probabilities only depend on $P(E_0)$ and $P(E_0 \cap
+E_1)$, which can be computed precisely as shown in sections~\ref{sec:cPE0}
+and \ref{sec:cPE0E1}.
+
+Note that the formulas also hold when the target has no duplicate
+sequence. In this case, only outcome $i.$ and $iii.$ are possible, but
+since $N=0$ we see that the second probability is 0, as expected.
+
+\section{Skip seeds}
+
+\begin{algorithm}[H]
+\label{alg:mcmcskip}
+\SetAlgoLined
+\KwResult {Sample a read at random. Return 1 if the read contains a
+good seed, otherwise return 0.}
+  $\lambda \leftarrow 0$ \Comment*[r]{Current read size.}
+  $s_+ \leftarrow 0$ \Comment*[r]{Current + match size.}
+  $s_- \leftarrow 0$ \Comment*[r]{Current - match size.}
+  \While {$\lambda < k$}{
+    $i \leftarrow geom(a)-1$
+                \Comment*[r]{Double-match nucleotides.}
+    \If {$i > k-\lambda$}{
+      $i \leftarrow k-\lambda$ \;
+    }
+    $s_+ \leftarrow s_+ + i$ \;
+    $s_- \leftarrow s_- + i$ \;
+    \If {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
+      return 1 \;
+    }
+    $\lambda \leftarrow \lambda + i+1$ \;
+    $r \leftarrow rand(0,1-a)$
+                \Comment*[r]{Uniform random number.}
+    \uIf {$r < b$}{
+      $s_+ \leftarrow s_++1$ \;
+      $s_- \leftarrow -\lambda \mod n+1$ \;
+    }\uElseIf{$r < b+c$}{
+      $s_+ \leftarrow -\lambda \mod n+1$ \;
+      $s_- \leftarrow s_-+1$ \;
+    }\uElse{
+      $s_+ \leftarrow -\lambda \mod n+1$ \;
+      $s_- \leftarrow -\lambda \mod n+1$ \;
+    }
+}
+\eIf {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
+  return 1 \;
+}{
+  return 0 \;
+}
+\end{algorithm}
+
+\section{MEM seeds}
+
+Before going further, we need to move a practical consideration out of the
+way. It stands to reason that very short matches between the read and the
+genome are uninformative. For instance, $99.7\%$ of all the possible
+12-mers are present in the reference human genome, so an exact match of
+size 12 or lower says nothing about the location of the DNA fragment. In
+what follows, we will assume that subsequences of the read can only match
+the original DNA sequence and/or its duplicates, but not the rest of the
+genome. This will greatly simplify the exposition of the theory. We will
+discuss how to relax this assumption in section~\ref{sec:short_MEMs}.
 
@@
 
-\section{Computational strategy}
-\label{sec:symbolic}
-
-We turn to the question of translating the extended MEM segment
-representation to a construction of reads without good seeds. For this, we
-use tools borrowed from analytic combinatorics. We give the minimum amount
-of background that is necessary to expose the theory. A more detailed
-account of the strategy to compute seeding probabilities with analytic
-combinatorics has been published
-elsewhere~\cite{filion2017analytic,filion2018analytic}. We refer the
-readers interested in learning more about analytic combinatorics to the
-textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.
-
-\subsection{Weighted generating functions}
-
-The central object of analytic combinatorics is the \emph{generating
-function}.
-
-\begin{definition}
-\label{def:GF}
-Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
-size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
-generating function of $\mathcal{A}$ is defined as
-
-\begin{equation}
-\label{eq:GF1}
-A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
-\end{equation}
-where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
-respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
-\geq 0}$ such that
-
-\begin{equation*}
-A(z) = \sum_{k=0}^\infty a_k z^k.
-\end{equation*}
-
-By definition $a_k = \sum_{a \in A_k}w(a)$, where $A_k$ is the class of
-objects of size $k$ in $\mathcal{A}$. The number $a_k$ is the
-total weight of objects of size $k$.
-\end{definition}
-
-To give a concrete example, if the probability of the symbol
-$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
-containing only this symbol is $qz$. In our case, we are interested in the
-weighted generating function of reads without good seeds, so that the
-coefficient $a_k$ is equal to the probability that a read of size $k$ has
-no good seed.
-
-The motivation for introducing weighted generating functions is that
-operations on combinatorial objects translate into operations on their
-weighted generating functions. If $A(z)$ and $B(z)$ are the weighted
-generating functions of two mutually exclusive sets $\mathcal{A}$ and
-$\mathcal{B}$, the weighted generating function of $\mathcal{A} \cup
-\mathcal{B}$ is $A(z) + B(z)$, as evident from expression (\ref{eq:GF1}).
-Size and weight can be defined for pairs of objects in $\mathcal{A} \times
-\mathcal{B}$ as $|(a,b)| = |a| + |b|$ and $w(a,b) = w(a)w(b)$. In other
-words the sizes are added and the weights are multiplied.  With this
-convention, the weighted generating function of the Cartesian product
-$\mathcal{A} \times \mathcal{B}$ is $A(z)B(z)$. This simply follows from
-expression (\ref{eq:GF1}) and from
-
-\begin{equation*}
-A(z)B(z) =
-\sum_{a\in \mathcal{A}}w(a)z^{|a|} \sum_{b\in \mathcal{B}}w(b)z^{|b|}
-= \sum_{(a,b) \in \mathcal{A} \times \mathcal{B}} w(a)w(b)z^{|a|+|b|}.
-\end{equation*}
-
-These two operations are all we need in order to compute the weighted
-generating functions of the reads of interest. Addition corresponds to
-creating a new family by merging reads from families $\mathcal{A}$ and
-$\mathcal{B}$. Multiplication corresponds to creating a new family by
-concatenating reads from families $\mathcal{A}$ and $\mathcal{B}$.
-
-\subsection{Analytic representation}
-
-The symbolic method is a strategy belonging to the general \emph{analytic
-combinatorics} framework. The idea is to combine simple combinatorial
-objects into more complex objects. Each combinatorial operation on the
-objects corresponds to a mathematical operation on their weighted
-generating functions. One can thus obtain the weighted generating function
-of complex objects, and from there extract their probability of
-occurrence.
-
-The combinatorial construction of reads using the extended MEM alphabet
-gives us a method to express their weighted generating function from the
-simpler generating function of segments. More specifically, we can
-describe complex sequences of segments through a so-called \emph{transfer
-matrix} $M(z)$ that specifies which pairs of consecutive segments are
-allowed and which are forbidden in the reads of interest.
-
-The segment types are identified by their terminators arranged in a
-predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
-weighted generating function of segments with the $j$-th terminator that
-can be inserted immediately after segments with the $i$-th terminator. The
-entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
-are the weighted generating functions of the reads of all sizes using all
-the allowed combinations of segments.
-
-The entry of interest from $M(z)$ is the one that corresponds to the
-terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
-any size whose first segment can be inserted after $\downarrow_{/0}$, and
-whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
-the end of the read, so it must be the final symbol of every valid read.
-As for the first segment, observe that all the segments that can follow
-the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
-postion of the read\footnote{One could have introduced a special start
-symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
-the reads.}.
-
-In summary, the transfer matrix $M(z)$ contains the weighted generating
-function of interest. Our next task is thus to specify $M(z)$.
-
 \section{The transfer matrix}
 \label{sec:transfer_mat}
-
-\subsection{Error model and divergence of the duplicates}
-
-We assume that the target sequence has $N$ duplicates, so that there are
-$N$ off-target sequences. We further assume that duplication was
-instantaenous and that all $N+1$ sequences diverge independently of each
-other at a constant rate. In other words, we ignore the complications due
-to the genealogy of the duplication events and we simply assume that at
-each position, any given duplicate is identical to the target with
-probability $1-\mu$. If it is not, we assume that the duplicate has any of
-the remaining three nucleotides with equal probability (\textit{i.e.} each
-is found with probability $\mu/3$).
-
-We also assume that the sequencing instrument has a constant substitution
-rate $p$, and that insertions and deletions never occur. In case of a
-substitution, we again assume that the remaining three nucleotides are
-equally likely.
-
-Read errors are always mismatches against the target (because we assume
-that the target is the true sequence), and they match each off-target
-sequence with probability $\mu/3$. Correct nucleotides are always a match
-for the target, and they match each off-target sequence with probability
-$1-\mu$.
-
-We are now ready to describe the transfer matrix $M(z)$. We do this entry
-by entry, in order of increasing complexity.
 
@@
 
-\subsection{The expression of the transfer matrix}
+\subsection{Expression of the transfer matrix}
 \label{sec:expression_of_M}
@@
 
+The special case $N=0$ is worth describing in detail. This implies that
+the target has no duplicate, so case $ii.$ is excluded (false positive),
+and only cases $i.$ and $iii.$ are possible (true positive and false
+negative, respectively). The transfer matrix $M(z)$ simplifies to the
+extent that we can directly compute the weighted generating function of
+reads without a good seed as
+\begin{equation}
+\tag{\ref{eq:F}}
+F(z) = \frac{1+qz+\ldots+(qz)^{\gamma-1}}
+  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
+\end{equation}
+
+Expression (\ref{eq:F}) was shown in~\cite{filion2018analytic} to be the
+weighted generating function of
+
 \section{Computing the probabilities}
@@
 $|$. The expression of $M(z)$ is too complex to compute this function
-directly. Instead, we return to the definition $M(z) \cdot (I-M(z))^{-1} =
+directly.
+
+Instead, we return to the definition $M(z) \cdot (I-M(z))^{-1} =
 M(z) + M(z)^2 + \ldots$ and observe that the terms $M(z)^{k+2},
@@
 \KwResult {Sample a read at random. Return 1 if the read contains a
-good seed, otherwise return 0.}
-  $\lambda \leftarrow k$ \Comment*[r]{Size remaining to generate.}
+good MEM seed, otherwise return 0.}
+  $\lambda \leftarrow 0$ \Comment*[r]{Current read size.}
   $m \leftarrow 0$ \Comment*[r]{Current number of hard masks.}
-  \While {$\lambda > 0$}{
+  \While {$\lambda < k$}{
     $i \leftarrow geom(p)-1$ \Comment*[r]{Error-free nucleotides.}
-    \eIf {$i \geq \lambda$}{
-      \eIf{$\lambda < \gamma$} {
+    \eIf {$i \geq k-\lambda$}{
+      \eIf{$k-\lambda < \gamma$} {
         return 0\; }{
-        $h \leftarrow binom(m,(1-\mu)^\lambda)$
+        $h \leftarrow binom(m,(1-\mu)^{k-\lambda})$
             \Comment*[r]{Surviving hard masks.}
@@
         $m \leftarrow s + binom(N-s, \mu/3)$\;
-        $\lambda \leftarrow \lambda - (i+1)$\;}
+        $\lambda \leftarrow \lambda + i+1$\;}
     }
@@
 
-\subsection{The probability of outcome $iii.$}
+\subsection{Probability of outcome $iii.$}
 
@@
 a read of size $k$ does not contain a match of size $\geq \gamma$ for the
-target. Then the following equation can be used to compute $a_k$ by
-recurrence:
-
-\begin{equation}
-\label{eq:PE0}
-a_k = 
-\begin{cases}
-1            &\quad\text{if } k < \gamma, \\
-1 -pq^\gamma &\quad\text{if } k = \gamma, \\
-a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
-\end{cases}
-\end{equation}
+target. This case was already treated in section~\ref{sec:example_exact}.
 
@@
 
-\subsection{Computing $P(E_0 \cap E_1)$}
-\label{sec:cPE0E1}
-
-The event $E_0 \cap E_1$ is that the read contains no match of size $\geq
-\gamma$ for either the target or the first off-target sequence (here
-``first'' refers to any fixed off-target sequence). We compute the
-probability of this event with the strategy described in
-section~\ref{sec:symbolic}: we first recode the reads using a different
-alphabet and then we specify the transfer matrix of the reads that have no
-match for any sequence. The powers of this matrix hold the weighted
-generating function of interest, from which we extract the coefficients
-that correspond to the terms $P(E_0 \cap E_1)$ for reads of different
-sizes.
-
-The reads are recoded as sequences of symbols from the so-called ``+/-
-alphabet'' $\mathcal{A}_{+/-} = \{\square, |, \downarrow_{/1}^-,
-\downarrow_{/2}^-, \ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots,
-\Downarrow\}$. The symbols $\downarrow_{/j}^-$ signify that the nucleotide
-is a mismatch against the off-target sequence, the symbols
-$\downarrow_{/j}^+$ signify that it is a mismatch against the target, and
-the symbol $\Downarrow$ signifies that it is a mismatch against both. As
-before, every other nucleotide is replaced by the symbol $\square$, and
-the terminator $|$ is appended at the end of the read. We again define
-reads as sequences of segments, except that now the terminators are the
-symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
-
-The number $j$ in the symbol $\downarrow_{/j}^-$ indicates the relative
-position of the last mismatch against the target. Likewise, the number $j$
-in the symbol $\downarrow_{/j}^+$ indicates the relative position of the
-last mismatch against the off-target sequence. In other words, the index
-$j$ is the size of the match for the sequence that does not have a
-mismatch. The terminators thus capture the local state of the read. For
-instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
-mismatch against the off-target sequence, that it is a match for the
-target, and that the six previous nucleotides were also a match for the
-target.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.85]{sketch_dual.pdf}
-\caption{\textbf{The +/- encoding.}
-A example read is represented in the +/- alphabet. The symbols
-$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
-the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
-and the symbol $\Downarrow$ represent a mismatch against both. The index
-$j$ is the number of matches since the last mismatch for the other
-sequence. Note the presence of the terminator $|$ at the end of the read.
-}
-\label{fig:dual}
-\end{figure}
-
-We use the same error model as before: for every nucleotide, the read
-differs from the target with a fixed probability $p$ and the sequences
-differ from each other with a fixed probability $\mu$. With these
-definitions, the probability of occurrence of the symbol $\square$ is $a =
-(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
-symbols $\downarrow_{/j}^-$ is $b = (1-p)\mu$, that of the symbols
-$\downarrow_{/j}^+$ is $c = p\mu/3$, and the probability of occurrence of
-the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
-$a+b+c+d=1$.
-
-These definitions allow us to specify the weighted generating function of
-the segments. For instance, the weighted generating function of the
-first segment of Fig.~\ref{fig:dual} is $(az)bz$, and that of the second is
-$(az)^5cz$.
-
-To specify the transfer matrix, we must ensure that we eliminate
-the concatenations of two segments that would create a match of size $\geq
-\gamma$ for any of the two sequences. After defining
-\begin{equation*}
-\begin{gathered}
-r_i^+(z) = (az)^icz, \\
-r_i^-(z) = (az)^ibz, \\
-R_i(z) = \big( 1 + az + \ldots + (az)^i \big)dz, \\
-F_i(z) = 1 + az + \ldots + (az)^i,
-\end{gathered}
-\end{equation*}
-we can verify that the expression of the transfer matrix $L(z)$ is
-
-\begin{equation*}
-\begin{blockarray}{ccccccccc}
-   & \scriptstyle{\Downarrow} & \scriptstyle{\downarrow_{/1}^+} & 
-    \ldots & \scriptstyle{\downarrow_{\gamma-1}^+} &
-    \scriptstyle{\downarrow_{/1}^-} & \ldots &
-    \scriptstyle{\downarrow_{/\gamma-1}^-} & \scriptstyle{|} \\
-\begin{block}{c[cccccccc]}
-\scriptstyle{\Downarrow} & R_{\gamma-1}(z)  & r_0^+(z) & \ldots &
-    r_{\gamma-2}^+(z) & r_0^-(z) & \ldots & r_{\gamma-2}^-(z) &
-    F_{\gamma-1}(z) \\
-\scriptstyle{\downarrow_{/1}^+} & R_{\gamma-2}(z) & & & & & & &
-    F_{\gamma-2}(z) \\
-\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
-    F_{\gamma-3}(z) \\
-\vdots & \vdots & & A(z) & & & B(z) & & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
-\scriptstyle{\downarrow_{/1}^-} & R_{\gamma-2}(z) & & & & & & &
-    F_{\gamma-1}(z) \\
-\scriptstyle{\downarrow_{/2}^+} & R_{\gamma-3}(z) & & & & & & &
-    F_{\gamma-2}(z) \\
-\vdots & \vdots & & C(z) & & & D(z) & & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
-\scriptstyle{|} & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
-\end{block}
-\end{blockarray},
-\end{equation*}
-where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are square matrices with
-$\gamma-1$ rows/columns, that are defined as
-
-\begin{equation*}
-A(z) = 
-\begin{blockarray}{cccccc}
-   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
-    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
-    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
-\begin{block}{c[ccccc]}
-\scriptstyle{\downarrow_{/1}^+} & 0 & r_0^+(z) & \ldots &
-    r_{\gamma-4}^+(z) & r_{\gamma-3}^+(z) \\
-\scriptstyle{\downarrow_{/2}^+} & 0 & 0 & \ldots &
-    r_{\gamma-5}^+(z) & r_{\gamma-4}^+(z) \\
-\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-2}^+} & 0 & 0 & \ldots & 0 & r_0^+(z) \\
-\scriptstyle{\downarrow_{/\gamma-1}^+} & 0 & 0 & \ldots & 0 & 0 \\
-\end{block}
-\end{blockarray},
-\end{equation*}
-
-\begin{equation*}
-B(z) = 
-\begin{blockarray}{cccccc}
-   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
-   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
-   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
-\begin{block}{c[ccccc]}
-\scriptstyle{\downarrow_{/1}^+} & r_0^-(z) & r_1^-(z) & \ldots &
-    r_{\gamma-3}^-(z) & r_{\gamma-2}^-(z) \\
-\scriptstyle{\downarrow_{/2}^+} & r_0^-(z) & r_1^-(z) & \ldots &
-    r_{\gamma-3}^-(z) & 0 \\
-\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-2}^+} & r_0^-(z) & r_1^-(z) &
-    \ldots & 0 & 0 \\
-\scriptstyle{\downarrow_{/\gamma-1}^+} & r_0^-(z) & 0 & \ldots & 0 & 0 \\
-\end{block}
-\end{blockarray},
-\end{equation*}
-
-\begin{equation*}
-C(z) = 
-\begin{blockarray}{cccccc}
-   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
-    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
-    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
-\begin{block}{c[ccccc]}
-\scriptstyle{\downarrow_{/1}^-} & r_0^+(z) & r_1^+(z) & \ldots &
-    r_{\gamma-3}^+(z) & r_{\gamma-2}^+(z) \\
-\scriptstyle{\downarrow_{/2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
-    r_{\gamma-3}^+(z) & 0 \\
-\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
-    0 & 0 \\
-\scriptstyle{\downarrow_{/\gamma-1}^-} & r_0^+(z) & 0 & \ldots & 0 & 0 \\
-\end{block}
-\end{blockarray},
-\end{equation*}
-
-\begin{equation*}
-D(z) = 
-\begin{blockarray}{cccccc}
-   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
-    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
-    \scriptstyle{\downarrow_{/\gamma-1}^-} \\
-\begin{block}{c[ccccc]}
-\scriptstyle{\downarrow_{/1}^-} & 0 & r_0^-(z) & \ldots &
-    r_{\gamma-4}^-(z) & r_{\gamma-3}^-(z) \\
-\scriptstyle{\downarrow_{/2}^-} & 0 & 0 & \ldots &
-    r_{\gamma-5}^-(z) & r_{\gamma-4}^-(z) \\
-\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
-\scriptstyle{\downarrow_{/\gamma-2}^-} & 0 & 0 & \ldots & 0 & r_0^-(z) \\
-\scriptstyle{\downarrow_{/\gamma-1}^-} & 0 & 0 & \ldots & 0 & 0 \\
-\end{block}
-\end{blockarray}.
-\end{equation*}
-
-The matrix $L(z)$ is relatively sparse and most non-zero entries are
-monomials. In addition, the number of rows/columns is approximately
-$2\gamma$, so the dimension of the matrix rarely exceeds 50. This means
-that computing the powers of $L(z)$ is not prohibitive and we do not need
-to design a Monte Carlo sampling scheme as we did before.
-
-That being said, we do not need to compute all the powers up to $k$
-because the terms vanish and become negligible after a certain iteration.
-Indeed, the terms of $L(z)^n$ are the weighted generating function of the
-reads without a match for either sequences and that contain exactly $n$
-segments. Of those $n$ segments, $n-1$ are terminated by a mismatch
-against at least one of the sequences. Following the rationale of
-section~\ref{sec:extract}, we define $\tilde{p}$ as the upper bound on the
-probability of a mismatch $\tilde{p} = \max\{b,c,d\}$, and updating
-formula (\ref{eq:bound}) we bound the probability that a read of size $k$
-contains $m = n-1$ or more mismatches
-
-\begin{equation*}
-Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-\tilde{p})} -m\log
-\frac{m}{k\tilde{p}} \right).
-\end{equation*}
-
-The strategy is the same as in section~\ref{sec:extract}: we compute the
-sum of matrices $L(z) + L(z)^2 + \ldots + L(z)^n$ where the weighted
-generating functions have been replaced by truncated polynomials and we
-extract the entry at the first column and the last row. When the bound is
-lower than a set fraction $\varepsilon$ of the coefficient of $z^k$, we
-stop the computations.
-
-\subsection{The probability of false positives}
+\subsection{Probability of false positives}
 
@@
 explained in sections~\ref{sec:cPE0E1} and \ref{sec:cPE0}.
+
+When the target has no duplicate sequence, there can be no false positive
+so this probability must be 0. In this case $N=0$ so according to formula
+(\ref{eq:PE0E1}) outcome $iii.$ has probability $P(E_0)$.
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
+
+XXXX
 
@@
 
-\subsection{MEM seeds at random locations}
+\subsection{MEM seeds for other sequences}
 \label{sec:short_MEMs}

### 2019-03-03 MEM.tex

@@
 
+The most popular heuristic for this problem is to first identify short
+regions of high similiarity between the read and the genome, and then run
+an exact alignment algorithm locally at those regions. The first step is a
+filtration called ``seeding'' and the regions of high similarity are
+called ``seeds''. This is for instance the general strategy of the popular
+alignment algorithm BLAST~\cite{pmid2231712}.
+
+If seeds can be discovered fast, we can quickly zoom in on a small set of
+candidate locations and reduce the input size of the (slow) alignment
+algorithm. The disadvantage is that the target location is not always in
+the candidate set. In this case, it cannot be discovered and the outcome
+is either $ii.$ or $iii.$, which are both erroneous.
+
+The slowness of the alignment algorithm imposes a trade-off between speed
+and accuracy. If the seeding step is set to produce many candidates, the
+target is likely to be among them and it will be discovered. The downside
+is that this will take long. Conversely, if the seeding step is set to
+produce few candidates, the process will run faster but the target is less
+likely to be in the candidate list so it is less likely to be discovered.
+
+Importantly, the terms of this trade-off depend on the algorithm. Some
+seeding methods are better than others (see~\cite{pmid20460430,
+sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
+brejova2003vector,pmid18684737,pmid15359419} for examples of design,
+see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
+see~\cite{navarro2001guided} for a more general review of filtration
+methods in pattern matching). The upshot is that we can develop faster
+mapping algorithms at no cost for accuracy if we find better seeding
+strategies.
+
+The seeding framework was developed for the best location problem, but it
+also applies to the true location problem. Actually, the process is
+exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
+location. In outcome $i.$ called a ``true positive'', the DNA fragment is
+mapped to the \emph{true} location. In outcome $ii.$ called a ``false
+positive'', the DNA fragment is mapped to an incorrect location. Outcome
+$iii.$ called a ``false negative'' is the same as before and means that
+the DNA fragment is not mapped at all.
+
+We will make two assumptions that will simplify the discussion. The first
+is that the true location is always the best. We will further comment on
+this and explain how to relax this assumption in section
+section~\ref{sec:not_the_best}. Until then, we consider that the true
+location problem is the same as the best location problem. The second
+assumption is that the best location is always discovered if it is in the
+candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
+determined by the seeding process. For this reason, we will sometimes
+refer to them as the outcomes of the seeding step, even though strictly
+speaking they are the outcomes of the whole mapping pipeline\footnote{This
+can be confusing because some authors refer to the off-target seeds as
+false positives. When we say that the outcome of the seeding step is a
+false positive, we mean that \emph{all} the seeds are off-target.}.
+
 That being said, we do not need to compute all the powers up to $k$
@@
 \end{equation*}
+
+In general, seeds do not have to be exact matches{}. For
+instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
+seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
+covered by a patent (reference US20070088510A1).}. However, in this
+document, we will always assume that a seed is an exact match with no gap.
+In other words, a seed is a perfect match between a part of the read and a
+genomic site.
+
+The rest of the document will focus on a seeding heuristic known as ``MEM
+seeding'' (where MEM stands for Maximal Exact Match). This approach gives
+good empirical results when mapping short reads (it is for instance the
+strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
+Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
+but there is presently no theory to justify its performance. A deeper
+understanding of MEM seeding would be useful to calibrate the
+speed/accuracy trade-off of this heurisitic.
 

### 2019-03-31 MEM.tex
First complete sketch with exact, skip and MEM seeds.

--- 2019-03-03 MEM.tex
@@
 
-\title{Estimating mismapping rates in seed-based algorithms
-for read mapping}
+\title{Estimating mapping quality when using seeding heuristics for
+DNA alignment}
 
@@
 ancient bones; identifying the donor of a biological sample; diagnosing
-genetic diseaes at birth; finding compatible graft donors \textit{etc}.
+genetic diseases at birth; finding compatible graft donors \textit{etc}.
 
@@
 
-Because sequencing errors can convert a DNA fragment into one of its
-duplicates, the \emph{true} location of a DNA fragment is not always the
-\emph{best} (as measured by the identity between the sequence and the
-candidate location). This leads to the following question: how can we
-identify the \emph{best} location of the fragment in the genome?
+Since repeats play a central role in the problem at hand, we give the
+terms ``targets'' and ``duplicates'' a meaning that will facilitate the
+exposition of the theory.
+
+\begin{definition}
+The target is the DNA fragment that was sequenced. Duplicates are
+sequences of the genome that are homologous to the target.
+\end{definition}
+
+Because sequencing errors can convert the sequence of the target into one
+of its duplicates, the \emph{true} location of a DNA fragment is not
+always the \emph{best} (as measured by the identity between the sequence
+and the candidate location). This leads to the following question: how can
+we identify the \emph{best} location of the fragment in the genome?
 
@@
 research has largely benefited from the improvement of computer hardware
-and from the deveopment of indexing algorithms. Literature on this topic
-is already abudant. Some examples of seeding algorithms are given in
-references~\cite{sun2005designing,pmid11934743,xu2006optimizing,
-pmid17044164,brejova2003vector,pmid18684737,pmid15359419}.
+and from the development of indexing algorithms. Literature on this topic
+is already abundant. Some examples of seeding algorithms are given in
+references~\cite{sun2005designing, pmid11934743, xu2006optimizing,
+kucherov2005multiseed, brejova2003vector, pmid18684737, pmid15359419}.
 References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
@@
 \subsection{The two types of seeding failure}
+\label{sec:twotypes}
 
@@
 \begin{enumerate}[i)]
-\item on target if the candidate set contains the target,
-\item off target if the candidate set contains a duplicate but
+\item ``on target'' if the candidate set contains the target,
+\item ``off target'' if the candidate set contains a duplicate but
 not the target, 
-\item null if the candidate set contains neither.
+\item ``null'' if the candidate set contains neither.
 \end{enumerate}
@@
 
-If a seed-based algorithm returns an incorrect location, then the best
-location was never tested with an exact alignment. If the error is given a
-high confidence, then it must be the best in the candidate set and it must
-be similar to the read. This narrows down the cases of mismapping, or more
-exactly high confidence mismapping to off-target seeding.
-
-This observation is fundamental. In order to estimate the mismapping rate
+If a seed-based mapping algorithm returns an incorrect location, then it
+must be the case that the best location was never tested in the second
+phase, otherwise it would have been singled out by the exact alignment
+algorithm.
+
+This observation is fundamental: in order to estimate the mismapping rate
 of an algorithm, we must know how often the associated seeding heuristic
-is off target. Yet, this information is missing for nearly all seeding
-methods, meaning that current mappers cannot be properly calibrated.
+is off target. For lack of theoretical framework, this information is
+missing for nearly all existing seeding methods, meaning that current
+mappers cannot be properly calibrated.
 
 The focus of this article is to develop a theory to estimate the
-probability of off-target seeding under the most popular seeding
-heuristics. Our previous work pioneered a method to compute seeding
-probabilities but it did not distinguish off-target from null
-seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
-not provide a way to control the mismapping rate. Other authors
+probability that seeding is off target under the most popular seeding
+heuristics for short reads. Our previous work pioneered a method to
+compute seeding probabilities but it did not distinguish off-target from
+null seeding~\cite{filion2017analytic,filion2018analytic}, and therefore
+did not provide a way to control the mismapping rate. Other authors
 investigated the reliability of mapping algorithms~\cite{pmid23872968},
@@
 hit, recognizing that solving the problem requires taking into
-consideration the repeat structure of the genome.
-
-\section{Types of seeds}
+consideration the repeat structure of the genome. Here we address this
+problem in general terms and we provide a practical way to estimate the
+mismapping rate in complex genomes.
+
+The rest of this document is organized to present the content in a
+didactic manner. The order of the examples is meant to facilitate the
+aquisition of the main concepts, meaning that we sometimes go back and
+forth between different types of seeding problems. We first give some
+background on the different types of seeding methods considered here, we
+then introduce the computational strategy based on transfer matrices, and
+then delve into each specific seeding method. Finally, we provide a
+statistical method to estimate the parameters of the problem from the
+data.
+
+\section{Seeds}
 
@@
 genome, a particular sequence motif or a structured pattern of matches.
-Also, the term is not always short for ``exact match''. For instance, the
-algorithm PatternHunter~\cite{pmid11934743} describes ``spaced seeds''
-that tolerate mismatches\footnote{For the anecdote, the method is covered
-by a patent (reference US20070088510A1).}. To avoid potential confusion,
+Also, the term does not always imply an exact match. For instance, the
+algorithm PatternHunter~\cite{pmid11934743} uses ``spaced seeds''
+that tolerate mismatches\footnote{Spaced seeds are covered
+by patent US20070088510A1.}. To avoid potential confusion,
 we will adopt the following convention.
@@
 When a seed matches a particular location of the genome, we will say that
-it is a seed \emph{for} that location. For instance, we will often refer
-to a seed for the target or a seed for one of its duplicates.
+it is a ``seed for that location''. For instance, we will often use
+expressions such as ``a seed for the target'' or ``a seed for a
+duplicate''.
 
@@
 exists practical methods to solve it, even when the reference genome is
-very large~\cite{ferragina2000opportunistic, ferragina2005indexing}. Such
-algorithms are crucial for the present theory, but describing them
-falls outside the scope of the document. Looking ahead to
-section~\ref{sec:est}, let us only mention that this can be done with an
-algorithm called the \emph{backward
-search}~\cite{ferragina2000opportunistic}. We refer the interested reader
-to the relevant
-literature~\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736} and
-here on we will assume that seeds are always known without ambiguity [ALL
-THE CITED LITERATURE IS FOR MEM SEEDS, WE NEED TO FIND ARTICLES THAT TREAT
-THE CASE OF OTHER SEEDS].
+very large.
+
+These algorithms are crucial for the present theory, but describing them
+in depth is outside the scope of this document. Let us just mention that
+all the methods belong to a family known as exact offline string matching
+algorithms, where ``offline'' means that sequences are looked up in an
+index instead of the genome itself. Online methods can be used when the
+reference genome is not indexed~\cite{faro2013exact}, but this case is of
+little relevance in the present context.
+
+The index is usually a hash table or a variant of the so-called
+FM-index~\cite{pmid20460430}. Hash tables are typically used to index
+$k$-mers, which makes them useful to search seeds of set size $k$,
+(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
+algorithms). In contrast, the FM-index has no set size so it can be used
+to search seeds of different lengths~\cite{ferragina2000opportunistic,
+ferragina2005indexing}. It is a compact data structure based on the suffix
+array~\cite{manber1993suffix} and on the Burrows-Wheeler
+transform~\cite{burrows1994block}, emulating a suffix trie with a much
+smaller memory footprint. Queries in the FM-index are performed with an
+algorithm known as the \emph{backward
+search}~\cite{ferragina2000opportunistic}.
+
+Other methods can be efficient (\textit{e.g.}, running a bisection on the
+suffix array~\cite{dobin2013star}) but the FM-index is currently the most
+popular choice for seeding methods. For MEM seeds described below, it is
+even the only practical option~\cite{pmid24336412, pmid25399029,
+pmid23349213, pmid19389736}, so we implicitly assume that seeding is
+performed by running the backward search. This will only have practical
+consequences in section~\ref{sec:est}, where we propose a method to
+estimate the parameters that are required to compute the probabilities of
+interest. For the rest, we simply assume that seeds are known at all times
+without ambiguity because this problem has several practical solutions.
 
@@
 set. This seeding heuristic was used in the first version of
-BLASTN~\cite{pmid2231712}, but is no longer used because they produce many
+BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
 overlapping matches in the same regions, as we are about to illustrate.
@@
 by an imperfect sequencer. The sequenced DNA fragment comes from a region
-of the genome that has three extra duplicates, and the read has three
-miscalled nucleotides (indicated by stars).
+of the genome that has three duplicates (there are four homologous
+sequences in total), and the read has three miscalled nucleotides
+(indicated by stars).
 
@@
 \caption{\textbf{Example of exact seeds.}
-A sequenced DNA fragment (Read) is shown above its original form in the
-genome (Target).}
+A sequenced DNA fragment (Read) is shown above the actual molecule
+(Target), and above homologous sequences also present in the genome
+(Duplicates). Sequencing errors are indicated by a star, each marked
+\texttt{T} was actually an \texttt{A} that was misread. The mismatches
+between the genomic sequences and the read are indicated in bold. The
+exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
+above the read. The matching regions in the genomic sequences are
+shadowed in grey. Several overlapping seeds accumulate at the center of
+the read, which is typical for this seeding method.}
 \label{fig:exact_seed_example}
@@
 
-Fig.~\ref{fig:exact_seed_example} also illustrates the seeding process
-with exact seeds of size $\gamma=7$. It is clear that erroneous
-nucleotides can also belong to exact seeds because they sometimes match
-one of the duplicates. However, this is not the case of erroneous
-nucleotides that are also mismatches for \emph{all} the duplicates, like
-the second sequencing error in this example.
-
-Thie middle of the read shows some cluttering where consecutive seeds
+The exact seeds with $\gamma=7$ are shown above the sequence of the read.
+Observe that erroneous nucleotides can belong to exact seeds because they
+sometimes match one of the duplicates. For instance, the first sequencing
+error matches all the duplicates and belongs to a seed for the first one.
+However, erroneous nucleotides that are mismatches for \emph{all} the
+duplicates cannot belong to a seed. This is the case of the second
+sequencing error in this example, where there is a local deficit of
+seeds.
+
+Thie middle of the read shows some clutter where consecutive seeds
 match consecutive sequences at the same location. This is typical for
@@
 skipped between every seed. This is a way to reduce the amount of
-overlapping matches at the same location, but the cost is a loss of
-sensitivity. This seeding heuristic is used by
-Bowtie2~\cite{pmid22388286}, where seeds have size $\gamma=16$ and fall
-every 10 nucleotides (9 nucleotides are skipped between seeds).
+overlapping matches at the same location, at the cost of sensitivity. This
+seeding heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds
+have size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
+skipped). We will refer to seeds where $n$ nucleotides are skipped as
+``skip-$n$ seeds''. For instance, Bowtie2 uses skip-9 seeds.
 
@@
 The sequences and the annotations are the same as in
-Fig.~\ref{fig:exact_seed_example}, but here one nucleotide is skipped
-betwen the seeds. In other words, seeds can never start on a nucleotide
-with an even index. To highlight the difference with
-Fig.~\ref{fig:exact_seed_example}, the missing seeds are represented by
-dotted lines. Skipping nucleotides reduces the clutter at the center of
-the read, but it also removes the only seed for the second duplicate.}
+Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
+words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
+highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
+missing seeds are represented by dotted lines. Skipping nucleotides
+reduces the clutter at the center of the read, but it also removes the
+only seed for the second duplicate.}
 \label{fig:skip_seed_example}
@@
 
-Fig.~\ref{fig:skip_seed_example} shows the seeding process on the read of
-Fig.~\ref{fig:exact_seed_example} when using skip seeds instead of exact
-seeds. Here the size is still $\gamma=7$ but 1 extra nucleotide is skipped
-between seeds. This amounts to removing every other seed. The consequence
-is that there are fewer overlapping matches at the center of the read, but
-the only seed for the second duplicate was lost. This is a rather positive
+Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
+replaced by skip-2 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
+Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
+seeds. This amounts to removing every second seed. The consequence is that
+there are fewer overlapping matches at the center of the read, but the
+only seed for the second duplicate is lost. This is a rather positive
 outcome because there is one off-target location fewer in the candidate
-set, but in general the same might happen to the target.
+set, but the same might happen to the target.
 
 It is clear that skipping nucleotides reduces the sensitivity of the
-seeding, but to what extent? One could test this empirically, but it
-depends on the seed length, the amount of nucleotides that are skipped,
-the error rate of the sequencer and the size of the reads. The present
-theory will allow us to get the answer automatically.
+seeding step, but to what extent? One could test this empirically, but the
+answer depends on the seed length, the number of nucleotides that are
+skipped, the error rate of the sequencer and the size of the reads. The
+present theory will allow us to make general statements about the
+performance of skip seeds against exact seeds in different contexts.
 
@@
 
-We will now focus on a seeding heuristic known as ``MEM seeding'' (where
-MEM stands for Maximal Exact Match). This approach gives good empirical
-results when mapping short reads (it is for instance the
-strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
-Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
-but there is presently no theory to justify its performance. A deeper
-understanding of MEM seeding would be useful to calibrate the
-speed/accuracy trade-off of this heurisitic.
-
-Before developing a formal theory, let us explain in more detail what
-Maximal Exact Matches are and how MEM seeding works.
+MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
+to define. Unlike exact seeds and skip seeds, their size is variable as we
+are about to see. They are used in BWA-MEM~\cite{li2013aligning} where
+they give good empirical results (but note that the actually seeding
+scheme uses more complex Super Maximal Exact Matches). To describe MEM
+seeds, let us first focus on MEMs.
 
 \begin{definition}
-A Maximal Exact Match (MEM) seed is a subsequence of a read that is
-present in the genome of interest and that cannot be extended --- either
-because the read ends or because the extended subsequence is not in the
-genome.
+A Maximal Exact Match (MEM) is a subsequence of the read that is present
+in the reference genome and that cannot be extended --- either because the
+read ends or because the extended subsequence is not in the genome.
 \end{definition}
 
-Fig.~\ref{fig:MEM_example} shows a concrete example where a read is
-aligned to candidate locations. The target (the true location of the
-read) is first potential location. The locations below correspond to
-duplicate sequences. The mismatches between the genome and the read are
-highlighted in bold. The read has three errors indicated by stars, so
-\texttt{T}, \texttt{T} and \texttt{C} should actually be \texttt{A},
-\texttt{A} and \texttt{G} (\textit{i.e.} as in the target sequence).
+A \emph{MEM seed} is simply a MEM of minimum size $\gamma$.
+Fig.~\ref{fig:MEM_example} shows what happens when exact seeds or skip
+seeds are replaced by MEM seeds on the read of
+Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the center
+of the read has disappeared because consecutive matches are fused into a
+single MEM seed.
 
@@
 \caption{\textbf{Example of MEM seeds.}
-A DNA fragment is sequenced with an imperfect instrument. It comes from
-the top genomic location (Truth), but three errors were introduced
-(indicated by stars). The sequence of the DNA fragment has three
-duplicates at other genomic locations, shown below the target. Mismatches
-between the read and the candidate sequence are highlighted in bold. MEM
-seeds are indicated by grey boxes.}
+The sequences and the annotations are the same as in
+Fig.~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
+size $\gamma=7$. The clutter at the center of the read has disappeared and
+there is at least one seed for each sequence.}
 \label{fig:MEM_example}
@@
 
-The MEM seeds are indicated by grey boxes. Technically, they are part of
-the read, but they are shown on the genome to highlight the position of
-the match. One MEM seed cannot cover a smaller one (otherwise the smaller
-would not be a MEM seed), but two consecutive MEM seeds can overlap, as in
-the case of the frst two from the left. Note that overlapping MEM seeds
-must match distinct sequences of the genome (otherwise neither of them
-would be a MEM seed). Consecutive MEM seeds can also be disjoint, as in
-the case of the second and the third from the left. In this case, they
-must be separated by a nucleotide that is a mismatch against \emph{all}
-the duplicates.
-
-\section{Model and computational strategy}
+Two consecutive MEM seeds can overlap, in which case they always match
+distinct sequences of the genome (otherwise neither of them would be a MEM
+seed). There does not have to be any overlap though, because a nucleotide
+can be a mismatch against \emph{all} the sequences, like the second read
+error for instance.
+
+Note that a MEM does not always match a single sequence of the genome. For
+instance, the rightmost MEM seed matches two distinct genomic sequences.
+This case motivates the following definition, which will play an important
+role later.
+
+\begin{definition}
+A \emph{strict} MEM seed has a single match in the genome.
+A \emph{shared} MEM seed has several matches in the genome.
+\end{definition}
+
+Compared to seeds of fixed size, MEM seeds have two counter-intuitive (and
+undesirable) properties. The first is that there are cases where there
+cannot be any good seed, even when changing the sensitivity parameter
+$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
+though there is a single sequencing error, the read has no MEM seed for
+the target. Lowering $\gamma$ does not change this, so there is no way to
+discover the true location in this case (even though it is the best
+location).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{full_masking_example.pdf}
+\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
+The read, the MEM seeds and the sequences are represented as in
+Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
+locations at the bottom effectively hide the target, so it cannot
+be discovered. This can occur even when the true location is the
+best and when there is a single error on the read.}
+\label{fig:full_masking_example}
+\end{figure}
+
+The second counter-intuitive property is that shortening a read can
+sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
+shows an example of this case. There is no MEM seed for the target, but
+there would be one if the read were two nucleotides shorter on the right
+side. Indeed, in this case there would be a shared MEM seed matching the
+first and the second locations (provided $\gamma \leq 12$).
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=1]{short_vs_long_example.pdf}
+\caption{\textbf{Issues with MEM seeds: too long reads.}
+The read, the MEM seeds and the sequences are represented as in
+Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
+read were two nucleotides shorter. The true location is hidden by
+the last two nucleotides.}
+\label{fig:short_vs_long}
+\end{figure}
+
+These examples show that MEM seeds can perform worse than seeds of fixed
+size. MEM seeds yield fewer candidates and therefore speed up the mapping
+process, but the question is at what cost for accuracy? The theory
+developed here will allow us to compute the probability that a read has
+no MEM seed for the target and thus that the true location is missed at
+the seeding stage.
+
+\section{Model and strategy}
 \label{sec:symbolic}
 
-\subsection{Sequencing error and duplications}
-
-Now that this is clarified, we need to model the sequencing and the
-duplication processes so that we can compute the probabilities of the
-events of interest. We assume that the sequencing instrument has a
-constant substitution rate $p$, and that insertions and deletions never
-occur. When a substitution occurs, we assume that the instrument is
-equally likely to output any one of the remaining three nucleotides. This
-is a coarse-grain description of the error spectrum of the Illumina
-technology~\cite{pmid21576222}.
+\subsection{Sequencing errors and divergence of the duplicates}
+\label{sec:error}
+
+We now need to model the sequencing and duplication processes so that
+we can compute the probabilities of the events of interest. We assume that
+the sequencing instrument has a constant substitution rate $p$, and that
+insertions and deletions never occur. When a substitution occurs, we
+assume that the instrument is equally likely to output any one of the
+remaining three nucleotides. This corresponds more or less to the error
+spectrum of the Illumina sequencing technology~\cite{pmid21576222}.
 
@@
 rate. In other words, we ignore the complications due to the genealogy of
-the duplication events and we simply assume that at each position, any
-given duplicate is identical to the target with probability $1-\mu$. If it
-is not, we assume that the duplicate has any of the remaining three
-nucleotides with equal probability (\textit{i.e.} each is found with
+the duplication events and we simply assume that at each nucleotide
+position, any given duplicate is identical to the target with probability
+$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
+one the remaining three nucleotides (\textit{i.e.} each is found with
 probability $\mu/3$).
@@
 assume that the target is the true sequence), and they match each
-off-target sequence with probability $\mu/3$. Correct nucleotides are
-always a match for the target, and they match each off-target sequence
-with probability $1-\mu$.
+off-target sequence with probability $\mu/3$. In contrast, correct
+nucleotides are always matches for the target, and they match each
+off-target sequence with probability $1-\mu$.
+
+Before going further, we also need to move a practical consideration out
+of the way. Seeds can match any sequence of the genome, not just the
+target or the duplicates. However, we will ignore matches in the rest of
+the genome because such random matches are unlikely to cause a mapping
+error when seeding is off target, contrary to matches in duplicates.
+Neglecting those will greatly simplify the exposition of the theory
+without loss of generality. We will explain in
+section~\ref{sec:short_MEMs} how to deal with this practical case and and
+how to identify those matches as off target. Until then, we will consider
+that the target and the duplicates are the only sequences in the reference
+genome.
 
@@
 The central object of analytic combinatorics is the generating function,
-and for our particular purpose we will use a special kind known as
-\emph{weighted generating function}.
+and for our purpose we will use a special kind known as \emph{weighted
+generating function}.
 
@@
 \label{def:GF}
-Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
-size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
-generating function of $\mathcal{A}$ is defined as
+Let $\mathcal{A}$ be a set of combinatorial objects such that $a \in
+\mathcal{A}$ has a size $|a| \in \mathbb{N}$ and a weight $w(a) \in
+\mathbb{R}^+$. The weighted generating function of $\mathcal{A}$ is
+defined as
 
@@
 \end{equation}
-where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
-respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
-\geq 0}$ such that
+Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k \geq 0}$ such
+that
 
@@
 
-To give a concrete example, if the probability of the symbol
-$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
-containing only this symbol is $qz$. In our case, we are interested in the
-weighted generating function of reads without good seeds, so that the
-coefficient $a_k$ is equal to the probability that a read of size $k$ has
-no good seed.
+To give an example, assume that a particular symbol, say $\Downarrow$, has
+a probability of occurence equal to $p$. The weighted generating function
+of the word containing only this symbol is $pz$. The weight of the word is
+its probability (here equal to $p$) and the size is its length in number
+of symbols (here 1).
+
+In this document we will focus on the weighted generating function $A(z)$
+of the set $\mathcal{A}$ of reads that do not have on-target seeds
+(\textit{i.e.} reads for which seeding is either off target or null). The
+weight of a read is its probability of occurrence and the size $k$ is its
+number of nucleotides. The coefficient $a_k$ is thus the proportion of
+reads of size $k$ that do not have an on-target seed, which is the
+quantity of interest.
 
@@
 
-The symbolic method is a strategy belonging to the general \emph{analytic
-combinatorics} framework. The idea is to combine simple combinatorial
-objects into more complex objects. Each combinatorial operation on the
-objects corresponds to a mathematical operation on their weighted
-generating functions. One can thus obtain the weighted generating function
-of complex objects, and from there extract their probability of
-occurrence.
-
-The combinatorial construction of reads using the extended MEM alphabet
-gives us a method to express their weighted generating function from the
-simpler generating function of segments. More specifically, we can
-describe complex sequences of segments through a so-called \emph{transfer
-matrix} $M(z)$ that specifies which pairs of consecutive segments are
-allowed and which are forbidden in the reads of interest.
-
-The segment types are identified by their terminators arranged in a
-predefined oder. The entry of $M(z)$ at coordinates $(i,j)$ is the
-weighted generating function of segments with the $j$-th terminator that
-can be inserted immediately after segments with the $i$-th terminator. The
-entries of the matrix $M(z) + M(z)^2 + \ldots = M(z) \cdot (I-M(z))^{-1}$
-are the weighted generating functions of the reads of all sizes using all
-the allowed combinations of segments.
-
-The entry of interest from $M(z)$ is the one that corresponds to the
-terminators $\downarrow_{/0}$ and $|$ because it represents the reads of
-any size whose first segment can be inserted after $\downarrow_{/0}$, and
-whose last segment is terminated by $|$. Indeed, the $|$ terminator marks
-the end of the read, so it must be the final symbol of every valid read.
-As for the first segment, observe that all the segments that can follow
-the $\downarrow_{/0}$ symbol are exactly those that can be at the frst
-postion of the read\footnote{One could have introduced a special start
-symbol. It is equivalent to an invisible $\downarrow_{/0}$ prepending
-the reads.}.
-
-In summary, the transfer matrix $M(z)$ contains the weighted generating
-function of interest. Our next task is thus to specify $M(z)$.
-
-More didactic descriptions of the strategy to compute seeding
-probabilities using analytic combinatorics have been published
-elsewhere~\cite{filion2017analytic,filion2018analytic}. The interested
-readers can also discover more about analytic combinatorics in the
-standard textbooks~\cite{flajolet2009analytic,sedgewick2013introduction}.
+The analytic combinatorics framework relies on a strategy referred to as
+the \emph{symbolice method}~\cite{sedgewick2013introduction}. The idea is
+to combine simple objects into more complex objects. Each combinatorial
+operation on the objects corresponds to a mathematical operation on their
+weighted generating functions. One can thus obtain the weighted generating
+function of complex objects, whose coefficients $a_k$ are the
+probabilities of interest.
+
+As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
+we recode the reads in alphabets of custom symbols and we specify a
+construction plan of the reads using a so-called \emph{transfer matrix}
+$M(z)$. The transfer matrix specifies which types of segments can follow
+each other in the reads of interest: the entry at coordinates $(i,j)$ is
+the weighted generating function of segments of type $j$ that can be
+appended to segments of type $i$.
+
+The entry of $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$
+at coordinates $(i,j)$ is the weighted generating function of all the
+reads that end with a segment of type $j$ and that can be appended to a
+segment of type $i$. If segments of types $i$ and $j$ are exactly those
+that can be placed at the beginning and at the end of the reads, then this
+entry is the weighted generating function of all the reads without
+on-target seed. The examples below will clarify the key steps of this
+strategy.
+
+A complete description of the strategy to compute seeding probabilities
+with the symbolic method can be found in~\cite{filion2017analytic,
+filion2018analytic}. The interested readers can also read more about
+analytic combinatorics in the popular
+textbooks~\cite{flajolet2009analytic, sedgewick2013introduction}.
 
 \subsection{Example 1: on-target exact seeds}
-\label{sec:sample_exact}
+\label{sec:example_exact}
 
@@
 example was described in detail in~\cite{filion2017analytic} and
-in~\cite{filion2018analytic}, but we repeat it here with a formalism that
-fits the present manuscript.
+in~\cite{filion2018analytic}, but we repeat it here with a different
+formalism to fit the present manuscript.
 
@@
 irrelevant. Indeed, the read has an on-target seed if and only if it
-contains a stretch of $\gamma$ nucleotides without error. So the main step
-is to recode the read as sequence of correct or erroneous nucleotides.
+contains a stretch of $\gamma$ nucleotides without error. For this reason,
+we recode reads as sequences of correct or erroneous nucleotides.
 
 We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
-\Downarrow,\}$, where $\square$ represents a correct nucleotide,
+\Downarrow\}$, where $\square$ represents a correct nucleotide,
 $\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
@@
 
-This recoding allows us to segment the read in \emph{segments} defined as
-a sequence of 0 or more symbols $\square$ followed by a symbol
+This recoding allows us to partition the read in \emph{segments} defined
+as a sequence of 0 or more symbols $\square$ followed by a symbol
 $\Downarrow$. The last part of the read is called the tail and it is
 terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
-not correspond to a nucleotide, it has size 0. This decomposition is
-unique, so we can view a read as a sequence of segments with a tail,
+not correspond to a nucleotide so it has size 0. Since the decomposition
+is unique, we can view a read as a sequence of segments with a tail,
 instead of a sequence of nucleotides.
@@
 Fig.~\ref{fig:simple} shows an example of decomposition in segments.
-On-target seeds cannot contain sequencing errors, therefore they must lie
-inside a segment or inside the tail. So the size of the segments and of
-the tail indicates whether the read contains an on-target seed.
+On-target seeds cannot contain sequencing errors, therefore they must be
+completely embedded inside a segment or inside the tail. So the size of
+the segments and of the tail indicates whether the read contains an
+on-target seed or not.
 
@@
 \caption{\textbf{The mismatch encoding.}
-A example read is represented in the match alphabet. The symbol
-$\Downarrow$ represents a mismatch against the target sequence and
-the symbol $\square$ represents a match.}
+An example read is represented in the mismatch alphabet. The symbol
+$\Downarrow$ represents a mismatch against the target (it is an erroneous
+nucleotide) and the symbol $\square$ represents a match (it is a correct
+nucleotide). The symbolic sequence of the target is represented below,
+where an open square stands for a match and a closed square stands for a
+mismatch.}
 \label{fig:simple}
@@
 concatenation, we see that the weighted generating function of a segment
-with $m$ symbols $\square$ is $(qz)^mpz$. The symbol $|$ has size 0, so
-the weighted generating function of a tail segment with $m$ symbols
-$\square$ is $(qz)^m$.
-
-The key insight is that reads without on-target seed correspond exactly
-with the reads that are made of segments and tail with fewer than $\gamma$
-symbols $\square$. The weighted generating function of such segments is
-$pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail
-is $1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can
-be encoded as a transfer matrix.
-
-Since there are two kinds of bjects, the segments and the tails, the
-dimensions of the transfer matrix are $2 \times 2$. A segment can be
-followed by another segment or by the tail. The tail can be followed by
-nothing. The expression of the transfer matrix $M(z)$ is thus
+with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
+the weighted generating function of a tail segment with $i$ symbols
+$\square$ is $(qz)^i$.
+
+The key insight is that reads without on-target seed are exactly the reads
+that are made of segments and tail with fewer than $\gamma$ symbols
+$\square$. The weighted generating function of such segments is $pz \cdot
+\big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
+$1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
+encoded in a transfer matrix.
+
+Since there are two kinds of objects, the segments and the tails, the
+dimension of the transfer matrix is $2 \times 2$. A segment can be
+followed by another segment or by the tail. The tail cannot be followed by
+anything. The expression of the transfer matrix $M_0(z)$ is thus
 \begin{equation*}
@@
 
-In this case, it is possible to compute $M(z) + M(z)^2 + \ldots = M(z)
-\cdot (I-M(z))^{-1}$ directly and to extract the entry of interest.
-Here it is the top right term because the read can start with a segment or
-the tail (as if the read were prepended by an invisible segment), and it
-must end with the tail. It is easy to check that the weighted generating
-function is equal to
+The types of objects are identified by their terminator, \textit{i.e.} by
+their last symbol. Segments are identified by the terminator $\Downarrow$
+and tails by the terminator $|$. The terminators are written in the
+margins of the transfer matrix for readability.
+
+We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
+(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
+term --- associated with terminators $\Downarrow$ and $|$. To see why,
+observe that every read can be prepended by a segment and only by a
+segment (not by a tail). Thus, reads are precisely the sequences of
+segments that can follow the symbol $\Downarrow$ and that are terminated
+by a tail, whose weighted generating function is the top right entry of
+the matrix. It is easy to check that this term is equal to
 \begin{equation}
@@
 \frac{1+qz+\ldots+(qz)^{\gamma-1}}
-  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
+  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} = a_0 + a_1z + a_2z^2 +
+\ldots
 \end{equation}
 
-The function above can be represented as a Taylor series $a_0 + a_1z +
-a_2z^2 + \ldots$ By construction, $a_k$ is the quantity of interest,
-\textit{i.e.} the probability that a read of size $k$ does not contain an
-on-target seed, so we now need to extract those coefficients from
-expression (\ref{eq:F}). There are several methods to do so, the one we
-choose here is to build a recurrence equation to compute $a_k$ from the
-previous terms of the series $a_0, a_1, \ldots, a_{k-1}$.
-
-For this, we observe that equation (\ref{eq:F}) is equivalent to
+The coefficients of the Taylor series $a_0 + a_1z + a_2z^2 + \ldots$ are
+unknown. But by construction, $a_k$ is the quantity of interest,
+\textit{i.e.} it is the probability that a read of size $k$ does not
+contain an on-target seed, so we now need to extract the coefficients
+from expression (\ref{eq:F}). There are several methods to do so; the one
+we choose here is to build a recurrence equation to compute $a_k$ from the
+previous terms of the series $a_0, a_1, \ldots, a_{k-1}$. For this, we
+observe that for $z \neq 1/q$, equation (\ref{eq:F}) is equivalent to
 \begin{equation*}
@@
 \end{equation*}
-Equating the terms of same degree on both side of the equal signs leads
+Balancing the terms of same degree on both sides of the equality leads
 to the following relationship between the coefficients of the series:
 \begin{equation}
-\label{eq:PE0}
+\label{eq:recur}
 a_k = 
@@
 
-The term of interest can be computed recursively using expression
-(\ref{eq:PE0}). This approach is very efficient because every iteration
+The terms of interest can be computed recursively using expression
+(\ref{eq:recur}). This approach is very efficient because every iteration
 involves at most one multiplication and one subtraction. Also, the default
@@
 
-This example shows how the symbolic approach gives a non-trivial and yet
-very simple method to compute the probability that a read of size $k$ does
+This example shows how the symbolic approach yields a non-trivial and yet
+simple algorithm to compute the probability that a read of size $k$ does
 not contain an on-target exact seed.
@@
 \subsection{Example 2: on-target skip seeds}
-
-Another example that will turn out to be essential is to compute the
-probability that a read contains no on-target skip seed. We tackle the
-question with the same strategy as in the previous example, so we first
-recode the reads using a more useful alphabet to solve this problem.
+\label{sec:example_skip}
+
+Here we go through another essential example, namely we devise a method to
+compute the probability that a read contains no on-target skip seed. Using
+the same strategy as in the previous example, we start by recoding the
+reads using a specialized alphabet to solve this problem.
 
@@
 \Downarrow_n \}$. As before, the symbol $\square$ represents a correct
-nucleotide and the symbol $|$ is a terminator appended at the end of the
-read. The symbols $\Downarrow_m$ ($m \in \mathbb{N}$) represent sequencing
-errors and $m$ indicates the number of nucleotides till the next
-non-skipped position (\textit{i.e.} $m=0$ for nucleotides immediately
-before a non-skipped position and $m=n$ for nucleotides at a non-skipped
+nucleotide and the symbol $|$ is a terminator appended to the end of the
+read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
+errors and $i$ indicates the number of nucleotides till the next
+non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
+before a non-skipped position and $i=n$ for nucleotides at a non-skipped
 position).
@@
 The segments in this alphabet are defined as sequences of 0 or more
-symbols $\square$ followed by any of the symbols $\Downarrow_m$. The tail
+symbols $\square$ followed by any of the symbols $\Downarrow_i$. The tail
 is again defined as a sequence of 0 or more symbols $\square$ followed by
@@
 a read as a sequence of segments with a tail. The example of
-Fig.~\ref{fig:simple} is showed again in Fig.~\ref{fig:skip}, where the
-segments in the mismatch alphabet have been replaced by the segments in
-the skip-3 alphabet,
+Fig.~\ref{fig:simple} is shown again in Fig.~\ref{fig:skip}, where
+segments in the mismatch alphabet have been replaced by segments in the
+skip-3 alphabet.
 
@@
 \caption{\textbf{The skip encoding.}
-Legend.}
+The read of Fig.~\ref{fig:simple} is represented in the skip-3 alphabet.
+The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
+target (they are erroneous nucleotides) and the symbol $\square$
+represents a match (it is a correct nucleotide). The positions of the
+reads with a vertical bar indicate non-skipped positions (the potential
+start of a seed). The number $i$ in $\Downarrow_i$ indicates the distance
+to the next non-skipped position in number of nucleotides. For $i=0$ the
+next nucleotide is at a non-skipped position. Other graphical
+representations are as in Fig.~\ref{fig:simple}.}
 \label{fig:skip}
@@
 
-The probability of occurrence of a sequencing error is $p$ so the weighted
-generating function of all the symbols $\Downarrow_m$ are identical and
-equal to $pz$. The weighted generating function of the symbol $\square$ is
-again $qz$, and so the weighted generating function of a segment with $m$
-symbols $\square$ is again $(qz)^mpz$, and the weighted generating
-function of a tail segment with $m$ symbols $\square$ is again $(qz)^m$.
-
-Reads that do not contain any on-target skip seed can still contain
-segments with $\gamma$ or more symbols $\square$, so this case is more
-complex than the previous one. For instance, if there is a sequencing
-error in phase $r$, the status of the next $r$ nucleotides does not have
-any influence on the presence of an on-target seed. Indeed, there is no
-on-target seed upstream of the error by construction, and the next seed is
-scheduled to start $r$ nucleotides downstream of it, so the following
-segment or tail can have up to $\gamma+r-1$ symbols $\square$. So in order
-to enforce the absence of on-target seed, we have to adjust the maximum
-size of the segments depending on the preceding terminator.
-
-In the skip-$n$ alphabet, reads are composed of $n+2$ different kinds of
-objects: the segments terminated by the $n+1$ symbols $\Downarrow_0,
-\Downarrow_1, \ldots, \Downarrow_n$, plus the tail. The dimensions of the
-transfer matrix are thus $(n+2) \times (n+2)$.
-
-As mentioned above, after the $\Downarrow_r$ terminator, we can append
-segments with up to $\gamma+r-1$ symbols $\square$. The terminators of
-those $\gamma+r$ possible segments are distributed according to the laws
-of the modulo $n+1$ arithmetic. For instance, if the next segment has no
-symbol $\square$ then it is terminated by $\Downarrow_{r+1}$, but also if
-it has $n+1$ symbols $\square$ (assuming $n+1 < \gamma$). Specifying the
-transfer matrix thus involves a substantial amount of bookkeeping.
-
+The probability of occurrence of a sequencing error is $p$, so every
+symbol $\Downarrow_i$ has the same weighted generating function $pz$ ---
+implicitly assuming that the next non-skipped position is at distance $i$,
+otherwise the weighted generating function is 0. The weighted generating
+function of the symbol $\square$ is again $qz$, and so the weighted
+generating function of a segment with $i$ symbols $\square$ is $(qz)^ipz$.
+Likewise, the weighted generating function of a tail with $i$ symbols
+$\square$ is $(qz)^i$.
+
+Reads that do not contain any on-target skip seed can contain segments
+with $\gamma$ or more symbols $\square$, so this case is more complex than
+the previous one. For instance, if there is a sequencing error $i$
+nucleotides before the next non-skipped position, the status of the next
+$i$ nucleotides does not have any influence on the presence of an
+on-target seed. Indeed, there is no on-target seed upstream of the error
+(by construction) and the next seed is scheduled to start $i$ nucleotides
+downstream, so there can be up to $\gamma+i-1$ symbols $\square$ in a row.
+In order to enforce the absence of on-target seeds, we thus have to adjust
+the maximum size of the segments depending on the preceding terminator.
+
+In the skip-$n$ alphabet, reads are composed segments terminated by the
+$n+1$ symbols $\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n$, plus the
+tail. The dimensions of the transfer matrix are thus $(n+2) \times (n+2)$.
+
+As mentioned above, after the terminator $\Downarrow_i$, we can append
+segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
+those $\gamma+i$ possible segments are distributed according to the laws
+of the arithmetic modulo $n+1$. For $i > 0$ for instance, if the next
+segment has no symbol $\square$ then it is terminated by
+$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
+$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
+involves a substantial amount of bookkeeping. One can check that the
+expression of the transfer matrix $M_n(z)$ is
 \begin{equation*}
@@
 \end{block}
-\end{blockarray}
+\end{blockarray},
 \end{equation*}
-
 \begin{eqnarray}
-H_{r,s}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
+H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
   \ldots + (qz)^{m(n+1)} \big), \\
 \notag
-  &\;& \text{where } x = r-s-1 \modulo{n+1},
+  &\;& \text{where } x = i-j-1 \modulo{n+1},
   \text{ and } m = \left\lfloor
-  \frac{\gamma+r-1-x}{n+1} \right\rfloor, \\
-J_r(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+r-1}.
+  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
+J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
 \end{eqnarray}
 
-The transfer matrix is now too complex to compute a close expression
-for $M(z)\cdot(I-M(z))^{-1}$. We return to the definition $M(z) + M(z)^2 +
-\ldots$ and we observe that the terms $M(z)^{k+2}, M(z)^{k+3}, \ldots$
-have no influence on the coefficients $a_0, a_1, \ldots, a_k$.
+The weighted generating function of interest is the top right entry of the
+matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
+why, observe that at the start of every read, the next nucleotide is a
+non-skipped position, so every read can be prepended by segments
+terminated with $\Downarrow_0$ and only by those. Thus, reads are
+precisely the sequences of segments that can follow the symbol
+$\Downarrow_0$ and that are terminated by a tail, whose weighted
+generating function is the top right entry of the matrix.
+
+$M_n(z)$ is too complex to compute a close expression of
+$M_n(z)\cdot(I-M_n(z))^{-1}$. We return to the definition $M_n(z) +
+M_n(z)^2 + \ldots$ and we observe that the terms $M_n(z)^{k+2},
+M_n(z)^{k+3}, \ldots$ have no influence on the coefficients of the
+weighted generating function $a_0, a_1, \ldots, a_k$.
 
@@
 contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
-+a_kz^k$ is $0$.
-
-So we can compute the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
++a_kz^k$ is strictly $0$.
+
+So we can compute the matrix $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$,
 extract the entry of interest and then compute the terms of the Taylor
@@
 interested in the coefficients up to order $k$, we can perform all
-algebraic operation on truncated polynomials of order $k$, \textit{i.e.},
+algebraic operations on truncated polynomials of order $k$, \textit{i.e.},
 we discard the coefficients of order $k+1$ or greater when multiplying two
@@
 
-But we can do even better. A read with $s$ segments contains $s$ errors.
-Since read errors typically have a small probability of occurrence, all
-the coefficients of $M(z)^{s+1}$ rapidly converge to $0$ as $s$ increases.
-So instead of computing the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$,
-we can interrupt the summation after a certain power of $M(z)$ because the
-terms become negligible.
+But we can do even better. A read with $s$ segments contains $s$ errors,
+so all the entries of $M_n(z)^{s+1}$ are dominated by $p^s$ and they
+rapidly converge to $0$ as $s$ increases. Instead of computing the matrix
+$M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$, we can interrupt the
+summation after a certain power of $M_n(z)$ because the terms become
+negligible.
 
@@
 Using the formula above, we can thus bound the probability that a read has
-$s$ segments or more. We compute the matrices $M(z) + M(z)^2 + \ldots
-+M(z)^{s+1}$ where the weighted generating functions have been replaced by
-truncated polynomials and we extract the entry at the first column and the
-last row. When the bound is lower than a set fraction $\varepsilon$ of the
-smallest coefficient, we stop the computations. Typically $\varepsilon =
-0.01$ so this method ensures that the probabilities that a read of size
-$k$ has no good seeds are accurate to within 1\%.
-
-\section{Exact seeds}
-
-\subsection{Computing $P(E_0 \cap E_1)$}
-\label{sec:cPE0E1}
-
-The event $E_0 \cap E_1$ is that the read contains no match of size $\geq
-\gamma$ for either the target or the first off-target sequence (here
-``first'' refers to any fixed off-target sequence). We compute the
-probability of this event with the strategy described in
-section~\ref{sec:symbolic}: we first recode the reads using a different
-alphabet and then we specify the transfer matrix of the reads that have no
-match for any sequence. The powers of this matrix hold the weighted
-generating function of interest, from which we extract the coefficients
-that correspond to the terms $P(E_0 \cap E_1)$ for reads of different
-sizes.
-
-The reads are recoded as sequences of symbols from the so-called ``+/-
-alphabet'' $\mathcal{A}_{+/-} = \{\square, |, \downarrow_{/1}^-,
-\downarrow_{/2}^-, \ldots, \downarrow_{/1}^+, \downarrow_{/2}^+, \ldots,
-\Downarrow\}$. The symbols $\downarrow_{/j}^-$ signify that the nucleotide
-is a mismatch against the off-target sequence, the symbols
-$\downarrow_{/j}^+$ signify that it is a mismatch against the target, and
-the symbol $\Downarrow$ signifies that it is a mismatch against both. As
-before, every other nucleotide is replaced by the symbol $\square$, and
-the terminator $|$ is appended at the end of the read. We again define
-reads as sequences of segments, except that now the terminators are the
-symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$, $\Downarrow$ and $|$.
-
-The number $j$ in the symbol $\downarrow_{/j}^-$ indicates the relative
-position of the last mismatch against the target. Likewise, the number $j$
-in the symbol $\downarrow_{/j}^+$ indicates the relative position of the
-last mismatch against the off-target sequence. In other words, the index
-$j$ is the size of the match for the sequence that does not have a
-mismatch. The terminators thus capture the local state of the read. For
+$s$ segments or more. We compute the matrices $M_n(z) + M_n(z)^2 + \ldots
++M_n(z)^{s+1}$ where the weighted generating functions have been replaced
+by truncated polynomials and we extract the entry at the first column and
+the last row. When the bound is lower than a set fraction $\varepsilon$ of
+the smallest coefficient, we stop the computations. Typically $\varepsilon
+= 0.01$ so this method ensures that the probabilities that a read of size
+$k$ has no on-target skip seed are accurate to within 1\%.
+
+\begin{remark}
+Observe that when $n=0$ the matrix $M_n(z)$ is identical to the matrix
+$M_0(z)$ of section~\ref{sec:example_exact}. This is consistent
+with the fact that exact seeds are skip-$n$ seeds for which $n=0$.
+\end{remark}
+
+\section{Off-target exact seeds}
+\label{sec:offexact}
+
+We now turn our attention to the problem of computing the probability that
+the seeding process is off target when using exact seeds --- recall from
+section~\ref{sec:twotypes} that off-target seeding means that the
+candidate set contains a duplicate but not the target.
+
+If there is $N = 0$ duplicate, seeding cannot be off target, it can only
+be on target or null. So from here we assume that the target has $N \geq
+1$ duplicates. Let $S_0$ denote the event that there is an on-target seed
+and let $S_j$ denote the event that there is a seed for the $j$-th
+duplicate. We are thus interested in computing $P\big(\overline{S}_0 \cap
+(S_1 \cup \ldots \cup S_N)\big)$, where $\overline{S}_j$ denotes the
+complement of the event $S_j$. For this, we first observe that
+\begin{equation}
+\label{eq:pcomp}
+P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
+P(\overline{S}_0) - P\big(\overline{S}_0 \cap \overline{S}_1 \cap \ldots
+\cap \overline{S}_N\big).
+\end{equation}
+
+Since the duplicates are assumed to evolve independently of each other and
+through the same mutagenesis process, the events are conditionally
+independent (we do not have full independence because $S_0$ depends on
+sequencing errors only whereas the events $S_j$ depend on sequencing
+errors and on the divergence between duplicates). We can thus write
+\begin{eqnarray}
+\notag
+P(\overline{S}_0 \cap \ldots \cap \overline{S}_N) &=&
+  P(\overline{S}_0) \cdot P(\overline{S}_1 \cap \ldots \cap
+  \overline{S}_N | \overline{S}_0) \\
+\label{eq:Pnull}
+  &=& P(\overline{S}_0) \cdot P(\overline{S}_1 | \overline{S}_0)^N =
+  P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0 \cap
+\overline{S}_1)}{P(\overline{S}_0)} \right)^N.
+\end{eqnarray}
+Combinging the two equations above, we obtain
+\begin{equation}
+\label{eq:Poff}
+P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
+P(\overline{S}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
+\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N.
+\end{equation}
+
+The probability that seeding is off target thus depends on only two
+quantitites. We have already seen how to compute $P(\overline{S}_0)$ in
+section~\ref{sec:example_exact} using recursive expression
+(\ref{eq:recur}). We now need to find a way to compute $P(\overline{S}_0
+\cap \overline{S}_1)$.
+
+\subsection{The dual encoding}
+\label{sec:dual}
+
+The event $\overline{S}_0 \cap \overline{S}_1$ is that the read contains
+no seed for either the target or the first duplicate --- numbering is
+arbitrary here, we simply chose a duplicate and stick to it.
+As illustrated in section~\ref{sec:symbolic}, we first recode the reads
+using a specialized alphabet to simplify the problem.
+
+It will be useful to consider a slightly more general problem where we
+have two sequences of interest labelled $+$ and $-$. One can say that the
+$+$ sequence stands for the target and that the $-$ sequence stands for
+its duplicate. We then define the dual alphabet $\tilde{\mathcal{A}}_0 =
+\{\square, |, \downarrow_{/1}^-, \downarrow_{/2}^-, \ldots,
+\downarrow_{/1}^+, \downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols
+$\downarrow_{/i}^-$ signify that the nucleotide is a mismatch against the
+$-$ sequence only, the symbols $\downarrow_{/i}^+$ signify that it is a
+mismatch against the $+$ only, and the symbol $\Downarrow$ signifies that
+it is a mismatch against both. As before, every other nucleotide is
+replaced by the symbol $\square$, and the terminator $|$ is appended to
+the end of the read. We again define reads as sequences of segments,
+except that now the terminators are the symbols $\downarrow_{/j}^-$,
+$\downarrow_{/j}^+$ and $\Downarrow$. The tail, as usual, is terminated by
+the symbol $|$.
+
+The number $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
+length of the $+$ sequence. Likewise, the number $i$ in the symbol
+$\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
 instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
-mismatch against the off-target sequence, that it is a match for the
-target, and that the six previous nucleotides were also a match for the
-target.
+mismatch against the $-$ sequence, that it is a match for the $+$
+sequence, and that the six previous nucleotides were also a match for the
+$+$ sequence. The terminators thus encode the local state of the read.
+
+Fig.~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
+and $-$ sequences are represented below the read, where the matches are
+represented by open squares and the mismatches by closed squares. It is
+visible from this example that symbols $\downarrow^-_{/i}$ and
+$\downarrow^+_{/i}$ alternate whenever the mismatches are on different
+sequences when going from left to right. The symbol $\Downarrow$ occurs
+only when a nucleotide is a double mismatch.
 
@@
 \includegraphics[scale=0.85]{sketch_dual.pdf}
-\caption{\textbf{The +/- encoding.}
-A example read is represented in the +/- alphabet. The symbols
-$\downarrow_{/j}^-$ represent a mismatch against the off-target sequence,
-the symbols $\downarrow_{/j}^+$ represent a mismatch against the target,
+\caption{\textbf{Example of dual encoding.}
+An example of read is represented in the dual alphabet. The symbols
+$\downarrow_{/i}^-$ represent a mismatch against the $-$ sequence, the
+symbols $\downarrow_{/i}^+$ represent a mismatch against the $+$ sequence,
 and the symbol $\Downarrow$ represent a mismatch against both. The index
-$j$ is the number of matches since the last mismatch for the other
-sequence. Note the presence of the terminator $|$ at the end of the read.
-}
+$i$ is the match length of the sequence that is not mismatched. The
+symbolic $+$ and $-$ sequences are represented below, where an open square
+stands for a match and a closed square stands for a mismatch.}
 \label{fig:dual}
@@
 
-We use the same error model as before: for every nucleotide, the read
-differs from the target with a fixed probability $p$ and the sequences
-differ from each other with a fixed probability $\mu$. With these
-definitions, the probability of occurrence of the symbol $\square$ is $a =
-(1-p)(1-\mu)$. Likewise, the collective probability of occurrence of the
-symbols $\downarrow_{/j}^-$ is $b = (1-p)\mu$, that of the symbols
-$\downarrow_{/j}^+$ is $c = p\mu/3$, and the probability of occurrence of
-the symbol $\Downarrow$ is $d = p(1-\mu/3)$. It is easy to verify that
-$a+b+c+d=1$.
-
+We assume that for each nucleotide, $a$ is the probability that the read
+matches both sequences, $b$ is the probability that it matches only the
+$+$ sequence, $c$ is the probability that it matches only the $-$ sequence
+and $d$ is the probability that it matches none. Obviously $a+b+c+d=1$.
 These definitions allow us to specify the weighted generating function of
-the segments. For instance, the weighted generating function of the
-first segment of Fig.~\ref{fig:dual} is $(az)bz$, and that of the second is
-$(az)^5cz$.
-
-To specify the transfer matrix, we must ensure that we eliminate
-the concatenations of two segments that would create a match of size $\geq
-\gamma$ for any of the two sequences. After defining
+the symbols and of the segments. To specify the transfer matrix, we only
+have to make sure that we eliminate the combinations that create a match
+of size $\gamma$ or more for any of the two sequences. For notational
+convenience, we define
 \begin{equation*}
 \begin{gathered}
-r_i^+(z) = (az)^icz, \\
-r_i^-(z) = (az)^ibz, \\
-R_i(z) = \big( 1 + az + \ldots + (az)^i \big)dz, \\
+r_i^+(z) = cz \cdot (az)^i, \\
+r_i^-(z) = bz \cdot (az)^i, \\
+R_i(z) = dz \cdot \big( 1 + az + \ldots + (az)^i \big), \\
 F_i(z) = 1 + az + \ldots + (az)^i,
@@
 \end{equation*}
-we can verify that the expression of the transfer matrix $L(z)$ is
-
+and we can verify that the expression of the transfer matrix
+$\tilde{M}_0(z)$ is
 \begin{equation*}
@@
     F_{\gamma-3}(z) \\
-\vdots & \vdots & & A(z) & & & B(z) & & \vdots \\
+\vdots & \vdots & & \tilde{A}(z) & & & \tilde{B}_0(z) & & \vdots \\
 \scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
@@
     F_{\gamma-2}(z) \\
-\vdots & \vdots & & C(z) & & & D(z) & & \vdots \\
+\vdots & \vdots & & \tilde{C}_0(z) & & & \tilde{D}(z) & & \vdots \\
 \scriptstyle{\downarrow_{/\gamma-1}^+} & R_0(z) & & & & & & & F_0(z) \\
@@
 \end{equation*}
-where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are square matrices with
-$\gamma-1$ rows/columns, that are defined as
-
+where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are matrices of dimensions
+$(\gamma-1) \times (\gamma-1)$ that are defined as
 \begin{equation*}
-A(z) = 
+\tilde{A}(z) = 
 \begin{blockarray}{cccccc}
@@
 \end{equation*}
-
 \begin{equation*}
-B(z) = 
+\tilde{B}_0(z) = 
 \begin{blockarray}{cccccc}
@@
 \end{equation*}
-
 \begin{equation*}
-C(z) = 
+\tilde{C}_0(z) = 
 \begin{blockarray}{cccccc}
@@
 \end{equation*}
-
 \begin{equation*}
-D(z) = 
+\tilde{D}(z) = 
 \begin{blockarray}{cccccc}
@@
 
-The matrix $L(z)$ is relatively sparse and most non-zero entries are
-monomials. In addition, the number of rows/columns is approximately
-$2\gamma$, so the dimension of the matrix rarely exceeds 50. This means
-that computing the powers of $L(z)$ is not prohibitive and we do not need
-to design a Monte Carlo sampling scheme as we did before.
-
-That being said, we do not need to compute all the powers up to $k$
-because the terms vanish and become negligible after a certain iteration.
-Indeed, the terms of $L(z)^n$ are the weighted generating function of the
-reads without a match for either sequences and that contain exactly $n$
-segments. Of those $n$ segments, $n-1$ are terminated by a mismatch
-against at least one of the sequences. Following the rationale of
-section~\ref{sec:extract}, we define $\tilde{p}$ as the upper bound on the
-probability of a mismatch $\tilde{p} = \max\{b,c,d\}$, and updating
-formula (\ref{eq:bound}) we bound the probability that a read of size $k$
-contains $m = n-1$ or more mismatches
-
+The term of interest is the top right entry of
+$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
+every read can be prepended by segments terminated by $\Downarrow$ and
+only by those (every other terminator would imply that one of the two
+sequences has a nonzero match size at the start of the read). Thus reads
+are precisely the sequences of segments that can follow the symbol
+$\Downarrow$ and that are terminated by a tail, whose weighted generating
+function is the top right entry of the matrix.
+
+It is possible to express $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$ in
+closed form, but it involves polynomials of very large degree. It is
+actually easier to proceed as in section~\ref{sec:example_skip} and to
+compute the powers of $\tilde{M}_0(z)$ (using the arithmetic of truncated
+polynomials) up to a given threshold value. Since each segment contains a
+mismatch against at least one sequence, we define $\tilde{p}$ as the upper
+bound on the probability of a mismatch $\tilde{p} = \max\{b,c,d\}$. The
+updated formula (\ref{eq:bound}) now gives an upper bound of the
+probability that a read of size $k$ contains $s$ or more mismatches as
 \begin{equation*}
-Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-\tilde{p})} -m\log
-\frac{m}{k\tilde{p}} \right).
+Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
+\frac{s}{k\tilde{p}} \right).
 \end{equation*}
 
-The strategy is the same as in section~\ref{sec:extract}: we compute the
-sum of matrices $L(z) + L(z)^2 + \ldots + L(z)^n$ where the weighted
-generating functions have been replaced by truncated polynomials and we
-extract the entry at the first column and the last row. When the bound is
-lower than a set fraction $\varepsilon$ of the coefficient of $z^k$, we
-stop the computations.
-
-\subsection{Corollary: this title sucks}
-
-The results of this section allow us to derive the probabilities of false
-positives when using seeds of fixed size $\gamma$.
-
-The complementary of the event $E_0$ is that the read contains at least
-one match of size $\geq \gamma$ for the target, implying that there is a
-seed for the target\footnote{This is the major difference with MEM seeds,
-where the logical implication does not hold.}. With our assumptions this
-means that the read is mapped to the correct location. The probability of
-outcome $i.$, a true positive, is thus $1-P(E_0)$.
-
-The event $E_0 \cap \ldots \cap E_N$ is that the read contains no match of
-size $\geq \gamma$ for either the targer or any of the off-target
-sequences. In other words, the probability of outome $iii.$, a false
-negative, is $P(E_0 \cap \ldots \cap E_N)$.
-
-The event $E_0 \cap (E_1 \cap \ldots \cap E_N)^c$ is that the read
-contains no match of size $\geq \gamma$ for the target, but contains a
-match of size $\geq \gamma$ for at least one of the off-target sequences.
-In other words, the probability of outcome $ii.$, a false positive, is
-$P(E_0) - P(E_0 \cap \ldots \cap E_N)$.
-
-It is easy to see that the three probabilities sum to 1, as expected.
-Below is a summary of the probabilities of interest.
-
-\begin{eqnarray}
-P(i.) &=& 1-P(E_0) \\
-P(ii.) &=& P(E_0) \cdot \left( 1- \left( \frac{P(E_0
-  \cap E_1)}{P(E_0)} \right)^N\right) \\
-P(iii.) &=& P(E_0) \cdot \left( \frac{P(E_0 \cap E_1)}{P(E_0)} \right)^N
-\end{eqnarray}
-
-In the expressions above, we have used equation (\ref{eq:PE0E1}) to
-highlight that these probabilities only depend on $P(E_0)$ and $P(E_0 \cap
-E_1)$, which can be computed precisely as shown in sections~\ref{sec:cPE0}
-and \ref{sec:cPE0E1}.
-
-Note that the formulas also hold when the target has no duplicate
-sequence. In this case, only outcome $i.$ and $iii.$ are possible, but
-since $N=0$ we see that the second probability is 0, as expected.
-
-\section{Skip seeds}
-
-\begin{algorithm}[H]
-\label{alg:mcmcskip}
-\SetAlgoLined
-\KwResult {Sample a read at random. Return 1 if the read contains a
-good seed, otherwise return 0.}
-  $\lambda \leftarrow 0$ \Comment*[r]{Current read size.}
-  $s_+ \leftarrow 0$ \Comment*[r]{Current + match size.}
-  $s_- \leftarrow 0$ \Comment*[r]{Current - match size.}
-  \While {$\lambda < k$}{
-    $i \leftarrow geom(a)-1$
-                \Comment*[r]{Double-match nucleotides.}
-    \If {$i > k-\lambda$}{
-      $i \leftarrow k-\lambda$ \;
-    }
-    $s_+ \leftarrow s_+ + i$ \;
-    $s_- \leftarrow s_- + i$ \;
-    \If {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
-      return 1 \;
-    }
-    $\lambda \leftarrow \lambda + i+1$ \;
-    $r \leftarrow rand(0,1-a)$
-                \Comment*[r]{Uniform random number.}
-    \uIf {$r < b$}{
-      $s_+ \leftarrow s_++1$ \;
-      $s_- \leftarrow -\lambda \mod n+1$ \;
-    }\uElseIf{$r < b+c$}{
-      $s_+ \leftarrow -\lambda \mod n+1$ \;
-      $s_- \leftarrow s_-+1$ \;
-    }\uElse{
-      $s_+ \leftarrow -\lambda \mod n+1$ \;
-      $s_- \leftarrow -\lambda \mod n+1$ \;
-    }
-}
-\eIf {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
-  return 1 \;
-}{
-  return 0 \;
-}
-\end{algorithm}
-
-\section{MEM seeds}
-
-Before going further, we need to move a practical consideration out of the
-way. It stands to reason that very short matches between the read and the
-genome are uninformative. For instance, $99.7\%$ of all the possible
-12-mers are present in the reference human genome, so an exact match of
-size 12 or lower says nothing about the location of the DNA fragment. In
-what follows, we will assume that subsequences of the read can only match
-the original DNA sequence and/or its duplicates, but not the rest of the
-genome. This will greatly simplify the exposition of the theory. We will
-discuss how to relax this assumption in section~\ref{sec:short_MEMs}.
-
-Observe that the rightmost MEM seed matches two genomic locations. This
-case motivates the following definition, which will play an important role
-for the development of the theory.
-
-\begin{definition}
-A \emph{strict} MEM seed has a single match in the genome.
-A \emph{shared} MEM seed has several matches in the genome.
-\end{definition}
-
-Without surprise, the principle of MEM seeding is to use the genomic
-matches of MEM seeds as candidate locations. Importantly, MEM seeds
-shorter than a certain threshold $\gamma$ are ignored (in line with the
-observation that short matches are uninformative). The parameter $\gamma$
-is the tuning knob on the speed/accuracy trade-off. In
-Fig.~\ref{fig:MEM_example} for instance, if $\gamma = 7$, there are three
-candidate locations (the first, the second and the fourth) and if $\gamma
-= 8$ there are only two (the first and the fourth).
-
-Not every MEM seed makes the target a candidate location, so we introduce
-the following terms to simplify the discussion.
-
-\begin{definition}
-Assume that the thresohld $\gamma$ is fixed. A \emph{good} seed is an
-on-target MEM seed of size $\gamma$ or greater. In other words, the true
-location is in the candidate set if and only if the read contains a
-good seed.
-\end{definition}
-
-Compared to seeds of fixed size, MEM seeds have at least two
-counter-intuitive (and undesirable) properties. The first is that there
-are cases where there cannot be any good seed, even when changing
-$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
-though there is a single sequencing error, the read has no good seed.
-Lowering $\gamma$ does not change this, so there is no way to discover the
-true location (even though it is the best location).
+Now returning to the problem of computing $P(\overline{S}_0 \cap
+\overline{S}_1)$, the $+$ sequence is interpreted as the target and the
+$-$ sequence as the duplicate. Based on the assumptions of the error model
+presented in section~\ref{sec:error}, this implies that $a =
+(1-p)(1-\mu)$, $b = (1-p)\mu$, $c = p\mu/3$, and $d = p(1-\mu/3)$.
+
+With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
+compute its powers until the upper bound above is lower than a set
+fraction $\varepsilon$ of the current value, which gives an accurate
+estimate of $P(\overline{S}_0 \cap \overline{S}_1)$.
+
+\subsection{Corollary: null seeding}
+
+Observe that expression (\ref{eq:Pnull}) is the probability of null
+seeding. The method described above thus allows us to compute this
+probability at no additional cost. This probablity is less relevant than
+the probabilities that the seeding process is on target or off target, but
+at times, it may be useful to know the probability that a read is
+not mappable, especially when reads are relatively short.
+
+\subsection{Illustration}
+\label{sec:illdual}
+
+Using the method above, we can compute the probability that a seeding
+scheme using exact seeds is off target. This depends on the exact
+specifications of the problem, so for illustration purposes, we choose
+reads of size $k=50$ sequenced with an instrument with error rate
+$p=0.01$, and se use exact seeds of size $\gamma=17$. We then vary the
+number of duplicates $N$ and their divergence rate $\mu$, which both
+depend on the particular read being sequenced.
+
+\section{Off-target skip seeds}
+
+To compute the probability that the seeding process is off target when
+using skip seeds, we observe that the logic of section~\ref{sec:offexact}
+can be transposed with very few modifications. In particular, the
+probability can be computed through expression (\ref{eq:Poff}), where
+$S_0$ is the event that the read has a \emph{skip} seed for the target and
+$S_1$ that it has a \emph{skip} seed for the first duplicate.
+
+We have already seen how to compute $P(\overline{S}_0)$ in
+section~\ref{sec:example_skip}, we now need to find a way to compute
+$P(\overline{S}_0 \cap \overline{S}_1)$ when using skip seeds.
+
+\subsection{The skip dual encoding}
+
+Once again, we start by recoding the reads in a specialized alphabet. We
+define the skip-$n$ dual alphabet $\tilde{\mathcal{A}}_n = \{\square, |,
+\Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
+\downarrow^-_{/1}, \downarrow^-_{/2}, \ldots, \downarrow^+_{/1},
+\downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
+$\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ have the same meaning as in
+the dual alphabet of section~\ref{sec:dual}. The symbols $\Downarrow_i$
+indicate that both sequences have match length 0 and that the next
+non-skipped position is $i$ nucleotides downstream.
+
+Fig.~\ref{fig:skip_dual} shows the same read as Fig.~\ref{fig:dual}, but
+here it is represented in the skip-3 dual encoding. It is important to
+note several differences with Fig.~\ref{fig:dual}. The first is that the
+symbols $\Downarrow_i$ are not always associated with double mismatches.
+For instance, the symbol $\Downarrow_0$ on the right side of the read
+corresponds to a mismatch for the $+$ sequence only. This happens whenever
+the $+$ and the $-$ sequences are mismatched in the same interval between
+non-skipped positions (the mismatches do not need to be on the same
+nucleotide).
+
+Another difference is that when the symbols $\downarrow^-_{/i}$ and
+$\downarrow^+_{/i}$ alternate, $i$ is no longer the distance to the
+previous terminator. This is visible for the symbol $\downarrow^+_{/4}$ on
+the left side of the read, where the previous terminator is 6 nucleotides
+upstream. When the symbols alternate, $i$ is the distance to the previous
+non-skipped position.
 
@@
 \centering
-\includegraphics[scale=1]{full_masking_example.pdf}
-\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
-The read, the MEM seeds and the sequences are represented as in
-Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
-locations at the bottom effectively hide the target, so it cannot
-be discovered. This can occur even when the true location is the
-best and when there is a single error on the read.}
-\label{fig:full_masking_example}
+\includegraphics[scale=0.85]{sketch_skip_dual.pdf}
+\caption{\textbf{Example skip dual encoding.}
+The read of Fig.~\ref{fig:dual} is represented in the dual alphabet. The
+vertical bars above the read indicate non-skipped positions. The symbols
+$\downarrow_{/i}^-$ and $\downarrow_{/i}^+$ have the same meaning as in
+the dual alphabet. The symbols $\Downarrow_i$ indicate that both sequences
+have match length 0 and that the next non-skipped position is located $i$
+nucleotides downstream. Other graphical representations are as in
+Fig.~\ref{fig:dual}.}
+\label{fig:skip_dual}
 \end{figure}
 
-The second counter-intuitive property of MEM seeding is that shortening
-a read can sometimes create a good seed. Fig.~\ref{fig:short_vs_long}
-shows an example of this case. Again, there is no good seed, but there
-would be one if the read were two nucleotides shorter on the right side.
-Indeed, in this case there would be a shared MEM seed matching the first
-and the second locations, provided $\gamma \leq 12$.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=1]{short_vs_long_example.pdf}
-\caption{\textbf{Issues with MEM seeds: too long reads.}
-The read, the MEM seeds and the sequences are represented as in
-Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
-read were two nucleotides shorter. The true location is hidden by
-the last two nucleotides.}
-\label{fig:short_vs_long}
-\end{figure}
-
-These examples show that MEM seeds can perform worse than seeds of fixed
-size. MEM seeds yield fewer candidates and therefore speed up the mapping
-process, but the question is at what cost for accuracy? The theory
-developed below will allow us to compute the probability that a read has
-no good seed and thus that the true location is missed. This probability
-is the key to evaluating and calibrating the MEM seeding heuristic.
-
-We first expose a strategy to compute the probablity of true positives
-(outcome $i.$), then show how to compute the probability of false
-negatives (outcome $iii.$), and combine these two quantities to compute
-the probability of false positives (outcome $ii.$). Finally, we provide a
-computationally efficient method for estimating the parameters that are
-necessary to estimate the probabilities of interest.
-
-\section{Combinatorial construction}
-
-In this section we represent reads as combinatorial ojects. The purpose of
-this construction is to give a generative model of reads without good
-seeds so that we can later compute their probability of occurrence.
+As in the case of the skip encoding, a fair amount of bookkeeping is
+required to specify the weighted generating functions of interest.
+We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
+section~\ref{sec:dual}. For notational convenience, we define
+\begin{eqnarray*}
+N_i(z) &=& 1+z+\ldots+z^i, \\
+W_j(z) &=& dz \cdot (az)^x \cdot \big( 1 + (az)^{n+1} + \ldots +
+  (az)^{(n+1)m} \big), \\
+  &\quad& \text{where } x = n-j \modulo{n+1},
+  \text{ and } m = \left\lfloor
+  \frac{\gamma-1-x}{n+1} \right\rfloor. \\
+U_{i,j}(z) &=& dz \cdot (az)^x \cdot
+  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
+\begin{cases}
+  bz\cdot(az)^x \text{ if } j < y, \\
+  0 \text{ otherwise}
+\end{cases} \\
+  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
+  y = n-i+1 \modulo{n+1}, \\
+  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
+    \right\rfloor, \\
+V_{i,j}(z) &=& dz \cdot (az)^x \cdot
+  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
+\begin{cases}
+  cz\cdot(az)^x \text{ if } j \leq x, \\
+  0 \text{ otherwise}
+\end{cases} \\
+  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
+  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
+\end{eqnarray*}
+
+With these definitions, one can check that the expression of the transfer
+matrix $\tilde{M}_n(z)$ is
+
+\begin{equation*}
+\rotatebox{90}{$
+\begin{blockarray}{cccccccccccc}
+   & \scriptstyle{\Downarrow_0} & \scriptstyle{\Downarrow_1} &
+   \ldots & \scriptstyle{\Downarrow_n} &
+    \scriptstyle{\downarrow_{/1}^+} & 
+    \ldots & \scriptstyle{\downarrow_{\gamma-1}^+} &
+    \scriptstyle{\downarrow_{/1}^-} & \ldots &
+    \scriptstyle{\downarrow_{/\gamma-1}^-} & \scriptstyle{|} \\
+\begin{block}{c[ccccccccccc]}
+\scriptstyle{\Downarrow_0} & W_0(z) & W_1(z) & \ldots & W_n(z) &
+    r_0^+(z) & \ldots & r_{\gamma-2}^+(z) & r_0^-(z) & \ldots &
+    r_{\gamma-2}^-(z) & F_{\gamma-1}(z) \\
+\scriptstyle{\Downarrow_1} & z  & 0 & \ldots &
+  0 & 0 & \ldots & 0 & 0 & \ldots & 0 & N_0(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
+  \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\Downarrow_n} & z^n & 0 & \ldots &
+  0 & 0 & \ldots & 0 & 0 & \ldots & 0 & N_{n-1}(z) \\
+\scriptstyle{\downarrow_{/1}^+} & U_{1,0}(z) & U_{1,1}(z) & \ldots &
+  U_{1,n}(z) & & & & & & & F_{\gamma-2}(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & & \tilde{A}(z) & & &
+  \tilde{B}_n(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & U_{\gamma-1,0}(z) &
+  U_{\gamma-1,1}(z) & \ldots & U_{\gamma-1,n}(z) & & & & & & & F_0(z) \\
+\scriptstyle{\downarrow_{/1}^-} & V_{1,0}(z) & V_{1,1}(z) & \ldots &
+    V_{1,n}(z) & & & & & & & F_{\gamma-2}(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & & \tilde{C}_n(z) &
+  & & \tilde{D}(z) & & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & V_{\gamma-1,0}(z) &
+    V_{\gamma-1,1}(z) & \ldots & V_{\gamma-1,n}(z) & & &
+    & & & & F_0(z) \\
+\scriptstyle{|} & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 &
+  \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}.
+$}
+\end{equation*}
+
+The matrices $\tilde{A}(z)$ and $\tilde{C}(z)$ in the expression of
+$\tilde{M}_n(z)$ are the same as in section~\ref{sec:dual}. The matrices
+$\tilde{B}_n(z)$ and $\tilde{C}_n(z)$ are defined as
+\begin{equation*}
+\tilde{B}_n(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
+   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
+   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & s_{1,1}(z) & s_{1,2}(z) &
+    \ldots & s_{1,\gamma-2}(z) & s_{1,\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/2}^+} & s_{2,1}(z) & s_{2,2}(z) &
+    \ldots & s_{2,\gamma-2}(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & s_{\gamma-2,1}(z) &
+    s_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & s_{\gamma-1,1}(z) & 0 &
+    \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+\begin{equation*}
+\tilde{C}_n(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & t_{1,1}(z) & t_{1,2}(z) &
+    \ldots & t_{1,\gamma-2}(z) & t_{1,\gamma-1}(z) \\
+\scriptstyle{\downarrow_{/2}^-} & t_{2,1}(z) & t_{2,2}(z) &
+    \ldots & t_{2,\gamma-2}(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & t_{\gamma-2,1}(z) &
+    t_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & t_{\gamma-1,1}(z) &
+    0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+with
+\begin{eqnarray*}
+s_{i,j} &=&
+  \begin{cases}
+  cz \cdot (az)^{x+j-1} &\text{ if } i+j+x \leq \gamma \\
+  0                     &\text{ otherwise,}
+  \end{cases} \\
+t_{i,j} &=&
+  \begin{cases}
+  bz \cdot (az)^{x+j-1} &\text{ if } i+j+x \leq \gamma \\
+  0                     &\text{ otherwise,}
+  \end{cases} \\
+  \text{where } x &=& -i \modulo{n+1}, \text{ in both cases.} \\
+\end{eqnarray*}
+
+The computation is performed exactly as described in
+section~\ref{sec:dual}. We compute the successive powers of
+$\tilde{M}_n(z)$ in the arithmetic of truncated polynomials and stop the
+iterations using the same criterion. So there is no change other than the
+definition of the transfer matrix.
+
+\begin{remark}
+Observe that when $n=0$ the matrix $\tilde{M}_n(z)$, is identical to the
+matrix $\tilde{M}_0(z)$ of section~\ref{sec:dual}, again consistent with
+the fact that exact seeds are skip-$n$ seeds for which $n=0$. The same
+applies to $\tilde{B}_n(z)$ and $\tilde{C}_n(z)$.
+\end{remark}
+
+% XXX The following algorithm works but it is not very useful    XXX
+% XXX because it is substantially slower than the WGF method.    XXX
+% XXX It has been used to check the accuracy of the calculations XXX
+% XXX but there is no reason to put it in the manuscript.        XXX
+
+%\begin{algorithm}[H]
+%\label{alg:mcmcskip}
+%\SetAlgoLined
+%\KwResult {Sample a read at random. Return 1 if the read contains a
+%good seed, otherwise return 0.}
+%  $\lambda \leftarrow 0$ \Comment*[r]{Current read size.}
+%  $s_+ \leftarrow 0$ \Comment*[r]{Current + match size.}
+%  $s_- \leftarrow 0$ \Comment*[r]{Current - match size.}
+%  \While {$\lambda < k$}{
+%    $i \leftarrow geom(1-a)-1$
+%                \Comment*[r]{Double-match nucleotides.}
+%    \If {$i \geq k-\lambda$}{
+%      $s_+ \leftarrow s_+ + k-\lambda$ \;
+%      $s_- \leftarrow s_- + k-\lambda$ \;
+%      break \;
+%    }
+%    $s_+ \leftarrow s_+ + i$ \;
+%    $s_- \leftarrow s_- + i$ \;
+%    \If {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
+%      return 1 \;
+%    }
+%    $\lambda \leftarrow \lambda + i+1$ \;
+%    $r \leftarrow rand(0,1-a)$
+%                \Comment*[r]{Uniform random number.}
+%    \uIf {$r < b$}{
+%      $s_+ \leftarrow s_++1$ \;
+%      $s_- \leftarrow -(-\lambda \mod n+1)$ \;
+%    }\uElseIf{$r < b+c$}{
+%      $s_+ \leftarrow -(-\lambda \mod n+1)$ \;
+%      $s_- \leftarrow s_-+1$ \;
+%    }\Else{
+%      $s_+ \leftarrow -(-\lambda \mod n+1)$ \;
+%      $s_- \leftarrow -(-\lambda \mod n+1)$ \;
+%    }
+%}
+%\eIf {$s_+ \geq \gamma$ or $s_- \geq \gamma$}{
+%  return 1 \;
+%}{
+%  return 0 \;
+%}
+%\end{algorithm}
+
+\subsection{Illustration}
+\label{sec:illskipdual}
+
+Using the method above, we can compute the probability that a seeding
+scheme using skip seeds is off target. Using the same settings as in
+section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), we set
+decide to use skip-9 seeds. We then vary the number of duplicates $N$ and
+their divergence rate $\mu$.
+
+\section{Off-target MEM seeds}
+
+MEM seeds are substantially more complex than exact seeds and skip seeds.
+As we will see below, the duplicates can prevent the existence of an
+on-target MEM seed, so we need to take them into account in the
+combinatorial construction.
 
@@
 
-For each potential location including the target, we define the
-\emph{match streak} at a given nucleotide as the number of nucleotides
-since the last mismatch on the left, or since the beginning of the read.
+We first introduce two important notions that will be the key of
+understanding the behavior of MEM seeds.
 
 \begin{definition}
-At a given position of the read, an off-target sequence is a \emph{hard
-mask} if its success streak is strictly longer than the match streak of
-the target. An off-target sequence is a \emph{soft mask} if its has the
-same match streak as the target.
+At a given position of the read, a duplicate is a \emph{hard mask} if its
+match length is strictly longer than the match length of the target. A
+duplicate is a \emph{soft mask} if its has the same match length as the
+target.
 \end{definition}
 
-Fig.~\ref{fig:hard_vs_soft_masks} illustrates hard and soft masks with an
-example. It is important to bear in mind that hard and soft masks depend
-on the position of interest: a sequence can be a mask at the left end of
-the read but not at the right end, or \textit{vice versa}. In the example
-of Fig.~\ref{fig:full_masking_example}, every nucleotide of the read has a
-hard mask, which is why the target cannot be discovered. In the example of
-Fig.~\ref{fig:short_vs_long}, the middle sequence is a soft mask on the
-right half of the read, but it turns into a hard mask two nucleotides
-before the right end, which is why the target could be discovered if the
-read were shorter.
+Fig.~\ref{fig:hard_vs_soft_masks} gives a graphical intuition of hard and
+soft masks. It is important to bear in mind that hard and soft masks
+depend on the position of interest: a sequence can be a mask at the left
+end of the read but not at the right end, or the opposite.
 
@@
 \caption{\textbf{Example of hard and soft masks.}
-The nucleotides that do not match the read are reprsented as black boxes.
-The nucleotides that contribute to the match streak are represented as
-grey boxes. The remaining nucleotides are represented as open boxes. At
-the given position, the match streak of the target is 7. The first
-off-target is a hard mask because its match streak is $13>7$. The second
-off-target sequence is a soft mask because its match streak is 7 like the
-target. The third off-target sequence is not a mask because its match
-streak is $2<7$.}
+Genomic sequencess are shown below a read where open squres represent
+nucleotides. In the sequences, the open squares represent matches and the
+closed squares represent mismatches. The nucleotides contributing to the
+match length are represented as grey boxes. At the focus position, the
+match length of the target is 7. The first duplicate is a hard mask
+because its match length is $13>7$. The second duplicate is a soft mask
+because its match length is 7, as the target. The third duplicate is not
+a mask because its match length is $2<7$.}
 \label{fig:hard_vs_soft_masks}
 \end{figure}
+
+Hard and soft masks explain the counter-intuitive of MEM seeds. For
+instance, in Fig.~\ref{fig:full_masking_example} the target cannot be
+discovered because every nucleotide of the read has a hard mask. In
+Fig.~\ref{fig:short_vs_long}, the target could be discovered if the read
+were shorter because a hard mask would turn into a soft one.
 
@@
 
-We recode the reads as sequences of letters from the so-called MEM
-alphabet $\mathcal{A} = \{\square, |, \uparrow_{/1}, \uparrow_{/2},
-\uparrow_{/3} \ldots, \downarrow_{/0}, \downarrow_{/1}, \downarrow_{/2},
-\ldots\}$.
+As before, we recode the reads as sequences of letters from a specialized
+alphabet called the MEM alphabet $\mathring{\mathcal{A}} = \{\square, |,
+\uparrow_{/1}, \uparrow_{/2}, \uparrow_{/3} \ldots, \downarrow_{/0},
+\downarrow_{/1}, \downarrow_{/2}, \ldots\}$.
 
 The symbols $\downarrow_{/m}$ indicate that the nucleotide is a sequencing
-error and $m \geq 0$ is the number of off-target locations that
-\emph{match} the nucleotide. Since a sequencing error is always a mismatch
-against the target, the symbol $\downarrow_{/0}$ indicates that the
-ncleotide is a mismatch against \emph{every} candidate.
-
-The symbols $\uparrow_{/j}$ indicate a change in masking status: the
-nucleotide is not masked but its left neighbor is. The index $j \geq 1$ is
-the number of nucleotides since the last mismatch or since the beginning
-of the read.
-
-All the other nucleotides are represented by the $\square$ symbol.
-Finally, the $|$ symbol is appended after the last nucleotide of the read.
-It is an obligatory terminator that does not correspond to an actual
-nucleotide.
-
-Note that the numbers decorating $\uparrow$ and $\downarrow$ are different
-in nature. In the case of the symbol $\downarrow$ it is an information
-about the number of sequences matching the read; in the case of the symbol
-$\uparrow$ it is the position of the previous sequencing error.
-Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
-Fig.~\ref{fig:MEM_example} in the MEM alphabet.
+error and $m \geq 0$ is the number of duplicates that \emph{match} the
+nucleotide. Since a sequencing error is always a mismatch against the
+target, the symbol $\downarrow_{/0}$ indicates that the ncleotide is a
+mismatch against \emph{every} sequence. The symbols $\uparrow_{/i}$
+indicate a change in masking status: the nucleotide is not masked but its
+left neighbor is. The index $i \geq 1$ is the number of nucleotides since
+the last mismatch or since the beginning of the read. All the other
+nucleotides are represented by the $\square$ symbol and the $|$ symbol is
+appended to the end of the read as usual.
+
+Note that in the symbols $\downarrow_{/m}$ and $\downarrow_{/i}$, the
+numbers $m$ and $i$ have different meanings. In the symbol
+$\downarrow_{/m}$, the index $m$ is a number of sequences ($0 \leq m \leq
+N$ where $N$ is the number of duplicates); in the symbol $\uparrow_{/i}$,
+the index $i$ is a number of nucleotides. Fig.~\ref{fig:sketch_extended}
+shows the encoding of the read from Fig.~\ref{fig:exact_seed_example} in
+the MEM alphabet.
 
@@
 \centering
-\includegraphics[scale=.85]{sketch_extended.pdf}
+\includegraphics[scale=.84]{sketch_extended.pdf}
 \caption{\textbf{The MEM encoding}.
-The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
-The arrows departing from those numbers help understand their meaning. The
-$\downarrow$ symbol is decorated by the number of locations that match the
-nucleotide. The $\uparrow$ symbol is decorated by the number of
-nucleotides from the last error or from the beginning of the read. Also
-note the presence of the terminator $|$ at the end of the read.}
+The read of Fig.~\ref{fig:exact_seed_example} is represented in the MEM
+alphabet. The arrows departing from the numbers help understand their
+meaning. The symbol $\downarrow_{/m}$ is indexed by the number $m$ of
+locations that match the nucleotide. The symbol $\uparrow_{/i}$ is
+indexed by the number $i$ of nucleotides from the last error or from the
+beginning of the read.}
 \label{fig:sketch_extended}
@@
 
-The MEM alphabet captures the masking status of the nucleotide as follows:
-If we denote the number of off-targets as $N$, the symbol
+The MEM alphabet captures the masking status of the nucleotide: the symbol
 $\downarrow_{/m}$ indicates that the nucleotide has $m$ hard masks and
-$N-m$ soft masks. In addition, the symbols $\uparrow_{/j}$ indicate that
-the nucleotide is unmasked.
+$N-m$ soft masks. The symbols $\uparrow_{/i}$ indicate that the nucleotide
+is unmasked and that the previous $i-1$ nucleotides were masked.
 
 In the MEM alphabet, strict on-target MEM seeds are the longest stretches
-of symbols containing $\uparrow$ and not containing $\downarrow$. Indeed,
-such a stretch matches the target, it matches the target only because it
-contains at least one unmasked nucleotide (marked by $\uparrow$), and it
-cannot be extended because it is flanked by sequencing errors or by the
-ends of the reads. Note that there is exactly one symbol $\uparrow$ per
-strict on-target MEM seed, and therefore that two symbols $\uparrow$ must
-be separated by at least one symbol $\downarrow$.
-
-Shared on-target MEM seeds are the longest stretches of $\square$ symbols
+of symbols containing some symbol $\uparrow_{/i}$ and not containing any
+symbol $\downarrow_{/m}$. Indeed, such a stretch is a match for the target
+because it does not contain any symbol $\downarrow_{/m}$, it is match for
+the target only because it contains at least one unmasked nucleotide
+(marked by $\uparrow_{/i}$), and it cannot be extended because it is
+flanked by sequencing errors or by the ends of the reads. Note that there
+is exactly one symbol $\uparrow_{/i}$ per strict on-target MEM seed, and
+therefore two symbols $\uparrow_{/i}$ must be separated by at least one
+symbol $\downarrow_{/m}$.
+
+Shared on-target MEM seeds are the longest stretches of symbols $\square$
 flanked by $\downarrow_{/0}$, or by the ends of the read. Indeed, such a
@@
 extended ($\downarrow_{/0}$ is a mismatch against every sequence). Also,
-it cannot be a strict on-target MEM seed because it does not contain the
-$\uparrow$ symbol, so it must be shared.
-
-\subsection{Reads as extended MEM segments}
-
-Reads can be viewed as sequences of symbols, or as sequences of
-\emph{segments}. For the purpose of constructing reads without good seeds,
-the second point of view turns out to be essential.
-
-\begin{definition}
-\label{def:segment}
-A segment is a maximal sequence of $\square$ symbols flanked on the right
-side by one of the symbols $\uparrow_{/j}$, $\downarrow_{/m}$, or $|$. The
-one and only segment flanked by $|$ is called the tail. It is the only
-segment that can have size 0.
-\end{definition}
-
-In line with definition~\ref{def:segment}, the symbols $\uparrow_{/j}$,
-$\downarrow_{/m}$ and $|$ are referrred to as \emph{terminators}.
-Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
-from Fig.~\ref{fig:sketch_extended}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.85]{sketch_segments.pdf}
-\caption{\textbf{The extended MEM segment encoding}.
-The read from Fig.~\ref{fig:sketch_extended} is represented as a sequence
-of segments. Each segment is terminated by either $\uparrow_{/j}$, or
-$\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
-below indicate the size of each segment in number of nucleotides.}
-\label{fig:sketch_segment}
-\end{figure}
-
-We now have a representation of reads as sequences of segments. Only some
-of these sequences are possible. For instance, a segment terminated by
-$\uparrow$ cannot follow another. The task, however, is not to give a
-construction of reads, but a construction of reads without good seeds. For
-this, we need to introduce a new set of tools.
-
-\section{The transfer matrix}
-\label{sec:transfer_mat}
-
-\subsection{Segments following $\uparrow_{/j}$}
-
-A segment terminated by $\uparrow_{/j}$ is the beginning of an on-target
-MEM seed of size at least $j$. The seed reaches the next sequencing
+it cannot be a strict on-target MEM seed because it does not contain any
+$\uparrow_{/i}$ symbol, so it must be a shared on-target MEM seed.
+
+As before, the read is converted from a sequence of symbols to a sequence
+of segments that consist of 0 or more symbols $\square$ followed by a
+terminator. We then specify the weighted generating functions of those
+segments and fill the tansfer matrix $\mathring{M}_N(z)$ of the reads that
+do not contain an on-target MEM seed. We introduce the terms of the matrix
+by increasing oder of complexity.
+
+\subsection{Segments following $\uparrow_{/i}$}
+
+A segment terminated by $\uparrow_{/i}$ is the beginning of an on-target
+MEM seed of size at least $i$. The seed reaches the next sequencing
 error or the end of the read, so the number of $\square$ symbols in the
-next segment must be at most $\gamma-j-1$ and it must be terminated by a
+next segment must be at most $\gamma-i-1$ and it must be terminated by a
 $\downarrow$ symbol or by the tail terminator $|$.
 
-On this segment, the matches between the read and the off-target sequences
-are irrelevant. The weighted genearting function of each $\square$ symbol
-is simply $qz$ ($q$ is the probability that the nucleotide is not an error
-and $z$ marks objects of size $1$). The following definition will simplify
-the notations.
+The following definition will simplify the notations.
 
@@
 
-The weighted generating function of the $\downarrow_{/m}$ terminator is
-thus $\omega_m pz$ (the probability of the symbol $\downarrow_{/m}$ is
-$\omega_m p$ and $z$ marks objects of size $1$).
-
-The entries of the transfer matrix that correspond to transitions from
-$\uparrow_{/j}$ to other segments are now easy to find. The weighted
-generating function of the segments terminated by $\downarrow_{/m}$
-following a segment terminated by $\uparrow_{/j}$ is
+Indeed, if the nucleotide is an error, the symbol is $\downarrow_{/m}$ for
+some $m$ between 0 and $N$. Each duplicate is a match with probability
+$\mu/3$, so $m$ has a Binomial distribution with parameters $(N, \mu/3)$.
+
+On this segment, the matches between the read and the duplicates are
+irrelevant, so the weighted generating function of a symbol $\square$ is
+simply $qz$ (recall that $p = 1-q$ is the probability of a sequencing
+error). The weighted generating function of the $\downarrow_{/m}$
+terminator is $\omega_m pz$, so the weighted generating function of the
+segments terminated by $\downarrow_{/m}$ following a segment terminated by
+$\uparrow_{/i}$ is
 \begin{equation}
 \label{eq:D}
-D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i.
+D_{i,m}(z) = \omega_m pz \sum_{j=0}^{\gamma-i-1} (qz)^j.
 \end{equation}
@@
 And the weighted generating function of the tail segments following
-segments terminated by $\uparrow_{/j}$ is
+segments terminated by $\uparrow_{/i}$ is
 \begin{equation}
 \label{eq:E}
-E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i.
+E_i(z) = \sum_{j=0}^{\gamma-i-1} (qz)^j.
 \end{equation}
@@
 masks and $N-m$ soft masks. If all the masks disappear before the first
-read error, the next terminator will be a $\uparrow$ symbol, otherwise it
-will be a $\downarrow$ symbol. We separate the cases based on the
-terminator of the segment.
-
-\subsubsection*{Case 1: the terminator is $\uparrow_{/i}$}
+read error, the next terminator will be one of the $\uparrow_{/j}$
+symbols, otherwise it will be one of the $\downarrow_{/m}$ symbols. We
+separate the cases based on the terminator of the segment.
+
+\subsubsection*{Case 1: the terminator is $\uparrow_{/j}$}
 
 \begin{definition}
-The probability that a given off-target sequence contains a mismatch in a
-sequence of $i$ error-free nucleotides is
+The probability that a given duplicate sequence contains a mismatch in a
+sequence of $j$ error-free nucleotides is
 \begin{equation}
 \label{eq:xi}
-\xi_i = 1-(1-\mu)^i.
+\xi_j = 1-(1-\mu)^j.
 \end{equation}
+This is the probability that a hard or soft mask vanishes within $j$
+correct nucleotides.
 \end{definition}
 
+Indeed, every nucleotide of the duplicate differs from the target
+with probability $\mu$. In the absence of sequencing error, this is also
+the probability that a nucleotide of the duplicate differs from the read.
+The probability that $j$ nucleotides in a row are identical to the read is
+thus $(1-\mu)^j$ and the probability that at least one of them is
+different is the complement $1-(1-\mu)^j$.
+
 With this notation, the probability that at least one mask survives a
-sequence of $i$ error-free nucleotides is thus $1-\xi_i^N$, and the
-probability that there remains a mask at the $i-1$-th but not at the 
-$i$-th error-free nucleotide is $\xi_i^N - \xi_{i-1}^N$. From this we
+sequence of $j$ error-free nucleotides is thus $1-\xi_j^N$, and the
+probability that there remains a mask at the $j-1$-th but not at the 
+$j$-th error-free nucleotide is $\xi_j^N - \xi_{j-1}^N$. From this we
 conclude that the weighted generating function of the segments terminated
-by $\uparrow_{/i}$ following a segment terminated by $\downarrow_{/m}$ is
+by $\uparrow_{/j}$ following a segment terminated by $\downarrow_{/m}$ is
 \begin{equation}
 \label{eq:B}
-B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i.
+B_j(z) = \Big( \xi_j^N-\xi_{j-1}^N \Big) (qz)^j.
 \end{equation}
 
-In the equation above, $q^i$ is the probability that there is no
-sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
-The fact that the reads have no good seeds imposes $i < \gamma$. Also note
-that this expression is the same for all $\downarrow$ symbols; it does not
-depend on $m$.
+The fact that the reads have no on-target seeds imposes $j < \gamma$. Also
+note that this expression is the same for all $\downarrow_{/m}$ symbols
+(it does not depend on $m$).
 
@@
 
-In this case there can be no good seed because the read terminates too
-early. We must just enforce the condition that at least one of the $N$
+In this case there can be no on-target seed because the read finishes
+too early. We must just enforce the condition that at least one of the $N$
 masks survives until the end of the segment in order to exclude the
-$\uparrow$ symbol. The weighted generating function is
+$\uparrow_{/j}$ symbols. The weighted generating function is
 \begin{equation*}
@@
 In this case, the soft masks do not hide the target: even if one of them
-survives until the end of the read, there will be a good seed (shared in
-this case). To make sure that the read has no good seed, we must enforce
-the condition that at least one hard mask survives until the end of the
+survives until the end of the read, there will be an on-target seed
+(shared in this case). To avoid on-target seeds, we must enforce the
+condition that at least one hard mask survives until the end of the
 segment (which is impossible if $m = 0$). The weighted generating function
@@
 
-Finally, the weighted generating function of the tail segments following
-segments terminated by $\downarrow_{/m}$ is
+Summing the expressions from cases 2a and 2b, we find that the
+weighted generating function of the tail following segments terminated by
+$\downarrow_{/m}$ is
 \begin{equation}
@@
 
-In this case, there can be no good seed and we must only exclude the
-$\uparrow$ terminator. The weighted generating function is
+In this case, there can be no on-target seed and we must only exclude the
+$\uparrow_{/j}$ terminators. As we have seen above, this implies that at
+least one of the $N$ masks survives until the terminator. For a read of
+size $j+1$, this occurs with probability $1-\xi_j^N$. Including the
+terminator and summing over $j+1 \leq \gamma$, we see that the weighted
+generating function is thus
 \begin{equation}
-\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
+\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - \xi_j^N \Big) (qz)^j.
 \end{equation}
@@
 This case is by far the most convoluted. Since the segment contains at
-least $\gamma$ error-free nucleotides, we must ensure that it does not
-contain a good seed. This will be the case if any of the two following
-conditions is validated: $i)$ at least one hard mask covers all the
-error-free nucleotides, or $ii)$ all the hard masks vanish but at least
-one soft mask covers the whole segment (including the terminator).
+least $\gamma$ error-free nucleotides, we must enforce the condition that
+it does not contain an on-target seed. This will be the case if any of the
+two following conditions is validated: $i)$ at least one hard mask covers
+all the error-free nucleotides, or $ii)$ all the hard masks vanish but at
+least one soft mask covers the whole segment (including the terminator).
 
@@
 to case $i)$ and the panel to case $ii)$. The top row represents the
-target, and the bottom rows represent off-target sequences (using the same
-symbols as in Fig.~\ref{fig:MEM_example}).
+target, and the bottom rows represent duplicates (using the same
+symbols as in Fig.~\ref{fig:sketch_extended}).
 \begin{inset}
@@
 
-Whenever a hard mask (here the first off-target sequence) covers the
-nucleotides as shown in the left panel, there can be no good seed,
-irrespective of the value of $\gamma$. The positions marked with a
-question mark are irrelevant, they cannot change the fact that there is no
-on-target MEM seed. If the hard masks vanish, as in the right panel, then
-it all depends on the soft masks. If a soft mask covers the whole segment,
-then there can be no good seed, irrespective of the value of $\gamma$.
-
-Condition $i)$ has probability $\big(1 - \xi_i^m \big)$ and the
-weighted generating function is thus
+Whenever a hard mask (here the first duplicate) covers the nucleotides as
+shown in the left panel, there can be no on-target seed. The positions
+marked with a question mark are irrelevant, they cannot change the fact
+that there is no on-target MEM seed. If the hard masks vanish, as in the
+right panel, then we need to look at the soft masks. If a soft mask covers
+the whole segment, then there can be no on-target seed. In all other cases
+there is an on-target MEM seed.
+
+For a segment of size $j+1$, condition $i)$ has probability $\big(1 -
+\xi_j^m \big)$. Summing over $j+1 > \gamma$ and including the terminator,
+we see that the associated weighted generating function is thus
 \begin{equation*}
-\omega_n pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
+\omega_n pz \sum_{j=\gamma}^\infty \Big(1 - \xi_j^m \Big) (qz)^j.
 \end{equation*}
@@
 \begin{definition}
-The probability that a given off-target sequence contains a mismatch in a
-sequence of $i$ error-free nucleotides followed by an error is
+The probability that a given duplicate sequence contains a mismatch in a
+sequence of $j$ error-free nucleotides followed by an error is
 \begin{equation}
 \label{eq:eta}
-\eta_i = 1-(1-\mu)^i\mu/3.
+\eta_j = 1-(1-\mu)^j\mu/3.
 \end{equation}
+This is the probability that a hard or soft mask vanishes within $j$
+correct nucleotides followed by a sequencing error.
 \end{definition}
 
-The probability of condition $ii)$ is
+Indeed, the probability that there are $j$ matches between the duplicate
+and the target is $(1-\mu)^j$. If there are no sequencing errors, this is
+also the probability that there are $j$ matches between the duplicate and
+the read. The probability that the duplicate matches the subsequent
+sequencing error is $\mu/3$, so the probability of that there are $j+1$
+matches including the sequencing error is $(1-\mu)^j\mu/3$. Finally, the
+probability that there is a mismatch is the complement $1-(1-\mu)^j\mu/3$.
+
+Let us for now consider a segment of fixed size $j+1$. From expressions
+(\ref{eq:xi}) and (\ref{eq:eta}), the probability of condition $ii)$ is
 \begin{equation*}
-\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
+\xi_j^m \Big(1 - \eta_j^{N-m} \Big),
 \end{equation*}
-but we need to break up this term among all the terminators
-$\downarrow_{/n}$, ($0 \leq n \leq N$). For this, we decompose the sum on
-the number of soft masks that run to the end of the segment, including the
-terminator. The probability that there are $r \geq 1$ such soft masks is
+but we need to break up this term among all the possible terminators
+$\downarrow_{/n}$ ($0 \leq n \leq N$) in order to fill the different
+entries of the transfer matrix. For this, we split this term on the number
+of soft masks that run until and including the terminator. From expression
+(\ref{eq:eta}), the probability that there are $r \geq 1$ such soft masks
+is
+\begin{equation}
+\label{eq:softmasks_r}
+{N-m \choose r} (1 - \eta_j)^r \eta_j^{N-m-r}.
+\end{equation}
+
+For now we consider $r$ fixed; we will compute the marginal probability at
+the final stage. By construction, each of those $r$ soft masks matches the
+terminator, so the total number of matches is $r$ plus the number of
+sequences that also match the last nucleotide, among the remaining $N-m-r$
+soft masks and the $m$ hard masks.
+
+Let us start with the $m$ hard masks. The probability that each of them
+matches the terminator is simply $\mu/3$, and the probably that they do
+not is $1-\mu/3$.
+
+The case of the $N-m-r$ soft masks is more complicated because they can
+vanish precisely on the terminator --- recall that in case $ii$) all the
+hard masks are assumed to vanish before. If the soft mask failed within
+the first $j$ nucleotides, then the $j+1$-th nucleotide can be anything
+and it will match the terminator with probability $\mu/3$. But if the soft
+mask survived the first $j$ nucleotides, then it \emph{must} fail on the
+$j+1$-th and it cannot match the terminator. From expressions
+(\ref{eq:xi}) and (\ref{eq:eta}), the probability that a given soft mask
+fails withtin the first $j$ nucleotides is $\xi_j/\eta_j$ --- this is the
+conditional probability that it fails within the first $j$ nucleotides
+given that it fails within the segment. Finally the probability that such
+a soft mask matches the terminator is $\mu/3 \cdot \xi_j / \eta_j$.
+
+Summing the contributions of the hard masks ($m$ in total, each matching
+the terminator with probability $\mu/3$) and of the soft masks ($N-m-r$ in
+total, each matching the terminator with probability
+$\mu/3\cdot\xi_j/\eta_j$), the probability that the total number of
+matches is $n-r$ appears as the convolution product
 \begin{equation*}
-{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
+\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_j^{N-m-r}}
+\psi_{j,m,n,r}\text{, where \;}
+\psi_{j,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
+\xi_j^{n-r-q}.
 \end{equation*}
 
-Each of them matches the nucleotide at the end of the segment, so the
-total number of matches is $r$ plus the number of sequences that also
-match the last nucleotide, among the remaining $N-m-r$ soft masks and $m$
-hard masks.
-
-Let us start with the $N-m-r$ that vanished somewhere in the segment. The
-conditional probability that they match the last nucleotide given that
-they vanish somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
-conditional probability that they do not match it is $(1-\mu/3) / \eta_i$.
-For the $m$ hard masks, the probability that they match the last
-nucleotide given that they vanish somewhere in the segment is simply
-$\mu/3$, and the probability that they do not match it is $1-\mu/3$.
-
-The probability that the last nucleotide of the segment matches $n-r$ of
-the $N-r$ masks that vanished before the terminator is
-\begin{equation*}
-\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_i^{N-m-r}}
-\psi_{i,m,n,r}\text{, where \;}
-\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
-\xi_i^{n-r-q}.
-\end{equation*}
-
-Finally, the probability that a total of $n$ off-target sequences match
-the terminator and thus that the terminator is the symbol
-$\downarrow_{/n}$ is
+Finally, we need to compute the marginal probability over the number $r$
+of soft masks that survive until the end of the read. Multiplying by the
+probability of $r$ from expression (\ref{eq:softmasks_r}) and summing over
+$r \geq 1$, the probability that $n$ duplicate sequences match the
+terminator appears as
 \begin{eqnarray*}
 &\;& \sum_{r\geq1} {N-m \choose r}
-(1 - \eta_i)^r (\mu/3)^{n-r} (1-\mu/3)^{N-n} \psi_{i,m,n,r} \\
+(1 - \eta_j)^r (\mu/3)^{n-r} (1-\mu/3)^{N-n} \psi_{j,m,n,r} \\
 &=& (\mu/3)^n(1-\mu/3)^{N-n} \sum_{r\geq1} {N-m \choose r}
-  (1 - \mu)^{ri} \psi_{i,m,n,r} \\
-&=& \omega_n \cdot \zeta_{i,m,n},
+  (1 - \mu)^{rj} \psi_{j,m,n,r} \\
+&=& \omega_n \cdot \zeta_{j,m,n},
 \end{eqnarray*}
@@
 \label{eq:zeta}
-\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
-(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n}.
+\zeta_{j,m,n} = \sum_{r\geq1} {N-m \choose r}
+(1-\mu)^{rj} \psi_{j,m,n,r} \bigg/ {N \choose n}.
 \end{equation}
 
-The weighted generating function of the segments terminated by
-$\downarrow_{/n}$ following segments terminated by $\downarrow_{/m}$ is
+This is the probability that the terminator is the symbol
+$\downarrow_{/n}$ given that the segment has size $j+1 > \gamma$, that
+the first sequencing error occurs on the last nucleotide, that the
+preceding terminator was $\downarrow_{/m}$ and that the $m$ hard masks
+fail before the end of the segment.
+
+Summing the terms from case 3a and from case 3b, we find that the weighted
+generating function of the segments terminated by $\downarrow_{/n}$
+following segments terminated by $\downarrow_{/m}$ is
 \begin{equation}
@@
 A_{m,n} =
-\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
-pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
-(1- \zeta_{i,m,n}) \Big) (qz)^i.
+\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - \xi_j^N \Big) (qz)^j + \omega_n
+pz \sum_{j=\gamma}^\infty \Big(1 - \xi_j^m \cdot
+(1- \zeta_{j,m,n}) \Big) (qz)^j.
 \end{equation}
 
-\subsection{Expression of the transfer matrix}
+\subsection{Expression of $\mathring{M}_N(z)$}
 \label{sec:expression_of_M}
 
-Collecting and arranging the results above, the final expression of the
-transfer matrix $M(z)$ appears as
+Collecting and arranging the results above, we can verify that the final
+expression of the transfer matrix $\mathring{M}_N(z)$ is
 \begin{equation*}
@@
 \tag{\ref{eq:xi}}
-\xi_i = 1-(1-\mu)^i \\
+\xi_j = 1-(1-\mu)^j \\
 \tag{\ref{eq:eta}}
-\eta_i = 1-(1-\mu)^i\mu/3 \\
+\eta_j = 1-(1-\mu)^j\mu/3 \\
 \tag{\ref{eq:zeta}}
-\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
-(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n} \\
+\zeta_{j,m,n} = \sum_{r\geq1} {N-m \choose r}
+(1-\mu)^{rj} \psi_{j,m,n,r} \bigg/ {N \choose n} \\
 \notag
-\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
-\xi_i^{n-r-q}.
+\psi_{j,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
+\xi_j^{n-r-q}.
 \end{gather}
 
-The special case $N=0$ is worth describing in detail. This implies that
-the target has no duplicate, so case $ii.$ is excluded (false positive),
-and only cases $i.$ and $iii.$ are possible (true positive and false
-negative, respectively). The transfer matrix $M(z)$ simplifies to the
-extent that we can directly compute the weighted generating function of
-reads without a good seed as
+\begin{remark}
+In the special case $N=0$, the transfer matrix simplifies to the extent
+that we can compute the weighted generating function of the reads without
+on-target MEM seed in closed form. The result is
 \begin{equation}
 \tag{\ref{eq:F}}
-F(z) = \frac{1+qz+\ldots+(qz)^{\gamma-1}}
+\frac{1+qz+\ldots+(qz)^{\gamma-1}}
   {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
@@
 
-Expression (\ref{eq:F}) was shown in~\cite{filion2018analytic} to be the
-weighted generating function of
-
-\section{Computing the probabilities}
-\label{sec:compute}
-
-\subsection{Coefficient extraction}
-\label{sec:extract}
-
-The purpose of constructing the weighted generating function $F(z) = a_0 +
-a_1z + a_2z^2 + \ldots$ is to extract the coefficient $a_k$, which
-represents the probability that a read of size $k$ does not contain a
-good seed.
-
-$F(z)$ is the entry of the matrix $M(z) \cdot (I-M(z))^{-1}$ at the first
-row and the last column, associated with terminators $\downarrow_{/0}$ and
-$|$. The expression of $M(z)$ is too complex to compute this function
-directly.
-
-Instead, we return to the definition $M(z) \cdot (I-M(z))^{-1} =
-M(z) + M(z)^2 + \ldots$ and observe that the terms $M(z)^{k+2},
-M(z)^{k+3}, \ldots$ have no influence on the coefficients $a_0, a_1,
-\ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments\footnote{Each
-segment contains a terminator, which has size 1, except the unique
-terminator $|$, which has size 0.}. Since the entries of $M(z)^n$ are the
-weighted generating functions of reads with exactly $n$ segments, $a_k$
-cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by
-induction that all the entries of $M(z)^k$ are divisible by $z^{k-1}$,
-showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to
-$a_0 + a_1z + \ldots +a_kz^k$ is $0$.
-
-So to extract the coefficients of $F(z)$ up to order $k$ we can compute
-the matrix $M(z) + M(z)^2 + \ldots + M(z)^{k+1}$, isolate the entry at the
-first row and the last column and then compute the terms of the Taylor
-expansion up to order $k$.
-
-But we can do better than that. Since we are only interested in the
-coefficients up to order $k$, we can replace the infinite power series
-of the terms $A_{m,n}(z)$ and $C_m(z)$ by their truncated versions where
-we keep only the terms up to order $k$. Likewise, we perform all
-algebraic operation on truncated polynomials of order $k$, \textit{i.e.},
-we discard at every stage the coefficients of order $k+1$ or greater.
-
-More formally, truncated polynomials are defined as polynomials whose
-coeffients of degree greater than $k$ are 0. Multiplying two truncated
-polynomials $P(z)$ and $Q(z)$ gives a truncated polynomial $R(z)$ whose
-coefficients up to order $k$ are those of $P(z)Q(z)$. Using truncated
-polynomials of order $k$, the entry of the matrix $M(z) + M(z)^2 + \ldots
-+ M(z)^{k+1}$ at the first row and the last column directly provides the
-coefficients of interest.
-
-But we can do even better than that. A read with $n$ segments contains at
-least $m=\lfloor{(n-1)/2}\rfloor$ errors. Indeed, two segments terminated
-by $\uparrow$ cannot follow each other, so at least (approximately) half
-of the segments must be terminated by $\downarrow$, which means that the
-segment contains a read error. Since read errors typically have a small
-probability of occurrence, all the coefficients of $M(z)^n$ rapidly
-converge to $0$ as $n$ increases. So instead of computing the matrix $M(z)
-+ M(z)^2 + \ldots + M(z)^{k+1}$, we can interrupt the summation after a
-certain power of $M(z)$ because the terms become negligible.
-
-The number of errors $X$ in a read of size $k$ has a Binomial
-distribution $X \sim B(k,p)$. From~\cite{arratia1989tutorial} we can bound
-the probabilities of the tail with the expression
-
-\begin{equation}
-\label{eq:bound}
-Pr(X \geq m) \leq \exp \left( (m-k)\log \frac{k-m}{k(1-p)} -m\log
-\frac{m}{kp} \right).
-\end{equation}
-
-Using the formula above, we can thus bound the probability that a read has
-$n$ segments or more. We compute the matrices $M(z) + M(z)^2 + \ldots
-+M(z)^n$ where the weighted generating functions have been replaced by
-truncated polynomials and we extract $F_n(z)$, the entry at the first
-column and the last row. When the bound is lower than a set fraction
-$\varepsilon$ of the coefficient of $z^k$, we stop the computations.
-Typically $\varepsilon = 0.01$ so this method ensures that the
-probabilities that a read of size $k$ has no good seeds are accurate to
-within 1\%.
-
-This approach is feasible when the matrix $M(z)$ is small, but recall that
-$M(z)$ has $N+\gamma+1$ rows/columns. Some sequences can have millions of
-duplicates in plant and animal genomes (\textit{i.e.} $N > 10^6$), which
-makes the computations with this method unfeasible because even a few
-matrix multiplications are prohibitive. For large $N$, another method is
-clearly needed.
+Expression (\ref{eq:F}) was shown in section~\ref{sec:example_exact} to be
+the weighted generating function of reads without on-target exact seed.
+This shows that when there are no duplicates, MEM seeds have exactly the
+same properties as exact seeds.
+\end{remark}
+
+\subsection{Computing MEM seeding probabilities}
+
+The matrix $\mathring{M}_N(z)\cdot(I-\mathring{M}_N(z))^{-1}$ contains the
+weighted generating functions of all the reads without on-target MEM
+seeds. The term of interest, as usual, is the top right entry. Indeed,
+every read can be prepended by segments terminated by $\downarrow_{/0}$
+and only by those, otherwise the read would start with fewer than $N$ soft
+masks. Thus, reads without an on-target MEM seed are precisely the
+sequences of segments that can be appended to the symbol $\downarrow_{/0}$
+and that are terimated by a tail.
+
+To compute this term, we proceed as in section~\ref{sec:example_skip},
+\textit{i.e.}, we compute the powers of $\mathring{M}_N(z)$ in the
+arithmetic of truncated polynomials and we stop the iterations when the
+terms are negligible. We bound the probability that the read contains more
+than $e$ sequencing errors using the expression
+\begin{equation*}
+\tag{\ref{eq:bound}}
+Pr(X \geq e) \leq \exp \left( (e-k)\log \frac{k-e}{k(1-p)} -e\log
+\frac{e}{kp} \right),
+\end{equation*}
+but here not every segment contains an error. However, there cannot be two
+symbols $\uparrow_{/j}$ in a row, so a read with $s$ segments so a tail
+must contain a minimum number of sequencing errors which is
+$e=\lfloor{s/2\rfloor}$. So we compute the powers of $\mathring{M}_N(z)$
+until the upper bound is less than a set fraction $\varepsilon$ of the
+current value.
+
+If we call $M_0$ the event that the read contains an on-target MEM seed,
+the method above gives us $P(\overline{M}_0)$. Calling $M_j$ the event
+that the read contains a MEM seed for the $j$-th duplicate, we are
+interested in the probability
+\begin{equation*}
+\tag{\ref{eq:pcomp}}
+P\big(\overline{M}_0 \cap (M_1 \cup \ldots \cup M_N)\big) =
+P(\overline{M}_0) - P\big(\overline{M}_0 \cap \overline{M}_1 \cap \ldots
+\cap \overline{M}_N\big).
+\end{equation*}
+
+The key insight to compute $P\big(\overline{M}_0 \cap \overline{M}_1 \cap
+\ldots \cap \overline{M}_N\big)$ is to realize that there is some MEM
+seed, on-target or not, if and only if the read contains a match of size
+$\gamma$ or more for any of the $N+1$ sequences. Therefore, this
+probability is the same as the term $P\big(\overline{S}_0 \cap
+\overline{S}_1 \cap \ldots \cap \overline{S}_N\big)$ computed in
+section~\ref{sec:offexact}.
+
+In conclusion, the probability that the MEM seeding process is off target
+is
+\begin{equation*}
+P(\overline{M}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
+\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N,
+\end{equation*}
+where $P(\overline{M}_0)$ is computed using $\mathring{M}_N(z)$ as
+explained in this section, $P(\overline{S}_0)$ is computed using a
+recursive equation as explained in section~\ref{sec:example_exact}, and
+$P(\overline{S}_0\cap\overline{S}_1)$ is computed using $\tilde{M}_0(z)$
+as explained in section~\ref{sec:dual}.
 
@@
 
+One potential difficulty in computing $P(\overline{M}_0)$ is that the
+matrix $\mathring{M}_N(z)$ has dimension $(N+\gamma+1) \times
+(N+\gamma+1)$. So for MEM seeds the problem can become computationally
+intractable because $N$ can be very large. For instance, the sequences
+called \emph{Alu} have more than one million duplicates in the human
+genome. There is no hope to compute the powers of $\mathring{M}_N(z)$ in
+these conditions and we need an alternative method.
+
 The symbolic representation as MEM segments can be used for an efficient
 method to sample reads. Instead of generating the nucleotides of the $N+1$
-sequences one by one, one can generate a single sequence of MEM segments.
+sequences one by one, one can generate a single sequence of segments.
 Since the number of segment does not depend on $N$, we can obtain a fast
 Monte Carlo method to sample millions of reads and count the proportion
-that contain a good seed.
+that contain an on-target MEM seed.
 
 The principle is to proceed in cycles of two steps. We first sample the
-position of the next read error, which gives the position of the next
-symbol $\downarrow$. The second step is to determine whether there is a
-symbol $\uparrow$ before that. For this we sample the number of masks that
-vanish before the symbol $\downarrow$. If they all vanish, the read
-contains a good seed, provided the next read error is at a distance
-greater than $\gamma$. Otherwise, the process is repeated until we
-generate a good seed, or until the read has size $k$ or greater (in which
-case it has no good seed).
-
-The method is summarized by the algorithm below. It requires efficient
-methods to generate random samples with a geometric distribution and a
-binomial distribution. Sampling from a geometric distribution can be done
-by computing the logarithm of a uniform $(0,1)$ random variable. Sampling
+position of the next sequencing error, which gives the position of the
+next symbol $\downarrow_{/m}$, where $m$ will be determined at a later
+stage. The second step is to determine whether there is a symbol
+$\uparrow_{/j}$ before that. For this we sample the number of masks that
+vanish before the symbol $\downarrow_{/m}$. If they all vanish, the read
+contains an on-target MEM seed, provided the next read error is at a
+distance greater than $\gamma$. Otherwise, we sample the number $m$ of
+hard masks at the sequencing error, and the process is repeated until we
+generate an on-target MEM seed, or until the read has size $k$ or greater
+(in which case it has no on-target MEM seed).
+
+The method is summarized in algorithm~\ref{alg:mcmc} below. It requires
+efficient methods to sample from the geometric and from the binomial
+distributions. Sampling from a geometric distribution can be done by
+computing the logarithm of a uniform $(0,1)$ random variable. Sampling
 from a binomial distribution can be done by the method of
 Kachitvichyanukul and Schmeiser~\cite{kachitvichyanukul1988binomial}.
+
+The running speed depends on the error rate of the instrument $p$, but not
+on the number of duplicates $N$. For values of $p$ around 0.01,
+the running time for 10,000,000 iterations is $1$-$10$ seconds on current
+hardware.
 
@@
 
-In practice, we use a slightly modified version of this algorithm, that
-returns the position of the good seeds if they are generated. This allows
-us to estimate the probabilities for \emph{all} read sizes up to $k$. We
-typically run this Monte Carlo sampling scheme 10,000,000 times, so that
-we can estimate probabilities as low as $\sim1$ in 1 million.
-
-The running speed depends on the error rate of the instrument $p$, but not
-on the number of off-target sequences $N$. For values of $p$ around 0.01,
-the running time for 10,000,000 iterations is $1$-$10$ seconds on current
-hardware.
-
-Algorithm~\ref{alg:mcmc} is a major result. It gives a compact solution to
-the problem of estimating the probability that a read can be mapped when
-using MEM seeds. The algorithm is also much faster than the naive approach
-of sampling every nucleotide of every sequence, because it samples many
-nucleotidea from \emph{all} the sequences at each iteration.
-
-\section{False positives}
-
-So far we have designed a method to estimate the probability that a
-sequencing read contains a good seed. We referred to this as outcome $i.$
-in section~\ref{sec:seedheur}. However, we are interested in the
-probability that a read is mapped at an incorrect location, which is
-outcome $ii.$
-
-Computing the probability of outcome $ii.$ is hard, but we can compute the
-probability that the read is not mapped at all, which is outcome $iii.$
-The probability of outcome $ii.$ is then found by subtraction.
-
-\subsection{Probability of outcome $iii.$}
-
-Outcome $iii.$ implies that the read contains no seed at all, neither
-on-target, nor off-target. This requires that there is no exact match of
-size $\geq \gamma$ for \emph{any} sequence. Indeed, if it were the case,
-then there would a MEM seed of size $\geq \gamma$ for at least one of the
-sequences.
-
-Let $E_0$ denote the event that there is no on-target match of size $\geq
-\gamma$, and $E_j$ denote the event that there is no match of size $\geq
-\gamma$ for off-target sequence $j$ ($1 \leq j \leq N$). We are thus
-interested in computing $P(E_0 \cap \ldots \cap E_N)$. Since we assume
-that the off-target sequences have evolved through the same mutagenesis
-process and independently of each other, we can write
-
-\begin{eqnarray}
-\notag
-P(E_0 \cap \ldots \cap E_N) &=&
-  P(E_0) \cdot P(E_1 \cap \ldots \cap E_N | E_0) \\
-\label{eq:PE0E1}
-  &=& P(E_0) \cdot P(E_1 | E_0)^N =
-  P(E_0) \cdot \left( \frac{P(E_0 \cap E_1)}{P(E_0)} \right)^N.
-\end{eqnarray}
-
-Expression (\ref{eq:PE0E1}) shows that we can compute the probability of
-outcome $iii.$ from two quantities, namely $P(E_0)$ and $P(E_0 \cap E_1)$.
-
-\subsection{Computing $P(E_0)$}
-\label{sec:cPE0}
-
-$P(E_0)$ is straightforward to evaluate because it involves a single
-sequence. Methods to do this efficiently have been explained in detail
-elsewhere~\cite{filion2018analytic}. Here we only give the recurrence
-equation to compute $P(E_0)$. Assume that the probability of a sequencing
-error $p$ is known, that $q = 1-p$, and that $a_k$ is the probability that
-a read of size $k$ does not contain a match of size $\geq \gamma$ for the
-target. This case was already treated in section~\ref{sec:example_exact}.
-
-Knowing the size of the read and the error rate of the sequencing
-instrument, we can thus compute $P(E_0)$ using expression (\ref{eq:PE0}).
-
-\subsection{Probability of false positives}
-
-For given values of the parameters $p$, $\mu$ and $N$, we can now compute
-the probability of a false positive, which is outcome $ii$.
-
-For this, we compute the probability of outcome $i.$, either by computing
-the powers of $M(z)$ or by Monte Carlo sampling as explained in
-section~\ref{sec:compute}. Next we compute the probability of outcome
-$iii.$ using expression (\ref{eq:PE0E1}), where the terms are computed as
-explained in sections~\ref{sec:cPE0E1} and \ref{sec:cPE0}.
-
-When the target has no duplicate sequence, there can be no false positive
-so this probability must be 0. In this case $N=0$ so according to formula
-(\ref{eq:PE0E1}) outcome $iii.$ has probability $P(E_0)$.
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-XXXX
-
-\section{Estimating the parameters}
-\label{sec:est}
-
-So far we have assumed that we know all the specifications of the problem,
-including the error rate of the instrument $p$, the number of off-target
-sequences $N$, and their rate of sequence divergence $\mu$ (or more
-accurately the probability that they differ from the target at every
-nucleotide).
-
-In general, $p$ is known from the vendor of the sequencing instrument, but
-$N$ and $\mu$ can change for every read and they are usually unknown. We
-thus need an estimation algorithm that can run fast enough that it does
-not slow the mapping process. The main idea here is to use the backward
-search, which is the method used to find the MEM seeds in the first place.
-
-The backward search is a method that was first proposed by Ferragina and
-Mazini~\cite{ferragina2000opportunistic} to identify exact matches between
-a text query and an index. The query is extended one letter/nucleotide at
-a time, and at each iteration the search returns the number of matches in
-the index/genome. This is thus a very efficient method to identify MEM
-seeds, because one can keep extending the query until there is no match in
-the genome or until the end of the read.
-
-What matters for us is that we can run the backward search at some
-arbitrary position of the read. The $i$-th iteration returns the number
-$N_i$ of matches in the genome for a query sequence of length $i$.
-Ignoring sequencing errors and focusing only on the $N$ off-target
-sequences, the probability that the count goes from $N_i$ to $N_{i+1}$ is
-
-\begin{equation*}
-P(N_{i+1}|N_i) = {N_i \choose N_{i+1}}
-\mu^{N_i-N_{i+1}}(1-\mu)^{N_{i+1}}.
-\end{equation*}
-
-Assuming that there are $N = N_0$ duplicates and that we run $m$
-iterations of the backward search, the probability of the observations
-$(N_1, N_2, \ldots, N_m)$ is thus
-
-\begin{equation*}
-P(N_1, \ldots, N_m|N_0) = 
-{N_0 \choose N_1 } \ldots {N_{m-1} \choose N_m }
-\mu^{N_0-N_m}(1-\mu)^{N_1+\ldots+N_m}.
-\end{equation*}
-
-This is a familiar distribution in disguise: by defining $\Delta_i =
-N_{i-1} - N_i$, the series of observations $(\Delta_1, \ldots, \Delta_m,
-N_m)$ given $N_0$ has a multinomial distribution with parameters
-$\big(\mu, \mu(1-\mu), \ldots, \mu(1-\mu)^{m-1}, (1-\mu)^m\big)$. More
-specifically, the probability of observing the sequence $(\Delta_1,
-\ldots, \Delta_m, N_m)$ given $N_0$ is
-
-\begin{align*}
-{N_0 \choose \Delta_1, \ldots, \Delta_m, N_m}
-\mu^{\Delta_1}\big(\mu(1-\mu)\big)^{\Delta_2}
-\ldots \big(\mu(1-\mu)^{m-1}\big)^{\Delta_m}(1-\mu)^{mN_m}.
-\end{align*}
-
-At that point, we need to bring back into the light the practical
-consideration that short matches are uninformative. This implies that the
-first iterations of the backward search say nothing about the number of
-duplicates of the sequence. In practice, we are better off ignoring the
-output of the first $n-1$ iterations by just discarding them. We observed
-that running 30 iterations of the backward search and discarding the
-output of the first 20 gives good results (\textit{i.e.} we set $n = 21$
-and $m=30$).
-
-Now, if $N_0, N_1, \ldots, N_{n-1}$ are missing, we obtain the probability
-of observing $(\Delta_{n+1}, \ldots, \Delta_m, N_m)$ from the marginal
-distribution as
-
-\begin{align*}
-\begin{split}
-{N_0 \choose N_0-N_n, \Delta_{n+1}, \ldots, \Delta_m, N_m}
-\big(1-(1-\mu)^n\big)^{N_0-N_n}
-\big(\mu(1-\mu)^n\big)^{\Delta_{n+1}} \times
-\ldots \\
-\ldots \times
-\big(\mu(1-\mu)^{m-1}\big)^{\Delta_m}(1-\mu)^{mN_m}.
-\end{split}
-\end{align*}
-
-Now replacing again $\Delta_i$ by $N_{i-1}-N_i$, the log-likelihood can be
-expressed as
-
-\begin{equation}
-\begin{split}
-\ell = cst + \log \Gamma(N_0+1) - \log \Gamma(N_0-N_n-1)
-+ (N_0-N_n) \log \big(1-(1-\mu)^n\big) + \\
-(N_n-N_m) \log(\mu) + (nN_n + N_{n+1} + \ldots + N_m) \log(1-\mu).
-\end{split}
-\end{equation}
-
-To estimate the parameters by maximum likelihood, we differentiate $\ell$
-with respect to $N_0$ and with respect to $\mu$, then we equate both terms
-to 0. We thus have to solve
-
-\begin{gather}
-\label{eq:psi}
-\Psi(N_0+1)-\Psi(N_0-N_n+1) +
-  \log\big( 1-(1-\mu)^n \big) = 0, \text{ and} \\
-\label{eq:N0}
-\frac{N_n-N_m}{\mu}
--\frac{nN_n + N_{n+1}+\ldots+N_m}{1-\mu} +
-(N_0-N_n)\frac{n(1-\mu)^{n-1}}{1-(1-\mu)^n} = 0.
-\end{gather}
-
-In the expressions above, $\Psi(\cdot)$ is the digamma function, defined
-as the derivative of $\log \Gamma(\cdot)$. We observe that equation
-(\ref{eq:N0}) gives an expression of $N_0$ as a function of $\mu$ and of
-the observations, namely
-
-\begin{equation*}
-N_0 = N_n +
-\left(\frac{nN_n+N_{n+1}+\ldots+N_m}{1-\mu}-
-\frac{N_n-N_m}{\mu} \right)
-\frac{1-(1-\mu)^n}{n(1-\mu)^{n-1}}.
-\end{equation*}
-
-To find the maximum likelihood estimate of $\mu$, we can substitute this
-expression in equation (\ref{eq:psi}), obtaining an equation depending on
-$\mu$ and on the observations, which can be solved numerically. One can
-then substitute the solution into equation (\ref{eq:N0}) to obtain the
-estimate of the number of duplicate sequences $N_0$.
-
-However, the estimations of $\mu$ and $N_0$ show large deviations from the
-target values on simulated data. In practice, one does not need to know
-the parameters with exquisite precision, so to avoid unrealistic
-estimates, we pick the best $\mu$ from a short list of candidate values
-(\textit{e.g.}, 0.02, 0.04 and 0.06). To do this, we choose a value
-$\mu^*$ from the candidate set, we compute $N_0(\mu^*)$ from equation
-(\ref{eq:N0}) and we evaluate $\ell$ with these values. Finally, we select
-the $\mu$ that gives the higest value of $\ell$, and the estimate for
-$N$ is set to $N_0(\mu)$.
-
-Finding the estimates of $N$ and $\mu$ can be performed in negligible time
-compared to running the backward search. We can repeat this method at
-different areas of the read and finally average the estimates to get
-reasonable values of $N$ and $\mu$ for the whole read. With this
-information, we can compute the probability that the read has no good seed
-using the methods presented above.
-
-\section{Neglected practical considerations}
-
-To keep the focus on the MEM seeding process, we have ignored some
-practicalities that we now have to revisit in order to evaluate the
-feasibility of our approach.
-
-\subsection{Is the true location the best one?}
-\label{sec:not_the_best}
-
-So far we have assumed that the true location is also the best one, so
-that it is always discovered when it is in the candidate set. But how
-likely is it that the true location is not the best one, and thus that it
-is virtually impossible to discover?
-
-Obviously, it is possible that the target has an exact duplicate. When
-that happens, the read cannot be mapped unambiguously, but this case is
-relatively easy to detect, so the mapper can return a low confidence score
-and thus avoid making an error.
-
-From now we assume that the read has no strict duplicate. The probability
-that the read of size $k$ has $m$ errors is
-\begin{equation*}
-{k \choose m}p^m(1-p)^{k-m}.
-\end{equation*} 
-A given duplicate has a better score if its Hamming distance from the read
-is less than $k$. In a first order approximation, this happens if the
-duplicate matches one of the $k$ errors, and also the $L-k$ correct
-nucleotides. The probability of this even is
-\begin{equation*}
-m(1-\mu)^{k-m}\mu/3.
-\end{equation*}
-
-There are $N$ duplicates that evolve independently, so the probability
-that one of them has a better score than the target is
-\begin{equation*}
-1-\left(1-m(1-\mu)^{k-m}\mu/3\right)^N \approx Nm(1-\mu)^{k-m}\mu/3.
-\end{equation*}
-
-Combining the first and the last expression, we find that the probability
-that the target does not have the best score is approximately
-\begin{equation}
-\label{eq:not_the_best}
-\sum_{m=0}^k {k \choose m}p^m(1-p)^{k-m} Nm\mu/3(1-\mu)^{k-m} = 
-Nk p \left(1-\mu(1-p)\right)^{k-1}\mu/3.
-\end{equation}
-
-For instance, for reads of size $k=100$ with error rate $p=0.01$, where
-the target has $N=5$ duplicates with divergence $\mu=0.05$ we obtain an
-estimate close to $10^{-4}$. In comparison, the probability of a false
-positive when using MEM seeds with a minimum size $\gamma=17$ is
-approximately $4.4\cdot10^{-4}$.
-
-The probability that the target does not have the best score is usually
-small, but it is of the same order of magnitude as the probabilities of
-false positives with MEM seeds. The reason is that a common source of
-false positives are the cases when the target does not have the best
-score. Indeed, in this case the target is much less likely to have a MEM
-seed.
-
-In short, the event that the true location is not the best one is already
-incorporated in the false positive rate and we do not recommend further
-updating the estimates. In the case that one is interested in this event
-\textit{per se}, formula (\ref{eq:not_the_best}) gives an approximate
-probability of its occurrence. When $\mu$ is unknown, we can replace it by
-its worst case value $\mu = 1/qL$, where $q=1-p$. But $N$ must be known at
-least approximately because the proability increases with $N$.
-
-In summary, the case that the true location is not the best one has a
-small but not negligible probability. It is partly redundant with the
-false positives generated during the MEM seeding process, and therefore it
-is not straightforward to adjust the estimates accordingly. We do not
-adjust them at all and consider that this case is already implicitly
-accounted for.
-
-\subsection{MEM seeds for other sequences}
+Algorithm~\ref{alg:mcmc} is an important result. It gives a compact
+solution to the problem of estimating the probability that a read can be
+mapped when using MEM seeds. The algorithm is also much faster than the
+naive approach of sampling every nucleotide of every sequence, because it
+is equivalent to sampling the nucleotide sequence of \emph{all} the
+duplicates.
+
+\subsection{Illustration}
+
+Here goes.
+
+\section{Random seeds}
 \label{sec:short_MEMs}
 
-To simplify the problem, we have assumed that MEM seeds can match only the
-target or one of its duplicates. In practice, MEM seeds that overlap a
-sequencing error often match random locations of the genome instead of the
-duplicates. When this occurs, the MEM seeds are typically small because 
-the match is random.
-
-We select MEM seeds of size at least $\gamma$, and it is important for the
-validity of the theory that all the MEM seeds above the threshold are
-considered candidate locations. If not, the estimates of the false
-positive rate are off. The hope is that spurious MEM seeds are shorter
+We have assumed throughout that seeds can match only the target or one of
+its duplicates. In practice, seeds that overlap a sequencing error often
+match random locations of the genome instead of the duplicates.
+
+It is important for the validity of the theory that all the sequences that
+have a seed are considered candidate locations (as long as the seed is
+above the threshold $\gamma$ of course). If not, the estimates of the
+mismapping rate are off. The hope is that spurious matches are shorter
 than $\gamma$ so that they are invisible, but this is not always the case.
-And if they qualify, we need to verify them.
+And if they qualify as a seed we need to verify the candidate location.
 
@@
 verifying them is wasted. Second, they can cause false positives.
-Fortunately, both issues can be addressed so that they do not constitute a
-major problem.
+Fortunately, both issues can be addressed.
 
@@
 distance 1 from the read, there is no need to verify candidates from MEM
-seeds in the middle of the read. Indeed, the hit must be at distance at
-least 2 because there is a mismatch before the MEM seed and another one
-after it.
-
-These considerations depend on the implementation of the mapper so we will
-not further develop them. What matters is that random hits do not have to
-impose a significant burden on the time needed to verify the candidates.
+seeds than span from neither end of the read. Indeed, the hit must be at
+distance at least 2 because there is a mismatch before the MEM seed and
+another one after it. These considerations depend on the implementation of
+the mapper so we will not delve deeper here. What matters is that random
+hits do not have to impose a significant burden on the time needed to
+verify the candidates.
 
 The second issue is that random hits can generate false positives. Such
-false positives happen when the read does not belong to the reference
-genome, or when it contains so many errors that no proper MEM seed was
-found. Since the hit is not homologous to the read, the alignment score is
-essentially random. The upshot is that it is easy to detect.
+cases occur when seeding is null (meaning that there is no seed for the
+target nor for its duplicates). Since the hits are not homologous to the
+read, the alignment score is essentially random, which makes it is easy to
+detect.
 
 Indeed, if the candidate location is random, mismatches occur with
-probability 3/4. If it is the right location, mismatches occur with
-probability $p$. We discard the MEM seed, which is an automatic match of
-size at least $\gamma$. Say that there remains $L$ nucleotides and that
-$m$ of them are mismatches for the candidate location. Using Bayes
-formula, we find that the probability that the location is random given
-the observed alignment score is
+probability 3/4. If it is the true location, mismatches occur with
+probability $p$. We discard the seed because it is an automatic match of
+size $\gamma$ (and possibly larger in the case of MEM seeds). Say that
+there remains $L$ nucleotides and that $m$ of them are mismatches for the
+candidate location. From Bayes formula, the probability that the location
+is random given the number of mismatches is
 \begin{equation}
@@
 
-One may have information about the level of contamination, allowing to
-choose a meaningful value for $\beta$. For instance, sequencing runs are
-often spiked by foreign DNA such as the phiX genome or other. If these
-reads are not filtered automatically, $\beta$ can be set to their expected
-frequency. For sequences that belong to the reference genome, random hits
-are rather rare because the read must contain so many error that neither
-the target nor its duplicates were discovered. The value of $\beta$ is
-much smaller in this case, but nevertheless difficult to estimate.
-
-That being said, the value of $\beta$ has little importance in practice.
-Say that a read generates a hit in the genome such that 33 nucleotides
-need to be aligned (the alignment status of the others was already known
-from the seeding step). If the hit is random there is a 99.99\% chance
-that at least 15 are mismatches. If $p = 0.01$, the denominator of
+The value of $\beta$ has little importance if $p$ is small. Say that
+$p=0.01$ and that a read generates a hit in the genome such that 33
+nucleotides need to be aligned after seeding. If the hit is random there
+is a 99.99\% chance that at least 15 are mismatches. The denominator of
 expression (\ref{eq:bayes}) is approximately $1 +
@@
 support for the hypothesis that the hit is random is overwhelming.
-
 Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
-less are mismathes. In this case, the denominator is approximately
-$1+6.8\cdot10^9(1-\beta)/\beta$, so unless $\beta > 1-10^9$, the support
-for the hypothesis that the hit is not random is overwhelming.
-
-In general, one does not need to worry about the value of $\beta$ because
-random hits are extremely visible when $p$ is small. For instance, one can
-choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears in
-expression (\ref{eq:bayes}).
-
-Because the probability that the best candidate is a random sequence is
-either very small or very large, it either has no influence on the
-probability of false positive, or completely dominates it. For every read,
-we thus take the maximum of this probability or the estimated probability
-that MEM seeding returned a false positive.
-
-In summary, MEM seeds in random sequences happen relatively frequently,
-but it is possible to minimize their computational burden. Also, they
-are no cause for concern regarding false positives because they can easily
-be detected using Bayes formula.
+fewer nucleotides are mismathes. In this case, the denominator is
+approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
+10^{-9}$, the support for the hypothesis that the hit is not random is
+overwhelming. If $p$ is small, we do not need to worry about the value of
+$\beta$, one can choose for instance $\beta=1/2$ so that the term
+$(1-/\beta)/\beta$ disappears in expression (\ref{eq:bayes}).
+
+Since the probability that the best candidate is a random sequence is
+either very small or very large, it either has no influence or dominates
+the probability of a false positive. For every read, we can use as
+confidence score the maximum of this probability of expression
+(\ref{eq:bayes}) and of the estimated probability that seeding is off
+target.
+
+In summary, seeds in random sequences happen relatively frequently, but it
+is possible to minimize their computational burden. Also, they are no
+cause for concern regarding false positives because they can easily be
+detected using Bayes formula, as shown in expression (\ref{eq:bayes}).
 
@@
 
-We have devised a method to compute the probability that the MEM seeding
-heuristic fails and sets up the mapping process for a false positive. This
-fills a gap of theoretical knowledge to understand and calibrate the
-performance of the MEM seeding heuristic. The pillars of our strategy are
-borrowed from analytic combinatorics~\cite{flajolet2009analytic,
-sedgewick2013introduction}, even though we do not follow the standard
-programme. Constructing generating functions usually serves the purpose of
-finding their singularities in order to approximate the solution. In our
-case, however, the weighted generating functions cannot be computed so
-this strategy is not applicable.
+We have devised a set of methods to compute the probability that seeding
+heuristics fail and sets the mapping process for a false positive. This
+fills a knowledge gap to understand and calibrate the performance of the
+seeding heuristics. The pillars of our strategy are borrowed from analytic
+combinatorics~\cite{flajolet2009analytic, sedgewick2013introduction}, even
+though we do not follow the complete programme. Constructing generating
+functions usually serves the purpose of finding their singularities in
+order to approximate the solution. In our case, however, the weighted
+generating functions cannot be computed so this strategy is not
+applicable.
 
@@
 weighted generating function, we compute only the first terms using
-iterative methods as in sections~\ref{sec:extract}, \ref{sec:cPE0} and
-\ref{sec:cPE0E1}, or using Monte Carlo sampling as in
+iterative methods as in explained in section~\ref{sec:example_skip}, 
+or using Monte Carlo sampling as explained in
 section~\ref{sec:montecarlo}. In this regard, the breakthrough is the
@@
 real genomes, the estimations are only approximately correct and this is
-but one of several simplifying assumptions. The theory is substantially
-intrincate; insertions and deletions would make it intractable.
-It is presently unclear how gracefully incorporate them in the theory.
-
-The second assumption is that all the candidate locations with a MEM
-seed of size $\geq \gamma$ are tested with the exact sequence alignment
-algorithm, so that the best location is always discovered if it is in the
-candidate set. This is not completely unrealistic, but it is important to
-realize that for plant and animal genomes, a shared MEM seed may have tens
-of thousands of hits. Therefore, most mappers impose a limit on the number
-of alignments so the best hit may not be found even if it is in the
-candidate set. This is again a minor impediment, since we can correct the
-probability that the location is a false positive when only some
-candidates have been tested.
+but one of several simplifying assumptions. Incorporating insertions and
+deletions would make the theory intractable, so it is presently unclear
+how to deal with this type of error.
+
+The second assumption is that all the candidate locations with a seed are
+tested with an exact sequence alignment algorithm, so that the best
+location is always discovered if it is in the candidate set. This is not
+completely unrealistic, but it is important to realize that for plant and
+animal genomes, a single seed may have tens of thousands of hits.
+Therefore, most mappers impose a limit on the number of alignments so the
+best hit may not be found even if it is in the candidate set. This is
+again a minor impediment, since we can correct the probability that the
+location is a false positive when only some candidates have been tested.
 
@@
 reconstruction techniques, but it would be challenging to incorporate this
-information in the present theory. The symbols of the MEM alphabet
-implicitly assume that the sequences are exchangeable and the complexity
-of the calculations explodes if it is not the case.
-
-The last assumption is that MEM seeds can match only the target or one of
-its duplicates. This is not the case but we have shown how to gracefully
-deal with this in section~\ref{sec:short_MEMs}.
+information in the present theory. The symbols of the alphabets developed
+above implicitly assume that the sequences are exchangeable and the
+complexity of the calculations explodes if it is not the case.
+
+The last assumption is that seeds can match only the target or one of its
+duplicates. This does not hold in general but we have shown how to deal
+with this possibility \textit{a posteriori} in
+section~\ref{sec:short_MEMs}.
 
@@
 straightforward, so one may wonder why the approach with weighted
-generating functions is necessary at all. The only reason is precision. To
-estimate the frequency of an event by Monte Carlo sampling, it must occur
-at least a few times in the simulation. For instance, with 10 million
-rounds of sampling, frequencies around $1/1,000,000$ or lower cannot be
-measured accurately. When not interested in rare events, it is thus a
-perfectly reasonable strategy to use only Monte Carlo sampling.
-
-For $N < 20$, the probability that a read does not contain a good seed is
-the smallest, so we need a method that is accurate in this range.
-Fortunately, the transfer matrix method is fast in this regime because the
-matrix $M(z)$ has few rows/columns.
-
-The proposed methods are compliant with the demand for speed of the
-mapping algorithms. One needs to compute the probabilities only once for a
-given value of $N$ and $\mu$ (the error rate $p$ is known and constant).
-For $N > 20$, the iterative method is usually prohibitive and we need to
-use Monte Carlo sampling instead. The running time depends on the desired
-number of iterations, on $p$ and on the size of the reads $k$. Since those
-are constant throughout the sequencing run, the method always takes the
-same time (around 1 second for reads of size $100$-$200$ nucleotides on
-modern hardware). The values of $N$ and $\mu$ can be binned in intervals
-of interest so that there are only around $100$ pairs for a total cost of
-a few minutes per run. Considering that mappers can process at most
-$10,000$ reads per second per core, the time of mapping a sequencing run
-of 250 million reads is over 7 hours, two orders of magnitude larger than
-the time required to estimate the probabilities of error.
+generating functions is necessary for MEM seeds. The only reason is
+precision. To estimate the frequency of an event by Monte Carlo sampling,
+it must occur at least a few times in the simulation. For instance, with
+10 million rounds of sampling, frequencies around $1/1,000,000$ or lower
+cannot be measured accurately. When not interested in rare events, it is
+thus a perfectly reasonable strategy.
+
+For $N < 20$, the probability that MEM seeding is null or off-target is
+relatively small, so we need a method that is accurate in this range.
+Fortunately, the transfer matrix method is fast because the dimensions of
+the matrix $\mathring{M}_N(z)$ are not prohibitive.
+
+The proposed methods meet the demand for speed. One needs to compute the
+probabilities only once for a given value of $N$ and $\mu$ (the error rate
+$p$ is known and constant). For $N > 20$, the iterative method is usually
+prohibitive and we need to use Monte Carlo sampling instead. The running
+time depends on $p$, on the size of the reads $k$, and on the desired
+number of iterations. Since those are constant throughout the sequencing
+run, the method always takes the same time (around 1-10 seconds for reads
+of size $100$-$200$ nucleotides on modern hardware). The values of $N$ and
+$\mu$ can be binned in intervals of interest so that there are only around
+$100$ pairs for a total cost of a few minutes per run. Considering that
+mappers can process at most $10,000$ reads per second per core, the time
+of mapping a sequencing run of 250 million reads is over 7 hours, two
+orders of magnitude larger than the time required to estimate the
+probabilities of error.
+
+Finally, one may wonder if our approach has any advantage over methods
+based on machine learning. This class of algorithms has already proved
+useful~\cite{lee2014mosaik} and the rapid progress in the field of deep
+learning suggests that it is possible to train algorithms to accurately
+estimate mapping quality. In time, such algorithms may prove faster and/or
+more robust because they could learn intrinsic biases of the mapping
+algorithms. Yet, the main benefit of our approach will remain: the
+combinatorial constructions are a direct access to the nature of the
+problem. For instance, viewing MEM seeds through the lens of hard and soft
+masks turns a seemingly intractable process into a relatively simple one
+(see algorithm~\ref{alg:mcmc}). The combinatorial stance is that there is
+value in clarity.
 
 In conclusion, we presented a practical solution to the problem of
-estimating the probability of false positives when using MEM seeds. This
-solution is adapted for mapping short reads sequenced with the Illumina
-technology. Being able to calibrate the MEM seeding heuristic not only
-allows the user to choose how to balance speed versus accuracy, but also
-opens new applications. For instance, one can map reads from contaminated
-samples in pools of closely related genomes (\textit{e.g.}, modern human
-and Neanderthal) in order to assign the reads to the organism they belong
-to. The probabilities of false positives give the right level of
-confidence in the assignment.
+estimating the probability of false positives when using seeding
+heuristics. This solution is adapted for mapping short reads sequenced
+with the Illumina technology. Being able to calibrate the seeding
+heuristic not only allows the user to choose how to balance speed versus
+accuracy, but also opens new applications. For instance, one can map reads
+from contaminated samples in pools of closely related genomes
+(\textit{e.g.}, modern human and Neanderthal) in order to assign the reads
+to the organism they belong to. The probabilities of false positives give
+the right level of confidence in the assignment.
 
@@
 computed. Using the same ideas, one could calibrate heuristics used in
-other alignment methods, especially in expanding fields such as long-read
+other alignment methods, especially the expanding field of long-read
 technologies.

### 2019-04-03 MEM.tex
First version ready to submit full manuscript (not copy-edited).

@@
 
-\title{Estimating mapping quality when using seeding heuristics for
+\title{On mapping errors and seeding heuristics for
 DNA alignment}
@@
 The increasing throughput of DNA sequencing technologies demands faster
-algorithms to analyze sequencing data. The fate of most reads is to be
-mapped to a reference sequence, typically a genome. Modern mappers rely on
+algorithms to analyze the data. The fate of most reads is to be mapped to
+a reference sequence, typically a genome. Modern mappers rely on
 heuristics to gain speed at a reasonable cost for accuracy. In the seeding
@@
 narrow the search to a set of candidate locations. Several seeding
-variants used in modern mappers show good empirical performance but
-theoretical results are still lacking, making them difficult to calibrate
-or top optimize.
+variants used in modern mappers show good empirical performance but they
+are difficult to calibrate or to optimize for lack of theoretical results.
 % Summary of the results.
 Here we develop a theory to estimate the probability that reads are
-mismapped due to a seeding failure. We describe the properties of simple
-exact seeds, skip-seeds used in Bowtie2 and MEM (Maximal Exact Match)
-seeds used in BWA-MEM. 
+mismapped due to failures at the seeding step. We describe the properties
+of simple exact seeds, skip-seeds and MEM seeds (Maximal Exact
+Match).
 % Innovation.
 The major innovation is to use concepts from analytic combinatorics to
-represent reads as abstract sequences, from which we estimate the
-probabilities of interest.
+represent reads as abstract sequences, and to specify their generative
+function to estimate the probabilities of interest.
 % Durability.
@@
 % Scope.
-These results significantly improve current mapping algorithms and lay the
+These results can improve current mapping algorithms and lay the
 foundation of a general strategy to tackle sequence alignment problems.
@@
 
-Suppose that a DNA fragment drawn at random from a known
-genome\footnote{We will assume that the sequence of the genome is known
-exactly and with absolute certainty. This is not the case in practical
-application, but this assumption will allow us to develop a theory
-centered on the read.} is sequenced with an imperfect instrument; how can
-we identify the location of the fragment in the genome?
+Suppose that we use and imperfect instrument to sequence a DNA fragment
+drawn at random from a known genome; how can we identify the location of
+the fragment in the genome?
 
@@
 selecting breeds of interest in agriculture; tracing human migrations from
-ancient bones; identifying the donor of a biological sample; diagnosing
-genetic diseases at birth; finding compatible graft donors \textit{etc}.
-
-At first sight, whether a read read can be mapepd to its correct location
-seems to depend only on its length and on the error rate of the sequencer.
+ancient bones; diagnosing genetic diseases at birth; finding compatible
+graft donors \textit{etc}.
+
+At first sight, whether a read can be mapped to its correct location seems
+to depend only on its length and on the error rate of the sequencer.
 However, there is another important factor to consider: most genomes
@@
 
-This case is rare though. Repeated sequences are in general not identical,
-but merely homologous\footnote{By that we mean that drawing nucleotides
-uniformly at random is very unlikely to produce sequences with the
-observed degree of identity, meaning that the similarity is due some
-biological duplication process.}. So when the DNA fragment is duplicated,
-we can still map it, as long as we can distinguish it from its duplicates.
-This in turn depends on their similarity with the target fragment and on
-the error rate of the sequencer.
+This case is rare though. Repeated sequences are in general not identical
+but merely homologous --- meaning that their similarity is unlikely to
+occur by chance. So when the DNA fragment has one or more duplicates in
+the genome, we can still map it, as long as we can distinguish the
+duplicates. This in turn depends on their similarity with the target
+fragment and on the error rate of the sequencer.
 
@@
 reason has received substantial attention in bioinformatics. For
-simplicity we will assume that the true location is always the best. We
-will relax this assumption in section section~\ref{sec:not_the_best}, but
-until then we will consider that the solution of the true location problem
-is the best location.
+simplicity we will assume that the true location is always the best, and
+we will often confound these two ideas.
 
@@
 
-The slowness of the alignment algorithm imposes a trade-off between speed
-and accuracy: If the seeding step is set to produce many candidates, it is
-likely to retrieve the candidate. The downside is that this will take
-long. Conversely, if the seeding step is set to produce few candidates,
-the process will run faster but the target is less likely to be in the
-candidate list.
+The slowness of exact alignment algorithms imposes a trade-off between
+speed and accuracy: If the seeding step is set to produce many candidates,
+it is likely to retrieve the candidate. The downside is that this will
+take long. Conversely, if the seeding step is set to produce few
+candidates, the process will run faster but the target is less likely to
+be in the candidate set.
 
@@
 second kind is easier to recognize because in this case the best hit is
-not similar to the read.
+not similar to the read, as we will highlight in
+section~\ref{sec:random_seeds}.
 
@@
 If a seed-based mapping algorithm returns an incorrect location, then it
-must be the case that the best location was never tested in the second
-phase, otherwise it would have been singled out by the exact alignment
-algorithm.
-
-This observation is fundamental: in order to estimate the mismapping rate
-of an algorithm, we must know how often the associated seeding heuristic
-is off target. For lack of theoretical framework, this information is
-missing for nearly all existing seeding methods, meaning that current
-mappers cannot be properly calibrated.
-
-The focus of this article is to develop a theory to estimate the
-probability that seeding is off target under the most popular seeding
-heuristics for short reads. Our previous work pioneered a method to
-compute seeding probabilities but it did not distinguish off-target from
-null seeding~\cite{filion2017analytic,filion2018analytic}, and therefore
-did not provide a way to control the mismapping rate. Other authors
-investigated the reliability of mapping algorithms~\cite{pmid23872968},
-but they focused on the probability that random sequences have a single
-hit, recognizing that solving the problem requires taking into
-consideration the repeat structure of the genome. Here we address this
-problem in general terms and we provide a practical way to estimate the
-mismapping rate in complex genomes.
+must be the case that the best location was never tested in the
+post-filtration phase (otherwise it would have been found by the exact
+alignment algorithm).
+
+This observation is fundamental: in order to estimate the error rate of of
+a mapping algorithm, we must know how often the associated seeding
+heuristic is off target. For lack of theoretical framework, this
+information is missing for nearly all existing seeding methods, meaning
+that current mappers cannot be properly calibrated.
+
+Our focus here is to develop a theory to estimate the probability that
+seeding is off target, when using the most popular seeding heuristics for
+short reads. Previous work pioneered a method to compute seeding
+probabilities but it did not distinguish off-target from null
+seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
+not provide a way to control mapping errors. Other authors investigated
+the reliability of mapping algorithms~\cite{pmid23872968}, but they
+focused on the probability that random sequences have a single hit,
+recognizing that solving the problem requires taking into consideration
+the repeat structure of the genome. Here we address this problem in
+general terms and we provide a practical way to estimate the mismapping
+rate in complex genomes.
 
@@
 then introduce the computational strategy based on transfer matrices, and
-then delve into each specific seeding method. Finally, we provide a
-statistical method to estimate the parameters of the problem from the
-data.
+then delve into each specific seeding method.
 
@@
 
-The term ``seed'' does not have exactly the same meaning according to
-different authors. It can designate a part of the read, a part of the
-genome, a particular sequence motif or a structured pattern of matches.
-Also, the term does not always imply an exact match. For instance, the
-algorithm PatternHunter~\cite{pmid11934743} uses ``spaced seeds''
-that tolerate mismatches\footnote{Spaced seeds are covered
-by patent US20070088510A1.}. To avoid potential confusion,
-we will adopt the following convention.
+The term ``seed'' changes meaning with different authors. It can designate
+a part of the read, a part of the genome, a particular sequence motif or a
+structured pattern of matches. Also, the term does not always imply an
+exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
+uses ``spaced seeds'' that tolerate mismatches. To avoid potential
+confusion, we will adopt the following convention.
 
@@
 transform~\cite{burrows1994block}, emulating a suffix trie with a much
-smaller memory footprint. Queries in the FM-index are performed with an
-algorithm known as the \emph{backward
-search}~\cite{ferragina2000opportunistic}.
+smaller memory footprint.
 
@@
 even the only practical option~\cite{pmid24336412, pmid25399029,
-pmid23349213, pmid19389736}, so we implicitly assume that seeding is
-performed by running the backward search. This will only have practical
-consequences in section~\ref{sec:est}, where we propose a method to
-estimate the parameters that are required to compute the probabilities of
-interest. For the rest, we simply assume that seeds are known at all times
-without ambiguity because this problem has several practical solutions.
+pmid23349213, pmid19389736}. However, the detail if of little interest
+here: we simply assume that seeds are known at all times without ambiguity
+because this problem has several practical solutions.
 
@@
 by an imperfect sequencer. The sequenced DNA fragment comes from a region
-of the genome that has three duplicates (there are four homologous
-sequences in total), and the read has three miscalled nucleotides
-(indicated by stars).
+of the genome that has three duplicates and the read has three miscalled
+nucleotides (indicated by stars).
 
@@
 sometimes match one of the duplicates. For instance, the first sequencing
-error matches all the duplicates and belongs to a seed for the first one.
+error matches all the duplicates and belongs to an off-target seed.
 However, erroneous nucleotides that are mismatches for \emph{all} the
-duplicates cannot belong to a seed. This is the case of the second
-sequencing error in this example, where there is a local deficit of
-seeds.
-
-Thie middle of the read shows some clutter where consecutive seeds
+sequences cannot belong to a seed. This is the case of the second
+sequencing error in this example, where there is a local deficit of seeds.
+
+The middle of the read shows some clutter where consecutive seeds
 match consecutive sequences at the same location. This is typical for
@@
 already in the candidate set. In addition, this seeding method is not
-particularly sensitive~\cite{pmid11934743} so it is used only for specific
-applications. Nevertheless, it will be extremely useful for the
-development of the present theory.
+particularly sensitive compared to spaced seed~\cite{pmid11934743} so it
+is used only in few specific applications. Nevertheless, it will be
+extremely useful for the development of the present theory.
 
@@
 Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
-start at every nucleotide. Instead, a fixed amount of nucleotides is
-skipped between every seed. This is a way to reduce the amount of
-overlapping matches at the same location, at the cost of sensitivity. This
-seeding heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds
-have size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
+start at every nucleotide. Instead, a certain amount of nucleotides is
+skipped between every seed. This is a way to reduce the overlapping
+matches at the same location, at the cost of sensitivity. This seeding
+heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds have
+size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
 skipped). We will refer to seeds where $n$ nucleotides are skipped as
@@
 Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
-replaced by skip-2 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
+replaced by skip-1 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
 Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
@@
 MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
-to define. Unlike exact seeds and skip seeds, their size is variable as we
-are about to see. They are used in BWA-MEM~\cite{li2013aligning} where
-they give good empirical results (but note that the actually seeding
-scheme uses more complex Super Maximal Exact Matches). To describe MEM
-seeds, let us first focus on MEMs.
+to define. Unlike exact seeds and skip seeds, their size is variable. They
+are used in BWA-MEM~\cite{li2013aligning} where they give good empirical
+results. To describe MEM seeds, let us first focus on MEMs.
 
@@
 
-A \emph{MEM seed} is simply a MEM of minimum size $\gamma$.
-Fig.~\ref{fig:MEM_example} shows what happens when exact seeds or skip
-seeds are replaced by MEM seeds on the read of
-Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the center
-of the read has disappeared because consecutive matches are fused into a
-single MEM seed.
+A \emph{MEM seed} is simply a MEM of size size $\gamma$ or greater.
+Fig.~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
+read of Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the
+center of the read has disappeared because consecutive matches are fused
+into a few MEM seeds.
 
@@
 
-Compared to seeds of fixed size, MEM seeds have two counter-intuitive (and
-undesirable) properties. The first is that there are cases where there
-cannot be any good seed, even when changing the sensitivity parameter
-$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
-though there is a single sequencing error, the read has no MEM seed for
-the target. Lowering $\gamma$ does not change this, so there is no way to
+Compared to seeds of fixed size, MEM seeds have two counter-intuitive
+properties. The first is that there are cases where there cannot be any
+on-target seed, even when changing the sensitivity parameter $\gamma$.
+Fig.~\ref{fig:full_masking_example} shows such an example. Even though
+there is a single sequencing error, the read has no MEM seed for the
+target. Lowering $\gamma$ does not change this, so there is no way to
 discover the true location in this case (even though it is the best
@@
 The read, the MEM seeds and the sequences are represented as in
-Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
-locations at the bottom effectively hide the target, so it cannot
-be discovered. This can occur even when the true location is the
-best and when there is a single error on the read.}
+Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
+at the bottom effectively hide the target, so it cannot be discovered.
+This can occur even when the true location is the best and when there is a
+single error on the read.}
 \label{fig:full_masking_example}
@@
 shows an example of this case. There is no MEM seed for the target, but
-there would be one if the read were two nucleotides shorter on the right
+there would be if the read were two nucleotides shorter on the right
 side. Indeed, in this case there would be a shared MEM seed matching the
-first and the second locations (provided $\gamma \leq 12$).
+the target and the first duplicate (provided $\gamma \leq 12$).
 
@@
 size. MEM seeds yield fewer candidates and therefore speed up the mapping
-process, but the question is at what cost for accuracy? The theory
-developed here will allow us to compute the probability that a read has
-no MEM seed for the target and thus that the true location is missed at
-the seeding stage.
+process, but the question is at what cost? The theory developed here will
+allow us to compute the probability that a read has no MEM seed for the
+target and thus that the true location is missed at the seeding stage.
 
@@
 that there are $N$ off-target sequences. We further assume that
-duplication happened instantaenously at some point in the past and that
+duplication happened instantaneously at some point in the past and that
 all $N+1$ sequences diverge independently of each other at a constant
@@
 assume that the target is the true sequence), and they match each
-off-target sequence with probability $\mu/3$. In contrast, correct
-nucleotides are always matches for the target, and they match each
-off-target sequence with probability $1-\mu$.
+duplicate with probability $\mu/3$. Correct nucleotides are always matches
+for the target, and they match each duplicate with probability $1-\mu$.
 
@@
 without loss of generality. We will explain in
-section~\ref{sec:short_MEMs} how to deal with this practical case and and
-how to identify those matches as off target. Until then, we will consider
-that the target and the duplicates are the only sequences in the reference
-genome.
+section~\ref{sec:random_seeds} how to deal with this practical case and
+and how to identify those matches as off target. Until then, we will
+consider that the target and the duplicates are the only sequences in the
+reference genome.
 
@@
 a probability of occurence equal to $p$. The weighted generating function
-of the word containing only this symbol is $pz$. The weight of the word is
+words containing only this symbol is $pz$. The weight of the word is
 its probability (here equal to $p$) and the size is its length in number
@@
 of the set $\mathcal{A}$ of reads that do not have on-target seeds
-(\textit{i.e.} reads for which seeding is either off target or null). The
+(\textit{i.e.} reads for which seeding is either null or off target). The
 weight of a read is its probability of occurrence and the size $k$ is its
@@
 The analytic combinatorics framework relies on a strategy referred to as
-the \emph{symbolice method}~\cite{sedgewick2013introduction}. The idea is
+the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
 to combine simple objects into more complex objects. Each combinatorial
@@
 
-The entry of $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$
-at coordinates $(i,j)$ is the weighted generating function of all the
-reads that end with a segment of type $j$ and that can be appended to a
-segment of type $i$. If segments of types $i$ and $j$ are exactly those
-that can be placed at the beginning and at the end of the reads, then this
-entry is the weighted generating function of all the reads without
-on-target seed. The examples below will clarify the key steps of this
+From the basic operations on weighted generating functions, the entry of
+$M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$ at
+coordinates $(i,j)$ is the weighted generating function of all the reads
+that end with a segment of type $j$ and that can be appended to a segment
+of type $i$. The examples below will clarify the key steps of this
 strategy.
 
-A complete description of the strategy to compute seeding probabilities
-with the symbolic method can be found in~\cite{filion2017analytic,
-filion2018analytic}. The interested readers can also read more about
+A complete description of how to compute seeding probabilities with the
+symbolic method can be found in~\cite{filion2017analytic,
+filion2018analytic}. The interested readers can also find more about
 analytic combinatorics in the popular
@@
 $\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
-appended after the last nucleotide to mark the end of the read.
-
-This recoding allows us to partition the read in \emph{segments} defined
-as a sequence of 0 or more symbols $\square$ followed by a symbol
-$\Downarrow$. The last part of the read is called the tail and it is
-terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
-not correspond to a nucleotide so it has size 0. Since the decomposition
-is unique, we can view a read as a sequence of segments with a tail,
-instead of a sequence of nucleotides.
-
-Fig.~\ref{fig:simple} shows an example of decomposition in segments.
-On-target seeds cannot contain sequencing errors, therefore they must be
-completely embedded inside a segment or inside the tail. So the size of
-the segments and of the tail indicates whether the read contains an
-on-target seed or not.
+appended after the last nucleotide to mark the end of the read. It is not
+associated to a nucleotide and therefore has size 0.
+
+This recoding allows us to partition the read in am important way.
+
+\begin{definition}
+\label{def:seg}
+Every symbol of the alphabet that is different from $\square$ is called a
+terminator. A segment is a sequence of 0 or more symbols $\square$
+followed by a terminator. The last segment of the read is terminated by
+the symbol $|$ and is called the tail.
+\end{definition}
+
+Since the decomposition of the read in segments is unique, we can view a
+read as a sequence of segments with a tail, instead of a sequence of
+nucleotides. Fig.~\ref{fig:simple} shows an example of decomposition in
+segments. On-target seeds cannot contain sequencing errors, therefore they
+must be completely embedded inside a segment. So the sizes of the segments
+indicate whether the read contains an on-target seed or not.
 
@@
 nucleotide) and the symbol $\square$ represents a match (it is a correct
-nucleotide). The symbolic sequence of the target is represented below,
-where an open square stands for a match and a closed square stands for a
-mismatch.}
+nucleotide). The symbol $|$ is appended to the end of the read. The
+symbolic sequence of the target is represented below, where an open square
+stands for a match and a closed square stands for a mismatch.}
 \label{fig:simple}
@@
 
-The probability of occurrence of the symbol $\Downarrow$ is $p$, the error
-rate of the sequencer, and the probability of occurrence of the symbol
+The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
+rate of the sequencer) and the probability of occurrence of the symbol
 $\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
@@
 The key insight is that reads without on-target seed are exactly the reads
-that are made of segments and tail with fewer than $\gamma$ symbols
-$\square$. The weighted generating function of such segments is $pz \cdot
-\big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
+that are made of segments with fewer than $\gamma$ symbols $\square$. The
+weighted generating function of such segments is $pz \cdot \big(
+1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
 $1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
@@
 
-Since there are two kinds of objects, the segments and the tails, the
-dimension of the transfer matrix is $2 \times 2$. A segment can be
-followed by another segment or by the tail. The tail cannot be followed by
-anything. The expression of the transfer matrix $M_0(z)$ is thus
+There are two kinds of objects: the $\Downarrow$ segments and the tails,
+so the dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$
+segment can be followed by another $\Downarrow$ segment or by the tail.
+The tail cannot be followed by anything. The expression of the transfer
+matrix $M_0(z)$ is thus
 \begin{equation*}
@@
 
-The types of objects are identified by their terminator, \textit{i.e.} by
-their last symbol. Segments are identified by the terminator $\Downarrow$
-and tails by the terminator $|$. The terminators are written in the
-margins of the transfer matrix for readability.
+In the representation above, the different types of objects are identified
+by their terminator. The terminators are written in the margins of the
+transfer matrix for readability.
 
@@
 term --- associated with terminators $\Downarrow$ and $|$. To see why,
-observe that every read can be prepended by a segment and only by a
-segment (not by a tail). Thus, reads are precisely the sequences of
+observe that every read can be prepended by $\Downarrow$ segments and only
+by those (not by a tail). Thus, reads are precisely the sequences of
 segments that can follow the symbol $\Downarrow$ and that are terminated
@@
 \end{equation*}
-Balancing the terms of same degree on both sides of the equality leads
+Balancing the terms of same degree on both sides of the equation leads
 to the following relationship between the coefficients of the series:
@@
 
-Here we go through another essential example, namely we devise a method to
+We go through another essential example, namely we devise a method to
 compute the probability that a read contains no on-target skip seed. Using
@@
 
-Once again, we need to know whether a nucleotide is a sequencing error,
-but this time we also need to know its phase in the repeated cycles of
-skipped positions. For this, we define the skip-$n$ alphabet
-$\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1, \ldots,
-\Downarrow_n \}$. As before, the symbol $\square$ represents a correct
-nucleotide and the symbol $|$ is a terminator appended to the end of the
-read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
-errors and $i$ indicates the number of nucleotides till the next
-non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
+As in the previous example, we need to know whether a nucleotide is a
+sequencing error, but this time we also need to know its phase in the
+repeated cycles of skipped positions. For this, we define the skip-$n$
+alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
+\ldots, \Downarrow_n \}$. The symbol $\square$ still represents a correct
+nucleotide and the symbol $|$ is still a terminator added at the end of
+the read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent
+sequencing errors and $i$ indicates the number of nucleotides till the
+next non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
 before a non-skipped position and $i=n$ for nucleotides at a non-skipped
@@
 
-The segments in this alphabet are defined as sequences of 0 or more
-symbols $\square$ followed by any of the symbols $\Downarrow_i$. The tail
-is again defined as a sequence of 0 or more symbols $\square$ followed by
-the symbol $|$. Given that this decomposition is unique, we can again view
-a read as a sequence of segments with a tail. The example of
-Fig.~\ref{fig:simple} is shown again in Fig.~\ref{fig:skip}, where
-segments in the mismatch alphabet have been replaced by segments in the
-skip-3 alphabet.
+As per definition~\ref{def:seg}, segments in this alphabet are sequences
+of 0 or more symbols $\square$ followed by any of the symbols
+$\Downarrow_i$ or by the symbol $|$. Given that this decomposition is
+unique, we can again view a read as a sequence of segments with a tail.
+The example of Fig.~\ref{fig:simple} is shown again in
+Fig.~\ref{fig:skip}, where segments in the mismatch alphabet have been
+replaced by segments in the skip-3 alphabet.
 
@@
 target (they are erroneous nucleotides) and the symbol $\square$
-represents a match (it is a correct nucleotide). The positions of the
-reads with a vertical bar indicate non-skipped positions (the potential
-start of a seed). The number $i$ in $\Downarrow_i$ indicates the distance
-to the next non-skipped position in number of nucleotides. For $i=0$ the
-next nucleotide is at a non-skipped position. Other graphical
-representations are as in Fig.~\ref{fig:simple}.}
+represents a match (it is a correct nucleotide). The vertical bars
+indicate non-skipped positions (the potential start of a seed). The number
+$i$ in $\Downarrow_i$ indicates the number of nucleotides till the next
+non-skipped position. For $i=0$ the next nucleotide is not skipped.
+Other features are as in Fig.~\ref{fig:simple}.}
 \label{fig:skip}
@@
 
-In the skip-$n$ alphabet, reads are composed segments terminated by the
+In the skip-$n$ alphabet, reads are composed of segments terminated by the
 $n+1$ symbols $\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n$, plus the
@@
 why, observe that at the start of every read, the next nucleotide is a
-non-skipped position, so every read can be prepended by segments
-terminated with $\Downarrow_0$ and only by those. Thus, reads are
-precisely the sequences of segments that can follow the symbol
-$\Downarrow_0$ and that are terminated by a tail, whose weighted
-generating function is the top right entry of the matrix.
-
-$M_n(z)$ is too complex to compute a close expression of
-$M_n(z)\cdot(I-M_n(z))^{-1}$. We return to the definition $M_n(z) +
-M_n(z)^2 + \ldots$ and we observe that the terms $M_n(z)^{k+2},
-M_n(z)^{k+3}, \ldots$ have no influence on the coefficients of the
-weighted generating function $a_0, a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k$ segments (plus one obligatory
-tail). Since $M(z)^{s+1}$ contains the weighted generating functions of
-reads with exactly $s$ segments and one tail, $a_k$ cannot depend on
-$M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by induction that
-all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
-contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
-+a_kz^k$ is strictly $0$.
+non-skipped position, so every read can be prepended by $\Downarrow_0$
+segments and only by those. Thus, reads are precisely the sequences of
+segments that can follow the symbol $\Downarrow_0$ and that are terminated
+by a tail, whose weighted generating function is the entry of the matrix
+associated with terminators $\Downarrow_0$ and $|$.
+
+$M_n(z)$ is too complex to compute a closed expression of $M_n(z) \cdot
+(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$
+and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no
+influence on the coefficients of the weighted generating function $a_0,
+a_1, \ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k+1$ segments (including the
+obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
+functions of reads with exactly $s+1$ segments including the tail, $a_k$
+cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by
+induction that all the entries of $M(z)^k$ are divisible by $z^{k-1}$,
+showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to
+$a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.
 
@@
 extract the entry of interest and then compute the terms of the Taylor
-expansion up to order $k$. But we can do better. Since we are only
+expansion up to order $k$. But we can do better: since we are only
 interested in the coefficients up to order $k$, we can perform all
-algebraic operations on truncated polynomials of order $k$, \textit{i.e.},
+algebraic operations on truncated polynomials of order $k$, \textit{i.e.}
 we discard the coefficients of order $k+1$ or greater when multiplying two
@@
 
-But we can do even better. A read with $s$ segments contains $s$ errors,
+But we can do even better: a read with $s+1$ segments contains $s$ errors,
 so all the entries of $M_n(z)^{s+1}$ are dominated by $p^s$ and they
@@
 Using the formula above, we can thus bound the probability that a read has
-$s$ segments or more. We compute the matrices $M_n(z) + M_n(z)^2 + \ldots
+$s+1$ segments or more. We compute $M_n(z) + M_n(z)^2 + \ldots
 +M_n(z)^{s+1}$ where the weighted generating functions have been replaced
-by truncated polynomials and we extract the entry at the first column and
-the last row. When the bound is lower than a set fraction $\varepsilon$ of
-the smallest coefficient, we stop the computations. Typically $\varepsilon
-= 0.01$ so this method ensures that the probabilities that a read of size
-$k$ has no on-target skip seed are accurate to within 1\%.
+by truncated polynomials and we extract the top right entry. When the
+bound is lower than a set fraction $\varepsilon$ of the current value of
+$a_k$, we stop the computations. Typically $\varepsilon = 0.01$ so this
+method ensures that the probabilities that a read of size $k$ has no
+on-target skip seed are accurate to within 1\%.
 
@@
 Observe that when $n=0$ the matrix $M_n(z)$ is identical to the matrix
-$M_0(z)$ of section~\ref{sec:example_exact}. This is consistent
-with the fact that exact seeds are skip-$n$ seeds for which $n=0$.
+$M_0(z)$ of section~\ref{sec:example_exact}. This is consistent with the
+fact that exact seeds are skip-$0$ seeds.
 \end{remark}
@@
 The event $\overline{S}_0 \cap \overline{S}_1$ is that the read contains
-no seed for either the target or the first duplicate --- numbering is
-arbitrary here, we simply chose a duplicate and stick to it.
-As illustrated in section~\ref{sec:symbolic}, we first recode the reads
-using a specialized alphabet to simplify the problem.
+no seed for either the target or for the first duplicate --- numbering is
+arbitrary here, we simply chose a duplicate and stick to it. As
+illustrated in section~\ref{sec:symbolic}, we first recode the reads using
+a specialized alphabet to simplify the problem.
 
@@
 Fig.~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
-and $-$ sequences are represented below the read, where the matches are
-represented by open squares and the mismatches by closed squares. It is
-visible from this example that symbols $\downarrow^-_{/i}$ and
-$\downarrow^+_{/i}$ alternate whenever the mismatches are on different
-sequences when going from left to right. The symbol $\Downarrow$ occurs
-only when a nucleotide is a double mismatch.
+and $-$ sequences are represented below the read, with matches represented
+as open squares and mismatches as closed squares. It is visible from this
+example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
+whenever the mismatches hit different sequences . The symbol $\Downarrow$
+occurs only when a nucleotide is a double mismatch.
 
@@
 \end{equation*}
-where $A(z)$, $B(z)$, $C(z)$ and $D(z)$ are matrices of dimensions
-$(\gamma-1) \times (\gamma-1)$ that are defined as
+where $\tilde{A}(z)$, $\tilde{B}_0(z)$, $\tilde{C}_0(z)$ and
+$\tilde{D}(z)$ are matrices of dimensions $(\gamma-1) \times (\gamma-1)$
+that are defined as
 \begin{equation*}
@@
 $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
-every read can be prepended by segments terminated by $\Downarrow$ and
-only by those (every other terminator would imply that one of the two
-sequences has a nonzero match size at the start of the read). Thus reads
-are precisely the sequences of segments that can follow the symbol
-$\Downarrow$ and that are terminated by a tail, whose weighted generating
-function is the top right entry of the matrix.
-
-It is possible to express $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$ in
-closed form, but it involves polynomials of very large degree. It is
-actually easier to proceed as in section~\ref{sec:example_skip} and to
-compute the powers of $\tilde{M}_0(z)$ (using the arithmetic of truncated
-polynomials) up to a given threshold value. Since each segment contains a
+every read can be prepended by $\Downarrow$ segments and only by those
+(every other terminator would imply that one of the two sequences has a
+nonzero match size at the start of the read). Thus reads are precisely the
+sequences of segments that can follow the symbol $\Downarrow$ and that are
+terminated by a tail, whose weighted generating function is the top right
+entry of the matrix.
+
+$\tilde{M}_0(z)$ is too complex to compute a closed expression of
+$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
+proceed as in section~\ref{sec:example_skip} and to compute the powers of
+$\tilde{M}_0(z)$ (using the arithmetic of truncated polynomials) up to a
+given threshold value. Since each segment except the tail contains a
 mismatch against at least one sequence, we define $\tilde{p}$ as the upper
@@
 With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
-compute its powers until the upper bound above is lower than a set
-fraction $\varepsilon$ of the current value, which gives an accurate
+compute its powers until the upper bound is lower than a set fraction
+$\varepsilon$ of the current value of $a_k$, which gives an accurate
 estimate of $P(\overline{S}_0 \cap \overline{S}_1)$.
 
-\subsection{Corollary: null seeding}
-
+\begin{remark}
 Observe that expression (\ref{eq:Pnull}) is the probability of null
@@
 not mappable, especially when reads are relatively short.
+\end{remark}
 
@@
 reads of size $k=50$ sequenced with an instrument with error rate
-$p=0.01$, and se use exact seeds of size $\gamma=17$. We then vary the
+$p=0.01$, and we use exact seeds of size $\gamma=17$. We then vary the
 number of duplicates $N$ and their divergence rate $\mu$, which both
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.65]{curves_exact.pdf}
+\caption{\textbf{Off target seeding probabilities (exact seeds).}
+The curves show the probability that seeding is off target for exact seeds
+of size $\gamma=17$ in reads of size $k=50$ nucleotides sequenced with an
+error rate $p=0.01$. Each line shows a different number of duplicate
+sequences $N$ from 1 to 10 and the x-axis shows the divergence rate $\mu$,
+defined as the probability that a given duplicate differs from the target
+at any nucleotide.}
+\label{fig:curves_exact}
+\end{figure}
+
+Fig.~\ref{fig:curves_exact} shows the result for $N$ from 1 to 10 and for
+$\mu$ from 0 to 0.20. The first observation is that the probability that
+seeding is off target increases with $N$. This is clearly visible from
+expression (\ref{eq:Poff}). This can also be understood intuitively
+because the probability of null seeding decreases when there are more
+potential candidates, and since the probability of on-target seeding does
+not change, the probability of off-target seeding must increase.
+
+The second observation is that there exists a ``worst'' value of $\mu$
+situated around 0.08. When $\mu$ is much smaller, the sequences of the
+duplicates are close to that of the target so it is unlikely that the
+candidate set contains one but not the other. When $\mu$ is much larger,
+the sequences of the duplicates are far from that of the target and they
+are unlikely to be in the candidate set. In expression (\ref{eq:Poff}),
+the only term that depends on $\mu$ is $P(\overline{S}_0 \cap
+\overline{S}_1)$, and it is obvious that the minimum of expression
+(\ref{eq:Poff}) corresponds to the maximum of $P(\overline{S}_0 \cap
+\overline{S}_1)$. This is why the worst value of $\mu$ is the same for
+every $N$.
+
+For comparison purposes, it is important to notice the order of magnitude
+of the probabilities. Up to $N=3$ duplicates, off-target seeding has
+probability lower than $10^{-4}$ and for $N=10$ duplicates the upper bound
+is $3\cdot10^{-4}$. This sets a baseline to compare with more elaborate
+seeding methods.
+
 \section{Off-target skip seeds}
@@
 using skip seeds, we observe that the logic of section~\ref{sec:offexact}
-can be transposed with very few modifications. In particular, the
-probability can be computed through expression (\ref{eq:Poff}), where
-$S_0$ is the event that the read has a \emph{skip} seed for the target and
-$S_1$ that it has a \emph{skip} seed for the first duplicate.
+can be transposed with few modifications. In particular, the probability
+can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
+event that the read has a \emph{skip} seed for the target and $S_1$ that
+it has a \emph{skip} seed for the first duplicate.
 
@@
 the left side of the read, where the previous terminator is 6 nucleotides
-upstream. When the symbols alternate, $i$ is the distance to the previous
-non-skipped position.
+upstream. In this case, $i$ is the distance to the previous non-skipped
+position.
 
@@
 \caption{\textbf{Example skip dual encoding.}
-The read of Fig.~\ref{fig:dual} is represented in the dual alphabet. The
-vertical bars above the read indicate non-skipped positions. The symbols
-$\downarrow_{/i}^-$ and $\downarrow_{/i}^+$ have the same meaning as in
-the dual alphabet. The symbols $\Downarrow_i$ indicate that both sequences
-have match length 0 and that the next non-skipped position is located $i$
-nucleotides downstream. Other graphical representations are as in
+The read of Fig.~\ref{fig:dual} is represented in the skip dual alphabet.
+The vertical bars above the read indicate non-skipped positions. The
+symbols $\downarrow_{/i}^-$ and $\downarrow_{/i}^+$ have the same meaning
+as in the dual alphabet. The symbols $\Downarrow_i$ indicate that both
+sequences have match length 0 and that the next non-skipped position is
+located $i$ nucleotides downstream. Other features are as in
 Fig.~\ref{fig:dual}.}
@@
 matrix $\tilde{M}_n(z)$ is
+
+\pagebreak
 
@@
 \end{equation*}
+
+\pagebreak
 
@@
 \begin{remark}
-Observe that when $n=0$ the matrix $\tilde{M}_n(z)$, is identical to the
+Observe that when $n=0$ the matrix $\tilde{M}_n(z)$ is identical to the
 matrix $\tilde{M}_0(z)$ of section~\ref{sec:dual}, again consistent with
-the fact that exact seeds are skip-$n$ seeds for which $n=0$. The same
-applies to $\tilde{B}_n(z)$ and $\tilde{C}_n(z)$.
+the fact that exact seeds are skip-$0$ seeds. The same applies to
+$\tilde{B}_n(z)$ and $\tilde{C}_n(z)$.
 \end{remark}
@@
 scheme using skip seeds is off target. Using the same settings as in
-section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), we set
-decide to use skip-9 seeds. We then vary the number of duplicates $N$ and
-their divergence rate $\mu$.
+section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), we decide to
+use skip-9 seeds and vary the number of duplicates $N$ and their
+divergence rate $\mu$.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.65]{curves_skip.pdf}
+\caption{\textbf{Off target seeding probabilities (skip seeds).}
+The curves show the probability that seeding is off target for skip-9
+seeds of size $\gamma=17$. Read size and erorr rate are the same as in
+Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
+shows a different number of duplicate sequences $N$ from 1 to 10 and the
+x-axis shows the divergence rate $\mu$, defined as the probability that a
+given duplicate differs from the target at any nucleotide.}
+\label{fig:curves_skip}
+\end{figure}
+
+Fig.~\ref{fig:curves_skip} shows the result for $N$ from 1 to 10 and for
+$\mu$ from 0 to 0.20. The curves have the same general aspect as those of
+Fig.~\ref{fig:curves_exact}. The probability that seeding is off target
+increases with $N$. There is again a  worst value of $\mu$, because the
+maximum of $P(\overline{S}_0 \cap \overline{S}_1)$ minimizes expression
+(\ref{eq:Poff}) for every value of $N$. But the value is different from
+that of Fig.~\ref{fig:curves_exact} because $P(\overline{S}_0 \cap
+\overline{S}_1)$ is computed for skip-9 seeds instead of exact seeds (here
+the worst value of $\mu$ is approximately 0.07).
+
+Importantly, Fig.~\ref{fig:curves_skip} reveals that skipping 9
+nucleotides increases the chances that seeding is off target by a factor
+approximately 4 (compare with Fig.~\ref{fig:curves_exact}). It is not
+obvious why this should be the case: skip seeds reduce the probability of
+on-target seeding (skipping positions implies fewer on-target seeds) but
+increase the probability of null seeding (skipping positions implies fewer
+seeds overall). So the net effect on off-target seeding is not intuitive.
+
+And indeed, there are cases that skip seeds are \emph{less} prone to
+off-target seeding than exact seeds. This is the case for instance when
+$p=0.1$, where the ranking is inverted for the whole range of values
+considered in Fig.~\ref{fig:curves_skip}.
+
+This information is critical for choosing the best seeding strategy. The
+off-target seeding probability is not the only criterion, though. Equally
+important considerations are the probability of on-target seeding and the
+computational resources required to implement a particular seeding
+strategy. The benefit of a theory to compute seeding probabilities is to
+have access to this knowledge.
 
@@
 
-MEM seeds are substantially more complex than exact seeds and skip seeds.
-As we will see below, the duplicates can prevent the existence of an
-on-target MEM seed, so we need to take them into account in the
+MEM seeds are substantially more complex than exact seeds and skip seeds
+because we need to take into account all the duplicates in the
 combinatorial construction.
@@
 match length is strictly longer than the match length of the target. A
-duplicate is a \emph{soft mask} if its has the same match length as the
+duplicate is a \emph{soft mask} if it has the same match length as the
 target.
@@
 depend on the position of interest: a sequence can be a mask at the left
-end of the read but not at the right end, or the opposite.
+end of the read and not at the right end, or the opposite.
 
@@
 \caption{\textbf{Example of hard and soft masks.}
-Genomic sequencess are shown below a read where open squres represent
+Genomic sequences are shown below a read where open squares represent
 nucleotides. In the sequences, the open squares represent matches and the
@@
 
-Hard and soft masks explain the counter-intuitive of MEM seeds. For
-instance, in Fig.~\ref{fig:full_masking_example} the target cannot be
+Hard and soft masks explain the counter-intuitive properties of MEM seeds.
+For instance, in Fig.~\ref{fig:full_masking_example} the target cannot be
 discovered because every nucleotide of the read has a hard mask. In
@@
 on-target MEM seed is always unmasked. Conversely, an unmasked nucleotide
-always belongs to exactly one strict on-target MEM seed. Also, the last
+always belongs to exactly one strict on-target MEM (not necessarily a seed
+because the size of the MEM can be less than $\gamma$). Also, the last
 nucleotide of every shared on-target MEM seed is always soft-masked, but a
-soft-masked nucleotide does not always belong to a shared on-target MEM
-seed.
+soft-masked nucleotide does not always belong to a shared on-target MEM.
 
@@
 mismatch against \emph{every} sequence. The symbols $\uparrow_{/i}$
-indicate a change in masking status: the nucleotide is not masked but its
-left neighbor is. The index $i \geq 1$ is the number of nucleotides since
-the last mismatch or since the beginning of the read. All the other
-nucleotides are represented by the $\square$ symbol and the $|$ symbol is
+indicate a change in masking status: the nucleotide is not masked but the
+previous is. The index $i \geq 1$ is the number of nucleotides since the
+last mismatch or since the beginning of the read. All the other
+nucleotides are represented by the symbol $\square$ and the symbol $|$ is
 appended to the end of the read as usual.
@@
 \caption{\textbf{The MEM encoding}.
-The read of Fig.~\ref{fig:exact_seed_example} is represented in the MEM
-alphabet. The arrows departing from the numbers help understand their
-meaning. The symbol $\downarrow_{/m}$ is indexed by the number $m$ of
-locations that match the nucleotide. The symbol $\uparrow_{/i}$ is
-indexed by the number $i$ of nucleotides from the last error or from the
-beginning of the read.}
+The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
+The arrows departing from the numbers help understand their meaning. The
+symbol $\downarrow_{/m}$ is indexed by the number $m$ of locations that
+match the nucleotide. The symbol $\uparrow_{/i}$ is indexed by the number
+$i$ of nucleotides from the last error or from the beginning of the read.
+The grey squares in the symbolic sequences represent MEM seed matches.
+Other features are as in Fig.~\ref{fig:dual}.}
 \label{fig:sketch_extended}
@@
 $N-m$ soft masks. The symbols $\uparrow_{/i}$ indicate that the nucleotide
-is unmasked and that the previous $i-1$ nucleotides were masked.
+is unmasked and that the previous nucleotide is masked.
 
@@
 (marked by $\uparrow_{/i}$), and it cannot be extended because it is
-flanked by sequencing errors or by the ends of the reads. Note that there
-is exactly one symbol $\uparrow_{/i}$ per strict on-target MEM seed, and
-therefore two symbols $\uparrow_{/i}$ must be separated by at least one
-symbol $\downarrow_{/m}$.
+flanked by sequencing errors (symbols $\downarrow_{/m}$) or by the ends of
+the reads. Note that there is exactly one symbol $\uparrow_{/i}$ per
+strict on-target MEM seed, and therefore two symbols $\uparrow_{/i}$ must
+be separated by at least one symbol $\downarrow_{/m}$.
 
@@
 
-A segment terminated by $\uparrow_{/i}$ is the beginning of an on-target
-MEM seed of size at least $i$. The seed reaches the next sequencing
-error or the end of the read, so the number of $\square$ symbols in the
+A segment terminated by $\uparrow_{/i}$ is the beginning of a strict
+on-target MEM of size at least $i$. The MEM reaches the next sequencing
+error or the end of the read, so the number of symbols $\square$ in the
 next segment must be at most $\gamma-i-1$ and it must be terminated by a
-$\downarrow$ symbol or by the tail terminator $|$.
+$\downarrow_{/m}$ symbol or by the tail terminator $|$.
 
@@
 
-Indeed, if the nucleotide is an error, the symbol is $\downarrow_{/m}$ for
-some $m$ between 0 and $N$. Each duplicate is a match with probability
-$\mu/3$, so $m$ has a Binomial distribution with parameters $(N, \mu/3)$.
+The definition of $\omega_m$ is correct: if the nucleotide is an error,
+the symbol is $\downarrow_{/m}$ for some $m$ between 0 and $N$. Each
+duplicate is a match with probability $\mu/3$, so $m$ has a Binomial
+distribution with parameters $(N, \mu/3)$.
 
@@
 simply $qz$ (recall that $p = 1-q$ is the probability of a sequencing
-error). The weighted generating function of the $\downarrow_{/m}$
-terminator is $\omega_m pz$, so the weighted generating function of the
-segments terminated by $\downarrow_{/m}$ following a segment terminated by
-$\uparrow_{/i}$ is
+error). The weighted generating function of the terminator
+$\downarrow_{/m}$ is $\omega_m pz$, so the weighted generating function of
+the $\downarrow_{/m}$ segments following $\uparrow_{/i}$ is
 \begin{equation}
@@
 And the weighted generating function of the tail segments following
-segments terminated by $\uparrow_{/i}$ is
+$\uparrow_{/i}$ is
 \begin{equation}
@@
 The symbol $\downarrow_{/m}$ signifies that the nucleotide has $m$ hard
-masks and $N-m$ soft masks. If all the masks disappear before the first
-read error, the next terminator will be one of the $\uparrow_{/j}$
-symbols, otherwise it will be one of the $\downarrow_{/m}$ symbols. We
-separate the cases based on the terminator of the segment.
+masks and $N-m$ soft masks. If all the masks vanish before the first read
+error, the next terminator will be a sybmol $\uparrow_{/j}$, otherwise it
+will be the sybmol $|$ or a symbol $\downarrow_{/m}$. We separate the
+cases based on the terminator of the segment.
 
@@
 \begin{definition}
-The probability that a given duplicate sequence contains a mismatch in a
-sequence of $j$ error-free nucleotides is
+The probability that a given duplicate contains a mismatch in a sequence
+of $j$ error-free nucleotides is
 \begin{equation}
@@
 
-Indeed, every nucleotide of the duplicate differs from the target
-with probability $\mu$. In the absence of sequencing error, this is also
-the probability that a nucleotide of the duplicate differs from the read.
-The probability that $j$ nucleotides in a row are identical to the read is
+The definition of $\xi_j$ is correct: every nucleotide of the duplicate
+differs from the target with probability $\mu$. In the absence of
+sequencing error, this is also the probability that a nucleotide of the
+duplicate differs from the read. Given that there is no error, the
+probability that $j$ nucleotides in a row are identical to the read is
 thus $(1-\mu)^j$ and the probability that at least one of them is
@@
 
-With this notation, the probability that at least one mask survives a
-sequence of $j$ error-free nucleotides is thus $1-\xi_j^N$, and the
-probability that there remains a mask at the $j-1$-th but not at the 
+With this notation, the probability that at least one of $N$ masks
+survives a sequence of $j$ error-free nucleotides is thus $1-\xi_j^N$, and
+the probability that there remains a mask at the $j-1$-th but not at the
 $j$-th error-free nucleotide is $\xi_j^N - \xi_{j-1}^N$. From this we
@@
 The fact that the reads have no on-target seeds imposes $j < \gamma$. Also
-note that this expression is the same for all $\downarrow_{/m}$ symbols
+note that this expression is the same for all symbols $\downarrow_{/m}$
 (it does not depend on $m$).
@@
 masks survives until the end of the segment in order to exclude the
-$\uparrow_{/j}$ symbols. The weighted generating function is
+symbols $\uparrow_{/j}$. The weighted generating function is
 \begin{equation*}
@@
 Summing the expressions from cases 2a and 2b, we find that the
-weighted generating function of the tail following segments terminated by
-$\downarrow_{/m}$ is
+weighted generating function of the tail following $\downarrow_{/m}$ is
 \begin{equation}
@@
 In this case, there can be no on-target seed and we must only exclude the
-$\uparrow_{/j}$ terminators. As we have seen above, this implies that at
+terminators $\uparrow_{/j}$. As we have seen above, this implies that at
 least one of the $N$ masks survives until the terminator. For a read of
@@
 terminator and summing over $j+1 \leq \gamma$, we see that the weighted
-generating function is thus
+generating function is
 \begin{equation}
@@
 graphically represented in the diagram below. The left panel corresponds
-to case $i)$ and the panel to case $ii)$. The top row represents the
-target, and the bottom rows represent duplicates (using the same
-symbols as in Fig.~\ref{fig:sketch_extended}).
+to case $i)$ and the right panel to case $ii)$. The top row represents the
+target, and the bottom rows represent duplicates (using the same symbols
+as in Fig.~\ref{fig:sketch_extended}).
 \begin{inset}
@@
 Whenever a hard mask (here the first duplicate) covers the nucleotides as
-shown in the left panel, there can be no on-target seed. The positions
-marked with a question mark are irrelevant, they cannot change the fact
-that there is no on-target MEM seed. If the hard masks vanish, as in the
-right panel, then we need to look at the soft masks. If a soft mask covers
-the whole segment, then there can be no on-target seed. In all other cases
-there is an on-target MEM seed.
+shown by the grey squres in the left panel, there can be no on-target
+seed. The positions marked with a question mark are irrelevant, they
+cannot change the fact that there is no on-target MEM seed. If the hard
+masks vanish, as in the right panel, then we need to look at the soft
+masks. If a soft mask covers the whole segment as indicated by the grey
+sqaures, then there can be no on-target seed. In all other cases there is
+an on-target MEM seed.
 
@@
 \xi_j^m \big)$. Summing over $j+1 > \gamma$ and including the terminator,
-we see that the associated weighted generating function is thus
+we see that the associated weighted generating function is
 \begin{equation*}
@@
 
-Indeed, the probability that there are $j$ matches between the duplicate
-and the target is $(1-\mu)^j$. If there are no sequencing errors, this is
-also the probability that there are $j$ matches between the duplicate and
-the read. The probability that the duplicate matches the subsequent
-sequencing error is $\mu/3$, so the probability of that there are $j+1$
-matches including the sequencing error is $(1-\mu)^j\mu/3$. Finally, the
-probability that there is a mismatch is the complement $1-(1-\mu)^j\mu/3$.
+The definition of $\eta_j$ is correct: the probability that there are $j$
+matches between the duplicate and the target is $(1-\mu)^j$. If there are
+no sequencing errors, this is also the probability that there are $j$
+matches between the duplicate and the read. The probability that the
+duplicate matches the subsequent error is $\mu/3$, so the probability of
+that there are $j+1$ matches including the sequencing error is
+$(1-\mu)^j\mu/3$. Finally, the probability that there is a mismatch is the
+complement $1-(1-\mu)^j\mu/3$.
 
@@
 terminator, so the total number of matches is $r$ plus the number of
-sequences that also match the last nucleotide, among the remaining $N-m-r$
-soft masks and the $m$ hard masks.
+sequences that also match the terminator, among the remaining $N-m-r$ soft
+masks and the $m$ hard masks.
 
 Let us start with the $m$ hard masks. The probability that each of them
-matches the terminator is simply $\mu/3$, and the probably that they do
-not is $1-\mu/3$.
+matches the terminator is simply $\mu/3$.
 
@@
 Summing the terms from case 3a and from case 3b, we find that the weighted
-generating function of the segments terminated by $\downarrow_{/n}$
-following segments terminated by $\downarrow_{/m}$ is
+generating function of the $\downarrow_{/n}$ segments following
+$\downarrow_{/m}$ is
 \begin{equation}
@@
 seeds. The term of interest, as usual, is the top right entry. Indeed,
-every read can be prepended by segments terminated by $\downarrow_{/0}$
-and only by those, otherwise the read would start with fewer than $N$ soft
-masks. Thus, reads without an on-target MEM seed are precisely the
-sequences of segments that can be appended to the symbol $\downarrow_{/0}$
-and that are terimated by a tail.
+every read can be prepended by $\downarrow_{/0}$ segments and only by
+those, otherwise the read would start with fewer than $N$ soft masks.
+Thus, reads without an on-target MEM seed are precisely the sequences of
+segments that can be appended to the symbol $\downarrow_{/0}$ and that are
+terimated by a tail.
 
@@
 \end{equation*}
-but here not every segment contains an error. However, there cannot be two
-symbols $\uparrow_{/j}$ in a row, so a read with $s$ segments so a tail
-must contain a minimum number of sequencing errors which is
-$e=\lfloor{s/2\rfloor}$. So we compute the powers of $\mathring{M}_N(z)$
-until the upper bound is less than a set fraction $\varepsilon$ of the
-current value.
+but here not every segment contains an error. There cannot be two symbols
+$\uparrow_{/j}$ in a row, so a read with $s+1$ segments must contain a
+minimum number of sequencing errors which is $e=\lfloor{s/2\rfloor}$. As
+before, we compute the powers of $\mathring{M}_N(z)$ until the upper bound
+is less than a set fraction $\varepsilon$ of the current value of $a_k$.
 
@@
 is
-\begin{equation*}
+\begin{equation}
+\label{eq:PMEM}
 P(\overline{M}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
 \cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N,
-\end{equation*}
+\end{equation}
 where $P(\overline{M}_0)$ is computed using $\mathring{M}_N(z)$ as
@@
 matrix $\mathring{M}_N(z)$ has dimension $(N+\gamma+1) \times
-(N+\gamma+1)$. So for MEM seeds the problem can become computationally
-intractable because $N$ can be very large. For instance, the sequences
-called \emph{Alu} have more than one million duplicates in the human
-genome. There is no hope to compute the powers of $\mathring{M}_N(z)$ in
-these conditions and we need an alternative method.
-
-The symbolic representation as MEM segments can be used for an efficient
-method to sample reads. Instead of generating the nucleotides of the $N+1$
-sequences one by one, one can generate a single sequence of segments.
-Since the number of segment does not depend on $N$, we can obtain a fast
-Monte Carlo method to sample millions of reads and count the proportion
-that contain an on-target MEM seed.
+(N+\gamma+1)$. The problem can become computationally intractable because
+$N$ can be very large. For instance, the sequences called \emph{Alu} have
+more than one million duplicates in the human genome. There is no hope to
+compute the powers of $\mathring{M}_N(z)$ in these conditions and we need
+an alternative method.
+
+The symbolic representation as MEM segments can be used to design an
+efficient method to sample reads. Instead of generating the nucleotides of
+the $N+1$ sequences one by one, one can generate a single sequence of
+segments. Since the number of segment does not depend on $N$, we can
+obtain a fast Monte Carlo method to sample millions of reads and count the
+proportion that contain an on-target MEM seed.
 
@@
 The method is summarized in algorithm~\ref{alg:mcmc} below. It requires
-efficient methods to sample from the geometric and from the binomial
+efficient algorithms to sample from the geometric and from the binomial
 distributions. Sampling from a geometric distribution can be done by
@@
 \subsection{Illustration}
-
-Here goes.
+\label{sec:illmem}
+
+Using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
+and $\gamma=17$), we decide to use MEM seeds and vary the number of
+duplicates $N$ and their divergence rate $\mu$.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.65]{curves_mem.pdf}
+\caption{\textbf{Off target seeding probabilities (MEM seeds).}
+The curves show the probability that seeding is off target for skip-9
+seeds of size $\gamma=17$. Read size and erorr rate are the same as in
+Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
+shows a different number of duplicate sequences $N$ from 1 to 10 and the
+x-axis shows the divergence rate $\mu$, defined as the probability that a
+given duplicate differs from the target at any nucleotide.}
+\label{fig:curves_mem}
+\end{figure}
+
+Fig.~\ref{fig:curves_mem} shows the result for $N$ from 1 to 10 and for
+$\mu$ from 0 to 0.20. The curves have the same general aspect as those of
+Fig.~\ref{fig:curves_exact}. The probability that seeding is off target
+increases with $N$, as shown by expression (\ref{eq:PMEM}).
+
+There is again a  worst value of $\mu$ but it is much lower than the
+previous two cases (here it is close to 0.026). In the case of MEM seeds,
+it is not obvious why the same value of $\mu$ maximizes (\ref{eq:PMEM})
+for all values of $N$ because both $P(\overline{M}_0)$ and
+$P(\overline{S}_0 \cap \overline{S}_1)$ depend on $\mu$.
+
+Fig.~\ref{fig:curves_mem} reveals that in this concrete case, MEM seeds
+increases the chances that seeding is off target by a factor up to 40
+compared to exact seeds (see Fig.~\ref{fig:curves_exact}). On this
+criterion, MEM seeds are always inferior to exact seeds with the same
+specifications. The reason is that every read triggering an off-target
+seeding with exact seeds will also trigger an off-target seeding with MEM
+seeds. Knowing that, it is still important to realize how large the
+increase is in practice. MEM seeds tremendously simplify the seeding
+process, but this comes at the cost of an increase in the false positive
+rate. The methods presented here are thus very useful to balance the
+risk when using MEM seeds.
 
 \section{Random seeds}
-\label{sec:short_MEMs}
+\label{sec:random_seeds}
 
 We have assumed throughout that seeds can match only the target or one of
-its duplicates. In practice, seeds that overlap a sequencing error often
-match random locations of the genome instead of the duplicates.
+its duplicates. In practice, seeds can match many random locations of the
+genome and not just the target and the duplicates.
 
@@
 have a seed are considered candidate locations (as long as the seed is
-above the threshold $\gamma$ of course). If not, the estimates of the
-mismapping rate are off. The hope is that spurious matches are shorter
-than $\gamma$ so that they are invisible, but this is not always the case.
-And if they qualify as a seed we need to verify the candidate location.
+above the threshold $\gamma$). If not, the estimates of the mismapping
+rate are off. The hope is that spurious matches are shorter than $\gamma$
+so that they are invisible, but this is not always the case. And if they
+qualify as a seed we need to verify the candidate location.
 
@@
 that the best hit is likely to be discovered first, allowing us to bail
-out from other alignments as early as possible. We can even bound
-the maximum quality of some candidates so that the alignment can be
-skipped altogether. For instance, if we already have a candidate at
-distance 1 from the read, there is no need to verify candidates from MEM
-seeds than span from neither end of the read. Indeed, the hit must be at
-distance at least 2 because there is a mismatch before the MEM seed and
-another one after it. These considerations depend on the implementation of
-the mapper so we will not delve deeper here. What matters is that random
+out from other alignments as early as possible. We can even bound the
+maximum quality of some candidates so that the alignment can be skipped
+altogether. These considerations depend on the implementation of the
+mapper so we will not develop this further. What matters is that random
 hits do not have to impose a significant burden on the time needed to
@@
 target nor for its duplicates). Since the hits are not homologous to the
-read, the alignment score is essentially random, which makes it is easy to
-detect.
+read, the alignment score is random, which makes it is easy to detect.
 
@@
 $\beta$, one can choose for instance $\beta=1/2$ so that the term
-$(1-/\beta)/\beta$ disappears in expression (\ref{eq:bayes}).
+$(1-/\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
 
 Since the probability that the best candidate is a random sequence is
-either very small or very large, it either has no influence or dominates
-the probability of a false positive. For every read, we can use as
-confidence score the maximum of this probability of expression
-(\ref{eq:bayes}) and of the estimated probability that seeding is off
-target.
+either very small or very large, it either has no influence or it
+dominates the probability of a false positive. For every read, we can use
+as confidence score the maximum of this probability and of the estimated
+probability that seeding is off target.
 
@@
 We have devised a set of methods to compute the probability that seeding
-heuristics fail and sets the mapping process for a false positive. This
+heuristics fail and commits the mapping process to an error. This
 fills a knowledge gap to understand and calibrate the performance of the
@@
 weighted generating function, we compute only the first terms using
-iterative methods as in explained in section~\ref{sec:example_skip}, 
-or using Monte Carlo sampling as explained in
-section~\ref{sec:montecarlo}. In this regard, the breakthrough is the
-encoding of reads as MEM segments, which is a form of implicit Markov
+iterative methods as explained in section~\ref{sec:example_skip}, or using
+Monte Carlo sampling as explained in section~\ref{sec:montecarlo}. In this
+regard, the breakthrough is the encoding of reads as segments in different
+alphabets, which is an implicit form of Markov
 imbedding~\cite{fu1994distribution}.
+
+The methods rely on the knowledge of two essential parameters: the number
+of duplicates $N$ and their divergence rate $\mu$. These quantities can be
+estimated efficiently using the FM-index~\cite{ferragina2005indexing} and
+we will publish the detail of this method elsewhere, together with a more
+complete overview of how to use the results presented here to implement
+practical mapping algorithms.
 
@@
 occur among duplicated sequences. This is obviously incorrect, but our
-initial tests with real data suggest that this is a minor impediment. On
-real genomes, the estimations are only approximately correct and this is
-but one of several simplifying assumptions. Incorporating insertions and
-deletions would make the theory intractable, so it is presently unclear
-how to deal with this type of error.
+initial tests with real data suggest that this is a minor impediment (most
+likely because this is only one of several simplififying assumptions).
+Incorporating insertions and deletions would make the theory intractable,
+so it is presently unclear how to deal with this type of error.
 
@@
 again a minor impediment, since we can correct the probability that the
-location is a false positive when only some candidates have been tested.
+location is false when only some candidates have been tested.
 
@@
 with this possibility \textit{a posteriori} in
-section~\ref{sec:short_MEMs}.
-
-The Monte Carlo approach of algorithm~\ref{alg:mcmc} is relatively
-straightforward, so one may wonder why the approach with weighted
-generating functions is necessary for MEM seeds. The only reason is
-precision. To estimate the frequency of an event by Monte Carlo sampling,
-it must occur at least a few times in the simulation. For instance, with
-10 million rounds of sampling, frequencies around $1/1,000,000$ or lower
-cannot be measured accurately. When not interested in rare events, it is
-thus a perfectly reasonable strategy.
+section~\ref{sec:random_seeds}.
+
+Being able to compute seeding probabilities revealed some interesting
+facts (sections~\ref{sec:illdual}, \ref{sec:illskipdual} and
+\ref{sec:illmem}). The first is that the seeding schemes considered here
+have a worst case scenario for a particular value of the divergence
+between duplicates $\mu$. Importantly, the worst value varies between
+different seeding methods, so it is possible in theory to construct
+opportunistic seeding strategies that pick the best method for every read.
+
+Another interesting fact is that skip seeds can have better performance
+than exact seeds in the sense that they can lower off-seeding
+probabilities. However, this always comes at the cost of accuracy because
+skipping nulceotides reduces the probability of on-target seeding.
+
+Finally, we also observed that MEM seeds cause a very serious increase in
+off-target seeding compared to exact seeds and skip seeds. This does not
+mean that MEM seeding is a bad strategy (it is usually faster than the
+other methods), but it is a good idea to keep the performance in check and
+switch method or even skip the read altogether if the chances of
+discovering the target are too low.
+
+Regarding the methodology, the Monte Carlo approach of
+algorithm~\ref{alg:mcmc} is relatively straightforward, so one may wonder
+why the approach with weighted generating functions is necessary for MEM
+seeds. The only reason is precision. To estimate the frequency of an event
+by Monte Carlo sampling, it must occur at least a few times in the
+simulation. For instance, with 10 million rounds of sampling, frequencies
+around $1/1,000,000$ or lower cannot be measured accurately. When
+interested only in frequent events, it is thus a perfectly reasonable
+strategy.
 
@@
 relatively small, so we need a method that is accurate in this range.
-Fortunately, the transfer matrix method is fast because the dimensions of
-the matrix $\mathring{M}_N(z)$ are not prohibitive.
+Fortunately, the transfer matrix method is fast because the dimension of
+the matrix $\mathring{M}_N(z)$ is small and the computations are not
+prohibitive.
 
@@
 $p$ is known and constant). For $N > 20$, the iterative method is usually
-prohibitive and we need to use Monte Carlo sampling instead. The running
+too slow and we need to use Monte Carlo sampling instead. The running
 time depends on $p$, on the size of the reads $k$, and on the desired
@@
 of size $100$-$200$ nucleotides on modern hardware). The values of $N$ and
-$\mu$ can be binned in intervals of interest so that there are only around
-$100$ pairs for a total cost of a few minutes per run. Considering that
-mappers can process at most $10,000$ reads per second per core, the time
-of mapping a sequencing run of 250 million reads is over 7 hours, two
-orders of magnitude larger than the time required to estimate the
-probabilities of error.
+$\mu$ can be binned in intervals so that there are only around $100$ pairs
+for a total cost of a few minutes per run. Considering that mappers can
+process at most $10,000$ reads per second per core, the time of mapping a
+sequencing run of 250 million reads is over 7 hours per core, two orders
+of magnitude larger than the time required to estimate the probabilities
+of error.
 
@@
 (\textit{e.g.}, modern human and Neanderthal) in order to assign the reads
-to the organism they belong to. The probabilities of false positives give
-the right level of confidence in the assignment.
+to the organism they belong to. In this case, the probabilities of false
+positives give the right level of confidence in the assignment.
 
@@
 computed. Using the same ideas, one could calibrate heuristics used in
-other alignment methods, especially the expanding field of long-read
+other alignment methods, especially in the expanding field of long-read
 technologies.

### 2019-04-08 MEM.tex
Ruggero on board
Removed myself from the acknowledgements.
Added myself in the authors list.
added a comment
On the role of the word "nearly" in section 1.2

@@
 \author[1,2]{Guillaume J. Filion}
+\author[1,2]{Ruggero Cortini}
 \author[1,2]{Eduard Zorita}
@@
 heuristic is off target. For lack of theoretical framework, this
-information is missing for nearly all existing seeding methods, meaning
+information is missing for nearly % nearly, or all? this is important because if
+% there is at least one seeding algorithm that actually estimates correctly the
+% probabilities for the seeding heuristics, then all of the rest is much less
+% meaningful.
+all existing seeding methods, meaning
 that current mappers cannot be properly calibrated.
@@
 
-We thank Ruggero Cortini for suggesting Monte Carlo methods as a way to
-tackle intractable cases. We acknowledge the financial support of the
-Spanish Ministry of Economy and Competitiveness (‘Centro de Excelencia
-Severo Ochoa 2013-2017’, Plan Nacional BFU2012-37168), of the CERCA
-Programme~/~Generalitat de Catalunya, and of the European Research Council
-(Synergy Grant 609989).
+We acknowledge the financial support of the Spanish Ministry of Economy and
+Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan Nacional
+BFU2012-37168), of the CERCA Programme~/~Generalitat de Catalunya, and of the
+European Research Council (Synergy Grant 609989).
 

### 2019-04-09 MEM.tex
fixed typos sections 1-4.

@@
 FM-index~\cite{pmid20460430}. Hash tables are typically used to index
-$k$-mers, which makes them useful to search seeds of set size $k$,
+$k$-mers, which makes them useful to search seeds of set size $k$
 (see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
@@
 Every symbol of the alphabet that is different from $\square$ is called a
-terminator. A segment is a sequence of 0 or more symbols $\square$
-followed by a terminator. The last segment of the read is terminated by
-the symbol $|$ and is called the tail.
+terminator. A segment is a terminator preceded by 0 or more symbols $\square$.
+The last segment of the read is terminated by the symbol $|$ and is called the
+tail.
 \end{definition}
@@
 matrix $M_n(z) + M_n(z)^2 + \ldots = M_n(z)\cdot(I-M_n(z))^{-1}$. To see
-why, observe that at the start of every read, the next nucleotide is a
+why, observe that, at the start of every read, the next nucleotide is a
 non-skipped position, so every read can be prepended by $\Downarrow_0$
@@
 
-$M_n(z)$ is too complex to compute a closed expression of $M_n(z) \cdot
-(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$
-and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no
-influence on the coefficients of the weighted generating function $a_0,
-a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments (including the
-obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
-functions of reads with exactly $s+1$ segments including the tail, $a_k$
-cannot depend on $M(z)^{k+2}, M(z)^{k+3}$. More formally, we can prove by
-induction that all the entries of $M(z)^k$ are divisible by $z^{k-1}$,
-showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to
-$a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.
+$M_n(z)$ is too complex to allow computing a closed expression of $M_n(z) \cdot
+(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$ and
+observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no influence on
+the coefficients of the weighted generating function $a_0, a_1, \ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k+1$ segments (including the obligatory
+tail). Since $M(z)^{s+1}$ contains the weighted generating functions of reads
+with exactly $s+1$ segments including the tail, $a_k$ cannot depend on
+$M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can prove by induction that
+all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
+contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
++a_kz^k$ is strictly $0$.
 
@@
 $\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
-instance the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
+instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
 mismatch against the $-$ sequence, that it is a match for the $+$

### 2019-04-10 MEM.tex
fixed typos sections 5 and 6

@@
 \end{eqnarray}
+where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.
 
@@
 
-$\tilde{M}_0(z)$ is too complex to compute a closed expression of
+$\tilde{M}_0(z)$ is too complex to allow computing a closed expression of
 $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
@@
 nucleotide. Since a sequencing error is always a mismatch against the
-target, the symbol $\downarrow_{/0}$ indicates that the ncleotide is a
+target, the symbol $\downarrow_{/0}$ indicates that the nucleotide is a
 mismatch against \emph{every} sequence. The symbols $\uparrow_{/i}$
@@
 
-Note that in the symbols $\downarrow_{/m}$ and $\downarrow_{/i}$, the
+Note that in the symbols $\downarrow_{/m}$ and $\uparrow_{/i}$, the
 numbers $m$ and $i$ have different meanings. In the symbol
@@
 terminator. We then specify the weighted generating functions of those
-segments and fill the tansfer matrix $\mathring{M}_N(z)$ of the reads that
+segments and fill the transfer matrix $\mathring{M}_N(z)$ of the reads that
 do not contain an on-target MEM seed. We introduce the terms of the matrix
-by increasing oder of complexity.
+by increasing order of complexity.
 
@@
 
-The definition of $\omega_m$ is correct: if the nucleotide is an error,
+The expression for $\omega_m$ is exact: if the nucleotide is an error,
 the symbol is $\downarrow_{/m}$ for some $m$ between 0 and $N$. Each
@@
 masks and $N-m$ soft masks. If all the masks vanish before the first read
-error, the next terminator will be a sybmol $\uparrow_{/j}$, otherwise it
-will be the sybmol $|$ or a symbol $\downarrow_{/m}$. We separate the
+error, the next terminator will be a symbol $\uparrow_{/j}$, otherwise it
+will be the symbol $|$ or a symbol $\downarrow_{/m}$. We separate the
 cases based on the terminator of the segment.
@@
 
-The definition of $\xi_j$ is correct: every nucleotide of the duplicate
+The expression for $\xi_j$ is exact: every nucleotide of the duplicate
 differs from the target with probability $\mu$. In the absence of
-sequencing error, this is also the probability that a nucleotide of the
+sequencing errors, this is also the probability that a nucleotide of the
 duplicate differs from the read. Given that there is no error, the
@@
 masks. If a soft mask covers the whole segment as indicated by the grey
-sqaures, then there can be no on-target seed. In all other cases there is
+squares, then there can be no on-target seed. In all other cases there is
 an on-target MEM seed.
@@
 
-The definition of $\eta_j$ is correct: the probability that there are $j$
+The expression for $\eta_j$ is exact: the probability that there are $j$
 matches between the duplicate and the target is $(1-\mu)^j$. If there are
@@
 $\downarrow_{/n}$ ($0 \leq n \leq N$) in order to fill the different
-entries of the transfer matrix. For this, we split this term on the number
+entries of the transfer matrix. For this, we split this term in the number
 of soft masks that run until and including the terminator. From expression
@@
 (\ref{eq:xi}) and (\ref{eq:eta}), the probability that a given soft mask
-fails withtin the first $j$ nucleotides is $\xi_j/\eta_j$ --- this is the
+fails within the first $j$ nucleotides is $\xi_j/\eta_j$ --- this is the
 conditional probability that it fails within the first $j$ nucleotides
@@
 \label{eq:A}
-A_{m,n} =
+A_{m,n}(z) =
 \omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - \xi_j^N \Big) (qz)^j + \omega_n
@@
 \tag{\ref{eq:A}}
-A_{m,n} =
+A_{m,n}(z) =
 \omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
@@
 segments that can be appended to the symbol $\downarrow_{/0}$ and that are
-terimated by a tail.
+terminated by a tail.
 
@@
 The curves show the probability that seeding is off target for skip-9
-seeds of size $\gamma=17$. Read size and erorr rate are the same as in
+seeds of size $\gamma=17$. Read size and error rate are the same as in
 Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line

### 2019-04-10 MEM.tex
finished correcting typos

@@
 $\beta$, one can choose for instance $\beta=1/2$ so that the term
-$(1-/\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
+$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
 
@@
 initial tests with real data suggest that this is a minor impediment (most
-likely because this is only one of several simplififying assumptions).
+likely because this is only one of several simplifying assumptions).
 Incorporating insertions and deletions would make the theory intractable,

### 2019-04-11 MEM.tex
added acknowledgements statement
Incorporated Ruggero's edits to the main text and figures.

@@
 heuristic is off target. For lack of theoretical framework, this
-information is missing for nearly % nearly, or all? this is important because if
-% there is at least one seeding algorithm that actually estimates correctly the
-% probabilities for the seeding heuristics, then all of the rest is much less
-% meaningful.
-all existing seeding methods, meaning
-that current mappers cannot be properly calibrated.
+information is missing, meaning that current mappers cannot be properly
+calibrated.
 
@@
   \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
-J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
+J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
 \end{eqnarray}
@@
 
-$M_n(z)$ is too complex to allow computing a closed expression of $M_n(z) \cdot
-(I-M_n(z))^{-1}$. We return to the definition $M_n(z) + M_n(z)^2 + \ldots$ and
-observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3}, \ldots$ have no influence on
-the coefficients of the weighted generating function $a_0, a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments (including the obligatory
-tail). Since $M(z)^{s+1}$ contains the weighted generating functions of reads
-with exactly $s+1$ segments including the tail, $a_k$ cannot depend on
-$M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can prove by induction that
-all the entries of $M(z)^k$ are divisible by $z^{k-1}$, showing that the
-contribution of $M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots
-+a_kz^k$ is strictly $0$.
+$M_n(z)$ is too complex to allow a closed expression of $M_n(z) \cdot
+(I-M_n(z))^{-1}$ to be computed. We return to the definition $M_n(z) +
+M_n(z)^2 + \ldots$ and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
+\ldots$ have no influence on the coefficients of the weighted generating
+function $a_0, a_1, \ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k+1$ segments (including the
+obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
+functions of reads with exactly $s+1$ segments including the tail, $a_k$
+cannot depend on $M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can
+prove by induction that all the entries of $M(z)^k$ are divisible by
+$z^{k-1}$, showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} +
+\ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.
 
@@
 
-$\tilde{M}_0(z)$ is too complex to allow computing a closed expression of
-$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
-proceed as in section~\ref{sec:example_skip} and to compute the powers of
-$\tilde{M}_0(z)$ (using the arithmetic of truncated polynomials) up to a
-given threshold value. Since each segment except the tail contains a
-mismatch against at least one sequence, we define $\tilde{p}$ as the upper
-bound on the probability of a mismatch $\tilde{p} = \max\{b,c,d\}$. The
-updated formula (\ref{eq:bound}) now gives an upper bound of the
-probability that a read of size $k$ contains $s$ or more mismatches as
+$\tilde{M}_0(z)$ is too complex to allow a closed expression of
+$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$ to be computed. It is
+actually easier to proceed as in section~\ref{sec:example_skip} and to
+compute the powers of $\tilde{M}_0(z)$ (using the arithmetic of truncated
+polynomials) up to a given threshold value. Since each segment except the
+tail contains a mismatch against at least one sequence, we define
+$\tilde{p}$ as the upper bound on the probability of a mismatch $\tilde{p}
+= \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now gives an upper
+bound of the probability that a read of size $k$ contains $s$ or more
+mismatches as
 \begin{equation*}
@@
 BFU2012-37168), of the CERCA Programme~/~Generalitat de Catalunya, and of the
-European Research Council (Synergy Grant 609989).
+European Research Council (Synergy Grant 609989). R.~C.~ was supported by the
+People Programme (Marie Curie Actions) of the European Union's Seventh Framework
+Programme (FP7/2007-2013) under REA grant agreement 608959.
 

### 2019-04-16 MEM.tex
Updated affiliations and acknowledgements according to instructions.

@@
 \author[1,2]{Guillaume J. Filion}
-\author[1,2]{Ruggero Cortini}
-\author[1,2]{Eduard Zorita}
-\affil[1]{Genome Architecture, Gene Regulation, Stem Cells and Cancer
-Programme, Center for Genomic Regulation (CRG), The Barcelona Institute of
+\author[1]{Ruggero Cortini}
+\author[1]{Eduard Zorita}
+\affil[1]{Center for Genomic Regulation (CRG), The Barcelona Institute of
 Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
-\affil[2]{University Pompeu Fabra, Doctor Aiguader, 08003 Barcelona,
-Spain.}
+\affil[2]{University Pompeu Fabra (UPF), Barcelona, Spain.}
 
@@
 
-We acknowledge the financial support of the Spanish Ministry of Economy and
-Competitiveness (‘Centro de Excelencia Severo Ochoa 2013-2017’, Plan Nacional
-BFU2012-37168), of the CERCA Programme~/~Generalitat de Catalunya, and of the
-European Research Council (Synergy Grant 609989). R.~C.~ was supported by the
-People Programme (Marie Curie Actions) of the European Union's Seventh Framework
-Programme (FP7/2007-2013) under REA grant agreement 608959.
+We acknowledge the financial support of the Spanish Ministry of Economy,
+Industry and Competitiveness (`Centro de Excelencia Severo Ochoa
+2013-2017', Plan Nacional BFU2012-37168), of the CERCA
+Programme~/~Generalitat de Catalunya, and of the European Research Council
+(Synergy Grant 609989). R.~C. was supported by the People Programme
+(Marie Curie Actions) of the European Union's Seventh Framework Programme
+(FP7/2007-2013) under REA grant agreement 608959. We also acknowledge
+support of the Spanish Ministry of Economy and Competitiveness (MEIC) to
+the EMBL partnership.
 

### 2019-04-25 manuscript.tex
Added edits from eduardvalera@gmail.com. Vesion submitted to bioRxiv.

--- 2019-04-16 MEM.tex
@@
 
-\title{On mapping errors and seeding heuristics for
-DNA alignment}
+\title{Calibrating seed-based heuristics to map short DNA reads}
 
@@
 % Statement of the problem.
-The increasing throughput of DNA sequencing technologies demands faster
-algorithms to analyze the data. The fate of most reads is to be mapped to
-a reference sequence, typically a genome. Modern mappers rely on
-heuristics to gain speed at a reasonable cost for accuracy. In the seeding
-approach, short matches between the reads and the genome are used to
-narrow the search to a set of candidate locations. Several seeding
-variants used in modern mappers show good empirical performance but they
-are difficult to calibrate or to optimize for lack of theoretical results.
+The increasing throughput of DNA sequencing technologies creates a need
+for faster algorithms. The fate of most reads is to be mapped to a
+reference sequence, typically a genome. Modern mappers rely on heuristics
+to gain speed at a reasonable cost for accuracy. In the seeding approach,
+short matches between the reads and the genome are used to narrow the
+search to a set of candidate locations. Several seeding variants used in
+modern mappers show good empirical performance but they are difficult to
+calibrate or to optimize for lack of theoretical results.
 % Summary of the results.
 Here we develop a theory to estimate the probability that reads are
-mismapped due to failures at the seeding step. We describe the properties
-of simple exact seeds, skip-seeds and MEM seeds (Maximal Exact
-Match).
+mapped to a wrong location due to limitations at the seeding step. We
+describe the properties of simple exact seeds, skip-seeds and MEM seeds
+(Maximal Exact Match).
 % Innovation.
-The major innovation is to use concepts from analytic combinatorics to
-represent reads as abstract sequences, and to specify their generative
-function to estimate the probabilities of interest.
+The main innovation of this work is to use concepts from analytic
+combinatorics to represent reads as abstract sequences, and to specify
+their generative function to estimate the probabilities of interest.
 % Durability.
-We provide several algorithms, which combined together give a practicial
-solution to the problem of calibrating seeding heuristic for short reads.
+We provide several algorithms, which combined together give a practical
+solution to the problem of calibrating seeding heuristics for short reads.
 % Scope.
@@
 of its duplicates, the \emph{true} location of a DNA fragment is not
-always the \emph{best} (as measured by the identity between the sequence
-and the candidate location). This leads to the following question: how can
-we identify the \emph{best} location of the fragment in the genome?
+always the \emph{best} candidate (as measured by the identity between the
+sequence and the candidate location). This leads to the following
+question: how can we identify the \emph{best} location of the fragment in
+the genome?
 
@@
 
-There exists algorithms to solve the best location
+There exist algorithms to solve the best location
 problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
 the large amount of data generated by modern sequencers. Instead, one uses
-heuristic methods, \textit{i.e.} algorithms that run fast, but that may
+heuristic methods, \textit{i.e.} algorithms that run faster, but that may
 return an incorrect result~\cite{Waterman1984}.
@@
 locations and reduce the input size of the (slow) alignment algorithm. The
-disadvantage is that the target location is not always in the candidate
-set. When that happens, the read cannot be mapped to the correct location
-because it has been filtered out at the seeding step.
+disadvantage is that the target location is not always detected and
+included in the candidate set. When that happens, the read cannot be
+mapped to the true location because it has been filtered out at the
+seeding step.
 
@@
 speed and accuracy: If the seeding step is set to produce many candidates,
-it is likely to retrieve the candidate. The downside is that this will
-take long. Conversely, if the seeding step is set to produce few
-candidates, the process will run faster but the target is less likely to
-be in the candidate set.
+it is likely to retrieve the true location. The downside is that
+processing all the candidates would take long. Conversely, if the seeding
+step is set to produce few candidates, the process would run faster but
+the target is less likely to be in the candidate set.
 
@@
 
-Before going further, we introduce some vocabulary that will be useful to
-simplify the exposition of the theory.
+Before going further, we introduce some nomenclature that will be useful
+to simplify the exposition of the theory.
 
@@
 
-This observation is fundamental: in order to estimate the error rate of of
-a mapping algorithm, we must know how often the associated seeding
-heuristic is off target. For lack of theoretical framework, this
-information is missing, meaning that current mappers cannot be properly
-calibrated.
+This observation is fundamental: in order to estimate the error rate of a
+mapping algorithm, we must know how often the associated seeding heuristic
+is off target. For lack of theoretical framework, this information is
+missing, meaning that current mappers are not properly calibrated.
 
@@
 seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
-not provide a way to control mapping errors. Other authors investigated
-the reliability of mapping algorithms~\cite{pmid23872968}, but they
-focused on the probability that random sequences have a single hit,
-recognizing that solving the problem requires taking into consideration
-the repeat structure of the genome. Here we address this problem in
-general terms and we provide a practical way to estimate the mismapping
-rate in complex genomes.
+not provide a way to estimate the frequency of mapping errors. Other
+authors investigated the reliability of mapping
+algorithms~\cite{pmid23872968}, but they focused on the probability that
+random sequences have a single hit, recognizing that solving the problem
+requires taking into consideration the repeat structure of the genome.
+Here we address this problem in general terms and we provide a practical
+way to estimate the mismapping rate in complex genomes.
 
@@
 didactic manner. The order of the examples is meant to facilitate the
-aquisition of the main concepts, meaning that we sometimes go back and
-forth between different types of seeding problems. We first give some
-background on the different types of seeding methods considered here, we
-then introduce the computational strategy based on transfer matrices, and
-then delve into each specific seeding method.
+acquisition of the main concepts, meaning that we sometimes go back and
+forth between different types of seeding strategies. We first give some
+background on the different types of seeding methods considered here; we
+then introduce the computational strategy based on transfer matrices;
+finally we delve into each specific seeding method.
 
@@
 uses ``spaced seeds'' that tolerate mismatches. To avoid potential
-confusion, we will adopt the following convention.
+confusion, we will adopt the following convention:
 
 \begin{definition}
-A seed is a subsequence of the read that is above a certain size $\gamma$
+A seed is a subsequence of the read that has size at least $\gamma$
 (defined by the context of the problem) and that has at least one perfect
@@
 
-When a seed matches a particular location of the genome, we will say that
-it is a ``seed for that location''. For instance, we will often use
-expressions such as ``a seed for the target'' or ``a seed for a
-duplicate''.
+When a seed matches a particular location of the genome, we say that it is
+a ``seed for that location''. For instance, we often use expressions such
+as ``a seed for the target'' or ``a seed for a duplicate''.
 
@@
 subsequence is a seed we need to know if it exists somewhere in the
-genome. This is a non trivial problem in itself, but fortunately there
-exists practical methods to solve it, even when the reference genome is
-very large.
+genome. This is a non trivial problem in itself, but fortunately we can
+use practical methods to solve it, even when the reference genome is very
+large.
 
@@
 FM-index~\cite{pmid20460430}. Hash tables are typically used to index
-$k$-mers, which makes them useful to search seeds of set size $k$
+$k$-mers, which makes them useful to search seeds of fixed size $k$
 (see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
-algorithms). In contrast, the FM-index has no set size so it can be used
-to search seeds of different lengths~\cite{ferragina2000opportunistic,
-ferragina2005indexing}. It is a compact data structure based on the suffix
+algorithms). In contrast, some text-indexing structures have no set size
+so they can be used to search seeds of different lengths. For instance,
+the FM-index is a compact data structure based on the suffix
 array~\cite{manber1993suffix} and on the Burrows-Wheeler
 transform~\cite{burrows1994block}, emulating a suffix trie with a much
-smaller memory footprint.
+smaller memory footprint~\cite{ferragina2000opportunistic,
+ferragina2005indexing}.
 
@@
 suffix array~\cite{dobin2013star}) but the FM-index is currently the most
-popular choice for seeding methods. For MEM seeds described below, it is
-even the only practical option~\cite{pmid24336412, pmid25399029,
-pmid23349213, pmid19389736}. However, the detail if of little interest
-here: we simply assume that seeds are known at all times without ambiguity
-because this problem has several practical solutions.
+popular choice for seeding methods. For MEM seeds, it is even the only
+practical option~\cite{pmid24336412, pmid25399029, pmid23349213,
+pmid19389736}. However, the detail is of little interest here: we simply
+assume that seeds are known at all times without ambiguity because this
+problem has several practical solutions.
 
@@
 BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
-overlapping matches in the same regions, as we are about to illustrate.
-
-Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
-by an imperfect sequencer. The sequenced DNA fragment comes from a region
-of the genome that has three duplicates and the read has three miscalled
-nucleotides (indicated by stars).
+overlapping matches within the same regions.
+
+Fig.~\ref{fig:exact_seed_example} shows a read produced by an imperfect
+sequencer. The sequenced DNA fragment comes from a region of the genome
+that has three duplicates and the read has three miscalled nucleotides
+(indicated by stars).
 
@@
 (Target), and above homologous sequences also present in the genome
-(Duplicates). Sequencing errors are indicated by a star, each marked
-\texttt{T} was actually an \texttt{A} that was misread. The mismatches
+(Duplicates). Sequencing errors are indicated by a star (each marked
+\texttt{T} was actually an \texttt{A} that was misread). The mismatches
 between the genomic sequences and the read are indicated in bold. The
@@
 
-The exact seeds with $\gamma=7$ are shown above the sequence of the read.
 Observe that erroneous nucleotides can belong to exact seeds because they
@@
 error matches all the duplicates and belongs to an off-target seed.
-However, erroneous nucleotides that are mismatches for \emph{all} the
+However, sequencing errors that are mismatches for \emph{all} the
 sequences cannot belong to a seed. This is the case of the second
@@
 already in the candidate set. In addition, this seeding method is not
-particularly sensitive compared to spaced seed~\cite{pmid11934743} so it
+particularly sensitive compared to spaced seeds~\cite{pmid11934743} so it
 is used only in few specific applications. Nevertheless, it will be
@@
 highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
-missing seeds are represented by dotted lines. Skipping nucleotides
-reduces the clutter at the center of the read, but it also removes the
-only seed for the second duplicate.}
+missing seeds are represented by dotted lines.}
 \label{fig:skip_seed_example}
@@
 are used in BWA-MEM~\cite{li2013aligning} where they give good empirical
-results. To describe MEM seeds, let us first focus on MEMs.
+results. To describe MEM seeds, let us first clarify the meaning of
+``Maximal Exact Match''.
 
@@
 target. Lowering $\gamma$ does not change this, so there is no way to
-discover the true location in this case (even though it is the best
+discover the true location using MEM seeds (even though it is the best
 location).
@@
 at the bottom effectively hide the target, so it cannot be discovered.
-This can occur even when the true location is the best and when there is a
-single error on the read.}
+This can occur even when the true location is the best candidate and when
+there is a single error on the read.}
 \label{fig:full_masking_example}
@@
 rate. In other words, we ignore the complications due to the genealogy of
-the duplication events and we simply assume that at each nucleotide
+the duplication events. Instead, we simply assume that at each nucleotide
 position, any given duplicate is identical to the target with probability
 $1-\mu$. If it is not, we assume that the duplicate sequence can be any of
-one the remaining three nucleotides (\textit{i.e.} each is found with
+one the three remaining nucleotides (\textit{i.e.} each is found with
 probability $\mu/3$).
@@
 we recode the reads in alphabets of custom symbols and we specify a
-construction plan of the reads using a so-called \emph{transfer matrix}
-$M(z)$. The transfer matrix specifies which types of segments can follow
-each other in the reads of interest: the entry at coordinates $(i,j)$ is
-the weighted generating function of segments of type $j$ that can be
-appended to segments of type $i$.
-
-From the basic operations on weighted generating functions, the entry of
-$M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$ at
-coordinates $(i,j)$ is the weighted generating function of all the reads
-that end with a segment of type $j$ and that can be appended to a segment
-of type $i$. The examples below will clarify the key steps of this
-strategy.
+construction plan of the reads using a \emph{transfer matrix} $M(z)$. The
+transfer matrix specifies which types of segments can follow each other in
+the reads of interest: the entry at coordinates $(i,j)$ is the weighted
+generating function of segments of type $j$ that can be appended to
+segments of type $i$.
+
+$M(z)$ contains the weighted generating functions of all the reads that
+consist of a single fragment. From the basic operations on weighted
+generating functions, $M(z)^s$ contains the weighted generating functions
+of all the reads that consist of $s$ segments. Therefore, the entry at
+coordinates $(i,j)$ of the matrix $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z)
+\cdot (I-M(z))^{-1}$ is the weighted generating function of the reads of
+any size and any number of segments, that end with a segment of type $j$
+and that can be appended to a segment of type $i$. The examples below will
+clarify the key steps of this strategy.
 
@@
 
-This recoding allows us to partition the read in am important way.
+This recoding allows us to partition the read in an important way.
 
@@
 An example read is represented in the mismatch alphabet. The symbol
-$\Downarrow$ represents a mismatch against the target (it is an erroneous
-nucleotide) and the symbol $\square$ represents a match (it is a correct
+$\Downarrow$ represents a mismatch against the target (an erroneous
+nucleotide) and the symbol $\square$ represents a match (a correct
 nucleotide). The symbol $|$ is appended to the end of the read. The
@@
 \frac{1+qz+\ldots+(qz)^{\gamma-1}}
-  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} = a_0 + a_1z + a_2z^2 +
-\ldots
+  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
 \end{equation}
 
-The coefficients of the Taylor series $a_0 + a_1z + a_2z^2 + \ldots$ are
-unknown. But by construction, $a_k$ is the quantity of interest,
-\textit{i.e.} it is the probability that a read of size $k$ does not
-contain an on-target seed, so we now need to extract the coefficients
-from expression (\ref{eq:F}). There are several methods to do so; the one
-we choose here is to build a recurrence equation to compute $a_k$ from the
-previous terms of the series $a_0, a_1, \ldots, a_{k-1}$. For this, we
-observe that for $z \neq 1/q$, equation (\ref{eq:F}) is equivalent to
+This function can be expanded as a Taylor series $a_0 + a_1z + a_2z^2 +
+\ldots$, but the coefficients $a_0, a_1, a_2, \ldots$ are unknown. By
+construction, $a_k$ is the quantity of interest, \textit{i.e.} it is the
+probability that a read of size $k$ does not contain an on-target seed, so
+we now need to extract the coefficients from expression (\ref{eq:F}).
+There are several methods to do so; the one we choose here is to build a
+recurrence equation to compute $a_k$ from the previous terms of the series
+$a_0, a_1, \ldots, a_{k-1}$. For this, we observe that for $z \neq 1/q$,
+equation
+\begin{equation*}
+\frac{1+qz+\ldots+(qz)^{\gamma-1}}
+  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} =
+  a_0 + a_1z +a_2z^2 + \ldots
+\end{equation*}
+is equivalent to
 \begin{equation*}
@@
 1            &\quad\text{if } k < \gamma, \\
-1 -pq^\gamma &\quad\text{if } k = \gamma, \\
+1 -q^\gamma &\quad\text{if } k = \gamma, \\
 a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
@@
 \end{equation}
+
+These results are consistent: for $k < \gamma$, the read is too short so
+the probability that it contains no seed is 1; for $k = \gamma$, the
+read contains a seed if and only if it has no error, which occurs with
+probability $q^\gamma$.
 
@@
 alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
-\ldots, \Downarrow_n \}$. The symbol $\square$ still represents a correct
-nucleotide and the symbol $|$ is still a terminator added at the end of
-the read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent
-sequencing errors and $i$ indicates the number of nucleotides till the
-next non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
+\ldots, \Downarrow_n \}$. Again, the symbol $\square$ represents a correct
+nucleotide and the symbol $|$ is a terminator added at the end of the
+read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
+errors and $i$ indicates the number of nucleotides till the next
+non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
 before a non-skipped position and $i=n$ for nucleotides at a non-skipped
@@
 those $\gamma+i$ possible segments are distributed according to the laws
-of the arithmetic modulo $n+1$. For $i > 0$ for instance, if the next
+of the arithmetic modulo $n+1$. When $i > 0$, for instance, if the next
 segment has no symbol $\square$ then it is terminated by
@@
 through the same mutagenesis process, the events are conditionally
-independent (we do not have full independence because $S_0$ depends on
-sequencing errors only whereas the events $S_j$ depend on sequencing
-errors and on the divergence between duplicates). We can thus write
+independent (we do not have full independence because $S_0$ only depends
+on sequencing errors whereas the events $S_j$ depend on sequencing errors
+and on the divergence between duplicates). We can thus write
 \begin{eqnarray}
@@
 
-The probability that seeding is off target thus depends on only two
+Hence, the probability that seeding is off target is a function of two
 quantitites. We have already seen how to compute $P(\overline{S}_0)$ in
@@
 
-It will be useful to consider a slightly more general problem where we
-have two sequences of interest labelled $+$ and $-$. One can say that the
-$+$ sequence stands for the target and that the $-$ sequence stands for
-its duplicate. We then define the dual alphabet $\tilde{\mathcal{A}}_0 =
-\{\square, |, \downarrow_{/1}^-, \downarrow_{/2}^-, \ldots,
-\downarrow_{/1}^+, \downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols
-$\downarrow_{/i}^-$ signify that the nucleotide is a mismatch against the
-$-$ sequence only, the symbols $\downarrow_{/i}^+$ signify that it is a
-mismatch against the $+$ only, and the symbol $\Downarrow$ signifies that
-it is a mismatch against both. As before, every other nucleotide is
-replaced by the symbol $\square$, and the terminator $|$ is appended to
-the end of the read. We again define reads as sequences of segments,
-except that now the terminators are the symbols $\downarrow_{/j}^-$,
-$\downarrow_{/j}^+$ and $\Downarrow$. The tail, as usual, is terminated by
-the symbol $|$.
+It will be useful to consider a more general problem where we
+have two sequences of interest labelled $+$ and $-$. The $+$ sequence
+stands for the target and that the $-$ sequence stands for its duplicate.
+We then define the dual alphabet $\tilde{\mathcal{A}}_0 = \{\square, |,
+\downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
+\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/i}^-$
+signify that the nucleotide is a mismatch against the $-$ sequence only,
+the symbols $\downarrow_{/i}^+$ signify that it is a mismatch against the
+$+$ only, and the symbol $\Downarrow$ signifies that it is a mismatch
+against both. As before, every other nucleotide is replaced by the symbol
+$\square$, and the terminator $|$ is appended to the end of the read. We
+again define reads as sequences of segments, except that now the
+terminators are the symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and
+$\Downarrow$. The tail, as usual, is terminated by the symbol $|$.
 
@@
 
-Fig.~\ref{fig:skip_dual} shows the same read as Fig.~\ref{fig:dual}, but
-here it is represented in the skip-3 dual encoding. It is important to
-note several differences with Fig.~\ref{fig:dual}. The first is that the
-symbols $\Downarrow_i$ are not always associated with double mismatches.
-For instance, the symbol $\Downarrow_0$ on the right side of the read
+Fig.~\ref{fig:skip_dual} shows the read from Fig.~\ref{fig:dual}
+represented in the skip-3 dual encoding. It is important to note several
+differences with Fig.~\ref{fig:dual}. The first is that the symbols
+$\Downarrow_i$ are not always associated with double mismatches. For
+instance, the symbol $\Downarrow_0$ on the right side of the read
 corresponds to a mismatch for the $+$ sequence only. This happens whenever
@@
 nucleotide).
-
-Another difference is that when the symbols $\downarrow^-_{/i}$ and
-$\downarrow^+_{/i}$ alternate, $i$ is no longer the distance to the
-previous terminator. This is visible for the symbol $\downarrow^+_{/4}$ on
-the left side of the read, where the previous terminator is 6 nucleotides
-upstream. In this case, $i$ is the distance to the previous non-skipped
-position.
 
@@
 At a given position of the read, a duplicate is a \emph{hard mask} if its
-match length is strictly longer than the match length of the target. A
-duplicate is a \emph{soft mask} if it has the same match length as the
-target.
+match length on the left side is strictly longer than the match length
+of the target. A duplicate is a \emph{soft mask} if it has the same match
+length as the target.
 \end{definition}
@@
 indicate a change in masking status: the nucleotide is not masked but the
-previous is. The index $i \geq 1$ is the number of nucleotides since the
+previous is --- this happens when all the duplicates fail to extend beyond
+this position. The index $i \geq 1$ is the number of nucleotides since the
 last mismatch or since the beginning of the read. All the other
@@
 the index $i$ is a number of nucleotides. Fig.~\ref{fig:sketch_extended}
-shows the encoding of the read from Fig.~\ref{fig:exact_seed_example} in
-the MEM alphabet.
+shows the encoding of a read in the MEM alphabet.
 
@@
 symbol $\downarrow_{/m}$. Indeed, such a stretch is a match for the target
-because it does not contain any symbol $\downarrow_{/m}$, it is match for
-the target only because it contains at least one unmasked nucleotide
-(marked by $\uparrow_{/i}$), and it cannot be extended because it is
-flanked by sequencing errors (symbols $\downarrow_{/m}$) or by the ends of
-the reads. Note that there is exactly one symbol $\uparrow_{/i}$ per
-strict on-target MEM seed, and therefore two symbols $\uparrow_{/i}$ must
-be separated by at least one symbol $\downarrow_{/m}$.
+because it does not contain any symbol $\downarrow_{/m}$, it only matches
+the target because it contains at least one unmasked nucleotide (marked by
+$\uparrow_{/i}$), and it cannot be extended because it is flanked by
+sequencing errors (symbols $\downarrow_{/m}$) or by the ends of the reads.
+Note that there is exactly one symbol $\uparrow_{/i}$ per strict on-target
+MEM seed, and therefore two symbols $\uparrow_{/i}$ must be separated by
+at least one symbol $\downarrow_{/m}$.
 
@@
 With this notation, the probability that at least one of $N$ masks
-survives a sequence of $j$ error-free nucleotides is thus $1-\xi_j^N$, and
-the probability that there remains a mask at the $j-1$-th but not at the
-$j$-th error-free nucleotide is $\xi_j^N - \xi_{j-1}^N$. From this we
-conclude that the weighted generating function of the segments terminated
-by $\uparrow_{/j}$ following a segment terminated by $\downarrow_{/m}$ is
+survives a sequence of $j$ error-free nucleotides is thus $1-(\xi_j)^N$,
+and the probability that there remains a mask at the $j-1$-th but not at
+the $j$-th error-free nucleotide is $(\xi_j)^N - (\xi_{j-1})^N$. From this
+we conclude that the weighted generating function of the segments
+terminated by $\uparrow_{/j}$ following a segment terminated by
+$\downarrow_{/m}$ is
 \begin{equation}
 \label{eq:B}
-B_j(z) = \Big( \xi_j^N-\xi_{j-1}^N \Big) (qz)^j.
+B_j(z) = \Big( (\xi_j)^N-(\xi_{j-1})^N \Big) (qz)^j.
 \end{equation}
@@
 
-In this case there can be no on-target seed because the read finishes
-too early. We must just enforce the condition that at least one of the $N$
-masks survives until the end of the segment in order to exclude the
-symbols $\uparrow_{/j}$. The weighted generating function is
+In this case there can be no on-target seed because the read finishes too
+early. However, we must enforce the condition that at least one of the $N$
+masks survives until the end, otherwise the segment would be terminated by
+one of the symbols $\uparrow_{/j}$. The weighted generating function is
 \begin{equation*}
-\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
+\sum_{i=0}^{\gamma-1} \Big(1 - (\xi_i)^N \Big) (qz)^i.
 \end{equation*}
@@
 
-In this case, the soft masks do not hide the target: even if one of them
+In this case, the soft masks do not hide the target. Even if a duplicate
 survives until the end of the read, there will be an on-target seed
-(shared in this case). To avoid on-target seeds, we must enforce the
+(shared in this case). To exclude on-target seeds, we must enforce the
 condition that at least one hard mask survives until the end of the
@@
 \begin{equation*}
-\sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
+\sum_{i=\gamma}^\infty \Big(1 - (\xi_i)^m \Big) (qz)^i.
 \end{equation*}
@@
 C_m(z) =
-\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
-  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
+\sum_{i=0}^{\gamma-1} \Big(1 - (\xi_i)^N \Big) (qz)^i +
+  \sum_{i=\gamma}^\infty \Big(1 - (\xi_i)^m \Big) (qz)^i.
 \end{equation}
@@
 least one of the $N$ masks survives until the terminator. For a read of
-size $j+1$, this occurs with probability $1-\xi_j^N$. Including the
+size $j+1$, this occurs with probability $1-(\xi_j)^N$. Including the
 terminator and summing over $j+1 \leq \gamma$, we see that the weighted
@@
 \begin{equation}
-\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - \xi_j^N \Big) (qz)^j.
+\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - (\xi_j)^N \Big) (qz)^j.
 \end{equation}
@@
 For a segment of size $j+1$, condition $i)$ has probability $\big(1 -
-\xi_j^m \big)$. Summing over $j+1 > \gamma$ and including the terminator,
-we see that the associated weighted generating function is
+(\xi_j)^m \big)$. Summing over $j+1 > \gamma$ and including the
+terminator, we see that the associated weighted generating function is
 \begin{equation*}
-\omega_n pz \sum_{j=\gamma}^\infty \Big(1 - \xi_j^m \Big) (qz)^j.
+\omega_n pz \sum_{j=\gamma}^\infty \Big(1 - (\xi_j)^m \Big) (qz)^j.
 \end{equation*}
@@
 matches between the duplicate and the read. The probability that the
-duplicate matches the subsequent error is $\mu/3$, so the probability of
+duplicate matches the subsequent error is $\mu/3$, so the probability
 that there are $j+1$ matches including the sequencing error is
@@
 \begin{equation*}
-\xi_j^m \Big(1 - \eta_j^{N-m} \Big),
+(\xi_j)^m \Big(1 - (\eta_j)^{N-m} \Big),
 \end{equation*}
@@
 \label{eq:softmasks_r}
-{N-m \choose r} (1 - \eta_j)^r \eta_j^{N-m-r}.
+{N-m \choose r} (1 - \eta_j)^r (\eta_j)^{N-m-r}.
 \end{equation}
@@
 matches is $n-r$ appears as the convolution product
-\begin{equation*}
-\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_j^{N-m-r}}
-\psi_{j,m,n,r}\text{, where \;}
-\psi_{j,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
-\xi_j^{n-r-q}.
-\end{equation*}
+\begin{eqnarray*}
+\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{(\eta_j)^{N-m-r}}
+&\psi_{j,m,n,r}&, \\
+\text{where }
+&\psi_{j,m,n,r}& = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
+(\xi_j)^{n-r-q}.
+\end{eqnarray*}
 
@@
 \begin{equation}
+\begin{split}
 \label{eq:A}
 A_{m,n}(z) =
-\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - \xi_j^N \Big) (qz)^j + \omega_n
-pz \sum_{j=\gamma}^\infty \Big(1 - \xi_j^m \cdot
+&\omega_n pz \sum_{j=0}^{\gamma-1} \Big(1 - (\xi_j)^N \Big)
+  (qz)^j \\
++ &\omega_n pz \sum_{j=\gamma}^\infty \Big(1 - (\xi_j)^m \cdot
 (1- \zeta_{j,m,n}) \Big) (qz)^j.
+\end{split}
 \end{equation}
@@
 where
+\begin{equation}
+\tag{\ref{eq:A}}
+\begin{split}
+A_{m,n}(z) =
+&\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - (\xi_i)^N \Big) (qz)^i \\
++ &\omega_n pz \sum_{i=\gamma}^\infty \Big(1 - (\xi_i)^m \cdot
+(1- \zeta_{i,m,n}) \Big) (qz)^i 
+\end{split}
+\end{equation}
 \begin{gather}
-\tag{\ref{eq:A}}
-A_{m,n}(z) =
-\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
-pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
-(1- \zeta_{i,m,n}) \Big) (qz)^i \\
 \tag{\ref{eq:B}}
-B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i \\
+B_i(z) = \Big( (\xi_i)^N-(\xi_{i-1})^N \Big) (qz)^i \\
 \tag{\ref{eq:C}}
 C_m(z) =
-\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
-  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i \\
+\sum_{i=0}^{\gamma-1} \Big(1 - (\xi_i)^N \Big) (qz)^i +
+  \sum_{i=\gamma}^\infty \Big(1 - (\xi_i)^m \Big) (qz)^i \\
 \tag{\ref{eq:D}}
@@
 \psi_{j,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
-\xi_j^{n-r-q}.
+(\xi_j)^{n-r-q}.
 \end{gather}
@@
 Kachitvichyanukul and Schmeiser~\cite{kachitvichyanukul1988binomial}.
-
-The running speed depends on the error rate of the instrument $p$, but not
-on the number of duplicates $N$. For values of $p$ around 0.01,
-the running time for 10,000,000 iterations is $1$-$10$ seconds on current
-hardware.
+Most importantly, the number of duplicates $N$ has little influence on
+the running speed of algorithm~\ref{alg:mcmc}.
 
 \begin{algorithm}[H]
+\SetKwInOut{Parameter}{Parameter}
 \label{alg:mcmc}
 \SetAlgoLined
+\Parameter {$k$ is the size of the reads.}
+\Parameter {$p$ is the error rate of the sequencer (substitutions only).}
+\Parameter {$N$ is the number of duplicates.}
+\Parameter {$\mu$ is the nucleotide-wise probability that duplicates
+differ.}
 \KwResult {Sample a read at random. Return 1 if the read contains a
@@
 Fig.~\ref{fig:curves_mem} reveals that in this concrete case, MEM seeds
-increases the chances that seeding is off target by a factor up to 40
+increase the chances that seeding is off target by a factor up to 40
 compared to exact seeds (see Fig.~\ref{fig:curves_exact}). On this
@@
 have a seed are considered candidate locations (as long as the seed is
-above the threshold $\gamma$). If not, the estimates of the mismapping
-rate are off. The hope is that spurious matches are shorter than $\gamma$
-so that they are invisible, but this is not always the case. And if they
-qualify as a seed we need to verify the candidate location.
+above the threshold $\gamma$). If not, the final estimates are biased. The
+hope is that spurious matches are shorter than $\gamma$ so that they are
+filtered out, but this is not always the case. And if they qualify as a
+seed we need to verify the candidate location.
 
@@
 mapper so we will not develop this further. What matters is that random
-hits do not have to impose a significant burden on the time needed to
-verify the candidates.
+hits do not impose a significant burden on the time needed to verify the
+candidates.
 
@@
 target nor for its duplicates). Since the hits are not homologous to the
-read, the alignment score is random, which makes it is easy to detect.
+read, the alignment score is extreme, which makes it easy to detect.
 
@@
 size $\gamma$ (and possibly larger in the case of MEM seeds). Say that
-there remains $L$ nucleotides and that $m$ of them are mismatches for the
+there remain $L$ nucleotides and that $m$ of them are mismatches for the
 candidate location. From Bayes formula, the probability that the location
@@
 Since the probability that the best candidate is a random sequence is
-either very small or very large, it either has no influence or it
-dominates the probability of a false positive. For every read, we can use
-as confidence score the maximum of this probability and of the estimated
-probability that seeding is off target.
+either very small or very large, it has no influence in the first case,
+or it dominates the probability of a false positive in the secondi case.
+For every read, we can use as confidence score the maximum of this
+probability and of the estimated probability that seeding is off target.
 
@@
 We have devised a set of methods to compute the probability that seeding
-heuristics fail and commits the mapping process to an error. This
+heuristics fail and commit the mapping process to an error. This
 fills a knowledge gap to understand and calibrate the performance of the
@@
 tested with an exact sequence alignment algorithm, so that the best
-location is always discovered if it is in the candidate set. This is not
-completely unrealistic, but it is important to realize that for plant and
-animal genomes, a single seed may have tens of thousands of hits.
-Therefore, most mappers impose a limit on the number of alignments so the
-best hit may not be found even if it is in the candidate set. This is
-again a minor impediment, since we can correct the probability that the
-location is false when only some candidates have been tested.
+location is always discovered if it is present in the candidate set. This
+is not completely unrealistic, but it is important to note that for plant
+and animal genomes, a single seed may have tens of thousands of hits.
+Therefore, most mappers impose a limit on the number of alignments per
+read, opening the possibility that the best hit is seeded but not aligned.
+This is again a minor impediment, since we can replace the probability
+that the location is false by the proportion of candidates that were not
+aligned.
 
@@
 simulation. For instance, with 10 million rounds of sampling, frequencies
-around $1/1,000,000$ or lower cannot be measured accurately. When
+around $1/1,000,000$ or lower cannot be measured accurately. When one is
 interested only in frequent events, it is thus a perfectly reasonable
-strategy.
-
-For $N < 20$, the probability that MEM seeding is null or off-target is
-relatively small, so we need a method that is accurate in this range.
-Fortunately, the transfer matrix method is fast because the dimension of
-the matrix $\mathring{M}_N(z)$ is small and the computations are not
-prohibitive.
+strategy. On the other hand, for $N < 20$, the probability that MEM
+seeding is null or off-target is relatively small, so we need a method
+that is accurate in this range. Fortunately, the transfer matrix method is
+fast because the dimension of the matrix $\mathring{M}_N(z)$ is small and
+the computations are not prohibitive.
 
@@
 Finally, one may wonder if our approach has any advantage over methods
-based on machine learning. This class of algorithms has already proved
+based on machine learning. Such algorithms have already proved
 useful~\cite{lee2014mosaik} and the rapid progress in the field of deep

### 2019-12-10 manuscript.tex
Add section about Sesame to the manuscript

@@
 \usepackage{graphicx}
+\usepackage{url}
 
@@
 
-\title{Calibrating seed-based heuristics to map short DNA reads}
+\title{Calibrating seed-based alignment heuristics with Sesame}
 
@@
 reference sequence, typically a genome. Modern mappers rely on heuristics
-to gain speed at a reasonable cost for accuracy. In the seeding approach,
+to gain speed at a reasonable cost for accuracy. In the seeding heuristic,
 short matches between the reads and the genome are used to narrow the
@@
 % Summary of the results.
-Here we develop a theory to estimate the probability that reads are
-mapped to a wrong location due to limitations at the seeding step. We
-describe the properties of simple exact seeds, skip-seeds and MEM seeds
-(Maximal Exact Match).
+Here we develop a theory to estimate the probability that the correct
+location of a read is filtered out during seeding, resulting in mapping
+errors. We describe the properties of simple exact seeds, skip-seeds and
+MEM seeds (Maximal Exact Match seeds).
 % Innovation.
@@
 % Durability.
-We provide several algorithms, which combined together give a practical
-solution to the problem of calibrating seeding heuristics for short reads.
+We provide several algorithms, which combined together give a workable
+solution for the problem of calibrating seeding heuristics for short
+reads. We also provide a C implementation of these algorithms in a library
+called Sesame.
 % Scope.
@@
 foundation of a general strategy to tackle sequence alignment problems.
+% Availability
+The Sesame library is open source and available for download at
+\url{https://github.com/gui11aume/sesame}.
 \end{abstract}
@@
 
-The rest of this document is organized to present the content in a
-didactic manner. The order of the examples is meant to facilitate the
-acquisition of the main concepts, meaning that we sometimes go back and
-forth between different types of seeding strategies. We first give some
-background on the different types of seeding methods considered here; we
-then introduce the computational strategy based on transfer matrices;
-finally we delve into each specific seeding method.
+The rest of this document is not organized by logical units but by
+increasing level of difficulty. The order of the examples is meant to
+ease the acquisition of the main concepts, meaning that we sometimes go
+back and forth between different types of seeding strategies. We first
+give some background on the different types of seeding methods considered
+here; we then introduce the computational strategy based on transfer
+matrices; finally we delve into each specific seeding method.
 
@@
 The index is usually a hash table or a variant of the so-called
-FM-index~\cite{pmid20460430}. Hash tables are typically used to index
-$k$-mers, which makes them useful to search seeds of fixed size $k$
-(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
-algorithms). In contrast, some text-indexing structures have no set size
-so they can be used to search seeds of different lengths. For instance,
-the FM-index is a compact data structure based on the suffix
-array~\cite{manber1993suffix} and on the Burrows-Wheeler
-transform~\cite{burrows1994block}, emulating a suffix trie with a much
-smaller memory footprint~\cite{ferragina2000opportunistic,
+FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}. Hash
+tables are typically used to index $k$-mers, which makes them useful to
+search seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
+benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
+structures have no set size so they can be used to search seeds of
+different lengths. For instance, the FM-index is a compact data structure
+based on the suffix array~\cite{manber1993suffix} and on the
+Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
+with a much smaller memory footprint~\cite{ferragina2000opportunistic,
 ferragina2005indexing}.
@@
 
-Using the method above, we can compute the probability that a seeding
-scheme using exact seeds is off target. This depends on the exact
-specifications of the problem, so for illustration purposes, we choose
-reads of size $k=50$ sequenced with an instrument with error rate
-$p=0.01$, and we use exact seeds of size $\gamma=17$. We then vary the
-number of duplicates $N$ and their divergence rate $\mu$, which both
-depend on the particular read being sequenced.
+We illustrate the strategy delineated above for reads of size $k=50$
+sequenced with an instrument with error rate $p=0.01$, when using exact
+seeds of size $\gamma=17$.
 
@@
 defined as the probability that a given duplicate differs from the target
-at any nucleotide.}
+at any given position.}
 \label{fig:curves_exact}
@@
 
-Fig.~\ref{fig:curves_exact} shows the result for $N$ from 1 to 10 and for
-$\mu$ from 0 to 0.20. The first observation is that the probability that
-seeding is off target increases with $N$. This is clearly visible from
-expression (\ref{eq:Poff}). This can also be understood intuitively
-because the probability of null seeding decreases when there are more
-potential candidates, and since the probability of on-target seeding does
-not change, the probability of off-target seeding must increase.
+Fig.~\ref{fig:curves_exact} shows the result for a number of duplicates
+$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
+observation is that the probability that seeding is off target increases
+with $N$. This is clearly visible from expression (\ref{eq:Poff}). This
+can also be understood intuitively because the probability of null seeding
+decreases when there are more potential candidates, and since the
+probability of on-target seeding does not change, the probability of
+off-target seeding must increase.
 
@@
 
-Using the method above, we can compute the probability that a seeding
-scheme using skip seeds is off target. Using the same settings as in
-section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), we decide to
-use skip-9 seeds and vary the number of duplicates $N$ and their
-divergence rate $\mu$.
+We illustrate the strategy delineated above Using the same settings as in
+section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), except that
+we use skip-9 seeds instead of exact seeds.
 
@@
 x-axis shows the divergence rate $\mu$, defined as the probability that a
-given duplicate differs from the target at any nucleotide.}
+given duplicate differs from the target at any given position.}
 \label{fig:curves_skip}
@@
 
-Fig.~\ref{fig:curves_skip} shows the result for $N$ from 1 to 10 and for
-$\mu$ from 0 to 0.20. The curves have the same general aspect as those of
-Fig.~\ref{fig:curves_exact}. The probability that seeding is off target
-increases with $N$. There is again a  worst value of $\mu$, because the
-maximum of $P(\overline{S}_0 \cap \overline{S}_1)$ minimizes expression
-(\ref{eq:Poff}) for every value of $N$. But the value is different from
-that of Fig.~\ref{fig:curves_exact} because $P(\overline{S}_0 \cap
-\overline{S}_1)$ is computed for skip-9 seeds instead of exact seeds (here
-the worst value of $\mu$ is approximately 0.07).
+Fig.~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
+from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
+have the same general aspect as those of Fig.~\ref{fig:curves_exact}. The
+probability that seeding is off target increases with $N$. There is again
+a  worst value of $\mu$, because the maximum of $P(\overline{S}_0 \cap
+\overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value of
+$N$. But the value is different from that of Fig.~\ref{fig:curves_exact}
+because $P(\overline{S}_0 \cap \overline{S}_1)$ is computed for skip-9
+seeds instead of exact seeds (here the worst value of $\mu$ is
+approximately 0.07).
 
@@
 
-Using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
-and $\gamma=17$), we decide to use MEM seeds and vary the number of
-duplicates $N$ and their divergence rate $\mu$.
+We illustrate the strategy delineated above 
+using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
+and $\gamma=17$), except that we replace exact seed by MEM seeds.
 
@@
 \caption{\textbf{Off target seeding probabilities (MEM seeds).}
-The curves show the probability that seeding is off target for skip-9
-seeds of size $\gamma=17$. Read size and error rate are the same as in
-Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
-shows a different number of duplicate sequences $N$ from 1 to 10 and the
-x-axis shows the divergence rate $\mu$, defined as the probability that a
-given duplicate differs from the target at any nucleotide.}
+The curves show the probability that seeding is off target for MEM
+seeds of minimum size $\gamma=17$. Read size and error rate are the same
+as in Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each
+line shows a different number of duplicate sequences $N$ from 1 to 10 and
+the x-axis shows the divergence rate $\mu$, defined as the probability
+that a given duplicate differs from the target at any given position.}
 \label{fig:curves_mem}
@@
 
-Fig.~\ref{fig:curves_mem} shows the result for $N$ from 1 to 10 and for
-$\mu$ from 0 to 0.20. The curves have the same general aspect as those of
-Fig.~\ref{fig:curves_exact}. The probability that seeding is off target
-increases with $N$, as shown by expression (\ref{eq:PMEM}).
+Fig.~\ref{fig:curves_mem} shows the result for a number of duplicates $N$
+from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
+have the same general aspect as those of Fig.~\ref{fig:curves_exact}. The
+probability that seeding is off target increases with $N$, as shown by
+expression (\ref{eq:PMEM}).
 
@@
 
+\section{The Sesame library}
+\label{sec:sesame}
+
+We implemented the methods and algorithms presented here in an open-source
+C library to compute seeding probabilities. The library is called Sesame
+and is available at \url{https://github.com/gui11aume/sesame}.
+
+\subsection{Main features of Sesame}
+
+Sesame contains functions to compute the seeding probabilities described
+here. All the functions were tested against simulations to ensure that the
+implementation is accurate and the code was checked extensively by static
+analysis and unit testing.
+
+Computating seeding probabilities can take up to a few seconds, even when
+replacing iterative methods by Monte Carlo simulations. This is
+incompatible with the speeed requirements of modern mappers, so Sesame has
+an interface for computations that must be performed online. In this mode,
+Sesame uses a type of lazy evaluation where the results are computed only
+the first time, and stored in memory for reuse on subsequent calls.
+
+Storing the results in memory is an efficient strategy because three
+parameters are constant throughout the sequencing run: the minimum seed
+size $\gamma$, the read size $k$ and the error rate of the sequencer $p$
+(and when using skip seeds, the amount of skipping $n$ is also constant).
+Only two parameters depend on the read: the number of duplicates $N$ and
+their divergence rate $\mu$. In this mode, Sesame automatically switches
+to Monte Carlo sampling when $N$ is large to save time. Also, the input
+parameters are ``snapped'' to a predefined grid of set values for $N$ and
+$\mu$, so that few computations are performed and most of the calls are
+actually memory lookups. Sesame can thus be integrated in short read
+mappers without being a bottleneck.
+
+Alternatively, the probabilities of interest can be computed offline,
+saved to disk and loaded at run time. This is particularly useful if the
+sequencing runs follow some standard conditions with a known error rate,
+because the computations can be recycled between runs.
+
+Finally, Sesame also has an offline interface, where seeding probabilities
+are computed exactly as requested by the users, \textit{i.e.} without
+modifying the algorithm or the parameters, and also without storing the
+results in memory.
+
+The Sesame manual, available from the repository, contains additional
+information and explains in detail how to use the library.
+
+\subsection{Using Sesame to compare seeding strategies}
+
+As a tool to compute seeding probabilities, Sesame can be used to compare
+the merits of different strategies. The kind of insight that we can gain
+from such calculations was already showcased in
+Fig.~\ref{fig:curves_exact}, Fig.~\ref{fig:curves_skip} and
+Fig.~\ref{fig:curves_mem}, where the numbers were computed using Sesame.
+
+To further showcase the potential benefits of computing seeding
+probabilities, we use Sesame to compare the default seeding strategies of
+BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286}. Note that
+both mappers use advanced techniques to refine the seeds, so this
+comparison does not reflect the true performance of the mappers. It is
+nevertheless useful to know the baseline of each strategy. The default of
+BWA-MEM it to use MEM seeds of minimum size 19; that of Bowtie2 is to use
+skip-9 seeds of size 16.
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[scale=.41]{mortal_kombat.pdf}
+\end{center}
+\caption{\textbf{Comparing seeding strategies with Sesame.} Left: MEM
+seeds with $\gamma=19$ and $\mu=0.06$. Right: skip-9 seeds with
+$\gamma=16$ and $\mu=0.06$. The probabilities that seeding is off target
+were computed using Sesame. Each curve represents the probability for a
+given number of duplicates ($N$). Estimates using iterative methods (skip
+seeds and MEM sees where $N \leq 20$) were computed to within 1\%
+accuracy. Estimates using Monte Carlo sampling (MEM seeds where $N > 20$)
+were computed as the average of 500 million simulations.
+}
+\label{fig_mortal_kombat}
+\end{figure*}
+
+The probabilities that seeding is off target for different read sizes $k$
+and different number of duplicates $N$ are plotted in
+Fig.~\ref{fig_mortal_kombat}. The left panel shows the results for MEM
+seeds and the right panel shows the results for skip seeds. Here the error
+rate $p$ is set to 1\%, close to the specifications the Illumina platform,
+and the divergence rate between duplicates $\mu$ is set to an arbirtarry
+value of 6\%.
+
+The behaviors of the two types of seeds are dramatically different. Let us
+start with MEM seeds. For every value of $N$, the probability initially
+increases with the read size, and then drops exponentially. The initial
+increase is a hallmark of MEM seeds; it is due to the fact that duplicates
+can mask the target. Note that the asymptotic decay depends on the number
+of duplicates $N$ because each duplicate can mask the target and thereby
+reduces the probability that it is discovered. Overall, these results show
+that the performance of MEM seeds is poor when the target has more than
+approximately 20 duplicates.
+
+Turning to skip seeds, we see that the curves have a staircase look with a
+drop every 10 nucleotides. This is so because seeding probabilities remain
+unchanged until there is space for another seed of size 16 on the read.
+The curves are otherwise decreasing with a steady exponential trend where
+the asymptotic decay does not depend on the number of duplicates $N$. The
+reason is that duplicates do not prevent the target from being discovered,
+they merely fool the mapper when the target was not found. This property
+makes the asymptotic decay of skip seeds substantially faster that of MEM
+seeds when $N$ is large.
+
+Those properties are self evident in retrospect, but they are not
+necessarily obvious from the definitions of MEM seeds and skip seeds.
+Keeping in mind that the asymptotic decay depends on $N$ for MEM seeds
+helps understand what their main limitation is.
+
+Fig.~\ref{fig_mortal_kombat} suggests that skip-9 seeds of size 16 are
+just better than MEM seeds of size 19. However, the gain in sensitivity
+comes at the cost of a larger candidate set, slowing down the mapping
+process. To remain competitive, Bowtie2 further filters the candidate set
+using a priority policy. But since some candidates are not checked, the
+probability that seeding is off target is larger than shown in
+Figure~\ref{fig_mortal_kombat}.
+
+\subsection{Key insights about MEM seeds}
+
+At least two key insights about MEM seeds can be gained from
+Fig.~\ref{fig_mortal_kombat}. The first is that it is probably not worth
+for a MEM-based mapper to check all the candidate loci when there are more
+than approximately 20 them. The mapper may find the correct location, but
+even if this is the case, the mapping quality will remain low because the
+prior chances of failure were high. A better strategy is to either bail
+out ot not waste time or to switch to a more sensitive seeding method (BWA
+opts for the second and uses a re-seeding policy). It is also important
+to note that this decision should be based on an estimated value of $N$
+and not, for instance, on the size of the seed or some other variable.
+
+A second insight is that for MEM seeds, the off-target rate is always
+above $10^{-3}$ for reads of 50 nucleotides or fewer. Here it is important
+to mention that the value $\mu=0.06$ is not even the worst for reads of
+this size when $p=0.01$ (according to Fig.~\ref{fig:curves_mem}, the worst
+value is around 0.02). So if $\mu$ is unknown and one wishes to be
+conservative, it seems that reads of 50 nucleotides cannot be mapped with
+confidence better than $1/1000$.
+
+However, this is not true for the reason that it is practically impossible
+to map an Illumina read of size 50 to the wrong location when $N=0$ (see
+section~\ref{sec:random_seeds}). This means that tweaking the seeding
+method to improve the mapping quality score will have less impact than
+checking whether $N$ is 0 (in other words, whether the target is a unique
+sequence).
+
+These insights suggest that there is a way to make MEM seeds more useful
+for short read mappers. We are currently working on a prototype based on
+these two ideas to increase the speed and the accuracy of short read
+mapping. This requires estimating $N$ and $\mu$ for every read and will be
+described in a separate work.
+
 \section{Discussion}
@@
 We have devised a set of methods to compute the probability that seeding
-heuristics fail and commit the mapping process to an error. This
-fills a knowledge gap to understand and calibrate the performance of the
-seeding heuristics. The pillars of our strategy are borrowed from analytic
-combinatorics~\cite{flajolet2009analytic, sedgewick2013introduction}, even
-though we do not follow the complete programme. Constructing generating
-functions usually serves the purpose of finding their singularities in
-order to approximate the solution. In our case, however, the weighted
-generating functions cannot be computed so this strategy is not
-applicable.
+heuristics fail and commit the mapping process to an error. We have also
+implemented the solution as an open source C library to perform the
+computations. This fills a knowledge gap to understand and calibrate the
+performance of the seeding heuristics. The pillars of our strategy are
+borrowed from analytic combinatorics~\cite{flajolet2009analytic,
+sedgewick2013introduction}, even though we do not follow the complete
+programme. Constructing generating functions usually serves the purpose of
+finding their singularities in order to approximate the solution. In our
+case, however, the weighted generating functions cannot be computed so
+this strategy is not applicable.
 
@@
 Another interesting fact is that skip seeds can have better performance
-than exact seeds in the sense that they can lower off-seeding
+than exact seeds in the sense that they can yield lower off-seeding
 probabilities. However, this always comes at the cost of accuracy because
@@
 mean that MEM seeding is a bad strategy (it is usually faster than the
-other methods), but it is a good idea to keep the performance in check and
+other methods), but it is a good idea to keep an eye on performance and
 switch method or even skip the read altogether if the chances of
@@
 by Monte Carlo sampling, it must occur at least a few times in the
-simulation. For instance, with 10 million rounds of sampling, frequencies
-around $1/1,000,000$ or lower cannot be measured accurately. When one is
-interested only in frequent events, it is thus a perfectly reasonable
-strategy. On the other hand, for $N < 20$, the probability that MEM
-seeding is null or off-target is relatively small, so we need a method
-that is accurate in this range. Fortunately, the transfer matrix method is
-fast because the dimension of the matrix $\mathring{M}_N(z)$ is small and
-the computations are not prohibitive.
+simulation. For instance, with 1 million rounds of sampling, frequencies
+around $1/100,000$ or lower cannot be measured accurately. When one is
+interested only in frequent events, it is thus a reasonable strategy. On
+the other hand, for $N < 20$, the probability that MEM seeding is null or
+off-target is relatively small, so we need a method that is accurate in
+this range. Fortunately, the transfer matrix method is fast because the
+dimension of the matrix $\mathring{M}_N(z)$ is small and the computations
+are not prohibitive.
 
@@
 Industry and Competitiveness (`Centro de Excelencia Severo Ochoa
-2013-2017', Plan Nacional BFU2012-37168), of the CERCA
+2013-2017', Plan Estatal PGC2018-099807-B-I00), of the CERCA
 Programme~/~Generalitat de Catalunya, and of the European Research Council

### 2019-12-10 make_figure_mortal_kombat.R

@@ -1,0 +1,65 @@
+mem = log10(as.matrix(read.table("results_mem.txt")))
+skip = log10(as.matrix(read.table("results_skip.txt")))
+
+k = 20:160
+ylim = c(-10,0)
+
+pdf("mortal_kombat.pdf", height=6, width=12, useDingbats=FALSE)
+
+par(mfrow=c(1,2))
+par(mar=c(3.5,3.5,1.5,1.5))
+
+plot(k, mem[1,], type="n",
+   panel.first=grid(), ylim=ylim,
+   bty="n", xaxt="n", yaxt="n", xlab="", ylab="")
+# Plot first so that it is below.
+legend(x="topright", inset=.02, bg="white", box.col="white",
+   legend="MEM seeds", text.col="grey25")
+lines(k, mem[1,], col="grey50", lwd=1.05)
+lines(k, mem[2,], col="grey45", lwd=1.10)
+lines(k, mem[3,], col="grey40", lwd=1.15)
+lines(k, mem[4,], col="grey35", lwd=1.20)
+lines(k, mem[5,], col="grey30", lwd=1.25)
+lines(k, mem[6,], col="grey25", lwd=1.30)
+lines(k, mem[7,], col="grey20", lwd=1.35)
+lines(k, mem[8,], col="grey15", lwd=1.40)
+lines(k, mem[9,], col="grey10", lwd=1.45)
+lines(k, mem[10,], col="grey5", lwd=1.50)
+
+title(xlab="Read size (nt)",
+   ylab="Probability that seeding is off target (log10)",
+   col.lab="grey25", line=2.2, cex.lab=1.2)
+axis(side=1, cex.axis=.8, col="grey50", col.axis="grey25")
+axis(side=2, cex.axis=.8, col="grey50", col.axis="grey25")
+
+legend(x="bottomleft", lwd=c(0, 1.05, 1.10, 1.15, 1.20, 1.25, 1.30,
+   1.35, 1.40, 1.45, 1.50), col=c("white", "grey50", "grey45", "grey40",
+   "grey35", "grey30", "grey25", "grey20", "grey15", "grey10", "grey5"),
+   legend=c("", "1","5","10","20","50","100","200","400", "800","1600"),
+   inset=.02, bg="white", box.col="white", text.col="grey25")
+text(x=20, y=-5.4, "Duplicates", col="grey25", adj=c(0,0))
+
+plot(k, skip[1,], type="n",
[…]

### 2020-02-20 manuscript.tex
Start to edit for revisions

@@
 
-\author[1,2]{Guillaume J. Filion}
+\author[1,2,3]{Guillaume J. Filion}
 \author[1]{Ruggero Cortini}
@@
 \affil[2]{University Pompeu Fabra (UPF), Barcelona, Spain.}
+\affil[3]{Present address: Dept. Biological Sciences, University of
+Toronto Scarborough}
 
@@
 % Durability.
-We provide several algorithms, which combined together give a workable
-solution for the problem of calibrating seeding heuristics for short
-reads. We also provide a C implementation of these algorithms in a library
-called Sesame.
+We provide several algorithms, which together give a workable solution for
+the problem of calibrating seeding heuristics for short reads. We also
+provide a C implementation of these algorithms in a library called Sesame.
 % Scope.
@@
 
-Suppose that we use and imperfect instrument to sequence a DNA fragment
-drawn at random from a known genome; how can we identify the location of
-the fragment in the genome?
-
-We will refer to this question as the \emph{true location problem}.
-It has applications in countless domains, like discovering
-disease-causing mutations; diagnosing viral, bacterial or fungal
-infections; diagnosing cancer and choosing appropriate therapies;
-selecting breeds of interest in agriculture; tracing human migrations from
-ancient bones; diagnosing genetic diseases at birth; finding compatible
-graft donors \textit{etc}.
-
-At first sight, whether a read can be mapped to its correct location seems
-to depend only on its length and on the error rate of the sequencer.
-However, there is another important factor to consider: most genomes
-contain repeated sequences, \textit{i.e.} relatively large subsequences
-that are present at more than one location. For instance, we cannot map
-the fragment with certainty if there exists an exact copy of the sequence
-somewhere else in the genome, because we cannot know which of the two
-copies was sequenced.
-
-This case is rare though. Repeated sequences are in general not identical
-but merely homologous --- meaning that their similarity is unlikely to
-occur by chance. So when the DNA fragment has one or more duplicates in
-the genome, we can still map it, as long as we can distinguish the
-duplicates. This in turn depends on their similarity with the target
-fragment and on the error rate of the sequencer.
-
-Since repeats play a central role in the problem at hand, we give the
+Say we use an imperfect instrument to sequence a small fragment of DNA. If
+we know its genome of origin, how can we find the location of the fragment
+in this genome?
+
+The advent of high-throughput sequencing has put this question at the
+center of countless applications in genetics, such as discovering
+disease-causing mutations, selecting breeds of interest in agriculture or
+tracing human migrations.
+
+We will refer to this as the \emph{true location problem}. The answer to
+the problem depends on the length of the read, on the error rate of the
+sequencer and on one key feature of the genome: its repeat structure.
+Most genomes contain repetitive sequences, \textit{i.e.}, relatively large
+subsequences that are present at multiple locations. If the sequenced DNA
+fragment comes from a repetitive sequence, it may be impossible to map the
+read with certainty.
+
+In general, repetitive sequences are not identical but merely
+homologous---meaning that their similarity is very unlikely to occur by
+chance. So if the sequenced DNA fragment originates from a repetitive
+sequence, we can map the read only if we can rule out all the candidates
+except one. This, in turn, depends on the similarity between the
+duplicates and on the error rate of the sequencer.
+
+Since repetitive sequences play a central role in the problem, we give the
 terms ``targets'' and ``duplicates'' a meaning that will facilitate the
@@
 \begin{definition}
-The target is the DNA fragment that was sequenced. Duplicates are
-sequences of the genome that are homologous to the target.
+The target is the DNA fragment that was actually sequenced. Duplicates are
+sequences of the genome that share homology with the target (in genetics
+they are often referred to as paralogs).
 \end{definition}
 
-Because sequencing errors can convert the sequence of the target into one
-of its duplicates, the \emph{true} location of a DNA fragment is not
-always the \emph{best} candidate (as measured by the identity between the
-sequence and the candidate location). This leads to the following
-question: how can we identify the \emph{best} location of the fragment in
-the genome?
-
-We will refer to this task as the \emph{best location problem}. It amounts
-to finding the optimal alignment between two sequences, and for this
-reason has received substantial attention in bioinformatics. For
-simplicity we will assume that the true location is always the best, and
-we will often confound these two ideas.
+The difficulty of the true location problem is due to sequencing errors.
+Occasionally, the sequence of a DNA fragment can be closer to one of the
+duplicates than to the target. So the \emph{true} location of a DNA
+fragment is not always the \emph{best} (as measured by the identity
+between the sequence and the candidate location). This naturally leads to
+ask how we can identify the \emph{best} location of the fragment in
+the genome.
+
+We will refer to this as the \emph{best location problem}. It amounts to
+finding the optimal alignment between two sequences, and for this reason
+has received substantial attention in bioinformatics. For the purpose of
+developing a theory rooted in statistics, our main concern is to address
+the true location problem, but for simplicity, we will assume that the
+true location is also the best. When applicable, we will clarify the
+implications of this hypothesis in the relevant sections, and we will show
+in section~\ref{sec:truevsbest} that this simplifying assumption is harmless for
+the present theory.
 
@@
 
-There exist algorithms to solve the best location
-problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
+Exact alignment algorithms were designed to solve the best location
+problem~\cite{pmid7265238, pmid5420325}, but they are too slow to process
 the large amount of data generated by modern sequencers. Instead, one uses
-heuristic methods, \textit{i.e.} algorithms that run faster, but that may
+heuristic methods, \textit{i.e.}, algorithms that run faster but may
 return an incorrect result~\cite{Waterman1984}.
 
-The most popular heuristic for the best location problem is a filtration
-method called ``seeding''. The principle is to first identify short
-regions of high similiarity between the read and the genome, and then run
-an exact alignment algorithm at the candidate locations. This is for
-instance the general strategy of the popular alignment algorithm
-BLAST~\cite{pmid2231712}.
-
-If seeding is fast, we can quickly zoom in on a small set of candidate
-locations and reduce the input size of the (slow) alignment algorithm. The
-disadvantage is that the target location is not always detected and
-included in the candidate set. When that happens, the read cannot be
-mapped to the true location because it has been filtered out at the
-seeding step.
-
-The slowness of exact alignment algorithms imposes a trade-off between
-speed and accuracy: If the seeding step is set to produce many candidates,
-it is likely to retrieve the true location. The downside is that
-processing all the candidates would take long. Conversely, if the seeding
-step is set to produce few candidates, the process would run faster but
-the target is less likely to be in the candidate set.
+The most popular heuristic for mapping DNA sequences is a filtration
+method called seeding. The principle is to first identify seeds,
+defined as short regions of high similarity between the read and the
+genome, and then use an exact alignment algorithm at the candidate
+locations to identify the optimum. This strategy was first proposed in
+FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to to search for
+homology in sequence databases.
+
+There are efficient methods to extract seeds, so it is possible to quickly
+hone in on a small set of candidates and to reduce the search space of the
+alignment algorithm. The disadvantage is that the target may not be in the
+candidate set, in which case the read cannot be mapped correctly.
+
+As a consequence, seeding methods induce a trade-off between speed and
+accuracy: If the filtration is set to produce a large candidate set, the
+target is likely to be discovered, with the downside that checking all the
+candidates with an exact alignment algorithm will take long. Conversely,
+if the filtration is set to produce a small candidate set, the process
+will run faster but the target is more likely to be missed.
 
 Importantly, the trade-off depends on the seeding method. This means that
-we can develop faster mapping algorithms at no cost for accuracy, as long
-as we can find better seeding strategies. Progress on this line of
-research has largely benefited from the improvement of computer hardware
-and from the development of indexing algorithms. Literature on this topic
-is already abundant. Some examples of seeding algorithms are given in
-references~\cite{sun2005designing, pmid11934743, xu2006optimizing,
-kucherov2005multiseed, brejova2003vector, pmid18684737, pmid15359419}.
-References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
-comparisons of different designs, and reference~\cite{navarro2001guided}
-gives a global overview of filtration methods in pattern matching.
+mapping algorithms can run faster at no cost for accuracy if we can find
+better seeding strategies. Progress on this line of research has largely
+benefited from the improvement of computer hardware and from the
+development of optimized data structures. There already exists a large
+body of literature on the design of seeding algorithms; the interested
+reader can find examples of those in references~\cite{sun2005designing,
+pmid11934743, xu2006optimizing, kucherov2005multiseed, brejova2003vector,
+pmid18684737, pmid15359419}. References~\cite{pmid16533404}
+and~\cite{pmid20460430} present high-level comparisons of different
+designs, and reference~\cite{navarro2001guided} gives a global overview of
+filtration methods in pattern matching.
 
@@
 
-Filtering methods are considered to fail whenever the target is not in the
-candidate set. For seeding heuristics, we need to distinguish two
-different kinds of failure. In the first kind, the candidate set contains 
-a duplicate of the target but does not contain the target itself; in the
-second kind, the candidate set contains neither.
-
-The distinction is important because duplicates of the target are similar
-to the read (due to their similarity to the target), so a failure of the
-first kind is difficult to distinguish from a success. A failure of the
-second kind is easier to recognize because in this case the best hit is
-not similar to the read, as we will highlight in
-section~\ref{sec:random_seeds}.
-
-Before going further, we introduce some nomenclature that will be useful
-to simplify the exposition of the theory.
+Filtering heuristics are considered to fail whenever the target is not in
+the candidate set, but here we must be more specific and distinguish two
+kinds of failure: In the first kind, the candidate set contains a
+duplicate of the target but does not contain the target itself; in the
+second kind, the candidate set contains neither the target nor any
+duplicate.
+
+The distinction is important because the duplicates of the target are
+similar to the read (due to their similarity to the target), so a failure
+of the first kind often looks like a success. In contrast, a failure of
+the second kind is easier to flag because in this case the candidates are
+not similar to the read. We will simply assume that seeding failures of
+the second kind are always detected (as explained in
+section~\ref{sec:random_seeds}), so that we can focus on the more
+difficult case of seeding failures of the first kind.
+
+Before going further, we introduce three terms that will simplify the
+exposition.
 
 \begin{definition}
-The seeding process is said to be
+The candidate set obtained at the seeding step is said to be
 \begin{enumerate}[i)]
-\item ``on target'' if the candidate set contains the target,
-\item ``off target'' if the candidate set contains a duplicate but
-not the target, 
-\item ``null'' if the candidate set contains neither.
+\item ``on target'' if it contains the target,
+\item ``off target'' if it contains a duplicate but not the target, 
+\item ``null'' if it contains neither.
 \end{enumerate}
@@
 
-If a seed-based mapping algorithm returns an incorrect location, then it
-must be the case that the best location was never tested in the
-post-filtration phase (otherwise it would have been found by the exact
-alignment algorithm).
-
-This observation is fundamental: in order to estimate the error rate of a
-mapping algorithm, we must know how often the associated seeding heuristic
-is off target. For lack of theoretical framework, this information is
-missing, meaning that current mappers are not properly calibrated.
-
-Our focus here is to develop a theory to estimate the probability that
-seeding is off target, when using the most popular seeding heuristics for
-short reads. Previous work pioneered a method to compute seeding
-probabilities but it did not distinguish off-target from null
+With our assumptions, a read is mapped to the wrong location if and only
+if the candidate set is off target. Indeed, if the candidate set is null,
+the read is not mapped, and if the candidate set is on target, the correct
+location is discovered at the alignment step. The equivalence is granted
+by the assumption that the true location is also the best, reducing the
+mapping problem to a seeding problem. That being said, the true location
+is not always the best in practical applications. We will show in
+section~\ref{sec:truevsbest} that off-target seeding is responsible for
+most of the mapping errors even without the assumption above, but for now
+we maintain the strict equivalence between mapping errors and off-target
+seeding.
+
+Focusing on popular heuristics to map short reads, our aim is to develop a
+method to estimate the probability that the candidate set is off target.
+Previous work pioneered a method to compute seeding probabilities but it
+did not distinguish off-target from null
 seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
@@
 authors investigated the reliability of mapping
-algorithms~\cite{pmid23872968}, but they focused on the probability that
-random sequences have a single hit, recognizing that solving the problem
+algorithms~\cite{pmid23872968}, but they focused on the probabilities of
+random hits, recognizing that addressing the problem of incorrect mapping
 requires taking into consideration the repeat structure of the genome.
-Here we address this problem in general terms and we provide a practical
-way to estimate the mismapping rate in complex genomes.
-
-The rest of this document is not organized by logical units but by
-increasing level of difficulty. The order of the examples is meant to
-ease the acquisition of the main concepts, meaning that we sometimes go
-back and forth between different types of seeding strategies. We first
-give some background on the different types of seeding methods considered
-here; we then introduce the computational strategy based on transfer
-matrices; finally we delve into each specific seeding method.
+
+The rest of this document is organized so as to present the concepts by
+increasing level of complexity, meaning that we sometimes go back and
+forth between different types of seeding strategies. We first provide some
+background knowledge on different types of seeding methods, we then
+introduce the computational strategy based on transfer matrices, and
+finally we delve into each specific seeding method.
 
@@
 
-The term ``seed'' changes meaning with different authors. It can designate
-a part of the read, a part of the genome, a particular sequence motif or a
-structured pattern of matches. Also, the term does not always imply an
-exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
-uses ``spaced seeds'' that tolerate mismatches. To avoid potential
-confusion, we will adopt the following convention:
+The term ``seed'' has different meanings in computational biology. It can
+designate a part of the read, a part of the genome, a particular sequence
+motif, or a structured pattern of matches. Also, a seed does not always
+refer to an exact match. For instance, the algorithm
+PatternHunter~\cite{pmid11934743} uses ``spaced seeds'' that tolerate
+mismatches. To avoid any confusion, we will adopt the convention below.
 
@@
 
-When a seed matches a particular location of the genome, we say that it is
-a ``seed for that location''. For instance, we often use expressions such
-as ``a seed for the target'' or ``a seed for a duplicate''.
+When a seed matches a given location of the genome, we say that it is a
+seed for that location. This is particularly useful in expressions such as
+``seed for the target'' or ``seed for a duplicate''.
 
 This definition presents a computational challenge: to know if a given
-subsequence is a seed we need to know if it exists somewhere in the
-genome. This is a non trivial problem in itself, but fortunately we can
-use practical methods to solve it, even when the reference genome is very
-large.
+subsequence of a read is a seed we need to know if it exists somewhere in
+the genome. This is a non trivial problem in itself, but fortunately we
+can use practical methods to solve it, even when the reference genome is
+very large.
 
@@
 tables are typically used to index $k$-mers, which makes them useful to
-search seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
+search for seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
 benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
-structures have no set size so they can be used to search seeds of
+structures have no set size so they can be used to search for seeds of
 different lengths. For instance, the FM-index is a compact data structure
@@
 suffix array~\cite{dobin2013star}) but the FM-index is currently the most
-popular choice for seeding methods. For MEM seeds, it is even the only
-practical option~\cite{pmid24336412, pmid25399029, pmid23349213,
-pmid19389736}. However, the detail is of little interest here: we simply
-assume that seeds are known at all times without ambiguity because this
-problem has several practical solutions.
+popular choice for seeding methods. For MEM seeds defined below, it is
+even the only practical option~\cite{pmid24336412, pmid25399029,
+pmid23349213, pmid19389736}. Overall, the detail is of little interest for
+the theory. We simply assume that seeds are known at all times without
+ambiguity because this problem has several practical solutions.
 
@@
 
-Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
-perfect matches of size $\gamma$ are collected to constitute the candidate
-set. This seeding heuristic was used in the first version of
-BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
-overlapping matches within the same regions.
-
-Fig.~\ref{fig:exact_seed_example} shows a read produced by an imperfect
-sequencer. The sequenced DNA fragment comes from a region of the genome
-that has three duplicates and the read has three miscalled nucleotides
-(indicated by stars).
+Exact seeds are seeds of fixed size $\gamma$. In other words, when using
+exact seeds, the candidate set consists of all the genomic locations for
+which there is a perfect match of size $\gamma$ in the read. This seeding
+heuristic was used in the first version of BLASTN~\cite{pmid2231712}, but
+it has become unpopular for producing many redundant hits within the same
+regions of the read and of the genome.
+
+Figure~\ref{fig:exact_seed_example} shows the exact seeds from an example
+a read with three miscalled nucleotides. The sequenced DNA fragment has
+three duplicates so the seeds can match four possible locations.
 
@@
 \includegraphics[scale=1]{exact_seed_example.pdf}
-\caption{\textbf{Example of exact seeds.}
+\caption{\textbf{Exact seeding.}
 A sequenced DNA fragment (Read) is shown above the actual molecule
-(Target), and above homologous sequences also present in the genome
-(Duplicates). Sequencing errors are indicated by a star (each marked
-\texttt{T} was actually an \texttt{A} that was misread). The mismatches
-between the genomic sequences and the read are indicated in bold. The
-exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
-above the read. The matching regions in the genomic sequences are
-shadowed in grey. Several overlapping seeds accumulate at the center of
-the read, which is typical for this seeding method.}
+(Target), and above three paralogs of the target (Duplicates). Sequencing
+errors are indicated by stars (all three are \texttt{A} misread as
+\texttt{T}). The mismatches between the genomic sequences and the read are
+indicated in bold. The exact seeds of size $\gamma = 7$ are indicated as
+horizontal grey lines above the read. Matching regions in the genomic
+sequences are shadowed in grey. Several overlapping seeds accumulate at
+the center of the read, which is typical of exact seeds.}
 \label{fig:exact_seed_example}
@@
 Observe that erroneous nucleotides can belong to exact seeds because they
-sometimes match one of the duplicates. For instance, the first sequencing
-error matches all the duplicates and belongs to an off-target seed.
-However, sequencing errors that are mismatches for \emph{all} the
-sequences cannot belong to a seed. This is the case of the second
-sequencing error in this example, where there is a local deficit of seeds.
-
-The middle of the read shows some clutter where consecutive seeds
-match consecutive sequences at the same location. This is typical for
-exact seeds and constitutes a nuisance for the implementation. Indeed, it
-is a waste of computer resources to discover matches in sequences that are
+sometimes match a duplicate. For instance, the first sequencing error
+matches all the duplicates and belongs to an off-target seed for the first
+duplicate. However, sequencing errors that are mismatches for \emph{all}
+the sequences cannot belong to a seed. This is the case of the second
+sequencing error in the example, creating a local deficit of seeds.
+
+Note the clutter in the middle of the read, where consecutive seeds match
+consecutive sequences at the same location. This is typical for exact
+seeds and is considered a nuisance for the implementation. Indeed, it is a
+waste of computer resources to discover matches in sequences that are
 already in the candidate set. In addition, this seeding method is not
 particularly sensitive compared to spaced seeds~\cite{pmid11934743} so it
-is used only in few specific applications. Nevertheless, it will be
-extremely useful for the development of the present theory.
+is used only in a few specific applications. Nevertheless, it will be
+useful for the development of the present theory.
 
@@
 The sequences and the annotations are the same as in
-Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
+Figure~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
 words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
-highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
+highlight the difference with Figure~\ref{fig:exact_seed_example}, the
 missing seeds are represented by dotted lines.}
@@
 
-Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
-replaced by skip-1 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
+Figure~\ref{fig:skip_seed_example} shows what happens when exact seeds are
+replaced by skip-1 seeds on the read of Figure~\ref{fig:exact_seed_example}.
 Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
@@
 A \emph{MEM seed} is simply a MEM of size size $\gamma$ or greater.
-Fig.~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
-read of Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the
+Figure~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
+read of Figure~\ref{fig:exact_seed_example}. Observe that the clutter at the
 center of the read has disappeared because consecutive matches are fused
@@
 The sequences and the annotations are the same as in
-Fig.~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
+Figure~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
 size $\gamma=7$. The clutter at the center of the read has disappeared and
@@
 on-target seed, even when changing the sensitivity parameter $\gamma$.
-Fig.~\ref{fig:full_masking_example} shows such an example. Even though
+Figure~\ref{fig:full_masking_example} shows such an example. Even though
 there is a single sequencing error, the read has no MEM seed for the
@@
 The read, the MEM seeds and the sequences are represented as in
-Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
+Figure~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
 at the bottom effectively hide the target, so it cannot be discovered.
@@
 The second counter-intuitive property is that shortening a read can
-sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
+sometimes generate a MEM seed for the target. Figure~\ref{fig:short_vs_long}
 shows an example of this case. There is no MEM seed for the target, but
@@
 The read, the MEM seeds and the sequences are represented as in
-Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
+Figure~\ref{fig:MEM_example}. There would be a good seed (shared) if the
 read were two nucleotides shorter. The true location is hidden by
@@
 read as a sequence of segments with a tail, instead of a sequence of
-nucleotides. Fig.~\ref{fig:simple} shows an example of decomposition in
+nucleotides. Figure~\ref{fig:simple} shows an example of decomposition in
 segments. On-target seeds cannot contain sequencing errors, therefore they
@@
 unique, we can again view a read as a sequence of segments with a tail.
-The example of Fig.~\ref{fig:simple} is shown again in
-Fig.~\ref{fig:skip}, where segments in the mismatch alphabet have been
+The example of Figure~\ref{fig:simple} is shown again in
+Figure~\ref{fig:skip}, where segments in the mismatch alphabet have been
 replaced by segments in the skip-3 alphabet.
@@
 \caption{\textbf{The skip encoding.}
-The read of Fig.~\ref{fig:simple} is represented in the skip-3 alphabet.
+The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
 The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
@@
 non-skipped position. For $i=0$ the next nucleotide is not skipped.
-Other features are as in Fig.~\ref{fig:simple}.}
+Other features are as in Figure~\ref{fig:simple}.}
 \label{fig:skip}
@@
 
-Fig.~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
+Figure~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
 and $-$ sequences are represented below the read, with matches represented
@@
 
-Fig.~\ref{fig:curves_exact} shows the result for a number of duplicates
+Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
 $N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
@@
 
-Fig.~\ref{fig:skip_dual} shows the read from Fig.~\ref{fig:dual}
+Figure~\ref{fig:skip_dual} shows the read from Figure~\ref{fig:dual}
 represented in the skip-3 dual encoding. It is important to note several
-differences with Fig.~\ref{fig:dual}. The first is that the symbols
+differences with Figure~\ref{fig:dual}. The first is that the symbols
 $\Downarrow_i$ are not always associated with double mismatches. For
@@
 \caption{\textbf{Example skip dual encoding.}
-The read of Fig.~\ref{fig:dual} is represented in the skip dual alphabet.
+The read of Figure~\ref{fig:dual} is represented in the skip dual alphabet.
 The vertical bars above the read indicate non-skipped positions. The
@@
 located $i$ nucleotides downstream. Other features are as in
-Fig.~\ref{fig:dual}.}
+Figure~\ref{fig:dual}.}
 \label{fig:skip_dual}
@@
 seeds of size $\gamma=17$. Read size and error rate are the same as in
-Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
+Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
 shows a different number of duplicate sequences $N$ from 1 to 10 and the
@@
 
-Fig.~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
+Figure~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
 from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
-have the same general aspect as those of Fig.~\ref{fig:curves_exact}. The
+have the same general aspect as those of Figure~\ref{fig:curves_exact}. The
 probability that seeding is off target increases with $N$. There is again
@@
 \overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value of
-$N$. But the value is different from that of Fig.~\ref{fig:curves_exact}
+$N$. But the value is different from that of Figure~\ref{fig:curves_exact}
 because $P(\overline{S}_0 \cap \overline{S}_1)$ is computed for skip-9
@@
 
-Importantly, Fig.~\ref{fig:curves_skip} reveals that skipping 9
+Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 9
 nucleotides increases the chances that seeding is off target by a factor
-approximately 4 (compare with Fig.~\ref{fig:curves_exact}). It is not
+approximately 4 (compare with Figure~\ref{fig:curves_exact}). It is not
 obvious why this should be the case: skip seeds reduce the probability of
@@
 $p=0.1$, where the ranking is inverted for the whole range of values
-considered in Fig.~\ref{fig:curves_skip}.
+considered in Figure~\ref{fig:curves_skip}.
 
@@
 
-Fig.~\ref{fig:hard_vs_soft_masks} gives a graphical intuition of hard and
+Figure~\ref{fig:hard_vs_soft_masks} gives a graphical intuition of hard and
 soft masks. It is important to bear in mind that hard and soft masks
@@
 Hard and soft masks explain the counter-intuitive properties of MEM seeds.
-For instance, in Fig.~\ref{fig:full_masking_example} the target cannot be
+For instance, in Figure~\ref{fig:full_masking_example} the target cannot be
 discovered because every nucleotide of the read has a hard mask. In
-Fig.~\ref{fig:short_vs_long}, the target could be discovered if the read
+Figure~\ref{fig:short_vs_long}, the target could be discovered if the read
 were shorter because a hard mask would turn into a soft one.
@@
 N$ where $N$ is the number of duplicates); in the symbol $\uparrow_{/i}$,
-the index $i$ is a number of nucleotides. Fig.~\ref{fig:sketch_extended}
+the index $i$ is a number of nucleotides. Figure~\ref{fig:sketch_extended}
 shows the encoding of a read in the MEM alphabet.
@@
 \caption{\textbf{The MEM encoding}.
-The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
+The read of Figure~\ref{fig:MEM_example} is represented in the MEM alphabet.
 The arrows departing from the numbers help understand their meaning. The
@@
 The grey squares in the symbolic sequences represent MEM seed matches.
-Other features are as in Fig.~\ref{fig:dual}.}
+Other features are as in Figure~\ref{fig:dual}.}
 \label{fig:sketch_extended}
@@
 target, and the bottom rows represent duplicates (using the same symbols
-as in Fig.~\ref{fig:sketch_extended}).
+as in Figure~\ref{fig:sketch_extended}).
 \begin{inset}
@@
 seeds of minimum size $\gamma=17$. Read size and error rate are the same
-as in Fig.~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each
+as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each
 line shows a different number of duplicate sequences $N$ from 1 to 10 and
@@
 
-Fig.~\ref{fig:curves_mem} shows the result for a number of duplicates $N$
+Figure~\ref{fig:curves_mem} shows the result for a number of duplicates $N$
 from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
-have the same general aspect as those of Fig.~\ref{fig:curves_exact}. The
+have the same general aspect as those of Figure~\ref{fig:curves_exact}. The
 probability that seeding is off target increases with $N$, as shown by
@@
 
-Fig.~\ref{fig:curves_mem} reveals that in this concrete case, MEM seeds
+Figure~\ref{fig:curves_mem} reveals that in this concrete case, MEM seeds
 increase the chances that seeding is off target by a factor up to 40
-compared to exact seeds (see Fig.~\ref{fig:curves_exact}). On this
+compared to exact seeds (see Figure~\ref{fig:curves_exact}). On this
 criterion, MEM seeds are always inferior to exact seeds with the same
@@
 
-We have assumed throughout that seeds can match only the target or one of
-its duplicates. In practice, seeds can match many random locations of the
+We have assumed that seeds can match only the target or one of its
+duplicates. In practice, seeds can match many random locations of the
 genome and not just the target and the duplicates.
@@
 
+\section{True versus best location}
+\label{sec:truevsbest}
+
+TODO: mention that this is MEM seeds only. What happens for other seed
+types by the way?
+
+Early in the development of the theory, we made the assumption that the
+true location is alo the best. This hypothesis was required to claim that
+a read is mapped to the true location whenever seeding is on target.
+Indeed, if the target is in the candidate set, the exact nature of the
+alignment algorithms ensures that the read will be mapped correctly. This
+assumption is thus critical because it says that all mapping errors are
+caused by off-target seeding.
+
+To know how reasonable this statement is, we have to evaluate how often
+the read is mapped incorrectly when seeding is on target. More accurately,
+we have to consider the probability that the candidate set contains the
+target plus a candidate that is more similar to the read than the target.
+
+The key insight is to realize that MEM seeds tend to exclude the target
+from the candidate set if it is not the best location. To develop an
+intuition as to why this is the case, consider a read with one sequencing
+error. If this is not the best location, then there exists a candidate
+location that is a perfect match for the read. But in that case, the read
+is covered by a single MEM seed and the target is masked so it cannot be
+in the candidate set.
+
+Using simulations, we can show that off-target seeding is much more
+frequent than mapping errors after on-target seeding (Table~XXX). In other
+words, for the vast majority of mapping errors, the target is not seeded.
+This means that, the initial assumption that the true location is always
+the best, although incorrect, does not appreciably modify the accuracy of
+the estimates for MEM seeds.
+
+For exact and skip-seeds, the conclusions are different.
+
 \section{The Sesame library}
@@
 replacing iterative methods by Monte Carlo simulations. This is
-incompatible with the speeed requirements of modern mappers, so Sesame has
+incompatible with the speed requirements of modern mappers, so Sesame has
 an interface for computations that must be performed online. In this mode,
@@
 from such calculations was already showcased in
-Fig.~\ref{fig:curves_exact}, Fig.~\ref{fig:curves_skip} and
-Fig.~\ref{fig:curves_mem}, where the numbers were computed using Sesame.
+Figure~\ref{fig:curves_exact}, Figure~\ref{fig:curves_skip} and
+Figure~\ref{fig:curves_mem}, where the numbers were computed using Sesame.
 
@@
 and different number of duplicates $N$ are plotted in
-Fig.~\ref{fig_mortal_kombat}. The left panel shows the results for MEM
+Figure~\ref{fig_mortal_kombat}. The left panel shows the results for MEM
 seeds and the right panel shows the results for skip seeds. Here the error
 rate $p$ is set to 1\%, close to the specifications the Illumina platform,
-and the divergence rate between duplicates $\mu$ is set to an arbirtarry
+and the divergence rate between duplicates $\mu$ is set to an arbitrary
 value of 6\%.
@@
 necessarily obvious from the definitions of MEM seeds and skip seeds.
-Keeping in mind that the asymptotic decay depends on $N$ for MEM seeds
-helps understand what their main limitation is.
-
-Fig.~\ref{fig_mortal_kombat} suggests that skip-9 seeds of size 16 are
+The main limitations of the MEM seeds are best understood by keeping in
+mind that their asymptotic decay bepends on $N$.
+
+Figure~\ref{fig_mortal_kombat} suggests that skip-9 seeds of size 16 are
 just better than MEM seeds of size 19. However, the gain in sensitivity
@@
 At least two key insights about MEM seeds can be gained from
-Fig.~\ref{fig_mortal_kombat}. The first is that it is probably not worth
+Figure~\ref{fig_mortal_kombat}. The first is that it is probably not worth
 for a MEM-based mapper to check all the candidate loci when there are more
@@
 to mention that the value $\mu=0.06$ is not even the worst for reads of
-this size when $p=0.01$ (according to Fig.~\ref{fig:curves_mem}, the worst
+this size when $p=0.01$ (according to Figure~\ref{fig:curves_mem}, the worst
 value is around 0.02). So if $\mu$ is unknown and one wishes to be

### 2020-02-22 manuscript.tex
Investigate the true vs best location

@@
 
-TODO: mention that this is MEM seeds only. What happens for other seed
-types by the way?
-
-Early in the development of the theory, we made the assumption that the
-true location is alo the best. This hypothesis was required to claim that
-a read is mapped to the true location whenever seeding is on target.
-Indeed, if the target is in the candidate set, the exact nature of the
-alignment algorithms ensures that the read will be mapped correctly. This
-assumption is thus critical because it says that all mapping errors are
-caused by off-target seeding.
-
-To know how reasonable this statement is, we have to evaluate how often
-the read is mapped incorrectly when seeding is on target. More accurately,
-we have to consider the probability that the candidate set contains the
-target plus a candidate that is more similar to the read than the target.
-
-The key insight is to realize that MEM seeds tend to exclude the target
-from the candidate set if it is not the best location. To develop an
-intuition as to why this is the case, consider a read with one sequencing
-error. If this is not the best location, then there exists a candidate
-location that is a perfect match for the read. But in that case, the read
-is covered by a single MEM seed and the target is masked so it cannot be
-in the candidate set.
-
-Using simulations, we can show that off-target seeding is much more
-frequent than mapping errors after on-target seeding (Table~XXX). In other
-words, for the vast majority of mapping errors, the target is not seeded.
-This means that, the initial assumption that the true location is always
-the best, although incorrect, does not appreciably modify the accuracy of
-the estimates for MEM seeds.
-
-For exact and skip-seeds, the conclusions are different.
+Early in the development of the theory, we distinguished the true location
+from the best location, but we assumed that they are identical in order to
+eliminate some practical considerations that would otherwise clutter the
+exposition. Here we have presented a framework to compute the probability
+that the \emph{true} location is in the candidate set. We have not said
+anything about the probability that the best location is in the candidate
+set, so our results do not help address the best location problem. Now it
+remains to see whether they help address the true location problem.
+
+The issue at hand is that the candidates are tested with an alignment
+algorithm that returns the best location. So even if the target is in the
+candidate set, the read may be mapped to another location. The question is
+thus how often this happens. If this case is rare, our results give a good
+approximation of the mapping errors (where errors are failures to map the
+reads to their true location). Otherwise the estimates may not be so
+useful in practice.
+
+The answer depends on the seed type. We start with MEM seeds because they
+will allow us to better explain what happens. The key insight is that MEM
+seeds tend to exclude the target from the candidate set if it is not the
+best location. To develop an intuition as to why this is the case,
+consider a read with a single sequencing error. We claim that the target
+was never in the candidate set if the read is mapped to an incorrect
+location (\textit{i.e.} not the true location).
+
+The argument goes as follows: if the target were in the candidate set,
+then there would exist a candidate that is a perfect match for the read.
+But then the read would be covered by a single MEM seed, the target would
+be masked and it could not be in the candidate set, leading to a
+contradiction.
+
+\begin{table}
+% MEM seeds.
+\renewcommand{\arraystretch}{1.1}
+\centering
+\begin{tabular}{ccccc}
+k & N & Prob. seeding error & Prob. other error & \% seeding \\
+\hline
+50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
+50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
+50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
+100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
+100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
+100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
+\end{tabular}
+\caption{My favorite table. $p = 0.01$ and $\mu = 0.06$.}
+\label{table_mem}
+\end{table}
+
+In short, either seeding is on target and the read is mapped to the true
+location, or seeding is off target (and the read is of course mapped
+somewhere else). When there are more than one errors, the strict
+equivalence breaks, but the trend remains the same. We can show this
+using simulations for realistic sequencing conditions
+(Table~\ref{table_mem}). For MEM seeds, the probability that the read is
+not mapped to the true location is approximately the same as the
+probability that the seeding is off target.
+
+For exact and skip-seeds, the conclusions are different. The situation is
+even the opposite if we consider reads with a single sequencing error. In
+this case the target is always in the candidate set (if the read size is
+greater than $2\gamma$, where $\gamma$ is the minimum seed length), so
+mapping errors are never due to failures of the seeding heuristic.
+Instead, mapping errors happen only when the true location is not the
+best.
+
+The trend can reverse for a high number of errors (most of the errors may
+be due to seeding errors), but because these reads are rare, the mapping
+errors are in general not due to failures of the seeding heuristic.
+
+\begin{table}
+% Exact seeds.
+\renewcommand{\arraystretch}{1.1}
+\centering
+\begin{tabular}{ccccc}
+k & N & Prob. seeding error & Prob. other error & \% seeding \\
+\hline
+50  &  1  & $3.3 \cdot 10^{-5}$  & $5.1 \cdot 10^{-4}$ & $6.0\%$ \\
+50  & 10  & $2.9 \cdot 10^{-4}$  & $5.0 \cdot 10^{-3}$ & $5.4\%$ \\
+50  & 100 & $1.0 \cdot 10^{-3}$  & $4.6 \cdot 10^{-2}$ & $2.2\%$ \\
+100 &  1  & $1.5 \cdot 10^{-10}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
+100 & 10  & $1.2 \cdot 10^{-9}$  & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
+100 & 100 & $2.6 \cdot 10^{-9}$  & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
+\end{tabular}
+\caption{My least favorite table.}
+\label{table_exact}
+\end{table}
+
+\begin{table}
+% Skip-9 seeds.
+\renewcommand{\arraystretch}{1.1}
+\centering
+\begin{tabular}{ccccc}
+k & N & Prob. seeding error & Prob. other error & \% seeding \\
+\hline
+50  &  1  & $1.3 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $20.1\%$ \\
+50  & 10  & $1.2 \cdot 10^{-3}$ & $4.8 \cdot 10^{-3}$ & $19.7\%$ \\
+50  & 100 & $5.5 \cdot 10^{-3}$ & $4.5 \cdot 10^{-2}$ & $10.9\%$ \\
+100 &  1  & $2.0 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
+100 & 10  & $1.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
+100 & 100 & $3.9 \cdot 10^{-7}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
+\end{tabular}
+\caption{My not so favorite table.}
+\label{table_skip}
+\end{table}
+
+Tables~\ref{table_exact} and \ref{table_skip} shows the result of numeric
+simulations for exact and skip-9 seeds, respectively. It appears that in
+these cases, the probability that probability of mapping the read to an
+incorrect location (\textit{i.e.} not the true location) is substantially
+higher than the probability that seeding is off target. As a consequence,
+seeding probabilities are not very useful to address true mapping problem
+when exact and skip seeds are used. In this case, it is more accurate to
+use the probabilities that the true location is not the best.
 
@@
 to mention that the value $\mu=0.06$ is not even the worst for reads of
-this size when $p=0.01$ (according to Figure~\ref{fig:curves_mem}, the worst
-value is around 0.02). So if $\mu$ is unknown and one wishes to be
+this size when $p=0.01$ (according to Figure~\ref{fig:curves_mem}, the
+worst value is around 0.02). So if $\mu$ is unknown and one wishes to be
 conservative, it seems that reads of 50 nucleotides cannot be mapped with

### 2020-02-26 manuscript.tex
Add appendix with proofs

@@
 properties. The first is that there are cases where there cannot be any
-on-target seed, even when changing the sensitivity parameter $\gamma$.
+on-target seed, even when changing the minimum seed size $\gamma$.
 Figure~\ref{fig:full_masking_example} shows such an example. Even though
@@
 \label{def:seg}
-Every symbol of the alphabet that is different from $\square$ is called a
-terminator. A segment is a terminator preceded by 0 or more symbols $\square$.
-The last segment of the read is terminated by the symbol $|$ and is called the
-tail.
+A terminator is any symbol that is different from the symbol $\square$. A
+segment is a sequence of 0 or more symbols $\square$ followed by a
+terminator. The tail is the last segment of the read, where the terminator
+is always the special symbol $|$.
 \end{definition}
@@
 segments. On-target seeds cannot contain sequencing errors, therefore they
-must be completely embedded inside a segment. So the sizes of the segments
+must be completely embedded in a segment. So the sizes of the segments
 indicate whether the read contains an on-target seed or not.
@@
 The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
-rate of the sequencer) and the probability of occurrence of the symbol
-$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
+rate of the sequencer) so the probability of occurrence of the symbol
+$\square$ is $1-p = q$. Both symbols have size 1, so their respective
 weighted generating function are $pz$ and $qz$. Using the rule for
-concatenation, we see that the weighted generating function of a segment
-with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
-the weighted generating function of a tail segment with $i$ symbols
-$\square$ is $(qz)^i$.
-
-The key insight is that reads without on-target seed are exactly the reads
-that are made of segments with fewer than $\gamma$ symbols $\square$. The
-weighted generating function of such segments is $pz \cdot \big(
-1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
-$1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
-encoded in a transfer matrix.
-
-There are two kinds of objects: the $\Downarrow$ segments and the tails,
-so the dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$
-segment can be followed by another $\Downarrow$ segment or by the tail.
-The tail cannot be followed by anything. The expression of the transfer
-matrix $M_0(z)$ is thus
+concatenation, we see that a segment of $i$ symbols $\square$ followed by
+a terminator has weighted generating function $(qz)^ipz$. The final symbol
+$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
+symbol $|$ has weighted generating function $(qz)^i$.
+
+The key insight is that the reads without on-target seed are exactly the
+reads that are made of segments with fewer than $\gamma$ symbols
+$\square$, where $\gamma$ is the minimum seed size. The weighted
+generating function of such segments is $\big( 1+qz+\ldots+(qz)^{\gamma-1}
+\big) pz$, and that of the tail is $1+qz+\ldots+(qz)^{\gamma-1}$. This
+gives a construction plan that can be encoded in a transfer matrix.
+
+Reads consist of only two kinds of objects: the segments terminated by
+$\Downarrow$ (or $\Downarrow$-segments for short) and the tails, so the
+dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$-segment
+can be followed by another $\Downarrow$-segment or by the tail. The tail
+cannot be followed by anything. The expression of the transfer matrix
+$M_0(z)$ is thus
 \begin{equation*}
@@
 \begin{block}{c[cc]}
-\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
-\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
+\scriptstyle{\Downarrow} & \big( 1+qz+\ldots+(qz)^{\gamma-1}
+\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
 \scriptstyle{|} & 0 & 0 \\
@@
 
-In the representation above, the different types of objects are identified
-by their terminator. The terminators are written in the margins of the
-transfer matrix for readability.
+In the representation above, the different types of segments are
+identified by their terminator. The terminators are written in the margins
+of the transfer matrix for readability.
 
@@
 (I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
-term --- associated with terminators $\Downarrow$ and $|$. To see why,
-observe that every read can be prepended by $\Downarrow$ segments and only
+term---associated with terminators $\Downarrow$ and $|$. To see why,
+observe that every read can be prepended by $\Downarrow$-segments and only
 by those (not by a tail). Thus, reads are precisely the sequences of
@@
 by a tail, whose weighted generating function is the top right entry of
-the matrix. It is easy to check that this term is equal to
+the matrix. In appendix~\ref{app_ex1}, we show that this term is equal to
 \begin{equation}
@@
 \frac{1+qz+\ldots+(qz)^{\gamma-1}}
-  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
+  {1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz}.
 \end{equation}
@@
 construction, $a_k$ is the quantity of interest, \textit{i.e.} it is the
-probability that a read of size $k$ does not contain an on-target seed, so
-we now need to extract the coefficients from expression (\ref{eq:F}).
-There are several methods to do so; the one we choose here is to build a
-recurrence equation to compute $a_k$ from the previous terms of the series
-$a_0, a_1, \ldots, a_{k-1}$. For this, we observe that for $z \neq 1/q$,
-equation
-\begin{equation*}
-\frac{1+qz+\ldots+(qz)^{\gamma-1}}
-  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} =
-  a_0 + a_1z +a_2z^2 + \ldots
-\end{equation*}
-is equivalent to
-\begin{equation*}
-(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
-(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
-\end{equation*}
-Balancing the terms of same degree on both sides of the equation leads
-to the following relationship between the coefficients of the series:
+probability that a read of size $k$ does not contain an on-target seed of
+minimum size $\gamma$, so we now need to extract the Taylor coefficients
+from expression (\ref{eq:F}). There are several methods to do so; the one
+we choose here is to build a recurrence equation. In
+appendix~\ref{app_rec}, we show that
+
 \begin{equation}
@@
 
-These results are consistent: for $k < \gamma$, the read is too short so
-the probability that it contains no seed is 1; for $k = \gamma$, the
-read contains a seed if and only if it has no error, which occurs with
-probability $q^\gamma$.
+Note that for $k < \gamma$, the read is too short so the probability that
+it contains no seed is 1; for $k = \gamma$, the read contains a seed if
+and only if it has no error, which occurs with probability $q^\gamma$.
 
@@
 \begin{tabular}{ccccc}
-k & N & Prob. seeding error & Prob. other error & \% seeding \\
+k & N & Not seeded & Seeded, not best & \% seed failure \\
 \hline
@@
 \end{tabular}
-\caption{My favorite table. $p = 0.01$ and $\mu = 0.06$.}
+\caption{Numerical simulating of mapping errors with MEM seeds. Reads are
+mapped incorrectly if either the true location is not in the candidate set
+(not seeded) or it is in the candidate set but is not the best location
+(seeded, not best). Ideal reads were randomly generated 1,000,000 times
+with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
+varying the read size $k$ and number $N$ of duplicates as indicated in the
+table. Mapping errors of both types were recorded, and the fraction of the
+first to the total was computed.}
 \label{table_mem}
@@
 when exact and skip seeds are used. In this case, it is more accurate to
-use the probabilities that the true location is not the best.
+use the probabilities that the true location is not the best, because
+they are in general much higher.
+
+In conclusion, the true location problem can be calibrated using the
+methods presented here when using MEM seeds, and by computing the
+probability that the true location is not the best when using exact or
+skip-seeds.
 
@@
 
+\appendix
+\section{Appendix}
+\subsection{Derivation of expression~(\ref{eq:F})}
+\label{app_ex1}
+
+In section~\ref{sec:example_exact}, we had to compute the top right entry
+of the matrix $M_*(z)$ defined as $M_*(z) = M_0(z) + M_0(z)^2 + \ldots =
+M_0(z) \cdot (I-M_0(z))^{-1}$, where
+\begin{equation*}
+M_0(z) =
+\begin{bmatrix}
+\big( 1+qz+\ldots+(qz)^{\gamma-1}
+\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
+0 & 0 \\
+\end{bmatrix}.
+\end{equation*}
+
+Since $I-M_0(z)$ is a $2\times2$ matrix, we can invert it analytically.
+Observing that the determinant of $I-M_0(z)$ is $\Delta(z) = 1-\big(
+1+qz+\ldots+(qz)^{\gamma-1}\big) pz$, we find
+\begin{equation*}
+(I-M_0(z))^{-1} = \frac{1}{\Delta(z)}
+\begin{bmatrix}
+1 & 1+qz+\ldots+(qz)^{\gamma-1} \\
+0 & 1-\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz
+\end{bmatrix}.
+\end{equation*}
+
+Premultiplying by $M_0(z)$, we finally obtain
+\begin{equation*}
+M_*(z) = \frac{1}{\Delta(z)}
+\begin{bmatrix}
+\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz &
+1+qz+\ldots+(qz)^{\gamma-1} \\
+0 & 0 \end{bmatrix},
+\end{equation*}
+where the top right term is equal to
+\begin{equation}
+\tag{\ref{eq:F}}
+\frac{1+qz+\ldots+(qz)^{\gamma-1}}
+{1-\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz}.
+\end{equation}
+
+\subsection{Derivation of expression~(\ref{eq:recur})}
+\label{app_rec}
+
+In section~\ref{sec:example_exact}, we had to extract the Taylor
+coefficients $a_0, a_1, a_2 \ldots$ of the function
+\begin{equation*}
+\frac{1+qz+\ldots+(qz)^{\gamma-1}}
+  {1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz} =
+  a_0 + a_1z +a_2z^2 + \ldots
+\end{equation*}
+
+Multipying both terms of the equation by $1-pz
+\big(1+qz+\ldots+(qz)^{\gamma-1} \big)$ we obtain
+\begin{equation*}
+1+qz+\ldots+(qz)^{\gamma-1} =
+\left(1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz\right)
+(a_0 + a_1z +a_2z^2 + \ldots).
+\end{equation*}
+
+Multiplying again both terms of the equation by $1-qz$, we find
+\begin{eqnarray*}
+1-(qz)^\gamma &=& (1-qz) \left(1- \big(1+qz+\ldots+(qz)^{\gamma-1}
+\big) pz\right) (a_0 + a_1z +a_2z^2 + \ldots) \\
+&=& (1-z+pq^\gamma \cdot z^{\gamma+1}) (a_0 + a_1z +a_2z^2 + \ldots) \\
+&=& a_0 + \sum_{k=1}^{\gamma} (a_k-a_{k-1})z^k +
+  \sum_{k=\gamma+1}^{\infty} (a_k-a_{k-1}+ pq^\gamma \cdot
+a_{k-\gamma-1})z^k.
+\end{eqnarray*}
+
+Balancing the terms of degree 0 on both sides of the equation forces $a_0
+= 1$. The terms of degree 1 to degree $\gamma-1$ are null on the left
+side, which implies $a_k - a_{k-1} = 0$ and consequently $a_k = a_0 = 1$
+for $k = 1, 2, \ldots, \gamma-1$. The term of degree $\gamma$ is
+$-q^\gamma$ on the left side, so $a_\gamma-a_{\gamma-1} = -q^\gamma$, or
+$a_\gamma = 1-q^\gamma$. Finally, all the other terms on the left side
+are null, which implies $a_k-a_{k-1}+ pq^\gamma \cdot a_{k-\gamma-1} = 0$,
+or $a_k = a_{k-1}-pq^\gamma\cdot a_{k-\gamma-1}$ for $k = \gamma+1,
+\gamma+2 \ldots$
+
+In comclusion, the Taylor coefficients $a_0, a_1, a_2, \ldots$ satisfy the
+recurrence equation
+\begin{equation}
+\tag{\ref{eq:recur}}
+a_k = 
+\begin{cases}
+1            &\quad\text{if } k < \gamma, \\
+1 -q^\gamma &\quad\text{if } k = \gamma, \\
+a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
+\end{cases}
+\end{equation}
+
 \end{document}

### 2020-02-27 manuscript.tex
Add explanations for skip seeds

@@
 location of a read is filtered out during seeding, resulting in mapping
-errors. We describe the properties of simple exact seeds, skip-seeds and
+errors. We describe the properties of simple exact seeds, skip seeds and
 MEM seeds (Maximal Exact Match seeds).
@@
 A terminator is any symbol that is different from the symbol $\square$. A
-segment is a sequence of 0 or more symbols $\square$ followed by a
+segment is a sequence of 0 or more $\square$ symbols followed by a
 terminator. The tail is the last segment of the read, where the terminator
@@
 The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
-rate of the sequencer) so the probability of occurrence of the symbol
-$\square$ is $1-p = q$. Both symbols have size 1, so their respective
+rate of the sequencer) so the probability of occurrence of the $\square$
+symbol is $1-p = q$. Both symbols have size 1, so their respective
 weighted generating function are $pz$ and $qz$. Using the rule for
@@
 In the representation above, the different types of segments are
-identified by their terminator. The terminators are written in the margins
-of the transfer matrix for readability.
+identified by their terminator, indicated in the margins for clarity.
 
@@
 
-As in the previous example, we need to know whether a nucleotide is a
-sequencing error, but this time we also need to know its phase in the
-repeated cycles of skipped positions. For this, we define the skip-$n$
-alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
-\ldots, \Downarrow_n \}$. Again, the symbol $\square$ represents a correct
-nucleotide and the symbol $|$ is a terminator added at the end of the
-read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
-errors and $i$ indicates the number of nucleotides till the next
-non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
-before a non-skipped position and $i=n$ for nucleotides at a non-skipped
-position).
+We need to know whether a nucleotide is a sequencing error, but this time
+we also need to know its phase in the repeated cycles of skipped
+positions. For this, we define the skip-$n$ alphabet $\mathcal{A}_n =
+\{\square, |, \Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$. Again,
+the symbol $\square$ represents a correct nucleotide and the symbol $|$ is
+a terminator added at the end of the read. The symbols $\Downarrow_j$ ($j
+\in \mathbb{N}$) represent sequencing errors and $j$ indicates the number
+of nucleotides till the next non-skipped position (\textit{i.e.} $j=0$ for
+nucleotides immediately before a non-skipped position and $j=n$ for
+nucleotides at a non-skipped position).
 
 As per definition~\ref{def:seg}, segments in this alphabet are sequences
-of 0 or more symbols $\square$ followed by any of the symbols
-$\Downarrow_i$ or by the symbol $|$. Given that this decomposition is
+of 0 or more $\square$ symbols followed by any of the symbols
+$\Downarrow_j$ or by the symbol $|$. Given that this decomposition is
 unique, we can again view a read as a sequence of segments with a tail.
@@
 The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
-The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
-target (they are erroneous nucleotides) and the symbol $\square$
+The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
+target (they are erroneous nucleotides) and the $\square$ symbol
 represents a match (it is a correct nucleotide). The vertical bars
 indicate non-skipped positions (the potential start of a seed). The number
-$i$ in $\Downarrow_i$ indicates the number of nucleotides till the next
-non-skipped position. For $i=0$ the next nucleotide is not skipped.
+$j$ in $\Downarrow_j$ indicates the number of nucleotides till the next
+non-skipped position. For $j=0$ the next nucleotide is not skipped.
 Other features are as in Figure~\ref{fig:simple}.}
@@
 The probability of occurrence of a sequencing error is $p$, so every
-symbol $\Downarrow_i$ has the same weighted generating function $pz$ ---
-implicitly assuming that the next non-skipped position is at distance $i$,
+symbol $\Downarrow_j$ has the same weighted generating function
+$pz$---provided the next non-skipped position is at distance $j$,
 otherwise the weighted generating function is 0. The weighted generating
-function of the symbol $\square$ is again $qz$, and so the weighted
-generating function of a segment with $i$ symbols $\square$ is $(qz)^ipz$.
-Likewise, the weighted generating function of a tail with $i$ symbols
-$\square$ is $(qz)^i$.
-
-Reads that do not contain any on-target skip seed can contain segments
-with $\gamma$ or more symbols $\square$, so this case is more complex than
-the previous one. For instance, if there is a sequencing error $i$
-nucleotides before the next non-skipped position, the status of the next
-$i$ nucleotides does not have any influence on the presence of an
-on-target seed. Indeed, there is no on-target seed upstream of the error
-(by construction) and the next seed is scheduled to start $i$ nucleotides
-downstream, so there can be up to $\gamma+i-1$ symbols $\square$ in a row.
-In order to enforce the absence of on-target seeds, we thus have to adjust
-the maximum size of the segments depending on the preceding terminator.
-
-In the skip-$n$ alphabet, reads are composed of segments terminated by the
-$n+1$ symbols $\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n$, plus the
-tail. The dimensions of the transfer matrix are thus $(n+2) \times (n+2)$.
-
-As mentioned above, after the terminator $\Downarrow_i$, we can append
-segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
-those $\gamma+i$ possible segments are distributed according to the laws
-of the arithmetic modulo $n+1$. When $i > 0$, for instance, if the next
-segment has no symbol $\square$ then it is terminated by
-$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
-$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
-involves a substantial amount of bookkeeping. One can check that the
-expression of the transfer matrix $M_n(z)$ is
+function of the symbol $\square$ is again $qz$, so a segment with $i$
+symbols $\square$ followed by the symbol $\Downarrow_j$ has weighted
+generating function $(qz)^ipz$---if the next non-skipped position is at
+distance $j$, otherwise it is 0. As in the previous example, a tail
+segment with $i$ symbols $\square$ followed by the symbol $|$ has weighted
+generating function $(qz)^i$.
+
+This case is more complex than the previous one: reads without on-target
+skip seed of minimum size $\gamma$ can contain segments with $\gamma$ or
+more $\square$ symbols. For instance, the read shown in
+Figure~\ref{fig:skip} contains a stretch of 9 nucleotides without errors
+but it has no seed of minimum size $\gamma=9$. More generally, if there is
+a sequencing error $i$ nucleotides before the next non-skipped position,
+there can be up to $\gamma+i-1$ symbols $\square$ in a row.
+
+Expressed in different words, it is possible to append segments with up to
+$\gamma+i-1$ symbols $\square$ after segments terminated by $\Downarrow_i$
+($\Downarrow_i$-segments for short). Each of those $\gamma+i$ possible
+segments is associated with a different terminator, depending on how far
+ahead the next non-skipped position lies. In Figure~\ref{fig:skip}, for
+instance, the second segment is terminated by $\Downarrow_1$ because there
+is 1 nucleotide before the next non-skipped position. If the segment were
+1 nucleotide longer, the terminator would have to be $\Downarrow_0$.
+
+The main issue is that segments of different lengths can be terminated by
+the same symbol. Going back to Figure~\ref{fig:skip}, the third segment
+has length 10 and is terminated by $\Downarrow_3$. It would also be
+terminated by $\Downarrow_3$ if it had length 2 or length 6. In the
+general case, the entries of the transfer matrix show some periodicity
+modulo $n+1$.
+
+Denote $H_{i,j}(z)$ the weighted generating function of
+$\Downarrow_j$-segments that can follow a $\Downarrow_i$-segment. The
+total number of segments that can follow a $\Downarrow_i$-segment is
+$\gamma+i$. Among them, the shortest $\Downarrow_j$-segment has size
+$\ell_0$ to be determined below, and the others have sizes $\ell_0+(n+1),
+\ell_0+2(n+1), \ldots, \ell_0+m(n+1)$, for some integer $m$. This $m$ is
+the largest number such that $\ell_0+m(n+1) \leq \gamma+i$, so $m =
+\lfloor (\gamma+i-\ell_0)/(n+1)\rfloor$, where $\lfloor \ldots \rfloor$ is
+the ``floor'' function.
+
+Those segments follow a $\Downarrow_i$-segment, so they start $i$
+nucleotides in front of the next non-skipped position. The shortest
+following segment that ends $j$ nucleotides before a non-skipped position
+has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
+equivalent to defining $x$ as the number of $\square$ symbols of the
+shortest $\Downarrow_j$-segment, with $x = i-j-1 \modulo{n+1}$ and $\ell_0
+= x+1$.
+
+The shortest $\Downarrow_j$-segment consists of $x$ symbols $\square$
+followed by the $\Downarrow_j$ terminator, so it has weighted generating
+function $(qz)^xpz$. The other segments have a multiple of $n+1$
+additional $\square$ symbols so their weighted generating functions are
+$(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots, (qz)^{x+m(n+1)}pz$. Summing
+over those cases, we finally obtain
+\begin{eqnarray}
+H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
+  \ldots + (qz)^{m(n+1)} \big), \\
+\notag
+  &\;& \text{where } x = i-j-1 \modulo{n+1},
+  \text{ and } m = \left\lfloor
+  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
+\end{eqnarray}
+
+Denote $J_i(z)$ the weighted generating function of tails that can follow
+a $\Downarrow_i$-segment. There are $\gamma+i$ of them, each consisting of
+0 to $\gamma+i-1$ symbols $\square$ followed by the special $|$
+terminator for the end of the read. The $|$ has size 0 so its weighted
+generating function is 1. Once again summing over the different cases, we
+obtain
+\begin{eqnarray}
+J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
+\end{eqnarray}
+
+Finally, the expression of $M_n(z)$ the transfer matrix of reads in the
+skip-$n$ alphabet is
 \begin{equation*}
@@
 \end{block}
-\end{blockarray},
+\end{blockarray}.
 \end{equation*}
-\begin{eqnarray}
-H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
-  \ldots + (qz)^{m(n+1)} \big), \\
-\notag
-  &\;& \text{where } x = i-j-1 \modulo{n+1},
-  \text{ and } m = \left\lfloor
-  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
-J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
-\end{eqnarray}
-where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.
 
@@
 why, observe that, at the start of every read, the next nucleotide is a
-non-skipped position, so every read can be prepended by $\Downarrow_0$
-segments and only by those. Thus, reads are precisely the sequences of
-segments that can follow the symbol $\Downarrow_0$ and that are terminated
-by a tail, whose weighted generating function is the entry of the matrix
-associated with terminators $\Downarrow_0$ and $|$.
-
-$M_n(z)$ is too complex to allow a closed expression of $M_n(z) \cdot
-(I-M_n(z))^{-1}$ to be computed. We return to the definition $M_n(z) +
-M_n(z)^2 + \ldots$ and observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
-\ldots$ have no influence on the coefficients of the weighted generating
-function $a_0, a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments (including the
-obligatory tail). Since $M(z)^{s+1}$ contains the weighted generating
-functions of reads with exactly $s+1$ segments including the tail, $a_k$
-cannot depend on $M(z)^{k+2}, M(z)^{k+3}, \ldots$. More formally, we can
-prove by induction that all the entries of $M(z)^k$ are divisible by
-$z^{k-1}$, showing that the contribution of $M(z)^{k+2} + M(z)^{k+3} +
-\ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is strictly $0$.
-
-So we can compute the matrix $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$,
-extract the entry of interest and then compute the terms of the Taylor
-expansion up to order $k$. But we can do better: since we are only
-interested in the coefficients up to order $k$, we can perform all
-algebraic operations on truncated polynomials of order $k$, \textit{i.e.}
-we discard the coefficients of order $k+1$ or greater when multiplying two
-polynomials.
+non-skipped position, so every read can be prepended by
+$\Downarrow_0$-segments and only by those. Thus, reads are precisely the
+sequences of segments that can follow the symbol $\Downarrow_0$ and that
+are terminated by a tail, whose weighted generating function is the entry
+of the matrix associated with terminators $\Downarrow_0$ and $|$.
+
+By construction, the Taylor expansion of the top right term of the matrix
+$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the the probabilities of interest.
+More specifically, $a_k$, the coefficient of $z^k$ in the series is the
+probability that a read of size $k$ contains no skip-$n$ seed of minimum
+size $\gamma$.
+
+But $M_n(z)$ is too complex to find a closed expression for $M_n(z) \cdot
+(I-M_n(z))^{-1}$. Instead, we return to the definition $M_n(z) + M_n(z)^2
++ \ldots$ and show in appendix~\ref{app_skip} that the coefficient of
+interest, $a_k$, only depends on the first $k+1$ terms of the sum. So for
+reads of size $k$ or lower, we only need to compute the matrix $M_n(z) +
+M_n(z)^2 + \ldots + M_n(z)^{k+1}$ and work out the Taylor expansion of the
+top right term.
+
+But we can do better: since we are only interested in the coefficients up
+to order $k$, we can perform all algebraic operations on truncated
+polynomials of order $k$, \textit{i.e.} we discard the coefficients of
+order $k+1$ or greater when multiplying two polynomials.
 
@@
 
-For exact and skip-seeds, the conclusions are different. The situation is
+For exact and skip seeds, the conclusions are different. The situation is
 even the opposite if we consider reads with a single sequencing error. In
@@
 probability that the true location is not the best when using exact or
-skip-seeds.
+skip seeds.
 
@@
 
+\subsection{Derivation of something important}
+\label{app_skip}
+
+Easy.
+
+observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
+\ldots$ have no influence on the coefficients of the weighted generating
+function $a_0, a_1, \ldots, a_k$.
+
+Indeed, a read of size $k$ has at most $k+1$ segments (including the
+obligatory tail, which can have size 0). Since $M(z)^{s+1}$ contains the
+weighted generating functions of reads with exactly $s+1$ segments
+including the tail, $a_k$ cannot depend on $M(z)^{k+2}, M(z)^{k+3},
+\ldots$ More formally, we can prove by induction that all the entries of
+$M(z)^k$ are divisible by $z^{k-1}$, showing that the contribution of
+$M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is
+strictly $0$.
+
 \end{document}

### 2020-03-09 manuscript.tex
Add section "Practical considerations"

@@
 the symbol $\square$ represents a correct nucleotide and the symbol $|$ is
-a terminator added at the end of the read. The symbols $\Downarrow_j$ ($j
-\in \mathbb{N}$) represent sequencing errors and $j$ indicates the number
+a terminator added at the end of the read. The symbols $\Downarrow_j$ ($ 0
+\leq j \leq n$) represent sequencing errors and $j$ indicates the number
 of nucleotides till the next non-skipped position (\textit{i.e.} $j=0$ for
@@
 skip-$n$ alphabet is
-\begin{equation*}
+\begin{equation}
+\label{eq_matrix_Mn}
 \begin{blockarray}{cccccc}
@@
 \end{blockarray}.
-\end{equation*}
+\end{equation}
 
@@
 
-By construction, the Taylor expansion of the top right term of the matrix
+By construction, the Taylor expansion of the top right term in the matrix
 $M_n(z)\cdot(I-M_n(z))^{-1}$ contains the the probabilities of interest.
-More specifically, $a_k$, the coefficient of $z^k$ in the series is the
-probability that a read of size $k$ contains no skip-$n$ seed of minimum
-size $\gamma$.
+More specifically, if the Taylor series of this term is $a_0 + a_1 z + a_2
+z^2 + \ldots$, then $a_k$ is the probability that a read of size $k$
+contains no skip-$n$ seed of minimum size $\gamma$.
 
 But $M_n(z)$ is too complex to find a closed expression for $M_n(z) \cdot
-(I-M_n(z))^{-1}$. Instead, we return to the definition $M_n(z) + M_n(z)^2
-+ \ldots$ and show in appendix~\ref{app_skip} that the coefficient of
-interest, $a_k$, only depends on the first $k+1$ terms of the sum. So for
-reads of size $k$ or lower, we only need to compute the matrix $M_n(z) +
-M_n(z)^2 + \ldots + M_n(z)^{k+1}$ and work out the Taylor expansion of the
-top right term.
+(I-M_n(z))^{-1}$ or its top right term. Instead, we return to the
+definition $M_n(z) + M_n(z)^2 + \ldots$ and show in
+appendix~\ref{app_skip} that the coefficient of interest, $a_k$, only
+depends on the first $k+1$ terms of the sum. So for reads of size $k$ or
+lower, we only need to compute the matrix $M_n(z) + M_n(z)^2 + \ldots +
+M_n(z)^{k+1}$ and work out the Taylor expansion of the top right term.
 
@@
 Using the formula above, we can thus bound the probability that a read has
-$s+1$ segments or more. We compute $M_n(z) + M_n(z)^2 + \ldots
+$s+1$ or more segments. We compute $M_n(z) + M_n(z)^2 + \ldots
 +M_n(z)^{s+1}$ where the weighted generating functions have been replaced
 by truncated polynomials and we extract the top right entry. When the
-bound is lower than a set fraction $\varepsilon$ of the current value of
-$a_k$, we stop the computations. Typically $\varepsilon = 0.01$ so this
-method ensures that the probabilities that a read of size $k$ has no
-on-target skip seed are accurate to within 1\%.
+right-hand side of expression (\ref{eq:bound}) is lower than a set
+fraction $\varepsilon$ of the current value of $a_k$, we stop the
+computations. Typically $\varepsilon = 0.01$ so this method ensures that
+the probabilities that a read of size $k$ has no on-target skip seed are
+accurate to within 1\%.
 
@@
 We now turn our attention to the problem of computing the probability that
-the seeding process is off target when using exact seeds --- recall from
+the seeding process is off target when using exact seeds---recall from
 section~\ref{sec:twotypes} that off-target seeding means that the
@@
 
-If there is $N = 0$ duplicate, seeding cannot be off target, it can only
-be on target or null. So from here we assume that the target has $N \geq
-1$ duplicates. Let $S_0$ denote the event that there is an on-target seed
-and let $S_j$ denote the event that there is a seed for the $j$-th
-duplicate. We are thus interested in computing $P\big(\overline{S}_0 \cap
-(S_1 \cup \ldots \cup S_N)\big)$, where $\overline{S}_j$ denotes the
-complement of the event $S_j$. For this, we first observe that
+If there is no duplicate (\textit{i.e.} $N = 0$), seeding cannot be off
+target, it can only be on target or null. So from here we assume that the
+target has $N \geq 1$ duplicates. Let $S_0$ denote the event that there is
+an on-target seed and let $S_j$ denote the event that there is a seed for
+the $j$-th duplicate. We are thus interested in computing
+$P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big)$, where
+$\overline{S}_j$ denotes the complement of the event $S_j$. First observe
+that
 \begin{equation}
@@
 Since the duplicates are assumed to evolve independently of each other and
-through the same mutagenesis process, the events are conditionally
-independent (we do not have full independence because $S_0$ only depends
-on sequencing errors whereas the events $S_j$ depend on sequencing errors
-and on the divergence between duplicates). We can thus write
+through the same mutagenesis process, the events $S_j$ ($1 \leq j \leq N$)
+are independent and identically distributed conditionally on
+$\overline{S}_0$. We can thus write
 \begin{eqnarray}
@@
 
-Hence, the probability that seeding is off target is a function of two
-quantitites. We have already seen how to compute $P(\overline{S}_0)$ in
+Hence, the probability that seeding is off target is a function of just
+two quantitites: $P(\overline{S}_0)$ and $P(\overline{S}_0 \cap
+\overline{S}_1)$. The first is the probability that the read has no seed
+for the target, which we have already computed in
 section~\ref{sec:example_exact} using recursive expression
@@
 
+\begin{remark}
+Observe that expression (\ref{eq:Pnull}) is the probability of null
+seeding (the read contains no seed for the target or any of its
+duplicates). Since it is also a function of just $P(\overline{S}_0)$ and
+$P(\overline{S}_0 \cap \overline{S}_1)$, it can be computed at no
+additional cost. This probablity is less relevant than the probabilities
+that the seeding process is on target or off target, but at times, it may
+be useful to know the probability that a read is not mappable, especially
+when reads are relatively short.
+\end{remark}
+
 \subsection{The dual encoding}
@@
 
-The event $\overline{S}_0 \cap \overline{S}_1$ is that the read contains
-no seed for either the target or for the first duplicate --- numbering is
-arbitrary here, we simply chose a duplicate and stick to it. As
+$P(\overline{S}_0 \cap \overline{S}_1)$ is the probability that the read
+has no seed for the target or for the first duplicate---numbering is
+arbitrary here, the first duplicate can be any fixed duplicate. As
 illustrated in section~\ref{sec:symbolic}, we first recode the reads using
@@
 \downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/i}^-$
-signify that the nucleotide is a mismatch against the $-$ sequence only,
-the symbols $\downarrow_{/i}^+$ signify that it is a mismatch against the
-$+$ only, and the symbol $\Downarrow$ signifies that it is a mismatch
-against both. As before, every other nucleotide is replaced by the symbol
-$\square$, and the terminator $|$ is appended to the end of the read. We
-again define reads as sequences of segments, except that now the
-terminators are the symbols $\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and
-$\Downarrow$. The tail, as usual, is terminated by the symbol $|$.
-
-The number $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
-length of the $+$ sequence. Likewise, the number $i$ in the symbol
+($i \geq 1$) signify that the nucleotide is a mismatch against the $-$
+sequence only, the symbols $\downarrow_{/i}^+$ ($i \geq 1$) signify that
+it is a mismatch against the $+$ sequence only, and the symbol
+$\Downarrow$ signifies that it is a mismatch against both. As before,
+every other nucleotide is replaced by the symbol $\square$, and the
+terminator $|$ is appended to the end of the read. We again define reads
+as sequences of segments, except that now the terminators are the symbols
+$\downarrow_{/i}^-$, $\downarrow_{/i}^+$ and $\Downarrow$ ($i \geq 1$).
+The tail, as usual, is terminated by the symbol $|$.
+
+The index $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
+length of the $+$ sequence. Likewise, the index $i$ in the symbol
 $\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
@@
 mismatch against the $-$ sequence, that it is a match for the $+$
-sequence, and that the six previous nucleotides were also a match for the
-$+$ sequence. The terminators thus encode the local state of the read.
-
-Figure~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
-and $-$ sequences are represented below the read, with matches represented
+sequence, and that the six preceding nucleotides were also a match for the
+$+$ sequence (but the nucleotide before that was a mismatch against the
+$+$ sequence). The terminators thus encode the local state of the read.
+
+Figure~\ref{fig:dual} shows an example of read in the dual encoding. The
+$+$ and $-$ sequences are shown below the read, with matches represented
 as open squares and mismatches as closed squares. It is visible from this
 example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
-whenever the mismatches hit different sequences . The symbol $\Downarrow$
+whenever the mismatches hit different sequences. The symbol $\Downarrow$
 occurs only when a nucleotide is a double mismatch.
@@
 
-We assume that for each nucleotide, $a$ is the probability that the read
-matches both sequences, $b$ is the probability that it matches only the
-$+$ sequence, $c$ is the probability that it matches only the $-$ sequence
-and $d$ is the probability that it matches none. Obviously $a+b+c+d=1$.
-These definitions allow us to specify the weighted generating function of
-the symbols and of the segments. To specify the transfer matrix, we only
-have to make sure that we eliminate the combinations that create a match
-of size $\gamma$ or more for any of the two sequences. For notational
-convenience, we define
+Let us assume that for each nucleotide, $a$ is the probability that the
+read matches both sequences, $b$ is the probability that it matches only
+the $+$ sequence, $c$ is the probability that it matches only the $-$
+sequence and $d$ is the probability that it matches neither. Obviously
+$a+b+c+d=1$. These definitions allow us to specify the weighted generating
+function of the symbols and of the segments. To specify the transfer
+matrix, we only have to make sure that we eliminate the combinations that
+create a match of size $\gamma$ or more for any of the two sequences. For
+notational convenience, we define
 \begin{equation*}
 \begin{gathered}
-r_i^+(z) = cz \cdot (az)^i, \\
-r_i^-(z) = bz \cdot (az)^i, \\
-R_i(z) = dz \cdot \big( 1 + az + \ldots + (az)^i \big), \\
+r_i^+(z) = (az)^i cz, \\
+r_i^-(z) = (az)^i bz, \\
+R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
 F_i(z) = 1 + az + \ldots + (az)^i,
@@
 
-The term of interest is the top right entry of
+As before, the term of interest is the top right entry of
 $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
@@
 sequences of segments that can follow the symbol $\Downarrow$ and that are
-terminated by a tail, whose weighted generating function is the top right
-entry of the matrix.
-
-$\tilde{M}_0(z)$ is too complex to allow a closed expression of
-$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$ to be computed. It is
-actually easier to proceed as in section~\ref{sec:example_skip} and to
-compute the powers of $\tilde{M}_0(z)$ (using the arithmetic of truncated
-polynomials) up to a given threshold value. Since each segment except the
-tail contains a mismatch against at least one sequence, we define
-$\tilde{p}$ as the upper bound on the probability of a mismatch $\tilde{p}
-= \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now gives an upper
-bound of the probability that a read of size $k$ contains $s$ or more
-mismatches as
+terminated by a tail, the weighted generating function of which is located
+in the top right entry of the matrix.
+
+$\tilde{M}_0(z)$ is too complex to compute a closed expression of
+$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
+proceed as in section~\ref{sec:example_skip} and to compute the powers of
+$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
+arithmetic of truncated polynomials. Since each segment except the tail
+contains a mismatch against at least one sequence, we define $\tilde{p}$
+as the upper bound on the probability of a mismatch, \textit{i.e.}
+$\tilde{p} = \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now
+gives an upper bound of the probability that a read of size $k$ contains
+$s$ or more mismatches as
 \begin{equation*}
@@
 With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
-compute its powers until the upper bound is lower than a set fraction
-$\varepsilon$ of the current value of $a_k$, which gives an accurate
-estimate of $P(\overline{S}_0 \cap \overline{S}_1)$.
-
-\begin{remark}
-Observe that expression (\ref{eq:Pnull}) is the probability of null
-seeding. The method described above thus allows us to compute this
-probability at no additional cost. This probablity is less relevant than
-the probabilities that the seeding process is on target or off target, but
-at times, it may be useful to know the probability that a read is
-not mappable, especially when reads are relatively short.
-\end{remark}
+compute its powers until the right-hand side of the expression above is
+lower than a set fraction $\varepsilon$ of the current value of $a_k$,
+which gives an accurate estimate of $P(\overline{S}_0 \cap
+\overline{S}_1)$.
 
@@
 can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
-event that the read has a \emph{skip} seed for the target and $S_1$ that
-it has a \emph{skip} seed for the first duplicate.
+event that the read has a \emph{skip} seed for the target (instead of an
+exact seed) and $S_1$ is the event that it has a \emph{skip} seed for the
+first duplicate (instead of an exact seed).
 
@@
 \downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
-$\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ have the same meaning as in
-the dual alphabet of section~\ref{sec:dual}. The symbols $\Downarrow_i$
-indicate that both sequences have match length 0 and that the next
-non-skipped position is $i$ nucleotides downstream.
+$\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ ($i \geq 1$) have the same
+meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
+$\Downarrow_i$ ($0 \leq i \leq n$) indicate that both sequences have match
+length 0 and that the next non-skipped position is $i$ nucleotides
+downstream.
 
@@
 
-\section{Random seeds}
-\label{sec:random_seeds}
-
-We have assumed that seeds can match only the target or one of its
-duplicates. In practice, seeds can match many random locations of the
-genome and not just the target and the duplicates.
-
-It is important for the validity of the theory that all the sequences that
-have a seed are considered candidate locations (as long as the seed is
-above the threshold $\gamma$). If not, the final estimates are biased. The
-hope is that spurious matches are shorter than $\gamma$ so that they are
-filtered out, but this is not always the case. And if they qualify as a
-seed we need to verify the candidate location.
-
-Such random hits can be problematic for two reasons: First, the time spent
-verifying them is wasted. Second, they can cause false positives.
-Fortunately, both issues can be addressed.
-
-To save time during the alignment phase, we can prioritize the seeds so
-that the best hit is likely to be discovered first, allowing us to bail
-out from other alignments as early as possible. We can even bound the
-maximum quality of some candidates so that the alignment can be skipped
-altogether. These considerations depend on the implementation of the
-mapper so we will not develop this further. What matters is that random
-hits do not impose a significant burden on the time needed to verify the
-candidates.
-
-The second issue is that random hits can generate false positives. Such
-cases occur when seeding is null (meaning that there is no seed for the
-target nor for its duplicates). Since the hits are not homologous to the
-read, the alignment score is extreme, which makes it easy to detect.
-
-Indeed, if the candidate location is random, mismatches occur with
-probability 3/4. If it is the true location, mismatches occur with
-probability $p$. We discard the seed because it is an automatic match of
-size $\gamma$ (and possibly larger in the case of MEM seeds). Say that
-there remain $L$ nucleotides and that $m$ of them are mismatches for the
-candidate location. From Bayes formula, the probability that the location
-is random given the number of mismatches is
-\begin{equation}
-\label{eq:bayes}
-\frac{1}{1 + 4^L(p/3)^mq^{L-m}(1-\beta)/\beta},
-\end{equation}
-where $q=1-p$ and $\beta$ is the prior probability that the hit is
-random.
-
-The value of $\beta$ has little importance if $p$ is small. Say that
-$p=0.01$ and that a read generates a hit in the genome such that 33
-nucleotides need to be aligned after seeding. If the hit is random there
-is a 99.99\% chance that at least 15 are mismatches. The denominator of
-expression (\ref{eq:bayes}) is approximately $1 +
-4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
-support for the hypothesis that the hit is random is overwhelming.
-Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
-fewer nucleotides are mismathes. In this case, the denominator is
-approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
-10^{-9}$, the support for the hypothesis that the hit is not random is
-overwhelming. If $p$ is small, we do not need to worry about the value of
-$\beta$, one can choose for instance $\beta=1/2$ so that the term
-$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
-
-Since the probability that the best candidate is a random sequence is
-either very small or very large, it has no influence in the first case,
-or it dominates the probability of a false positive in the secondi case.
-For every read, we can use as confidence score the maximum of this
-probability and of the estimated probability that seeding is off target.
-
-In summary, seeds in random sequences happen relatively frequently, but it
-is possible to minimize their computational burden. Also, they are no
-cause for concern regarding false positives because they can easily be
-detected using Bayes formula, as shown in expression (\ref{eq:bayes}).
-
-\section{True versus best location}
+\section{Practical considerations}
+
+\subsection{True versus best location}
 \label{sec:truevsbest}
@@
 
+\subsection{Non-specific seeds}
+\label{sec:random_seeds}
+
+We have assumed that seeds can match only the target or one of its
+duplicates. In practice, seeds can match many random locations of the
+genome and not just the target and the duplicates.
+
+It is important for the validity of the theory that all the sequences that
+have a seed are considered candidate locations (as long as the seed is
+above the threshold $\gamma$). If not, the final estimates are biased. The
+hope is that spurious matches are shorter than $\gamma$ so that they are
+filtered out, but this is not always the case. And if they qualify as a
+seed we need to verify the candidate location.
+
+Such random hits can be problematic for two reasons: First, the time spent
+verifying them is wasted. Second, they can cause false positives.
+Fortunately, both issues can be addressed.
+
+To save time during the alignment phase, we can prioritize the seeds so
+that the best hit is likely to be discovered first, allowing us to bail
+out from other alignments as early as possible. We can even bound the
+maximum quality of some candidates so that the alignment can be skipped
+altogether. These considerations depend on the implementation of the
+mapper so we will not develop this further. What matters is that random
+hits do not impose a significant burden on the time needed to verify the
+candidates.
+
+The second issue is that random hits can generate false positives. Such
+cases occur when seeding is null (meaning that there is no seed for the
+target nor for its duplicates). Since the hits are not homologous to the
+read, the alignment score is extreme, which makes it easy to detect.
+
+Indeed, if the candidate location is random, mismatches occur with
+probability 3/4. If it is the true location, mismatches occur with
+probability $p$. We discard the seed because it is an automatic match of
+size $\gamma$ (and possibly larger in the case of MEM seeds). Say that
+there remain $L$ nucleotides and that $m$ of them are mismatches for the
+candidate location. From Bayes formula, the probability that the location
+is random given the number of mismatches is
+\begin{equation}
+\label{eq:bayes}
+\frac{1}{1 + 4^L(p/3)^mq^{L-m}(1-\beta)/\beta},
+\end{equation}
+where $q=1-p$ and $\beta$ is the prior probability that the hit is
+random.
+
+The value of $\beta$ has little importance if $p$ is small. Say that
+$p=0.01$ and that a read generates a hit in the genome such that 33
+nucleotides need to be aligned after seeding. If the hit is random there
+is a 99.99\% chance that at least 15 are mismatches. The denominator of
+expression (\ref{eq:bayes}) is approximately $1 +
+4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
+support for the hypothesis that the hit is random is overwhelming.
+Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
+fewer nucleotides are mismathes. In this case, the denominator is
+approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
+10^{-9}$, the support for the hypothesis that the hit is not random is
+overwhelming. If $p$ is small, we do not need to worry about the value of
+$\beta$, one can choose for instance $\beta=1/2$ so that the term
+$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
+
+Since the probability that the best candidate is a random sequence is
+either very small or very large, it has no influence in the first case,
+or it dominates the probability of a false positive in the secondi case.
+For every read, we can use as confidence score the maximum of this
+probability and of the estimated probability that seeding is off target.
+
+In summary, seeds in random sequences happen relatively frequently, but it
+is possible to minimize their computational burden. Also, they are no
+cause for concern regarding false positives because they can easily be
+detected using Bayes formula, as shown in expression (\ref{eq:bayes}).
+
+\subsection{Realistic error patterns}
+
+This section will come later.
+
 \section{The Sesame library}
@@
 
-\subsection{Derivation of something important}
+\subsection{Proof that $a_k$ is independent of $M_n(z)^m$ for $m > k+1$}
 \label{app_skip}
 
-Easy.
-
-observe that the terms $M_n(z)^{k+2}, M_n(z)^{k+3},
-\ldots$ have no influence on the coefficients of the weighted generating
-function $a_0, a_1, \ldots, a_k$.
-
-Indeed, a read of size $k$ has at most $k+1$ segments (including the
-obligatory tail, which can have size 0). Since $M(z)^{s+1}$ contains the
-weighted generating functions of reads with exactly $s+1$ segments
-including the tail, $a_k$ cannot depend on $M(z)^{k+2}, M(z)^{k+3},
-\ldots$ More formally, we can prove by induction that all the entries of
-$M(z)^k$ are divisible by $z^{k-1}$, showing that the contribution of
-$M(z)^{k+2} + M(z)^{k+3} + \ldots$ to $a_0 + a_1z + \ldots +a_kz^k$ is
-strictly $0$.
+In section~\ref{sec:example_skip} we considered the transfer matrix
+$M_n(z)$ of reads containing no skip-$n$ seed. The quantity of interest
+$a_k$ is the coefficent of $z^k$ in the Taylor expansion of the top right
+term of $M_n(z) + M_n(z)^2 + M_n(z)^3 + \ldots$ We claimed that the
+coefficients $a_0, a_1, \ldots, a_k$ are independent of $M_n(z)^{k+2},
+M_n(z)^{k+3}, \ldots$ so that they can be computed from the finite sum
+$M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$.
+
+We proceed by induction and show that for $k \geq 1$ all the entries of
+$M_n(z)^k$ are mutliples of $z^{k-1}$. For $k = 1$ the statement is
+trivial and there is nothing to show. Now assume that for some $k > 1$,
+all the entries of $M_n(z)^k$ are multiples of $z^{k-1}$. Observe that
+from the definition of $M_n(z)$ in expression (\ref{eq_matrix_Mn}), the
+last row of $M_n(z)^m$ consists of zeros only for all $m \geq 1$. So in
+the matrix product $M_n(z) \cdot M_n(z)^k$, the polynomials $J_0(z),
+\ldots, J_n(z)$ in the last column of $M_n(z)$ are all multiplied by 0.
+The remaining terms of $M_n(z)$ are multiples of $z$, so applying the
+induction hypothesis shows that all the terms of $M_n(z)^{k+1}$ are
+multiples of $z^k$.
+
+Since all the terms of $M(z)^m$ are multiples of $z^{k+1}$ for $m \geq
+k+1$, the coefficient of $z^k$ is 0 in the top right term of every matrix
+$M(z)^m$ for $m \geq k+1$. As a result, the coefficient of $z^k$ in the
+top right entry of $M_n(z) + M_n(z)^2 + M_n(z)^3 + \ldots$ is the same as
+in $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$, as we had to demonstrate.
+The same rationale holds for the coefficients of $z^m$ $(0 \leq m < k)$,
+so finally the coefficients $a_0, a_1, \ldots, a_k$ can all be computed
+from the top right entry of $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$ as
+claimed.
 

### 2020-03-11 manuscript.tex
About to simplify the skip transfer matrix

@@
 \downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
-\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/i}^-$
-($i \geq 1$) signify that the nucleotide is a mismatch against the $-$
-sequence only, the symbols $\downarrow_{/i}^+$ ($i \geq 1$) signify that
+\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/j}^-$
+($j \geq 1$) signify that the nucleotide is a mismatch against the $-$
+sequence only, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) signify that
 it is a mismatch against the $+$ sequence only, and the symbol
@@
 terminator $|$ is appended to the end of the read. We again define reads
-as sequences of segments, except that now the terminators are the symbols
-$\downarrow_{/i}^-$, $\downarrow_{/i}^+$ and $\Downarrow$ ($i \geq 1$).
+as sequences of segments (zero or more $\square$ symbols followed by a
+terminator), except that now the terminators are the symbols
+$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and $\Downarrow$ ($j \geq 1$).
 The tail, as usual, is terminated by the symbol $|$.
 
-The index $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
-length of the $+$ sequence. Likewise, the index $i$ in the symbol
-$\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
-instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
-mismatch against the $-$ sequence, that it is a match for the $+$
+The index $j$ in the symbol $\downarrow_{/j}^-$ indicates the match length
+of the $+$ sequence (note that this is not the same as the number of
+$\square$ symbols in the segment). Likewise, the index $j$ in the symbol
+$\downarrow_{/j}^+$ indicates the match length of the $-$ sequence. For
+instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is
+a mismatch against the $-$ sequence, that it is a match for the $+$
 sequence, and that the six preceding nucleotides were also a match for the
@@
 as open squares and mismatches as closed squares. It is visible from this
-example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
+example that symbols $\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ alternate
 whenever the mismatches hit different sequences. The symbol $\Downarrow$
@@
 An example of read is represented in the dual alphabet. The symbols
-$\downarrow_{/i}^-$ represent a mismatch against the $-$ sequence, the
-symbols $\downarrow_{/i}^+$ represent a mismatch against the $+$ sequence,
-and the symbol $\Downarrow$ represent a mismatch against both. The index
-$i$ is the match length of the sequence that is not mismatched. The
-symbolic $+$ and $-$ sequences are represented below, where an open square
-stands for a match and a closed square stands for a mismatch.}
+$\downarrow_{/j}^-$ ($j \geq 1$) represent a mismatch against the $-$
+sequence, the symbols $\downarrow_{/j}^+$ ($i \geq 1$) represent a
+mismatch against the $+$ sequence, and the symbol $\Downarrow$ represents
+a mismatch against both. The index $i$ is the match length of the sequence
+that is not mismatched. The symbolic $+$ and $-$ sequences are represented
+below, where an open square stands for a match and a closed square stands
+for a mismatch.}
 \label{fig:dual}
@@
 the $+$ sequence, $c$ is the probability that it matches only the $-$
-sequence and $d$ is the probability that it matches neither. Obviously
-$a+b+c+d=1$. These definitions allow us to specify the weighted generating
-function of the symbols and of the segments. To specify the transfer
-matrix, we only have to make sure that we eliminate the combinations that
-create a match of size $\gamma$ or more for any of the two sequences. For
-notational convenience, we define
-\begin{equation*}
+sequence and $d$ is the probability that it matches neither. Since there
+are no other cases, we have $a+b+c+d=1$.
+
+With these definitions, $a$, $b$, $c$ and $d$ are the probabilities of
+occurrence of the symbols $\square$, $\downarrow_{/j}^-$,
+$\downarrow_{/j}^+$ and $\Downarrow$, respectively ($j \geq 1$). The next
+sections clarify how this is used to compute the weighted gerating
+functions of interest.
+
+\subsection{Segments following $\Downarrow$}
+
+After a $\Downarrow$ terminator, the match counter for both sequences is
+reset; the following segment can thus have up to $\gamma-1$ matches for
+any of the two sequences. Each match corresponds to the $\square$ symbol
+with weighted generating function $az$. The terminators $\Downarrow$ and
+$|$ have respective generating function $dz$ and $1$ (recall that the tail
+symbol has size 0), so if the next terminator is $\Downarrow$ or $|$, the
+segments have weighted generating functions $(1 + az + \ldots +
+(az)^{\gamma-1})dz$ or $1 + az + \ldots + (az)^{\gamma-1}$, respectively.
+
+If the next terminator is $\downarrow_{/i}^+$, there is a match of size
+$i$ for the $-$ sequence, so the segment contains $i-1$ symbols
+$\square$ (one of the matches is accounted for by the terminator). The
+weighted generating function is thus $(az)^{i-1}cz$. Likewise, if the next
+terminator is $\downarrow_{/i}^-$, the weighted generating function of the
+segment is $(az)^{i-1}bz$. The terminators $\downarrow_{/i}^+$ and
+$\downarrow_{/i}^-$ are disallowed for $j \geq \gamma$ because this would
+create a match of size $\gamma$ or greater for one of the sequences.
+
+\subsection{Segments following $\downarrow_{/j}^+$}
+
+At a $\downarrow_{/j}^+$ terminator, the match length of the $+$ sequence
+is 0 and the match length of the $-$ sequence is $j$. The next segment can
+thus have up to $\gamma-1$ matches for the $+$ sequence, but only
+$\gamma-j-1$ matches for the $-$ sequence, lowering the maximum size of
+the segment. If the next terminator is $\Downarrow$ or $|$, the segments
+have weighted generating functions $(1 + az + \ldots +
+(az)^{\gamma-j-1})dz$ and $1 + az + \ldots + (az)^{\gamma-j-1}$,
+respectively.
+
+If the next terminator is $\downarrow_{/i}^-$, the weighted generating
+function is $(az)^{i-1}bz$ as in the previous section. The difference is
+that the terminators $\downarrow_{/i}^-$ are allowed only for $1 \leq i
+\leq \gamma-j$, otherwise this would create a match of size $\gamma$ or
+greater for the $-$ sequence.
+
+If the next terminator is $\downarrow_{/i}^+$, the situation is slightly
+more complicated, because this imposes $j < i \leq \gamma-1$. Indeed,
+there were already $j$ matches for the $-$ sequence at the terminator
+$\downarrow_{/j}^+$, and there will be more at the end of the segment
+because it has no mismatch for the $-$ sequence. Taking this into account,
+we see that the weighted generating function of those segments is
+$(az)^{i-j-1}cz$.
+
+\subsection{Segments following $\downarrow_{/j}^-$}
+
+We can find the weighted generating functions by just reversing the $+$
+and $-$ signs in the previous section. This way, we see that the weighted
+generating function of the segments terminated by $\Downarrow$ or $|$ are
+$(1 + az + \ldots + (az)^{\gamma-j-1})dz$ and $1 + az + \ldots +
+(az)^{\gamma-j-1}$, respectively.
+
+Likewise, the weighted generating function of the segments terminated by
+$\downarrow_{/i}^+$ is $(az)^{i-1}cz$, where $1 \leq i \leq \gamma-j$; and
+the weighted generating function of the segments terminated by
+$\downarrow_{/i}^-$ is $(az)^{i-j-1}bz$, where $j < i \leq \gamma-1$.
+
+\subsection{Transfer matrix}
+
+We now have all the elements to specify the transfer matrix or reads with
+no seed for either sequence. Recall that $a$ is the probability of a
+double match, $b$ is the probability of a mismatch only against the $-$
+sequence, $c$ is the probability of a mismatch only against the $+$
+sequence and $d$ is the probability of a double mismatch. For notational
+convenience, we define
+\begin{equation}
+% These expressions are not needed for reference in the text, but
+% they are still numbered so that the code can reference to it.
 \begin{gathered}
@@
 R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
-F_i(z) = 1 + az + \ldots + (az)^i,
+F_i(z) = 1 + az + \ldots + (az)^i.
 \end{gathered}
-\end{equation*}
-and we can verify that the expression of the transfer matrix
-$\tilde{M}_0(z)$ is
+\end{equation}
+
+With these notations, the information from the previous sections can be
+summarized in the transfer matrix $\tilde{M}_0(z)$ equal to
 \begin{equation*}
@@
 $\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
-every read can be prepended by $\Downarrow$ segments and only by those
+every read can be prepended by $\Downarrow$-segments and only by those
 (every other terminator would imply that one of the two sequences has a
@@
 $\tilde{M}_0(z)$ is too complex to compute a closed expression of
-$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
-proceed as in section~\ref{sec:example_skip} and to compute the powers of
+$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
+in section~\ref{sec:example_skip} and to compute the powers of
 $\tilde{M}_0(z)$ up to a finite value. This is done once again using the
@@
 sequenced with an instrument with error rate $p=0.01$, when using exact
-seeds of size $\gamma=17$.
+seeds of size $\gamma=19$.
 
@@
 \includegraphics[scale=0.65]{curves_exact.pdf}
-\caption{\textbf{Off target seeding probabilities (exact seeds).}
-The curves show the probability that seeding is off target for exact seeds
-of size $\gamma=17$ in reads of size $k=50$ nucleotides sequenced with an
-error rate $p=0.01$. Each line shows a different number of duplicate
-sequences $N$ from 1 to 10 and the x-axis shows the divergence rate $\mu$,
-defined as the probability that a given duplicate differs from the target
-at any given position.}
+\caption{\textbf{Off-target seeding probabilities (exact seeds).}
+The surfaces show the probability that seeding is off target for exact seeds
+of size $\gamma=19$ in reads of size $k=50$ nucleotides sequenced with an
+error rate $p=0.01$. The divergence rate $\mu$, defined as the probability
+that a given duplicate differs from the target at any given position. Note
+the factor 1000 in the scale of the z-axis.}
 \label{fig:curves_exact}
@@
 observation is that the probability that seeding is off target increases
-with $N$. This is clearly visible from expression (\ref{eq:Poff}). This
-can also be understood intuitively because the probability of null seeding
+with $N$. This is also clearly from expression (\ref{eq:Poff}). This can
+also be understood intuitively because the probability of null seeding
 decreases when there are more potential candidates, and since the
@@
 The second observation is that there exists a ``worst'' value of $\mu$
-situated around 0.08. When $\mu$ is much smaller, the sequences of the
-duplicates are close to that of the target so it is unlikely that the
-candidate set contains one but not the other. When $\mu$ is much larger,
-the sequences of the duplicates are far from that of the target and they
-are unlikely to be in the candidate set. In expression (\ref{eq:Poff}),
-the only term that depends on $\mu$ is $P(\overline{S}_0 \cap
-\overline{S}_1)$, and it is obvious that the minimum of expression
-(\ref{eq:Poff}) corresponds to the maximum of $P(\overline{S}_0 \cap
-\overline{S}_1)$. This is why the worst value of $\mu$ is the same for
-every $N$.
-
-For comparison purposes, it is important to notice the order of magnitude
-of the probabilities. Up to $N=3$ duplicates, off-target seeding has
-probability lower than $10^{-4}$ and for $N=10$ duplicates the upper bound
-is $3\cdot10^{-4}$. This sets a baseline to compare with more elaborate
-seeding methods.
+situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
+be exactly identical to the target, making it impossible that there is a
+seed for a duplicate but no seed for the target. When $\mu$ is much
+larger, the duplicates are far from the target and they are unlikely to be
+in the candidate set. In expression (\ref{eq:Poff}), the only term that
+depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it is
+clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
+maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
+value of $\mu$ is the same for every $N$.
 
@@
 event that the read has a \emph{skip} seed for the target (instead of an
-exact seed) and $S_1$ is the event that it has a \emph{skip} seed for the
-first duplicate (instead of an exact seed).
+exact seed) and $S_1$ is the event that the read has a \emph{skip} seed
+for the first duplicate (instead of an exact seed).
 
@@
 
-Once again, we start by recoding the reads in a specialized alphabet. We
-define the skip-$n$ dual alphabet $\tilde{\mathcal{A}}_n = \{\square, |,
+Again we recode the reads in a specialized alphabet. We define the
+skip-$n$ dual alphabet as $\tilde{\mathcal{A}}_n = \{\square, *, |,
 \Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
@@
 \downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
-$\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ ($i \geq 1$) have the same
+$\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ ($j \geq 1$) have the same
 meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
-$\Downarrow_i$ ($0 \leq i \leq n$) indicate that both sequences have match
-length 0 and that the next non-skipped position is $i$ nucleotides
-downstream.
+$\Downarrow_j$ ($0 \leq j \leq n$) indicate that both sequences have match
+length 0 and that the next non-skipped position is $j$ nucleotides
+further. The new symbol $*$ indicates that the nature of the nucleotide
+does not matter, as we will explain below.
 
@@
 differences with Figure~\ref{fig:dual}. The first is that the symbols
-$\Downarrow_i$ are not always associated with double mismatches. For
-instance, the symbol $\Downarrow_0$ on the right side of the read
-corresponds to a mismatch for the $+$ sequence only. This happens whenever
-the $+$ and the $-$ sequences are mismatched in the same interval between
-non-skipped positions (the mismatches do not need to be on the same
-nucleotide).
+$\Downarrow_j$ ($0 \leq j \leq n$) are not always associated with double
+mismatches. For instance, the symbol $\Downarrow_0$ on the right side of
+the read corresponds to a mismatch for the $+$ sequence only. This happens
+whenever the $+$ and the $-$ sequences are mismatched in the same interval
+between non-skipped positions (the mismatches do not need to be on the
+same nucleotide).
+
+Also observe that the symbol $\Downarrow_1$ is followed by a $*$ to
+indicate that this nucleotide has no influence on the presence of a seed,
+regardless 
+. The reason is that there
+is a double mismatch at the $\Downarrow_1$ symbol so there can be no seed.
+But since the next potential seed starts only at the next tick mark,
+\textit{i.e.} one nucleotide further downstream,
 
@@
 \includegraphics[scale=0.85]{sketch_skip_dual.pdf}
-\caption{\textbf{Example skip dual encoding.}
+\caption{\textbf{Example of skip dual encoding.}
 The read of Figure~\ref{fig:dual} is represented in the skip dual alphabet.
 The vertical bars above the read indicate non-skipped positions. The
-symbols $\downarrow_{/i}^-$ and $\downarrow_{/i}^+$ have the same meaning
-as in the dual alphabet. The symbols $\Downarrow_i$ indicate that both
-sequences have match length 0 and that the next non-skipped position is
-located $i$ nucleotides downstream. Other features are as in
+symbols $\downarrow_{/j}^-$ and $\downarrow_{/j}^+$ ($j \geq 1$) have the
+same meaning as in the dual alphabet. The symbols $\Downarrow_j$ indicate
+that both sequences have match length 0 and that the next non-skipped
+position is located $j$ nucleotides downstream. Other features are as in
 Figure~\ref{fig:dual}.}
@@
 \end{figure}
+
+As in the previous section, let $a$, $b$, $c$ and $d$ denote the
+probabilities of a double match, a mismatch against $-$, a mismatch
+against $+$ and a double mismatch, respectively. This also corresponds to
+the probabilities of occurrence of the symbols $\square$,
+$\downarrow_{/j}^+$, $\downarrow_{/j}^-$ ($j \geq 1$) and $\Downarrow_j$
+($0 \leq j \leq n$), respectively. With this information, we compute the
+weighted generating functions of interest in the sections below.
+
+\subsection{Segments following $\Downarrow_j$ ($1 \leq j \leq n)$}
+
+After a $\Downarrow_j$ terminator, the match counter for both sequences is
+at zero and the next potential seed starts only $j \geq 1$ nucleotides
+further downstream. This means that the next $j$ nucleotides are just
+filler so it does not matter which sequence they match or mismatch.
+
+There are only two scenarios to consider: if the end of the read is within
+the next $j$ nucleotides, then
+
+\subsection{Segments following $\Downarrow_0$}
+
+Hello world!
+
+\subsection{Transfer matrix}
 
@@
 We illustrate the strategy delineated above Using the same settings as in
-section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), except that
+section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=19$), except that
 we use skip-9 seeds instead of exact seeds.
@@
 \centering
-\includegraphics[scale=0.65]{curves_skip.pdf}
-\caption{\textbf{Off target seeding probabilities (skip seeds).}
-The curves show the probability that seeding is off target for skip-9
-seeds of size $\gamma=17$. Read size and error rate are the same as in
-Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
-shows a different number of duplicate sequences $N$ from 1 to 10 and the
-x-axis shows the divergence rate $\mu$, defined as the probability that a
-given duplicate differs from the target at any given position.}
+\includegraphics[scale=0.47]{curves_skip.pdf}
+\caption{\textbf{Off-target seeding probabilities (skip seeds).}
+The surfaces show the probability that seeding is off target for skip-9
+seeds of size $\gamma=19$. Read size and error rate are the same as in
+Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. The
+divergence rate $\mu$, defined as the probability that a given duplicate
+differs from the target at any given position. Note the factor 1000 in
+the scale of the z-axis.}
 \label{fig:curves_skip}
@@
 using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
-and $\gamma=17$), except that we replace exact seed by MEM seeds.
+and $\gamma=19$), except that we replace exact seed by MEM seeds.
 
@@
 \centering
-\includegraphics[scale=0.65]{curves_mem.pdf}
-\caption{\textbf{Off target seeding probabilities (MEM seeds).}
-The curves show the probability that seeding is off target for MEM
-seeds of minimum size $\gamma=17$. Read size and error rate are the same
-as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each
-line shows a different number of duplicate sequences $N$ from 1 to 10 and
-the x-axis shows the divergence rate $\mu$, defined as the probability
-that a given duplicate differs from the target at any given position.}
+\includegraphics[scale=.48]{curves_mem.pdf}
+\caption{\textbf{Off-target seeding probabilities (MEM seeds).}
+The surfaces show the probability that seeding is off target for MEM seeds
+of indicated minimum size $\gamma$. Read size and error rate are the same
+as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$.
+The divergence rate $\mu$ is the probability that a given duplicate
+differs from the target at any given position. Note the factor 1000 in the
+scale of the z-axis.}
 \label{fig:curves_mem}
@@
 Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
-fewer nucleotides are mismathes. In this case, the denominator is
+fewer nucleotides are mismatches. In this case, the denominator is
 approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
@@
 either very small or very large, it has no influence in the first case,
-or it dominates the probability of a false positive in the secondi case.
+or it dominates the probability of a false positive in the second case.
 For every read, we can use as confidence score the maximum of this

### 2020-03-16 manuscript.tex
Simplify skip-n transfer matrix (in appendix)

@@
 \end{block}
-\end{blockarray}.
+\end{blockarray},
 \end{equation*}
-
-In the representation above, the different types of segments are
-identified by their terminator, indicated in the margins for clarity.
+where $p$ is the error rate of the sequencer, $q=1-p$ and $\gamma$ is the
+minimum seed length. In the representation above, the different types of
+segments are identified by their terminator, indicated in the margins for
+clarity.
 
@@
 has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
-equivalent to defining $x$ as the number of $\square$ symbols of the
-shortest $\Downarrow_j$-segment, with $x = i-j-1 \modulo{n+1}$ and $\ell_0
-= x+1$.
-
-The shortest $\Downarrow_j$-segment consists of $x$ symbols $\square$
-followed by the $\Downarrow_j$ terminator, so it has weighted generating
-function $(qz)^xpz$. The other segments have a multiple of $n+1$
-additional $\square$ symbols so their weighted generating functions are
-$(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots, (qz)^{x+m(n+1)}pz$. Summing
-over those cases, we finally obtain
+equivalent to defining $\ell_0$ as $x+1$ where $x$ is the number of
+$\square$ symbols of the shortest $\Downarrow_j$-segment, \textit{i.e.} $x
+= i-j-1 \modulo{n+1}$.
+
+With these notations, he shortest $\Downarrow_j$-segment consists of $x$
+symbols $\square$ followed by the $\Downarrow_j$ terminator, so it has
+weighted generating function $(qz)^xpz$. The other segments have a
+multiple of $n+1$ extra $\square$ symbols so their weighted
+generating functions are $(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots,
+(qz)^{x+m(n+1)}pz$. Summing over those cases, we finally obtain
 \begin{eqnarray}
+\label{eq_H}
 H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
@@
 
-Denote $J_i(z)$ the weighted generating function of tails that can follow
-a $\Downarrow_i$-segment. There are $\gamma+i$ of them, each consisting of
-0 to $\gamma+i-1$ symbols $\square$ followed by the special $|$
-terminator for the end of the read. The $|$ has size 0 so its weighted
+Denote $J_i(z)$ the weighted generating function of tail segments that can
+follow a $\Downarrow_i$-segment. There are $\gamma+i$ such tails, each
+consisting of 0 to $\gamma+i-1$ symbols $\square$ followed by the special
+$|$ terminator for the end of the read. The $|$ has size 0 so its weighted
 generating function is 1. Once again summing over the different cases, we
 obtain
-\begin{eqnarray}
-J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
-\end{eqnarray}
+\begin{equation}
+\label{eq_J}
+J_i(z) = 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1}.
+\end{equation}
 
@@
 \end{block}
-\end{blockarray}.
+\end{blockarray},
 \end{equation}
+where $p$ is the error rate of the sequencer, $q=1-p$, $n$ is the number of
+skipped nucleotides between potential seeds, $\gamma$ is the minimum seed
+length, and polynomials $H_{i,j}(z)$ and $J_i(z)$ are defined in
+expressions (\ref{eq_H}) and (\ref{eq_J}).
+
+In appendix~\ref{app_Mn}, we present a transfer matrix with a simpler
+expression that will prove useful in section~\ref{sec_offskip}. The
+version of appendix~\ref{app_Mn} is simpler, but the expression above has
+some advantages that will be explained below.
 
@@
 But we can do even better: a read with $s+1$ segments contains $s$ errors,
-so all the entries of $M_n(z)^{s+1}$ are dominated by $p^s$ and they
-rapidly converge to $0$ as $s$ increases. Instead of computing the matrix
-$M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$, we can interrupt the
-summation after a certain power of $M_n(z)$ because the terms become
-negligible.
+so the top right entry of $M_n(z)^{s+1}$ is the weighted generating
+function of reads with $s$ errors that have no seed of minimum size
+$\gamma$. Computing the partial sum $M_n(z) + M_n(z)^2 + \ldots +
+M_n(z)^s$ instead of $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{k+1}$
+corresponds to neglecting reads with $s$ or more errors. For $s$
+sufficiently large, such reads are exceedingly rare so we can obtain
+accurate estimates without computing all the powers of $M_n(z)$ up to
+order $k$.
 
@@
 accurate to within 1\%.
+
+If we used $M_n^*(z)$the transfer matrix of appendix~\ref{app_Mn}, the top
+right entry of $M_n(z)^{s+1}$ is not the weighted generating function of
+reads with $s$ errors, so (\ref{eq:bound}) is not an upper bound for the
+neglected terms of the sum. As a consequence, one would have to compute
+more terms in the partial sum $M_n(z) + M_n(z)^2 + \ldots + M_n(z)^{s+1}$
+to reach the same accuracy. The transfer matrix shown in expression
+(\ref{eq_matrix_Mn}) is not the simplest, but it has the benefit of
+requiring fewer iterations.
 
@@
 \includegraphics[scale=0.65]{curves_exact.pdf}
-\caption{\textbf{Off-target seeding probabilities (exact seeds).}
-The surfaces show the probability that seeding is off target for exact seeds
+\caption{\textbf{Off-target seeding probabilities (exact seeds).} The
+surfaces show the probability that seeding is off target for exact seeds
 of size $\gamma=19$ in reads of size $k=50$ nucleotides sequenced with an
@@
 that a given duplicate differs from the target at any given position. Note
-the factor 1000 in the scale of the z-axis.}
-\label{fig:curves_exact}
+the factor 1000 in the scale of the z-axis.} \label{fig:curves_exact}
 \end{figure}
@@
 \section{Off-target skip seeds}
+\label{sec_offskip}
 
@@
 section~\ref{sec:dual}. For notational convenience, we define
-\begin{eqnarray*}
+\begin{eqnarray}
+\label{eq_N}
 N_i(z) &=& 1+z+\ldots+z^i, \\
@@
   m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
-\end{eqnarray*}
+\end{eqnarray}
 
@@
 
+\subsection{An alternative construction of $M_n(z)$}
+\label{app_Mn}
+
+The alphabets described here are constructed so that there is exactly one
+way to decompose reads in symbols or in segments. For a given alphabet
+there exists only one transfer matrix that describes the reads that have
+no seed of minimum size $\gamma$. However, if we choose a different
+alphabet, we can obtain a different transfer matrix describing exactly the
+same reads.
+
+In section~\ref{sec:example_skip}, we decomposed the reads in
+the so-called skip-$n$ alphabet $\mathcal{A}_n = \{\square, |,
+\Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$, from which we
+derived the transfer matrix $M_n(z)$ shown in expression
+(\ref{eq_matrix_Mn}). We now introduce the alternative skip-$n$ alphabet
+$\mathcal{A}^*_n = \{\square, *, |, \Downarrow_0, \Downarrow_1, \ldots,
+\Downarrow_n \}$, where we add the symbol $*$.
+
+The $*$ symbol stands for the nucleotides between a $\Downarrow_j$
+terminator ($0 \leq j \leq n)$ and the following non-skipped position.
+Recall that seed can only start at non-skipped positions occurring every
+$n+1$ nucleotides, and that the $\Downarrow_j$ indicates a sequencing
+error $j$ nucleotides before a non-skipped position.
+Figure~\ref{fig_alt_skip} shows the read of Figure~\ref{fig:skip} in the
+alternative skip-3 alphabet. The $*$ symbols appear after the
+$\Downarrow_1$ and $\Downarrow_3$ terminators. They are not found after a
+$\Downarrow_0$ terminator because there are no nucleotides between this
+terminator and the following non-skipped position.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.85]{alternative_sketch_skip.pdf}
+\caption{\textbf{The alternative skip encoding.}
+The read of Figure~\ref{fig:skip} is represented in the alternative skip-3
+alphabet. The only difference is that the nucleotides between a
+$\Downarrow_j$ terminator ($0 \leq j \leq n$) and the following
+non-skipped position are replaced by the $*$ symbol.}
+\label{fig_alt_skip}
+\end{figure}
+
+The match status of the $*$ nucleotides is irrelevant. Indeed, after a
+$\Downarrow_j$ symbol one has to ``wait''until the next potential seed can
+start, $j$ nucleotides further down. The weighted generating function of a
+$*$ nucleotide is thus $z$, as it stands of any nucleotide.
+
+A sequence of $*$ symbols can end in only two ways: either with the end of
+the read (as in the case of Figure~\ref{fig_alt_skip}), or with a
+non-skipped position. If the end of the read comes first, the sequence of
+$*$ symbols is terminated by the usual $|$ terminator. Otherwise, the
+sequence has no terminator but it puts the read in exactly the same state
+as after a $\Downarrow_0$ terminator.
+
+This means that in the alternative transfer matrix, the entries associated
+to $\Downarrow_j$ ($1 \leq j \leq n)$ simply ``fast forward'' to the
+entries associated with $\Downarrow_0$ and $|$. It takes exactly $j$
+nucleotides of any kind to go from a $\Downarrow_j$ symbol to the next
+non-skipped position, with weighted generating function $z^j$. If the end
+of the read occurs before the next non-skipped position, the
+$\Downarrow_j$ will be followed by a tail segment of up to $j-1$ symbols
+$*$, with weighted generating expression $1 + z + \ldots + z^{j-1}$. The
+segments following the $\Downarrow_0$ terminator are unchanged compared to
+the original expression of the transfer matrix.
+
+Finally, the expression of  the transfer matrix $M_n^*(z)$ in the
+alternative skip-$n$ alphabet is
+\begin{equation}
+\label{eq_alt_Mn}
+\begin{blockarray}{cccccc}
+   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
+\begin{block}{c[ccccc]}
+\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
+\Dn{1} & z & 0 & \ldots & 0 & N_0(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\Dn{n} & z^n & 0 & \ldots & 0 & N_{n-1}(z) \\
+\nd & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation}
+where $p$ is the error rate of the sequencer, $q=1-p$, where $n$ is the
+number of skipped nucleotides between potential seeds, where $\gamma$ is
+the minimum seed length, and where
+\begin{gather}
+\tag{\ref{eq_H}}
+H_{0,j}(z) = (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
+  \ldots + (qz)^{m(n+1)} \big), \\
+\notag
+  \text{with } x = n-j \modulo{n+1},
+  \text{ and } m = \left\lfloor
+  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
+\tag{\ref{eq_J}}
+J_0(z) = 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma-1}, \\
+\tag{\ref{eq_N}}
+N_i(z) = 1+z+\ldots+z^i.
+\end{gather}
+
+Observe that this expression is simpler than (\ref{eq_matrix_Mn}). It
+could have been used in section \ref{sec:example_skip} because it is also
+easier to derive, but the burden of computations with expression
+(\ref{eq_alt_Mn}) is heavier. The reason is that the top right entry of
+$M^*_n(z)^{s+1}$ is not the weighted generating function of reads with $s$
+sequencing errors ($s \geq 1$), so we cannot use the probability that the
+reads contains more than $s$ errors as a bound on the neglected terms. As
+a consequence, we would need to perform more iterations in order to
+ascertain that the estimates are accurate to within chosen $\varepsilon$.
+
 \subsection{Proof that $a_k$ is independent of $M_n(z)^m$ for $m > k+1$}

### 2020-03-20 manuscript.tex
Add detail for off-traget skip seeds

@@
 \label{eq_H}
-H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
-  \ldots + (qz)^{m(n+1)} \big), \\
+H_{i,j}(z) &=& (qz)^x \big( 1 + (qz)^{n+1} +
+  \ldots + (qz)^{m(n+1)} \big)pz, \\
 \notag
@@
 has no seed for the target or for the first duplicate---numbering is
-arbitrary here, the first duplicate can be any fixed duplicate. As
-illustrated in section~\ref{sec:symbolic}, we first recode the reads using
-a specialized alphabet to simplify the problem.
+arbitrary here, the first duplicate can be any fixed duplicate. As in
+section~\ref{sec:symbolic}, we first recode the reads using a specialized
+alphabet to simplify the problem.
 
@@
 terminator), except that now the terminators are the symbols
-$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and $\Downarrow$ ($j \geq 1$).
+$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ ($j \geq 1$) and $\Downarrow$.
 The tail, as usual, is terminated by the symbol $|$.
@@
 $\downarrow_{/j}^-$ ($j \geq 1$) represent a mismatch against the $-$
-sequence, the symbols $\downarrow_{/j}^+$ ($i \geq 1$) represent a
+sequence, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) represent a
 mismatch against the $+$ sequence, and the symbol $\Downarrow$ represents
@@
 
-With these definitions, $a$, $b$, $c$ and $d$ are the probabilities of
-occurrence of the symbols $\square$, $\downarrow_{/j}^-$,
-$\downarrow_{/j}^+$ and $\Downarrow$, respectively ($j \geq 1$). The next
-sections clarify how this is used to compute the weighted gerating
+With these definitions, the weighted generating functions of the symbols
+$\square$, $\downarrow_{/j}^-$, $\downarrow_{/j}^+$ ($j \geq 1$) and
+$\Downarrow$ are $az$, $bz$, $cz$ and $dz$, respectively. The next
+sections clarify how this is used to compute the weighted generating
 functions of interest.
@@
 
-If the next terminator is $\downarrow_{/i}^+$, there is a match of size
-$i$ for the $-$ sequence, so the segment contains $i-1$ symbols
-$\square$ (one of the matches is accounted for by the terminator). The
-weighted generating function is thus $(az)^{i-1}cz$. Likewise, if the next
-terminator is $\downarrow_{/i}^-$, the weighted generating function of the
-segment is $(az)^{i-1}bz$. The terminators $\downarrow_{/i}^+$ and
-$\downarrow_{/i}^-$ are disallowed for $j \geq \gamma$ because this would
-create a match of size $\gamma$ or greater for one of the sequences.
-
-\subsection{Segments following $\downarrow_{/j}^+$}
-
-At a $\downarrow_{/j}^+$ terminator, the match length of the $+$ sequence
-is 0 and the match length of the $-$ sequence is $j$. The next segment can
+If the next terminator is $\downarrow_{/j}^-$, there is a match of length
+$j$ for the $+$ sequence, so the segment contains $j-1$ symbols $\square$
+plus the terminator (which also matches the $+$ sequence). The weighted
+generating function is thus $(az)^{j-1}bz$. By the same rationale, if the
+next terminator is $\downarrow_{/j}^+$, the weighted generating function
+of the segment is $(az)^{j-1}cz$.
+
+The terminators $\downarrow_{/j}^+$ and $\downarrow_{/j}^-$ are disallowed
+for $j \geq \gamma$ because this would create a seed for at least one of
+the sequences.
+
+\subsection{Segments following $\downarrow_{/i}^+$}
+
+At a $\downarrow_{/i}^+$ terminator, the match length of the $+$ sequence
+is 0 and the match length of the $-$ sequence is $i$. The next segment can
 thus have up to $\gamma-1$ matches for the $+$ sequence, but only
-$\gamma-j-1$ matches for the $-$ sequence, lowering the maximum size of
+$\gamma-i-1$ matches for the $-$ sequence, lowering the maximum size of
 the segment. If the next terminator is $\Downarrow$ or $|$, the segments
@@
 
-If the next terminator is $\downarrow_{/i}^-$, the weighted generating
-function is $(az)^{i-1}bz$ as in the previous section. The difference is
-that the terminators $\downarrow_{/i}^-$ are allowed only for $1 \leq i
-\leq \gamma-j$, otherwise this would create a match of size $\gamma$ or
-greater for the $-$ sequence.
-
-If the next terminator is $\downarrow_{/i}^+$, the situation is slightly
-more complicated, because this imposes $j < i \leq \gamma-1$. Indeed,
-there were already $j$ matches for the $-$ sequence at the terminator
-$\downarrow_{/j}^+$, and there will be more at the end of the segment
-because it has no mismatch for the $-$ sequence. Taking this into account,
-we see that the weighted generating function of those segments is
-$(az)^{i-j-1}cz$.
-
-\subsection{Segments following $\downarrow_{/j}^-$}
+If the next terminator is $\downarrow_{/j}^-$, the weighted generating
+function is $(az)^{j-1}bz$ as in the previous section. The difference is
+that the terminators $\downarrow_{/j}^-$ are allowed only for $1 \leq j
+\leq \gamma-i$, otherwise this would create seed for the $-$ sequence.
+
+If the next terminator is $\downarrow_{/j}^+$, the situation is slightly
+more complex because this imposes $i < j \leq \gamma-1$. Indeed,
+there were already $i$ matches for the $-$ sequence at the terminator
+$\downarrow_{/i}^+$, and there will be more at the end of the following
+segment because it has no mismatch for the $-$ sequence. Taking this into
+account, we see that the weighted generating function of those segments is
+$(az)^{j-i-1}cz$ with $i < j \leq \gamma-1$.
+
+\subsection{Segments following $\downarrow_{/i}^-$}
 
@@
 generating function of the segments terminated by $\Downarrow$ or $|$ are
-$(1 + az + \ldots + (az)^{\gamma-j-1})dz$ and $1 + az + \ldots +
-(az)^{\gamma-j-1}$, respectively.
+$(1 + az + \ldots + (az)^{\gamma-i-1})dz$ and $1 + az + \ldots +
+(az)^{\gamma-i-1}$, respectively.
 
 Likewise, the weighted generating function of the segments terminated by
-$\downarrow_{/i}^+$ is $(az)^{i-1}cz$, where $1 \leq i \leq \gamma-j$; and
+$\downarrow_{/j}^+$ is $(az)^{j-1}cz$, where $1 \leq j \leq \gamma-i$; and
 the weighted generating function of the segments terminated by
-$\downarrow_{/i}^-$ is $(az)^{i-j-1}bz$, where $j < i \leq \gamma-1$.
+$\downarrow_{/j}^-$ is $(az)^{j-i-1}bz$ where $i < j \leq \gamma-1$.
 
@@
 
-We now have all the elements to specify the transfer matrix or reads with
+We now have all the elements to specify the transfer matrix of reads with
 no seed for either sequence. Recall that $a$ is the probability of a
@@
 \end{equation*}
-where $\tilde{A}(z)$, $\tilde{B}_0(z)$, $\tilde{C}_0(z)$ and
-$\tilde{D}(z)$ are matrices of dimensions $(\gamma-1) \times (\gamma-1)$
-that are defined as
+where $\gamma$ is the minimum seed length, and where $\tilde{A}(z)$,
+$\tilde{B}_0(z)$, $\tilde{C}_0(z)$ and $\tilde{D}(z)$ are matrices of
+dimensions $(\gamma-1) \times (\gamma-1)$ that are defined as
 \begin{equation*}
@@
 sequences of segments that can follow the symbol $\Downarrow$ and that are
-terminated by a tail, the weighted generating function of which is located
-in the top right entry of the matrix.
+terminated by a tail, the weighted generating function of which is the top
+right entry of the matrix.
 
@@
 arithmetic of truncated polynomials. Since each segment except the tail
-contains a mismatch against at least one sequence, we define $\tilde{p}$
-as the upper bound on the probability of a mismatch, \textit{i.e.}
-$\tilde{p} = \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now
-gives an upper bound of the probability that a read of size $k$ contains
-$s$ or more mismatches as
+contains a mismatch against at least one sequence, the top right entry of
+$\tilde{M}_0(z)^{s+1}$ is the weighted generating function of reads that
+contain $s$ mismatches (against any of the two sequences). We thus define
+$\tilde{p}$ as the upper bound on the probability of a mismatch,
+\textit{i.e.} $\tilde{p} = \max\{b,c,d\}$. The updated formula
+(\ref{eq:bound}) now gives an upper bound of the probability that a read
+of size $k$ contains $s$ or more mismatches as
 \begin{equation*}
@@
 \end{equation*}
+
+With this upper bound, we can compute the terms of the matrix partial sums
+$\tilde{M}_0(z) + \tilde{M}_0(z)^2 + \ldots + \tilde{M}_0(z)^s$ until the
+neglected terms become negligible, \textit{i.e.} until we can be sure that
+the coefficient of interest $a_k$ is accurate to within chosen
+$\varepsilon$.
 
@@
 With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
-compute its powers until the right-hand side of the expression above is
-lower than a set fraction $\varepsilon$ of the current value of $a_k$,
-which gives an accurate estimate of $P(\overline{S}_0 \cap
-\overline{S}_1)$.
+compute its powers until the estimate of the coefficient of interest $a_k$
+is accurate enough, finally giving a numerical value for $P(\overline{S}_0
+\cap \overline{S}_1)$.
 
@@
 \caption{\textbf{Off-target seeding probabilities (exact seeds).} The
-surfaces show the probability that seeding is off target for exact seeds
-of size $\gamma=19$ in reads of size $k=50$ nucleotides sequenced with an
-error rate $p=0.01$. The divergence rate $\mu$, defined as the probability
-that a given duplicate differs from the target at any given position. Note
-the factor 1000 in the scale of the z-axis.} \label{fig:curves_exact}
+surfaces show the probability that seeding is off target as a function of
+the divergence rate $\mu$ and of the number $N$ of duplicates. The results
+are shown for exact seeds of size $\gamma=19$ and for reads of size $k=50$
+nucleotides sequenced with an error rate $p=0.01$. The divergence rate
+$\mu$ is defined as the probability that a given duplicate differs from
+the target at any given position. Note the factor 1000 in the scale of the
+z-axis.}
+\label{fig:curves_exact}
 \end{figure}
@@
 observation is that the probability that seeding is off target increases
-with $N$. This is also clearly from expression (\ref{eq:Poff}). This can
-also be understood intuitively because the probability of null seeding
-decreases when there are more potential candidates, and since the
-probability of on-target seeding does not change, the probability of
-off-target seeding must increase.
+with $N$. This is also clear from expression (\ref{eq:Poff}). This can
+also be understood intuitively because the probability of not seeding the
+target is fixed and the probability of having an empty candidate set
+decreases as $N$ increases. As a result the probability that the candidate
+set contains only invalid candidates increases with $N$.
 
@@
 larger, the duplicates are far from the target and they are unlikely to be
-in the candidate set. In expression (\ref{eq:Poff}), the only term that
-depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it is
-clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
+in the candidate set at all. In expression (\ref{eq:Poff}), the only term
+that depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it
+is clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
 maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
@@
 
-Again we recode the reads in a specialized alphabet. We define the
+As before, we recode the reads in a specialized alphabet. We define the
 skip-$n$ dual alphabet as $\tilde{\mathcal{A}}_n = \{\square, *, |,
@@
 $\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ ($j \geq 1$) have the same
-meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
-$\Downarrow_j$ ($0 \leq j \leq n$) indicate that both sequences have match
-length 0 and that the next non-skipped position is $j$ nucleotides
-further. The new symbol $*$ indicates that the nature of the nucleotide
-does not matter, as we will explain below.
+meaning as in the dual alphabet of section~\ref{sec:dual}, \textit{i.e.}
+the $\square$ symbol stands for a double match, the $|$ terminator marks
+the end of the read, the $\downarrow^-_{/j}$ symbol indicates a mismatch
+against the $-$ sequence with a match length of size $j$ for the $+$
+sequence, and conversely the $\downarrow^+_{/j}$ symbol indicates a
+mismatch against the $+$ sequence with a match of length $j$ for the $-$
+sequence. The symbols $\Downarrow_j$ ($0 \leq j \leq n$) indicate that
+both sequences have match length 0 and that the next non-skipped position
+is $j$ nucleotides further. The symbol $*$ indicates that it does not
+matter whether the nucleotide is a match or a mismatch, as we will explain
+below.
 
@@
 same nucleotide).
-
-Also observe that the symbol $\Downarrow_1$ is followed by a $*$ to
-indicate that this nucleotide has no influence on the presence of a seed,
-regardless 
-. The reason is that there
-is a double mismatch at the $\Downarrow_1$ symbol so there can be no seed.
-But since the next potential seed starts only at the next tick mark,
-\textit{i.e.} one nucleotide further downstream,
 
@@
 
+Also observe that the symbol $\Downarrow_1$ is followed by a $*$ symbol,
+indicating that it does not matter whether the nucleotide is a match or a
+mismatch for any of the two sequences. After the $\Downarrow_1$ symbol,
+both sequences have match length 0 and one has to ``wait'' for a potential
+new seed one nucleotide further downstream. As in appendix~\ref{app_Mn},
+the $\Downarrow_j$ symbols are followed by $j$ symbols $*$, unless the end
+of the read comes before the next non-skipped position. After a sequence
+of $*$ symbols, the read is either finished or at a non-skipped position.
+In the second case, the is in the same state as after a $\Downarrow_0$
+symbol, so we can consider that the $*$ symbol allows the terminators
+$\Downarrow_j$ ($1 \leq j \leq n$) to just ``fast forward'' to either
+$\Downarrow_0$ or $|$, as shown in appendix~\ref{app_Mn}.
+
 As in the previous section, let $a$, $b$, $c$ and $d$ denote the
 probabilities of a double match, a mismatch against $-$, a mismatch
-against $+$ and a double mismatch, respectively. This also corresponds to
-the probabilities of occurrence of the symbols $\square$,
-$\downarrow_{/j}^+$, $\downarrow_{/j}^-$ ($j \geq 1$) and $\Downarrow_j$
-($0 \leq j \leq n$), respectively. With this information, we compute the
-weighted generating functions of interest in the sections below.
-
-\subsection{Segments following $\Downarrow_j$ ($1 \leq j \leq n)$}
-
-After a $\Downarrow_j$ terminator, the match counter for both sequences is
-at zero and the next potential seed starts only $j \geq 1$ nucleotides
-further downstream. This means that the next $j$ nucleotides are just
-filler so it does not matter which sequence they match or mismatch.
-
-There are only two scenarios to consider: if the end of the read is within
-the next $j$ nucleotides, then
+against $+$ and a double mismatch, respectively. With these definitions,
+the weighted generating functions of the symbols $\square$,
+$\downarrow_{/j}^-$ and $\downarrow_{/j}^+$ ($j \geq 1$) are $az$, $bz$
+and $cz$, respectively. The weighted generating function of the $*$ symbol
+is $z$, and that of the symbols $\Downarrow_j$ ($0 \leq j \leq n$) will be
+worked out in the sections below.
+
+\subsection{Segments following $\Downarrow_i$ ($1 \leq i \leq n$)}
+
+A $\Downarrow_i$ terminator is followed by up to $i$ symbols $*$. If there
+are fewer than $i$ symbols $*$, the read is finished so the segment must
+be a tail. Recall that the weighted generating function of the $|$
+terminator is 1, so the weighted generating function of tail segments is
+$1 + z + \ldots + z^{i-1}$.
+
+If there are $i$ symbols $*$, the sequence ends at a non-skipped position,
+\textit{i.e.} in the same state as after a $\Downarrow_0$ terminator. This
+is not a segment proper, because there is no terminator, but in the
+transfer matrix the symbols $\Downarrow_i$ ($1 \leq i \leq n$) project
+directly to the symbol $\Downarrow_0$ with weighted generating function
+$z^i$.
 
@@
 
-Hello world!
+After a $\Downarrow_0$ symbol, all the counters are reset as in the
+beginning of the read. If the next terminator is $|$, the segment is a
+tail and we only have to make sure that it contains fewer than $\gamma$
+symbols $\square$, otherwise this would create a seed. The weighted
+generating function of tail segments following $\Downarrow_0$ is thus $1 +
+qz + \ldots + (qz)^{\gamma-1}$.
+
+If the next terminator is a $\downarrow_{/j}^-$ symbol ($j \geq 1$), the
+segment is a match of size $j$ for the $+$ sequence. This imposes $j <
+\gamma$ otherwise the segment would create a seed. Such segments consist
+of $j-1$ symbols $\square$ followed by the terminator, so their weighted
+generating function is $(az)^{j-1}bz$ with $1 \leq j < \gamma$.
+
+Conversely, if the next terminator is a $\downarrow_{/j}^+$ symbol ($j
+\geq 1$), we can apply the same rationale to see that the weighted
+generating function is $(az)^{j-1}cz$, where $1 \leq j < \gamma$.
+
+Finally, if the next terminator is a $\Downarrow_j$ symbol ($0 \leq j \leq
+n$), we must not only make sure that the segment contains fewer than
+$\gamma$ symbols $\square$, but also keep track of the position of the
+next non-skipped position (stored in index $j$). Here we can follow
+verbatim the rationale of section~\ref{sec:example_skip} where we replace
+the probability of a match (previously $q$) by that of a double match (now
+$a$), and the probability of a mismatch (previously $p$) by that a double
+mismatch (now $d$). Replacing the symbols in expression (\ref{eq_H}), we
+see that the weighted generating function is $(az)^x \cdot \big( 1 +
+(az)^{n+1} + \ldots + (az)^{(n+1)m} \big) dz$, where  $x = n-j
+\modulo{n+1}$ and $m = \left\lfloor (\gamma-1-x) / (n+1) \right\rfloor$.
+
+\subsection{Segments following $\downarrow_{/i}^-$}
+
+At a $\downarrow_{/i}^-$ terminator ($i \geq 1$), the read contains a
+match of length $i$ for the $+$ sequence. If the next segment is a tail,
+we must make sure that the match length for the $+$ sequence does not
+exceed $\gamma-1$. This means that we can have up to $\gamma-i-1$ symbols
+$\square$ and thus that the weighted generating function of the tail
+segments is $1 + (qz) + \ldots + (qz)^{\gamma-j-1}$.
+
+If the next terminator is a $\downarrow_{/j}^-$ symbol, we must have $j >
+i$ because there is no mismatch against the $+$ sequence. In this case, we
+must only make sure that the total match length for the $+$ sequence
+remains lower than $\gamma$. Such segments contain $j-i-1$ symbols
+$\square$ followed by the terminator so their weighted generating function
+is $(qz)^{i-j-1}bz$ with $i < j \leq \gamma-1$.
+
+For the last two types of terminators, we must pay attention to the fact
+that in general, a mismatch against $+$ that follows a mismatch against
+$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
+n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
+$\Downarrow_j$ if the two mismatches are within the same interval between
+non-skipped positions because both sequences have match length 0. If the
+mismatch against $+$ is in another interval, the terminator is
+$\downarrow_{/j}^+$ because the $-$ sequence has a positive match length.
+
+If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it cannot
+be in the same interval as the preceding $\downarrow_{/i}^-$ terminator,
+which imposes a lower bound on the size of the segment. Since at the
+$\downarrow_{/i}^-$ terminator the match length for $+$ was $i$, there
+must be a non-skipped position $i$ nucleotides before. The number of
+nucleotides from the $\downarrow_{/i}^-$ terminator to the next
+non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum segment
+length is $y+1$ There is also an upper bound on the size of the segment
+because the match length for the $+$ sequence cannot become higher than
+$\gamma-1$, imposing the length to be lower than $\gamma-i$. The shortest
+segment is terminated by $\downarrow_{/1}^+$, and the longest by
+$\downarrow_{/{\gamma-y-i}}^+$. Finally, the weighted generating function
+of segments terminated by $\downarrow_{/j}^+$ following the
+$\downarrow_{/i}^-$ terminator is $(az)^{y+j-1}cz$ with $1 \leq j \leq
+\gamma-y-i$. Note that for some value of $i$ and $y$, no $j$ can satisfy
+the last inequality.
+
+Finally, if the next terminator is $\Downarrow_j$ ($0 \leq j \leq n$) we
+have to distinguish two cases. The first is that the segment length is $y$
+nucleotides or shorter, where we have seen that a mismatch against $+$ is
+sufficient to produce the $\Downarrow_j$ terminator. The second case is
+that the segment length is above $y$ nucleotides, where a double mismatch
+is necessary to produce the the $\Downarrow_j$ terminator. We have seen
+above that the $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
+before the next non-skipped position, so in the first case, only the
+terminators from $\Downarrow_0$ to $\Downarrow_{y-1}$ are possible (this
+case is impossible if $y=0$). More specifically, short segment terminated
+by $\Downarrow_j$ have $y-j-1$ matches followed by a mismatch for the $+$
+sequence, with weighted generating function $(az)^{y-j-1}bz$.
 
@@
 N_i(z) &=& 1+z+\ldots+z^i, \\
-W_j(z) &=& dz \cdot (az)^x \cdot \big( 1 + (az)^{n+1} + \ldots +
-  (az)^{(n+1)m} \big), \\
-  &\quad& \text{where } x = n-j \modulo{n+1},
+W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
+  (az)^{(n+1)m} \big) dz, \\
+  \notag
+  &\quad& \text{where } x = -j-1 \modulo{n+1},
   \text{ and } m = \left\lfloor
-  \frac{\gamma-1-x}{n+1} \right\rfloor. \\
-U_{i,j}(z) &=& dz \cdot (az)^x \cdot
-  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
+  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
+U_{i,j}(z) &=& (az)^x
+  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
 \begin{cases}
@@
 \end{cases} \\
+  \notag
   &\quad& \text{where } x = -i-j-1 \modulo{n+1},
-  y = n-i+1 \modulo{n+1}, \\
+  y = -i \modulo{n+1}, \\
+  \notag
   &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
     \right\rfloor, \\
-V_{i,j}(z) &=& dz \cdot (az)^x \cdot
-  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
+V_{i,j}(z) &=& (az)^x 
+  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
 \begin{cases}
@@
 \end{cases} \\
+  \notag
   &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
@@
   \begin{cases}
-  cz \cdot (az)^{x+j-1} &\text{ if } i+j+x \leq \gamma \\
+  cz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
   0                     &\text{ otherwise,}
@@
   \begin{cases}
-  bz \cdot (az)^{x+j-1} &\text{ if } i+j+x \leq \gamma \\
+  bz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
   0                     &\text{ otherwise,}
   \end{cases} \\
-  \text{where } x &=& -i \modulo{n+1}, \text{ in both cases.} \\
+  \text{where } y &=& -i \modulo{n+1}, \text{ in both cases.} \\
 \end{eqnarray*}
@@
 \tag{\ref{eq_H}}
-H_{0,j}(z) = (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
-  \ldots + (qz)^{m(n+1)} \big), \\
+H_{0,j}(z) = (qz)^x\big( 1 + (qz)^{n+1} +
+  \ldots + (qz)^{m(n+1)} \big) pz, \\
 \notag
-  \text{with } x = n-j \modulo{n+1},
+  \text{with } x = -j-1 \modulo{n+1},
   \text{ and } m = \left\lfloor

### 2020-03-22 manuscript.tex
Add background on spaced seeds

@@
 
-\title{Calibrating seed-based alignment heuristics with Sesame}
+\title{Calibrating seed-based mapping heuristics with Sesame}
 
@@
 \begin{definition}
+\label{def_dup}
 The target is the DNA fragment that was actually sequenced. Duplicates are
@@
 The most popular heuristic for mapping DNA sequences is a filtration
-method called seeding. The principle is to first identify seeds,
-defined as short regions of high similarity between the read and the
-genome, and then use an exact alignment algorithm at the candidate
-locations to identify the optimum. This strategy was first proposed in
-FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to to search for
-homology in sequence databases.
+method called ``seed and extend''. The principle is to first identify
+seeds, defined as short regions of high similarity between the read and
+the genome, and then use an exact alignment algorithm at the seeded
+locations to evaluate the candidates and identify the optimum. This
+strategy was first proposed in FASTA~\cite{pmid2983426} and
+BLAST~\cite{pmid2231712} to to search for homology in sequence databases
+of proteins and DNA.
 
@@
 
-The rest of this document is organized so as to present the concepts by
-increasing level of complexity, meaning that we sometimes go back and
-forth between different types of seeding strategies. We first provide some
-background knowledge on different types of seeding methods, we then
-introduce the computational strategy based on transfer matrices, and
-finally we delve into each specific seeding method.
+The rest of the article is organized as follows: section~\ref{sec_seeds}
+presents common seeding strategies used in bioinformatics,
+section~\ref{sec:symbolic} presents the basic concepts of analytic
+combinatorics that will be required to compute seeding probabilities,
+sections~\ref{sec:offexact} to \ref{sec_offMEM} treat the cases of exact
+seeds, skip seeds and MEM seeds, three common seeding heuristic used for
+mapping, section~\ref{sec_practical} returns to the mapping problem and
+revisits the assumptions of the model, section~\ref{sec:sesame} presents
+Sesame, a C library implementing the main results of the theory, and
+finally section~\ref{sec_disc} provides some perspectives on the present
+work.
 
 \section{Seeds}
+\label{sec_seeds}
 
@@
 refer to an exact match. For instance, the algorithm
-PatternHunter~\cite{pmid11934743} uses ``spaced seeds'' that tolerate
+PatternHunter~\cite{pmid11934743} uses spaced seeds that tolerate
 mismatches. To avoid any confusion, we will adopt the convention below.
@@
 structures have no set size so they can be used to search for seeds of
-different lengths. For instance, the FM-index is a compact data structure
-based on the suffix array~\cite{manber1993suffix} and on the
+different lengths. This is the case of the FM-index, a compact data
+structure based on the suffix array~\cite{manber1993suffix} and on the
 Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
@@
 ambiguity because this problem has several practical solutions.
+
+It is important to mention the case of spaced seeds. Originally introduced
+in~\cite{califano1993flash} and popularized by
+PatternHunter~\cite{pmid11934743}, spaced seeds feature ``don't-care''
+positions allowing them to detect imperfect matches. Spaced seeds are
+represented by models such as `11011', here indicating that the seeds have
+length 5 and that the middle position is disregarded. At index time, the
+genome is scanned with the model so that nucleotides labelled `1' are
+concatenated and indexed. At search time, the query is scanned with the
+model and the concatenated nucleotides labelled `1' are looked up in the
+index.
+
+Spaced seeds of weight $w$ (the number of 1s in the model) have the same
+memory requirements as continuous seeds of length $w$ but they are more
+sensitive~\cite{li2006superiority}, making them very attractive for
+homology search. They were used in the first generation of short-read
+mappers~\cite{pmid18684737, pmid19461883, pmid19675096, pmid2945724}, but
+nowadays they find more applications in genome
+comparisons~\cite{pmid21209072, healy2016flak},
+metagenomics~\cite{pmid26209798, pmid27540266}, genome
+assembly~\cite{pmid26539459} and long-read mapping~\cite{pmid27079541}.
+
+The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
+Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
+supports continuous seeds. Some workarounds are
+available~\cite{horton2008dislex, gagie2017compressed} but they increase
+the memory footprint, explaining that short reads are typically mapped
+using continuous seeds. More generally, computing the sensitivity of
+spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
+martin2017faster}. It is theoreticlally possible to do this using the
+tools introduced below, but the complexity becomes prohibitive for seed
+models with several don't-care positions. For these reasons we will not
+treat the case of spaced seeds and we will give more detail in
+section~\ref{sec_disc} bout the kind of difficulties they represent for
+our approach.
 
@@
 A Maximal Exact Match (MEM) is a subsequence of the read that is present
-in the reference genome and that cannot be extended --- either because the
+in the reference genome and that cannot be extended---either because the
 read ends or because the extended subsequence is not in the genome.
@@
 there would be if the read were two nucleotides shorter on the right
-side. Indeed, in this case there would be a shared MEM seed matching the
+side. Indeed, in this case there would be a shared MEM seed matching
 the target and the first duplicate (provided $\gamma \leq 12$).
@@
 \leq j \leq n$) represent sequencing errors and $j$ indicates the number
-of nucleotides till the next non-skipped position (\textit{i.e.} $j=0$ for
-nucleotides immediately before a non-skipped position and $j=n$ for
+of nucleotides until the next non-skipped position (\textit{i.e.} $j=0$
+for nucleotides immediately before a non-skipped position and $j=n$ for
 nucleotides at a non-skipped position).
@@
 indicate non-skipped positions (the potential start of a seed). The number
-$j$ in $\Downarrow_j$ indicates the number of nucleotides till the next
+$j$ in $\Downarrow_j$ indicates the number of nucleotidesu until the next
 non-skipped position. For $j=0$ the next nucleotide is not skipped.
@@
 Those segments follow a $\Downarrow_i$-segment, so they start $i$
-nucleotides in front of the next non-skipped position. The shortest
-following segment that ends $j$ nucleotides before a non-skipped position
-has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
-equivalent to defining $\ell_0$ as $x+1$ where $x$ is the number of
-$\square$ symbols of the shortest $\Downarrow_j$-segment, \textit{i.e.} $x
-= i-j-1 \modulo{n+1}$.
+nucleotides before the next non-skipped position. The shortest following
+segment that ends $j$ nucleotides before a non-skipped position has length
+$\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is equivalent to
+defining $\ell_0$ as $x+1$ where $x$ is the number of $\square$ symbols of
+the shortest $\Downarrow_j$-segment, \textit{i.e.} $x = i-j-1
+\modulo{n+1}$.
 
@@
 \subsection{Transfer matrix}
+\label{sec_transmat_skip}
 
@@
 $\tilde{M}_0(z)^{s+1}$ is the weighted generating function of reads that
-contain $s$ mismatches (against any of the two sequences). We thus define
-$\tilde{p}$ as the upper bound on the probability of a mismatch,
+contain $s$ mismatches (where doule mismatches count as one). We thus
+define $\tilde{p}$ as the upper bound on the probability of a mismatch,
 \textit{i.e.} $\tilde{p} = \max\{b,c,d\}$. The updated formula
@@
 The second observation is that there exists a ``worst'' value of $\mu$
-situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
+situated around 0.070. When $\mu$ is much smaller, the duplicates tend to
 be exactly identical to the target, making it impossible that there is a
@@
 $\square$ and thus that the weighted generating function of the tail
-segments is $1 + (qz) + \ldots + (qz)^{\gamma-j-1}$.
+segments is $1 + (qz) + \ldots + (qz)^{\gamma-i-1}$.
 
@@
 $\Downarrow_j$ if the two mismatches are within the same interval between
-non-skipped positions because both sequences have match length 0. If the
-mismatch against $+$ is in another interval, the terminator is
-$\downarrow_{/j}^+$ because the $-$ sequence has a positive match length.
-
-If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it cannot
-be in the same interval as the preceding $\downarrow_{/i}^-$ terminator,
-which imposes a lower bound on the size of the segment. Since at the
-$\downarrow_{/i}^-$ terminator the match length for $+$ was $i$, there
-must be a non-skipped position $i$ nucleotides before. The number of
-nucleotides from the $\downarrow_{/i}^-$ terminator to the next
-non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum segment
-length is $y+1$ There is also an upper bound on the size of the segment
-because the match length for the $+$ sequence cannot become higher than
-$\gamma-1$, imposing the length to be lower than $\gamma-i$. The shortest
-segment is terminated by $\downarrow_{/1}^+$, and the longest by
+non-skipped positions because both sequences lovally have match length 0.
+If the mismatch against $+$ is in another interval, the terminator is
+$\downarrow_{/j}^+$ because in that case the $-$ sequence has a positive
+match length. For now, bear in mind that a mismatch against the $+$
+sequence only can produce a $\Downarrow_j$ terminator ($0 \leq j \leq n$),
+this will be important later.
+
+If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it must
+be separated from the  preceding $\downarrow_{/i}^-$ terminator by a
+non-skipped position, which imposes a lower bound on the size of the
+segment. Since at the $\downarrow_{/i}^-$ terminator the match length for
+$+$ was $i$, there must be a non-skipped position $i$ nucleotides before.
+The number of nucleotides from the $\downarrow_{/i}^-$ terminator to the
+next non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum
+segment length is $y+1$ There is also an upper bound on the size of the
+segment because the match length for the $+$ sequence cannot become higher
+than $\gamma-1$, imposing the length to be lower than $\gamma-i$. The
+shortest segment is terminated by $\downarrow_{/1}^+$, and the longest by
 $\downarrow_{/{\gamma-y-i}}^+$. Finally, the weighted generating function
@@
 Finally, if the next terminator is $\Downarrow_j$ ($0 \leq j \leq n$) we
-have to distinguish two cases. The first is that the segment length is $y$
-nucleotides or shorter, where we have seen that a mismatch against $+$ is
-sufficient to produce the $\Downarrow_j$ terminator. The second case is
-that the segment length is above $y$ nucleotides, where a double mismatch
-is necessary to produce the the $\Downarrow_j$ terminator. We have seen
-above that the $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
-before the next non-skipped position, so in the first case, only the
-terminators from $\Downarrow_0$ to $\Downarrow_{y-1}$ are possible (this
-case is impossible if $y=0$). More specifically, short segment terminated
-by $\Downarrow_j$ have $y-j-1$ matches followed by a mismatch for the $+$
-sequence, with weighted generating function $(az)^{y-j-1}bz$.
+have to distinguish two cases, depending on whether the segment ends with
+a double mismatch, or with a single mismatch against the $+$ sequence. For
+the first case we can apply the rationale of
+section~\ref{sec:example_skip}. Recall that the index $j$ indicates the
+number of nucleotides until the next non-skipped position. Let $\ell_0$ be
+the length of the shortest possible segment. The lengths of the other
+segments are of the form $\ell_0 + n+1, \ldots, \ell_0 + m(n+1)$, where
+the integer $m$ must be chosen so that the segment does not create a seed.
+At the $\downarrow_{/i}^-$ terminator, there is a match of length $i$ for
+the $+$ sequence so $m$ is the largest integer such that $i+\ell_0+m(n+1)
+< \gamma$, \textit{i.e.} $m = \lfloor(\gamma-i-\ell_0) / (n+1)\rfloor$.
+
+The $\downarrow_{/i}^-$ terminator is $y = -i \modulo{n+1}$ nucleotides
+before the next non-skipped position so the shortest segment has length
+$\ell_0 = y-j$ if $y > j$, and $n+1-j+y$ otherwise. This is equivalent to
+defining $\ell_0$ as $x+1$ where $x$ is the number of $\square$ symbols in
+the shortest segment, \textit{i.e.} $x=-i-j-1 \modulo{n+1}$. Summing the
+weighted generating functions of the individual segments, we find
+$(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz$.
+
+The last remaining issue is that a mismatch against the $+$ sequence only
+can produce a $\Downarrow_j$ terminator. This happens when the preceding
+$\downarrow_{/i}^-$ terminator is not separated from the mismatch by a
+non-skipped position (in this case both sequences locally have match
+length 0). The $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
+before the next non-skipped position, so if the terminator is from
+$\Downarrow_0$ to $\Downarrow_{y-1}$ we need to add the term
+$(az)^xbz$ to the previous weighted generating function. In
+conclusion, the weighted generating function of segments terminated by
+$\Downarrow_j$ is $(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz +
+\delta^+_{i,j}(z)$, where $\delta^+_{i,j}(z) = (az)^xbz$ if $j < y$ and
+$0$ otherwise.
+
+\subsection{Segments following $\downarrow_{/i}^+$}
+
+We can find the weighted generating functions by just reversing the $+$
+and $-$ signs in the previous section. This way we can see that the
+weighted generating function of tail segments is $1 + (qz) + \ldots +
+(qz)^{\gamma-i-1}$.
+
+Likewise, the weighted generating function of segments terminated by
+$\downarrow_{/j}^+$ is $(qz)^{i-j-1}bz$ with $i < j \leq \gamma-1$.
+
+The weighted generating function of segments terminated by
+$\downarrow_{/j}^-$ is $(az)^{y+j-1}dz$ with $y = -i \modulo{n+1}$ and $1
+\leq j \leq \gamma-y-i$.
+
+Finally, the weighted generating function of segments terminated by
+$\Downarrow_j$ is $(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz +
+\delta^-_{i,j}(z)$, where $\delta^-_{i,j}(z) = (az)^xcz$ if $j < y$ and
+$0$ otherwise.
 
@@
 
-As in the case of the skip encoding, a fair amount of bookkeeping is
-required to specify the weighted generating functions of interest.
-We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
-section~\ref{sec:dual}. For notational convenience, we define
+We now have all the elements to specify the transfer matrix of reads with
+no skip-$n$ seed for either sequence. Recall that $n$ is the number of
+skipped nucleotides, $\gamma$ is the minimum seed length, $a$ is the
+probabiliy of a double match, $b$ is the probability of a mismatch against
+the $-$ sequence only, $c$ is the probability of a mismatch against the
+$+$ sequence only and $d$ is the probability of a double mismatch. For
+notational convenience we define
 \begin{eqnarray}
@@
 
-With these definitions, one can check that the expression of the transfer
-matrix $\tilde{M}_n(z)$ is
+With these notations, the information from the previous sections can be
+summarized in the transfer matrix $\tilde{M}_n(z)$ equal to
 
@@
 The matrices $\tilde{A}(z)$ and $\tilde{C}(z)$ in the expression of
-$\tilde{M}_n(z)$ are the same as in section~\ref{sec:dual}. The matrices
-$\tilde{B}_n(z)$ and $\tilde{C}_n(z)$ are defined as
+$\tilde{M}_n(z)$ are the same as in section~\ref{sec:dual}. They are
+reproduced here for convenience.
+\begin{equation*}
+\tilde{A}(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^+} & 0 & r_0^+(z) & \ldots &
+    r_{\gamma-4}^+(z) & r_{\gamma-3}^+(z) \\
+\scriptstyle{\downarrow_{/2}^+} & 0 & 0 & \ldots &
+    r_{\gamma-5}^+(z) & r_{\gamma-4}^+(z) \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^+} & 0 & 0 & \ldots & 0 & r_0^+(z) \\
+\scriptstyle{\downarrow_{/\gamma-1}^+} & 0 & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray},
+\end{equation*}
+\begin{equation*}
+\tilde{C}_0(z) = 
+\begin{blockarray}{cccccc}
+   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
+    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
+    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
+\begin{block}{c[ccccc]}
+\scriptstyle{\downarrow_{/1}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & r_{\gamma-2}^+(z) \\
+\scriptstyle{\downarrow_{/2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    r_{\gamma-3}^+(z) & 0 \\
+\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
+\scriptstyle{\downarrow_{/\gamma-2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
+    0 & 0 \\
+\scriptstyle{\downarrow_{/\gamma-1}^-} & r_0^+(z) & 0 & \ldots & 0 & 0 \\
+\end{block}
+\end{blockarray}.
+\end{equation*}
+
+The matrices $\tilde{B}_n(z)$ and $\tilde{C}_n(z)$ are defined as
 \begin{equation*}
@@
 
-The computation is performed exactly as described in
-section~\ref{sec:dual}. We compute the successive powers of
-$\tilde{M}_n(z)$ in the arithmetic of truncated polynomials and stop the
-iterations using the same criterion. So there is no change other than the
-definition of the transfer matrix.
+Now returning to the problem of computing $P(\overline{S}_0 \cap
+\overline{S}_1)$, the $+$ sequence is interpreted as the target and the
+$-$ sequence as the duplicate. Based on the assumptions of the error model
+presented in section~\ref{sec:error}, this implies that $a =
+(1-p)(1-\mu)$, $b = (1-p)\mu$, $c = p\mu/3$, and $d = p(1-\mu/3)$.
+
+The computation is performed as described in section~\ref{sec:dual}. We
+compute the successive powers of $\tilde{M}_n(z)$ in the arithmetic of
+truncated polynomials and stop the iterations using the same criterion.
+The only modification is that the top right term of $\tilde{M}_n(z)^{s+1}$
+is not the weighted generating function of reads with $s$ mismatches
+(where double mismatches count as one). The reason is that the sequences
+of $*$ symbols are not terminated by a mismatch. However, that can be at
+most one sequence of $*$ symbols for each mismatch or double mismatch, so
+the reads described by the top right term of $\tilde{M}_n(z)^{s+1}$ have
+at least $\lfloor s/2 \rfloor$ mismatches.
+
+We thus define $\tilde{p}$ as the upper bound on the probability of a
+mismatch, \textit{i.e.} $\tilde{p} = \max\{b,c,d\}$ and use the updated
+formula (\ref{eq:bound}) from section~\ref{sec_transmat_skip}
+\begin{equation*}
+Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
+\frac{s}{k\tilde{p}} \right).
+\end{equation*}
+
+With this upper bound, we can compute the terms of the matrix partial sums
+$\tilde{M}_0(z) + \tilde{M}_0(z)^2 + \ldots + \tilde{M}_0(z)^{2s+1}$ until
+the neglected terms become negligible, \textit{i.e.} until we can be sure
+that the coefficient of interest $a_k$ is accurate to within chosen
+$\varepsilon$.
 
@@
 
-We illustrate the strategy delineated above Using the same settings as in
-section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=19$), except that
-we use skip-9 seeds instead of exact seeds.
+We illustrate the strategy delineated above using the same settings as in
+section~\ref{sec:illdual} (reads of length $k=50$, probability of
+sequencing error $p=0.01$ and seeds of minimum size $\gamma=19$), except
+that we replace exact seeds by skip-5 and skip-9 seeds.
 
@@
 \caption{\textbf{Off-target seeding probabilities (skip seeds).}
-The surfaces show the probability that seeding is off target for skip-9
-seeds of size $\gamma=19$. Read size and error rate are the same as in
-Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. The
-divergence rate $\mu$, defined as the probability that a given duplicate
-differs from the target at any given position. Note the factor 1000 in
-the scale of the z-axis.}
+The surfaces show the probability that seeding is off target for skip-5
+and skip-9 seeds of size $\gamma=19$. Read size and error rate are the
+same as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and
+$p=0.01$. The divergence rate $\mu$, defined as the probability that a
+given duplicate differs from the target at any given position. Note the
+factor 1000 in the scale of the z-axis.}
 \label{fig:curves_skip}
@@
 Figure~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
-from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
-have the same general aspect as those of Figure~\ref{fig:curves_exact}. The
-probability that seeding is off target increases with $N$. There is again
-a  worst value of $\mu$, because the maximum of $P(\overline{S}_0 \cap
-\overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value of
-$N$. But the value is different from that of Figure~\ref{fig:curves_exact}
-because $P(\overline{S}_0 \cap \overline{S}_1)$ is computed for skip-9
-seeds instead of exact seeds (here the worst value of $\mu$ is
-approximately 0.07).
-
-Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 9
+from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The surfaces
+have the same general aspect as in Figure~\ref{fig:curves_exact}. The
+probability that seeding is off target increases with $N$ and there is
+again a  worst value of $\mu$, because the maximum of $P(\overline{S}_0
+\cap \overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value
+of $N$. However, those values of $\mu$ are not constant (they are
+approximately 0.065 and 0.060 for skip-5 and skip-9 seeds, respectively).
+
+Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 5
 nucleotides increases the chances that seeding is off target by a factor
-approximately 4 (compare with Figure~\ref{fig:curves_exact}). It is not
-obvious why this should be the case: skip seeds reduce the probability of
-on-target seeding (skipping positions implies fewer on-target seeds) but
-increase the probability of null seeding (skipping positions implies fewer
-seeds overall). So the net effect on off-target seeding is not intuitive.
-
-And indeed, there are cases that skip seeds are \emph{less} prone to
-off-target seeding than exact seeds. This is the case for instance when
-$p=0.1$, where the ranking is inverted for the whole range of values
-considered in Figure~\ref{fig:curves_skip}.
-
-This information is critical for choosing the best seeding strategy. The
-off-target seeding probability is not the only criterion, though. Equally
-important considerations are the probability of on-target seeding and the
-computational resources required to implement a particular seeding
-strategy. The benefit of a theory to compute seeding probabilities is to
-have access to this knowledge.
+approximately 1.5, and skipping 9 nucleotides by a factor approximately
+1.8 (compare with Figure~\ref{fig:curves_exact}). Skipping nucleotides
+seems to have little effect, but this is not a general conclusion. Indeed,
+skipping nucleotides decreases the probability of on-target seeding
+(skipping positions implies fewer on-target seeds) but increases the
+probability of null seeding (skipping positions implies fewer seeds
+overall), so the effects must be evaluated on a case-by-case basis.
+Skipping nucleotides can even decrease the probability that seeding is
+off target. Concretely, for exact seeds of size $\gamma=19$ in reads of
+size $k=50$ with $N=10$ duplicates at divergence $\mu=0.1$ with a
+sequencing error $p=0.1$, the probability of off-target seeding is
+approximately 0.178 with exact seeds and 0.035 with skip-9 seeds, showing
+skipping nucleotides can have different effects.
+
+This kind of information is critical for choosing the best seeding
+strategy. Yet, the off-target seeding probability is not the only
+criterion. Equally important considerations are the probability of
+on-target seeding, the computational resources required to implement a
+particular seeding strategy and other sources existing of mapping errors
+(see discussion in section~\ref{sec:truevsbest}). The benefit of a theory
+to compute seeding probabilities is to have access to this knowledge.
 
 \section{Off-target MEM seeds}
+\label{sec_offMEM}
 
@@
 
-The symbols $\downarrow_{/m}$ indicate that the nucleotide is a sequencing
-error and $m \geq 0$ is the number of duplicates that \emph{match} the
-nucleotide. Since a sequencing error is always a mismatch against the
+The symbols $\downarrow_{/m}$ ($m \geq 0$) indicate that the nucleotide is
+a sequencing error and $m$ is the number of duplicates that \emph{match}
+the nucleotide. Since a sequencing error is always a mismatch against the
 target, the symbol $\downarrow_{/0}$ indicates that the nucleotide is a
@@
 indicate a change in masking status: the nucleotide is not masked but the
-previous is --- this happens when all the duplicates fail to extend beyond
+previous is---this happens when all the duplicates fail to extend beyond
 this position. The index $i \geq 1$ is the number of nucleotides since the
 last mismatch or since the beginning of the read. All the other
-nucleotides are represented by the symbol $\square$ and the symbol $|$ is
-appended to the end of the read as usual.
+nucleotides are represented by the symbol $\square$, implying that
+$\square$ symbols are never sequencing errors and always match the
+target. The symbol $|$ is appended to the end of the read as before.
 
@@
 \begin{definition}
-The probability that a symbol is $\downarrow_{/m}$ given that the
-nucleotide is a read error is
+Given the divergence rate $\mu$ and the number of duplicates $N$, the
+probability that a symbol is $\downarrow_{/m}$ given that the nucleotide
+is a read error is
 \begin{equation}
@@
 \begin{definition}
-The probability that a given duplicate contains a mismatch in a sequence
-of $j$ error-free nucleotides is
+Given the divergence rate $\mu$, the probability that a given duplicate
+contains a mismatch in a sequence of $j$ error-free nucleotides is
 \begin{equation}
@@
 \begin{definition}
-The probability that a given duplicate sequence contains a mismatch in a
-sequence of $j$ error-free nucleotides followed by an error is
+Given the divergence rate $\mu$, the probability that a duplicate sequence
+contains a mismatch in a sequence of $j$ error-free nucleotides followed
+by an error is
 \begin{equation}
@@
 The case of the $N-m-r$ soft masks is more complicated because they can
-vanish precisely on the terminator --- recall that in case $ii$) all the
+vanish precisely on the terminator---recall that in case $ii$) all the
 hard masks are assumed to vanish before. If the soft mask failed within
@@
 (\ref{eq:xi}) and (\ref{eq:eta}), the probability that a given soft mask
-fails within the first $j$ nucleotides is $\xi_j/\eta_j$ --- this is the
+fails within the first $j$ nucleotides is $\xi_j/\eta_j$---this is the
 conditional probability that it fails within the first $j$ nucleotides
@@
 \section{Practical considerations}
+\label{sec_practical}
 
@@
 \section{Discussion}
+\label{sec_disc}
 
@@
 \appendix
+
+\section{Definitions}
+\setcounter{definition}{0}
+
+For convenience we reproduce here the definitions introduced in the text.
+
+\begin{definition}
+The target is the DNA fragment that was actually sequenced. Duplicates are
+sequences of the genome that share homology with the target (in genetics
+they are often referred to as paralogs).
+\end{definition}
+
+\begin{definition}
+The candidate set obtained at the seeding step is said to be
+\begin{enumerate}[i)]
+\item ``on target'' if it contains the target,
+\item ``off target'' if it contains a duplicate but not the target, 
+\item ``null'' if it contains neither.
+\end{enumerate}
+\end{definition}
+
+\begin{definition}
+A seed is a subsequence of the read that has size at least $\gamma$
+(defined by the context of the problem) and that has at least one perfect
+match in the reference genome.
+\end{definition}
+
+\begin{definition}
+A Maximal Exact Match (MEM) is a subsequence of the read that is present
+in the reference genome and that cannot be extended---either because the
+read ends or because the extended subsequence is not in the genome.
+\end{definition}
+
+\begin{definition}
+A \emph{strict} MEM seed has a single match in the genome.
+A \emph{shared} MEM seed has several matches in the genome.
+\end{definition}
+
+\begin{definition}
+Let $\mathcal{A}$ be a set of combinatorial objects such that $a \in
+\mathcal{A}$ has a size $|a| \in \mathbb{N}$ and a weight $w(a) \in
+\mathbb{R}^+$. The weighted generating function of $\mathcal{A}$ is
+defined as
+
+\begin{equation}
+\tag{\ref{eq:GF1}}
+A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
+\end{equation}
+Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k \geq 0}$ such
+that
+
+\begin{equation*}
+A(z) = \sum_{k=0}^\infty a_k z^k.
+\end{equation*}
+
+By definition $a_k = \sum_{a \in A_k}w(a)$, where $A_k$ is the class of
+objects of size $k$ in $\mathcal{A}$. The number $a_k$ is the
+total weight of objects of size $k$.
+\end{definition}
+
+\begin{definition}
+\label{def:seg}
+A terminator is any symbol that is different from the symbol $\square$. A
+segment is a sequence of 0 or more $\square$ symbols followed by a
+terminator. The tail is the last segment of the read, where the terminator
+is always the special symbol $|$.
+\end{definition}
+
+\begin{definition}
+At a given position of the read, a duplicate is a \emph{hard mask} if its
+match length on the left side is strictly longer than the match length
+of the target. A duplicate is a \emph{soft mask} if it has the same match
+length as the target.
+\end{definition}
+
+\begin{definition}
+Given the divergence rate $\mu$ and the number of duplicates $N$, the
+probability that a symbol is $\downarrow_{/m}$ given that the nucleotide
+is a read error is
+\begin{equation}
+\tag{\ref{eq:omega}}
+\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m.
+\end{equation}
+\end{definition}
+
+\begin{definition}
+Given the divergence rate $\mu$, the probability that a given duplicate
+contains a mismatch in a sequence of $j$ error-free nucleotides is
+\begin{equation}
+\tag{\ref{eq:xi}}
+\xi_j = 1-(1-\mu)^j.
+\end{equation}
+This is the probability that a hard or soft mask vanishes within $j$
+correct nucleotides.
+\end{definition}
+
+\begin{definition}
+Given the divergence rate $\mu$, the probability that a duplicate sequence
+contains a mismatch in a sequence of $j$ error-free nucleotides followed
+by an error is
+\begin{equation}
+\tag{\ref{eq:eta}}
+\eta_j = 1-(1-\mu)^j\mu/3.
+\end{equation}
+This is the probability that a hard or soft mask vanishes within $j$
+correct nucleotides followed by a sequencing error.
+\end{definition}
+
 \section{Appendix}

### 2020-03-25 manuscript.tex
Add section about realistic error models

@@
 Spaced seeds of weight $w$ (the number of 1s in the model) have the same
-memory requirements as continuous seeds of length $w$ but they are more
+memory requirements as contiguous seeds of length $w$ but they are more
 sensitive~\cite{li2006superiority}, making them very attractive for
@@
 Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
-supports continuous seeds. Some workarounds are
+supports contiguous seeds. Some workarounds are
 available~\cite{horton2008dislex, gagie2017compressed} but they increase
 the memory footprint, explaining that short reads are typically mapped
-using continuous seeds. More generally, computing the sensitivity of
+using contiguous seeds. More generally, computing the sensitivity of
 spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
@@
 function of complex objects, whose coefficients $a_k$ are the
-probabilities of interest.
+probabilities of interest~\cite{regnier2000unified, nicodeme2002motif,
+sedgewick2013introduction}.
 
@@
 we recode the reads in alphabets of custom symbols and we specify a
-construction plan of the reads using a \emph{transfer matrix} $M(z)$. The
-transfer matrix specifies which types of segments can follow each other in
-the reads of interest: the entry at coordinates $(i,j)$ is the weighted
-generating function of segments of type $j$ that can be appended to
-segments of type $i$.
+construction plan of the reads using a so-called \emph{transfer matrix}
+$M(z)$. The transfer matrix specifies which types of segments can follow
+each other in the reads of interest: the entry at coordinates $(i,j)$ is
+the weighted generating function of segments of type $j$ that can be
+appended to segments of type $i$.
 
@@
 \hline
-50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
-50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
-50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
-100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
-100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
-100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
+50  &  1  & $9.2 \cdot 10^{-4}$ & $8.4 \cdot 10^{-5}$ & $91.7\%$ \\
+50  & 10  & $9.0 \cdot 10^{-3}$ & $7.9 \cdot 10^{-4}$ & $92.0\%$ \\
+50  & 100 & $8.1 \cdot 10^{-2}$ & $5.1 \cdot 10^{-3}$ & $94.1\%$ \\
+100 &  1  & $2.4 \cdot 10^{-5}$ & $7.3 \cdot 10^{-6}$ & $76.7\%$ \\
+100 & 10  & $2.9 \cdot 10^{-4}$ & $6.9 \cdot 10^{-5}$ & $80.1\%$ \\
+100 & 100 & $5.5 \cdot 10^{-3}$ & $5.7 \cdot 10^{-4}$ & $90.7\%$
 \end{tabular}
-\caption{Numerical simulating of mapping errors with MEM seeds. Reads are
+\caption{Numerical simulation of mapping errors with MEM seeds. Reads are
 mapped incorrectly if either the true location is not in the candidate set
@@
 varying the read size $k$ and number $N$ of duplicates as indicated in the
-table. Mapping errors of both types were recorded, and the fraction of the
-first to the total was computed.}
+table and using MEM seed of minimum size $\gamma = 19$. Mapping errors of
+both types were recorded, and the fraction of the first to the total was
+computed.}
 \label{table_mem}
@@
 \begin{table}
-% Exact seeds.
-\renewcommand{\arraystretch}{1.1}
-\centering
-\begin{tabular}{ccccc}
-k & N & Prob. seeding error & Prob. other error & \% seeding \\
-\hline
-50  &  1  & $3.3 \cdot 10^{-5}$  & $5.1 \cdot 10^{-4}$ & $6.0\%$ \\
-50  & 10  & $2.9 \cdot 10^{-4}$  & $5.0 \cdot 10^{-3}$ & $5.4\%$ \\
-50  & 100 & $1.0 \cdot 10^{-3}$  & $4.6 \cdot 10^{-2}$ & $2.2\%$ \\
-100 &  1  & $1.5 \cdot 10^{-10}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
-100 & 10  & $1.2 \cdot 10^{-9}$  & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
-100 & 100 & $2.6 \cdot 10^{-9}$  & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
-\end{tabular}
-\caption{My least favorite table.}
-\label{table_exact}
-\end{table}
-
-\begin{table}
 % Skip-9 seeds.
@@
 \hline
-50  &  1  & $1.3 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $20.1\%$ \\
-50  & 10  & $1.2 \cdot 10^{-3}$ & $4.8 \cdot 10^{-3}$ & $19.7\%$ \\
-50  & 100 & $5.5 \cdot 10^{-3}$ & $4.5 \cdot 10^{-2}$ & $10.9\%$ \\
-100 &  1  & $2.0 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
-100 & 10  & $1.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
-100 & 100 & $3.9 \cdot 10^{-7}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
+50  &  1  & $1.7 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $25.4\%$ \\
+50  & 10  & $1.6 \cdot 10^{-3}$ & $4.7 \cdot 10^{-3}$ & $24.9\%$ \\
+50  & 100 & $7.9 \cdot 10^{-3}$ & $4.4 \cdot 10^{-2}$ & $15.2\%$ \\
+100 &  1  & $4.4 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $0.2\%$ \\
+100 & 10  & $3.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $0.2\%$ \\
+100 & 100 & $1.0 \cdot 10^{-6}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
 \end{tabular}
-\caption{My not so favorite table.}
+\caption{Numerical simulation of mapping errors with skip-9 seeds. Reads
+are mapped incorrectly if either the true location is not in the candidate
+set (not seeded) or it is in the candidate set but is not the best
+location (seeded, not best). Ideal reads were randomly generated 1,000,000
+times with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
+varying the read size $k$ and number $N$ of duplicates as indicated in the
+table and using skip-9 seed of minimum size $\gamma = 19$. Mapping errors
+of both types were recorded, and the fraction of the first to the total
+was computed.}
 \label{table_skip}
@@
 skip seeds.
+
+\subsection{More realistic error models}
+
+In order to make the model tractable, we had to make some simplifying
+assumptions regarding the distribution of errors
+(section~\ref{sec:error}). In particular, we assumed that sequencing
+errors are uniform on the read, which is known to not be the case with
+current instruments. The errors are typically more frequent at the ends of
+the reads, meaning that seeds may be longer than expected, impacting the
+probability that seeding is off target.
+
+To test whether this is the case, we used the ART
+simulator~\cite{pmid22199392} to emulate the error distribution of
+Illumina reads as per technology standards of 2016. We used the settings
+for the Illumina HiSeq 2000 and generated random reads of size 50 and 100
+from the human genome. We estimated the probability that seeding is off
+target from simulations and compared the results to calculations performed
+by Sesame (ART suggests error rates $p=0.0052$ and $p=0.0075$ for reads of
+size 50 and 100, respectively).
+
+Table~\ref{table_MEM_Illumina} shows the results for MEM seeds of minimum
+size $\gamma=19$. The estimates show some discrepancy but Sesame is
+accurate to within 15\% in these conditions.
+Table~\ref{table_skip_Illumina} shows the same type of comparison for
+skip-9 seeds of size $\gamma=19$. The values are substantially lower than
+for MEM seeds because skip-9 seeds are more sensitive. In these
+conditions, Sesame is accurate to within 35\% but it is important to
+highlight that simulations are unreliable for events with very low
+probability.
+
+\begin{table}
+% MEM seeds.
+\renewcommand{\arraystretch}{1.1}
+\centering
+\begin{tabular}{cccc}
+k & N & Simulation & Sesame \\
+\hline
+50  &  1  & $3.8 \cdot 10^{-4}$ & $4.5 \cdot 10^{-4}$ \\
+50  & 10  & $4.4 \cdot 10^{-3}$ & $4.5 \cdot 10^{-3}$ \\
+50  & 100 & $4.2 \cdot 10^{-2}$ & $4.0 \cdot 10^{-2}$ \\
+100 &  1  & $3.5 \cdot 10^{-5}$ & $3.7 \cdot 10^{-5}$ \\
+100 & 10  & $4.1 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
+100 & 100 & $8.6 \cdot 10^{-3}$ & $8.9 \cdot 10^{-3}$ 
+\end{tabular}
+\caption{Numerical simulation of ``seeding'' errors with MEM seeds.
+Realistic reads were randomly generated 1,000,000 times with sequencing
+error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
+$k$ and number $N$ of duplicates as indicated in the table and using
+skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
+were recorded, and the fraction of the first to the total was computed.}
+\label{table_MEM_Illumina}
+\end{table}
+
+\begin{table}
+% Skip seeds.
+\renewcommand{\arraystretch}{1.1}
+\centering
+\begin{tabular}{cccc}
+k & N & Simulation & Sesame \\
+\hline
+50  &  1  & $4.5 \cdot 10^{-5}$ & $4.5 \cdot 10^{-5}$ \\
+50  & 10  & $3.9 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
+50  & 100 & $1.9 \cdot 10^{-3}$ & $2.1 \cdot 10^{-3}$ \\
+100 &  1  & $< 10^{-6}$         & $< 10^{-6}$         \\
+100 & 10  & $2 \cdot 10^{-6}$   & $3 \cdot 10^{-5}$ \\
+100 & 100 & $6 \cdot 10^{-6}$   & $8 \cdot 10^{-4}$ 
+\end{tabular}
+\caption{Numerical simulation of ``seeding'' errors with skip-9 seeds.
+Realistic reads were randomly generated 1,000,000 times with sequencing
+error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
+$k$ and number $N$ of duplicates as indicated in the table and using
+skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
+were recorded, and the fraction of the first to the total was computed.}
+\label{table_skip_Illumina}
+\end{table}
+
+Overall, these results suggest that the relatively strong assumptions
+regarding the distribution of errors are not a major issue in practice.
+The answer of course depends on the type of sequencer that is used and on
+the particulars of the mapping problem.
+
+In practical applications, it is likely that other factors will be more
+detrimental for the precision of the estimates. For instance, the number
+of duplicates $N$ and their divergence rate $\mu$ are unknown, introducing
+an uncertainty that propagates to the Sesame estimates. Importantly, we
+assumed that duplicates evolve independently of each other and with
+uniform substitutions even in the simulations featured in
+Table~\ref{table_MEM_Illumina} and \ref{table_skip_Illumina} because more
+realistic models are tedious to implement. The burden of this assumption
+on the precision of the estimate in practical settings is difficult to
+gauge. Overall, Sesame estimates should be considered approximate in all
+applications with real biological data.
 
@@
 detected using Bayes formula, as shown in expression (\ref{eq:bayes}).
-
-\subsection{Realistic error patterns}
-
-This section will come later.
 
@@
 section~\ref{sec:random_seeds}.
+
+We have not treated the case of spaced seeds here. It is possible to find
+a transfer matrix for reads that contain no match for a given spaced seed,
+but as the number of don't-care positions increases, the dimensions of the
+matrix become prohibitively large. This is a well known difficulty. The
+analytic combinatorics approach presented here is equivalent to defining
+finite automata that recognize seed-containing reads. Automata can be used
+to compute the sensitivity of spaced seeds~\cite{kucherov2005multiseed,
+buhler2005designing}, with the caveat that such automata can contain a
+very large amount of states. Overall, the transfer matrix method does not
+provide any significant advantage over existing alternatives for spaced
+seeds.
 

### 2020-03-25 C/sim_mem.c

@@ -1,0 +1,173 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+#include <time.h>
+#include "mt.h"
+
+#define ITER 1000000
+#define GAMMA 19
+#define K 100
+#define N 100
+
+#define randombp() (1 + (randomMT() / 1431655765))
+
+static int POS[151] = {0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,
+   21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,
+   45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,67,68,
+   69,70,71,72,73,74,75,76,77,78,79,80,81,82,83,84,85,86,87,88,89,90,91,92,
+   93,94,95,96,97,98,99,100,101,102,103,104,105,106,107,108,109,110,111,112,
+   113,114,115,116,117,118,119,120,121,122,123,124,125,126,127,128,129,130,
+   131,132,133,134,135,136,137,138,139,140,141,142,143,144,145,146,147,148,
+   149,150};
+
+int shuffle (const void * a, const void * b) {
+  return (randomMT() < 2147483648) ? -1 : 1;
+}
+
+void print (char * seq) {
+  for (int i = 0 ; i < K ; i++) {
+    fprintf(stdout, "%d", seq[i]);
+  }
+  fprintf(stdout, "\n");
+}
+
+
+int main(int argc, char **argv) {
+
+//  char **ignore;
+//
+  size_t E    = atoi(argv[1]);
+//  double prob = strtod(argv[2], ignore); // p
+//
+  if (E == 0) {
+    fprintf(stderr, "argument error\n");
+    exit(EXIT_FAILURE);
+  }
+
+//  const double prob = 0.01;
+  const double mu   = 0.06;
+
+//  const unsigned long int p = (prob * 4294967295);
+  const unsigned long int m = (mu   * 4294967295);
+
+  // Set the random seed.
+  seedMT(123);
+
+  char dup[N+1][K] = {0};
+  char * read = dup[0];
+
+  int  err[N+1] = {0};      // Total errors.
+  int  str[N+1] = {0};      // Streak score.
+
+  int  falls[N+1] = {0};    // Which threads fall.
+  int  has_seed[N+1] = {0}; // Seeded duplicates.
+
[…]

### 2020-03-25 C/sim_mem_realistic.c

@@ -1,0 +1,164 @@
+#define _GNU_SOURCE
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+#include <time.h>
+#include "mt.h"
+
+#define GAMMA 19
+#define K 100
+#define N 1
+
+// Random number from 1 to 3 included.
+#define randombp() (1 + (randomMT() / 1431655765))
+
+void print (char * seq) {
+  for (int i = 0 ; i < K ; i++) {
+    fprintf(stdout, "%d", seq[i]);
+  }
+  fprintf(stdout, "\n");
+}
+
+
+int main(int argc, char **argv) {
+
+  const double mu   = 0.06;
+  const unsigned long int m = (mu * 4294967295);
+
+  FILE * mutfile = fopen(argv[1], "r");
+  if (mutfile == NULL) {
+     fprintf(stderr, "cannot open file %s\n", argv[1]);
+     exit(EXIT_FAILURE);
+  }
+
+  // Set the random seed.
+  seedMT(123);
+
+  char dup[N+1][K] = {0};
+  char * read = dup[0];
+
+  int  err[N+1] = {0};      // Total errors.
+  int  str[N+1] = {0};      // Streak score.
+
+  int  falls[N+1] = {0};    // Which threads fall.
+  int  has_seed[N+1] = {0}; // Seeded duplicates.
+
+  int total_case_1 = 0;
+  int total_case_2 = 0;
+
+
+  size_t sz = 151;
+  ssize_t len;
+  char * mut = malloc(sz);
+
+  // Run the simulation.
+  int ITER = 0;
+  while ((len = getline(&mut, &sz, mutfile)) != -1) {
+
+    ITER++;
+
+    // Erase seed info.
+    bzero(has_seed, (N+1) * sizeof(int));
+    
+    // Erase read and duplicates.
+    bzero(dup, (N+1) * K);
+    bzero(err, (N+1) * sizeof(int));
+    bzero(str, (N+1) * sizeof(int));
+
+    // Introduce errors in the read.
+    bzero(read, K);
+    int E = 0; // Number of errors.
+    for (int i = 0 ; i < K ; i++) {
+      if (mut[i] == '1') {
+         read[i] = randombp();
+         E++;
+      }
+    }
+
+    int dominant = 0;
+
+    for (int i = 0 ; i < K ; i++) {
+      // See which threads fall.
+      falls[0] = read[i] != 0;
[…]

### 2020-03-25 C/sim_skip.c

@@ -1,0 +1,139 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+#include <time.h>
+#include "mt.h"
+
+#define ITER 1000000
+#define GAMMA 19
+#define K 100
+#define N 100
+#define skip 9
+
+#define randombp() (1 + (randomMT() / 1431655765))
+
+static int POS[151] = {0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,
+   21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,
+   45,46,47,48,49,50,51,52,53,54,55,56,57,58,59,60,61,62,63,64,65,66,67,68,
+   69,70,71,72,73,74,75,76,77,78,79,80,81,82,83,84,85,86,87,88,89,90,91,92,
+   93,94,95,96,97,98,99,100,101,102,103,104,105,106,107,108,109,110,111,112,
+   113,114,115,116,117,118,119,120,121,122,123,124,125,126,127,128,129,130,
+   131,132,133,134,135,136,137,138,139,140,141,142,143,144,145,146,147,148,
+   149,150};
+
+int shuffle (const void * a, const void * b) {
+  return (randomMT() < 2147483648) ? -1 : 1;
+}
+
+void print (char * seq) {
+  for (int i = 0 ; i < K ; i++) {
+    fprintf(stdout, "%d", seq[i]);
+  }
+  fprintf(stdout, "\n");
+}
+
+
+int main(int argc, char **argv) {
+
+//  char **ignore;
+//
+  size_t E    = atoi(argv[1]);
+//  double prob = strtod(argv[2], ignore); // p
+//
+  if (E == 0) {
+    fprintf(stderr, "argument error\n");
+    exit(EXIT_FAILURE);
+  }
+
+//  const double prob = 0.01;
+  const double mu   = 0.06;
+
+//  const unsigned long int p = (prob * 4294967295);
+  const unsigned long int m = (mu   * 4294967295);
+
+  // Set the random seed.
+  seedMT(123);
+
+  char dup[N+1][K] = {0};
+  char * read = dup[0];
+
+  int  err[N+1] = {0};      // Total errors.
+  int  str[N+1] = {0};      // Streak score.
+
+  int  has_seed[N+1] = {0}; // Seeded duplicates.
+
+  int total_case_1 = 0;
+  int total_case_2 = 0;
+
+  // Run the simulation.
[…]

### 2020-03-25 C/sim_skip_realistic.c

@@ -1,0 +1,130 @@
+#define _GNU_SOURCE
+#include <stdio.h>
+#include <stdlib.h>
+#include <strings.h>
+#include <time.h>
+#include "mt.h"
+
+#define GAMMA 19
+#define K 100
+#define N 100
+#define skip 9
+
+// Random number from 1 to 3 included.
+#define randombp() (1 + (randomMT() / 1431655765))
+
+void print (char * seq) {
+  for (int i = 0 ; i < K ; i++) {
+    fprintf(stdout, "%d", seq[i]);
+  }
+  fprintf(stdout, "\n");
+}
+
+
+int main(int argc, char **argv) {
+
+  const double mu   = 0.06;
+  const unsigned long int m = (mu * 4294967295);
+
+  FILE * mutfile = fopen(argv[1], "r");
+  if (mutfile == NULL) {
+     fprintf(stderr, "cannot open file %s\n", argv[1]);
+     exit(EXIT_FAILURE);
+  }
+
+  // Set the random seed.
+  seedMT(123);
+
+  char dup[N+1][K] = {0};
+  char * read = dup[0];
+
+  int  err[N+1] = {0};      // Total errors.
+  int  str[N+1] = {0};      // Streak score.
+
+  int  has_seed[N+1] = {0}; // Seeded duplicates.
+
+  int total_case_1 = 0;
+  int total_case_2 = 0;
+
+
+  size_t sz = 151;
+  ssize_t len;
+  char * mut = malloc(sz);
+
+  // Run the simulation.
+  int ITER = 0;
+  while ((len = getline(&mut, &sz, mutfile)) != -1) {
+
+    ITER++;
+
+    // Erase seed info.
+    bzero(has_seed, (N+1) * sizeof(int));
+    
+    // Erase read and duplicates.
+    bzero(dup, (N+1) * K);
+    bzero(err, (N+1) * sizeof(int));
+    bzero(str, (N+1) * sizeof(int));
+
+    // Introduce errors in the read.
+    bzero(read, K);
+    int E = 0; // Number of errors.
+    for (int i = 0 ; i < K ; i++) {
+      if (mut[i] == '1') {
+         read[i] = randombp();
+         E++;
+      }
+    }
+
+    for (int i = 0 ; i < K ; i++) {
+      if (read[i] != 0) {
+        str[0] = (i % (skip+1)) - skip;
+        err[0]++;
+      }
+      else {
+        str[0]++;
+      }
[…]

### 2020-03-26 manuscript.tex
State of the manuscript on resubmission

@@
 
-\title{Calibrating seed-based mapping heuristics with Sesame}
+\title{Calibrating seed-based heuristics to map short reads with Sesame}
 
@@
 sequences of the genome that share homology with the target (in genetics
-they are often referred to as paralogs).
+they are often referred to as paralogs). In this article we will focus on
+short reads from complex eukaryotic genomes, so for concreteness the
+reader can assume that fragments are 30-300 bp long and that duplicates
+have above 75\% identity with the target.
 \end{definition}
@@
 between the sequence and the candidate location). This naturally leads to
-ask how we can identify the \emph{best} location of the fragment in
+asking how we can identify the \emph{best} location of the fragment in
 the genome.
@@
 Exact alignment algorithms were designed to solve the best location
-problem~\cite{pmid7265238, pmid5420325}, but they are too slow to process
+problem~\cite{pmid5420325, pmid7265238}, but they are too slow to process
 the large amount of data generated by modern sequencers. Instead, one uses
@@
 The most popular heuristic for mapping DNA sequences is a filtration
-method called ``seed and extend''. The principle is to first identify
+method called ``seed-and-extend''. The principle is to first identify
 seeds, defined as short regions of high similarity between the read and
 the genome, and then use an exact alignment algorithm at the seeded
-locations to evaluate the candidates and identify the optimum. This
-strategy was first proposed in FASTA~\cite{pmid2983426} and
-BLAST~\cite{pmid2231712} to to search for homology in sequence databases
-of proteins and DNA.
+locations to evaluate the candidates and identify the optimum. Exact
+alignment algorithms, such as the Needleman-Wunsch~\cite{pmid5420325} or
+the Smith-Waterman~\cite{pmid7265238} algorithms return the distance or
+the similarity between two sequences according to some quantitative
+criterion. They are exact in the sense that they always return the correct
+answer, unlike heuristic algorithms. The seed-and-extend strategy was
+first proposed in FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to
+search for homology in sequence databases of proteins and DNA.
 
@@
 \begin{definition}
-The candidate set obtained at the seeding step is said to be
+The output of the seeding step is the candidate set. The candidate set is
+the list of genomic locations where the read can be potentially mapped.
+The read is always mapped to one element of the candidate set. The seeding
+step is said to be
 \begin{enumerate}[i)]
-\item ``on target'' if it contains the target,
-\item ``off target'' if it contains a duplicate but not the target, 
-\item ``null'' if it contains neither.
+\item ``on target'' if the candidate set contains the target,
+\item ``off-target'' if the candidate set contains a duplicate but not the target, 
+\item ``null'' if the candidate set contains neither.
 \end{enumerate}
+In this article, we will always consider that a genomic location is in the
+candidate set if and only if the read contains at least one seed with a
+perfect match for this genomic location.
 \end{definition}
@@
 With our assumptions, a read is mapped to the wrong location if and only
-if the candidate set is off target. Indeed, if the candidate set is null,
+if the candidate set is off-target. Indeed, if the candidate set is null,
 the read is not mapped, and if the candidate set is on target, the correct
@@
 Focusing on popular heuristics to map short reads, our aim is to develop a
-method to estimate the probability that the candidate set is off target.
+method to estimate the probability that the candidate set is off-target.
 Previous work pioneered a method to compute seeding probabilities but it
@@
 seeds, skip seeds and MEM seeds, three common seeding heuristic used for
-mapping, section~\ref{sec_practical} returns to the mapping problem and
-revisits the assumptions of the model, section~\ref{sec:sesame} presents
-Sesame, a C library implementing the main results of the theory, and
-finally section~\ref{sec_disc} provides some perspectives on the present
-work.
+mapping, section~\ref{sec:sesame} presents Sesame, a C library
+implementing the main results of the theory, section~\ref{sec_practical}
+returns to the mapping problem and revisits the assumptions of the model,
+and finally section~\ref{sec_disc} provides some perspectives on the
+present work. Appendix~\ref{app_def} gathers for reference all the
+definitions encountered in the text, and appendix~\ref{app_extra} contains
+proofs and complements omitted from the main text.
 
@@
 (defined by the context of the problem) and that has at least one perfect
-match in the reference genome.
+match in the reference genome. Every genomic match of every seed is in the
+candidate set. 
 \end{definition}
@@
 spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
-martin2017faster}. It is theoreticlally possible to do this using the
+martin2017faster}. It is theoretically possible to do this using the
 tools introduced below, but the complexity becomes prohibitive for seed
@@
 treat the case of spaced seeds and we will give more detail in
-section~\ref{sec_disc} bout the kind of difficulties they represent for
+section~\ref{sec_disc} about the kind of difficulties they represent for
 our approach.
@@
 heuristic was used in the first version of BLASTN~\cite{pmid2231712}, but
-it has become unpopular for producing many redundant hits within the same
-regions of the read and of the genome.
+it has become unpopular for producing many short spurious hits.
 
 Figure~\ref{fig:exact_seed_example} shows the exact seeds from an example
-a read with three miscalled nucleotides. The sequenced DNA fragment has
+read with three miscalled nucleotides. The sequenced DNA fragment has
 three duplicates so the seeds can match four possible locations.
@@
 the genome because such random matches are unlikely to cause a mapping
-error when seeding is off target, contrary to matches in duplicates.
+error when seeding is off-target, contrary to matches in duplicates.
 Neglecting those will greatly simplify the exposition of the theory
@@
 section~\ref{sec:random_seeds} how to deal with this practical case and
-and how to identify those matches as off target. Until then, we will
-consider that the target and the duplicates are the only sequences in the
-reference genome.
+how to identify those matches as off-target. Until then, we will consider
+that the target and the duplicates are the only sequences in the reference
+genome.
 
@@
 To give an example, assume that a particular symbol, say $\Downarrow$, has
-a probability of occurence equal to $p$. The weighted generating function
-words containing only this symbol is $pz$. The weight of the word is
-its probability (here equal to $p$) and the size is its length in number
-of symbols (here 1).
+a probability of occurrence equal to $p$. The weighted generating function
+words containing only this symbol is $pz$. The weight of the word is its
+probability (here equal to $p$) and the size is its length (here 1).
 
@@
 of the set $\mathcal{A}$ of reads that do not have on-target seeds
-(\textit{i.e.} reads for which seeding is either null or off target). The
+(\textit{i.e.} reads for which seeding is either null or off-target). The
 weight of a read is its probability of occurrence and the size $k$ is its
@@
 weighted generating functions. One can thus obtain the weighted generating
-function of complex objects, whose coefficients $a_k$ are the
+function of complex objects, whose coefficients $a_k$ ($k \geq 0$) are the
 probabilities of interest~\cite{regnier2000unified, nicodeme2002motif,
@@
 \Downarrow\}$, where $\square$ represents a correct nucleotide,
-$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
-appended after the last nucleotide to mark the end of the read. It is not
-associated to a nucleotide and therefore has size 0.
+$\Downarrow$ represents an erroneous nucleotide and $|$ is a special
+symbol appended after the last nucleotide to mark the end of the read. It
+is not associated with a nucleotide and therefore has size 0.
 
@@
 symbol is $1-p = q$. Both symbols have size 1, so their respective
-weighted generating function are $pz$ and $qz$. Using the rule for
+weighted generating functions are $pz$ and $qz$. Using the rule for
 concatenation, we see that a segment of $i$ symbols $\square$ followed by
@@
 indicate non-skipped positions (the potential start of a seed). The number
-$j$ in $\Downarrow_j$ indicates the number of nucleotidesu until the next
+$j$ in $\Downarrow_j$ indicates the number of nucleotides until the next
 non-skipped position. For $j=0$ the next nucleotide is not skipped.
@@
 By construction, the Taylor expansion of the top right term in the matrix
-$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the the probabilities of interest.
-More specifically, if the Taylor series of this term is $a_0 + a_1 z + a_2
-z^2 + \ldots$, then $a_k$ is the probability that a read of size $k$
-contains no skip-$n$ seed of minimum size $\gamma$.
+$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the probabilities of interest. More
+specifically, if the Taylor series of this term is $a_0 + a_1 z + a_2 z^2
++ \ldots$, then $a_k$ is the probability that a read of size $k$ contains
+no skip-$n$ seed of minimum size $\gamma$.
 
@@
 We now turn our attention to the problem of computing the probability that
-the seeding process is off target when using exact seeds---recall from
+the seeding process is off-target when using exact seeds---recall from
 section~\ref{sec:twotypes} that off-target seeding means that the
@@
 
-If there is no duplicate (\textit{i.e.} $N = 0$), seeding cannot be off
-target, it can only be on target or null. So from here we assume that the
-target has $N \geq 1$ duplicates. Let $S_0$ denote the event that there is
-an on-target seed and let $S_j$ denote the event that there is a seed for
-the $j$-th duplicate. We are thus interested in computing
+If there is no duplicate (\textit{i.e.} $N = 0$), seeding cannot be
+off-target, it can only be on target or null. So from here we assume that
+the target has $N \geq 1$ duplicates. Let $S_0$ denote the event that
+there is an on-target seed and let $S_j$ denote the event that there is a
+seed for the $j$-th duplicate. We are thus interested in computing
 $P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big)$, where
@@
 \end{eqnarray}
-Combinging the two equations above, we obtain
+Combining the two equations above, we obtain
 \begin{equation}
@@
 
-Hence, the probability that seeding is off target is a function of just
-two quantitites: $P(\overline{S}_0)$ and $P(\overline{S}_0 \cap
+Hence, the probability that seeding is off-target is a function of just
+two quantities: $P(\overline{S}_0)$ and $P(\overline{S}_0 \cap
 \overline{S}_1)$. The first is the probability that the read has no seed
@@
 $P(\overline{S}_0 \cap \overline{S}_1)$, it can be computed at no
-additional cost. This probablity is less relevant than the probabilities
-that the seeding process is on target or off target, but at times, it may
+additional cost. This probability is less relevant than the probabilities
+that the seeding process is on target or off-target, but at times, it may
 be useful to know the probability that a read is not mappable, especially
@@
 \caption{\textbf{Off-target seeding probabilities (exact seeds).} The
-surfaces show the probability that seeding is off target as a function of
+surfaces show the probability that seeding is off-target as a function of
 the divergence rate $\mu$ and of the number $N$ of duplicates. The results
@@
 $N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
-observation is that the probability that seeding is off target increases
+observation is that the probability that seeding is off-target increases
 with $N$. This is also clear from expression (\ref{eq:Poff}). This can
@@
 
-To compute the probability that the seeding process is off target when
+To compute the probability that the seeding process is off-target when
 using skip seeds, we observe that the logic of section~\ref{sec:offexact}
@@
 $\Downarrow_j$ if the two mismatches are within the same interval between
-non-skipped positions because both sequences lovally have match length 0.
+non-skipped positions because both sequences locally have match length 0.
 If the mismatch against $+$ is in another interval, the terminator is
@@
 skipped nucleotides, $\gamma$ is the minimum seed length, $a$ is the
-probabiliy of a double match, $b$ is the probability of a mismatch against
-the $-$ sequence only, $c$ is the probability of a mismatch against the
-$+$ sequence only and $d$ is the probability of a double mismatch. For
-notational convenience we define
+probability of a double match, $b$ is the probability of a mismatch
+against the $-$ sequence only, $c$ is the probability of a mismatch
+against the $+$ sequence only and $d$ is the probability of a double
+mismatch. For notational convenience we define
 \begin{eqnarray}
@@
 \caption{\textbf{Off-target seeding probabilities (skip seeds).}
-The surfaces show the probability that seeding is off target for skip-5
+The surfaces show the probability that seeding is off-target for skip-5
 and skip-9 seeds of size $\gamma=19$. Read size and error rate are the
@@
 have the same general aspect as in Figure~\ref{fig:curves_exact}. The
-probability that seeding is off target increases with $N$ and there is
+probability that seeding is off-target increases with $N$ and there is
 again a  worst value of $\mu$, because the maximum of $P(\overline{S}_0
@@
 Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 5
-nucleotides increases the chances that seeding is off target by a factor
+nucleotides increases the chances that seeding is off-target by a factor
 approximately 1.5, and skipping 9 nucleotides by a factor approximately
@@
 Skipping nucleotides can even decrease the probability that seeding is
-off target. Concretely, for exact seeds of size $\gamma=19$ in reads of
+off-target. Concretely, for exact seeds of size $\gamma=19$ in reads of
 size $k=50$ with $N=10$ duplicates at divergence $\mu=0.1$ with a
@@
 
-We first introduce two important notions that will be the key of
+We first introduce two important notions that will be the key to
 understanding the behavior of MEM seeds.
@@
 \begin{definition}
+\label{def_mask}
 At a given position of the read, a duplicate is a \emph{hard mask} if its
@@
 Whenever a hard mask (here the first duplicate) covers the nucleotides as
-shown by the grey squres in the left panel, there can be no on-target
+shown by the grey squares in the left panel, there can be no on-target
 seed. The positions marked with a question mark are irrelevant, they
@@
 
-In conclusion, the probability that the MEM seeding process is off target
+In conclusion, the probability that the MEM seeding process is off-target
 is
@@
 the $N+1$ sequences one by one, one can generate a single sequence of
-segments. Since the number of segment does not depend on $N$, we can
+segments. Since the number of segments does not depend on $N$, we can
 obtain a fast Monte Carlo method to sample millions of reads and count the
@@
 
-We illustrate the strategy delineated above 
-using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
-and $\gamma=19$), except that we replace exact seed by MEM seeds.
+We illustrate the strategy delineated above using the same settings as in
+section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=19$), except that
+we replace exact seed by MEM seeds.
 
@@
 \caption{\textbf{Off-target seeding probabilities (MEM seeds).}
-The surfaces show the probability that seeding is off target for MEM seeds
+The surfaces show the probability that seeding is off-target for MEM seeds
 of indicated minimum size $\gamma$. Read size and error rate are the same
@@
 
-Figure~\ref{fig:curves_mem} shows the result for a number of duplicates $N$
-from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The curves
-have the same general aspect as those of Figure~\ref{fig:curves_exact}. The
-probability that seeding is off target increases with $N$, as shown by
-expression (\ref{eq:PMEM}).
-
-There is again a  worst value of $\mu$ but it is much lower than the
-previous two cases (here it is close to 0.026). In the case of MEM seeds,
-it is not obvious why the same value of $\mu$ maximizes (\ref{eq:PMEM})
-for all values of $N$ because both $P(\overline{M}_0)$ and
+Figure~\ref{fig:curves_mem} shows the result for a number of duplicates
+$N$ from 1 to 10, for a divergence rate $\mu$ from 0 to 0.20 and for MEM
+seeds of different minimum size $\gamma$. The surfaces have the same
+general aspect as those of Figure~\ref{fig:curves_exact}. The probability
+that seeding is off-target increases with $N$, as shown by expression
+(\ref{eq:PMEM}).
+
+There is again a  worst value of $\mu$ in each plot, but it is much lower
+than the previous two cases (it is close to 0.028 for $\gamma=14$i and
+$\gamma=19$, 0.030 for $gamma=27$ and .035 for $\gamma=34$). In the case
+of MEM seeds, it is not obvious why the same value of $\mu$ maximizes
+(\ref{eq:PMEM}) for all values of $N$ because both $P(\overline{M}_0)$ and
 $P(\overline{S}_0 \cap \overline{S}_1)$ depend on $\mu$.
@@
 Figure~\ref{fig:curves_mem} reveals that in this concrete case, MEM seeds
-increase the chances that seeding is off target by a factor up to 40
+increase the chances that seeding is off-target by a factor 15
 compared to exact seeds (see Figure~\ref{fig:curves_exact}). On this
 criterion, MEM seeds are always inferior to exact seeds with the same
-specifications. The reason is that every read triggering an off-target
-seeding with exact seeds will also trigger an off-target seeding with MEM
-seeds. Knowing that, it is still important to realize how large the
-increase is in practice. MEM seeds tremendously simplify the seeding
-process, but this comes at the cost of an increase in the false positive
-rate. The methods presented here are thus very useful to balance the
-risk when using MEM seeds.
+specifications. The reason is that a MEM seed always contains at least one
+exact seed of size $\gamma$. MEM seeds tremendously simplify the seeding
+process, but this comes at the cost of an increase of the probability that
+seeding is off-target.
+
+Also note that the values change less than one might expect when varying
+the minimum seed length $\gamma$. Part of the reason is that MEM seeds are
+typically longer than their minimum size $\gamma$. In any event, this
+example shows that reducing $\gamma$ to gain sensitivity provides very
+modest benefits, with potentially high costs in terms of short spurious
+hits.
+
+\section{The Sesame library}
+\label{sec:sesame}
+
+We implemented the methods and algorithms presented here in an open-source
+C library to compute seeding probabilities. The library is called Sesame
+and is available at \url{https://github.com/gui11aume/sesame}.
+
+\subsection{Main features of Sesame}
+
+Sesame contains functions to compute the seeding probabilities described
+here. All the functions were tested against simulations to ensure that the
+implementation is accurate and the code was checked extensively by static
+analysis and unit testing.
+
+Computing seeding probabilities can take up to a few seconds, even when
+replacing iterative methods by Monte Carlo simulations. This is
+incompatible with the speed requirements of modern mappers, so Sesame has
+an interface for this type of application. In this mode, Sesame computes
+the results only the first time and stores them in memory for reuse on
+subsequent calls.
+
+Storing the results in memory is an efficient strategy because three
+parameters are constant throughout the sequencing run: the minimum seed
+size $\gamma$, the read size $k$ and the error rate of the sequencer $p$
+(and for skip seeds, the number of skipped nucleotides $n$ is also
+constant). Only two parameters depend on the read: the number of
+duplicates $N$ and their divergence rate $\mu$. In this mode, Sesame
+automatically switches to Monte Carlo sampling when $N$ is large to save
+time. Also, the input parameters are ``snapped'' to a predefined grid of
+set values for $N$ and $\mu$, so that few computations are performed and
+most of the calls are actually memory lookups. Sesame can thus be
+integrated in short read mappers without being a bottleneck.
+
+Alternatively, the probabilities of interest can be computed offline,
+saved to disk and loaded at run time. This is particularly useful if the
+sequencing runs follow some standard conditions with a known error rate,
+because the computations can be recycled between runs.
+
+Finally, Sesame also has an offline interface, where seeding probabilities
+are computed exactly as requested by the users, \textit{i.e.} without
+modifying the algorithm or the parameters, and also without storing the
+results in memory.
+
+The Sesame manual, available from the repository, contains additional
+information and explains in detail how to use the library.
+
+\subsection{Using Sesame to compare seeding strategies}
+
+As a tool to compute seeding probabilities, Sesame can be used to compare
+the merits of different strategies. The kind of insight that we can gain
+from such calculations was already showcased in
+Figure~\ref{fig:curves_exact}, Figure~\ref{fig:curves_skip} and
+Figure~\ref{fig:curves_mem}, where the numbers were computed using Sesame.
+
+To further showcase the potential benefits of computing seeding
+probabilities, we use Sesame to compare the default seeding strategies of
+BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286}. Note that
+both mappers use advanced techniques to refine the seeds, so this
+comparison does not reflect the true performance of the mappers. It is
+nevertheless useful to know the baseline of each strategy. The default of
+BWA-MEM is to use MEM seeds of minimum size 19; that of Bowtie2 is to use
+skip-9 seeds of size 16.
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[scale=.41]{mortal_kombat.pdf}
+\end{center}
+\caption{\textbf{Comparing seeding strategies with Sesame.} Left: MEM
+seeds with $\gamma=19$ and $\mu=0.06$. Right: skip-9 seeds with
+$\gamma=16$ and $\mu=0.06$. The probabilities that seeding is off-target
+were computed using Sesame. Each curve represents the probability for a
+given number of duplicates ($N$). Estimates using iterative methods (skip
+seeds and MEM seeds where $N \leq 20$) were computed to within 1\%
+accuracy. Estimates using Monte Carlo sampling (MEM seeds where $N > 20$)
+were computed as the average of 500 million simulations.
+}
+\label{fig_mortal_kombat}
+\end{figure*}
+
+The probabilities that seeding is off-target for different read sizes $k$
+and different number of duplicates $N$ are plotted in
+Figure~\ref{fig_mortal_kombat}. The left panel shows the results for MEM
+seeds and the right panel shows the results for skip seeds. Here the error
+rate $p$ is set to 1\%, close to the specifications of the Illumina
+platform~\cite{pmid21576222}, and the divergence rate between duplicates
+$\mu$ is set to an arbitrary value of 6\%.
+
+The behaviors of the two types of seeds are dramatically different. Let us
+start with MEM seeds. For every value of $N$, the probability initially
+increases with the read size, and then drops exponentially. The initial
+increase is a hallmark of MEM seeds; it is due to the fact that duplicates
+can mask the target. Note that the asymptotic decay depends on the number
+of duplicates $N$ because each duplicate can mask the target and thereby
+reduces the probability that it is discovered. Overall, these results show
+that the performance of MEM seeds is poor when the target has more than
+approximately 20 duplicates.
+
+Turning to skip seeds, we see that the curves have a staircase look with a
+drop every 10 nucleotides. This is so because seeding probabilities remain
+unchanged until there is space for another seed of size 16 on the read.
+The curves are otherwise decreasing with a steady exponential trend where
+the asymptotic decay does not depend on the number of duplicates $N$. The
+reason is that duplicates do not prevent the target from being discovered,
+they merely fool the mapper when the target was not found. This property
+makes the asymptotic decay of skip seeds substantially faster that of MEM
+seeds when $N$ is large.
+
+Those properties are self-evident in retrospect, but they are not
+necessarily obvious from the definitions of MEM seeds and skip seeds.
+The main limitations of the MEM seeds are best understood by keeping in
+mind that their asymptotic decay depends on $N$.
+
+Figure~\ref{fig_mortal_kombat} suggests that skip-9 seeds of size 16 are
+just better than MEM seeds of size 19. However, the gain in sensitivity
+comes at the cost of a larger candidate set, slowing down the mapping
+process. To remain competitive, Bowtie2 further filters the candidate set
+using a priority policy. But since some candidates are not checked, the
+probability that seeding is off-target is larger than shown in
+Figure~\ref{fig_mortal_kombat}. Also, we will show in
+section~\ref{sec:truevsbest} that the higher sensitivity of skip seeds is
+not as big an advantage as it looks because seeding is not the only source
+of mapping errors.
+
+\subsection{Key insights about MEM seeds}
+
+At least two key insights about MEM seeds can be gained from
+Figure~\ref{fig_mortal_kombat}. The first is that it is not worth it for a
+MEM-based mapper to check all the candidate loci when there are more than
+approximately 20 of them. The mapper may find the correct location, but
+even if this is the case, the mapping quality will remain low because the
+prior chances of failure were high. A better strategy is to either bail
+out to not waste time, or to switch to a more sensitive seeding method
+(BWA opts for the second and uses a re-seeding policy). It is also
+important to note that this decision should be based on an estimated value
+of $N$ and not, for instance, on the size of the seed or some other
+variable.
+
+A second insight is that for MEM seeds, the off-target rate is always
+above $10^{-3}$ for reads of 50 nucleotides or fewer. Here it is important
+to mention that the value $\mu=0.06$ is not even the worst for reads of
+this size when $p=0.01$ (according to Figure~\ref{fig:curves_mem}, the
+worst value is around 0.02-0.03). So if $\mu$ is unknown and one wishes to
+be conservative, it seems that reads of 50 nucleotides cannot be mapped
+with confidence better than $1/1000$.
+
+However, this is not true for the reason that it is practically impossible
+to map an Illumina read of size 50 to the wrong location when $N=0$ (see
+section~\ref{sec:random_seeds}). Indeed, incorrect locations are unrelated
+sequences in this case, and it is easy to recognize that two sequences of
+size 50 are not homologous. This means that the highest impact one can
+have on the mapping quality is to check whether $N$ is 0, or in other
+words, whether the target is a unique sequence.
+
+These insights suggest that there is a way to make MEM seeds more useful
+for short read mappers. Following these principles, we have implemented a
+prototype mapper based on Sesame that shows good overall
+performance~\cite{zorita2020mapping}.
 
@@
 Early in the development of the theory, we distinguished the true location
-from the best location, but we assumed that they are identical in order to
-eliminate some practical considerations that would otherwise clutter the
-exposition. Here we have presented a framework to compute the probability
-that the \emph{true} location is in the candidate set. We have not said
-anything about the probability that the best location is in the candidate
-set, so our results do not help address the best location problem. Now it
-remains to see whether they help address the true location problem.
+from the best location. We swiftly assumed that they are identical in
+order to eliminate some practical considerations that would otherwise
+clutter the exposition. Throughout the article we have developed a
+framework to compute the probability that the \emph{true} location is in
+the candidate set. We have not mentioned anything about the probability
+that the \emph{best} location is in the candidate set, so our results do
+not address the best location problem. It remains to establish how they
+contribute to addressing the \emph{true} location problem.
 
 The issue at hand is that the candidates are tested with an alignment
-algorithm that returns the best location. So even if the target is in the
-candidate set, the read may be mapped to another location. The question is
-thus how often this happens. If this case is rare, our results give a good
-approximation of the mapping errors (where errors are failures to map the
-reads to their true location). Otherwise the estimates may not be so
-useful in practice.
+algorithm that returns the best location. So even if the true location is
+in the candidate set, the read may be mapped somewhere else because some
+other candidate has a better alignment score. The question is how often
+this happens. If this is a rare event, our results give a good
+approximation of the probability that a read is mapped to the true
+location. Otherwise the estimates may not be so useful in practice.
 
 The answer depends on the seed type. We start with MEM seeds because they
-will allow us to better explain what happens. The key insight is that MEM
-seeds tend to exclude the target from the candidate set if it is not the
-best location. To develop an intuition as to why this is the case,
-consider a read with a single sequencing error. We claim that the target
-was never in the candidate set if the read is mapped to an incorrect
-location (\textit{i.e.} not the true location).
-
-The argument goes as follows: if the target were in the candidate set,
-then there would exist a candidate that is a perfect match for the read.
-But then the read would be covered by a single MEM seed, the target would
-be masked and it could not be in the candidate set, leading to a
-contradiction.
+will allow us to highlight an important phenomenon. The key insight is
+that MEM seeds tend to exclude the target from the candidate set if it is
+not the best location. To develop an intuition as to why this is the case,
+consider a read with a single sequencing error. The only way a candidate
+can have a better score is to be a perfect match for the read. But in this
+case the true location is hard-masked (see definition~\ref{def_mask}) and
+it is not in the candidate set.
 
@@
 \begin{tabular}{ccccc}
-k & N & Not seeded & Seeded, not best & \% seed failure \\
+k & N & Not seeded & Seeded, not best & Relative error \\
 \hline
-50  &  1  & $9.2 \cdot 10^{-4}$ & $8.4 \cdot 10^{-5}$ & $91.7\%$ \\
-50  & 10  & $9.0 \cdot 10^{-3}$ & $7.9 \cdot 10^{-4}$ & $92.0\%$ \\
-50  & 100 & $8.1 \cdot 10^{-2}$ & $5.1 \cdot 10^{-3}$ & $94.1\%$ \\
-100 &  1  & $2.4 \cdot 10^{-5}$ & $7.3 \cdot 10^{-6}$ & $76.7\%$ \\
-100 & 10  & $2.9 \cdot 10^{-4}$ & $6.9 \cdot 10^{-5}$ & $80.1\%$ \\
-100 & 100 & $5.5 \cdot 10^{-3}$ & $5.7 \cdot 10^{-4}$ & $90.7\%$
+50  &  1  & $9.2 \cdot 10^{-4}$ & $8.4 \cdot 10^{-5}$ & $1.09$ \\
+50  & 10  & $9.0 \cdot 10^{-3}$ & $7.9 \cdot 10^{-4}$ & $1.09$ \\
+50  & 100 & $8.1 \cdot 10^{-2}$ & $5.1 \cdot 10^{-3}$ & $1.06$ \\
+100 &  1  & $2.4 \cdot 10^{-5}$ & $7.3 \cdot 10^{-6}$ & $1.30$ \\
+100 & 10  & $2.9 \cdot 10^{-4}$ & $6.9 \cdot 10^{-5}$ & $1.24$ \\
+100 & 100 & $5.5 \cdot 10^{-3}$ & $5.7 \cdot 10^{-4}$ & $1.10$
 \end{tabular}
-\caption{Numerical simulation of mapping errors with MEM seeds. Reads are
-mapped incorrectly if either the true location is not in the candidate set
-(not seeded) or it is in the candidate set but is not the best location
-(seeded, not best). Ideal reads were randomly generated 1,000,000 times
-with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
-varying the read size $k$ and number $N$ of duplicates as indicated in the
-table and using MEM seed of minimum size $\gamma = 19$. Mapping errors of
-both types were recorded, and the fraction of the first to the total was
-computed.}
+\caption{Simulations of mapping errors with MEM seeds. Ideal reads were
+randomly generated 1,000,000 times with sequencing error rate $p = 0.01$,
+divergence rate $\mu = 0.06$, varying the read size $k$ and number $N$ of
+duplicates as indicated, using MEM seeds of minimum size $\gamma = 19$.
+\textit{Not seeded} shows how often the read had no seed for the target,
+\textit{Seeded, not best} shows how often the read had a seed for the
+target but the target was not the best candidate, and \textit{Relative
+error} shows the factor between the probability that the read is not
+mapped to the true location and the probability that seeding is
+off-target.}
 \label{table_mem}
@@
 
-In short, either seeding is on target and the read is mapped to the true
-location, or seeding is off target (and the read is of course mapped
-somewhere else). When there are more than one errors, the strict
-equivalence breaks, but the trend remains the same. We can show this
-using simulations for realistic sequencing conditions
-(Table~\ref{table_mem}). For MEM seeds, the probability that the read is
-not mapped to the true location is approximately the same as the
-probability that the seeding is off target.
+Excluding ties, there are two possibilities: either the true location is
+the best, in which case it is seeded and the read is correctly mapped; or
+the true location is not the best, in which case it is not seeded and the
+read is mapped incorrectly. The read is mapped to the true location if and
+only if seeding is on target. In practice there are ties, and reads can
+have more than one error, but this is a general trend for MEM seeds.
+
+Table~\ref{table_mem} shows the results of simulations with MEM seeds
+where we dissociate the true from the best location. In each tested
+condition, we record how often the reads are mapped incorrectly because
+seeding was off-target. Importantly, we also record the cases where reads
+are mapped incorrectly despite the fact that seeding was on target. The
+results show that the second case occurs a minor fraction of the time,
+meaning that the probability that seeding is on target is close to the
+probability that the read is mapped to the true location.
 
@@
 
-The trend can reverse for a high number of errors (most of the errors may
-be due to seeding errors), but because these reads are rare, the mapping
-errors are in general not due to failures of the seeding heuristic.
-
 \begin{table}
@@
 \begin{tabular}{ccccc}
-k & N & Prob. seeding error & Prob. other error & \% seeding \\
+k & N & Not seeded & Seeded, not best & Relative error \\
 \hline
-50  &  1  & $1.7 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $25.4\%$ \\
-50  & 10  & $1.6 \cdot 10^{-3}$ & $4.7 \cdot 10^{-3}$ & $24.9\%$ \\
-50  & 100 & $7.9 \cdot 10^{-3}$ & $4.4 \cdot 10^{-2}$ & $15.2\%$ \\
-100 &  1  & $4.4 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $0.2\%$ \\
-100 & 10  & $3.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $0.2\%$ \\
-100 & 100 & $1.0 \cdot 10^{-6}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
+50  &  1  & $1.7 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $3.9$ \\
+50  & 10  & $1.6 \cdot 10^{-3}$ & $4.7 \cdot 10^{-3}$ & $3.9$ \\
+50  & 100 & $7.9 \cdot 10^{-3}$ & $4.4 \cdot 10^{-2}$ & $6.6$ \\
+100 &  1  & $4.4 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $523.7$ \\
+100 & 10  & $3.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $658.1$ \\
+100 & 100 & $1.0 \cdot 10^{-6}$ & $2.3 \cdot 10^{-3}$ & $2301.0$
 \end{tabular}
-\caption{Numerical simulation of mapping errors with skip-9 seeds. Reads
-are mapped incorrectly if either the true location is not in the candidate
-set (not seeded) or it is in the candidate set but is not the best
-location (seeded, not best). Ideal reads were randomly generated 1,000,000
-times with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
-varying the read size $k$ and number $N$ of duplicates as indicated in the
-table and using skip-9 seed of minimum size $\gamma = 19$. Mapping errors
-of both types were recorded, and the fraction of the first to the total
-was computed.}
+\caption{Simulations of mapping errors with MEM seeds. Ideal reads were
+randomly generated 1,000,000 times with sequencing error rate $p = 0.01$,
+divergence rate $\mu = 0.06$, varying the read size $k$ and number $N$ of
+duplicates as indicated, using skip-9 seeds of size $\gamma = 19$. The
+meanings of the columns are the same as in Table~\ref{table_mem}.}
 \label{table_skip}
@@
 
-Tables~\ref{table_exact} and \ref{table_skip} shows the result of numeric
-simulations for exact and skip-9 seeds, respectively. It appears that in
-these cases, the probability that probability of mapping the read to an
-incorrect location (\textit{i.e.} not the true location) is substantially
-higher than the probability that seeding is off target. As a consequence,
-seeding probabilities are not very useful to address true mapping problem
-when exact and skip seeds are used. In this case, it is more accurate to
-use the probabilities that the true location is not the best, because
-they are in general much higher.
-
-In conclusion, the true location problem can be calibrated using the
-methods presented here when using MEM seeds, and by computing the
-probability that the true location is not the best when using exact or
-skip seeds.
-
-\subsection{More realistic error models}
+Table~\ref{table_skip} shows the results of similar simulations with
+skip-9 seeds. In all the tested cases the probability that the true
+location is not the best is substantially higher than the probability that
+seeding is off-target. As a result, the Sesame estimates are very far from
+the probability that the read is not mapped to the true location.
+The results for exact seeds are omitted, but they are qualitatively
+similar to those obtained for skip-9 seeds.
+
+In conclusion, the off-target probabilities computed by Sesame are close
+to the probability that the read is not mapped to the true location when
+using MEM seeds. In contrast, the estimates are very far when using exact
+seeds and skip seeds. In this case, it is more accurate to use an estimate
+of the probability that the true location is not the best, which is
+relatively easy to compute given the error rate of the sequencer $p$, the
+number of duplicates $N$ and their divergence rate $\mu$.
+
+\subsection{Realistic sequencing errors}
 
@@
 current instruments. The errors are typically more frequent at the ends of
-the reads, meaning that seeds may be longer than expected, impacting the
-probability that seeding is off target.
+the reads~\cite{pmid21576222}, meaning that seeds may be longer than
+expected, impacting the probability that seeding is off-target.
 
@@
 for the Illumina HiSeq 2000 and generated random reads of size 50 and 100
-from the human genome. We estimated the probability that seeding is off
-target from simulations and compared the results to calculations performed
-by Sesame (ART suggests error rates $p=0.0052$ and $p=0.0075$ for reads of
-size 50 and 100, respectively).
+from the human genome. We estimated the probability that seeding is
+off-target from simulations and compared the results to calculations
+performed by Sesame (ART suggests error rates $p=0.0052$ and $p=0.0075$
+for reads of size 50 and 100, respectively).
 
 Table~\ref{table_MEM_Illumina} shows the results for MEM seeds of minimum
-size $\gamma=19$. The estimates show some discrepancy but Sesame is
-accurate to within 15\% in these conditions.
-Table~\ref{table_skip_Illumina} shows the same type of comparison for
-skip-9 seeds of size $\gamma=19$. The values are substantially lower than
-for MEM seeds because skip-9 seeds are more sensitive. In these
-conditions, Sesame is accurate to within 35\% but it is important to
-highlight that simulations are unreliable for events with very low
-probability.
+size $\gamma=19$. The discrepancy with Sesame estimates is always within a
+factor 1.2. Table~\ref{table_skip_Illumina} shows the same type of
+comparison for skip-9 seeds of size $\gamma=19$. The values are
+substantially lower than for MEM seeds because skip-9 seeds are more
+sensitive. In these conditions, the discrepancy with the Sesame estimates
+is within a factor 1.5, but it is important to highlight that simulations
+are unreliable for events with very low probability.
 
@@
 \centering
-\begin{tabular}{cccc}
-k & N & Simulation & Sesame \\
+\begin{tabular}{ccccc}
+k & N & Sesame & Simulation & Ratio \\
 \hline
-50  &  1  & $3.8 \cdot 10^{-4}$ & $4.5 \cdot 10^{-4}$ \\
-50  & 10  & $4.4 \cdot 10^{-3}$ & $4.5 \cdot 10^{-3}$ \\
-50  & 100 & $4.2 \cdot 10^{-2}$ & $4.0 \cdot 10^{-2}$ \\
-100 &  1  & $3.5 \cdot 10^{-5}$ & $3.7 \cdot 10^{-5}$ \\
-100 & 10  & $4.1 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
-100 & 100 & $8.6 \cdot 10^{-3}$ & $8.9 \cdot 10^{-3}$ 
+50  &  1  & $4.5 \cdot 10^{-4}$ & $3.8 \cdot 10^{-4}$ & 1.18 \\
+50  & 10  & $4.5 \cdot 10^{-3}$ & $4.4 \cdot 10^{-3}$ & 1.02 \\
+50  & 100 & $4.0 \cdot 10^{-2}$ & $4.2 \cdot 10^{-2}$ & 0.95 \\
+100 &  1  & $3.7 \cdot 10^{-5}$ & $3.5 \cdot 10^{-5}$ & 1.05 \\
+100 & 10  & $4.2 \cdot 10^{-4}$ & $4.1 \cdot 10^{-4}$ & 1.02 \\
+100 & 100 & $8.9 \cdot 10^{-3}$ & $8.6 \cdot 10^{-3}$ & 1.03
 \end{tabular}
-\caption{Numerical simulation of ``seeding'' errors with MEM seeds.
-Realistic reads were randomly generated 1,000,000 times with sequencing
-error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
-$k$ and number $N$ of duplicates as indicated in the table and using
-skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
-were recorded, and the fraction of the first to the total was computed.}
+\caption{Simulations of seeding errors with MEM seeds. Realistic Illumina
+HiSeq 2000 reads were randomly generated 1,000,000 times varying the read
+size $k$ and number $N$ of duplicates as indicated, and fixing the
+divergence rate to $\mu = 0.06$ and using MEM seeds of minimum size
+$\gamma=19$. \textit{Sesame} shows the probability that seeding is
+off-target as computed with Sesame, \textit{Simulation} shows this
+probability estimated by the simulation, and \textit{Ratio} shows their
+ratio.}
 \label{table_MEM_Illumina}
@@
 \centering
-\begin{tabular}{cccc}
-k & N & Simulation & Sesame \\
+\begin{tabular}{ccccc}
+k & N & Sesame & Simulation & Ratio \\
 \hline
-50  &  1  & $4.5 \cdot 10^{-5}$ & $4.5 \cdot 10^{-5}$ \\
-50  & 10  & $3.9 \cdot 10^{-4}$ & $4.2 \cdot 10^{-4}$ \\
-50  & 100 & $1.9 \cdot 10^{-3}$ & $2.1 \cdot 10^{-3}$ \\
-100 &  1  & $< 10^{-6}$         & $< 10^{-6}$         \\
-100 & 10  & $2 \cdot 10^{-6}$   & $3 \cdot 10^{-5}$ \\
-100 & 100 & $6 \cdot 10^{-6}$   & $8 \cdot 10^{-4}$ 
+50  &  1  & $4.5 \cdot 10^{-5}$ & $4.5 \cdot 10^{-5}$ & 1.00 \\
+50  & 10  & $4.2 \cdot 10^{-4}$ & $3.9 \cdot 10^{-4}$ & 1.07 \\
+50  & 100 & $2.1 \cdot 10^{-3}$ & $1.9 \cdot 10^{-3}$ & 1.11 \\
+100 &  1  & $< 10^{-6}$         & $< 10^{-6}$         & NA   \\
+100 & 10  & $3 \cdot 10^{-6}$   & $2 \cdot 10^{-5}$   & 1.50 \\
+100 & 100 & $8 \cdot 10^{-6}$   & $6 \cdot 10^{-4}$   & 1.33
 \end{tabular}
-\caption{Numerical simulation of ``seeding'' errors with skip-9 seeds.
-Realistic reads were randomly generated 1,000,000 times with sequencing
-error rate $p = 0.01$, divergence rate $\mu = 0.06$, varying the read size
-$k$ and number $N$ of duplicates as indicated in the table and using
-skip-9 seed of minimum size $\gamma = 19$. Mapping errors of both types
-were recorded, and the fraction of the first to the total was computed.}
+\caption{Simulations of seeding errors with skip-9 seeds. Realistic Illumina
+HiSeq 2000 reads were randomly generated 1,000,000 times varying the read
+size $k$ and number $N$ of duplicates as indicated, and fixing the
+divergence rate to $\mu = 0.06$ and using MEM seeds of minimum size
+$\gamma=19$. The meanings of the columns are the same as in
+Table~\ref{table_MEM_Illumina}.}
 \label{table_skip_Illumina}
@@
 
-\subsection{Non-specific seeds}
+\subsection{Spurious random hits}
 \label{sec:random_seeds}
@@
 We have assumed that seeds can match only the target or one of its
-duplicates. In practice, seeds can match many random locations of the
-genome and not just the target and the duplicates.
-
-It is important for the validity of the theory that all the sequences that
-have a seed are considered candidate locations (as long as the seed is
-above the threshold $\gamma$). If not, the final estimates are biased. The
-hope is that spurious matches are shorter than $\gamma$ so that they are
-filtered out, but this is not always the case. And if they qualify as a
-seed we need to verify the candidate location.
+duplicates. In reality, seeds have a small chance to match any sequence of
+the genome.
+
+We have assumed throughout that all the genomic sequences that have a
+perfect match for a seed are considered candidate locations. The hope is
+that spurious random matches are shorter than the minimum seed length
+$\gamma$, but this is not always the case. And as mentioned above, every
+candidate location must be verified by an exact alignment algorithm.
 
@@
 To save time during the alignment phase, we can prioritize the seeds so
-that the best hit is likely to be discovered first, allowing us to bail
-out from other alignments as early as possible. We can even bound the
-maximum quality of some candidates so that the alignment can be skipped
-altogether. These considerations depend on the implementation of the
-mapper so we will not develop this further. What matters is that random
-hits do not impose a significant burden on the time needed to verify the
-candidates.
+that the best location is likely to be discovered first, allowing us to
+bail out from other alignments as early as possible. In some cases, it is
+even possible to give an upper bound on the alignment score of a candidate
+location so that the alignment can be skipped altogether. These
+considerations depend on the implementation of the mapper so we will not
+develop this further. What matters is that random hits do not impose a
+significant burden on the time needed to verify the candidates.
 
 The second issue is that random hits can generate false positives. Such
-cases occur when seeding is null (meaning that there is no seed for the
-target nor for its duplicates). Since the hits are not homologous to the
-read, the alignment score is extreme, which makes it easy to detect.
+cases occur when seeding is null (meaning that there is no seed for either
+the target or any of its duplicates). Even when seeding is null, the
+candidate set is in general not empty because of spurious random hits. But
+since such hits are not homologous to the read, the alignment score is
+noticeably low, and this case is easy to detect.
 
@@
 where $q=1-p$ and $\beta$ is the prior probability that the hit is
-random.
-
-The value of $\beta$ has little importance if $p$ is small. Say that
-$p=0.01$ and that a read generates a hit in the genome such that 33
-nucleotides need to be aligned after seeding. If the hit is random there
-is a 99.99\% chance that at least 15 are mismatches. The denominator of
-expression (\ref{eq:bayes}) is approximately $1 +
-4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
+a spurious.
+
+The value of $\beta$ has little importance if $p$ is small. To give an
+example, say that $p=0.01$ and that a read generates a hit in the genome
+such that 33 nucleotides need to be aligned after seeding. If the hit is
+random there is a 99.99\% chance that at least 15 are mismatches. For
+$m=15$, the denominator of expression (\ref{eq:bayes}) is approximately $1
++ 4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
 support for the hypothesis that the hit is random is overwhelming.
 Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
-fewer nucleotides are mismatches. In this case, the denominator is
+fewer nucleotides are mismatches. For $m=4$, the denominator is
 approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
 10^{-9}$, the support for the hypothesis that the hit is not random is
-overwhelming. If $p$ is small, we do not need to worry about the value of
-$\beta$, one can choose for instance $\beta=1/2$ so that the term
-$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
+overwhelming. In summary, if $p$ is small, we do not need to worry about
+the value of $\beta$, one can choose for instance $\beta=1/2$ so that the
+term $(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).
 
@@
 either very small or very large, it has no influence in the first case,
-or it dominates the probability of a false positive in the second case.
-For every read, we can use as confidence score the maximum of this
-probability and of the estimated probability that seeding is off target.
-
-In summary, seeds in random sequences happen relatively frequently, but it
-is possible to minimize their computational burden. Also, they are no
-cause for concern regarding false positives because they can easily be
-detected using Bayes formula, as shown in expression (\ref{eq:bayes}).
-
-\section{The Sesame library}
-\label{sec:sesame}
-
-We implemented the methods and algorithms presented here in an open-source
-C library to compute seeding probabilities. The library is called Sesame
-and is available at \url{https://github.com/gui11aume/sesame}.
-
-\subsection{Main features of Sesame}
-
-Sesame contains functions to compute the seeding probabilities described
-here. All the functions were tested against simulations to ensure that the
-implementation is accurate and the code was checked extensively by static
-analysis and unit testing.
-
-Computating seeding probabilities can take up to a few seconds, even when
-replacing iterative methods by Monte Carlo simulations. This is
-incompatible with the speed requirements of modern mappers, so Sesame has
-an interface for computations that must be performed online. In this mode,
-Sesame uses a type of lazy evaluation where the results are computed only
-the first time, and stored in memory for reuse on subsequent calls.
-
-Storing the results in memory is an efficient strategy because three
-parameters are constant throughout the sequencing run: the minimum seed
-size $\gamma$, the read size $k$ and the error rate of the sequencer $p$
-(and when using skip seeds, the amount of skipping $n$ is also constant).
-Only two parameters depend on the read: the number of duplicates $N$ and
-their divergence rate $\mu$. In this mode, Sesame automatically switches
-to Monte Carlo sampling when $N$ is large to save time. Also, the input
-parameters are ``snapped'' to a predefined grid of set values for $N$ and
-$\mu$, so that few computations are performed and most of the calls are
-actually memory lookups. Sesame can thus be integrated in short read
-mappers without being a bottleneck.
-
-Alternatively, the probabilities of interest can be computed offline,
-saved to disk and loaded at run time. This is particularly useful if the
-sequencing runs follow some standard conditions with a known error rate,
-because the computations can be recycled between runs.
-
-Finally, Sesame also has an offline interface, where seeding probabilities
-are computed exactly as requested by the users, \textit{i.e.} without
-modifying the algorithm or the parameters, and also without storing the
-results in memory.
-
-The Sesame manual, available from the repository, contains additional
-information and explains in detail how to use the library.
-
-\subsection{Using Sesame to compare seeding strategies}
-
-As a tool to compute seeding probabilities, Sesame can be used to compare
-the merits of different strategies. The kind of insight that we can gain
-from such calculations was already showcased in
-Figure~\ref{fig:curves_exact}, Figure~\ref{fig:curves_skip} and
-Figure~\ref{fig:curves_mem}, where the numbers were computed using Sesame.
-
-To further showcase the potential benefits of computing seeding
-probabilities, we use Sesame to compare the default seeding strategies of
-BWA-MEM~\cite{li2013aligning} and Bowtie2~\cite{pmid22388286}. Note that
-both mappers use advanced techniques to refine the seeds, so this
-comparison does not reflect the true performance of the mappers. It is
-nevertheless useful to know the baseline of each strategy. The default of
-BWA-MEM it to use MEM seeds of minimum size 19; that of Bowtie2 is to use
-skip-9 seeds of size 16.
-
-\begin{figure*}[t]
-\begin{center}
-\includegraphics[scale=.41]{mortal_kombat.pdf}
-\end{center}
-\caption{\textbf{Comparing seeding strategies with Sesame.} Left: MEM
-seeds with $\gamma=19$ and $\mu=0.06$. Right: skip-9 seeds with
-$\gamma=16$ and $\mu=0.06$. The probabilities that seeding is off target
-were computed using Sesame. Each curve represents the probability for a
-given number of duplicates ($N$). Estimates using iterative methods (skip
-seeds and MEM sees where $N \leq 20$) were computed to within 1\%
-accuracy. Estimates using Monte Carlo sampling (MEM seeds where $N > 20$)
-were computed as the average of 500 million simulations.
-}
-\label{fig_mortal_kombat}
-\end{figure*}
-
-The probabilities that seeding is off target for different read sizes $k$
-and different number of duplicates $N$ are plotted in
-Figure~\ref{fig_mortal_kombat}. The left panel shows the results for MEM
-seeds and the right panel shows the results for skip seeds. Here the error
-rate $p$ is set to 1\%, close to the specifications the Illumina platform,
-and the divergence rate between duplicates $\mu$ is set to an arbitrary
-value of 6\%.
-
-The behaviors of the two types of seeds are dramatically different. Let us
-start with MEM seeds. For every value of $N$, the probability initially
-increases with the read size, and then drops exponentially. The initial
-increase is a hallmark of MEM seeds; it is due to the fact that duplicates
-can mask the target. Note that the asymptotic decay depends on the number
-of duplicates $N$ because each duplicate can mask the target and thereby
-reduces the probability that it is discovered. Overall, these results show
-that the performance of MEM seeds is poor when the target has more than
-approximately 20 duplicates.
-
-Turning to skip seeds, we see that the curves have a staircase look with a
-drop every 10 nucleotides. This is so because seeding probabilities remain
-unchanged until there is space for another seed of size 16 on the read.
-The curves are otherwise decreasing with a steady exponential trend where
-the asymptotic decay does not depend on the number of duplicates $N$. The
-reason is that duplicates do not prevent the target from being discovered,
-they merely fool the mapper when the target was not found. This property
-makes the asymptotic decay of skip seeds substantially faster that of MEM
-seeds when $N$ is large.
-
-Those properties are self evident in retrospect, but they are not
-necessarily obvious from the definitions of MEM seeds and skip seeds.
-The main limitations of the MEM seeds are best understood by keeping in
-mind that their asymptotic decay bepends on $N$.
-
-Figure~\ref{fig_mortal_kombat} suggests that skip-9 seeds of size 16 are
-just better than MEM seeds of size 19. However, the gain in sensitivity
-comes at the cost of a larger candidate set, slowing down the mapping
-process. To remain competitive, Bowtie2 further filters the candidate set
-using a priority policy. But since some candidates are not checked, the
-probability that seeding is off target is larger than shown in
-Figure~\ref{fig_mortal_kombat}.
-
-\subsection{Key insights about MEM seeds}
-
-At least two key insights about MEM seeds can be gained from
-Figure~\ref{fig_mortal_kombat}. The first is that it is probably not worth
-for a MEM-based mapper to check all the candidate loci when there are more
-than approximately 20 them. The mapper may find the correct location, but
-even if this is the case, the mapping quality will remain low because the
-prior chances of failure were high. A better strategy is to either bail
-out ot not waste time or to switch to a more sensitive seeding method (BWA
-opts for the second and uses a re-seeding policy). It is also important
-to note that this decision should be based on an estimated value of $N$
-and not, for instance, on the size of the seed or some other variable.
-
-A second insight is that for MEM seeds, the off-target rate is always
-above $10^{-3}$ for reads of 50 nucleotides or fewer. Here it is important
-to mention that the value $\mu=0.06$ is not even the worst for reads of
-this size when $p=0.01$ (according to Figure~\ref{fig:curves_mem}, the
-worst value is around 0.02). So if $\mu$ is unknown and one wishes to be
-conservative, it seems that reads of 50 nucleotides cannot be mapped with
-confidence better than $1/1000$.
-
-However, this is not true for the reason that it is practically impossible
-to map an Illumina read of size 50 to the wrong location when $N=0$ (see
-section~\ref{sec:random_seeds}). This means that tweaking the seeding
-method to improve the mapping quality score will have less impact than
-checking whether $N$ is 0 (in other words, whether the target is a unique
-sequence).
-
-These insights suggest that there is a way to make MEM seeds more useful
-for short read mappers. We are currently working on a prototype based on
-these two ideas to increase the speed and the accuracy of short read
-mapping. This requires estimating $N$ and $\mu$ for every read and will be
-described in a separate work.
+and it dominates the probability of a false positive in the second case.
+Checking the value of expression (\ref{eq:bayes}) after mapping reveals
+whether the hit should be discarded and the read should be considered
+unmapped.
+
+In summary, spurious random hits occur regularly, but it is possible to
+minimize their computational burden. Also, they are no cause for concern
+regarding false positives because they can easily be detected using Bayes'
+formula, as shown in expression (\ref{eq:bayes}).
 
@@
 heuristics fail and commit the mapping process to an error. We have also
-implemented the solution as an open source C library to perform the
+implemented the algorithms as an open source C library to perform the
 computations. This fills a knowledge gap to understand and calibrate the
@@
 borrowed from analytic combinatorics~\cite{flajolet2009analytic,
-sedgewick2013introduction}, even though we do not follow the complete
-programme. Constructing generating functions usually serves the purpose of
-finding their singularities in order to approximate the solution. In our
-case, however, the weighted generating functions cannot be computed so
-this strategy is not applicable.
+sedgewick2013introduction, regnier2000unified, nicodeme2002motif}, even
+though we do not follow the complete programme. Constructing generating
+functions usually serves the purpose of finding their singularities in
+order to approximate the solution. In our case, however, the weighted
+generating functions cannot be computed so this strategy is not
+applicable.
 
 To find the probabilities of interest in the absence of a fully specified
-weighted generating function, we compute only the first terms using
-iterative methods as explained in section~\ref{sec:example_skip}, or using
-Monte Carlo sampling as explained in section~\ref{sec:montecarlo}. In this
-regard, the breakthrough is the encoding of reads as segments in different
-alphabets, which is an implicit form of Markov
-imbedding~\cite{fu1994distribution}.
-
-The methods rely on the knowledge of two essential parameters: the number
-of duplicates $N$ and their divergence rate $\mu$. These quantities can be
-estimated efficiently using the FM-index~\cite{ferragina2005indexing} and
-we will publish the detail of this method elsewhere, together with a more
-complete overview of how to use the results presented here to implement
-practical mapping algorithms.
-
-The method is general, but it is important to clearly state the
-assumptions it depends on. First and most importantly, we have ignored
+weighted generating function, we compute only the first terms of the
+Taylor series using iterative methods as explained in
+section~\ref{sec:example_skip}, or using Monte Carlo sampling as explained
+in section~\ref{sec:montecarlo}. In this regard, the breakthrough is the
+encoding of reads as segments in different alphabets, which is an implicit
+form of Markov imbedding~\cite{fu1994distribution}.
+
+Our strategy relies on the knowledge of two essential parameters: the
+number of duplicates $N$ and their divergence rate $\mu$. These quantities
+can be estimated efficiently using the
+FM-index~\cite{ferragina2005indexing} as shown in our related work on a
+prototype mapper based on the concepts developed in this
+article~\cite{zorita2020mapping}.
+
+The method presented here is general, but it is important to clearly state
+the assumptions it depends on. First and most importantly, we have ignored
 insertions and deletions. We assume that the sequencing errors are
@@
 occur among duplicated sequences. This is obviously incorrect, but our
-initial tests with real data suggest that this is a minor impediment (most
-likely because this is only one of several simplifying assumptions).
+initial tests with real data suggest that this is a minor impediment.
 Incorporating insertions and deletions would make the theory intractable,
@@
 
-The second assumption is that all the candidate locations with a seed are
-tested with an exact sequence alignment algorithm, so that the best
-location is always discovered if it is present in the candidate set. This
-is not completely unrealistic, but it is important to note that for plant
-and animal genomes, a single seed may have tens of thousands of hits.
-Therefore, most mappers impose a limit on the number of alignments per
-read, opening the possibility that the best hit is seeded but not aligned.
-This is again a minor impediment, since we can replace the probability
-that the location is false by the proportion of candidates that were not
-aligned.
+The second assumption is that the candidate set consists of all the
+genomic locations that have a perfect match for at least one seed, and
+that all the elements of the candidate set are tested with an exact
+sequence alignment algorithm. This is possible, but it is important to
+note that for plant and animal genomes, a single seed may have tens of
+thousands of hits. Therefore, most mappers impose a limit on the number of
+alignments per read, opening the possibility that the best hit is seeded
+but not aligned. It is clear that in this case, the probability that the
+target is in the candidate set has little to do with the probability that
+the read is mapped correctly. This is again a minor impediment, since the
+probability of mapping such reads correctly is low either way. Mappers can
+have a lower range of confidence score when the candidate set is too
+large to check every sequence.
 
 The third assumption is that all the duplicate sequences evolve
-independently of each other and at the same rate. This is again incorrect,
+independently of each other and at the same rate. This is again incorrect
 because duplication events can happen continuously, creating complex
@@
 
-The last assumption is that seeds can match only the target or one of its
-duplicates. This does not hold in general but we have shown how to deal
-with this possibility \textit{a posteriori} in
-section~\ref{sec:random_seeds}.
-
-We have not treated the case of spaced seeds here. It is possible to find
-a transfer matrix for reads that contain no match for a given spaced seed,
+The last assumption is that seeds can match only the target or its
+duplicates. This does not hold in general because the candidate set
+usually contains spurious random hits, but we have shown how to deal with
+this possibility \textit{a posteriori} in section~\ref{sec:random_seeds}.
+
+We have not treated the case of spaced seeds. It is possible to find a
+transfer matrix for reads that contain no match for a given spaced seed,
 but as the number of don't-care positions increases, the dimensions of the
@@
 buhler2005designing}, with the caveat that such automata can contain a
-very large amount of states. Overall, the transfer matrix method does not
-provide any significant advantage over existing alternatives for spaced
-seeds.
+very large amount of states (equivalent to the rows of the transfer
+matrix). To our knowledge, the transfer matrix method does not provide any
+significant advantage over existing alternatives for spaced seeds.
 
@@
 \ref{sec:illmem}). The first is that the seeding schemes considered here
-have a worst case scenario for a particular value of the divergence
-between duplicates $\mu$. Importantly, the worst value varies between
-different seeding methods, so it is possible in theory to construct
-opportunistic seeding strategies that pick the best method for every read.
-
-Another interesting fact is that skip seeds can have better performance
-than exact seeds in the sense that they can yield lower off-seeding
-probabilities. However, this always comes at the cost of accuracy because
-skipping nulceotides reduces the probability of on-target seeding.
-
-Finally, we also observed that MEM seeds cause a very serious increase in
-off-target seeding compared to exact seeds and skip seeds. This does not
-mean that MEM seeding is a bad strategy (it is usually faster than the
-other methods), but it is a good idea to keep an eye on performance and
-switch method or even skip the read altogether if the chances of
-discovering the target are too low.
+have a worst case scenario for a particular value of $\mu$, the divergence
+between duplicates. Importantly, the worst value varies between different
+seeding methods, so it is possible in theory to construct opportunistic
+seeding strategies that pick the best method for every read, depending on
+the value of $\mu$. Another interesting fact is that skip seeds can have
+better performance than exact seeds in the sense that they can yield lower
+off-seeding probabilities (section~\ref{sec:illskipdual}). However, this
+always comes at the cost of accuracy because skipping nucleotides reduces
+the probability of on-target seeding.
+
+We also observed that MEM seeds have a significantly higher off-target
+seeding rate compared to exact seeds and skip seeds
+(section~\ref{sec:illmem}). This does not mean that MEM seeding is a bad
+strategy (it is usually faster than the other methods), but it is good
+practice to keep an eye on performance and switch methods or even skip the
+read altogether if the chances of discovering the target are too low. We
+also showed in section~\ref{sec:truevsbest} that for MEM seeds, the
+on-target seeding rate is close to the probability that the read is mapped
+to the true location, which was not the case for exact seeds and skip
+seeds. In this regard, the present theory is most useful when using MEM
+seeds.
 
@@
 algorithm~\ref{alg:mcmc} is relatively straightforward, so one may wonder
-why the approach with weighted generating functions is necessary for MEM
-seeds. The only reason is precision. To estimate the frequency of an event
-by Monte Carlo sampling, it must occur at least a few times in the
-simulation. For instance, with 1 million rounds of sampling, frequencies
-around $1/100,000$ or lower cannot be measured accurately. When one is
-interested only in frequent events, it is thus a reasonable strategy. On
-the other hand, for $N < 20$, the probability that MEM seeding is null or
-off-target is relatively small, so we need a method that is accurate in
-this range. Fortunately, the transfer matrix method is fast because the
-dimension of the matrix $\mathring{M}_N(z)$ is small and the computations
-are not prohibitive.
+why the approach with weighted generating functions would be necessary at
+all for MEM seeds. The only reason is precision. To estimate the frequency
+of an event by Monte Carlo sampling, this event must occur at least a few
+times in the simulation. For instance, with 1 million rounds of sampling,
+frequencies around $1/100,000$ or lower cannot be measured accurately.
+When one is interested only in frequent events, it is thus a reasonable
+strategy. On the other hand, for $N < 20$, the probability that MEM
+seeding is null or off-target is relatively small, so we need a method
+that is accurate in this range. Fortunately, the transfer matrix method is
+fast because the dimension of the matrix $\mathring{M}_N(z)$ is small and
+the computations are not prohibitive for small values of $N$.
 
@@
 $p$ is known and constant). For $N > 20$, the iterative method is usually
-too slow and we need to use Monte Carlo sampling instead. The running
-time depends on $p$, on the size of the reads $k$, and on the desired
-number of iterations. Since those are constant throughout the sequencing
-run, the method always takes the same time (around 1-10 seconds for reads
-of size $100$-$200$ nucleotides on modern hardware). The values of $N$ and
-$\mu$ can be binned in intervals so that there are only around $100$ pairs
-for a total cost of a few minutes per run. Considering that mappers can
-process at most $10,000$ reads per second per core, the time of mapping a
-sequencing run of 250 million reads is over 7 hours per core, two orders
-of magnitude larger than the time required to estimate the probabilities
-of error.
+too slow and we need to use Monte Carlo sampling instead. The running time
+depends on $p$, on the size of the reads $k$, and on the desired number of
+iterations. Since those are constant throughout the sequencing run, the
+method always takes the same amount of time (around 1-10 seconds for
+1,000,000 simulations of reads of size around $100$ nucleotides on modern
+hardware). The values of $N$ and $\mu$ can be binned in intervals so that
+there are only around $100$ pairs for a total cost of a few minutes per
+run. Considering that mappers seem to process at most $10,000$ reads per
+second per core, the time of mapping a sequencing run of 250 million reads
+is over 7 hours per core, two orders of magnitude larger than the time
+required to estimate the probabilities of error.
 
@@
 (see algorithm~\ref{alg:mcmc}). The combinatorial stance is that there is
-value in clarity.
+value in the models themselves.
 
@@
 \section{Definitions}
+\label{app_def}
 \setcounter{definition}{0}
@@
 \section{Appendix}
+\label{app_extra}
+
 \subsection{Derivation of expression~(\ref{eq:F})}
@@
 
-In comclusion, the Taylor coefficients $a_0, a_1, a_2, \ldots$ satisfy the
+In conclusion, the Taylor coefficients $a_0, a_1, a_2, \ldots$ satisfy the
 recurrence equation
@@
 start, $j$ nucleotides further down. The weighted generating function of a
-$*$ nucleotide is thus $z$, as it stands of any nucleotide.
+$*$ nucleotide is thus $z$, as it stands for any nucleotide.
 
@@
 We proceed by induction and show that for $k \geq 1$ all the entries of
-$M_n(z)^k$ are mutliples of $z^{k-1}$. For $k = 1$ the statement is
+$M_n(z)^k$ are multiples of $z^{k-1}$. For $k = 1$ the statement is
 trivial and there is nothing to show. Now assume that for some $k > 1$,
